### 2020-03-13 src/s_lda.Rmd
added vehtari paper

---
title: "R Notebook"
output: html_notebook
---

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.

```{r}
plot(cars)
```

### 2020-03-13 src/s_lda.Rmd
added start of abstract

@@
 ---
-title: "R Notebook"
+title: "Fake data"
 output: html_notebook
@@
 
-This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.
+```{r}
+rm(list=ls())
+library(tidyverse)
+```
 
-Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.
+# AR1 examples
+```{r}
+f_ar1 <- function(rho, sigma, L){
+  x <- vector(length = L)
+  x[1] <- rnorm(1, 0, sd=sigma)
+  for(i in 2:L)
+    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
+  return(x)
+}
+plot(f_ar1(1.0, 1, 100), type='l')
 
+f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, 4] <- z
+  return(x)
+}
+
+f_generate_lower_var_four(1/10, 0.3, 1, 1000) %>% 
+  melt() %>% 
+  rename(iter=Var1,
+         chain=Var2) %>% 
+  ggplot(aes(x=value, fill=as.factor(chain))) +
+  geom_histogram(position="identity", alpha=0.3)
+```
+
+Test LDA on classifying chains
 ```{r}
-plot(cars)
+full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
+  melt() %>% 
+  rename(iter=Var1,
+         chain=Var2) %>% 
+  dplyr::select(-iter)
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+require(MASS)
+fitter <- lda(formula = chain ~ ., 
+         data = training_data, 
+         prior = c(1,1,1,1)/4)
+plda <- predict(object=fitter, newdata=testing_data)
+tibble(predicted=plda$class, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
 ```
+
+Try Caret
+```{r}
+library(caret)
+library(gbm)
+# add in extra column of nonsense values since the 
+full_data <- full_data %>% 
+  mutate(chain=as.factor(chain)) %>% 
+  mutate(ones=rnorm(nrow(full_data)))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid)
+# plot(gbmFit1)
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+# plot predictions as fn of value
+testing_data %>% 
+  arrange(value) %>% 
+  mutate(predictions=predict(object=gbmFit1, .)) %>% 
+  ggplot(aes(x=value, y=predictions)) +
+  geom_jitter()
+```

### 2020-03-13 write_up/ms.tex
added start of abstract

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2018

% ready for submission
% \usepackage{neurips_2018}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2018}

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2018}

% to avoid loading the natbib package, add option nonatbib:
%     \usepackage[nonatbib]{neurips_2018}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}

\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}
\newcommand{\Dir}[0]{\textrm{Dirichlet}}
\newcommand{\Ray}[0]{\textrm{Rayleigh}}
\newcommand{\gam}[0]{\textrm{Gamma}}
\newcommand{\dgamma}[0]{\textrm{Gamma}}
\newcommand{\dpoisson}[0]{\textrm{Poisson}}
\newcommand{\dbeta}[0]{\textrm{Beta}}
\newcommand{\dbern}[0]{\textrm{Bernoulli}}
\newcommand{\dunif}[0]{\mathrm{Uniform}}
\newcommand{\dgig}[0]{\textrm{GIG}}
\newcommand{\dnormal}[0]{\mathrm{Normal}}
\newcommand{\dt}[0]{\mathrm{t}}
\newcommand{\igamma}[0]{\textrm{Gamma}^{-1}}
\newcommand{\rayl}[0]{\textrm{Rayleigh}}
\newcommand{\Exp}[0]{\textrm{Exponential}}
\newcommand{\Bet}[0]{\textrm{Beta}}
\newcommand{\GEM}[0]{\textrm{GEM}}
\newcommand{\DP}[0]{\textrm{DP}}
\newcommand{\ESS}[0]{\mathrm{ESS}}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\bbeta}{\bm{\beta}}
\newcommand{\bpi}{\bm{\pi}}
\newcommand{\bomega}{\bm{\omega}}
\newcommand{\bgamma}{\bm{\gamma}}
\newcommand{\blambda}{\bm{\lambda}}
\newcommand{\bphi}{\bm{\phi}}
\newcommand{\btheta}{\bm{\theta}}
\newcommand{\bmu}{\bm{\mu}}
\newcommand{\bb}{\bm{b}}
\newcommand{\bk}{\bm{k}}
\newcommand{\bl}{\bm{l}}
\newcommand{\bn}{\bm{n}}
\newcommand{\bw}{\bm{w}}
\newcommand{\bz}{\bm{z}}
\newcommand{\bx}{\bm{x}}
\newcommand{\bX}{\bm{X}}
\newcommand{\by}{\bm{y}}
\newcommand{\bZ}{\bm{Z}}
\newcommand{\bW}{\bm{W}}
\newcommand{\bS}{\bm{S}}
\newcommand{\bH}{\bm{H}}
\newcommand{\Mult}{\textrm{Multinomial}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\NEW}{\textrm{\tiny new}}
\newcommand{\OLD}{\textrm{\tiny old}}
\newcommand{\sigmat}{\sigma^2}
\newcommand{\IBP}{\textrm{IBP}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Eq}{\mathbb{E}_q}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\cHq}{\mathcal{H}_q}
\newcommand{\test}[1]{\mbox{$#1$}^{\small \mbox{test}}}
\newcommand{\alphaW}{\alpha^{(W)}}
\newcommand{\alphaH}{\alpha^{(H)}}
\newcommand{\betaW}{\beta^{(W)}}
\newcommand{\betaH}{\beta^{(H)}}
\newcommand{\gammaW}{\gamma^{(W)}}
\newcommand{\gammaH}{\gamma^{(H)}}
\newcommand{\gammaT}{\gamma^{(\theta)}}
\newcommand{\rhoW}{\rho^{(W)}}
\newcommand{\rhoH}{\rho^{(H)}}
\newcommand{\rhoT}{\rho^{(\theta)}}
\newcommand{\tauW}{\tau^{(W)}}
\newcommand{\tauH}{\tau^{(H)}}
\newcommand{\tauT}{\tau^{(\theta)}}
\newcommand{\muW}{\hat{W}}
\newcommand{\muH}{\hat{H}}
\newcommand{\Var}{\textrm{Var}}
\newcommand{\LF}{\mathrm{Leapfrog}}

\title{A single MCMC convergence diagnostic across all parameters using machine learning}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
	 Ben Lambert\\
	 MRC Centre for Global Infectious Disease Analysis\\
	 School of Public Health\\
	 Imperial College London\\
	 W2 1PG, United Kingdom\\
	 \texttt{ben.c.lambert@gmail.com} \\
}

\begin{document}
% \nipsfinalcopy is no longer used

\maketitle

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed inference of Bayesian models over the past three decades: taking the theory from theoretical fancy to a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians working in this area often echo that MCMC "should be used with caution, almost as a last resort". The problem is that we are often at the "last resort for interesting models".
\end{abstract}

\section{Introduction}

Chris Holmes quoate

\bibliographystyle{unsrt}
\bibliography{bibliography}

\end{document}

### 2020-03-14 write_up/ms.tex
Update ms.tex

@@
 Despite its importance, MCMC is a notoriously subtle beast.
-Indeed, the words of statisticians working in this area often echo that MCMC "should be used with caution, almost as a last resort".
-The problem is that we are often at the "last resort for interesting models".
+Indeed, the words of statisticians working in this area often echo that MCMC "should be used with caution, almost as a last resort"; the problem is that we are usually at the "last resort" for interesting models.
+Central to the concerns about MCMC is the difficulty in determining whether Markov chains have converged to the posterior distribution.
+Under general conditions, MCMC sampling will asymptotically converge to the posterior distribution but this makes no guarantees about its finite sample performance.
+The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population of chains as a whole: if the within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution (although this says nothing of convergence to the posterior).
+However, collectively, these statistics aim to determine whether or not it is possible to differentiate between the chains on the basis of a particular sample: if you can, then the chains have not mixed and convergence hasn't occurred.
+
+	Over the past 25 years, there have been changes to the particular statistics calculated but they tend to have one thing in common, that these are calculated per parameter and, increasingly, that there is a zoo of statistics that a researcher must keep track of.
 \end{abstract}
@@
 
-Chris Holmes quoate
+Chris Holmes quote reference.
 

### 2020-03-14 src/s_lda.Rmd
added 8 schools example

@@
 library(tidyverse)
+library(reshape2)
+library(rstan)
+library(latex2exp)
+library(caret)
+library(gbm)
+options(mc.cores=4)
+rstan_options(auto_write = TRUE)
 ```
@@
 ```
+
+Perform 1000 replicates
+```{r}
+f_replicate <- function(){
+  full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
+  melt() %>% 
+  rename(iter=Var1,
+         chain=Var2) %>% 
+  dplyr::select(-iter)
+  rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+  training_data <- full_data[rand_samples, ]
+  testing_data <- full_data[-rand_samples, ]
+
+  # add in extra column of nonsense values since the method needs at least 2 vars 
+  full_data <- full_data %>% 
+    mutate(chain=as.factor(chain)) %>% 
+    mutate(ones=rnorm(nrow(full_data)))
+  training_data <- full_data[rand_samples, ]
+  testing_data <- full_data[-rand_samples, ]
+
+  gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+  plda <- predict(object=gbmFit1, newdata=testing_data)
+  a_accuracy <- 
+    tibble(predicted=plda, actual=testing_data$chain) %>%
+    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+    summarise(mean(correct)) %>% 
+    pull()
+  return(a_accuracy)
+}
+
+nreplicates <- 1000
+accuracies <- vector(length = nreplicates)
+for(i in 1:nreplicates){
+  if(i%%100==0)
+    print(i)
+  accuracies[i] <- f_replicate()
+}
+```
+
+Plot results
+```{r}
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  library(grid)
+
+  # Make a list from the ... arguments and plotlist
+  plots <- c(list(...), plotlist)
+
+  numPlots = length(plots)
+
+  # If layout is NULL, then use 'cols' to determine layout
+  if (is.null(layout)) {
+    # Make the panel
+    # ncol: Number of columns of plots
+    # nrow: Number of rows needed, calculated from # of cols
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                    ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+
+ if (numPlots==1) {
+    print(plots[[1]])
+
+  } else {
+    # Set up the page
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+    # Make each plot, in the correct location
+    for (i in 1:numPlots) {
+      # Get the i,j matrix positions of the regions that contain this subplot
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+  }
+}
+
+g1 <- tibble(accuracy=accuracies/0.25, iteration=seq(1, nreplicates, 1)) %>% 
+  ggplot(aes(x=accuracy)) +
+  geom_histogram() +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  ggtitle("A.") +
+  theme(text = element_text(size=14, colour="black")) +
+  xlim(1, NA)
+
+g2 <- testing_data %>% 
+  arrange(value) %>% 
+  mutate(predictions=predict(object=gbmFit1, .)) %>% 
+  ggplot(aes(x=value, y=predictions)) +
+  geom_jitter(height = 0.3) +
+  xlab("Value") +
+  ylab("Chain classification") +
+  theme(text = element_text(size=14, colour="black")) +
+  coord_flip() +
+  ggtitle("B.")
+```
+
+Trying probabilistic prediction
+```{r}
+full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
+  melt() %>% 
+  rename(iter=Var1,
+         chain=Var2) %>% 
+  dplyr::select(-iter)
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+full_data <- full_data %>% 
+    mutate(chain=as.factor(chain)) %>% 
+    mutate(ones=rnorm(nrow(full_data)))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/0.25)
+
+f_generate_all_same_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:4)
+    x[, i] <- f_ar1(rho, sigma, L)
+  return(x)
+}
+
+full_data <- f_generate_all_same_var_four(1/3, 0.3, 1, 1000) %>% 
+  melt() %>% 
+  rename(iter=Var1,
+         chain=Var2) %>% 
+  dplyr::select(-iter)
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+full_data <- full_data %>% 
+    mutate(chain=as.factor(chain)) %>% 
+    mutate(ones=rnorm(nrow(full_data)))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy1 <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy1[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy1)/0.25)
+
+# saveRDS(tibble(unconverged=colMeans(mAccuracy)/0.25,
+#        converged=colMeans(mAccuracy1)/0.25),
+#        "../output/ar1_two_replicates.rds")
+mAR1 <- readRDS("../output/ar1_two_replicates.rds")
+g3 <- 
+  mAR1 %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series") +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+
+pdf("../output/ar1.pdf", width = 12, height = 6)
+multiplot(g1, g2, g3, cols = 3)
+dev.off()
+```
+
+# Eight schools
+```{r}
+source("eight_schools.data.R")
+eight_schools <- list(J=J, y=y, sigma=sigma)
+model_cp <- stan_model("eight_schools_cp.stan")
+model_ncp <- stan_model("eight_schools_ncp.stan")
+```
+
+Run model
+```{r}
+fit_cp <- sampling(
+  model_cp, data = eight_schools,
+  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
+  control = list(adapt_delta = 0.95)
+)
+
+fit_ncp <- sampling(
+  model_ncp, data = eight_schools,
+  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
+  control = list(adapt_delta = 0.95)
+)
+```
+
+# Break each chain into two then make flattened data
+## Centered
+```{r}
+full_data_all <- rstan::extract(fit_cp, permuted=F)
+full_data <- array(dim=c(500, 8, 11))
+k <- 1
+for(i in 1:4){
+  first_half <- full_data_all[1:500, i, ]
+  second_half <- full_data_all[501:1000, i, ]
+  full_data[, k, ] <- first_half
+  k <- k + 1
+  full_data[, k, ] <- second_half
+  k <- k + 1
+}
+
+m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = 12)
+k <- 1
+for(i in 1:8){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:11] <- full_data[j, i, ]
+    m_flattened[k, 12] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V12) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/0.125)
+mAccuracy_centered <- mAccuracy
+```
+
+```{r}
+full_data_all <- rstan::extract(fit_ncp, permuted=F)
+nparams <- 19
+full_data <- array(dim=c(500, 8, nparams))
+k <- 1
+for(i in 1:4){
+  first_half <- full_data_all[1:500, i, ]
+  second_half <- full_data_all[501:1000, i, ]
+  full_data[, k, ] <- first_half
+  k <- k + 1
+  full_data[, k, ] <- second_half
+  k <- k + 1
+}
+
+m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = (nparams+1))
+k <- 1
+for(i in 1:8){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V20) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/0.125)
+mAccuracy_noncentered <- mAccuracy
+```
+
+```{r}
+# saveRDS(tibble(centered=colMeans(mAccuracy_centered)/0.125,
+#        non_centered=colMeans(mAccuracy_noncentered)/0.125),
+#        "../output/eight_schools_replicates.rds")
+tibble(centered=colMeans(mAccuracy_centered)/0.125,
+       non_centered=colMeans(mAccuracy_noncentered)/0.125) %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+```
+
+## Trying with original data (i.e. 4 chains)
+```{r}
+full_data <- rstan::extract(fit_cp, permuted=F)
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 12)
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:11] <- full_data[j, i, ]
+    m_flattened[k, 12] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V12) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_centered <- mAccuracy
+
+full_data <- rstan::extract(fit_ncp, permuted=F)
+k <- 1
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams+1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V20) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_noncentered <- mAccuracy
+```
+
+```{r}
+a_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
+       non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/eight_schools_replicates_4.rds")
+a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+```
+
+Trying original data with index variables
+```{r}
+full_data <- rstan::extract(fit_cp, permuted=F)
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 13)
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:11] <- full_data[j, i, ]
+    if(j < (nrow(full_data)/2))
+      m_flattened[k, 12] <- 0
+    else
+      m_flattened[k, 12] <- 1
+    m_flattened[k, 13] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V13) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_centered <- mAccuracy
+
+full_data <- rstan::extract(fit_ncp, permuted=F)
+k <- 1
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams+2))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    if(j < (nrow(full_data)/2))
+      m_flattened[k, (nparams+1)] <- 0
+    else
+      m_flattened[k, (nparams+1)] <- 1
+    m_flattened[k, (nparams+2)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V21) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_noncentered <- mAccuracy
+```
+
+```{r}
+a_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
+       non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/eight_schools_replicates_4_index.rds")
+a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+```
+
+Three uncertainty distributions
+```{r}
+a_df <- readRDS("../output/eight_schools_replicates.rds")
+g1 <- 
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  xlim(0.78, 1.6) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("A.") +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = "none")
+a_df <- readRDS("../output/eight_schools_replicates_4.rds")
+g2 <- 
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlim(0.78, 1.6) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("B.") +
+  theme(text = element_text(size=14, colour="black"))
+a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
+g3 <- 
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+
+pdf("../output/eight_schools.pdf", width = 12, height = 6)
+multiplot(g1, g2, cols = 2)
+dev.off()
+```

### 2020-03-14 write_up/ms.tex
added 8 schools example

--- 2020-03-13 write_up/ms.tex
@@
 
-\title{A single MCMC convergence diagnostic across all parameters using machine learning}
+\title{MCMC convergence diagnostics with uncertainty using gradient-boosted machines}
 
@@
 \begin{abstract}
-	Markov chain Monte Carlo (MCMC) has transformed inference of Bayesian models over the past three decades: taking the theory from theoretical fancy to a workhorse of applied scientists.
+	Markov chain Monte Carlo (MCMC) has transformed inference of Bayesian models over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
 Despite its importance, MCMC is a notoriously subtle beast.
-Indeed, the words of statisticians working in this area often echo that MCMC "should be used with caution, almost as a last resort".
-The problem is that we are often at the "last resort for interesting models".
+Indeed, the words of statisticians often echo that MCMC "should be used with caution, almost as a last resort"; the problem is that we are usually at the "last resort" for interesting models.
+Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution.
+Under general conditions, MCMC sampling will converge asymptotically to the posterior distribution, but this makes no guarantees about its finite sample performance.
+The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population of chains as a whole: if the within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution.
+Collectively, these summary statistics aim to determine whether or not it is possible to determine the chain that generated a particular sample: if you can, then the chains have not mixed and convergence has not occurred.
+Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them.
+A benefit of this approach is that it provides a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
+Additionally, this measure of convergence is not based on any single characteristic of the sampling distribution (for example, the mean, as for the predominant $\hat{R}$); instead using all the information in the chain.
+The method is straightforward to implement and could be an additional check on convergence for applied analyses.
 \end{abstract}
@@
 
-Chris Holmes quoate
+Chris Holmes quote reference.
+
+Can we determine uncertainty in convergence?
+
+\section{Method}
+
+\section{Results}
+To illustrate the utility of our approach, we use examples taken from \cite{vehtari2019rank}.
+
+\subsection{Heterogeneity in chain variance}\label{sec:heterogeneity}
+We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
+%
+\begin{equation}
+X_t = \rho X_{t-1} + \epsilon_t
+\end{equation}
+%
+where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$.
+Three of the chains share the same $\sigma=\bar{\sigma}$, whereas the other chain has $\sigma=1/3\bar{\sigma}$, so that it has $1/3$ of the (unconditional) variance of the others.
+In each replicate, we simulate 1000 iterations from all four chains and fit a gradient-boosted model to a labelled training set of 2800 randomly-chosen samples, and use it to classify samples according to the chain which generated them in an independent test set comprising the remaining 1200 observations.
+We perform 1000 replicates and, in each case, we report the ratio of accuracy of the predictions to the null model which assigns observations to chains uniformly at random (i.e.
+having an accuracy of 25\%).
+
+In Fig.
+\ref{fig:ar1}A, we show how an example model fit to training data classifies observations by chain according to the sample's value.
+Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
+In Fig.
+\ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates.
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
+	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text.
+See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
+	\label{fig:ar1}
+\end{figure}
+
+Since the gradient-boosted models return a probability simplex for each sample, which indicates the probability it was generated by each chain, we are also able to generate uncertainty in our measure of convergence.
+For each sample, we randomly sample a chain classification from a multinomial distribution with class probabilities given by the outputs from the machine learning model then check whether this matches the true chain identity.
+The accuracy is then given by the percentage of samples where the prediction is correct.
+By repeating this exercise across a number of replicates, we obtain an uncertainty in the accuracy, which we convert to an accuracy ratio as before.
+
+We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the "unconverged data"); and another, where all chains sample from the same distribution (we call this the "converged data").
+In Fig.
+\ref{fig:ar1}C, we show the uncertainty distributions for the accuracy ratio that result from sampling from the underlying fit of a single gradient-boosted model to each dataset.
+For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
+For the converged data, the distribution is centred on 1, indicating convergence.
+
+\subsection{Infinite chain variance}
+Use Cauchy example
+
+\subsection{Hierarchical model: Eight schools}
+In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+
+The model can be parameterised two ways, as described in \cite{vehtari2019rank}.
+The simplest way is referred to as the "centered" parameterisation and exactly mirrors the underlying statistical model,
+%
+\begin{align*}
+\theta_j &\sim \N(\mu, \tau) \\
+y_j &\sim \N(\theta_j, \sigma_j).
+\end{align*}
+%
+The non-centered parameterisation recodes this model in a way which does not affect the joint distribution of $(\tilde{\theta}, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables.
+This can be written as,
+%
+\begin{align*}
+\tilde{\theta}_j &\sim \N(0, 1) \\
+\theta_j &= \mu + \tau \tilde{\theta}_j \\
+y_j &\sim \N(\theta_j, \sigma_j).
+\end{align*}
+%
+In both cases, $\theta_j$ are the treatment effects in the eight schools, and $\mu, \tau$ represent the population mean and standard deviation 
+of the distribution of these effects.
+In the centered parameterization, the $\theta$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}$ are parameters and $\theta$ is a derived quantity.
+
+We first used Stan \cite{carpenter2017stan} to sample from the centered model using 4 chains.
+Like \cite{vehtari2019rank}, we use settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
+We also used Stan with the same algorithm settings to sample from the non-centered model.
+
+To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \cite{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
+Next, we trained the gradient-boosted model on a training set comprising 2800 samples, and used it to classify the causative chain for 1200 independent test samples.
+Following the same approach as in \S\ref{sec:heterogeneity}, we generated uncertainty in $R^*$ for both the centered and non-centered models.
+The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
+In this plot, it is clear that whereas the centered model shows signs of convergence, the non-centered model does not: intuitively, the chains have not sufficiently mixed with themselves nor the others meaning that the causative chain can be predicted with reasonable accuracy from the sample location.
+
+In gradient-boosted tree models (GBMs), it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
+For GBM fitted to the centered model, the most important variable was $\tau$, followed by Stan's $lp$ variable (we also include this in our training and testing data), followed by $\mu$ and $\theta_2$.
+Interestingly, these were slightly differently ordered according to Stan's $\hat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).
+
+In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two.
+The results are shown in Fig.\ref{fig:eight_schools}B.
+In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each sample, meaning that the centered model $R^*$ values are shifted leftwards.
+Despite this, however, the distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged.
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
+	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains; B shows the same but using the 4 original chains.
+In both cases, the plots show 1000 samples.
+See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
+	\label{fig:eight_schools}
+\end{figure}
+
+\section{Discussion}
+Add in index variable into set of predictors.
+Relative insensitivity of GBM to settings (in all examples we used the same hyperparameters)
 

### 2020-03-14 src/s_lda.Rmd
started cauchy example

@@
 
-# Break each chain into two then make flattened data
+## Break each chain into two then make flattened data
 ## Centered
@@
 ```
+
+# Cauchy example
+```{r}
+fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0)
+full_data <- rstan::extract(fit_nom, permuted=F)
+nparams <- 52
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V53) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_nom <- mAccuracy
+```
+
+```{r}
+fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0)
+full_data <- rstan::extract(fit_alt1, permuted=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V153) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_alt_1 <- mAccuracy
+```
+
+```{r}
+a_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
+               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/cauchy.rds")
+a_df <- readRDS("../output/cauchy.rds")
+
+g <-
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = "none")
+g
+
+ggsave("../output/cauchy.pdf", g, width = 8, height = 6)
+```

### 2020-03-14 write_up/ms.tex
started cauchy example

@@
 
-\title{MCMC convergence diagnostics with uncertainty using gradient-boosted machines}
+\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty via gradient-boosted machines}
 
@@
 \begin{abstract}
-	Markov chain Monte Carlo (MCMC) has transformed inference of Bayesian models over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
+	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
 Despite its importance, MCMC is a notoriously subtle beast.
-Indeed, the words of statisticians often echo that MCMC "should be used with caution, almost as a last resort"; the problem is that we are usually at the "last resort" for interesting models.
+Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models.
 Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution.
-Under general conditions, MCMC sampling will converge asymptotically to the posterior distribution, but this makes no guarantees about its finite sample performance.
-The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population of chains as a whole: if the within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution.
-Collectively, these summary statistics aim to determine whether or not it is possible to determine the chain that generated a particular sample: if you can, then the chains have not mixed and convergence has not occurred.
-Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them.
-A benefit of this approach is that it provides a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
-Additionally, this measure of convergence is not based on any single characteristic of the sampling distribution (for example, the mean, as for the predominant $\hat{R}$); instead using all the information in the chain.
-The method is straightforward to implement and could be an additional check on convergence for applied analyses.
+Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
+The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution.
+Collectively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if this predictions are accurate, then the chains have not mixed and convergence has not occurred.
+Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$.
+In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
+Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain.
+By taking advantage of the uncertainty in chain predictions from the fitted machine learning model, our method also produces a measure of uncertainty in $R^*$.
+The method is straightforward to implement, robust to machine learning hyperparameter choice, and could be a complimentary additional check on MCMC convergence for applied analyses.
 \end{abstract}
@@
 \section{Introduction}
+Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
+In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through its many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpinned the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags}, to the more recent, and also popular, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no} that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
 
@@
 
-\subsection{Infinite chain variance}
-Use Cauchy example
+\subsection{Infinite variance: Cauchy example}
+We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
+Like \cite{vehtari2019rank}, we use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
+%
+\begin{equation}
+x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
+\end{equation}
+%
+and monitor convergence of each $x_j$ separately.
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.6\textwidth]{../output/cauchy.pdf}}
+	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations.
+See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
+	\label{fig:cauchy}
+\end{figure}
 

### 2020-03-14 src/s_lda.Rmd
finished cauchy section

@@
 ```
+
+## More iterations
+```{r}
+fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0, iter = 10000,
+                thin = 5)
+full_data <- rstan::extract(fit_nom, permuted=F)
+nparams <- 52
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V53) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_nom <- mAccuracy
+```
+
+```{r}
+fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0, iter = 10000,
+                thin = 5)
+full_data <- rstan::extract(fit_alt1, permuted=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V153) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_alt_1 <- mAccuracy
+```
+
+```{r}
+a_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
+               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/cauchy_more_iter.rds")
+a_df <- readRDS("../output/cauchy.rds")
+
+g1 <-
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = "none") +
+  ggtitle("A.")
+g1
+
+a_df <- readRDS("../output/cauchy_more_iter.rds")
+
+g2 <-
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black")) +
+  ggtitle("B.")
+g2
+
+ggsave("../output/cauchy.pdf", g, width = 8, height = 6)
+```

### 2020-03-14 write_up/ms.tex
finished cauchy section

@@
 In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through its many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpinned the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags}, to the more recent, and also popular, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no} that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
+MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
 
-Chris Holmes quote reference.
+Its importance in applied scientists' toolkits means it is essential the MCMC is used properly and with adequate care.
+A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised depictions of the posterior.
+In spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
+We repeat here a quote from Chris Holmes to this effect,
 
-Can we determine uncertainty in convergence?
+\begin{quote}
+	MCMC is most definitely not a panacea and should be used with caution, almost as a last resort. It is just that often we are at the last resort for interesting models.
+\end{quote}
+
+Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
 
@@
 We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
-Like \cite{vehtari2019rank}, we use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
+Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
 %
@@
 %
-and monitor convergence of each $x_j$ separately.
+and monitor convergence of each $x_j$ separately. We call this parameterisation the ``nominal'' version of this model.
+
+In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \cite{vehtari2019rank},
+%
+\begin{align}
+a_j \sim  \N(0,1), \qquad
+b_j \sim  \text{Gamma}(0.5, 0.5), \qquad
+x_j =  a_j/\sqrt{b_j}.
+\end{align}
+%
+The distribution of the $x$ vector is the same under both parameterisations, although the thin-tailed $(a,b)$ vectors define a higher dimensional posterior that improves sampling efficiency.
+
+In Fig.
+\ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations.
+As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, the samples still contain information about the chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$.
+Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig.
+\ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5).
+In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has been reached.
+Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
 
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.6\textwidth]{../output/cauchy.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
 	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations.

### 2020-03-14 write_up/ms.tex
added more words to intro

@@
 
-Its importance in applied scientists' toolkits means it is essential the MCMC is used properly and with adequate care.
+Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care.
 A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised depictions of the posterior.
-In spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
-We repeat here a quote from Chris Holmes to this effect,
+Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
+Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
+We repeat here a quote from Chris Holmes to this effect \cite{holmes},
 
@@
 
-Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
+The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \cite{lambert2018Student}.
+MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions).
+In the face of only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often practically converge to the posterior in only a (relatively small) finite number of samples.
+
+The predominant diagnostic method for determining whether practical convergence has occurred relies on the idea that the posterior distribution is the unique stationary distribution for an MCMC sampler.
+Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
+Unfortunately, anyone who uses MCMC knows that the method is full of false dawns: the Markov chains can easily become stuck in areas of parameter space and observation over short intervals can mean that the sampling distribution \textit{appears} converged.
+Like furious bees trapped in a room of a house [cite video], MCMC samplers may fail to move because of narrow gaps joining neighbouring areas.
+With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.
 

### 2020-03-16 write_up/ms.tex
first draft of intro

@@
 The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution.
-Collectively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if this predictions are accurate, then the chains have not mixed and convergence has not occurred.
+Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
 Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$.
@@
 Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain.
-By taking advantage of the uncertainty in chain predictions from the fitted machine learning model, our method also produces a measure of uncertainty in $R^*$.
+Since our machine learning method, gradient-boosted regression trees, provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
 The method is straightforward to implement, robust to machine learning hyperparameter choice, and could be a complimentary additional check on MCMC convergence for applied analyses.
@@
 Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through its many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpinned the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags}, to the more recent, and also popular, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no} that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to the more recent algorithms: Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
 MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care.
-A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised depictions of the posterior.
+A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior.
 Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
@@
 
-The predominant diagnostic method for determining whether practical convergence has occurred relies on the idea that the posterior distribution is the unique stationary distribution for an MCMC sampler.
+The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
 Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
-Unfortunately, anyone who uses MCMC knows that the method is full of false dawns: the Markov chains can easily become stuck in areas of parameter space and observation over short intervals can mean that the sampling distribution \textit{appears} converged.
-Like furious bees trapped in a room of a house [cite video], MCMC samplers may fail to move because of narrow gaps joining neighbouring areas.
+Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space and observation over short intervals mean the sampling distribution \textit{appears} converged.
+Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
 With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.
 
-\section{Method}
+To combat this curse of hindsight, running multiple, independent, chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}.
+If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge.
+This approach makes it less likely that convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard recommended practice \cite{lambert2018Student}.
+The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}.
+If these variances are similar, $\hat{R}\approx 1$ and chains are deemed to have mixed.
+Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves \cite{vehtari2019rank}.
+Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}.
+Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.
 
-\section{Results}
+Here, we introduce $R^*$, a new convergence metric.
+This statistic is built on the intuition that, if the chains are mixed, it should not be possible to discern from a sample the chain that generated it.
+Rephrased, it should not be possible to predict the chain that \textit{caused} a sample.
+In this vein, we use a supervised machine learning (ML) approach to measure convergence.
+Specifically, we train a ML model to classify the chain that caused each observation.
+By evaluating the performance of the model on a held-out test set, this provides a new convergence metric.
+To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across all quantiles and parameters -- any of which signal that convergence hasn't occurred.
+This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$ whereas $\hat{R}$ has separate values for each parameter.
+However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
+For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}.
+For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated and is insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
+In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful status of MCMC convergence.
+$R^*$ can straightforwardly be incorporated into existing software libraries and we argue provides a complementary convergence metric alongside more established measures.
+
+The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank}.
+Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
+
+%$R^*$ calculation is enabled in Pints inference software \cite{Clerx2019Pints}.
+
+\section{Method}\label{sec:method}
+
+\section{Results}\label{sec:results}
 To illustrate the utility of our approach, we use examples taken from \cite{vehtari2019rank}.
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text.
+%\begin{figure}[h]
+%	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
+%	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text.
 See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
-	\label{fig:ar1}
-\end{figure}
+%	\label{fig:ar1}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations.
+%\begin{figure}[h]
+%	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
+%	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations.
 See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
-	\label{fig:cauchy}
-\end{figure}
+%	\label{fig:cauchy}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains; B shows the same but using the 4 original chains.
+%\begin{figure}[h]
+%	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
+%	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains; B shows the same but using the 4 original chains.
 In both cases, the plots show 1000 samples.
 See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
-	\label{fig:eight_schools}
-\end{figure}
+%	\label{fig:eight_schools}
+%\end{figure}
 

### 2020-03-16 write_up/ms.tex
added mixture plots for mcmc
draft of method (without algorithms)

@@
 \usepackage{graphicx}
+\usepackage{bbm}
 
@@
 \section{Method}\label{sec:method}
+If Markov chains have not mixed, it is possible to determine to which chain a sample belongs from the sample's value.
+This is possible if there are differences in the sampling distribution for any dimension, $\theta$, in the target distribution (Fig.
+\ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a sample belongs to.
+It  is also possible to predict the chain that generated a given sample if there are differences in the joint distribution of two (or more) dimensions of the target, even if the marginal distributions are the same (Fig.
+\ref{fig:joint}).
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.6\textwidth]{../output/unmixed_1.pdf}}
+	\caption{\textbf{Chain prediction based on marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to right of panel); B shows two chains that have not mixed.}
+	\label{fig:marginal}
+\end{figure}
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.4\textwidth]{../output/unmixed_2.pdf}}
+	\caption{\textbf{Chain prediction based on joint distribution of two parameters.} Here, the marginal distributions of each dimensions are similar but the joint distributions differ.}
+	\label{fig:joint}
+\end{figure}
+
+These two cases, whilst simple, illustrate the basis of our approach.
+To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs.
+By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$ where $N$ is the number of chains.
+By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}.
+In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
+Algorithm X shows how this statistic is constructed.
+
+The ML model we use here is a gradient-boosted regression tree (GBM), which was introduced in \cite{friedman2001greedy}, has been shown to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours.
+Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s Caret package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}.
+The data for each chain has dimensions: $X\in \mathbb{R}^{SK}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
+We split each chain's samples into training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
+The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for models with 1000 samples) and its predictive performance was insensitive to its hyperparameters.
+In what follows, our hyperparameter settings were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations would be the minimum for each node and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).
+
+A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$.
+For each sample, $s$, in our testing set, our GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a single chain prediction, $c^{\{s\}}$.
+By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^s=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
+We repeat this sampling process for each sample in the testing set, generating $\boldsymbol{a}=(a^1,a^2,...,a^{N_\text{test}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$.
+We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
+Algorithm Y shows how to calculate this distribution for a given set of MCMC samples.
 
 \section{Results}\label{sec:results}
-To illustrate the utility of our approach, we use examples taken from \cite{vehtari2019rank}.
+To illustrate the utility of our approach, we use examples from \cite{vehtari2019rank} with a few additional cases that further illustrate the utility of $R^*$.
 
@@
 For the converged data, the distribution is centred on 1, indicating convergence.
+
+\subsection{Poor convergence in joint distribution}
+Simple AR1 example followed by 250 MVT normal.
 

### 2020-03-16 write_up/ms.tex
added algorithm blocks

@@
 
+Add another panel for the joint case
+
 \begin{figure}[h]
@@
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
-Algorithm X shows how this statistic is constructed.
+Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.
+
+\begin{algorithm}[tb]
+	\caption{$R^*$ calculation algorithm}
+	\label{alg:R_star}
+	\begin{algorithmic}
+		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
+		\FOR{$m=1$ to $N$}
+		\STATE Sample $k^0\sim\N(0, M)$.
+		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
+		\FOR{$i=1$ to $L$}
+		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
+		\ENDFOR
+		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
+		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
+		\ENDFOR
+		\STATE
+		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
+		\STATE Set $\tilde k \leftarrow k +
+		(\epsilon/2)\nabla_\theta\cL(\theta)$.
+		\STATE Set $\tilde \theta \leftarrow \theta +
+		\epsilon M^{-1}k'$.
+		\STATE Set $\tilde k \leftarrow \tilde k +
+		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
+		\RETURN $\tilde \theta, \tilde k$.
+	\end{algorithmic}
+\end{algorithm}
 
@@
 A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$.
-For each sample, $s$, in our testing set, our GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a single chain prediction, $c^{\{s\}}$.
-By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^s=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this sampling process for each sample in the testing set, generating $\boldsymbol{a}=(a^1,a^2,...,a^{N_\text{test}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$.
+Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words.
+For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
+By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
+We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$.
 We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
-Algorithm Y shows how to calculate this distribution for a given set of MCMC samples.
+
+\begin{algorithm}[tb]
+	\caption{Procedure to generate a single $R^*$ sample}
+	\label{alg:R_star_uncertainty}
+	\begin{algorithmic}
+		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
+		\FOR{$m=1$ to $N$}
+		\STATE Sample $k^0\sim\N(0, M)$.
+		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
+		\FOR{$i=1$ to $L$}
+		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
+		\ENDFOR
+		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
+		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
+		\ENDFOR
+		\STATE
+		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
+		\STATE Set $\tilde k \leftarrow k +
+		(\epsilon/2)\nabla_\theta\cL(\theta)$.
+		\STATE Set $\tilde \theta \leftarrow \theta +
+		\epsilon M^{-1}k'$.
+		\STATE Set $\tilde k \leftarrow \tilde k +
+		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
+		\RETURN $\tilde \theta, \tilde k$.
+	\end{algorithmic}
+\end{algorithm}
 

### 2020-03-16 src/s_examples.Rmd
started discussion

--- 2020-03-14 src/s_lda.Rmd
@@
 ---
-title: "Fake data"
+title: "Examples in R* paper"
+author: "Ben Lambert"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: html_notebook
@@
 
-# AR1 examples
+# AR1 example
+Functions to generate simulated data
 ```{r}
@@
 }
-plot(f_ar1(1.0, 1, 100), type='l')
 
@@
 
-Test LDA on classifying chains
+Use gbm model in Caret to predict chain identity
 ```{r}
+library(caret)
+library(gbm)
+
+# generate data
 full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
@@
   dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-require(MASS)
-fitter <- lda(formula = chain ~ ., 
-         data = training_data, 
-         prior = c(1,1,1,1)/4)
-plda <- predict(object=fitter, newdata=testing_data)
-tibble(predicted=plda$class, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-```
 
-Try Caret
-```{r}
-library(caret)
-library(gbm)
-# add in extra column of nonsense values since the 
+# add in extra column of nonsense values since gbm needs 2+ vars
 full_data <- full_data %>% 
@@
   mutate(ones=rnorm(nrow(full_data)))
+
+# create training/testing data
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
 training_data <- full_data[rand_samples, ]
@@
   summarise(mean(correct))
-varImp(gbmFit1)
 
@@
   ggplot(aes(x=value, y=predictions)) +
-  geom_jitter()
+  geom_jitter() +
+  coord_flip()
 ```
@@
 
-Trying probabilistic prediction
+Generate R* distribution based on 1000 iterations
 ```{r}
@@
 
-nsim <- 1000
+nsim <- 10
 mAccuracy <- matrix(nrow = nrow(plda),
@@
 }
-hist(colMeans(mAccuracy)/0.25)
 
@@
 
-# saveRDS(tibble(unconverged=colMeans(mAccuracy)/0.25,
-#        converged=colMeans(mAccuracy1)/0.25),
-#        "../output/ar1_two_replicates.rds")
-mAR1 <- readRDS("../output/ar1_two_replicates.rds")
+mAR1 <- tibble(unconverged=colMeans(mAccuracy)/0.25,
+               converged=colMeans(mAccuracy1)/0.25)
 g3 <- 
@@
 
+# Multivariate normal
+
+Fit model in Stan
+```{r}
+N <- 250
+A <- rWishart(1, 250, diag(N))[,,1]
+model <- stan_model("mvt_250.stan")
+fit <- sampling(model, data=list(N=N, A=A), iter=200, chains=4)
+print(fit)
+```
+
+```{r}
+full_data <- rstan::extract(fit, permuted=F)
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:251] <- full_data[j, i, ]
+    m_flattened[k, 252] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V252) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+```
+
+# Cauchy example
+
+Generate R* distribution from nominal Cauchy fit
+```{r}
+nsim <- 1000
+fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0)
+full_data <- rstan::extract(fit_nom, permuted=F)
+nparams <- 52
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V53) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+mAccuracy_nom <- mAccuracy
+```
+
+Generate R* distribution from alternative Cauchy parameterisation
+```{r}
+fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0)
+full_data <- rstan::extract(fit_alt1, permuted=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V153) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+mAccuracy_alt_1 <- mAccuracy
+```
+
+```{r}
+a_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
+               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
+
+g <-
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = "none")
+g
+
+ggsave("../output/cauchy.pdf", g, width = 8, height = 6)
+```
+
+## More iterations
+Nominal model with 10000 iterations and thinning by 10
+```{r}
+fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0, iter = 10000,
+                thin = 5)
+full_data <- rstan::extract(fit_nom, permuted=F)
+nparams <- 52
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V53) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+mAccuracy_nom <- mAccuracy
+```
+
+Alternative model with 10000 iterations and thinning by 10
+```{r}
+fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0, iter = 10000,
+                thin = 5)
+full_data <- rstan::extract(fit_alt1, permuted=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams+1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V153) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+mAccuracy_alt_1 <- mAccuracy
+```
+
+Plot
+```{r}
+b_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
+               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
+
+g1 <-
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = "none") +
+  ggtitle("A.")
+g1
+
+g2 <-
+  b_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black")) +
+  ggtitle("B.")
+g2
+
+pdf("../output/cauchy.pdf", width = 12, height = 6)
+multiplot(g1, g2, cols = 2)
+dev.off()
+```
+
 # Eight schools
@@
 
-Run model
+Run models
 ```{r}
@@
 
-# Break each chain into two then make flattened data
-## Centered
+## Break each chain into two then make flattened data
+Centered
 ```{r}
@@
 }
-hist(colMeans(mAccuracy)/0.125)
 mAccuracy_centered <- mAccuracy
@@
 
+Noncentered
 ```{r}
@@
 }
-hist(colMeans(mAccuracy)/0.125)
 mAccuracy_noncentered <- mAccuracy
@@
 
+Store both
 ```{r}
-# saveRDS(tibble(centered=colMeans(mAccuracy_centered)/0.125,
-#        non_centered=colMeans(mAccuracy_noncentered)/0.125),
-#        "../output/eight_schools_replicates.rds")
-tibble(centered=colMeans(mAccuracy_centered)/0.125,
-       non_centered=colMeans(mAccuracy_noncentered)/0.125) %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
+c_df <- tibble(centered=colMeans(mAccuracy_centered)/0.125,
+       non_centered=colMeans(mAccuracy_noncentered)/0.125)
 ```
@@
 }
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
 mAccuracy_centered <- mAccuracy
@@
 }
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
 mAccuracy_noncentered <- mAccuracy
@@
 ```{r}
-a_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
+d_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
        non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
-saveRDS(a_df, "../output/eight_schools_replicates_4.rds")
-a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
 ```
 
-Trying original data with index variables
+Plot two R* distributions
 ```{r}
-full_data <- rstan::extract(fit_cp, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 13)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    if(j < (nrow(full_data)/2))
-      m_flattened[k, 12] <- 0
-    else
-      m_flattened[k, 12] <- 1
-    m_flattened[k, 13] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V13) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-mAccuracy_centered <- mAccuracy
-
-full_data <- rstan::extract(fit_ncp, permuted=F)
-k <- 1
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams+2))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    if(j < (nrow(full_data)/2))
-      m_flattened[k, (nparams+1)] <- 0
-    else
-      m_flattened[k, (nparams+1)] <- 1
-    m_flattened[k, (nparams+2)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V21) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-mAccuracy_noncentered <- mAccuracy
-```
-
-```{r}
-a_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
-       non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
-saveRDS(a_df, "../output/eight_schools_replicates_4_index.rds")
-a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
-```
-
-Three uncertainty distributions
-```{r}
-a_df <- readRDS("../output/eight_schools_replicates.rds")
 g1 <- 
-  a_df %>% 
+  c_df %>% 
   melt() %>% 
@@
         legend.position = "none")
-a_df <- readRDS("../output/eight_schools_replicates_4.rds")
 g2 <- 
-  a_df %>% 
+  d_df %>% 
   melt() %>% 
@@
 a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
-g3 <- 
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
 

### 2020-03-16 write_up/ms.tex
started discussion

@@
 
-Add another panel for the joint case
-
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.6\textwidth]{../output/unmixed_1.pdf}}
-	\caption{\textbf{Chain prediction based on marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to right of panel); B shows two chains that have not mixed.}
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
+	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right of panel); B shows two chains that have not mixed.}
 	\label{fig:marginal}
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.4\textwidth]{../output/unmixed_2.pdf}}
-	\caption{\textbf{Chain prediction based on joint distribution of two parameters.} Here, the marginal distributions of each dimensions are similar but the joint distributions differ.}
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_2.pdf}}
+	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
 	\label{fig:joint}
@@
 To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs.
-By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$ where $N$ is the number of chains.
+By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
 By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}.
@@
 \begin{algorithm}[tb]
-	\caption{$R^*$ calculation algorithm}
+	\caption{$R^*$ calculation}
 	\label{alg:R_star}
 	\begin{algorithmic}
-		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
+		\STATE Given $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}, S_\text{test}$:
 		\FOR{$m=1$ to $N$}
-		\STATE Sample $k^0\sim\N(0, M)$.
-		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
-		\FOR{$i=1$ to $L$}
-		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
+		\STATE Partition, $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
 		\ENDFOR
-		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
-		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
+		\STATE $X_\text{train} = \{X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train}\}$
+		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
+		\FOR{$s=1$ to $S_\text{test}$}
+		\STATE Predict chain id, $c^\{s\} = \text{ML}(x^\{s\}|X_\text{train})$
+		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
 		\ENDFOR
-		\STATE
-		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
-		\STATE Set $\tilde k \leftarrow k +
-		(\epsilon/2)\nabla_\theta\cL(\theta)$.
-		\STATE Set $\tilde \theta \leftarrow \theta +
-		\epsilon M^{-1}k'$.
-		\STATE Set $\tilde k \leftarrow \tilde k +
-		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
-		\RETURN $\tilde \theta, \tilde k$.
+		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
+		\STATE Calculate ratio to null model, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
+		\RETURN $R^*$
 	\end{algorithmic}
@@
 The data for each chain has dimensions: $X\in \mathbb{R}^{SK}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
-We split each chain's samples into training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
+We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
 The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for models with 1000 samples) and its predictive performance was insensitive to its hyperparameters.
@@
 \begin{algorithm}[tb]
-	\caption{Procedure to generate a single $R^*$ sample}
+	\caption{Procedure to generate $I$ samples of $R^*$}
 	\label{alg:R_star_uncertainty}
 	\begin{algorithmic}
-		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
-		\FOR{$m=1$ to $N$}
-		\STATE Sample $k^0\sim\N(0, M)$.
-		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
-		\FOR{$i=1$ to $L$}
-		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
+		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
+		\STATE model, $\text{ML}(x^{\{s\}}|X_\text{train})\rightarrow(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$:
+		\FOR{$i=1$ to $I$}
+		\FOR{$s=1$ to $S_\text{test}$}
+		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^\{s\}|X_\text{train})$
+		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
+		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
 		\ENDFOR
-		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
-		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
+		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
+		\STATE Calculate ratio to null model, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
 		\ENDFOR
-		\STATE
-		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
-		\STATE Set $\tilde k \leftarrow k +
-		(\epsilon/2)\nabla_\theta\cL(\theta)$.
-		\STATE Set $\tilde \theta \leftarrow \theta +
-		\epsilon M^{-1}k'$.
-		\STATE Set $\tilde k \leftarrow \tilde k +
-		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
-		\RETURN $\tilde \theta, \tilde k$.
+		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
 	\end{algorithmic}
@@
 
-\subsection{Heterogeneity in chain variance}\label{sec:heterogeneity}
+\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
 We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
@@
 
-\subsection{Poor convergence in joint distribution}
+\subsection{Lack of convergence of the joint distribution: multivariate normal model}\label{sec:multivariate_normal}
 Simple AR1 example followed by 250 MVT normal.
@@
 
-\subsection{Hierarchical model: Eight schools}
+\subsection{Hierarchical model: Eight schools model}
 In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
@@
 \section{Discussion}
+If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated that particular observation.
+Based on this observation, used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in the samples.
+By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
+By extracting the chain identity probabilities from each prediction in the test set, this allowed us to generate an uncertainty distribution for $R^*$.
+Across a range of previously published examples, $R^*$ was shown to be predictive of whether or not chains had converged.
+
+The predominant methods for diagnosing MCMC convergence rely heavily on looking for differences in the marginal distributions of each dimension between chains.
+$R^*$ naturally includes this information in building a model capable of predicting the chain which generated each sample.
+It also naturally includes information about the joint distribution across all dimensions of the target.
+Since both the marginal and joint distributions will be approximately equal across converged chains, any measure of convergence should account for both of these aspects.
+Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise.
+This is not a slight on existing measures, more that we argue this illustrates the complementarity of $R^*$ to them.
+
 Add in index variable into set of predictors.
+
 Relative insensitivity of GBM to settings (in all examples we used the same hyperparameters)

### 2020-03-16 write_up/ms.tex
draft of discussion

@@
 
-\subsection{Hierarchical model: Eight schools model}
+\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
 In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
@@
 \section{Discussion}
-If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated that particular observation.
-Based on this observation, used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in the samples.
+If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated it.
+Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in the samples.
 By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
-By extracting the chain identity probabilities from each prediction in the test set, this allowed us to generate an uncertainty distribution for $R^*$.
+By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
 Across a range of previously published examples, $R^*$ was shown to be predictive of whether or not chains had converged.
@@
 Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise.
-This is not a slight on existing measures, more that we argue this illustrates the complementarity of $R^*$ to them.
+This is not a sleight on existing measures, more that we argue this illustrates the complementarity of $R^*$ to them.
 
-Add in index variable into set of predictors.
+When first starting to develop $R^*$, it was unclear to us whether \textit{any} metric aligned with a supervised ML method would be overly sensitive to its hyperparameters.
+With gradient-boosted regression tree models, we found this not to be the case.
+Indeed, after a little experimenting, we found that a single set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples.
+To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and we recommend that applied users consider including this step in their workflow.
 
-Relative insensitivity of GBM to settings (in all examples we used the same hyperparameters)
+Many implementations of $\hat{R}$ suggest splitting chains in two before calculating it.
+In the Eight Schools example in \S\ref{sec:eight_shools}, we trial this before calculating $R^*$, and find that this approach leads to more accurate chain prediction.
+As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that $R^*$ is maximised.
+Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' categorical indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
+If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information may boost $R^*$.
+
+MCMC enables inference across a large range of models encountered across the social, biological and physical sciences.
+Its ease of implementation, however, masks important underlying fragilities in the method.
+Namely, that unless the chains have converged to a truly stationary distribution, the samples generated are not faithful depictions of the posterior.
+In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far.
+$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a complementary measure of convergence to existing metrics.
 

### 2020-03-16 write_up/ms.tex
draft of text waiting for mvt normal case

@@
 \subsection{Lack of convergence of the joint distribution: multivariate normal model}\label{sec:multivariate_normal}
-Simple AR1 example followed by 250 MVT normal.
+In this example, we use the 250-dimensional multivariate normal target, where its precision matrix is generated from a Wishart distribution \cite{hoffman2014no}.
+We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions.
+We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for X iterations: at this point, all the $\hat{R}<1.01$ and the effective sample size for each dimension exceeded 400 (recommendations from \cite{vehtari2019rank}).
+Using these samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig.
+X.
 

### 2020-03-17 src/s_examples.Rmd
ms nearly there -- just needs to have mvt normal section completed

@@
 # Multivariate normal
-
-Fit model in Stan
+Fit model in Stan with 10000 iterations
 ```{r}
 N <- 250
-A <- rWishart(1, 250, diag(N))[,,1]
+# A <- rWishart(1, 250, diag(N))[,,1]
 model <- stan_model("mvt_250.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=200, chains=4)
+fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
 print(fit)
-```
+saveRDS(fit, "../output/mvt_fit_10000.rds")
 
-```{r}
 full_data <- rstan::extract(fit, permuted=F)
@@
 varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- tibble(samples_5000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+```
+
+With 400 iterations
+```{r}
+fit <- sampling(model, data=list(N=N, A=A), iter=400, chains=4)
+print(fit)
+saveRDS(fit, "../output/mvt_fit_400.rds")
+
+full_data <- rstan::extract(fit, permuted=F)
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:251] <- full_data[j, i, ]
+    m_flattened[k, 252] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V252) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- a_df %>%
+  mutate(samples_400=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+```
+
+```{r}
+model <- stan_model("mvt_250_ncp.stan")
+fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
+print(fit)
+saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
+
+full_data <- rstan::extract(fit, permuted=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V502) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- a_df %>%
+  mutate(samples_10000_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+
+saveRDS(a_df, "../output/mvt_fit_all_accuracy.rds")
+
+g <- 
+  a_df %>% 
+  select(samples_400, samples_10000_ncp) %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8, bins=20) +
+  scale_fill_grey("# of samples", labels=c("400", "2000", "5000")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"))
+# ggsave("../output/mvt_three.pdf", g, width = 8, height = 6)
 ```

### 2020-03-17 write_up/ms.tex
ms nearly there -- just needs to have mvt normal section completed

@@
 
-\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty via gradient-boosted machines}
+\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines}
 
@@
 In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain.
-Since our machine learning method, gradient-boosted regression trees, provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
-The method is straightforward to implement, robust to machine learning hyperparameter choice, and could be a complimentary additional check on MCMC convergence for applied analyses.
+Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely uncovered by existing approaches.
+Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
+The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
 \end{abstract}
@@
 Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic} that underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to the more recent algorithms: Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, that Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
 MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 
-To combat this curse of hindsight, running multiple, independent, chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}.
+To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}.
 If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge.
-This approach makes it less likely that convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard recommended practice \cite{lambert2018Student}.
+This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \cite{lambert2018Student}.
 The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}.
-If these variances are similar, $\hat{R}\approx 1$ and chains are deemed to have mixed.
-Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves \cite{vehtari2019rank}.
+If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed.
+Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \cite{vehtari2019rank}.
 Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}.
@@
 Here, we introduce $R^*$, a new convergence metric.
-This statistic is built on the intuition that, if the chains are mixed, it should not be possible to discern from a sample the chain that generated it.
+This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample the chain that generated it.
 Rephrased, it should not be possible to predict the chain that \textit{caused} a sample.
 In this vein, we use a supervised machine learning (ML) approach to measure convergence.
-Specifically, we train a ML model to classify the chain that caused each observation.
+Specifically, we train a ML model to classify the chain that generated each observation.
 By evaluating the performance of the model on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across all quantiles and parameters -- any of which signal that convergence hasn't occurred.
-This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$ whereas $\hat{R}$ has separate values for each parameter.
+To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches.
+This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter.
 However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
 For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated and is insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
-In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful status of MCMC convergence.
-$R^*$ can straightforwardly be incorporated into existing software libraries and we argue provides a complementary convergence metric alongside more established measures.
+For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test).
+It is also insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
+In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
+$R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
-	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right of panel); B shows two chains that have not mixed.}
+	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
 	\label{fig:marginal}
@@
 	\begin{algorithmic}
-		\STATE Given $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}, S_\text{test}$:
+		\STATE Given chain-wise samples from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
 		\FOR{$m=1$ to $N$}
-		\STATE Partition, $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
+		\STATE Create train and test sets by random-sampling (w/o replacement), $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
 		\ENDFOR
-		\STATE $X_\text{train} = \{X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train}\}$
+		\STATE Stack $X_\text{train} = (X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train})^T$
+		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
 		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
 		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Predict chain id, $c^\{s\} = \text{ML}(x^\{s\}|X_\text{train})$
+		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
+		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
 		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
@@
 		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
+		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
 		\RETURN $R^*$
@@
 
-The ML model we use here is a gradient-boosted regression tree (GBM), which was introduced in \cite{friedman2001greedy}, has been shown to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours.
-Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s Caret package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}.
-The data for each chain has dimensions: $X\in \mathbb{R}^{SK}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
+The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours.
+Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}.
+The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
 We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
-The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for models with 1000 samples) and its predictive performance was insensitive to its hyperparameters.
-In what follows, our hyperparameter settings were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations would be the minimum for each node and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).
+The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for models with 1000 samples), and its predictive performance was insensitive to its hyperparameters.
+In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).
 
@@
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=\bar{a}$ estimate for iteration $i$.
+We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N/\bar{a}$ estimate for iteration $i$.
 We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
@@
 		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
-		\STATE model, $\text{ML}(x^{\{s\}}|X_\text{train})\rightarrow(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$:
+		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
 		\FOR{$i=1$ to $I$}
 		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^\{s\}|X_\text{train})$
+		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
+		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
 		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
@@
 		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
+		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
 		\ENDFOR
@@
 \section{Results}\label{sec:results}
-To illustrate the utility of our approach, we use examples from \cite{vehtari2019rank} with a few additional cases that further illustrate the utility of $R^*$.
+To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} with one additional example: the multivariate normal model described in \cite{hoffman2014no}.
 
@@
 %
-where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$.
-Three of the chains share the same $\sigma=\bar{\sigma}$, whereas the other chain has $\sigma=1/3\bar{\sigma}$, so that it has $1/3$ of the (unconditional) variance of the others.
-In each replicate, we simulate 1000 iterations from all four chains and fit a gradient-boosted model to a labelled training set of 2800 randomly-chosen samples, and use it to classify samples according to the chain which generated them in an independent test set comprising the remaining 1200 observations.
-We perform 1000 replicates and, in each case, we report the ratio of accuracy of the predictions to the null model which assigns observations to chains uniformly at random (i.e.
-having an accuracy of 25\%).
+where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$ and $t=1,2,...,1000$.
+Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others.
+In Fig.
+\ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a sample's value.
+Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
 
-In Fig.
-\ref{fig:ar1}A, we show how an example model fit to training data classifies observations by chain according to the sample's value.
-Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
+To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e.
+with one chain with a lower variance).
+We then fit a gradient-boosted model to a labelled training set of 2800 randomly chosen samples.
+The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them.
+For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}.
 In Fig.
@@
 
-%\begin{figure}[h]
-%	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-%	\caption{\textbf{Autoregressive example.} A shows how the gradient-boosted model's classifications vary according to the sample's value for an example model fit; B shows the ratio of predictive accuracy to the null model accuracy across 1000 replicates; C shows the uncertainty in accuracy ratio for two series described in the text.
-See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
-%	\label{fig:ar1}
-%\end{figure}
+\begin{figure}[h]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the sample's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:ar1}
+\end{figure}
 
-Since the gradient-boosted models return a probability simplex for each sample, which indicates the probability it was generated by each chain, we are also able to generate uncertainty in our measure of convergence.
-For each sample, we randomly sample a chain classification from a multinomial distribution with class probabilities given by the outputs from the machine learning model then check whether this matches the true chain identity.
-The accuracy is then given by the percentage of samples where the prediction is correct.
-By repeating this exercise across a number of replicates, we obtain an uncertainty in the accuracy, which we convert to an accuracy ratio as before.
-
-We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the "unconverged data"); and another, where all chains sample from the same distribution (we call this the "converged data").
+Since GBMs return a probability simplex for each sample, which, for each chain, indicates the probability that a sample was generated by it, we can also generate a measure of uncertainty in $R^*$.
+Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done.
+We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data'); and another, where all chains sample from the same distribution (we call this the ``converged'' data).
 In Fig.
-\ref{fig:ar1}C, we show the uncertainty distributions for the accuracy ratio that result from sampling from the underlying fit of a single gradient-boosted model to each dataset.
+\ref{fig:ar1}C, we show the $R^*$ distributions in each case.
 For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
@@
 We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for X iterations: at this point, all the $\hat{R}<1.01$ and the effective sample size for each dimension exceeded 400 (recommendations from \cite{vehtari2019rank}).
-Using these samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig.
-X.
+We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for three different iteration counts: 400, 2000 and 5000.
+For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig.
+\ref{fig:mvt}A.
+From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
+Even in the 5000 iteration case, the $R^*$ distribution remains stubbornly shifted somewhat rightwards of $R^*=1$ (coincidently, the $\hat{R}$ values ...).
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.6\textwidth]{../output/mvt_three.pdf}}
+	\caption{\textbf{Multivariate normal example.} Plot shows $R^*$ distributions obtained for three different MCMC sample sizes.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:mvt}
+\end{figure}
 
@@
 %
-and monitor convergence of each $x_j$ separately. We call this parameterisation the ``nominal'' version of this model.
+We call this parameterisation the ``nominal'' version of this model.
 
@@
 \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, the samples still contain information about the chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
 The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$.
@@
 
-%\begin{figure}[h]
-%	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
-%	\caption{\textbf{Cauchy example.} A shows 1000 samples from the $R^*$ distribution generated under two model parameterisations.
-See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
-%	\label{fig:cauchy}
-%\end{figure}
+\begin{figure}[h]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
+	\caption{\textbf{Cauchy example.} A shows 1000 MCMC samples from the $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC samples used (which are then thinned by a factor of 5).
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:cauchy}
+\end{figure}
 
 \subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
 
 The model can be parameterised two ways, as described in \cite{vehtari2019rank}.
-The simplest way is referred to as the "centered" parameterisation and exactly mirrors the underlying statistical model,
+The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
 %
@@
 %
-The non-centered parameterisation recodes this model in a way which does not affect the joint distribution of $(\tilde{\theta}, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables.
+The ``non-centered'' parameterisation recodes this model in a way which does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
 This can be written as,
@@
 %
-In both cases, $\theta_j$ are the treatment effects in the eight schools, and $\mu, \tau$ represent the population mean and standard deviation 
+In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
 of the distribution of these effects.
-In the centered parameterization, the $\theta$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}$ are parameters and $\theta$ is a derived quantity.
+In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
 
@@
 To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \cite{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Next, we trained the gradient-boosted model on a training set comprising 2800 samples, and used it to classify the causative chain for 1200 independent test samples.
-Following the same approach as in \S\ref{sec:heterogeneity}, we generated uncertainty in $R^*$ for both the centered and non-centered models.
+Next, we trained the GBM on a training set comprising 2800 samples, and used it to classify the causative chain for 1200 independent test samples.
+Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
 The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
@@
 
-In gradient-boosted tree models (GBMs), it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
+In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
 For GBM fitted to the centered model, the most important variable was $\tau$, followed by Stan's $lp$ variable (we also include this in our training and testing data), followed by $\mu$ and $\theta_2$.
-Interestingly, these were slightly differently ordered according to Stan's $\hat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).
+Interestingly, these were ordered differently according to Stan's $\hat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).
 
@@
 
-%\begin{figure}[h]
-%	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-%	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains; B shows the same but using the 4 original chains.
-In both cases, the plots show 1000 samples.
-See the R markdown file in the SOM for hyperparameters and code for fitting the gradient-boosted model.}
-%	\label{fig:eight_schools}
-%\end{figure}
+\begin{figure}[h]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
+	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
+In both cases, the plots show 1000 $R^*$ samples.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:eight_schools}
+\end{figure}
 
@@
 If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated it.
-Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in the samples.
+Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in samples.
 By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
@@
 
-The predominant methods for diagnosing MCMC convergence rely heavily on looking for differences in the marginal distributions of each dimension between chains.
+The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension.
 $R^*$ naturally includes this information in building a model capable of predicting the chain which generated each sample.
 It also naturally includes information about the joint distribution across all dimensions of the target.
-Since both the marginal and joint distributions will be approximately equal across converged chains, any measure of convergence should account for both of these aspects.
+Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects.
 Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise.
@@
 In the Eight Schools example in \S\ref{sec:eight_shools}, we trial this before calculating $R^*$, and find that this approach leads to more accurate chain prediction.
-As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that $R^*$ is maximised.
-Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' categorical indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
-If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information may boost $R^*$.
+As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
+Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
+If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.
 
@@
 In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far.
-$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a complementary measure of convergence to existing metrics.
+$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
 

### 2020-03-17 src/s_examples.Rmd
updated fig 3

@@
 
-g1 <- tibble(accuracy=accuracies/0.25, iteration=seq(1, nreplicates, 1)) %>% 
+g2 <- tibble(accuracy=accuracies/0.25, iteration=seq(1, nreplicates, 1)) %>% 
   ggplot(aes(x=accuracy)) +
@@
   ylab("Count") +
-  ggtitle("A.") +
+  ggtitle("B.") +
   theme(text = element_text(size=14, colour="black")) +
@@
 
-g2 <- testing_data %>% 
+g1 <- testing_data %>% 
   arrange(value) %>% 
@@
   coord_flip() +
-  ggtitle("B.")
+  ggtitle("A.")
 ```
@@
 
-nsim <- 10
+nsim <- 1000
 mAccuracy <- matrix(nrow = nrow(plda),
@@
 
+nsim <- 1000
 mAccuracy1 <- matrix(nrow = nrow(plda),

### 2020-03-17 src/s_examples.Rmd
added mvt normal example

@@
 library(gbm)
+library(mvtnorm)
 options(mc.cores=4)
@@
 
-# Multivariate normal
+# Bivariate normal
+```{r}
+rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
+  return(rmvnorm(n, c(mux, muy),
+                 matrix(c(sigmax^2, sigmax * sigmay * rho,
+                          sigmax * sigmay * rho, sigmay^2),
+                        ncol = 2)))
+}
+# calling function
+n <- 1000
+chain1 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
+  as.data.frame() %>% 
+  mutate(chain=1)
+chain2 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
+  as.data.frame() %>% 
+  mutate(chain=2)
+chain3 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
+  as.data.frame() %>% 
+  mutate(chain=3)
+chain4 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.9) %>% 
+  as.data.frame() %>% 
+  mutate(chain=4)
+
+chains_stacked <- rbind.data.frame(chain1, chain2, chain3, chain4) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- chains_stacked
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+```
+
+# 250-dimensional multivariate normal
 Fit model in Stan with 10000 iterations
@@
 N <- 250
-# A <- rWishart(1, 250, diag(N))[,,1]
+A <- rWishart(1, 250, diag(N))[,,1]
+saveRDS(A, "../output/A_matrix.rds")
 model <- stan_model("mvt_250.stan")
@@
 print(fit)
-saveRDS(fit, "../output/mvt_fit_10000.rds")
-
+# saveRDS(fit, "../output/mvt_fit_10000.rds")
+fit <- readRDS("../output/mvt_fit_10000.rds")
 full_data <- rstan::extract(fit, permuted=F)
@@
 
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- tibble(samples_5000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- tibble(samples_10000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+
+importance <- 100-varImp(gbmFit1)$importance$Overall
+rhat <- summary(fit)$summary[, 10]
+ess <- summary(fit)$summary[, 9]
+plot(importance, rhat)
+plot(importance, ess)
+b_df <- tibble(importance, ess, rhat, lp=c(rep(0, 250), 1))
+sum(rhat<1.01)
+sum(ess>400)
+gc <- 
+  b_df %>%
+  ggplot(aes(importance, ess)) +
+  geom_jitter(width=1, aes(shape=as.factor(lp))) +
+  geom_smooth(span=1) +
+  xlab("Variable importance") +
+  ylab("ESS") +
+  ggtitle("C.") +
+  theme(legend.position = "none")
+
+gb <- b_df %>%
+  ggplot(aes(importance, rhat)) +
+  geom_jitter(width = 1, aes(shape=as.factor(lp))) +
+  geom_smooth(span=1) +
+  xlab("Variable importance") +
+  ylab(TeX("$\\hat{R}$")) +
+  ggtitle("B.") +
+  theme(legend.position = "none")
 ```
@@
 print(fit)
-saveRDS(fit, "../output/mvt_fit_400.rds")
+# saveRDS(fit, "../output/mvt_fit_400.rds")
+fit <- readRDS("../output/mvt_fit_400.rds")
 
@@
 print(fit)
-saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
+# saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
+fit <- readRDS("../output/mvt_ncp_fit_10000.rds")
 
@@
 
-g <- 
+ga <- 
   a_df %>% 
-  select(samples_400, samples_10000_ncp) %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
+  melt() %>% 
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
   geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of samples", labels=c("400", "2000", "5000")) +
+  scale_fill_grey("# of samples", labels=c("400", "10000, cp", "10000, ncp")) +
   xlab(TeX("$R*$")) +
@@
   geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"))
-# ggsave("../output/mvt_three.pdf", g, width = 8, height = 6)
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8)) +
+  ggtitle("A.")
+
+pdf("../output/mvt_three.pdf", width = 12, height = 8)
+multiplot(ga, gb, gc, cols = 3)
+dev.off()
 ```

### 2020-03-17 write_up/ms.tex
added mvt normal example

@@
 Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
-The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this may indicate that the chains have converged to a stationary distribution.
+The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a stationary distribution.
 Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
@@
 In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely uncovered by existing approaches.
+Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely overlooked by existing approaches.
 Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
@@
 Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a wide community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
 MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test).
-It is also insensitive to the GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
-In addition, since the outputs from GBMs are class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
+It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
+In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
 $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
@@
 
-\subsection{Lack of convergence of the joint distribution: multivariate normal model}\label{sec:multivariate_normal}
-In this example, we use the 250-dimensional multivariate normal target, where its precision matrix is generated from a Wishart distribution \cite{hoffman2014no}.
+\subsection{Diagnosing unconvergence in joint distribution: bivariate normal model}\label{sec:bivariate_normal}
+
+\subsection{Slow joint distribution convergence: 250-dimensional multivariate normal model}\label{sec:multivariate_normal}
+In this example, we use a 250-dimensional multivariate normal target, where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}.
 We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for three different iteration counts: 400, 2000 and 5000.
+We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (thinned by a factor of 5).
+First, we used the ``centered'' parameterisation of this model, which is of the form,
+%
+\begin{equation}
+\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
+\end{equation}
+%
+where $\boldsymbol{x}\in\mathbb{R}^{250}$.
 For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig.
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 5000 iteration case, the $R^*$ distribution remains stubbornly shifted somewhat rightwards of $R^*=1$ (coincidently, the $\hat{R}$ values ...).
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \cite{vehtari2019rank}.
+
+Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it.
+This model has the form,
+%
+\begin{align}
+\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
+\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
+z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
+\end{align}
+%
+where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$.
+Fig.
+\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is centred on $R^*=1$ (mean $R^*=1.00$) indicating convergence.
+This was also reiterated by the fact that $\hat{R}<1.01$ and $\text{ESS}>400$ for all parameters.
+
+In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
+We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan.
+For a GBM fitted to the centered model, we plot in Fig.
+\ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
+From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$).
+In Fig.
+\ref{fig:mvt}C, we plot variable importance versus ESS, which exhibit a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
+Together, these plots illustrate that the GBM's variable importance measure provides complementary information to $\hat{R}$ and ESS.
 
 \begin{figure}[h]
-	\centerline{\includegraphics[width=0.6\textwidth]{../output/mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example.} Plot shows $R^*$ distributions obtained for three different MCMC sample sizes.
+	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
+	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
+In plots B and C, horizontal jitter was added to the points.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
-For GBM fitted to the centered model, the most important variable was $\tau$, followed by Stan's $lp$ variable (we also include this in our training and testing data), followed by $\mu$ and $\theta_2$.
+As in \S\ref{sec:multivariate_normal}, we calculated a measure of variable importance to understand which variables contributed most to the predictive power of the model.
+For the GBM fitted to the centered model, the most important variable was $\tau$, followed by Stan's $lp$ variable (we also include this in our training and testing data), followed by $\mu$ and $\theta_2$.
 Interestingly, these were ordered differently according to Stan's $\hat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).

### 2020-03-17 src/s_examples.Rmd
first draft of paper

@@
 rstan_options(auto_write = TRUE)
+source("monitornew.R")
 ```
@@
   as.data.frame() %>% 
-  mutate(chain=1)
+  mutate(chain=1,
+         iter=seq_along(chain))
 chain2 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
   as.data.frame() %>% 
-  mutate(chain=2)
+  mutate(chain=2,
+         iter=seq_along(chain))
 chain3 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
   as.data.frame() %>% 
-  mutate(chain=3)
+  mutate(chain=3,
+         iter=seq_along(chain))
 chain4 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.9) %>% 
   as.data.frame() %>% 
-  mutate(chain=4)
+  mutate(chain=4,
+         iter=seq_along(chain))
 
@@
   mutate(chain=as.factor(chain))
-
-full_data <- chains_stacked
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
+chains_stacked1 <- chains_stacked %>% 
+  select(-iter)
+
+full_data <- chains_stacked1
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+# 
+# # use Vehtari et al. script to calculate Rhat
+# chains_wide1 <- chains_stacked %>% 
+#   pivot_wider(values_from = c("V1"), names_from = c("chain"), id_cols = "iter")
+# rhat_rfun(as.matrix(chains_wide1[, 2:5]))
+# chains_wide2 <- chains_stacked %>% 
+#   pivot_wider(values_from = c("V2"), names_from = c("chain"), id_cols = "iter")
+# rhat_rfun(as.matrix(chains_wide2[, 2:5]))
+# 
+# # Use Vehtari et al. to calculate ess
+# ess_rfun(as.matrix(chains_wide1[, 2:5]))
+# ess_rfun(as.matrix(chains_wide2[, 2:5]))
+# ess_bulk(as.matrix(chains_wide1[, 2:5]))
+# ess_bulk(as.matrix(chains_wide2[, 2:5]))
+# ess_tail(as.matrix(chains_wide1[, 2:5]))
+# ess_tail(as.matrix(chains_wide2[, 2:5]))
+# ess_quantile(as.matrix(chains_wide1[, 2:5]), 0.95)
+# ess_tail(as.matrix(chains_wide2[, 2:5]))
+
+# Try monitor functionality -- split chains first
+half_iter <- max(chains_stacked$iter) / 2
+m_samples <- array(dim = c(half_iter,
+                           2*n_distinct(chains_stacked$chain),
+                           2))
+k <- 1
+chains_stacked_first <- chains_stacked %>% 
+  filter(iter<=half_iter)
+chains_stacked_second <- chains_stacked %>% 
+  filter(iter>half_iter)
+
+for(i in 1:half_iter){
+  temp <- chains_stacked_first %>% 
+    filter(iter==i)
+  for(j in 1:4){
+    temp1 <- temp %>%
+      filter(chain==j)
+    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
+    k <- k + 1
+  }
+}
+
+for(i in 1:half_iter){
+  temp <- chains_stacked_second %>% 
+    filter(iter==(i + 500))
+  for(j in 5:8){
+    temp1 <- temp %>%
+      filter(chain==(j - 4))
+    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
+    k <- k + 1
+  }
+}
+
+g <- 
+  qplot(4*colMeans(mAccuracy)) +
+  xlab(TeX("$R*$")) +
+  xlim(0.9, NA) +
+  geom_vline(xintercept = 1, linetype=2) +
+  ylab("Count") +
+  theme(text=element_text(size=14, colour="black"))
+ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
+mon <- monitor(m_samples)
+
 ```

### 2020-03-17 write_up/ms.tex
first draft of paper

@@
 		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
+		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N \bar{a}$
 		\RETURN $R^*$
@@
 		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
+		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} = N\bar{a}$
 		\ENDFOR
@@
 \section{Results}\label{sec:results}
-To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} with one additional example: the multivariate normal model described in \cite{hoffman2014no}.
+To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} two additional examples based on a multivariate normal model.
 
@@
 
-\subsection{Diagnosing unconvergence in joint distribution: bivariate normal model}\label{sec:bivariate_normal}
+\subsection{Diagnosing convergence in joint distributions: bivariate normal model}\label{sec:bivariate_normal}
+To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density.
+In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
+By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.
+
+First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
+In all cases, the various quantities were calculated based on chains split into halves.
+For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.
+
+Next, we estimate the $R^*$ distribution which is shown in Fig.
+\ref{fig:bivariate}.
+The mean of this distribution is 1.18 and all the samples are above 1 -- indicating that the sampling distribution has not converged.
+By taking account of all the information in the chains,
+
+\begin{figure}[h]
+	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
+	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 samples. See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:bivariate}
+\end{figure}
 

### 2020-03-17 src/monitornew.R

@@ -1,0 +1,741 @@
+fft_next_good_size <- function(N) {
+  # Find the optimal next size for the FFT so that
+  # a minimum number of zeros are padded.
+  if (N <= 2)
+    return(2)
+  while (TRUE) {
+    m = N
+    while ((m %% 2) == 0) m = m / 2
+    while ((m %% 3) == 0) m = m / 3
+    while ((m %% 5) == 0) m = m / 5
+    if (m <= 1)
+      return(N)
+    N = N + 1
+  }
+}
+
+#' Autocovariance estimates
+#'
+#' Compute autocovariance estimates for every lag for the specified
+#' input sequence using a fast Fourier transform approach. Estimate
+#' for lag t is scaled by N-t.
+#'
+#' @param y A numeric vector forming a sequence of values.
+#'
+#' @return A numeric vector of autocovariances at every lag (scaled by N-lag).
+autocovariance <- function(y) {
+  N <- length(y)
+  M <- fft_next_good_size(N)
+  Mt2 <- 2 * M
+  yc <- y - mean(y)
+  yc <- c(yc, rep.int(0, Mt2 - N))
+  transform <- fft(yc)
+  ac <- fft(Conj(transform) * transform, inverse = TRUE)
+  # use "biased" estimate as recommended by Geyer (1992)
+  ac <- Re(ac)[1:N] / (N^2 * 2)
+  ac
+}
+
+#' Autocorrelation estimates
+#'
+#' Compute autocorrelation estimates for every lag for the specified
+#' input sequence using a fast Fourier transform approach. Estimate
+#' for lag t is scaled by N-t.
+#'
+#' @param y A numeric vector forming a sequence of values.
+#'
+#' @return A numeric vector of autocorrelations at every lag (scaled by N-lag).
+autocorrelation <- function(y) {
+  ac <- autocovariance(y)
+  ac <- ac / ac[1]
+}
+
+#' Rank normalization
+#'
+#' Compute rank normalization for a numeric array. First replace each
+#' value by its rank. Average rank for ties are used to conserve the
+#' number of unique values of discrete quantities. Second, normalize
+#' ranks via the inverse normal transformation.
+#'
+#' @param x A numeric array of values.
+#'
+#' @return A numeric array of rank normalized values with the same
+#'     size as input.
+z_scale <- function(x) {
[…]

### 2020-03-17 write_up/ms_arxiv.tex
added arxiv version of the paper

--- 2020-03-17 write_up/ms.tex
@@
 \documentclass{article}
-
+\pdfoutput=1
 % if you need to pass options to natbib, use, e.g.:
@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage[final]{nips_2018}
+%\usepackage[final]{nips_2018}
 
 % to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2018}
+% \usepackage[nonatbib]{nips_2018}
 
@@
 Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
-Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call our convergence measure $R^*$.
+Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call this convergence measure $R^*$.
 In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
@@
 In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
-MCMC methods are currently the most effective methods for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
+MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
 
@@
 Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
-Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space and observation over short intervals mean the sampling distribution \textit{appears} converged.
+Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged.
 Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
@@
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_1.pdf}}
 	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
@@
 \begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_2.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_2.pdf}}
 	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
@@
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N/\bar{a}$ estimate for iteration $i$.
+We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
 We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
@@
 		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} = N\bar{a}$
+		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N\bar{a}$
 		\ENDFOR
@@
 \section{Results}\label{sec:results}
-To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} two additional examples based on a multivariate normal model.
+To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} and two additional examples based on a multivariate normal model.
 
@@
 In Fig.
-\ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates.
+\ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates indicating that the chains have not converged.
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
 	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the sample's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text.
@@
 
-\subsection{Diagnosing convergence in joint distributions: bivariate normal model}\label{sec:bivariate_normal}
+\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
 To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density.
-In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
+In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
 By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.
@@
 \ref{fig:bivariate}.
-The mean of this distribution is 1.18 and all the samples are above 1 -- indicating that the sampling distribution has not converged.
-By taking account of all the information in the chains,
+The mean of this distribution is 1.18, and all the samples are above 1, indicating that the sampling distribution has not converged.
+By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals.
+As such, we see a role for $R^*$ as an additional convergence metric.
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.6\textwidth]{bivariate.pdf}}
 	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 samples. See the R markdown file on the Github repository for code to reproduce this example.}
@@
 
-\subsection{Slow joint distribution convergence: 250-dimensional multivariate normal model}\label{sec:multivariate_normal}
-In this example, we use a 250-dimensional multivariate normal target, where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}.
+We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}.
 We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (thinned by a factor of 5).
+We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
 First, we used the ``centered'' parameterisation of this model, which is of the form,
@@
 In Fig.
-\ref{fig:mvt}C, we plot variable importance versus ESS, which exhibit a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
-Together, these plots illustrate that the GBM's variable importance measure provides complementary information to $\hat{R}$ and ESS.
+\ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
+Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{mvt_three.pdf}}
+	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
 In plots B and C, horizontal jitter was added to the points.
@@
 \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5).
-In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has been reached.
+In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
 Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
 	\caption{\textbf{Cauchy example.} A shows 1000 MCMC samples from the $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC samples used (which are then thinned by a factor of 5).
@@
 \begin{align*}
-\theta_j &\sim \N(\mu, \tau) \\
+\theta_j &\sim \N(\mu, \tau), \\
 y_j &\sim \N(\theta_j, \sigma_j).
@@
 \begin{align*}
-\tilde{\theta}_j &\sim \N(0, 1) \\
-\theta_j &= \mu + \tau \tilde{\theta}_j \\
+\tilde{\theta}_j &\sim \N(0, 1), \\
+\theta_j &= \mu + \tau \tilde{\theta}_j,\\
 y_j &\sim \N(\theta_j, \sigma_j).
@@
 We first used Stan \cite{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we use settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
+Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
 We also used Stan with the same algorithm settings to sample from the non-centered model.
@@
 In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each sample, meaning that the centered model $R^*$ values are shifted leftwards.
-Despite this, however, the distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged.
+Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears converged.
 
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools.pdf}}
 	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
-In both cases, the plots show 1000 $R^*$ samples.
+In all cases, the plots show 1000 $R^*$ samples.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 \section{Discussion}
-If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated it.
+If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a sample it should be impossible to discern which chain generated it.
 Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in samples.
@@
 
-The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension.
-$R^*$ naturally includes this information in building a model capable of predicting the chain which generated each sample.
+The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target.
+$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each sample.
 It also naturally includes information about the joint distribution across all dimensions of the target.
@@
 Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise.
-This is not a sleight on existing measures, more that we argue this illustrates the complementarity of $R^*$ to them.
+This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
 
@@
 
-MCMC enables inference across a large range of models encountered across the social, biological and physical sciences.
+MCMC enables inference across a wide range of models encountered across the social, biological and physical sciences.
 Its ease of implementation, however, masks important underlying fragilities in the method.

### 2020-03-19 src/s_examples.Rmd
added Aki's suggestions about exploring wider data

@@
 source("monitornew.R")
+
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
 ```
@@
 ```{r}
-library(caret)
-library(gbm)
-
 # generate data
@@
 
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
 gbmFit1 <- train(chain ~ ., data = training_data, 
@@
 dev.off()
+```
+
+## Aki Vehtari's suggestion: try measure for similar numbers of iterations to posterior
+Centered
+```{r}
+N <- 250
+A <- readRDS("../output/A_matrix.rds")
+model <- stan_model("mvt_250.stan")
+fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
+saveRDS(fit, "../output/mvt_fit_500.rds")
+fit <- readRDS("../output/mvt_fit_500.rds")
+
+full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:251] <- full_data[j, i, ]
+    m_flattened[k, 252] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V252) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+varImp(gbmFit1)
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- tibble(samples_cp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+
+mon <- monitor(fit)
+mean(mon$Rhat>1.01)
+mean(mon$Bulk_ESS<400)
+mean(mon$Tail_ESS<400)
+```
+
+Non-centered
+```{r}
+A <- readRDS("../output/A_matrix.rds")
+model <- stan_model("mvt_250_ncp.stan")
+fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
+saveRDS(fit, "../output/mvt_ncp_fit_500.rds")
+
+full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V502) %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- a_df %>%
+  mutate(samples_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/mvt_500_all.rds")
+g <- 
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8, bins=20) +
+  scale_fill_grey("Model", labels=c("centered", "noncentered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+ggsave("../output/wide_data.pdf", g, width = 12, height = 8)
+```
+
+# Wide dataset - standard normal
+```{r}
+N <- 10000
+A <- diag(nrow=N)
+model <- stan_model("mvt_wide.stan")
+fit <- sampling(model, data=list(N=N), iter=400, chains=4)
+saveRDS(fit, "../output/mvt_wide_400.rds")
+
+full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame()
+colnames(m_flattened)[nparams + 1] <- "chain"
+m_flattened <- m_flattened %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- tibble(samples_400=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
+mon <- monitor(fit)
+mean(mon$Rhat>1.01)
+mean(mon$Bulk_ESS<400)
+mean(mon$Tail_ESS<400)
+```
+
+```{r}
+N <- 10000
+A <- diag(nrow=N)
+model <- stan_model("mvt_wide.stan")
+fit <- sampling(model, data=list(N=N), iter=1000, chains=4)
+saveRDS(fit, "../output/mvt_wide_1000.rds")
+
+full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
+nparams <- dim(full_data)[3]
+m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
+k <- 1
+for(i in 1:4){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:nparams] <- full_data[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame()
+colnames(m_flattened)[nparams + 1] <- "chain"
+m_flattened <- m_flattened %>% 
+  mutate(chain=as.factor(chain))
+n_distinct(m_flattened$chain)
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
+plda <- predict(object=gbmFit1, newdata=testing_data)
+tibble(predicted=plda, actual=testing_data$chain) %>% 
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct))
+
+plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+
+nsim <- 1000
+mAccuracy <- matrix(nrow = nrow(plda),
+                    ncol = nsim)
+for(i in 1:nsim){
+  for(j in 1:nrow(plda)){
+    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
+  }
+}
+
+a_df <- a_df %>%
+  tibble(samples_1000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+saveRDS(a_df, "../output/mvt_wide_summaries.rds")
+mon <- monitor(fit)
+mean(mon$Rhat>1.01)
+mean(mon$Bulk_ESS<400)
+mean(mon$Tail_ESS<400)
+
+g <- 
+  a_df %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8, bins=20) +
+  scale_fill_grey("Model", labels=c("400", "1000")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = c(0.8, 0.8))
+ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
 ```

### 2020-03-19 src/s_examples.Rmd
Update s_examples.Rmd

@@
 saveRDS(fit, "../output/mvt_ncp_fit_500.rds")
+fit <- readRDS("../output/mvt_ncp_fit_500.rds")
 
@@
 ggsave("../output/wide_data.pdf", g, width = 12, height = 8)
+
+mon <- monitor(fit)
+mean(mon$Rhat>1.01)
+mean(mon$Bulk_ESS<400)
+mean(mon$Tail_ESS<400)
 ```

### 2020-03-21 src/s_trends.Rmd
started trend experiments

---
title: "Examining the effect of a trend on R*"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(mvtnorm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")

caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)

f_fit_gbm_predict <- function(r){
  rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
  training_data <- r[rand_samples, ]
  testing_data <- r[-rand_samples, ]

  gbmFit1 <- train(chain ~ ., data = training_data, 
                 method = "gbm",
                 trControl = trainControl(method = 'none'), 
                    tuneGrid = caretGrid, verbose=FALSE)
  plda <- predict(object=gbmFit1, newdata=testing_data)
  a_accuracy <- 
    tibble(predicted=plda, actual=testing_data$chain) %>%
    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
    summarise(mean(correct)) %>% 
    pull()
  return(a_accuracy)
}
```

# Consistent trend across all chains
Repeating analysis from Vehtari et a. appendix B
```{r}
conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  trend = c(0, 0.25, 0.5, 0.75, 1),
  rep = 1:10
)
res <- vector("list", nrow(conds))
chains = 4
for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  trend <- conds[i, "trend"]
  rep <- conds[i, "rep"]
  r <- array(rnorm(iters * chains), c(iters, chains))
  r <- r + seq(-trend, trend, length.out = iters)
  rs <- as.data.frame(monitor_extra(r))

  # R* calculation
  r1 <- r %>% 
    as.data.frame() %>% 
    melt(id.vars = NULL) %>% 
    mutate(chain=as.factor(variable)) %>% 
    select(-variable) %>% 
  mutate(ones=rnorm(length(chain)))
  a_accuracy <- f_fit_gbm_predict(r1)

  # Split R* calculation
  half_rows <- nrow(r) / 2
  r2 <- cbind(r[1:half_rows, ], r[(half_rows+1):nrow(r), ]) %>% 
    as.data.frame() %>% 
    melt(id.vars = NULL) %>% 
    mutate(chain=as.factor(variable)) %>% 
    select(-variable) %>% 
  mutate(ones=rnorm(length(chain)))
  b_accuracy <- f_fit_gbm_predict(r2)

  res[[i]] <- cbind(iters, trend, rep, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
}
res <- bind_rows(res)
saveRDS(res, "../output/experiment_trends_all_dim.rds")
```

```{r}
res <- readRDS("../output/experiment_trends_all_dim.rds")
ggplot(data = res, aes(y = Rstar_split, x = trend)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1) + 
  ggtitle('Split-R*')
```

# Trend in a single dimension
```{r}
conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  trend = c(0, 0.25, 0.5, 0.75, 1),
  rep = 1:10,
  dimension = c(1, 5, 10, 20)
)
res <- vector("list", nrow(conds))
chains = 4

for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  trend <- conds[i, "trend"]
  rep <- conds[i, "rep"]
  dimension <- conds[i, "dimension"]
  r <- array(rnorm(iters * chains * dimension), c(iters, chains, dimension))
  r[, , 1] <- r[, , 1] + seq(-trend, trend, length.out = iters)
  rs <- as.data.frame(monitor_extra(r))
  r_2d <- matrix(nrow = iters * chains, ncol=(dimension + 2))
  k <- 1
  for(i in 1:iters){
    for(j in 1:chains){
      r_2d[k, ] <- c(i, j, r[i, j, ])
      k <- k + 1
    }
  }
  r_2d <- r_2d %>%
    as.data.frame()
  colnames(r_2d)[1:2] <- c("iter", "chain")

  # R* calculation
  r1 <- r_2d %>% 
    arrange(chain, iter) %>% 
    select(-iter) %>% 
    mutate(chain=as.factor(chain))
  if(dimension==1)
    r1 <- r1 %>% mutate(ones=rnorm(length(chain)))
  a_accuracy <- f_fit_gbm_predict(r1)

  # Split R* calculation
  half_rows <- iters / 2
  r_1h <- r_2d %>% 
    filter(iter<=half_rows)
  r_2h <- r_2d %>% 
    filter(iter>half_rows) %>% 
    mutate(chain=chain + chains)
  r2 <- bind_rows(r_1h, r_2h) %>% 
    select(-iter) %>% 
    mutate(chain=as.factor(chain))
  if(dimension==1)
    r2 <- r2 %>% mutate(ones=rnorm(length(chain)))
  b_accuracy <- f_fit_gbm_predict(r2)

  res[[i]] <- cbind(iters, trend, rep, dimension, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
}
res <- bind_rows(res)
saveRDS(res, "../output/experiment_trends_one_dim.rds")
```

# Stationary -> non-stationary data
Vary rho of AR1 process from stationary->non-stationary generating process
```{r}
f_ar1 <- function(rho, sigma, L){
  x <- vector(length = L)
  x[1] <- rnorm(1, 0, sd=sigma)
  for(i in 2:L)
    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
  return(x)
}

conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  trend = c(0, 0.25, 0.5, 0.75, 1),
  rep = 1:10,
  rho = c(0, 0.25, 0.5, 0.75, 1)
)
res <- vector("list", nrow(conds))
chains = 4
sigma <- 1

for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  trend <- conds[i, "trend"]
  rep <- conds[i, "rep"]
  rho <- conds[i, "rho"]
  r <- array(rnorm(iters * chains), c(iters, chains))
  for(i in 1:chains)
    r[, i] <- f_ar1(rho, sigma, iters)
  rs <- as.data.frame(monitor_extra(r))

  # R* calculation
  r1 <- r %>% 
    as.data.frame() %>% 
    melt(id.vars = NULL) %>% 
    mutate(chain=as.factor(variable)) %>% 
    select(-variable)
  if(dimension==1)
    r1 <- r1 %>% mutate(ones=rnorm(length(chain)))
  a_accuracy <- f_fit_gbm_predict(r1)

  # Split R* calculation
  half_rows <- nrow(r) / 2
  r2 <- cbind(r[1:half_rows, ], r[(half_rows+1):nrow(r), ]) %>% 
    as.data.frame() %>% 
    melt(id.vars = NULL) %>% 
    mutate(chain=as.factor(variable)) %>% 
    select(-variable)
  if(dimension==1)
    r2 <- r2 %>% mutate(ones=rnorm(length(chain)))
  b_accuracy <- f_fit_gbm_predict(r2)

  res[[i]] <- cbind(iters, trend, rep, rho, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
}
res <- bind_rows(res)
saveRDS(res, "../output/experiment_trends_rho.rds")

```

### 2020-03-21 src/s_eight_schools.Rmd
added sims of trends

--- 2020-03-17 src/s_examples.Rmd
@@
 ---
-title: "Examples in R* paper"
-author: "Ben Lambert"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+title: "R Notebook"
 output: html_notebook
@@
 source("monitornew.R")
-```
-
-# AR1 example
-Functions to generate simulated data
-```{r}
-f_ar1 <- function(rho, sigma, L){
-  x <- vector(length = L)
-  x[1] <- rnorm(1, 0, sd=sigma)
-  for(i in 2:L)
-    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
-  return(x)
-}
-
-f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:3)
-    x[, i] <- f_ar1(rho, sigma, L)
-  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
-  x[, 4] <- z
-  return(x)
-}
-
-f_generate_lower_var_four(1/10, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  ggplot(aes(x=value, fill=as.factor(chain))) +
-  geom_histogram(position="identity", alpha=0.3)
-```
-
-Use gbm model in Caret to predict chain identity
-```{r}
-library(caret)
-library(gbm)
-
-# generate data
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-
-# add in extra column of nonsense values since gbm needs 2+ vars
-full_data <- full_data %>% 
-  mutate(chain=as.factor(chain)) %>% 
-  mutate(ones=rnorm(nrow(full_data)))
-
-# create training/testing data
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
 
@@
                    n.minobsinnode=10)
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid)
-# plot(gbmFit1)
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
 
-# plot predictions as fn of value
-testing_data %>% 
-  arrange(value) %>% 
-  mutate(predictions=predict(object=gbmFit1, .)) %>% 
-  ggplot(aes(x=value, y=predictions)) +
-  geom_jitter() +
-  coord_flip()
-```
-
-Perform 1000 replicates
-```{r}
-f_replicate <- function(){
-  full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-  rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-  training_data <- full_data[rand_samples, ]
-  testing_data <- full_data[-rand_samples, ]
-
-  # add in extra column of nonsense values since the method needs at least 2 vars 
-  full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-  training_data <- full_data[rand_samples, ]
-  testing_data <- full_data[-rand_samples, ]
+f_fit_gbm_predict <- function(r){
+  rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
+  training_data <- r[rand_samples, ]
+  testing_data <- r[-rand_samples, ]
 
@@
   plda <- predict(object=gbmFit1, newdata=testing_data)
+  plda_prob <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
   a_accuracy <- 
@@
     pull()
-  return(a_accuracy)
+  return(list(accuracy=a_accuracy, predicted_prob=plda_prob, testing_data=testing_data))
 }
 
-nreplicates <- 1000
-accuracies <- vector(length = nreplicates)
-for(i in 1:nreplicates){
-  if(i%%100==0)
-    print(i)
-  accuracies[i] <- f_replicate()
+f_r_star_distribution <- function(nsim, pred_test_prob, testing_data){
+
+  mAccuracy <- matrix(nrow = nrow(pred_test_prob),
+                    ncol = nsim)
+  for(j in 1:nrow(pred_test_prob)){
+    test <- apply(rmultinom(nsim, 1, prob = pred_test_prob[j, ]), 2, function(x) which(x==1))
+    mAccuracy[j, ] <- if_else(test==testing_data$chain[j], 1, 0)
   }
-```
-
-Plot results
-```{r}
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
-  library(grid)
-
-  # Make a list from the ... arguments and plotlist
-  plots <- c(list(...), plotlist)
-
-  numPlots = length(plots)
-
-  # If layout is NULL, then use 'cols' to determine layout
-  if (is.null(layout)) {
-    # Make the panel
-    # ncol: Number of columns of plots
-    # nrow: Number of rows needed, calculated from # of cols
-    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                    ncol = cols, nrow = ceiling(numPlots/cols))
+  return(colMeans(mAccuracy) * n_distinct(testing_data$chain))
 }
 
- if (numPlots==1) {
-    print(plots[[1]])
-
-  } else {
-    # Set up the page
-    grid.newpage()
-    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
-
-    # Make each plot, in the correct location
-    for (i in 1:numPlots) {
-      # Get the i,j matrix positions of the regions that contain this subplot
-      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-
-      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                      layout.pos.col = matchidx$col))
-    }
-  }
-}
-
-g2 <- tibble(accuracy=accuracies/0.25, iteration=seq(1, nreplicates, 1)) %>% 
-  ggplot(aes(x=accuracy)) +
-  geom_histogram() +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black")) +
-  xlim(1, NA)
-
-g1 <- testing_data %>% 
-  arrange(value) %>% 
-  mutate(predictions=predict(object=gbmFit1, .)) %>% 
-  ggplot(aes(x=value, y=predictions)) +
-  geom_jitter(height = 0.3) +
-  xlab("Value") +
-  ylab("Chain classification") +
-  theme(text = element_text(size=14, colour="black")) +
-  coord_flip() +
-  ggtitle("A.")
-```
-
-Generate R* distribution based on 1000 iterations
-```{r}
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-f_generate_all_same_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:4)
-    x[, i] <- f_ar1(rho, sigma, L)
-  return(x)
-}
-
-full_data <- f_generate_all_same_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy1 <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy1[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy1)/0.25)
-
-mAR1 <- tibble(unconverged=colMeans(mAccuracy)/0.25,
-               converged=colMeans(mAccuracy1)/0.25)
-g3 <- 
-  mAR1 %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series") +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
-
-pdf("../output/ar1.pdf", width = 12, height = 6)
-multiplot(g1, g2, g3, cols = 3)
-dev.off()
-```
-
-# Bivariate normal
-```{r}
-rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
-  return(rmvnorm(n, c(mux, muy),
-                 matrix(c(sigmax^2, sigmax * sigmay * rho,
-                          sigmax * sigmay * rho, sigmay^2),
-                        ncol = 2)))
-}
-# calling function
-n <- 1000
-chain1 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
-  as.data.frame() %>% 
-  mutate(chain=1,
-         iter=seq_along(chain))
-chain2 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
-  as.data.frame() %>% 
-  mutate(chain=2,
-         iter=seq_along(chain))
-chain3 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
-  as.data.frame() %>% 
-  mutate(chain=3,
-         iter=seq_along(chain))
-chain4 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.9) %>% 
-  as.data.frame() %>% 
-  mutate(chain=4,
-         iter=seq_along(chain))
-
-chains_stacked <- rbind.data.frame(chain1, chain2, chain3, chain4) %>% 
-  mutate(chain=as.factor(chain))
-chains_stacked1 <- chains_stacked %>% 
-  select(-iter)
-
-full_data <- chains_stacked1
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-# 
-# # use Vehtari et al. script to calculate Rhat
-# chains_wide1 <- chains_stacked %>% 
-#   pivot_wider(values_from = c("V1"), names_from = c("chain"), id_cols = "iter")
-# rhat_rfun(as.matrix(chains_wide1[, 2:5]))
-# chains_wide2 <- chains_stacked %>% 
-#   pivot_wider(values_from = c("V2"), names_from = c("chain"), id_cols = "iter")
-# rhat_rfun(as.matrix(chains_wide2[, 2:5]))
-# 
-# # Use Vehtari et al. to calculate ess
-# ess_rfun(as.matrix(chains_wide1[, 2:5]))
-# ess_rfun(as.matrix(chains_wide2[, 2:5]))
-# ess_bulk(as.matrix(chains_wide1[, 2:5]))
-# ess_bulk(as.matrix(chains_wide2[, 2:5]))
-# ess_tail(as.matrix(chains_wide1[, 2:5]))
-# ess_tail(as.matrix(chains_wide2[, 2:5]))
-# ess_quantile(as.matrix(chains_wide1[, 2:5]), 0.95)
-# ess_tail(as.matrix(chains_wide2[, 2:5]))
-
-# Try monitor functionality -- split chains first
-half_iter <- max(chains_stacked$iter) / 2
-m_samples <- array(dim = c(half_iter,
-                           2*n_distinct(chains_stacked$chain),
-                           2))
-k <- 1
-chains_stacked_first <- chains_stacked %>% 
-  filter(iter<=half_iter)
-chains_stacked_second <- chains_stacked %>% 
-  filter(iter>half_iter)
-
-for(i in 1:half_iter){
-  temp <- chains_stacked_first %>% 
-    filter(iter==i)
-  for(j in 1:4){
-    temp1 <- temp %>%
-      filter(chain==j)
-    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
-    k <- k + 1
-  }
-}
-
-for(i in 1:half_iter){
-  temp <- chains_stacked_second %>% 
-    filter(iter==(i + 500))
-  for(j in 5:8){
-    temp1 <- temp %>%
-      filter(chain==(j - 4))
-    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
-    k <- k + 1
-  }
-}
-
-g <- 
-  qplot(4*colMeans(mAccuracy)) +
-  xlab(TeX("$R*$")) +
-  xlim(0.9, NA) +
-  geom_vline(xintercept = 1, linetype=2) +
-  ylab("Count") +
-  theme(text=element_text(size=14, colour="black"))
-ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
-mon <- monitor(m_samples)
-
-```
-
-# 250-dimensional multivariate normal
-Fit model in Stan with 10000 iterations
-```{r}
-N <- 250
-A <- rWishart(1, 250, diag(N))[,,1]
-saveRDS(A, "../output/A_matrix.rds")
-model <- stan_model("mvt_250.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
-print(fit)
-# saveRDS(fit, "../output/mvt_fit_10000.rds")
-fit <- readRDS("../output/mvt_fit_10000.rds")
-full_data <- rstan::extract(fit, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:251] <- full_data[j, i, ]
-    m_flattened[k, 252] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V252) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- tibble(samples_10000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-
-importance <- 100-varImp(gbmFit1)$importance$Overall
-rhat <- summary(fit)$summary[, 10]
-ess <- summary(fit)$summary[, 9]
-plot(importance, rhat)
-plot(importance, ess)
-b_df <- tibble(importance, ess, rhat, lp=c(rep(0, 250), 1))
-sum(rhat<1.01)
-sum(ess>400)
-gc <- 
-  b_df %>%
-  ggplot(aes(importance, ess)) +
-  geom_jitter(width=1, aes(shape=as.factor(lp))) +
-  geom_smooth(span=1) +
-  xlab("Variable importance") +
-  ylab("ESS") +
-  ggtitle("C.") +
-  theme(legend.position = "none")
-
-gb <- b_df %>%
-  ggplot(aes(importance, rhat)) +
-  geom_jitter(width = 1, aes(shape=as.factor(lp))) +
-  geom_smooth(span=1) +
-  xlab("Variable importance") +
-  ylab(TeX("$\\hat{R}$")) +
-  ggtitle("B.") +
-  theme(legend.position = "none")
-```
-
-With 400 iterations
-```{r}
-fit <- sampling(model, data=list(N=N, A=A), iter=400, chains=4)
-print(fit)
-# saveRDS(fit, "../output/mvt_fit_400.rds")
-fit <- readRDS("../output/mvt_fit_400.rds")
-
-full_data <- rstan::extract(fit, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:251] <- full_data[j, i, ]
-    m_flattened[k, 252] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V252) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- a_df %>%
-  mutate(samples_400=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-```
-
-```{r}
-model <- stan_model("mvt_250_ncp.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
-print(fit)
-# saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
-fit <- readRDS("../output/mvt_ncp_fit_10000.rds")
-
-full_data <- rstan::extract(fit, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams + 1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V502) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- a_df %>%
-  mutate(samples_10000_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-
-saveRDS(a_df, "../output/mvt_fit_all_accuracy.rds")
-
-ga <- 
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of samples", labels=c("400", "10000, cp", "10000, ncp")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8)) +
-  ggtitle("A.")
-
-pdf("../output/mvt_three.pdf", width = 12, height = 8)
-multiplot(ga, gb, gc, cols = 3)
-dev.off()
-```
-
-# Cauchy example
-
-Generate R* distribution from nominal Cauchy fit
-```{r}
-nsim <- 1000
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0)
-full_data <- rstan::extract(fit_nom, permuted=F)
-nparams <- 52
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V53) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_nom <- mAccuracy
-```
-
-Generate R* distribution from alternative Cauchy parameterisation
-```{r}
-fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0)
-full_data <- rstan::extract(fit_alt1, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V153) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_alt_1 <- mAccuracy
-```
-
-```{r}
-a_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
-               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
-
-g <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none")
-g
-
-ggsave("../output/cauchy.pdf", g, width = 8, height = 6)
-```
-
-## More iterations
-Nominal model with 10000 iterations and thinning by 10
-```{r}
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0, iter = 10000,
-                thin = 5)
-full_data <- rstan::extract(fit_nom, permuted=F)
-nparams <- 52
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V53) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-mAccuracy_nom <- mAccuracy
-```
-
-Alternative model with 10000 iterations and thinning by 10
-```{r}
-fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0, iter = 10000,
-                thin = 5)
-full_data <- rstan::extract(fit_alt1, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V153) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_alt_1 <- mAccuracy
-```
-
-Plot
-```{r}
-b_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
-               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
-
-g1 <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none") +
-  ggtitle("A.")
-g1
-
-g2 <-
-  b_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black")) +
-  ggtitle("B.")
-g2
-
-pdf("../output/cauchy.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
-```
-
-# Eight schools
-```{r}
 source("eight_schools.data.R")
@@
 full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
+both <- f_fit_gbm_predict(full_data)
 
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+plda <- both$predicted_prob
+testing_data <- both$testing_data
 
 nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_centered <- mAccuracy
+r_star_centered <- f_r_star_distribution(nsim, plda, testing_data)
 ```
@@
 full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
+both <- f_fit_gbm_predict(full_data)
 
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
+plda <- both$predicted_prob
+testing_data <- both$testing_data
 
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_noncentered <- mAccuracy
+nsim <- 1000
+r_star_noncentered <- f_r_star_distribution(nsim, plda, testing_data)
 ```
@@
 ```{r}
-c_df <- tibble(centered=colMeans(mAccuracy_centered)/0.125,
-       non_centered=colMeans(mAccuracy_noncentered)/0.125)
+c_df <- tibble(centered=r_star_centered,
+               non_centered=r_star_noncentered)
 ```
@@
 ```
+```

### 2020-03-21 src/s_trends.Rmd
added sims of trends

@@
   k <- 1
-  for(i in 1:iters){
+  for(ii in 1:iters){
     for(j in 1:chains){
-      r_2d[k, ] <- c(i, j, r[i, j, ])
+      r_2d[k, ] <- c(ii, j, r[ii, j, ])
       k <- k + 1
@@
 
+```{r}
+res <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
+  select(trend, iters, rep, dimension, Rstar) %>% 
+  unique()
+ggplot(data = res, aes(x = trend, y = Rstar)) + 
+  geom_point() + 
+  geom_jitter() + 
+  facet_grid(vars(iters), vars(dimension)) + 
+  geom_hline(yintercept = 1) + 
+  ggtitle('R* without splitting')
+```
+
+```{r}
+res <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
+  select(trend, iters, rep, dimension, Rstar_split) %>% 
+  unique()
+ggplot(data = res, aes(x = trend, y = Rstar_split)) + 
+  geom_point() + 
+  geom_jitter() + 
+  facet_grid(vars(iters), vars(dimension)) + 
+  geom_hline(yintercept = 1) + 
+  ggtitle('Split-R*')
+```
+
 # Stationary -> non-stationary data
@@
   rep = 1:10,
-  rho = c(0, 0.25, 0.5, 0.75, 1)
+  rho = c(0.8, 0.85, 0.9, 0.95, 0.975, 1)
 )
@@
   r <- array(rnorm(iters * chains), c(iters, chains))
-  for(i in 1:chains)
-    r[, i] <- f_ar1(rho, sigma, iters)
+  for(j in 1:chains)
+    r[, j] <- f_ar1(rho, sigma, iters)
   rs <- as.data.frame(monitor_extra(r))
@@
     select(-variable)
-  if(dimension==1)
+  if(ncol(r1)==2)
     r1 <- r1 %>% mutate(ones=rnorm(length(chain)))
@@
     select(-variable)
-  if(dimension==1)
+  if(ncol(r2)==2)
     r2 <- r2 %>% mutate(ones=rnorm(length(chain)))
@@
 res <- bind_rows(res)
-saveRDS(res, "../output/experiment_trends_rho.rds")
+saveRDS(res, "../output/experiment_trends_rho_nearer1.rds")
 
 ```
+
+```{r}
+res <- readRDS("../output/experiment_trends_rho_nearer1.rds")
+
+ggplot(data = res, aes(x = rho, y = Rstar)) + 
+  geom_jitter() + 
+  facet_grid(vars(iters)) + 
+  geom_hline(yintercept = 1) + 
+  ggtitle('R* without splitting') +
+  ylim(0, 6)
+
+ggplot(data = res, aes(x = rho, y = Rstar_split)) + 
+  geom_jitter() + 
+  facet_grid(vars(iters)) + 
+  geom_hline(yintercept = 1) + 
+  ggtitle('Split-R*') +
+  ylim(0, 6)
+```

### 2020-03-23 src/s_cauchy.Rmd
added multivariate nonparametric test

---
title: "Cauchy example"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(mvtnorm)
library(goftest)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

## Few iterations
Generate split-R* distribution from nominal Cauchy fit
```{r}
fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878,
                refresh = 0)
x <- rstan::extract(fit_nom, permuted=F)

r_star_nom <- r_star_dist(x)
```

Generate split-R* distribution from alternative Cauchy parameterisation
```{r}
fit_alt1 <- stan(file = 'cauchy_alt_1.stan',
                 seed = 7878, refresh = 0)
full_data <- rstan::extract(fit_alt1, permuted=F)

r_star_alt <- r_star_dist(full_data)
```

Plot split-R* values for each parameterisation
```{r}
a_df <- tibble(nominal=r_star_nom,
               alt=r_star_alt)
g <-
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = "none")
g
```

## Calculate KL divergence from true density by fitting kernel density estimator to points then integrating
nominal parameterisation
```{r}
x <- rstan::extract(fit_nom, permuted=F)

sample <- as.vector(x[, , i])
dens <- density(sample)
f_approx <- approxfun(dens$x, dens$y, rule=2)
integrate(
    function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -20, 20, subdivisions = 1000)

t <- seq(-20, 20, 0.01)
tibble(t,
  true=map_dbl(t, ~dcauchy(.)),
  approx=map_dbl(t, ~f_approx(.))) %>%
  melt(id.vars="t") %>%
  ggplot(aes(x=t, y=value, colour=as.factor(variable))) +
  geom_line()

f_kl_approx <- function(sample, limit=20){
  dens <- density(sample)
  f_approx <- approxfun(dens$x, dens$y, rule=2)
  kl <- integrate(function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -limit, limit, subdivisions=1000)$value
  if(kl < 0)
    return(0)
  else
    return(kl)
}

kl <- vector(length = 50)
for(i in seq_along(kl))
  kl[i] <- f_kl_approx(x[, , i], limit=20)
```

Density estimation fails for alternative parameterisation because of fat tails
```{r}
names(extract(fit_alt1))
x <- rstan::extract(fit_alt1, permuted=F, pars="x")
sample <- as.vector(x[, , 3])
dens <- density(sample, 0.001)
f_approx <- approxfun(dens$x, dens$y, rule=2)
integrate(function(x) f_approx(x), -10, 10)
integrate(
    function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -100, 100, subdivisions = 1000)

t <- seq(-20, 20, 0.01)
tibble(t,
  true=map_dbl(t, ~dcauchy(.)),
  approx=map_dbl(t, ~f_approx(.))) %>%
  melt(id.vars="t") %>%
  ggplot(aes(x=t, y=value, colour=as.factor(variable))) +
  geom_line()

f_kl_approx <- function(sample, limit=20){
  dens <- density(sample)
  f_approx <- approxfun(dens$x, dens$y, rule=2)
  kl <- integrate(function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -limit, limit, subdivisions=1000)$value
  if(kl < 0)
    return(0)
  else
    return(kl)
}

kl <- vector(length = 50)
for(i in seq_along(kl))
  kl[i] <- f_kl_approx(x[, , i], limit=20)
```

Look at quantiles instead using KS test statistic: problem with KS test statistic is that it is most sensitive to middle values: https://asaip.psu.edu/articles/beware-the-kolmogorov-smirnov-test/
```{r}
x <- rstan::extract(fit_alt1, permuted=F, pars="x")
sample <- as.vector(x[, , 1])
ks.test(sample, pcauchy)
ks <- vector(length = 50)
for(i in 1:50)
  ks[i] <- ks.test(x[, , i], pcauchy)[1]$statistic
```

Anderson-darling test -- by comparing with an independent sampler we get a gauge of the quality of MCMC
```{r}
x <- rstan::extract(fit, permuted=F, pars="x")

f_calculate_ad <- function(x){
  dims <- dim(x)
  nparams <- dims[3]
  ad <- vector(length = nparams)
  for(i in 1:nparams)
    ad[i] <- ad.test(x[, , i], pcauchy)[2]$p.value
  return(ad)
}

# keep sample size fixed by thinning
nmax <- 5000
iter <- seq(500, nmax, 50)
ad <- vector(length = length(iter))
for(i in seq_along(iter)){
  a_iter <- iter[i]
  ele <- round(seq(1, a_iter, length.out = 500))
  ad[i] <- mean(f_calculate_ad(x[ele, ,]))
}

e_df <- tibble(iter, ad, source="MCMC")

# compare with independent samples
# ps <- map_dbl(seq(1, 1000, 1), function(i) ad.test(rcauchy(1000), pcauchy)[2]$p.value)
# hist(ps)

x <- array(rcauchy(2 * nmax * 50), c(nmax, 4, 50))

iter <- seq(500, nmax, 50)
ad <- vector(length = length(iter))
for(i in seq_along(iter)){
  a_iter <- iter[i]
  ele <- round(seq(1, a_iter, length.out = 500))
  ad[i] <- mean(f_calculate_ad(x[ele, ,]))
}
f_df <- e_df %>% 
  bind_rows(tibble(iter, ad, source="independent"))
f_df %>% 
  ggplot(aes(x=iter, y=ad, colour=as.factor(source))) +
  geom_line()
```

Try R* with independent sampler
```{r}
nmax <- 100
x <- array(rcauchy(2 * nmax * 50), c(nmax, 4, 50))
r_star_inde <- r_star_dist(x)
hist(r_star_inde)
```

Possible to tell whether sample came from MCMC or independent?
```{r}
x_mcmc <- rstan::extract(fit, permuted=F, pars="x")
x_mcmc <- array(rcauchy(2000 * 50), c(2000, 50)) %>% 
  as.data.frame() %>% 
  mutate(source="mcmc")

f_distribution <- function(x_mcmc){
  x_inde <- array(rcauchy(2000 * 50), c(2000, 50))
  x_inde <- x_inde %>% 
  as.data.frame() %>% 
  mutate(source="independent")
  r <- bind_rows(x_mcmc,
                 x_inde) %>% 
    mutate(source=as.factor(source))
  rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
  training_data <- r[rand_samples, ]
  testing_data <- r[-rand_samples, ]

  gbmFit1 <- train(source ~ ., data = training_data, 
                     method = "gbm",
                     trControl = trainControl(method = 'none'), 
                     tuneGrid = caretGrid, verbose=FALSE)

  plda <- predict(object=gbmFit1, newdata=testing_data)
  a_accuracy <- 
    tibble(predicted=plda, actual=testing_data$source) %>%
    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
    summarise(mean(correct)) %>% 
    pull()
  return(a_accuracy)
}
library(furrr)
future::plan(multiprocess)
laccuracy <- future_map_dbl(seq(1, 1000, 1), function(i) f_distribution(x_mcmc))
hist(laccuracy, 50)
```

Could do same test with simply empirical distributions, also, by bootstrapping

## 10,000 iterations
Nominal model with 10000 iterations and thinning by 10
```{r}
fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
            refresh = 0, iter = 1000)
full_data <- rstan::extract(fit, permuted=F)

r_star_nom <- r_star_dist(full_data)
```

Alternative model with 10000 iterations and thinning by 10
```{r}
fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
            refresh = 0, iter = 1000)
full_data <- rstan::extract(fit, permuted=F)

r_star_alt <- r_star_dist(full_data)
```

Plot
```{r}
b_df <- tibble(nominal=r_star_nom,
               alt=r_star_alt)

g1 <-
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = "none") +
  ggtitle("A.")
g1

g2 <-
  b_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8) +
  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black")) +
  ggtitle("B.")
g2

pdf("../output/cauchy.pdf", width = 12, height = 6)
multiplot(g1, g2, cols = 2)
dev.off()
```

### 2020-03-23 src/s_eight_schools.Rmd
added multivariate nonparametric test

@@
 
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
+both <- f_fit_gbm_predict(m_flattened)
 
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+plda <- both$predicted_prob
+testing_data <- both$testing_data
 
 nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_centered <- mAccuracy
+r_star_centered <- f_r_star_distribution(nsim, plda, testing_data)
 
+# noncentered
 full_data <- rstan::extract(fit_ncp, permuted=F)
@@
   mutate(chain=as.factor(chain))
+both <- f_fit_gbm_predict(m_flattened)
+plda <- both$predicted_prob
+testing_data <- both$testing_data
 
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_noncentered <- mAccuracy
+nsim <- 1000
+r_star_noncentered <- f_r_star_distribution(nsim, plda, testing_data)
 ```
@@
 ```{r}
-d_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
-       non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
+d_df <- tibble(centered=r_star_centered,
+               non_centered=r_star_noncentered)
 ```
@@
   c_df %>% 
-  melt() %>% 
+  melt(id.vars=NULL) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   d_df %>% 
-  melt() %>% 
+  melt(id.vars=NULL) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   theme(text = element_text(size=14, colour="black"))
-a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
+# a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
 
-pdf("../output/eight_schools.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
+# pdf("../output/eight_schools.pdf", width = 12, height = 6)
+# multiplot(g1, g2, cols = 2)
+# dev.off()
 ```
+
+# Plot tau prediction accuracy across quantiles
+```{r}
+full_data_all <- rstan::extract(fit_cp, permuted=F)
+full_data <- array(dim=c(500, 8, 11))
+k <- 1
+for(i in 1:4){
+  first_half <- full_data_all[1:500, i, ]
+  second_half <- full_data_all[501:1000, i, ]
+  full_data[, k, ] <- first_half
+  k <- k + 1
+  full_data[, k, ] <- second_half
+  k <- k + 1
+}
+
+m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = 12)
+k <- 1
+for(i in 1:8){
+  for(j in 1:nrow(full_data)){
+    m_flattened[k, 1:11] <- full_data[j, i, ]
+    m_flattened[k, 12] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V12) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+both <- f_fit_gbm_predict(full_data)
+
+plda <- both$predicted_prob
+testing_data <- both$testing_data
+
+f_r_star_accuracy <- function(nsim, pred_test_prob, testing_data){
+
+  mAccuracy <- matrix(nrow = nrow(pred_test_prob),
+                    ncol = nsim)
+  for(j in 1:nrow(pred_test_prob)){
+    test <- apply(rmultinom(nsim, 1, prob = pred_test_prob[j, ]), 2, function(x) which(x==1))
+    mAccuracy[j, ] <- if_else(test==testing_data$chain[j], 1, 0)
+  }
+  return(mAccuracy)
+}
+
+nsim <- 20
+mAccuracy <- f_r_star_accuracy(nsim, plda, testing_data)
+colnames(testing_data) <- c("mu", "tau",
+                            "theta.1", "theta.2",
+                            "theta.3", "theta.4",
+                            "theta.5", "theta.6",
+                            "theta.7", "theta.8",
+                            "lp", "chain")
+colnames(mAccuracy) <- map_chr(seq(1, nsim, 1), ~paste0("V", .))
+mAccuracy <- mAccuracy %>% 
+  as.data.frame() %>% 
+  bind_cols(testing_data)
+
+lbreaks <- quantile(mAccuracy$tau, seq(0, 1, 0.1))
+mAccuracy$tau_cuts <- cut(mAccuracy$tau, breaks=lbreaks, include.lowest = T, right=T, ordered_result = T)
+cuts <- sort(unique(mAccuracy$tau_cuts))
+r_star <- matrix(ncol = length(cuts),
+                 nrow=nsim)
+for(i in seq_along(cuts)){
+  temp <- mAccuracy %>% 
+    filter(tau_cuts==cuts[i])
+  r_star[, i] <- colMeans(temp[, 1:nsim]) * n_distinct(testing_data$chain)
+}
+colnames(r_star) <- seq(0.1, 1, 0.1)
+r_mean <- 
+  r_star %>% 
+  melt(id.vars=NULL) %>%  
+  group_by(Var2) %>% 
+  summarise(value=mean(value))
+r_full <- 
+  r_star %>%
+  melt(id.vars=NULL) %>%
+  bind_rows(r_mean) %>% 
+  mutate(variable=as.numeric(as.character(Var2)))
+
+ggplot(filter(r_full, !is.na(Var1)),
+       aes(x=variable, y=value)) +
+  geom_jitter(height = 0, width = 0.02, colour="grey") +
+  ylab(TeX("$R*$")) +
+  xlab("Quantile") +
+  geom_hline(yintercept = 1, linetype=2) +
+  theme(text = element_text(colour="black",
+                            size=12)) +
+  geom_line(data=filter(r_full, is.na(Var1)))
 ```

### 2020-03-23 src/r_star_monitor.R

@@ -1,0 +1,175 @@
+library(caret)
+library(gbm)
+library(tidyverse)
+
+#' Estimate R star
+#'
+#' Compute estimate of R* using a gradient-boosted model (GBM)
+#'
+#' @param x a 3D array of samples of dimensions: # iter x # chains x # parameters
+#'
+#' @param split_chains a Boolean (defaults to true) indicating whether to split
+#' chains into two equal halves
+#'
+#' @param training_percent proportion of iterations used to train GBM model
+#' (default is 0.7)
+#'
+#' @param caret_default GBM hyperparameter settings given as a list (defaults
+#' are interaction.depth=3, n.trees=50, shrinkage=0.1, nminobsinmode=10)
+#'
+#' @return estimated R* value
+#' 
+#' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
+#' diagnostic with uncertainty using gradient-boosted machines
+#' \emph{arXiv preprint} \code{arXiv:TBD}
+r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
+  
+  if(split_data)
+    x <- split_data(x)
+  
+  nparams <- dim(x)[3]
+  nchains <- dim(x)[2]
+  niter <- dim(x)[1]
+  m_flattened <- matrix(nrow = nrow(x) * nchains,
+                        ncol = (nparams + 1))
+  k <- 1
+  for(i in 1:nchains){
+    for(j in 1:nrow(x)){
+      m_flattened[k, 1:nparams] <- x[j, i, ]
+      m_flattened[k, (nparams + 1)] <- i
+      k <- k + 1
+    }
+  }
+  m_flattened <- m_flattened %>% 
+    as.data.frame()
+  colnames(m_flattened)[nparams + 1] <- "chain"
+  r <- m_flattened %>% 
+    mutate(chain=as.factor(chain))
+  
+  rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
+  training_data <- r[rand_samples, ]
+  testing_data <- r[-rand_samples, ]
+  
+  if(is.na(caret_grid_default))
+    caretGrid <- expand.grid(interaction.depth=c(3), 
+                             n.trees = 50,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)
+  else
+    caretGrid <- expand.grid(caret_grid_default)
+  
+  gbmFit <- train(chain ~ ., data = training_data, 
[…]

### 2020-03-24 src/s_cauchy.Rmd
added cauchy convergence plots

@@
 source("r_star_monitor.R")
+
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  library(grid)
+
+  # Make a list from the ... arguments and plotlist
+  plots <- c(list(...), plotlist)
+
+  numPlots = length(plots)
+
+  # If layout is NULL, then use 'cols' to determine layout
+  if (is.null(layout)) {
+    # Make the panel
+    # ncol: Number of columns of plots
+    # nrow: Number of rows needed, calculated from # of cols
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                    ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+
+ if (numPlots==1) {
+    print(plots[[1]])
+
+  } else {
+    # Set up the page
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+    # Make each plot, in the correct location
+    for (i in 1:numPlots) {
+      # Get the i,j matrix positions of the regions that contain this subplot
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+  }
+}
+
 ```
@@
 fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
-            refresh = 0, iter = 1000)
+            refresh = 0, iter = 50000)
+saveRDS(fit, "../output/fit_nominal_50000.rds")
 full_data <- rstan::extract(fit, permuted=F)
@@
 fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
-            refresh = 0, iter = 1000)
+            refresh = 0, iter = 50000)
+saveRDS(fit, "../output/fit_alt_50000.rds")
 full_data <- rstan::extract(fit, permuted=F)
@@
 ```
+
+Compare quantiles
+```{r}
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+r_squared <- function(x, lower=0.001){
+  qs <- seq(lower, 1-lower, lower)
+  q_actual <- map_dbl(qs, ~qcauchy(.))
+  r_vals <- vector(length = 50)
+  for(i in seq_along(r_vals)){
+    q_approx <- quantile(x[, , i], qs)
+    aLM <- lm(data=tibble(q_approx, q_actual), formula = q_actual~q_approx)
+    a_sum <- summary(aLM)
+    r_vals[i] <- a_sum$r.squared
+  }
+  return(r_vals)
+}
+
+log_last <- log10(25000)
+log_first <- log10(100)
+iter <- round(10^(seq(log_first, log_last, length.out = 20)))
+r_vals1 <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
+r_vals1_lower <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.025))
+r_vals1_upper <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.975))
+
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+r_vals <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
+r_vals_lower <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.025))
+r_vals_upper <- map_dbl(iter, ~quantile(r_squared(x[1:., , ]), 0.975))
+```
+
+Compare nominal and alt parameterisations
+```{r}
+b_df <- 
+  tibble(iteration=iter,
+         middle=r_vals,
+         lower=r_vals_lower,
+         upper=r_vals_upper,
+         variable="alt") %>% 
+  bind_rows(tibble(iteration=iter,
+         middle=r_vals1,
+         lower=r_vals1_lower,
+         upper=r_vals1_upper,
+         variable="nom"))
+# saveRDS(b_df, "../output/cauchy_r_squared.rds")
+g1 <- b_df %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=middle,colour=as.factor(variable))) +
+  geom_line(aes(y=middle,colour=as.factor(variable))) +
+  ylim(NA, 1) +
+  ylab(TeX("$R^2$")) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  scale_x_log10() +
+  ggtitle(TeX("Quantile $R^2$"))
+g1
+```
+
+Progression of R*
+```{r}
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+
+library(furrr)
+future::plan(multiprocess)
+laccuracy <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
+
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+
+laccuracy1 <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
+
+a_df <- 
+  tibble(iteration=iter,
+         alt=laccuracy1,
+       nom=laccuracy)
+# saveRDS(a_df, "../output/cauchy_r_star.rds")
+```
+
+```{r}
+g2 <- 
+  a_df %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration, y=value, colour=as.factor(variable))) +
+  geom_point() +
+  geom_line() +
+  ylab(TeX("$R^*$")) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = c(0.8, 0.8),
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_x_log10() +
+  scale_y_continuous(limits=c(0.8, NA), breaks = seq(1, 7, 1)) +
+  ggtitle(TeX("Split-$R^*$"))
+g2
+```
+
+Rhat
+```{r}
+r_hat_all_params <- function(x){
+ nparams <- dim(x)[3] 
+ rhat_vals <- map_dbl(seq(1, nparams, 1), ~rhat(split_chains(x[ , , .])))
+ return(rhat_vals)
+}
+
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+future::plan(multiprocess)
+r_hat <- future_map_dbl(iter, ~max(r_hat_all_params(x[1:., , ])))
+r_hat_lower <- future_map_dbl(iter, ~quantile(r_hat_all_params(x[1:., , ]), 0.025))
+r_hat_upper <- future_map_dbl(iter, ~quantile(r_hat_all_params(x[1:., , ]), 0.975))
+
+fit1 <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+r_hat1 <- future_map_dbl(iter, ~max(r_hat_all_params(x[1:., , ])))
+r_hat1_lower <- future_map_dbl(iter, ~quantile(r_hat_all_params(x[1:., , ]), 0.025))
+r_hat1_upper <- future_map_dbl(iter, ~quantile(r_hat_all_params(x[1:., , ]), 0.975))
+
+c_df <- 
+  tibble(iteration=iter,
+         middle=r_hat1,
+         lower=r_hat1_lower,
+         upper=r_hat1_upper,
+         variable="alt") %>% 
+  bind_rows(tibble(iteration=iter,
+         middle=r_hat,
+         lower=r_hat_lower,
+         upper=r_hat_upper,
+         variable="nom"))
+# saveRDS(c_df, "../output/cauchy_r_hat.rds")
+g3 <- 
+  c_df %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=middle, colour=as.factor(variable))) +
+  geom_line(aes(y=middle, colour=as.factor(variable))) +
+  ylab(TeX("$\\hat{R}$")) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_x_log10() +
+  scale_y_continuous(limits=c(0.99, NA), breaks = seq(1, 6, 0.2)) +
+  ggtitle(TeX("Max split-$\\hat{R}$"))
+g3
+```
+
+Combine
+```{r}
+pdf("../output/cauchy_convergence.pdf", width = 12, height = 4)
+multiplot(g1, g3, g2, cols = 3)
+dev.off()
+```
+
+Look at quantile accuracy for models converged by R* and Rhat
+```{r}
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+quantile_errors <- function(x, lower=0.001){
+  print(dim(x))
+  qs <- seq(lower, 1-lower, lower)
+  q_actual <- map_dbl(qs, ~qcauchy(.))
+  errors <- matrix(nrow = length(qs),
+                   ncol=dim(x)[3])
+  for(i in 1:50){
+    q_approx <- quantile(x[, , i], qs)
+    errors[, i] <- abs(q_actual - q_approx)
+  }
+  errors <- errors %>% 
+    as.data.frame() %>% 
+    mutate(q=0.5-abs(0.5-qs)) %>% 
+    melt(id.vars="q")
+  return(errors)
+}
+
+errors_df <- quantile_errors(x, lower=0.001)
+errors_df %>% 
+  ggplot(aes(x=q, y=value)) +
+  geom_jitter(width = 0.05) +
+  scale_x_log10() +
+  scale_y_log10() +
+  geom_smooth(method="lm") +
+  xlab("Distance from centre") +
+  ylab("Error")
+
+aLM <- lm(data=errors_df, formula="log(value)~log(q)")
+summary(aLM)
+
+errors_df <- quantile_errors(x[1:iter[8], , ], lower=0.001)
+errors_df %>% 
+  ggplot(aes(x=q, y=value)) +
+  geom_jitter(width = 0.05) +
+  scale_x_log10() +
+  scale_y_log10() +
+  geom_smooth(method="lm") +
+  xlab("Distance from centre") +
+  ylab("Error")
+
+bLM <- lm(data=errors_df, formula="log(value)~log(q)")
+summary(bLM)
+```

### 2020-03-24 src/r_star_monitor.R

@@ -24,7 +24,7 @@ library(tidyverse)
 #' \emph{arXiv preprint} \code{arXiv:TBD}
 r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
   
-  if(split_data)
+  if(split_chains)
     x <- split_data(x)
   
   nparams <- dim(x)[3]
@@ -50,7 +50,7 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]
   
-  if(is.na(caret_grid_default))
+  if(is.na(caret_default))
     caretGrid <- expand.grid(interaction.depth=c(3), 
                              n.trees = 50,
                              shrinkage=c(0.1),
@@ -92,7 +92,7 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
 #' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
-r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_grid_default=NA, nsim=1000){
+r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_default=NA, nsim=1000){
   
   if(split_chains)
     x <- split_data(x)
@@ -120,7 +120,7 @@ r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_grid_defa
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]
   
-  if(is.na(caret_grid_default))
+  if(is.na(caret_default))
     caretGrid <- expand.grid(interaction.depth=c(3), 
                              n.trees = 50,
                              shrinkage=c(0.1),
@@ -158,7 +158,8 @@ split_data <- function(x){
   if(half%%1!=0){
     print("Warning: dropping first iteration size since # iter
           is not an even integer")
-    full_data <- full_data[-1]
+    x <- x[-1, , ]
+    niter <- niter - 1
   }
   half <- floor(half)
   full_data <- array(dim=c(half, 2 * nchains, nparams))

### 2020-03-25 src/s_cauchy.Rmd
added cauchy convergence example

@@
 ```{r}
+a_df <- readRDS("../output/cauchy_r_star.rds")
 g2 <- 
@@
 # saveRDS(c_df, "../output/cauchy_r_hat.rds")
+c_df <- readRDS("../output/cauchy_r_hat.rds")
 g3 <- 

### 2020-03-25 write_up/ms.tex
added cauchy convergence example

@@
 	 \texttt{ben.c.lambert@gmail.com} \\
+	 \And
+	 Aki Vehtari \\
+	 Department of Computer Science\\
+	 Aalto University\\
+	 Finland\\
+	 \texttt{aki.vehtari@aalto.fi}
 }
@@
 
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
@@
 In Fig.
-\ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates.
+\ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates indicating that the chains have not converged.
 
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
@@
 \ref{fig:bivariate}.
-The mean of this distribution is 1.18 and all the samples are above 1 -- indicating that the sampling distribution has not converged.
-By taking account of all the information in the chains,
+The mean of this distribution is 1.18, and all the samples are above 1, indicating that the sampling distribution has not converged.
+By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals.
+As such, we see a role for $R^*$ as an additional convergence metric.
 
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
@@
 In Fig.
-\ref{fig:mvt}C, we plot variable importance versus ESS, which exhibit a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
-Together, these plots illustrate that the GBM's variable importance measure provides complementary information to $\hat{R}$ and ESS.
+\ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
+Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.
 
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
+	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
 In plots B and C, horizontal jitter was added to the points.
@@
 \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5).
-In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has been reached.
+In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
 Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
 
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
@@
 	\label{fig:cauchy}
+\end{figure}
+
+To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how close the sampling distribution is to the target.
+One measure of distributional "closeness" is the KL-divergence, that in this case could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated.
+The trouble is, for distributions like the Cauchy that have fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails.
+This means that approximate KL-divergence is unreliable for these types of model.
+Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-derived quantities.
+Specifically, we calculate the $R^2$ for the regression of the actual quantile values on the sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
+In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 $x_j$.
+
+In the left-hand panel of Fig.
+\ref{fig:cauchy_convergence}, we plot this ``quantile $R^2$'' as a function of MCMC sample size for both the nominal and alternative parameterisations of the Cauchy model.
+This shows that after only c.10,000 iterations the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
+In the middle and right panels of Fig.
+\ref{fig:cauchy_convergence}, we show how the maximum value $\hat{R}$ across all 50 dimensions of the target and $R^*$ as calculated by Algorithm \ref{alg:R_star} (both measures are calculated on chains split into equal halves) perform as the number of MCMC samples increases.
+After c.550 iterations, the alternative parameterisation achieves split-$\hat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum split-$\hat{R}$ value: in both cases, these suggest convergence.
+By contrast, after 25,000 iterations, $R^*\approx 1.05$ for the nominal model, and $R^*>2$ for the nominal parameterisation: in both cases, these $R^*$ values suggest lack of convergence.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
+	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} The left-hand panel shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; the middle panel, shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; the right shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:cauchy_convergence}
 \end{figure}
@@
 In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each sample, meaning that the centered model $R^*$ values are shifted leftwards.
-Despite this, however, the distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged.
+Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears converged.
 
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
 	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
-In both cases, the plots show 1000 $R^*$ samples.
+In all cases, the plots show 1000 $R^*$ samples.
 See the R markdown file on the Github repository for code to reproduce this example.}

### 2020-03-26 src/s_cauchy.Rmd
finished draft of Cauchy section

--- 2020-03-24 src/s_cauchy.Rmd
@@
 # saveRDS(b_df, "../output/cauchy_r_squared.rds")
+b_df <- readRDS("../output/cauchy_r_squared.rds")
 g1 <- b_df %>% 
@@
   scale_x_log10() +
-  ggtitle(TeX("Quantile $R^2$"))
+  ggtitle(TeX("A. Quantile $R^2$"))
 g1
@@
   scale_y_continuous(limits=c(0.8, NA), breaks = seq(1, 7, 1)) +
-  ggtitle(TeX("Split-$R^*$"))
+  ggtitle(TeX("D. Split-$R^*$"))
 g2
@@
   scale_y_continuous(limits=c(0.99, NA), breaks = seq(1, 6, 0.2)) +
-  ggtitle(TeX("Max split-$\\hat{R}$"))
+  ggtitle(TeX("B. Max split-$\\hat{R}$"))
 g3
+```
+
+Multivariate Rhat
+```{r}
+fit1 <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+Rhat_alt <- map_dbl(iter, ~r_hat_multivariate(x[1:., , ]))
+
+fit1 <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+Rhat_nom <- map_dbl(iter, ~r_hat_multivariate(x[1:., , ]))
+g_df <- tibble(iteration=iter,
+       alt=Rhat_alt,
+       nom=Rhat_nom) 
+g4 <- 
+  tibble(iteration=iter,
+       alt=Rhat_alt,
+       nom=Rhat_nom) %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=value, colour=as.factor(variable))) +
+  geom_line(aes(y=value, colour=as.factor(variable))) +
+  ylab(TeX("$\\hat{R}$")) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_x_log10() +
+  scale_y_continuous(limits=c(0.99, NA), breaks = seq(1, 6, 0.2)) +
+  ggtitle(TeX("C. Multivariate split-$\\hat{R}$"))
+```
+
+Tail ESS
+```{r}
+fit1 <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+tail_ess_alt <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_tail(x[1:i, , .]))))
+bulk_ess_alt <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_bulk(x[1:i, , .]))))
+
+fit1 <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+tail_ess_nom <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_tail(x[1:i, , .]))))
+bulk_ess_nom <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_bulk(x[1:i, , .]))))
+
+e_df <- tibble(iteration=iter,
+               alt=tail_ess_alt,
+               nom=tail_ess_nom)
+saveRDS(e_df, "../output/cauchy_tail_ess.rds")
+f_df <- tibble(iteration=iter,
+               alt=bulk_ess_alt,
+               nom=bulk_ess_nom)
+saveRDS(f_df, "../output/cauchy_bulk_ess.rds")
+g5 <- 
+  e_df %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=value, colour=as.factor(variable))) +
+  geom_line(aes(y=value, colour=as.factor(variable))) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 400, linetype=2) +
+  scale_x_log10() +
+  scale_y_log10() +
+  ggtitle(TeX("E. Tail ESS"))
+
+g6 <- 
+  f_df %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=value, colour=as.factor(variable))) +
+  geom_line(aes(y=value, colour=as.factor(variable))) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 400, linetype=2) +
+  scale_x_log10() +
+  scale_y_log10() +
+  ggtitle(TeX("F. Bulk ESS"))
 ```
@@
 ```{r}
-pdf("../output/cauchy_convergence.pdf", width = 12, height = 4)
-multiplot(g1, g3, g2, cols = 3)
+pdf("../output/cauchy_convergence.pdf", width = 12, height = 8)
+multiplot(g1, g4, g5, g3, g2, g6, cols = 2)
 dev.off()

### 2020-03-26 write_up/ms.tex
finished draft of Cauchy section

@@
 Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \cite{geman1984stochastic}, which underpins the popular BUGS \cite{lunn2000winbugs} and JAGS \cite{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \cite{neal2011mcmc} and the No U-Turn Sampler (NUTS) \cite{hoffman2014no}, which Stan \cite{carpenter2017stan} and PyMC3 \cite{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} and the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement.
 MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior.
-Because of this, software packages (Stan in particular \cite{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
+Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
 Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
-We repeat here a quote from Chris Holmes to this effect \cite{holmes},
+We repeat here a quote from Chris Holmes to this effect \citep{holmes},
 
@@
 
-The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \cite{lambert2018Student}.
+The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}.
 MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions).
@@
 Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged.
-Like furious bees trapped in a room of a house \cite{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
+Like furious bees trapped in a room of a house \citep{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
 With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.
 
-To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \cite{gelman1992inference}.
+To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}.
 If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge.
-This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \cite{lambert2018Student}.
-The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \cite{gelman2013bayesian}.
+This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}.
+The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman2013bayesian}.
 If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed.
-Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \cite{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \cite{vehtari2019rank}.
-Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \cite{vehtari2019rank} and adopted \cite{carpenter2017stan}.
+Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \citep{vehtari2019rank}.
+Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
 Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.
@@
 However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
-For our machine learning model, we use gradient-boosted regression trees \cite{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \cite{chollet2018}.
+For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
 For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test).
@@
 
-The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank}.
+The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \citep{vehtari2019rank}.
 Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
 
-%$R^*$ calculation is enabled in Pints inference software \cite{Clerx2019Pints}.
+%$R^*$ calculation is enabled in Pints inference software \citep{Clerx2019Pints}.
 
@@
 By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}.
+By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \citep{gelman2013bayesian}.
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
@@
 
-The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours.
-Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}.
+The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \citep{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
+Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
 The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
@@
 \section{Results}\label{sec:results}
-To illustrate the versatility of $R^*$, we use examples from \cite{vehtari2019rank} and two additional examples based on a multivariate normal model.
+To illustrate the versatility of $R^*$, we use examples from \citep{vehtari2019rank} and two additional examples based on a multivariate normal model.
 
@@
 
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
+First, we use the code provided in \citep{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
 In all cases, the various quantities were calculated based on chains split into halves.
@@
 
-We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \cite{hoffman2014no}.
+We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}.
 We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions.
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \cite{vehtari2019rank}.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
 
@@
 
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
+In GBMs, it is possible to calculate variable importance (see, for example, \citep{friedman2001greedy} and \citep{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
 We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan.
@@
 We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
-Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
+Like \citep{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
 %
@@
 
-In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \cite{vehtari2019rank},
+In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \citep{vehtari2019rank},
 %
@@
 \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+As shown in \citep{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
 The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$.
-Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig.
+Interesting, however, in contrast to the measures presented in \citep{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig.
 \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5).
@@
 
-To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how close the sampling distribution is to the target.
-One measure of distributional "closeness" is the KL-divergence, that in this case could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated.
-The trouble is, for distributions like the Cauchy that have fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails.
+To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target.
+One measure of distributional "closeness" is the KL-divergence, which, in this case, could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated.
+The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails.
 This means that approximate KL-divergence is unreliable for these types of model.
-Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-derived quantities.
-Specifically, we calculate the $R^2$ for the regression of the actual quantile values on the sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
-In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 $x_j$.
+Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
+Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
+In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.
 
-In the left-hand panel of Fig.
-\ref{fig:cauchy_convergence}, we plot this ``quantile $R^2$'' as a function of MCMC sample size for both the nominal and alternative parameterisations of the Cauchy model.
-This shows that after only c.10,000 iterations the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
-In the middle and right panels of Fig.
-\ref{fig:cauchy_convergence}, we show how the maximum value $\hat{R}$ across all 50 dimensions of the target and $R^*$ as calculated by Algorithm \ref{alg:R_star} (both measures are calculated on chains split into equal halves) perform as the number of MCMC samples increases.
-After c.550 iterations, the alternative parameterisation achieves split-$\hat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum split-$\hat{R}$ value: in both cases, these suggest convergence.
-By contrast, after 25,000 iterations, $R^*\approx 1.05$ for the nominal model, and $R^*>2$ for the nominal parameterisation: in both cases, these $R^*$ values suggest lack of convergence.
+In Fig.
+\ref{fig:cauchy_convergence}A, we plot this ``quantile $R^2$'' as a function of MCMC sample size for both parameterisations of the Cauchy model.
+This shows that after c.10,000 iterations, the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
+Next, in Fig.
+\ref{fig:cauchy_convergence}B and Fig.
+\ref{fig:cauchy_convergence}C, we plot two measures of $\hat{R}$, each calculated from splitting the 4 original chains into two equal halves.
+The first of these measures is the rank-normalised $\hat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig.
+\ref{fig:cauchy_convergence}B, we show how the maximum of this measurement across all 50 dimensions changes with sample size.
+After c.550 iterations, the alternative parameterisation achieves $\hat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum $\hat{R}$ value: in both cases, these suggest convergence.
+The second measure is multivariate $\hat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig.
+\ref{fig:cauchy_convergence}C shows how this metric changes with sample size for both Cauchy model parameterisations.
+After c.
+1800 iterations, multivariate $\hat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\hat{R}>1.07$ indicating more samples are needed.
+In Fig.
+\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the nominal model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence.
+Finally, in Figs.
+\ref{fig:cauchy_convergence}E\&F, we plot the minimum across all the dimensions of tail- and bulk-ESS calculated as described in \cite{vehtari2019rank}.
+After c.180 iterations, the alternative parameterisation surpassed a tail-ESS of 400; after c.18,700, the nominal parameterisation did the same.
+Both models were quicker to pass 400 bulk-ESSs.
+
+Comparing our ``omniscient'' measure of convergence shown in Fig.
+\ref{fig:cauchy_convergence}A, with the various derived measures, it is clear that all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
+The rate at which these converge differs though, and $R^*$ (Fig.
+\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to our ``omniscient'' measure.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
-	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} The left-hand panel shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; the middle panel, shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; the right shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}.
+	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split $\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 \subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \citep{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
 
-The model can be parameterised two ways, as described in \cite{vehtari2019rank}.
+The model can be parameterised two ways, as described in \citep{vehtari2019rank}.
 The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
@@
 
-We first used Stan \cite{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \cite{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
+We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
+Like \citep{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
 We also used Stan with the same algorithm settings to sample from the non-centered model.
 
-To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \cite{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
+To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \citep{vehtari2019rank}, resulting in 500 iterations across 8 chains.
 Next, we trained the GBM on a training set comprising 2800 samples, and used it to classify the causative chain for 1200 independent test samples.
@@
 
-\bibliographystyle{unsrt}
+\bibliographystyle{chicago}
 \bibliography{bibliography}

### 2020-03-26 src/r_star_monitor.R

@@ -174,3 +174,51 @@ split_data <- function(x){
   }
   return(full_data)
 }
+
+#' Compute estimate of multivariate Rhat distribution
+#'
+#' @param x a 3D array of samples of dimensions: # iter x # chains x # parameters
+#'
+#' @param split_chains a Boolean (defaults to true) indicating whether to split
+#' chains into two equal halves
+#'
+#' @return single Rhat value
+#' 
+#' @references Stephen Brooks, Andrew Gelman (1998), General Methods for Monitoring
+#' Convergence of Iterative Simulations  
+#' Journal of Computational and Graphical Statistics, Volume 7, Number 4, Pages 434455
+r_hat_multivariate <- function(x, split_chains=T){
+  
+  if(split_chains)
+    x <- split_data(x)
+  
+  niter <- dim(x)[1]
+  nchains <- dim(x)[2]
+  nparams <- dim(x)[3]
+  
+  chains_l <- seq(1, nchains, 1)
+  phi_j_bar <- map(chains_l, ~colMeans(x[, ., ]))
+  phi_diff <- map(chains_l, ~x[ , ., ] - phi_j_bar[[.]])
+  phi_m <- map(chains_l, ~t(phi_diff[[.]])%*%phi_diff[[.]])
+  for(i in chains_l){
+    if(i==1)
+      phi_tot <- phi_m[[i]]
+    else
+      phi_tot <- phi_tot + phi_m[[i]]
+  }
+  W <- (1/(nchains * (niter - 1))) * phi_tot
+  phi_bar <- apply(x, c(3), mean)
+  phi_diff1 <- map(chains_l, ~phi_j_bar[[.]] - phi_bar)
+  temp <- map(chains_l, ~as.matrix(phi_diff1[[.]], nrow=nparams) %*% phi_diff1[[.]])
+  for(i in chains_l){
+    if(i==1)
+      phi_tot1 <- temp[[i]]
+    else
+      phi_tot1 <- phi_tot1 + temp[[i]]
+  }
+  B <- niter / (nchains - 1) * phi_tot1
+  W_minus_1_B_n <- solve(W) %*% B / niter
+  y <- eigen(W_minus_1_B_n)
+  R_hat <- (nchains + 1) / nchains * max(abs(y$values)) + (niter - 1) / niter
+  return(R_hat)
+}

### 2020-03-27 src/s_cauchy.Rmd
added in plot of r star by quantile for 8 schools example

--- 2020-03-24 src/s_cauchy.Rmd
@@
 source("r_star_monitor.R")
-
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
-  library(grid)
-
-  # Make a list from the ... arguments and plotlist
-  plots <- c(list(...), plotlist)
-
-  numPlots = length(plots)
-
-  # If layout is NULL, then use 'cols' to determine layout
-  if (is.null(layout)) {
-    # Make the panel
-    # ncol: Number of columns of plots
-    # nrow: Number of rows needed, calculated from # of cols
-    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                    ncol = cols, nrow = ceiling(numPlots/cols))
-  }
-
- if (numPlots==1) {
-    print(plots[[1]])
-
-  } else {
-    # Set up the page
-    grid.newpage()
-    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
-
-    # Make each plot, in the correct location
-    for (i in 1:numPlots) {
-      # Get the i,j matrix positions of the regions that contain this subplot
-      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-
-      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                      layout.pos.col = matchidx$col))
-    }
-  }
-}
-
 ```
@@
   ylab(TeX("$R^2$")) +
-  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
   xlab("Iteration") +
@@
   ylab(TeX("$R^*$")) +
-  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
   xlab("Iteration") +
@@
   ylab(TeX("$\\hat{R}$")) +
-  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
-  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  scale_fill_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
   xlab("Iteration") +
@@
   scale_y_continuous(limits=c(0.99, NA), breaks = seq(1, 6, 0.2)) +
-  ggtitle(TeX("Max split-$\\hat{R}$"))
+  ggtitle(TeX("B. Max split-$\\hat{R}$"))
 g3
+```
+
+Multivariate Rhat
+```{r}
+fit1 <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+Rhat_alt <- map_dbl(iter, ~r_hat_multivariate(x[1:., , ]))
+
+fit1 <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+Rhat_nom <- map_dbl(iter, ~r_hat_multivariate(x[1:., , ]))
+g_df <- tibble(iteration=iter,
+       alt=Rhat_alt,
+       nom=Rhat_nom) 
+g4 <- 
+  tibble(iteration=iter,
+       alt=Rhat_alt,
+       nom=Rhat_nom) %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=value, colour=as.factor(variable))) +
+  geom_line(aes(y=value, colour=as.factor(variable))) +
+  ylab(TeX("$\\hat{R}$")) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  scale_fill_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_x_log10() +
+  scale_y_continuous(limits=c(0.99, NA), breaks = seq(1, 6, 0.2)) +
+  ggtitle(TeX("C. Multivariate split-$\\hat{R}$"))
+```
+
+Tail ESS
+```{r}
+fit1 <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+tail_ess_alt <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_tail(x[1:i, , .]))))
+bulk_ess_alt <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_bulk(x[1:i, , .]))))
+
+fit1 <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit1, permuted=F, pars="x")
+tail_ess_nom <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_tail(x[1:i, , .]))))
+bulk_ess_nom <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_bulk(x[1:i, , .]))))
+
+e_df <- tibble(iteration=iter,
+               alt=tail_ess_alt,
+               nom=tail_ess_nom)
+saveRDS(e_df, "../output/cauchy_tail_ess.rds")
+f_df <- tibble(iteration=iter,
+               alt=bulk_ess_alt,
+               nom=bulk_ess_nom)
+saveRDS(f_df, "../output/cauchy_bulk_ess.rds")
+e_df <- readRDS("../output/cauchy_tail_ess.rds")
+f_df <- readRDS("../output/cauchy_bulk_ess.rds")
+g5 <- 
+  e_df %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=value, colour=as.factor(variable))) +
+  geom_line(aes(y=value, colour=as.factor(variable))) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  scale_fill_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 400, linetype=2) +
+  scale_x_log10() +
+  scale_y_log10() +
+  ggtitle(TeX("E. Tail ESS"))
+
+g6 <- 
+  f_df %>% 
+  melt(id.vars="iteration") %>% 
+  ggplot(aes(x=iteration)) +
+  geom_point(aes(y=value, colour=as.factor(variable))) +
+  geom_line(aes(y=value, colour=as.factor(variable))) +
+  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=11, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.position = "none",
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 400, linetype=2) +
+  scale_x_log10() +
+  scale_y_log10() +
+  ggtitle(TeX("F. Bulk ESS"))
 ```

### 2020-03-27 src/s_eight_schools.Rmd
added in plot of r star by quantile for 8 schools example

@@
 ```{r}
-full_data_all <- rstan::extract(fit_cp, permuted=F)
-full_data <- array(dim=c(500, 8, 11))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
+x <- rstan::extract(fit_cp, permuted=F)
 
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = 12)
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-both <- f_fit_gbm_predict(full_data)
-
-plda <- both$predicted_prob
-testing_data <- both$testing_data
-
-nsim <- 1000
-r_star_centered <- f_r_star_distribution(nsim, plda, testing_data)
+r_star_centered <- r_star_dist(x)
 ```
@@
 ```{r}
-full_data_all <- rstan::extract(fit_ncp, permuted=F)
-nparams <- 19
-full_data <- array(dim=c(500, 8, nparams))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
-
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = (nparams+1))
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V20) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-both <- f_fit_gbm_predict(full_data)
-
-plda <- both$predicted_prob
-testing_data <- both$testing_data
-
-nsim <- 1000
-r_star_noncentered <- f_r_star_distribution(nsim, plda, testing_data)
+x <- rstan::extract(fit_ncp, permuted=F)
+r_star_noncentered <- r_star_dist(x)
 ```
@@
 ```{r}
-full_data <- rstan::extract(fit_cp, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 12)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
-
-both <- f_fit_gbm_predict(m_flattened)
-
-plda <- both$predicted_prob
-testing_data <- both$testing_data
-
-nsim <- 1000
-r_star_centered <- f_r_star_distribution(nsim, plda, testing_data)
+x <- rstan::extract(fit_cp, permuted=F)
+r_star_centered <- r_star_dist(x)
 
 # noncentered
-full_data <- rstan::extract(fit_ncp, permuted=F)
-k <- 1
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams+1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V20) %>% 
-  mutate(chain=as.factor(chain))
-both <- f_fit_gbm_predict(m_flattened)
-plda <- both$predicted_prob
-testing_data <- both$testing_data
-
-nsim <- 1000
-r_star_noncentered <- f_r_star_distribution(nsim, plda, testing_data)
+x <- rstan::extract(fit_ncp, permuted=F)
+r_star_noncentered <- r_star_dist(x)
 ```
@@
 
-# Plot tau prediction accuracy across quantiles
+# Plot R* versus quantiles
 ```{r}
-full_data_all <- rstan::extract(fit_cp, permuted=F)
-full_data <- array(dim=c(500, 8, 11))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
+x <- rstan::extract(fit_cp, permuted=F)
+g1 <- plot_r_star_quantiles(1, x, nsim=40, split_chains = F) +
+  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6))
 
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = 12)
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
+g2 <- plot_r_star_quantiles(2, x, nsim=40, split_chains = F) +
+  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6))
 
-full_data <- m_flattened
-both <- f_fit_gbm_predict(full_data)
+library(ggExtra)
+f <- ecdf(x[, , 2])
+p <- x[, , 2] %>% 
+  melt(id.vars="NULL") %>% 
+  mutate(value=f(value)) %>% 
+  ggplot(aes(x=iterations, y=value, colour=as.factor(chains))) +
+  geom_line(alpha=0.8) +
+  geom_point(aes(size=NA)) +
+  scale_size_identity() +
+  scale_color_brewer(palette = "Dark2") +
+  theme(text = element_text(colour="black", size=11),
+          axis.text = element_text(colour="black", size=10),
+        legend.position = "none",
+        axis.title.x = element_blank()) +
+  xlab("Iteration") +
+  ylab("Quantile") +
+  ylim(0, 1) +
+  coord_flip()
+ga <- ggMarginal(p, type="density", margins="x", groupColour = T,
+                 xparams = list(bw=0.05))
 
-plda <- both$predicted_prob
-testing_data <- both$testing_data
+f <- ecdf(x[, , 1])
+p <- x[, , 1] %>% 
+  melt(id.vars="NULL") %>% 
+  mutate(value=f(value)) %>% 
+  ggplot(aes(x=iterations, y=value, colour=as.factor(chains))) +
+  geom_line(alpha=0.8) +
+  geom_point(aes(size=NA)) +
+  scale_size_identity() +
+  scale_color_brewer(palette = "Dark2") +
+  theme(text = element_text(colour="black", size=11),
+          axis.text = element_text(colour="black", size=10),
+        legend.position = "none",
+        axis.title.x = element_blank()) +
+  xlab("Iteration") +
+  ylab("Quantile") +
+  ylim(0, 1) +
+  coord_flip()
+gb <- ggMarginal(p, type="density", margins="x", groupColour = T,
+                 xparams = list(bw=0.05))
 
-f_r_star_accuracy <- function(nsim, pred_test_prob, testing_data){
+library(cowplot)
 
-  mAccuracy <- matrix(nrow = nrow(pred_test_prob),
-                    ncol = nsim)
-  for(j in 1:nrow(pred_test_prob)){
-    test <- apply(rmultinom(nsim, 1, prob = pred_test_prob[j, ]), 2, function(x) which(x==1))
-    mAccuracy[j, ] <- if_else(test==testing_data$chain[j], 1, 0)
-  }
-  return(mAccuracy)
-}
-
-nsim <- 20
-mAccuracy <- f_r_star_accuracy(nsim, plda, testing_data)
-colnames(testing_data) <- c("mu", "tau",
-                            "theta.1", "theta.2",
-                            "theta.3", "theta.4",
-                            "theta.5", "theta.6",
-                            "theta.7", "theta.8",
-                            "lp", "chain")
-colnames(mAccuracy) <- map_chr(seq(1, nsim, 1), ~paste0("V", .))
-mAccuracy <- mAccuracy %>% 
-  as.data.frame() %>% 
-  bind_cols(testing_data)
-
-lbreaks <- quantile(mAccuracy$tau, seq(0, 1, 0.1))
-mAccuracy$tau_cuts <- cut(mAccuracy$tau, breaks=lbreaks, include.lowest = T, right=T, ordered_result = T)
-cuts <- sort(unique(mAccuracy$tau_cuts))
-r_star <- matrix(ncol = length(cuts),
-                 nrow=nsim)
-for(i in seq_along(cuts)){
-  temp <- mAccuracy %>% 
-    filter(tau_cuts==cuts[i])
-  r_star[, i] <- colMeans(temp[, 1:nsim]) * n_distinct(testing_data$chain)
-}
-colnames(r_star) <- seq(0.1, 1, 0.1)
-r_mean <- 
-  r_star %>% 
-  melt(id.vars=NULL) %>%  
-  group_by(Var2) %>% 
-  summarise(value=mean(value))
-r_full <- 
-  r_star %>%
-  melt(id.vars=NULL) %>%
-  bind_rows(r_mean) %>% 
-  mutate(variable=as.numeric(as.character(Var2)))
-
-ggplot(filter(r_full, !is.na(Var1)),
-       aes(x=variable, y=value)) +
-  geom_jitter(height = 0, width = 0.02, colour="grey") +
-  ylab(TeX("$R*$")) +
-  xlab("Quantile") +
-  geom_hline(yintercept = 1, linetype=2) +
-  theme(text = element_text(colour="black",
-                            size=12)) +
-  geom_line(data=filter(r_full, is.na(Var1)))
+pdf("../output/eight_schools_r_star_quantiles.pdf", width = 6, height = 6)
+plot_grid(arrangeGrob(gb, top="A."), arrangeGrob(ga, top="B."), g1, g2,nrow = 2, rel_heights = c(1/2, 1/4))
+dev.off()
 ```

### 2020-03-27 write_up/ms.tex
added in plot of r star by quantile for 8 schools example

@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-	\caption{\textbf{Eight schools example.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
+	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
 In all cases, the plots show 1000 $R^*$ samples.
@@
 	\label{fig:eight_schools}
+\end{figure}
+
+To understand the predictive power of the ML better, we investigate how predictive accuracy varies across parameter space.
+After fitting the GBM model, we group MCMC samples into deciles and sample from the $R^*$ distribution for each decile.
+In Fig.
+\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
+In the top row of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
+Above the top panel, we show the marginal distributions for each chain.
+In the bottom row, we show 40 $R^*$ samples for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all the samples.
+In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
+For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
+Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved ML predictive accuracy.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools_r_star_quantiles.pdf}}
+	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
+In each case, we show the path of the four individual chains above and 40 $R^*$ samples by parameter quantile below.
+Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
+Note that, in the bottom plots, jitter has been added to the data points.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:eight_schools_r_star_quantiles}
 \end{figure}

### 2020-03-27 src/r_star_monitor.R

@@ -222,3 +222,146 @@ r_hat_multivariate <- function(x, split_chains=T){
   R_hat <- (nchains + 1) / nchains * max(abs(y$values)) + (niter - 1) / niter
   return(R_hat)
 }
+
+#' Plot R* versus parameter quantiles
+#'
+#' @param index index of parameter in x
+#' 
+#' @param x a 3D array of samples of dimensions: # iter x # chains x # parameters
+#' 
+#' @param split_chains a Boolean (defaults to true) indicating whether to split
+#' chains into two equal halves
+#'
+#' @param training_percent proportion of iterations used to train GBM model
+#' (default is 0.7)
+#'
+#' @param caret_default GBM hyperparameter settings given as a list (defaults
+#' are interaction.depth=3, n.trees=50, shrinkage=0.1, nminobsinmode=10)
+#'
+#' @param nsim number of samples of R* to generate from GBM fit (defaults to
+#' 20)
+#'
+#' @return distribution of R* values
+#' 
+#' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
+#' diagnostic with uncertainty using gradient-boosted machines
+#' \emph{arXiv preprint} \code{arXiv:TBD}
+plot_r_star_quantiles <- function(index, x, split_chains=T, training_percent=0.7, caret_default=NA, nsim=20){
+  if(split_chains)
+    x <- split_data(x)
+  
+  nparams <- dim(x)[3]
+  nchains <- dim(x)[2]
+  niter <- dim(x)[1]
+  m_flattened <- matrix(nrow = nrow(x) * nchains,
+                        ncol = (nparams + 1))
+  k <- 1
+  for(i in 1:nchains){
+    for(j in 1:nrow(x)){
+      m_flattened[k, 1:nparams] <- x[j, i, ]
+      m_flattened[k, (nparams + 1)] <- i
+      k <- k + 1
+    }
+  }
+  m_flattened <- m_flattened %>% 
+    as.data.frame()
+  colnames(m_flattened)[nparams + 1] <- "chain"
+  r <- m_flattened %>% 
+    mutate(chain=as.factor(chain))
+  
+  rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
+  training_data <- r[rand_samples, ]
+  testing_data <- r[-rand_samples, ]
+  
+  if(is.na(caret_default))
+    caretGrid <- expand.grid(interaction.depth=c(3), 
[…]

### 2020-03-27 src/s_cauchy.Rmd
added more notes

@@
         axis.text = element_text(size=10, colour="black"),
-        legend.position = c(0.8, 0.8),
+        legend.position = c(0.8, 0.7),
         axis.title.y = element_blank()) +
@@
   geom_line(aes(y=value, colour=as.factor(variable))) +
-  scale_color_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
-  scale_fill_brewer("Parameterisation", palette = "Dark2", labels=c("Alternative", "Nominal")) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  scale_fill_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
   xlab("Iteration") +

### 2020-03-27 src/s_eight_schools.Rmd
added more notes

@@
 source("monitornew.R")
-
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-
-f_fit_gbm_predict <- function(r){
-  rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
-  training_data <- r[rand_samples, ]
-  testing_data <- r[-rand_samples, ]
-
-  gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-  plda <- predict(object=gbmFit1, newdata=testing_data)
-  plda_prob <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-  a_accuracy <- 
-    tibble(predicted=plda, actual=testing_data$chain) %>%
-    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-    summarise(mean(correct)) %>% 
-    pull()
-  return(list(accuracy=a_accuracy, predicted_prob=plda_prob, testing_data=testing_data))
-}
-
-f_r_star_distribution <- function(nsim, pred_test_prob, testing_data){
-
-  mAccuracy <- matrix(nrow = nrow(pred_test_prob),
-                    ncol = nsim)
-  for(j in 1:nrow(pred_test_prob)){
-    test <- apply(rmultinom(nsim, 1, prob = pred_test_prob[j, ]), 2, function(x) which(x==1))
-    mAccuracy[j, ] <- if_else(test==testing_data$chain[j], 1, 0)
-  }
-  return(colMeans(mAccuracy) * n_distinct(testing_data$chain))
-}
-
+source("r_star_monitor.R")
 source("eight_schools.data.R")

### 2020-03-27 src/s_trends.Rmd
added in trends examples

@@
 res <- readRDS("../output/experiment_trends_all_dim.rds")
-ggplot(data = res, aes(y = Rstar_split, x = trend)) + 
+g <- 
+  res %>%
+  select(iters, trend, Rstar, Rstar_split) %>% 
+  melt(id.vars=c("iters", "trend")) %>% 
+  mutate(variable=case_when(variable=="Rstar"~"A. R*",
+                            variable=="Rstar_split"~"B. split-R*")) %>% 
+  ggplot(aes(y = value, x = trend)) + 
   geom_point() + 
   geom_jitter() + 
-  facet_grid(. ~ iters) + 
+  facet_grid(vars(iters), vars(variable)) + 
   geom_hline(yintercept = 1) +
-  ggtitle('Split-R*')
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black")) +
+  xlab("Trend") +
+  ylab("Value")
+ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)
 ```
@@
 ```{r}
-res <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
-  select(trend, iters, rep, dimension, Rstar) %>% 
-  unique()
-ggplot(data = res, aes(x = trend, y = Rstar)) + 
+g <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
+  select(trend, iters, rep, dimension, Rstar_split) %>% 
+  unique() %>%
+  select(-rep) %>% 
+  melt(c("trend", "iters", "dimension")) %>% 
+  filter(variable=="Rstar_split") %>% 
+  ggplot(aes(y = value, x = trend)) + 
   geom_point() + 
@@
   geom_hline(yintercept = 1) +
-  ggtitle('R* without splitting')
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black")) +
+  xlab("Trend") +
+  ylab("Value")
+g
+
+ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
+```
+
+# Trend in the joint distribution over time
+```{r}
+library(mvtnorm)
+rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
+  return(rmvnorm(n, c(mux, muy),
+                 matrix(c(sigmax^2, sigmax * sigmay * rho,
+                          sigmax * sigmay * rho, sigmay^2),
+                        ncol = 2)))
+}
+
+conds <- expand.grid(
+  iters = c(250, 1000, 4000), 
+  rep = 1:10,
+  rho = c(0, 0.2, 0.4, 0.6, 0.8)
+)
+res <- vector("list", nrow(conds))
+chains = 4
+
+for (i in 1:nrow(conds)) {
+  iters <- conds[i, "iters"]
+  rep <- conds[i, "rep"]
+  rho <- conds[i, "rho"]
+  rho_trend <- seq(-rho, rho, length.out = iters)
+  r <- array(dim=c(iters, chains, 2))
+  for(j in 1:chains)
+    for(jj in 1:iters)
+      r[jj, j, ] <- rmvrnorm2D(1, 0, 0, 1, 1, rho_trend[jj])
+  # take max as just using the Rhat values
+  rs <- as.data.frame(monitor_extra(r)) %>% 
+    apply(2, max)
+
+ r_2d <- matrix(nrow = iters * chains, ncol=(2 + 2))
+  k <- 1
+  for(ii in 1:iters){
+    for(j in 1:chains){
+      r_2d[k, ] <- c(ii, j, r[ii, j, ])
+      k <- k + 1
+    }
+  }
+  r_2d <- r_2d %>%
+    as.data.frame()
+  colnames(r_2d)[1:2] <- c("iter", "chain")
+
+  # R* calculation
+  r1 <- r_2d %>% 
+    arrange(chain, iter) %>% 
+    select(-iter) %>% 
+    mutate(chain=as.factor(chain))
+  a_accuracy <- f_fit_gbm_predict(r1)
+
+  # Split R* calculation
+  half_rows <- iters / 2
+  r_1h <- r_2d %>% 
+    filter(iter<=half_rows)
+  r_2h <- r_2d %>% 
+    filter(iter>half_rows) %>% 
+    mutate(chain=chain + chains)
+  r2 <- bind_rows(r_1h, r_2h) %>% 
+    select(-iter) %>% 
+    mutate(chain=as.factor(chain))
+  b_accuracy <- f_fit_gbm_predict(r2)
+
+  res[[i]] <- cbind(iters, rep, rho, t(rs), "Rstar"=a_accuracy * n_distinct(r1$chain),
+                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
+}
+res1 <- do.call(rbind, res)
+saveRDS(res1, "../output/experiment_trends_joint_distribution.rds")
 ```
@@
 ```{r}
-res <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
-  select(trend, iters, rep, dimension, Rstar_split) %>% 
-  unique()
-ggplot(data = res, aes(x = trend, y = Rstar_split)) + 
-  geom_point() + 
+g <- readRDS("../output/experiment_trends_joint_distribution.rds") %>%
+  as.data.frame() %>% 
+  select(iters, rho, rep, zsRhat, Rstar_split) %>% 
+  unique() %>%
+  select(-rep) %>% 
+  melt(c("rho", "iters"))  %>% 
+  mutate(variable=case_when(variable=="zsRhat"~"A. split-Rhat",
+                            variable=="Rstar_split"~"B. split-R*")) %>% 
+  ggplot(aes(y = value, x = rho)) + 
   geom_jitter() + 
-  facet_grid(vars(iters), vars(dimension)) + 
+  facet_grid(vars(iters), vars(variable)) + 
   geom_hline(yintercept = 1) +
-  ggtitle('Split-R*')
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black")) +
+  ylab("Value") +
+  xlab(TeX("$\\rho$"))
+g
+
+ggsave("../output/trends_joint_distribution.pdf", g, width = 8, height = 6)
 ```
@@
 
-ggplot(data = res, aes(x = rho, y = Rstar)) + 
+ggplot(data = res, aes(x = rho, y = Rhat)) + 
   geom_jitter() + 

### 2020-03-27 write_up/ms.tex
added in trends examples

@@
 \subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
-We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
+We generate four Markov chains, where each samples from an autoregressive order 1 (AR(1)) process of the form,
 %
@@
 %
-where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$ and $t=1,2,...,1000$.
+where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,1000$.
 Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others.
@@
 
+\subsection{Non-stationary marginals}
+To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}.
+This example showed that split-$\hat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
+Specifically, they first generated four chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling.
+We first repeat this analysis but using $R^*$ rather than $\hat{R}$: in Fig.
+\ref{fig:trends_all_dim}, we show these results.
+In column A, we show the results for $R^*$ calculated on the four chains that were run; column B shows the same calculation but after the chains are split into two equal halves.
+The different rows show the different sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
+This plot mirrors Fig.
+18 in \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does.
+As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.
+
+These results are what would intuitively be expected: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation.
+In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals.
+By splitting chains into two -- the first half being the early phase and, the second half being the later phase with higher mean -- this forces differences in the marginals.
+This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one.
+As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \cite{carpenter2017stan,vehtari2019rank}.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_all_dim.pdf}}
+	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the four chains; column B  shows the same calculation after each chain is split into two halves.
+The rows present the differing sample sizes.
+The horizontal axis measures (half) the change in mean across the whole sample: so a value ``1'' indicates the mean increases by 2 units from the start to end of sampling.
+At each parameter set, 10 replicates were run and jitter was added to the points.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:trends_all_dim}
+\end{figure}
+
+We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends.
+In Fig.
+\ref{fig:trends_one_dim}, we show the effect on $R^*$ of trends of different magnitudes across a range of target dimensions.
+In the simulations here, all dimensions bar one are stationary; the remaining dimension has a trend added to it.
+In all cases, split-$R^*$ increased with trend.
+Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered.
+This suggests split-$R^*$ can robustly determine chain identity if only a single dimension has a non-stationary sampling distribution.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_one_dim.pdf}}
+	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns represent different dimensionalities of the target distribution; the rows present different sample sizes.
+The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
+At each parameter set, 10 replicates were run and jitter was added to the points.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:trends_one_dim}
+\end{figure}
+
+Means that trends over time is one form of non-stationarity; another is a time-varying covariance.
+Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time.
+Specifically, we allow the correlation to increase from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
+samples from the process across 4 ``chains''.
+Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution.
+Similarly, split-$\hat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig.
+\ref{fig:trends_joint_distribution}A).
+By contrast, split-$R^*$ can (Fig.
+\ref{fig:trends_joint_distribution}B), since it uses all information in the samples, including the covariance structure.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_joint_distribution.pdf}}
+	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\hat{R}$; column B for $R^*$.
+The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
+At each parameter set, 10 replicates were run and jitter was added to the points.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:trends_joint_distribution}
+\end{figure}
+
+A subtler form of non-stationarity can occur as the autocorrelation of a series increases.
+For example, if consider an AR(1) process, $X_t = \rho X_{t-1} + \epsilon_t$, where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$.
+For $|\rho|<1$, the process has a stationary variance: $\sigma^2 / (1 - \rho^2)$.
+If $|\rho|=1$, however, the process becomes a random walk, with a variance that increases linearly with time.
+Any measure of convergence should, therefore, determine that the sampling distribution of an AR(1) process with $|\rho=1|$ is not converged since, by construction, such processes do not have a stationary target distribution.
+Any $\rho<1$ has a stationary distribution and, if sample sizes suffice,
+
 \subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}

### 2020-03-27 write_up/ms.tex
removed AR1 trends example

@@
 
-A subtler form of non-stationarity can occur as the autocorrelation of a series increases.
-For example, if consider an AR(1) process, $X_t = \rho X_{t-1} + \epsilon_t$, where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$.
-For $|\rho|<1$, the process has a stationary variance: $\sigma^2 / (1 - \rho^2)$.
-If $|\rho|=1$, however, the process becomes a random walk, with a variance that increases linearly with time.
-Any measure of convergence should, therefore, determine that the sampling distribution of an AR(1) process with $|\rho=1|$ is not converged since, by construction, such processes do not have a stationary target distribution.
-Any $\rho<1$ has a stationary distribution and, if sample sizes suffice,
-
 \subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}

### 2020-03-28 src/s_ar1.Rmd
redraft of article

--- 2020-03-17 src/s_examples.Rmd
@@
 source("monitornew.R")
+source("r_star_monitor.R")
+
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
 ```
@@
 }
-
-f_generate_lower_var_four(1/10, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  ggplot(aes(x=value, fill=as.factor(chain))) +
-  geom_histogram(position="identity", alpha=0.3)
-```
-
-Use gbm model in Caret to predict chain identity
-```{r}
-library(caret)
-library(gbm)
-
-# generate data
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-
-# add in extra column of nonsense values since gbm needs 2+ vars
-full_data <- full_data %>% 
-  mutate(chain=as.factor(chain)) %>% 
-  mutate(ones=rnorm(nrow(full_data)))
-
-# create training/testing data
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid)
-# plot(gbmFit1)
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-# plot predictions as fn of value
-testing_data %>% 
-  arrange(value) %>% 
-  mutate(predictions=predict(object=gbmFit1, .)) %>% 
-  ggplot(aes(x=value, y=predictions)) +
-  geom_jitter() +
-  coord_flip()
 ```
@@
 ```{r}
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
-  library(grid)
-
-  # Make a list from the ... arguments and plotlist
-  plots <- c(list(...), plotlist)
-
-  numPlots = length(plots)
-
-  # If layout is NULL, then use 'cols' to determine layout
-  if (is.null(layout)) {
-    # Make the panel
-    # ncol: Number of columns of plots
-    # nrow: Number of rows needed, calculated from # of cols
-    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                    ncol = cols, nrow = ceiling(numPlots/cols))
-  }
-
- if (numPlots==1) {
-    print(plots[[1]])
-
-  } else {
-    # Set up the page
-    grid.newpage()
-    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
-
-    # Make each plot, in the correct location
-    for (i in 1:numPlots) {
-      # Get the i,j matrix positions of the regions that contain this subplot
-      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-
-      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                      layout.pos.col = matchidx$col))
-    }
-  }
-}
 
@@
 }
-hist(colMeans(mAccuracy1)/0.25)
 
@@
 
-pdf("../output/ar1.pdf", width = 12, height = 6)
+# pdf("../output/ar1.pdf", width = 12, height = 6)
 multiplot(g1, g2, g3, cols = 3)
-dev.off()
+# dev.off()
 ```
-
-# Bivariate normal
-```{r}
-rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
-  return(rmvnorm(n, c(mux, muy),
-                 matrix(c(sigmax^2, sigmax * sigmay * rho,
-                          sigmax * sigmay * rho, sigmay^2),
-                        ncol = 2)))
-}
-# calling function
-n <- 1000
-chain1 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
-  as.data.frame() %>% 
-  mutate(chain=1,
-         iter=seq_along(chain))
-chain2 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
-  as.data.frame() %>% 
-  mutate(chain=2,
-         iter=seq_along(chain))
-chain3 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
-  as.data.frame() %>% 
-  mutate(chain=3,
-         iter=seq_along(chain))
-chain4 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.9) %>% 
-  as.data.frame() %>% 
-  mutate(chain=4,
-         iter=seq_along(chain))
-
-chains_stacked <- rbind.data.frame(chain1, chain2, chain3, chain4) %>% 
-  mutate(chain=as.factor(chain))
-chains_stacked1 <- chains_stacked %>% 
-  select(-iter)
-
-full_data <- chains_stacked1
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-# 
-# # use Vehtari et al. script to calculate Rhat
-# chains_wide1 <- chains_stacked %>% 
-#   pivot_wider(values_from = c("V1"), names_from = c("chain"), id_cols = "iter")
-# rhat_rfun(as.matrix(chains_wide1[, 2:5]))
-# chains_wide2 <- chains_stacked %>% 
-#   pivot_wider(values_from = c("V2"), names_from = c("chain"), id_cols = "iter")
-# rhat_rfun(as.matrix(chains_wide2[, 2:5]))
-# 
-# # Use Vehtari et al. to calculate ess
-# ess_rfun(as.matrix(chains_wide1[, 2:5]))
-# ess_rfun(as.matrix(chains_wide2[, 2:5]))
-# ess_bulk(as.matrix(chains_wide1[, 2:5]))
-# ess_bulk(as.matrix(chains_wide2[, 2:5]))
-# ess_tail(as.matrix(chains_wide1[, 2:5]))
-# ess_tail(as.matrix(chains_wide2[, 2:5]))
-# ess_quantile(as.matrix(chains_wide1[, 2:5]), 0.95)
-# ess_tail(as.matrix(chains_wide2[, 2:5]))
-
-# Try monitor functionality -- split chains first
-half_iter <- max(chains_stacked$iter) / 2
-m_samples <- array(dim = c(half_iter,
-                           2*n_distinct(chains_stacked$chain),
-                           2))
-k <- 1
-chains_stacked_first <- chains_stacked %>% 
-  filter(iter<=half_iter)
-chains_stacked_second <- chains_stacked %>% 
-  filter(iter>half_iter)
-
-for(i in 1:half_iter){
-  temp <- chains_stacked_first %>% 
-    filter(iter==i)
-  for(j in 1:4){
-    temp1 <- temp %>%
-      filter(chain==j)
-    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
-    k <- k + 1
-  }
-}
-
-for(i in 1:half_iter){
-  temp <- chains_stacked_second %>% 
-    filter(iter==(i + 500))
-  for(j in 5:8){
-    temp1 <- temp %>%
-      filter(chain==(j - 4))
-    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
-    k <- k + 1
-  }
-}
-
-g <- 
-  qplot(4*colMeans(mAccuracy)) +
-  xlab(TeX("$R*$")) +
-  xlim(0.9, NA) +
-  geom_vline(xintercept = 1, linetype=2) +
-  ylab("Count") +
-  theme(text=element_text(size=14, colour="black"))
-ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
-mon <- monitor(m_samples)
-
-```
-
-# 250-dimensional multivariate normal
-Fit model in Stan with 10000 iterations
-```{r}
-N <- 250
-A <- rWishart(1, 250, diag(N))[,,1]
-saveRDS(A, "../output/A_matrix.rds")
-model <- stan_model("mvt_250.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
-print(fit)
-# saveRDS(fit, "../output/mvt_fit_10000.rds")
-fit <- readRDS("../output/mvt_fit_10000.rds")
-full_data <- rstan::extract(fit, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:251] <- full_data[j, i, ]
-    m_flattened[k, 252] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V252) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- tibble(samples_10000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-
-importance <- 100-varImp(gbmFit1)$importance$Overall
-rhat <- summary(fit)$summary[, 10]
-ess <- summary(fit)$summary[, 9]
-plot(importance, rhat)
-plot(importance, ess)
-b_df <- tibble(importance, ess, rhat, lp=c(rep(0, 250), 1))
-sum(rhat<1.01)
-sum(ess>400)
-gc <- 
-  b_df %>%
-  ggplot(aes(importance, ess)) +
-  geom_jitter(width=1, aes(shape=as.factor(lp))) +
-  geom_smooth(span=1) +
-  xlab("Variable importance") +
-  ylab("ESS") +
-  ggtitle("C.") +
-  theme(legend.position = "none")
-
-gb <- b_df %>%
-  ggplot(aes(importance, rhat)) +
-  geom_jitter(width = 1, aes(shape=as.factor(lp))) +
-  geom_smooth(span=1) +
-  xlab("Variable importance") +
-  ylab(TeX("$\\hat{R}$")) +
-  ggtitle("B.") +
-  theme(legend.position = "none")
-```
-
-With 400 iterations
-```{r}
-fit <- sampling(model, data=list(N=N, A=A), iter=400, chains=4)
-print(fit)
-# saveRDS(fit, "../output/mvt_fit_400.rds")
-fit <- readRDS("../output/mvt_fit_400.rds")
-
-full_data <- rstan::extract(fit, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:251] <- full_data[j, i, ]
-    m_flattened[k, 252] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V252) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- a_df %>%
-  mutate(samples_400=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-```
-
-```{r}
-model <- stan_model("mvt_250_ncp.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
-print(fit)
-# saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
-fit <- readRDS("../output/mvt_ncp_fit_10000.rds")
-
-full_data <- rstan::extract(fit, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams + 1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V502) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- a_df %>%
-  mutate(samples_10000_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-
-saveRDS(a_df, "../output/mvt_fit_all_accuracy.rds")
-
-ga <- 
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of samples", labels=c("400", "10000, cp", "10000, ncp")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8)) +
-  ggtitle("A.")
-
-pdf("../output/mvt_three.pdf", width = 12, height = 8)
-multiplot(ga, gb, gc, cols = 3)
-dev.off()
-```
-
-# Cauchy example
-
-Generate R* distribution from nominal Cauchy fit
-```{r}
-nsim <- 1000
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0)
-full_data <- rstan::extract(fit_nom, permuted=F)
-nparams <- 52
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V53) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_nom <- mAccuracy
-```
-
-Generate R* distribution from alternative Cauchy parameterisation
-```{r}
-fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0)
-full_data <- rstan::extract(fit_alt1, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V153) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_alt_1 <- mAccuracy
-```
-
-```{r}
-a_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
-               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
-
-g <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none")
-g
-
-ggsave("../output/cauchy.pdf", g, width = 8, height = 6)
-```
-
-## More iterations
-Nominal model with 10000 iterations and thinning by 10
-```{r}
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0, iter = 10000,
-                thin = 5)
-full_data <- rstan::extract(fit_nom, permuted=F)
-nparams <- 52
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V53) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-mAccuracy_nom <- mAccuracy
-```
-
-Alternative model with 10000 iterations and thinning by 10
-```{r}
-fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0, iter = 10000,
-                thin = 5)
-full_data <- rstan::extract(fit_alt1, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V153) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_alt_1 <- mAccuracy
-```
-
-Plot
-```{r}
-b_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
-               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
-
-g1 <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none") +
-  ggtitle("A.")
-g1
-
-g2 <-
-  b_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black")) +
-  ggtitle("B.")
-g2
-
-pdf("../output/cauchy.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
-```
-
-# Eight schools
-```{r}
-source("eight_schools.data.R")
-eight_schools <- list(J=J, y=y, sigma=sigma)
-model_cp <- stan_model("eight_schools_cp.stan")
-model_ncp <- stan_model("eight_schools_ncp.stan")
-```
-
-Run models
-```{r}
-fit_cp <- sampling(
-  model_cp, data = eight_schools,
-  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
-  control = list(adapt_delta = 0.95)
-)
-
-fit_ncp <- sampling(
-  model_ncp, data = eight_schools,
-  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
-  control = list(adapt_delta = 0.95)
-)
-```
-
-## Break each chain into two then make flattened data
-Centered
-```{r}
-full_data_all <- rstan::extract(fit_cp, permuted=F)
-full_data <- array(dim=c(500, 8, 11))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
-
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = 12)
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_centered <- mAccuracy
-```
-
-Noncentered
-```{r}
-full_data_all <- rstan::extract(fit_ncp, permuted=F)
-nparams <- 19
-full_data <- array(dim=c(500, 8, nparams))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
-
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = (nparams+1))
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V20) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_noncentered <- mAccuracy
-```
-
-Store both
-```{r}
-c_df <- tibble(centered=colMeans(mAccuracy_centered)/0.125,
-       non_centered=colMeans(mAccuracy_noncentered)/0.125)
-```
-
-## Trying with original data (i.e. 4 chains)
-```{r}
-full_data <- rstan::extract(fit_cp, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 12)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_centered <- mAccuracy
-
-full_data <- rstan::extract(fit_ncp, permuted=F)
-k <- 1
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams+1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V20) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_noncentered <- mAccuracy
-```
-
-```{r}
-d_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
-       non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
-```
-
-Plot two R* distributions
-```{r}
-g1 <- 
-  c_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  xlim(0.78, 1.6) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("A.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none")
-g2 <- 
-  d_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlim(0.78, 1.6) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black"))
-a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
-
-pdf("../output/eight_schools.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
-```

### 2020-03-28 src/s_bivariate_normal.Rmd
redraft of article

--- 2020-03-17 src/s_examples.Rmd
@@
 ---
-title: "Examples in R* paper"
-author: "Ben Lambert"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+title: "Bivariate normal"
 output: html_notebook
@@
 source("monitornew.R")
-```
-
-# AR1 example
-Functions to generate simulated data
-```{r}
-f_ar1 <- function(rho, sigma, L){
-  x <- vector(length = L)
-  x[1] <- rnorm(1, 0, sd=sigma)
-  for(i in 2:L)
-    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
-  return(x)
-}
-
-f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:3)
-    x[, i] <- f_ar1(rho, sigma, L)
-  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
-  x[, 4] <- z
-  return(x)
-}
-
-f_generate_lower_var_four(1/10, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  ggplot(aes(x=value, fill=as.factor(chain))) +
-  geom_histogram(position="identity", alpha=0.3)
-```
-
-Use gbm model in Caret to predict chain identity
-```{r}
-library(caret)
-library(gbm)
-
-# generate data
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-
-# add in extra column of nonsense values since gbm needs 2+ vars
-full_data <- full_data %>% 
-  mutate(chain=as.factor(chain)) %>% 
-  mutate(ones=rnorm(nrow(full_data)))
-
-# create training/testing data
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
+source("r_star_monitor.R")
 
@@
                    n.minobsinnode=10)
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid)
-# plot(gbmFit1)
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-# plot predictions as fn of value
-testing_data %>% 
-  arrange(value) %>% 
-  mutate(predictions=predict(object=gbmFit1, .)) %>% 
-  ggplot(aes(x=value, y=predictions)) +
-  geom_jitter() +
-  coord_flip()
 ```
 
-Perform 1000 replicates
-```{r}
-f_replicate <- function(){
-  full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-  rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-  training_data <- full_data[rand_samples, ]
-  testing_data <- full_data[-rand_samples, ]
-
-  # add in extra column of nonsense values since the method needs at least 2 vars 
-  full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-  training_data <- full_data[rand_samples, ]
-  testing_data <- full_data[-rand_samples, ]
-
-  gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-  plda <- predict(object=gbmFit1, newdata=testing_data)
-  a_accuracy <- 
-    tibble(predicted=plda, actual=testing_data$chain) %>%
-    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-    summarise(mean(correct)) %>% 
-    pull()
-  return(a_accuracy)
-}
-
-nreplicates <- 1000
-accuracies <- vector(length = nreplicates)
-for(i in 1:nreplicates){
-  if(i%%100==0)
-    print(i)
-  accuracies[i] <- f_replicate()
-}
-```
-
-Plot results
-```{r}
-multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
-  library(grid)
-
-  # Make a list from the ... arguments and plotlist
-  plots <- c(list(...), plotlist)
-
-  numPlots = length(plots)
-
-  # If layout is NULL, then use 'cols' to determine layout
-  if (is.null(layout)) {
-    # Make the panel
-    # ncol: Number of columns of plots
-    # nrow: Number of rows needed, calculated from # of cols
-    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
-                    ncol = cols, nrow = ceiling(numPlots/cols))
-  }
-
- if (numPlots==1) {
-    print(plots[[1]])
-
-  } else {
-    # Set up the page
-    grid.newpage()
-    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
-
-    # Make each plot, in the correct location
-    for (i in 1:numPlots) {
-      # Get the i,j matrix positions of the regions that contain this subplot
-      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
-
-      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
-                                      layout.pos.col = matchidx$col))
-    }
-  }
-}
-
-g2 <- tibble(accuracy=accuracies/0.25, iteration=seq(1, nreplicates, 1)) %>% 
-  ggplot(aes(x=accuracy)) +
-  geom_histogram() +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black")) +
-  xlim(1, NA)
-
-g1 <- testing_data %>% 
-  arrange(value) %>% 
-  mutate(predictions=predict(object=gbmFit1, .)) %>% 
-  ggplot(aes(x=value, y=predictions)) +
-  geom_jitter(height = 0.3) +
-  xlab("Value") +
-  ylab("Chain classification") +
-  theme(text = element_text(size=14, colour="black")) +
-  coord_flip() +
-  ggtitle("A.")
-```
-
-Generate R* distribution based on 1000 iterations
-```{r}
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-f_generate_all_same_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:4)
-    x[, i] <- f_ar1(rho, sigma, L)
-  return(x)
-}
-
-full_data <- f_generate_all_same_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy1 <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy1[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy1)/0.25)
-
-mAR1 <- tibble(unconverged=colMeans(mAccuracy)/0.25,
-               converged=colMeans(mAccuracy1)/0.25)
-g3 <- 
-  mAR1 %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series") +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
-
-pdf("../output/ar1.pdf", width = 12, height = 6)
-multiplot(g1, g2, g3, cols = 3)
-dev.off()
-```
-
-# Bivariate normal
 ```{r}
@@
 
-# 
-# # use Vehtari et al. script to calculate Rhat
-# chains_wide1 <- chains_stacked %>% 
-#   pivot_wider(values_from = c("V1"), names_from = c("chain"), id_cols = "iter")
-# rhat_rfun(as.matrix(chains_wide1[, 2:5]))
-# chains_wide2 <- chains_stacked %>% 
-#   pivot_wider(values_from = c("V2"), names_from = c("chain"), id_cols = "iter")
-# rhat_rfun(as.matrix(chains_wide2[, 2:5]))
-# 
-# # Use Vehtari et al. to calculate ess
-# ess_rfun(as.matrix(chains_wide1[, 2:5]))
-# ess_rfun(as.matrix(chains_wide2[, 2:5]))
-# ess_bulk(as.matrix(chains_wide1[, 2:5]))
-# ess_bulk(as.matrix(chains_wide2[, 2:5]))
-# ess_tail(as.matrix(chains_wide1[, 2:5]))
-# ess_tail(as.matrix(chains_wide2[, 2:5]))
-# ess_quantile(as.matrix(chains_wide1[, 2:5]), 0.95)
-# ess_tail(as.matrix(chains_wide2[, 2:5]))
-
 # Try monitor functionality -- split chains first
@@
   theme(text=element_text(size=14, colour="black"))
-ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
+# ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
 mon <- monitor(m_samples)
@@
 ```
-
-# 250-dimensional multivariate normal
-Fit model in Stan with 10000 iterations
-```{r}
-N <- 250
-A <- rWishart(1, 250, diag(N))[,,1]
-saveRDS(A, "../output/A_matrix.rds")
-model <- stan_model("mvt_250.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
-print(fit)
-# saveRDS(fit, "../output/mvt_fit_10000.rds")
-fit <- readRDS("../output/mvt_fit_10000.rds")
-full_data <- rstan::extract(fit, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:251] <- full_data[j, i, ]
-    m_flattened[k, 252] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V252) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- tibble(samples_10000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-
-importance <- 100-varImp(gbmFit1)$importance$Overall
-rhat <- summary(fit)$summary[, 10]
-ess <- summary(fit)$summary[, 9]
-plot(importance, rhat)
-plot(importance, ess)
-b_df <- tibble(importance, ess, rhat, lp=c(rep(0, 250), 1))
-sum(rhat<1.01)
-sum(ess>400)
-gc <- 
-  b_df %>%
-  ggplot(aes(importance, ess)) +
-  geom_jitter(width=1, aes(shape=as.factor(lp))) +
-  geom_smooth(span=1) +
-  xlab("Variable importance") +
-  ylab("ESS") +
-  ggtitle("C.") +
-  theme(legend.position = "none")
-
-gb <- b_df %>%
-  ggplot(aes(importance, rhat)) +
-  geom_jitter(width = 1, aes(shape=as.factor(lp))) +
-  geom_smooth(span=1) +
-  xlab("Variable importance") +
-  ylab(TeX("$\\hat{R}$")) +
-  ggtitle("B.") +
-  theme(legend.position = "none")
-```
-
-With 400 iterations
-```{r}
-fit <- sampling(model, data=list(N=N, A=A), iter=400, chains=4)
-print(fit)
-# saveRDS(fit, "../output/mvt_fit_400.rds")
-fit <- readRDS("../output/mvt_fit_400.rds")
-
-full_data <- rstan::extract(fit, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:251] <- full_data[j, i, ]
-    m_flattened[k, 252] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V252) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- a_df %>%
-  mutate(samples_400=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-```
-
-```{r}
-model <- stan_model("mvt_250_ncp.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
-print(fit)
-# saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
-fit <- readRDS("../output/mvt_ncp_fit_10000.rds")
-
-full_data <- rstan::extract(fit, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams + 1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V502) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-a_df <- a_df %>%
-  mutate(samples_10000_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-
-saveRDS(a_df, "../output/mvt_fit_all_accuracy.rds")
-
-ga <- 
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of samples", labels=c("400", "10000, cp", "10000, ncp")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8)) +
-  ggtitle("A.")
-
-pdf("../output/mvt_three.pdf", width = 12, height = 8)
-multiplot(ga, gb, gc, cols = 3)
-dev.off()
-```
-
-# Cauchy example
-
-Generate R* distribution from nominal Cauchy fit
-```{r}
-nsim <- 1000
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0)
-full_data <- rstan::extract(fit_nom, permuted=F)
-nparams <- 52
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V53) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_nom <- mAccuracy
-```
-
-Generate R* distribution from alternative Cauchy parameterisation
-```{r}
-fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0)
-full_data <- rstan::extract(fit_alt1, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V153) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_alt_1 <- mAccuracy
-```
-
-```{r}
-a_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
-               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
-
-g <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none")
-g
-
-ggsave("../output/cauchy.pdf", g, width = 8, height = 6)
-```
-
-## More iterations
-Nominal model with 10000 iterations and thinning by 10
-```{r}
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0, iter = 10000,
-                thin = 5)
-full_data <- rstan::extract(fit_nom, permuted=F)
-nparams <- 52
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V53) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-hist(colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
-mAccuracy_nom <- mAccuracy
-```
-
-Alternative model with 10000 iterations and thinning by 10
-```{r}
-fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0, iter = 10000,
-                thin = 5)
-full_data <- rstan::extract(fit_alt1, permuted=F)
-nparams <- dim(full_data)[3]
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams + 1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V153) %>% 
-  mutate(chain=as.factor(chain))
-n_distinct(m_flattened$chain)
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_alt_1 <- mAccuracy
-```
-
-Plot
-```{r}
-b_df <- tibble(nominal=colMeans(mAccuracy_nom)/(1/n_distinct(training_data$chain)),
-               alt=colMeans(mAccuracy_alt_1)/(1/n_distinct(training_data$chain)))
-
-g1 <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none") +
-  ggtitle("A.")
-g1
-
-g2 <-
-  b_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black")) +
-  ggtitle("B.")
-g2
-
-pdf("../output/cauchy.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
-```
-
-# Eight schools
-```{r}
-source("eight_schools.data.R")
-eight_schools <- list(J=J, y=y, sigma=sigma)
-model_cp <- stan_model("eight_schools_cp.stan")
-model_ncp <- stan_model("eight_schools_ncp.stan")
-```
-
-Run models
-```{r}
-fit_cp <- sampling(
-  model_cp, data = eight_schools,
-  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
-  control = list(adapt_delta = 0.95)
-)
-
-fit_ncp <- sampling(
-  model_ncp, data = eight_schools,
-  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
-  control = list(adapt_delta = 0.95)
-)
-```
-
-## Break each chain into two then make flattened data
-Centered
-```{r}
-full_data_all <- rstan::extract(fit_cp, permuted=F)
-full_data <- array(dim=c(500, 8, 11))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
-
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = 12)
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_centered <- mAccuracy
-```
-
-Noncentered
-```{r}
-full_data_all <- rstan::extract(fit_ncp, permuted=F)
-nparams <- 19
-full_data <- array(dim=c(500, 8, nparams))
-k <- 1
-for(i in 1:4){
-  first_half <- full_data_all[1:500, i, ]
-  second_half <- full_data_all[501:1000, i, ]
-  full_data[, k, ] <- first_half
-  k <- k + 1
-  full_data[, k, ] <- second_half
-  k <- k + 1
-}
-
-m_flattened <- matrix(nrow = nrow(full_data) * 8, ncol = (nparams+1))
-k <- 1
-for(i in 1:8){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V20) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_noncentered <- mAccuracy
-```
-
-Store both
-```{r}
-c_df <- tibble(centered=colMeans(mAccuracy_centered)/0.125,
-       non_centered=colMeans(mAccuracy_noncentered)/0.125)
-```
-
-## Trying with original data (i.e. 4 chains)
-```{r}
-full_data <- rstan::extract(fit_cp, permuted=F)
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 12)
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:11] <- full_data[j, i, ]
-    m_flattened[k, 12] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V12) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_centered <- mAccuracy
-
-full_data <- rstan::extract(fit_ncp, permuted=F)
-k <- 1
-m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = (nparams+1))
-k <- 1
-for(i in 1:4){
-  for(j in 1:nrow(full_data)){
-    m_flattened[k, 1:nparams] <- full_data[j, i, ]
-    m_flattened[k, (nparams+1)] <- i
-    k <- k + 1
-  }
-}
-m_flattened <- m_flattened %>% 
-  as.data.frame() %>% 
-  rename(chain=V20) %>% 
-  mutate(chain=as.factor(chain))
-
-full_data <- m_flattened
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-mAccuracy_noncentered <- mAccuracy
-```
-
-```{r}
-d_df <- tibble(centered=colMeans(mAccuracy_centered)/(1/n_distinct(training_data$chain)),
-       non_centered=colMeans(mAccuracy_noncentered)/(1/n_distinct(training_data$chain)))
-```
-
-Plot two R* distributions
-```{r}
-g1 <- 
-  c_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  xlim(0.78, 1.6) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("A.") +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none")
-g2 <- 
-  d_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlim(0.78, 1.6) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black"))
-a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
-
-pdf("../output/eight_schools.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
-```

### 2020-03-28 src/s_mvt_normal.Rmd
redraft of article

---
title: "R Notebook"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(mvtnorm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")

caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
```

Fit model in Stan with 10000 iterations
```{r}
N <- 250
# A <- rWishart(1, 250, diag(N))[,,1]
# saveRDS(A, "../output/A_matrix.rds")
A <- readRDS("../output/A_matrix.rds")
model <- stan_model("mvt_250.stan")
# fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
# saveRDS(fit, "../output/mvt_fit_10000.rds")
fit <- readRDS("../output/mvt_fit_10000.rds")
full_data <- rstan::extract(fit, permuted=F)
m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
k <- 1
for(i in 1:4){
  for(j in 1:nrow(full_data)){
    m_flattened[k, 1:251] <- full_data[j, i, ]
    m_flattened[k, 252] <- i
    k <- k + 1
  }
}
m_flattened <- m_flattened %>% 
  as.data.frame() %>% 
  rename(chain=V252) %>% 
  mutate(chain=as.factor(chain))

full_data <- m_flattened
rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
training_data <- full_data[rand_samples, ]
testing_data <- full_data[-rand_samples, ]
gbmFit1 <- train(chain ~ ., data = training_data, 
                 method = "gbm",
                 trControl = trainControl(method = 'none'), 
                    tuneGrid = caretGrid, verbose=FALSE)

plda <- predict(object=gbmFit1, newdata=testing_data)
tibble(predicted=plda, actual=testing_data$chain) %>% 
  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
  summarise(mean(correct))
varImp(gbmFit1)

plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")

nsim <- 1000
mAccuracy <- matrix(nrow = nrow(plda),
                      ncol = nsim)
  for(j in 1:nrow(plda)){
    test <- apply(rmultinom(nsim, 1, prob = plda[j, ]), 2, function(x) which(x==1))
    mAccuracy[j, ] <- if_else(test==testing_data$chain[j], 1, 0)
  }

a_df <- tibble(samples_10000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))

importance <- 100-varImp(gbmFit1)$importance$Overall
rhat <- summary(fit)$summary[, 10]
ess <- summary(fit)$summary[, 9]
plot(importance, rhat)
plot(importance, ess)
b_df <- tibble(importance, ess, rhat, lp=c(rep(0, 250), 1))
sum(rhat<1.01)
sum(ess>400)
gc <- 
  b_df %>%
  ggplot(aes(importance, ess)) +
  geom_jitter(width=1, aes(shape=as.factor(lp)), colour='grey') +
  geom_smooth(span=1, colour="black") +
  xlab("Variable importance") +
  ylab("ESS") +
  ggtitle("C.") +
  theme(legend.position = "none")

gb <- b_df %>%
  ggplot(aes(importance, rhat)) +
  geom_jitter(width = 1, aes(shape=as.factor(lp)), colour='grey') +
  geom_smooth(span=1, colour="black") +
  xlab("Variable importance") +
  ylab(TeX("$\\hat{R}$")) +
  ggtitle("B.") +
  theme(legend.position = "none")
```

With 400 iterations
```{r}
# fit <- sampling(model, data=list(N=N, A=A), iter=400, chains=4)
print(fit)
# saveRDS(fit, "../output/mvt_fit_400.rds")
fit <- readRDS("../output/mvt_fit_400.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star2 <- r_star_dist(full_data)

a_df <- a_df %>%
  mutate(samples_400=r_star2)
```

10,000 iterations
```{r}
# model <- stan_model("mvt_250_ncp.stan")
# fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
# print(fit)
# saveRDS(fit, "../output/mvt_ncp_fit_10000.rds")
fit <- readRDS("../output/mvt_ncp_fit_10000.rds")

full_data <- rstan::extract(fit, permuted=F)
r_star3 <- r_star_dist(full_data)

a_df <- a_df %>%
  mutate(samples_10000=r_star3)

a_df <- a_df %>%
  mutate(samples_10000_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))

saveRDS(a_df, "../output/mvt_fit_all_accuracy.rds")
a_df <- readRDS("../output/mvt_fit_all_accuracy.rds")

ga <- 
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
  geom_histogram(position="identity", alpha=0.8, bins=20) +
  scale_fill_grey("# of samples", labels=c("400, cp", "10000, cp", "10000, ncp")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = c(0.8, 0.8)) +
  ggtitle("A.")

pdf("../output/mvt_three.pdf", width = 12, height = 8)
multiplot(ga, gb, gc, cols = 3)
dev.off()
```

## Try measure for similar numbers of iterations to posterior
Centered
```{r}
N <- 250
A <- readRDS("../output/A_matrix.rds")
model <- stan_model("mvt_250.stan")
fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
saveRDS(fit, "../output/mvt_fit_500.rds")
fit <- readRDS("../output/mvt_fit_500.rds")

full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star <- r_star_dist(full_data)

a_df <- tibble(samples_cp=r_star)

mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)

ha <- tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat) %>% 
  melt() %>% 
  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
                            variable=="tail_ess"~"Tail ESS",
                            variable=="rhat"~"Split-Rhat")) %>% 
  ggplot(aes(x=value)) +
  geom_histogram() + 
  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
  ylab("Count") +
  theme(text=element_text(size=12, colour="black"),
        axis.text=element_text(size=12, colour="black"),
        strip.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank())
ha
aa_df <- tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat,
       source="Centered")
ggsave("../output/wide_cp_diagnostics.pdf", ha, width = 12, height = 8)
```

Non-centered
```{r}
A <- readRDS("../output/A_matrix.rds")
model <- stan_model("mvt_250_ncp.stan")
fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
saveRDS(fit, "../output/mvt_ncp_fit_500.rds")
fit <- readRDS("../output/mvt_ncp_fit_500.rds")
full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star <- r_star_dist(full_data)

a_df <- a_df %>%
  mutate(samples_ncp=r_star)
saveRDS(a_df, "../output/mvt_500_all.rds")
g <- 
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8, bins=20) +
  scale_fill_grey("Model", labels=c("centered", "noncentered")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = c(0.8, 0.8))
# ggsave("../output/wide_data.pdf", g, width = 12, height = 8)

mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)

h <- tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat) %>% 
  melt() %>% 
  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
                            variable=="tail_ess"~"Tail ESS",
                            variable=="rhat"~"Split-Rhat")) %>% 
  ggplot(aes(x=value)) +
  geom_histogram() + 
  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
  ylab("Count") +
  theme(text=element_text(size=12, colour="black"),
        axis.text=element_text(size=12, colour="black"),
        strip.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank())
ggsave("../output/wide_ncp_diagnostics.pdf", h, width = 12, height = 8)

aa_df <- aa_df %>%
  bind_rows(tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat,
       source="Non-centered"))

h1 <- aa_df %>% 
  melt(id.vars="source") %>% 
  mutate(variable=case_when(variable=="bulk_ess"~"B. Bulk ESS",
                            variable=="tail_ess"~"C. Tail ESS",
                            variable=="rhat"~"A. Split-Rhat")) %>% 
  ggplot(aes(x=value)) +
  geom_histogram() + 
  facet_grid(vars(source), vars(relevel(as.factor(variable), "A. Split-Rhat", "B. Bulk ESS", "C. Tail ESS")), scales="free") + xlab("Value") +
  ylab("Count") +
  theme(text=element_text(size=12, colour="black"),
        axis.text=element_text(size=12, colour="black"),
        strip.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank())
ggsave("../output/wide_both_diagnostics.pdf", h1, width = 12, height = 8)
```

# Wide dataset - standard normal with 10,000 dimensions
```{r}
# N <- 10000
# A <- diag(nrow=N)
# model <- stan_model("mvt_wide.stan")
# fit <- sampling(model, data=list(N=N), iter=400, chains=4)
# saveRDS(fit, "../output/mvt_wide_400.rds")
fit <- readRDS("../output/mvt_wide_400.rds")

full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star <- r_star_dist(full_data)
a_df <- tibble(samples_400=r_star)
# saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)
```

1,000 samples
```{r}
N <- 10000
A <- diag(nrow=N)
model <- stan_model("mvt_wide.stan")
fit <- sampling(model, data=list(N=N), iter=1000, chains=4)
saveRDS(fit, "../output/mvt_wide_1000.rds")

fit <- readRDS("../output/mvt_wide_1000.rds")
full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star1 <- r_star_dist(full_data)

a_df <- a_df %>%
  tibble(samples_1000=r_star1)
saveRDS(a_df, "../output/mvt_wide_summaries.rds")
mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)

g1 <- 
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8, bins=20) +
  scale_fill_grey("Model", labels=c("400", "1000")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = c(0.8, 0.8))
# ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
```

```{r}
pdf("../output/mvt_wide_both.pdf", width = 12, height = 8)
multiplot(g, g1, cols = 2)
dev.off()
```

### 2020-03-28 write_up/ms.tex
redraft of article

@@
 MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions).
-In the face of only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often practically converge to the posterior in only a (relatively small) finite number of samples.
+With only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often approximate the posterior with only a (relatively small) finite number of samples.
 
@@
 Here, we introduce $R^*$, a new convergence metric.
-This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample the chain that generated it.
+This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample's value the chain that generated it.
 Rephrased, it should not be possible to predict the chain that \textit{caused} a sample.
@@
 For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test).
+For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to calculate).
 It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
@@
 These two cases, whilst simple, illustrate the basis of our approach.
-To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs.
+To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs.
 By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
@@
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
-Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.
+Algorithm \ref{alg:R_star} gives a recipe for calculating $R^*$.
 
@@
 
-The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \citep{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
+The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
 Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
@@
 We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
-The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for models with 1000 samples), and its predictive performance was insensitive to its hyperparameters.
+Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters.
 In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).
@@
 \section{Results}\label{sec:results}
-To illustrate the versatility of $R^*$, we use examples from \citep{vehtari2019rank} and two additional examples based on a multivariate normal model.
+To illustrate the versatility of $R^*$, we use a range of examples that demonstrate how this statistic fares across a range of scenarios.
 
@@
 To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e.
-with one chain with a lower variance).
-We then fit a gradient-boosted model to a labelled training set of 2800 randomly chosen samples.
+where one chain has a lower variance).
+We then fit a GBM to a labelled training set of 2800 randomly chosen samples.
 The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them.
@@
 In Fig.
-\ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates indicating that the chains have not converged.
+\ref{fig:ar1}B, we show that $R^*>1$ for all replicates indicating that the chains have not converged.
 
@@
 Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done.
-We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data'); and another, where all chains sample from the same distribution (we call this the ``converged'' data).
+We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data); and another, where all chains sample from the same distribution (we call this the ``converged'' data).
 In Fig.
@@
 
-First, we use the code provided in \citep{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
+First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
 In all cases, the various quantities were calculated based on chains split into halves.
@@
 
-Next, we estimate the $R^*$ distribution which is shown in Fig.
+Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty}, which is shown in Fig.
 \ref{fig:bivariate}.
-The mean of this distribution is 1.18, and all the samples are above 1, indicating that the sampling distribution has not converged.
+The mean of this distribution is 1.18, and all $R^*$ samples are above 1, indicating that the sampling distribution has not converged.
 By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals.
-As such, we see a role for $R^*$ as an additional convergence metric.
 
@@
 %
-\begin{equation}
+\begin{equation}\label{eq:mvt_normal_250}
 \boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
 
@@
 
-In GBMs, it is possible to calculate variable importance (see, for example, \citep{friedman2001greedy} and \citep{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
+In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
 We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan.
@@
 \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
-Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.
+Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\hat{R}$ and ESS.
 
@@
 	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
-In plots B and C, horizontal jitter was added to the points.
+In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 
+\subsection{Wide datasets: multivariate normal}\label{sec:wide}
+As the number of parameter dimensions increases, it could be hypthesised that ML algorithms will overfit the data, and, hence, testing set classification would be poor; in being so, this could lead us to falsely conclude that convergence had occurred.
+To test this hypothesis, we investigated two scenarios based around the multivariate normal.
+In the first of these, we used the 250-dimensional multivariate normal of eq.
+\ref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
+Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
+\ref{fig:mvt_wide_both}A.
+In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
+The same conclusion is reached if split-$\hat{R}$ is used instead (Fig.
+\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\hat{R}>1.01$.
+Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
+\ref{fig:wide_both_diagnostics}B\&C).
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_wide_both.pdf}}
+	\caption{\textbf{Wide data examples: $R^*$ distributions.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations from Stan's NUTS algorithm across both model parameterisations.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:mvt_wide_both}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
+	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
+Column A shows split-$\hat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:wide_both_diagnostics}
+\end{figure}
+
+We next consider a more challenging example -- a target distribution with 10,000 dimensions.
+In this case, we assume independent standard normals for each dimension.
+In Fig.
+\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distributions: one with 400 iterations, the other with 1000.
+In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
+These results were also echoed by split-$\hat{R}$, with 28\% of dimensions having $\hat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
+
+Overall, the examples in this section suggest that $R^*$ is robust to wide datasets; we also note that the statistic took comparable time to calculate relative to existing convergence diagnostics on a desktop computer.
+
 \subsection{Infinite variance: Cauchy example}
 We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
-Like \citep{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
+Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-dimensional vector $x$,
 %
 \begin{equation}
-x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
+x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50.
 \end{equation}
@@
 \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations.
-As shown in \citep{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
 The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$.
-Interesting, however, in contrast to the measures presented in \citep{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig.
+Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig.
 \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5).
@@
 To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target.
-One measure of distributional "closeness" is the KL-divergence, which, in this case, could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated.
+One measure of distributional ``closeness'' is the KL-divergence, which, in this case, could be used to measure the divergence from target to sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated for each dimension.
 The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails.
@@
 Comparing our ``omniscient'' measure of convergence shown in Fig.
-\ref{fig:cauchy_convergence}A, with the various derived measures, it is clear that all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
+\ref{fig:cauchy_convergence}A, with the various derived measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
 The rate at which these converge differs though, and $R^*$ (Fig.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
-	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split $\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
+	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 \ref{fig:trends_all_dim}, we show these results.
-In column A, we show the results for $R^*$ calculated on the four chains that were run; column B shows the same calculation but after the chains are split into two equal halves.
-The different rows show the different sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
+In column A, we show the results for $R^*$ calculated on the four chains that ran; column B shows the same calculation but after the chains are split into two equal halves.
+The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
 This plot mirrors Fig.
@@
 
-These results are what would intuitively be expected: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation.
+These results make intuitive sense: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation.
 In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals.
-By splitting chains into two -- the first half being the early phase and, the second half being the later phase with higher mean -- this forces differences in the marginals.
+By splitting chains into two -- the first half being the early phase, and the second half being the later phase with higher mean -- this forces differences in the marginals.
 This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one.
-As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \cite{carpenter2017stan,vehtari2019rank}.
+As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \citep{carpenter2017stan,vehtari2019rank}.
 
@@
 In Fig.
-\ref{fig:trends_one_dim}, we show the effect on $R^*$ of trends of different magnitudes across a range of target dimensions.
-In the simulations here, all dimensions bar one are stationary; the remaining dimension has a trend added to it.
+\ref{fig:trends_one_dim}, we show how $R^*$ performs across a range of target dimensions.
+In the simulations here, all dimensions bar one are stationary; the remaining dimension has a linear trend added to it.
 In all cases, split-$R^*$ increased with trend.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_one_dim.pdf}}
-	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns represent different dimensionalities of the target distribution; the rows present different sample sizes.
+	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
 The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
@@
 
-Means that trends over time is one form of non-stationarity; another is a time-varying covariance.
+Means that trend over time is one form of non-stationarity; another is a time-varying covariance.
 Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time.
-Specifically, we allow the correlation to increase from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
+Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations, and use i.i.d.
 samples from the process across 4 ``chains''.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_joint_distribution.pdf}}
-	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\hat{R}$; column B for $R^*$.
+	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\hat{R}$; column B for $R^*$; the rows present the differing sample sizes.
 The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
@@
 
-The model can be parameterised two ways, as described in \citep{vehtari2019rank}.
+The model can be parameterised two ways, as described in \cite{vehtari2019rank}.
 The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
@@
 We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \citep{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
+Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
 We also used Stan with the same algorithm settings to sample from the non-centered model.
 
-To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \citep{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Next, we trained the GBM on a training set comprising 2800 samples, and used it to classify the causative chain for 1200 independent test samples.
+To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
 Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
 The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
-In this plot, it is clear that whereas the centered model shows signs of convergence, the non-centered model does not: intuitively, the chains have not sufficiently mixed with themselves nor the others meaning that the causative chain can be predicted with reasonable accuracy from the sample location.
+In this plot, the centered model shows signs of convergence, whereas the non-centered model does not.
 
@@
 
-To understand the predictive power of the ML better, we investigate how predictive accuracy varies across parameter space.
+To understand the predictive power of the ML better, we investigated how predictive accuracy varies across parameter space.
 After fitting the GBM model, we group MCMC samples into deciles and sample from the $R^*$ distribution for each decile.
@@
 	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
-In each case, we show the path of the four individual chains above and 40 $R^*$ samples by parameter quantile below.
+In each column, we show the path of the four individual chains above and 40 $R^*$ samples for each parameter quantile below.
 Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
@@
 By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
-Across a range of previously published examples, $R^*$ was shown to be predictive of whether or not chains had converged.
+Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
 
@@
 Many implementations of $\hat{R}$ suggest splitting chains in two before calculating it.
-In the Eight Schools example in \S\ref{sec:eight_shools}, we trial this before calculating $R^*$, and find that this approach leads to more accurate chain prediction.
+In a number of examples, we trial this before calculating $R^*$, and find that this approach leads to more accurate chain prediction.
 As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.

### 2020-04-10 write_up/ms.tex
edited abstract and sections 1-2

@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage[final]{nips_2018}
+\usepackage[final]{neurips_2018}
 
@@
 \usepackage{hyperref}       % hyperlinks
+\hypersetup{unicode=true,
+            pdfborder={0 0 0},
+            breaklinks=true,
+            colorlinks=true,
+            linkcolor=blue,
+            citecolor=blue,
+            filecolor=blue,
+            urlcolor=blue}
 \usepackage{url}            % simple URL typesetting
@@
 \usepackage{bbm}
+
+\newcommand{\aki}[1]{\textcolor{red}{[Aki: #1]}}
 
@@
   Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
-Despite its importance, MCMC is a notoriously subtle beast.
+% Despite its importance, MCMC is a notoriously subtle beast.
 Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models.
@@
   Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
-The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a stationary distribution.
-Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
-Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call this convergence measure $R^*$.
-In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely overlooked by existing approaches.
-Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
+The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
+%  Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
+Here, we introduce a new method for diagnosing convergence based whether machine learning classifier model can successfully discriminate the individual chains.
+We call this convergence measure $R^*$.
+In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined.
+Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
+Since the used machine learning classifier, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
 The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
@@
 Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} and the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement.
 MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
-Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
+% Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
 We repeat here a quote from Chris Holmes to this effect \citep{holmes},
+% \begin{quote}
+% 	MCMC is most definitely not a panacea and should be used with caution, almost as a last resort. It is just that often we are at the last resort for interesting models.
+% \end{quote}
 
-\begin{quote}
-	MCMC is most definitely not a panacea and should be used with caution, almost as a last resort. It is just that often we are at the last resort for interesting models.
-\end{quote}
-
-The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}.
-MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions).
-With only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often approximate the posterior with only a (relatively small) finite number of samples.
+The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}.\aki{Maybe some other reference, too?} MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions).
+Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small finite number of samples.
 
@@
 This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}.
-The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman2013bayesian}.
-If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed.
-Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \citep{vehtari2019rank}.
+The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\widehat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman1992inference}.
+If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed.
+Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}.
 Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
@@
 Rephrased, it should not be possible to predict the chain that \textit{caused} a sample.
-In this vein, we use a supervised machine learning (ML) approach to measure convergence.
-Specifically, we train a ML model to classify the chain that generated each observation.
+In this vein, we use a machine learning (ML) classifier to measure convergence.
+Specifically, we train a ML model to predict the chain that generated each observation.
 By evaluating the performance of the model on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches.
-This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter.
-However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
-For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to calculate).
+To maximise predictive accuracy, the used ML classifier naturally exploit differences between the full joint distributions of each chain, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
+This statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
+However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
+For our machine learning classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
+For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically less than 1 second to calculate).
 It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
@@
 By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \citep{gelman2013bayesian}.
+By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{gelman2013bayesian}.
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
@@
 The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
-We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
+We split each chain's samples into randomly divided training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
 Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters.
@@
 
-A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$.
+We can also get class probabilities, which we leverage to produce an uncertainty distribution for $R^*$.
 Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words.
@@
 
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
+First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
 In all cases, the various quantities were calculated based on chains split into halves.
-For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.
+For both dimensions, the two ESS measures were above 3700 and $\widehat{R}<1.001$, indicating no issues with convergence.
 
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\widehat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
 
@@
 \ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is centred on $R^*=1$ (mean $R^*=1.00$) indicating convergence.
-This was also reiterated by the fact that $\hat{R}<1.01$ and $\text{ESS}>400$ for all parameters.
+This was also reiterated by the fact that $\widehat{R}<1.01$ and $\text{ESS}>400$ for all parameters.
 
 In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
-We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan.
+We now compare these with the more established metrics $\widehat{R}$ and ESS as calculated in Stan.
 For a GBM fitted to the centered model, we plot in Fig.
-\ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$).
+\ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
+From this plot, there appears a weak positive association between GBM's variable importance and $\widehat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$).
 In Fig.
 \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
-Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\hat{R}$ and ESS.
+Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
 
@@
 	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
+	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\widehat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
 In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.
@@
 In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
-The same conclusion is reached if split-$\hat{R}$ is used instead (Fig.
-\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\hat{R}>1.01$.
+The same conclusion is reached if split-$\widehat{R}$ is used instead (Fig.
+\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\widehat{R}>1.01$.
 Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
@@
 	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
-Column A shows split-$\hat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm.
+Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by split-$\hat{R}$, with 28\% of dimensions having $\hat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
+These results were also echoed by split-$\widehat{R}$, with 28\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
 
@@
 \ref{fig:cauchy_convergence}B and Fig.
-\ref{fig:cauchy_convergence}C, we plot two measures of $\hat{R}$, each calculated from splitting the 4 original chains into two equal halves.
-The first of these measures is the rank-normalised $\hat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig.
+\ref{fig:cauchy_convergence}C, we plot two measures of $\widehat{R}$, each calculated from splitting the 4 original chains into two equal halves.
+The first of these measures is the rank-normalised $\widehat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig.
 \ref{fig:cauchy_convergence}B, we show how the maximum of this measurement across all 50 dimensions changes with sample size.
-After c.550 iterations, the alternative parameterisation achieves $\hat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum $\hat{R}$ value: in both cases, these suggest convergence.
-The second measure is multivariate $\hat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig.
+After c.550 iterations, the alternative parameterisation achieves $\widehat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum $\widehat{R}$ value: in both cases, these suggest convergence.
+The second measure is multivariate $\widehat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig.
 \ref{fig:cauchy_convergence}C shows how this metric changes with sample size for both Cauchy model parameterisations.
 After c.
-1800 iterations, multivariate $\hat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\hat{R}>1.07$ indicating more samples are needed.
+1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more samples are needed.
 In Fig.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
-	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
+	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}.
-This example showed that split-$\hat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
+This example showed that split-$\widehat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
 Specifically, they first generated four chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling.
-We first repeat this analysis but using $R^*$ rather than $\hat{R}$: in Fig.
+We first repeat this analysis but using $R^*$ rather than $\widehat{R}$: in Fig.
 \ref{fig:trends_all_dim}, we show these results.
@@
 This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one.
-As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \citep{carpenter2017stan,vehtari2019rank}.
+As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\widehat{R}$ \citep{carpenter2017stan,vehtari2019rank}.
 
@@
 Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution.
-Similarly, split-$\hat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig.
+Similarly, split-$\widehat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig.
 \ref{fig:trends_joint_distribution}A).
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_joint_distribution.pdf}}
-	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\hat{R}$; column B for $R^*$; the rows present the differing sample sizes.
+	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
 The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
@@
 For the GBM fitted to the centered model, the most important variable was $\tau$, followed by Stan's $lp$ variable (we also include this in our training and testing data), followed by $\mu$ and $\theta_2$.
-Interestingly, these were ordered differently according to Stan's $\hat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).
+Interestingly, these were ordered differently according to Stan's $\widehat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).
 
@@
 
-Many implementations of $\hat{R}$ suggest splitting chains in two before calculating it.
+Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
 In a number of examples, we trial this before calculating $R^*$, and find that this approach leads to more accurate chain prediction.

### 2020-04-11 write_up/ms.tex
added mvt_wide_both and updated draw/sample ambiguity in text

@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage[final]{neurips_2018}
+\usepackage{neurips_2019}
 
@@
   Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
-% Despite its importance, MCMC is a notoriously subtle beast.
-Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models.
-Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution.
 Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
 The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
-%  Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
-Here, we introduce a new method for diagnosing convergence based whether machine learning classifier model can successfully discriminate the individual chains.
+Here, we introduce a new method for diagnosing convergence based on whether a machine learning classifier model can successfully discriminate the individual chains.
 We call this convergence measure $R^*$.
@@
 Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
-% Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use.
-We repeat here a quote from Chris Holmes to this effect \citep{holmes},
-% \begin{quote}
-% 	MCMC is most definitely not a panacea and should be used with caution, almost as a last resort. It is just that often we are at the last resort for interesting models.
-% \end{quote}
 
-The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}.\aki{Maybe some other reference, too?} MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions).
-Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small finite number of samples.
+The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
+MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions).
+Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small finite number of draws.
 
@@
 If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed.
-Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}.
+Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter draws rather than the raw values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}.
 Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
@@
 Here, we introduce $R^*$, a new convergence metric.
-This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample's value the chain that generated it.
-Rephrased, it should not be possible to predict the chain that \textit{caused} a sample.
+This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
+Rephrased, it should not be possible to predict the chain that \textit{caused} a draw.
 In this vein, we use a machine learning (ML) classifier to measure convergence.
-Specifically, we train a ML model to predict the chain that generated each observation.
-By evaluating the performance of the model on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, the used ML classifier naturally exploit differences between the full joint distributions of each chain, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
-This statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
+Specifically, we train a ML classifier to predict the chain that generated each observation.
+By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric.
+To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between the chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
+Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
 However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
-For our machine learning classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
+For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
 For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically less than 1 second to calculate).
@@
 
-The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \citep{vehtari2019rank}.
+The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \citep{vehtari2019rank} and elsewhere.
 Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
 
-%$R^*$ calculation is enabled in Pints inference software \citep{Clerx2019Pints}.
-
 \section{Method}\label{sec:method}
-If Markov chains have not mixed, it is possible to determine to which chain a sample belongs from the sample's value.
+If Markov chains have not mixed, it is possible to determine to which chain a draw belongs from its value.
 This is possible if there are differences in the sampling distribution for any dimension, $\theta$, in the target distribution (Fig.
-\ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a sample belongs to.
-It  is also possible to predict the chain that generated a given sample if there are differences in the joint distribution of two (or more) dimensions of the target, even if the marginal distributions are the same (Fig.
+\ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a draw belongs to.
+It  is also possible to predict the chain that generated a given draw if there are differences in the joint distribution of two (or more) dimensions of the target, even if the marginal distributions are the same (Fig.
 \ref{fig:joint}).
@@
 These two cases, whilst simple, illustrate the basis of our approach.
-To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs.
+To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each draw belongs.
 By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
@@
 	\begin{algorithmic}
-		\STATE Given chain-wise samples from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
+		\STATE Given chain-wise draws from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
 		\FOR{$m=1$ to $N$}
@@
 		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
-		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
+		\STATE Train ML model to classify chain id from any draw, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
 		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
+		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
 		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
@@
 
-The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
+The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
 Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
-The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
-We split each chain's samples into randomly divided training and testing tests: here, we use 70\% of samples for training and 30\% for testing.
+The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
+We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing.
 Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters.
@@
 We can also get class probabilities, which we leverage to produce an uncertainty distribution for $R^*$.
-Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words.
-For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
+Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
+For each draw, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
+We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
 We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
@@
 \begin{algorithm}[tb]
-	\caption{Procedure to generate $I$ samples of $R^*$}
+	\caption{Procedure to generate $I$ draws of $R^*$}
 	\label{alg:R_star_uncertainty}
@@
 		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Obtain test sample, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
+		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
 		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
@@
 In Fig.
-\ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a sample's value.
+\ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a draw's value.
 Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
@@
 where one chain has a lower variance).
-We then fit a GBM to a labelled training set of 2800 randomly chosen samples.
-The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them.
+We then fit a GBM to a labelled training set.
+The fitted model is then used to classify draws in an independent test set according to the chain which generated them.
 For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the sample's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text.
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 
-Since GBMs return a probability simplex for each sample, which, for each chain, indicates the probability that a sample was generated by it, we can also generate a measure of uncertainty in $R^*$.
+Since GBMs return a probability simplex for each draw, which, for each chain, indicates the probability that the draw was generated by it, we can also generate a measure of uncertainty in $R^*$.
 Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done.
@@
 In all cases, the various quantities were calculated based on chains split into halves.
-For both dimensions, the two ESS measures were above 3700 and $\widehat{R}<1.001$, indicating no issues with convergence.
+For both dimensions, the two ESS measures were above 3700, and $\widehat{R}<1.001$, indicating no issues with convergence.
 
@@
 \ref{fig:bivariate}.
-The mean of this distribution is 1.18, and all $R^*$ samples are above 1, indicating that the sampling distribution has not converged.
+The mean of this distribution is 1.18, and all $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
 By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals.
@@
 	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
-	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 samples. See the R markdown file on the Github repository for code to reproduce this example.}
+	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws. See the R markdown file on the Github repository for code to reproduce this example.}
 	\label{fig:bivariate}
@@
 where $\boldsymbol{x}\in\mathbb{R}^{250}$.
-For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig.
+For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig.
 \ref{fig:mvt}A.
@@
 	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
-Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm.
+Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
 The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} A shows 1000 MCMC samples from the $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC samples used (which are then thinned by a factor of 5).
+	\caption{\textbf{Cauchy example.} A shows 1000 MCMC draws and associated $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC draws used (which are then thinned by a factor of 5).
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 After c.
-1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more samples are needed.
+1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more draws are needed.
 In Fig.
@@
 Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations, and use i.i.d.
-samples from the process across 4 ``chains''.
+draws from the process across 4 ``chains''.
 Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution.
@@
 The results are shown in Fig.\ref{fig:eight_schools}B.
-In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each sample, meaning that the centered model $R^*$ values are shifted leftwards.
+In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each draw, meaning that the centered model $R^*$ values are shifted leftwards.
 Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears converged.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows samples from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
-In all cases, the plots show 1000 $R^*$ samples.
+	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
+In all cases, the plots show 1000 $R^*$ draws.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 To understand the predictive power of the ML better, we investigated how predictive accuracy varies across parameter space.
-After fitting the GBM model, we group MCMC samples into deciles and sample from the $R^*$ distribution for each decile.
+After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
 In Fig.
@@
 Above the top panel, we show the marginal distributions for each chain.
-In the bottom row, we show 40 $R^*$ samples for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all the samples.
+In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all the draws.
 In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
@@
 	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
-In each column, we show the path of the four individual chains above and 40 $R^*$ samples for each parameter quantile below.
+In each column, we show the path of the four individual chains above and 40 $R^*$ draws for each parameter quantile below.
 Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
@@
 \section{Discussion}
-If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a sample it should be impossible to discern which chain generated it.
-Based on this observation, we used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in samples.
+If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a draw, it should be impossible to discern which chain generated it.
+Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
 By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
@@
 The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target.
-$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each sample.
+$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each draw.
 It also naturally includes information about the joint distribution across all dimensions of the target.
@@
 Its ease of implementation, however, masks important underlying fragilities in the method.
-Namely, that unless the chains have converged to a truly stationary distribution, the samples generated are not faithful depictions of the posterior.
+Namely, that unless the chains have converged to a truly stationary distribution, the draws generated are not faithful depictions of the posterior.
 In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far.

### 2020-04-14 src/s_ovarian_prostate.Rmd
added ovarian and prostate data and model

---
title: "Ovarian and prostate models"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(mvtnorm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

# Ovarian data
```{r}
ovarian <- read.csv("../data/ovarian.csv", header=FALSE)

y<-ovarian[,1537]
x<-ovarian[,1:1536]

n = nrow(x);
p = ncol(x);

## regularized horseshoe prior
p0 <- 5
sigma <- 1
tau0 <- p0/(p-p0)*sigma/sqrt(n)
## data
data <- list(n = n, d = ncol(x), x = x, y = y,
             nu_local = 1, nu_global = 1, scale_global = tau0,
             scale_icept=5, slab_scale=2, slab_df=100)
```

### 2020-04-14 src/s_ar1.Rmd
started to add ovarian plots

@@
 ---
-title: "R Notebook"
+title: "Autoregressive example"
 output: html_notebook
@@
 
-# AR1 example
 Functions to generate simulated data
@@
 
+# Generates three chains with same var; one with a different var
 f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
@@
 
-Perform 1000 replicates
+Calculate R* and Rhat across 1000 replicates
 ```{r}
 f_replicate <- function(){
-  full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-  rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-  training_data <- full_data[rand_samples, ]
-  testing_data <- full_data[-rand_samples, ]
-
-  # add in extra column of nonsense values since the method needs at least 2 vars 
-  full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-  training_data <- full_data[rand_samples, ]
-  testing_data <- full_data[-rand_samples, ]
-
-  gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-  plda <- predict(object=gbmFit1, newdata=testing_data)
-  a_accuracy <- 
-    tibble(predicted=plda, actual=testing_data$chain) %>%
-    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-    summarise(mean(correct)) %>% 
-    pull()
-  return(a_accuracy)
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  a_accuracy <- r_star(a_array)
+  mon <- monitor(temp)
+  return(list(r_star=a_accuracy, r_hat=mon$Rhat))
 }
@@
 nreplicates <- 1000
-accuracies <- vector(length = nreplicates)
+r_star_vals <- vector(length = nreplicates)
+r_hat_vals <- vector(length = nreplicates)
 for(i in 1:nreplicates){
@@
     print(i)
-  accuracies[i] <- f_replicate()
+
+  temp <- f_replicate()
+  r_star_vals[i] <- temp$r_star
+  r_hat_vals[i] <- temp$r_hat
 }
@@
 ```{r}
-
-g2 <- tibble(accuracy=accuracies/0.25, iteration=seq(1, nreplicates, 1)) %>% 
+g2 <- tibble(accuracy=r_star_vals, iteration=seq(1, nreplicates, 1)) %>% 
   ggplot(aes(x=accuracy)) +
@@
   xlim(1, NA)
+```
 
+Look at predictions of model
+```{r}
 g1 <- testing_data %>% 
@@
 
-Generate R* distribution based on 1000 iterations
+Generate R* distribution based on 1000 iterations for run where chain has lower variance and another where all four chains have same variance
 ```{r}
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
+full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) 
+a_array <- array(dim=c(1000, 4, 1))
+a_array[,,1] <- temp
+unmixed <- r_star_dist(a_array)
 
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
+full_data <- f_generate_lower_var_four(1, 0.3, 1, 1000)
+a_array <- array(dim=c(1000, 4, 1))
+a_array[,,1] <- temp
+mixed <- r_star_dist(a_array)
 
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-f_generate_all_same_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:4)
-    x[, i] <- f_ar1(rho, sigma, L)
-  return(x)
-}
-
-full_data <- f_generate_all_same_var_four(1/3, 0.3, 1, 1000) %>% 
-  melt() %>% 
-  rename(iter=Var1,
-         chain=Var2) %>% 
-  dplyr::select(-iter)
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-full_data <- full_data %>% 
-    mutate(chain=as.factor(chain)) %>% 
-    mutate(ones=rnorm(nrow(full_data)))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy1 <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy1[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-mAR1 <- tibble(unconverged=colMeans(mAccuracy)/0.25,
-               converged=colMeans(mAccuracy1)/0.25)
+mAR1 <- tibble(unmixed, mixed)
 g3 <- 

### 2020-04-14 src/s_bivariate_normal.Rmd
started to add ovarian plots

@@
 source("r_star_monitor.R")
-
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
 ```
 
+Generate data
 ```{r}
@@
          iter=seq_along(chain))
+```
 
+Determine R* distribution
+```{r}
+full_samples <- array(dim = c(n, 4, 2))
+full_samples[, 1, ] <- as.matrix(chain1[, 1:2])
+full_samples[, 2, ] <- as.matrix(chain2[, 1:2])
+full_samples[, 3, ] <- as.matrix(chain3[, 1:2])
+full_samples[, 4, ] <- as.matrix(chain4[, 1:2])
+
+# Calculate R* dist
+r_star_vals <- r_star_dist(full_samples)
+
+g <- 
+  qplot(r_star_vals) +
+  xlab(TeX("$R*$")) +
+  xlim(0.9, NA) +
+  geom_vline(xintercept = 1, linetype=2) +
+  ylab("Count") +
+  theme(text=element_text(size=14, colour="black"))
+# ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
+```
+
+Determine Rhat
+```{r}
 chains_stacked <- rbind.data.frame(chain1, chain2, chain3, chain4) %>% 
   mutate(chain=as.factor(chain))
-chains_stacked1 <- chains_stacked %>% 
-  select(-iter)
 
-full_data <- chains_stacked1
-rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
-training_data <- full_data[rand_samples, ]
-testing_data <- full_data[-rand_samples, ]
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
-                 trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                    ncol = nsim)
-for(i in 1:nsim){
-  for(j in 1:nrow(plda)){
-    mAccuracy[j, i] <- if_else(which(rmultinom(1, 1, prob = plda[j, ])==1)==testing_data$chain[j], 1, 0)
-  }
-}
-
-# Try monitor functionality -- split chains first
+# Use monitor package functionality -- split chains first
 half_iter <- max(chains_stacked$iter) / 2
@@
 
-g <- 
-  qplot(4*colMeans(mAccuracy)) +
-  xlab(TeX("$R*$")) +
-  xlim(0.9, NA) +
-  geom_vline(xintercept = 1, linetype=2) +
-  ylab("Count") +
-  theme(text=element_text(size=14, colour="black"))
-# ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
 mon <- monitor(m_samples)
-
 ```

### 2020-04-14 src/s_ovarian_prostate.Rmd
started to add ovarian plots

@@
              scale_icept=5, slab_scale=2, slab_df=100)
+
+model <- stan_model("bernoulli_logit_glm_rhs.stan")
 ```
+
+With 4 long chains
+```{r}
+fit <- sampling(model, data=data, chains=4, iter=9000, warmup=1000,
+                control=list(adapt_delta = 0.8, max_treedepth=10), cores=4, thin=10)
+saveRDS(fit, "../data/ovarian_long.rds")
+fit <- readRDS("../data/ovarian_long.rds")
+
+mon <- monitor(fit)
+saveRDS(mon, "../data/ovarian_long_mon.rds")
+full_data <- rstan::extract(fit, permuted=F)
+r_star <- r_star_dist(full_data)
+saveRDS(r_star, "../data/ovarian_long_r_star.rds")
+```
+
+With 16, shorter, chains
+```{r}
+fit <- sampling(model, data=data, chains=16, iter=1000, warmup=500,
+                control=list(adapt_delta = 0.8, max_treedepth=10), cores=4)
+saveRDS(fit, "../data/ovarian_short.rds")
+fit <- readRDS("../data/ovarian_short.rds")
+
+mon <- monitor(fit)
+saveRDS(mon, "../data/ovarian_short_mon.rds")
+full_data <- rstan::extract(fit, permuted=F)
+r_star <- r_star_dist(full_data)
+saveRDS(r_star, "../data/ovarian_short_r_star.rds")
+```
+
+Plot R* and Rhat
+```{r}
+r_star_longer <- readRDS("../data/ovarian_long_r_star.rds")
+r_star_shorter <- readRDS("../data/ovarian_short_r_star.rds")
+
+mon_longer <- readRDS("../data/ovarian_long_mon.rds")
+mon_shorter <- readRDS("../data/ovarian_short_mon.rds")
+
+g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
+  bind_rows(tibble(long=mon_longer$Rhat, short=mon_shorter$Rhat, metric="B. Rhat")) %>% 
+  bind_rows(tibble(long=mon_longer$Bulk_ESS, short=mon_shorter$Bulk_ESS, metric="C. Bulk ESS")) %>% 
+  bind_rows(tibble(long=mon_longer$Tail_ESS, short=mon_shorter$Tail_ESS, metric="D. Tail ESS")) %>% 
+  melt(id.vars="metric") %>% 
+  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.6) +
+  facet_wrap(~metric, scales="free") +
+  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
+  theme(strip.text.x = element_text(size = 10, colour="black"),
+        text = element_text(size=12, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.text = element_text(size=12, colour="black"),
+        axis.title.x = element_blank()) +
+  ylab("Count")
+ggsave("../output/ovarian.pdf", g, width = 12, height = 8)
+```
+
+# Prostate
+```{r}
+prostate <- read.csv("../data/prostate.csv", header=FALSE)
+
+y<-prostate[,5967]
+x<-prostate[,1:5966]
+
+n = nrow(x);
+p = ncol(x);
+
+## regularized horseshoe prior
+p0 <- 5
+sigma <- 1
+tau0 <- p0/(p-p0)*sigma/sqrt(n)
+## data
+data <- list(n = n, d = ncol(x), x = x, y = y,
+             nu_local = 1, nu_global = 1, scale_global = tau0,
+             scale_icept=5, slab_scale=2, slab_df=100)
+
+model <- stan_model("bernoulli_logit_glm_rhs.stan")
+```
+
+With 4 long chains
+```{r}
+fit <- sampling(model, data=data, chains=4, iter=9000, warmup=1000,
+                control=list(adapt_delta = 0.8, max_treedepth=10), cores=4, thin=10)
+saveRDS(fit, "../data/prostate_long.rds")
+fit <- readRDS("../data/prostate_long.rds")
+
+mon <- monitor(fit)
+saveRDS(mon, "../data/prostate_long_mon.rds")
+full_data <- rstan::extract(fit, permuted=F)
+r_star <- r_star_dist(full_data)
+saveRDS(r_star, "../data/prostate_long_r_star.rds")
+```
+
+With 16, shorter, chains
+```{r}
+fit <- sampling(model, data=data, chains=16, iter=1000, warmup=500,
+                control=list(adapt_delta = 0.8, max_treedepth=10), cores=4)
+saveRDS(fit, "../data/prostate_short.rds")
+fit <- readRDS("../data/prostate_short.rds")
+
+mon <- monitor(fit)
+saveRDS(mon, "../data/prostate_short_mon.rds")
+full_data <- rstan::extract(fit, permuted=F)
+r_star <- r_star_dist(full_data)
+saveRDS(r_star, "../data/prostate_short_r_star.rds")
+```
+
+Plot R* and Rhat
+```{r}
+r_star_longer <- readRDS("../data/prostate_long_r_star.rds")
+r_star_shorter <- readRDS("../data/prostate_short_r_star.rds")
+
+mon_longer <- readRDS("../data/prostate_long_mon.rds")
+mon_shorter <- readRDS("../data/prostate_short_mon.rds")
+
+g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
+  bind_rows(tibble(long=mon_longer$Rhat, short=mon_shorter$Rhat, metric="B. Rhat")) %>% 
+  bind_rows(tibble(long=mon_longer$Bulk_ESS, short=mon_shorter$Bulk_ESS, metric="C. Bulk ESS")) %>% 
+  bind_rows(tibble(long=mon_longer$Tail_ESS, short=mon_shorter$Tail_ESS, metric="D. Tail ESS")) %>% 
+  melt(id.vars="metric") %>% 
+  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.6) +
+  facet_wrap(~metric, scales="free") +
+  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
+  theme(strip.text.x = element_text(size = 10, colour="black"),
+        text = element_text(size=12, colour="black"),
+        axis.text = element_text(size=10, colour="black"),
+        legend.text = element_text(size=12, colour="black"),
+        axis.title.x = element_blank()) +
+  ylab("Count")
+ggsave("../output/prostate.pdf", g, width = 12, height = 8)
+```

### 2020-04-14 src/r_star_monitor.R

@@ -45,6 +45,10 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
   colnames(m_flattened)[nparams + 1] <- "chain"
   r <- m_flattened %>% 
     mutate(chain=as.factor(chain))
+  # if only 1 param, add in a column of random noise since gbm requires >1 dims
+  if(nparams==1)
+    r <- r %>% 
+      mutate(V_new=rnorm(nrow(r)))
   
   rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
   training_data <- r[rand_samples, ]

### 2020-04-15 src/s_ovarian_prostate.Rmd
added prostate and ovary plots

@@
 # Prostate
+
 ```{r}
@@
 full_data <- rstan::extract(fit, permuted=F)
-r_star <- r_star_dist(full_data)
+
+# Thin data again
+r_star <- r_star_dist(full_data[, , seq(1, 18106, 5)])
 saveRDS(r_star, "../data/prostate_long_r_star.rds")
@@
 full_data <- rstan::extract(fit, permuted=F)
-r_star <- r_star_dist(full_data)
+r_star <- r_star_dist(full_data[, , seq(1, 18106, 5)])
 saveRDS(r_star, "../data/prostate_short_r_star.rds")
@@
 mon_shorter <- readRDS("../data/prostate_short_mon.rds")
+
+my_breaks <- function(x) { if (max(x) < 10) c(0.95, max(x)) else c(0, max(x)) }
 
@@
         axis.title.x = element_blank()) +
-  ylab("Count")
+  ylab("Count") +
+  scale_x_continuous(limits=my_breaks)
 ggsave("../output/prostate.pdf", g, width = 12, height = 8)

### 2020-05-03 src/s_ar1.Rmd
Up to end of fig. 8

@@
 source("r_star_monitor.R")
-
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
 ```
@@
 }
+
+saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/ar1_r_star_hat.rds")
 ```
@@
   theme(text = element_text(size=14, colour="black")) +
+  xlim(0.9, NA)
+
+g2a <- tibble(accuracy=r_hat_vals, iteration=seq(1, nreplicates, 1)) %>% 
+  ggplot(aes(x=accuracy)) +
+  geom_histogram() +
+  xlab(TeX("split-$\\hat{R}$")) +
+  ylab("Count") +
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black")) +
   xlim(1, NA)
@@
 ```{r}
+# add in a column of noise since gbm requires >1 variables
+full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
+  melt() %>% 
+  select(-Var1) %>% 
+  rename(chain=Var2) %>% 
+  mutate(ran_val=rnorm(4000)) %>% 
+  mutate(chain=as.factor(chain))
+
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+
 g1 <- testing_data %>% 
@@
 ```{r}
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) 
-a_array <- array(dim=c(1000, 4, 1))
+temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) 
+temp1 <- matrix(rnorm(4000), nrow = 1000)
+a_array <- array(dim=c(1000, 4, 2))
 a_array[,,1] <- temp
+a_array[,,2] <- temp1
 unmixed <- r_star_dist(a_array)
 
-full_data <- f_generate_lower_var_four(1, 0.3, 1, 1000)
-a_array <- array(dim=c(1000, 4, 1))
+temp <- f_generate_lower_var_four(1, 0.3, 1, 1000)
+temp1 <- matrix(rnorm(4000), nrow = 1000)
+a_array <- array(dim=c(1000, 4, 2))
 a_array[,,1] <- temp
+a_array[,,2] <- temp1
 mixed <- r_star_dist(a_array)
@@
   geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series") +
+  scale_fill_grey("Series", labels=c("un-mixed", "mixed")) +
   xlab(TeX("$R*$")) +
@@
   geom_vline(xintercept = 1, linetype=2) +
-  ggtitle("C.") +
+  ggtitle("D.") +
   theme(text = element_text(size=14, colour="black"),
@@
 
-# pdf("../output/ar1.pdf", width = 12, height = 6)
-multiplot(g1, g2, g3, cols = 3)
-# dev.off()
+pdf("../output/ar1.pdf", width = 12, height = 6)
+multiplot(g1, g2a, g2, g3, cols = 2)
+dev.off()
 ```

### 2020-05-03 src/s_cauchy.Rmd
Up to end of fig. 8

@@
 r_star_nom <- r_star_dist(x)
+
+mon <- monitor(fit_nom)
+a_rhat_df <- tibble(nominal=mon$Rhat[1:50])
 ```
@@
 r_star_alt <- r_star_dist(full_data)
+mon <- monitor(fit_alt1)
+a_rhat_df <-a_rhat_df %>% 
+  mutate(alt=mon$Rhat[101:150],
+         source="r_hat")
 ```
@@
 a_df <- tibble(nominal=r_star_nom,
-               alt=r_star_alt)
+               alt=r_star_alt,
+               source="r_star") %>% 
+  bind_rows(a_rhat_df)
 g <-
@@
   theme(text = element_text(size=14, colour="black"),
-        legend.position = "none")
+        legend.position = "none") +
+  facet_wrap(~source, scales = "free")
 g
@@
 
-## Calculate KL divergence from true density by fitting kernel density estimator to points then integrating
-nominal parameterisation
-```{r}
-x <- rstan::extract(fit_nom, permuted=F)
-
-sample <- as.vector(x[, , i])
-dens <- density(sample)
-f_approx <- approxfun(dens$x, dens$y, rule=2)
-integrate(
-    function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -20, 20, subdivisions = 1000)
-
-t <- seq(-20, 20, 0.01)
-tibble(t,
-  true=map_dbl(t, ~dcauchy(.)),
-  approx=map_dbl(t, ~f_approx(.))) %>%
-  melt(id.vars="t") %>%
-  ggplot(aes(x=t, y=value, colour=as.factor(variable))) +
-  geom_line()
-
-f_kl_approx <- function(sample, limit=20){
-  dens <- density(sample)
-  f_approx <- approxfun(dens$x, dens$y, rule=2)
-  kl <- integrate(function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -limit, limit, subdivisions=1000)$value
-  if(kl < 0)
-    return(0)
-  else
-    return(kl)
-}
-
-kl <- vector(length = 50)
-for(i in seq_along(kl))
-  kl[i] <- f_kl_approx(x[, , i], limit=20)
-```
-
-Density estimation fails for alternative parameterisation because of fat tails
-```{r}
-names(extract(fit_alt1))
-x <- rstan::extract(fit_alt1, permuted=F, pars="x")
-sample <- as.vector(x[, , 3])
-dens <- density(sample, 0.001)
-f_approx <- approxfun(dens$x, dens$y, rule=2)
-integrate(function(x) f_approx(x), -10, 10)
-integrate(
-    function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -100, 100, subdivisions = 1000)
-
-t <- seq(-20, 20, 0.01)
-tibble(t,
-  true=map_dbl(t, ~dcauchy(.)),
-  approx=map_dbl(t, ~f_approx(.))) %>%
-  melt(id.vars="t") %>%
-  ggplot(aes(x=t, y=value, colour=as.factor(variable))) +
-  geom_line()
-
-f_kl_approx <- function(sample, limit=20){
-  dens <- density(sample)
-  f_approx <- approxfun(dens$x, dens$y, rule=2)
-  kl <- integrate(function(x) f_approx(x) * log(f_approx(x) / dcauchy(x)),      -limit, limit, subdivisions=1000)$value
-  if(kl < 0)
-    return(0)
-  else
-    return(kl)
-}
-
-kl <- vector(length = 50)
-for(i in seq_along(kl))
-  kl[i] <- f_kl_approx(x[, , i], limit=20)
-```
-
-Look at quantiles instead using KS test statistic: problem with KS test statistic is that it is most sensitive to middle values: https://asaip.psu.edu/articles/beware-the-kolmogorov-smirnov-test/
-```{r}
-x <- rstan::extract(fit_alt1, permuted=F, pars="x")
-sample <- as.vector(x[, , 1])
-ks.test(sample, pcauchy)
-ks <- vector(length = 50)
-for(i in 1:50)
-  ks[i] <- ks.test(x[, , i], pcauchy)[1]$statistic
-```
-
-Anderson-darling test -- by comparing with an independent sampler we get a gauge of the quality of MCMC
-```{r}
-x <- rstan::extract(fit, permuted=F, pars="x")
-
-f_calculate_ad <- function(x){
-  dims <- dim(x)
-  nparams <- dims[3]
-  ad <- vector(length = nparams)
-  for(i in 1:nparams)
-    ad[i] <- ad.test(x[, , i], pcauchy)[2]$p.value
-  return(ad)
-}
-
-# keep sample size fixed by thinning
-nmax <- 5000
-iter <- seq(500, nmax, 50)
-ad <- vector(length = length(iter))
-for(i in seq_along(iter)){
-  a_iter <- iter[i]
-  ele <- round(seq(1, a_iter, length.out = 500))
-  ad[i] <- mean(f_calculate_ad(x[ele, ,]))
-}
-
-e_df <- tibble(iter, ad, source="MCMC")
-
-# compare with independent samples
-# ps <- map_dbl(seq(1, 1000, 1), function(i) ad.test(rcauchy(1000), pcauchy)[2]$p.value)
-# hist(ps)
-
-x <- array(rcauchy(2 * nmax * 50), c(nmax, 4, 50))
-
-iter <- seq(500, nmax, 50)
-ad <- vector(length = length(iter))
-for(i in seq_along(iter)){
-  a_iter <- iter[i]
-  ele <- round(seq(1, a_iter, length.out = 500))
-  ad[i] <- mean(f_calculate_ad(x[ele, ,]))
-}
-f_df <- e_df %>% 
-  bind_rows(tibble(iter, ad, source="independent"))
-f_df %>% 
-  ggplot(aes(x=iter, y=ad, colour=as.factor(source))) +
-  geom_line()
-```
-
-Try R* with independent sampler
-```{r}
-nmax <- 100
-x <- array(rcauchy(2 * nmax * 50), c(nmax, 4, 50))
-r_star_inde <- r_star_dist(x)
-hist(r_star_inde)
-```
-
-Possible to tell whether sample came from MCMC or independent?
-```{r}
-x_mcmc <- rstan::extract(fit, permuted=F, pars="x")
-x_mcmc <- array(rcauchy(2000 * 50), c(2000, 50)) %>% 
-  as.data.frame() %>% 
-  mutate(source="mcmc")
-
-f_distribution <- function(x_mcmc){
-  x_inde <- array(rcauchy(2000 * 50), c(2000, 50))
-  x_inde <- x_inde %>% 
-  as.data.frame() %>% 
-  mutate(source="independent")
-  r <- bind_rows(x_mcmc,
-                 x_inde) %>% 
-    mutate(source=as.factor(source))
-  rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
-  training_data <- r[rand_samples, ]
-  testing_data <- r[-rand_samples, ]
-
-  gbmFit1 <- train(source ~ ., data = training_data, 
-                     method = "gbm",
-                     trControl = trainControl(method = 'none'), 
-                     tuneGrid = caretGrid, verbose=FALSE)
-
-  plda <- predict(object=gbmFit1, newdata=testing_data)
-  a_accuracy <- 
-    tibble(predicted=plda, actual=testing_data$source) %>%
-    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-    summarise(mean(correct)) %>% 
-    pull()
-  return(a_accuracy)
-}
-library(furrr)
-future::plan(multiprocess)
-laccuracy <- future_map_dbl(seq(1, 1000, 1), function(i) f_distribution(x_mcmc))
-hist(laccuracy, 50)
-```
-
-Could do same test with simply empirical distributions, also, by bootstrapping
-
 ## 10,000 iterations
-Nominal model with 10000 iterations and thinning by 10
+Nominal model with 50000 iterations and thinning by 5
 ```{r}
 fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
-            refresh = 0, iter = 50000)
+            refresh = 0, iter = 50000, thin=5)
 saveRDS(fit, "../output/fit_nominal_50000.rds")
+# fit <- readRDS("../output/fit_nominal_50000.rds")
 full_data <- rstan::extract(fit, permuted=F)
@@
 r_star_nom <- r_star_dist(full_data)
+mon_nom <- monitor(fit)
 ```
 
-Alternative model with 10000 iterations and thinning by 10
+Alternative model with 50000 iterations and thinning by 5
 ```{r}
 fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
-            refresh = 0, iter = 50000)
+            refresh = 0, iter = 60000, thin=3)
 saveRDS(fit, "../output/fit_alt_50000.rds")
@@
 r_star_alt <- r_star_dist(full_data)
+mon_alt <- monitor(fit)
 ```
@@
 b_df <- tibble(nominal=r_star_nom,
-               alt=r_star_alt)
+               alt=r_star_alt,
+               source="r_star") %>% 
+  bind_rows(tibble(nominal=mon_nom$Rhat[1:50],
+                alt=mon_alt$Rhat[101:150],
+                source="r_hat")) %>% 
+  mutate(iter=10000)
 
-g1 <-
-  a_df %>% 
-  melt() %>% 
+c_df <- a_df %>% mutate(iter=1000) %>% 
+  bind_rows(b_df) %>% 
+  mutate(source=if_else(source=="r_hat", "split-Rhat", source)) %>% 
+  mutate(source=if_else(source=="r_star", "R*", source))
+
+my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.03) else c(0.9, 3)}
+
+g <- c_df %>% 
+  melt(id.vars=c("source", "iter")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
   ylab("Count") +
   geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none") +
-  ggtitle("A.")
-g1
+  facet_wrap(~source+iter, scales="free") +
+  scale_x_continuous(limits=my_lims) +
+  theme(strip.text = element_text(size=14, colour="black"),
+        axis.text = element_text(size=12, colour="black"),
+        axis.title = element_text(size=14, colour="black"),
+        legend.text = element_text(size=14, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  xlab("Value")
 
-g2 <-
-  b_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black")) +
-  ggtitle("B.")
-g2
-
-pdf("../output/cauchy.pdf", width = 12, height = 6)
-multiplot(g1, g2, cols = 2)
-dev.off()
+ggsave("../output/cauchy.pdf", g, width = 12, height = 8)
 ```

### 2020-05-03 src/s_mvt_normal.Rmd
Up to end of fig. 8

@@
 ---
-title: "R Notebook"
+title: "Multivariate normal example"
 output: html_notebook
@@
 
-Fit model in Stan with 10000 iterations
+# Centered and non-centered parameterisations
+Fit centered model in Stan with 10000 iterations and look at variable importance
 ```{r}
@@
 fit <- readRDS("../output/mvt_fit_10000.rds")
+
+# Collect Rhat
+a_sum <- monitor(fit)
+rhat <- a_sum$Rhat
+
+# Reshape data into form for R* estimation
 full_data <- rstan::extract(fit, permuted=F)
@@
 testing_data <- full_data[-rand_samples, ]
+
+# Fit gradient-boosed model
 gbmFit1 <- train(chain ~ ., data = training_data, 
@@
 
-plda <- predict(object=gbmFit1, newdata=testing_data)
-tibble(predicted=plda, actual=testing_data$chain) %>% 
-  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-  summarise(mean(correct))
-varImp(gbmFit1)
-
-plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
-
-nsim <- 1000
-mAccuracy <- matrix(nrow = nrow(plda),
-                      ncol = nsim)
-  for(j in 1:nrow(plda)){
-    test <- apply(rmultinom(nsim, 1, prob = plda[j, ]), 2, function(x) which(x==1))
-    mAccuracy[j, ] <- if_else(test==testing_data$chain[j], 1, 0)
-  }
-
-a_df <- tibble(samples_10000=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+# Get R* distribution
+full_data <- rstan::extract(fit, permuted=F)
+a_df <- tibble(samples_10000_cp=r_star_dist(full_data))
 
 importance <- 100-varImp(gbmFit1)$importance$Overall
-rhat <- summary(fit)$summary[, 10]
-ess <- summary(fit)$summary[, 9]
-plot(importance, rhat)
-plot(importance, ess)
-b_df <- tibble(importance, ess, rhat, lp=c(rep(0, 250), 1))
+ess_b <- a_sum$Bulk_ESS
+ess_t <- a_sum$Tail_ESS
+b_df <- tibble(importance, ess_b, ess_t, rhat, lp=c(rep(0, 250), 1))
 sum(rhat<1.01)
-sum(ess>400)
-gc <- 
-  b_df %>%
-  ggplot(aes(importance, ess)) +
-  geom_jitter(width=1, aes(shape=as.factor(lp)), colour='grey') +
-  geom_smooth(span=1, colour="black") +
+mean(ess_b>400)
+mean(ess_t>400)
+cor.test(b_df$importance, b_df$rhat, method="spearman")
+cor.test(b_df$importance, b_df$ess_b, method="spearman")
+cor.test(b_df$importance, b_df$ess_t, method="spearman")
+gc <- b_df %>%
+  select(importance, ess_b, ess_t, lp) %>% 
+  melt(id.vars=c("importance", "lp")) %>% 
+  ggplot(aes(importance, value)) +
+  geom_jitter(width=1, aes(shape=as.factor(lp), colour=as.factor(variable))) +
+  geom_smooth(span=1, aes(colour=as.factor(variable))) +
   xlab("Variable importance") +
   ylab("ESS") +
-  ggtitle("C.") +
-  theme(legend.position = "none")
+  ggtitle("D.") +
+  scale_color_grey("ESS type", labels=c("Bulk", "Tail")) +
+  scale_shape("LP param", labels=c("False", "True")) +
+  theme(legend.position = c(0.85, 0.7))
 
@@
   xlab("Variable importance") +
-  ylab(TeX("$\\hat{R}$")) +
-  ggtitle("B.") +
+  ylab(TeX("split-$\\hat{R}$")) +
+  ggtitle("C.") +
   theme(legend.position = "none")
+b_df <- tibble(samples_10000_cp=a_sum$Rhat[1:250])
 ```
@@
 fit <- readRDS("../output/mvt_fit_400.rds")
+
 full_data <- rstan::extract(fit, permuted=F)
@@
   mutate(samples_400=r_star2)
+
+a_sum <- monitor(fit)
+b_df <- b_df %>%
+  mutate(samples_400=a_sum$Rhat[1:250])
 ```
 
-10,000 iterations
+10,000 iterations for non-centered parameterisation
 ```{r}
@@
 fit <- readRDS("../output/mvt_ncp_fit_10000.rds")
+a_sum <- monitor(fit)
+sum(a_sum$Bulk_ESS>400)
+sum(a_sum$Tail_ESS>400)
+b_df <- b_df %>%
+  mutate(samples_10000_ncp=a_sum$Rhat[251:500])
+saveRDS(b_df, "../output/mvt_fit_all_rhat.rds")
+b_df <- readRDS("../output/mvt_fit_all_rhat.rds")
 
@@
 a_df <- a_df %>%
-  mutate(samples_10000=r_star3)
-
-a_df <- a_df %>%
-  mutate(samples_10000_ncp=colMeans(mAccuracy)/(1/n_distinct(training_data$chain)))
+  mutate(samples_10000_ncp=r_star3)
 
@@
   geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of samples", labels=c("400, cp", "10000, cp", "10000, ncp")) +
+  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
   xlab(TeX("$R*$")) +
@@
 
-pdf("../output/mvt_three.pdf", width = 12, height = 8)
-multiplot(ga, gb, gc, cols = 3)
-dev.off()
-```
-
-## Try measure for similar numbers of iterations to posterior
-Centered
-```{r}
-N <- 250
-A <- readRDS("../output/A_matrix.rds")
-model <- stan_model("mvt_250.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
-saveRDS(fit, "../output/mvt_fit_500.rds")
-fit <- readRDS("../output/mvt_fit_500.rds")
-
-full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star <- r_star_dist(full_data)
-
-a_df <- tibble(samples_cp=r_star)
-
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
-
-ha <- tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat) %>% 
+ga2 <- 
+  b_df %>% 
   melt() %>% 
-  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
-                            variable=="tail_ess"~"Tail ESS",
-                            variable=="rhat"~"Split-Rhat")) %>% 
-  ggplot(aes(x=value)) +
-  geom_histogram() + 
-  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
-  ylab("Count") +
-  theme(text=element_text(size=12, colour="black"),
-        axis.text=element_text(size=12, colour="black"),
-        strip.text = element_text(size=12, colour="black"),
-        axis.title.x = element_blank())
-ha
-aa_df <- tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat,
-       source="Centered")
-ggsave("../output/wide_cp_diagnostics.pdf", ha, width = 12, height = 8)
-```
-
-Non-centered
-```{r}
-A <- readRDS("../output/A_matrix.rds")
-model <- stan_model("mvt_250_ncp.stan")
-fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
-saveRDS(fit, "../output/mvt_ncp_fit_500.rds")
-fit <- readRDS("../output/mvt_ncp_fit_500.rds")
-full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star <- r_star_dist(full_data)
-
-a_df <- a_df %>%
-  mutate(samples_ncp=r_star)
-saveRDS(a_df, "../output/mvt_500_all.rds")
-g <- 
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
   geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("Model", labels=c("centered", "noncentered")) +
-  xlab(TeX("$R*$")) +
+  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
+  xlab(TeX("split-$\\hat{R}$")) +
   ylab("Count") +
@@
   theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
-# ggsave("../output/wide_data.pdf", g, width = 12, height = 8)
+        legend.position = c(0.8, 0.8)) +
+  ggtitle("B.")
 
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
-
-h <- tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat) %>% 
-  melt() %>% 
-  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
-                            variable=="tail_ess"~"Tail ESS",
-                            variable=="rhat"~"Split-Rhat")) %>% 
-  ggplot(aes(x=value)) +
-  geom_histogram() + 
-  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
-  ylab("Count") +
-  theme(text=element_text(size=12, colour="black"),
-        axis.text=element_text(size=12, colour="black"),
-        strip.text = element_text(size=12, colour="black"),
-        axis.title.x = element_blank())
-ggsave("../output/wide_ncp_diagnostics.pdf", h, width = 12, height = 8)
-
-aa_df <- aa_df %>%
-  bind_rows(tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat,
-       source="Non-centered"))
-
-h1 <- aa_df %>% 
-  melt(id.vars="source") %>% 
-  mutate(variable=case_when(variable=="bulk_ess"~"B. Bulk ESS",
-                            variable=="tail_ess"~"C. Tail ESS",
-                            variable=="rhat"~"A. Split-Rhat")) %>% 
-  ggplot(aes(x=value)) +
-  geom_histogram() + 
-  facet_grid(vars(source), vars(relevel(as.factor(variable), "A. Split-Rhat", "B. Bulk ESS", "C. Tail ESS")), scales="free") + xlab("Value") +
-  ylab("Count") +
-  theme(text=element_text(size=12, colour="black"),
-        axis.text=element_text(size=12, colour="black"),
-        strip.text = element_text(size=12, colour="black"),
-        axis.title.x = element_blank())
-ggsave("../output/wide_both_diagnostics.pdf", h1, width = 12, height = 8)
-```
-
-# Wide dataset - standard normal with 10,000 dimensions
-```{r}
-# N <- 10000
-# A <- diag(nrow=N)
-# model <- stan_model("mvt_wide.stan")
-# fit <- sampling(model, data=list(N=N), iter=400, chains=4)
-# saveRDS(fit, "../output/mvt_wide_400.rds")
-fit <- readRDS("../output/mvt_wide_400.rds")
-
-full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star <- r_star_dist(full_data)
-a_df <- tibble(samples_400=r_star)
-# saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
-```
-
-1,000 samples
-```{r}
-N <- 10000
-A <- diag(nrow=N)
-model <- stan_model("mvt_wide.stan")
-fit <- sampling(model, data=list(N=N), iter=1000, chains=4)
-saveRDS(fit, "../output/mvt_wide_1000.rds")
-
-fit <- readRDS("../output/mvt_wide_1000.rds")
-full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star1 <- r_star_dist(full_data)
-
-a_df <- a_df %>%
-  tibble(samples_1000=r_star1)
-saveRDS(a_df, "../output/mvt_wide_summaries.rds")
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
-
-g1 <- 
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("Model", labels=c("400", "1000")) +
-  xlab(TeX("$R*$")) +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8))
-# ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
-```
-
-```{r}
-pdf("../output/mvt_wide_both.pdf", width = 12, height = 8)
-multiplot(g, g1, cols = 2)
+pdf("../output/mvt_three.pdf", width = 12, height = 8)
+multiplot(ga, gb, ga2, gc, cols = 2)
 dev.off()

### 2020-05-03 src/s_wide_distributions.Rmd
Up to end of fig. 8

---
title: "MVT Normal: wide"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(mvtnorm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

## Similar numbers of iterations to target dims
Centered
```{r}
# N <- 250
# A <- readRDS("../output/A_matrix.rds")
# model <- stan_model("mvt_250.stan")
# fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
# saveRDS(fit, "../output/mvt_fit_500.rds")
fit <- readRDS("../output/mvt_fit_500.rds")

full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star <- r_star_dist(full_data)

a_df <- tibble(samples_cp=r_star)

mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)

b_df <- tibble(samples_cp=mon$Rhat[1:250])

ha <- tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat) %>% 
  melt() %>% 
  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
                            variable=="tail_ess"~"Tail ESS",
                            variable=="rhat"~"Split-Rhat")) %>% 
  ggplot(aes(x=value)) +
  geom_histogram() + 
  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
  ylab("Count") +
  theme(text=element_text(size=12, colour="black"),
        axis.text=element_text(size=12, colour="black"),
        strip.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank())
ha
aa_df <- tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat,
       source="Centered")
# ggsave("../output/wide_cp_diagnostics.pdf", ha, width = 12, height = 8)
```

Non-centered
```{r}
# A <- readRDS("../output/A_matrix.rds")
# model <- stan_model("mvt_250_ncp.stan")
# fit <- sampling(model, data=list(N=N, A=A), iter=500, chains=4, thin=1)
# saveRDS(fit, "../output/mvt_ncp_fit_500.rds")
fit <- readRDS("../output/mvt_ncp_fit_500.rds")
full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star <- r_star_dist(full_data)

a_df <- a_df %>%
  mutate(samples_ncp=r_star)
saveRDS(a_df, "../output/mvt_500_all.rds")
a_df <- readRDS("../output/mvt_500_all.rds")
g <- 
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8, bins=20) +
  scale_fill_grey("Model", labels=c("centered", "non-centered")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = c(0.8, 0.8)) +
  ggtitle("A.")
# ggsave("../output/wide_data.pdf", g, width = 12, height = 8)

mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)

h <- tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat) %>% 
  melt() %>% 
  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
                            variable=="tail_ess"~"Tail ESS",
                            variable=="rhat"~"Split-Rhat")) %>% 
  ggplot(aes(x=value)) +
  geom_histogram() + 
  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
  ylab("Count") +
  theme(text=element_text(size=12, colour="black"),
        axis.text=element_text(size=12, colour="black"),
        strip.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank())
ggsave("../output/wide_ncp_diagnostics.pdf", h, width = 12, height = 8)

aa_df <- aa_df %>%
  bind_rows(tibble(bulk_ess=mon$Bulk_ESS,
       tail_ess=mon$Tail_ESS,
       rhat=mon$Rhat,
       source="Non-centered"))

h1 <- aa_df %>% 
  melt(id.vars="source") %>% 
  mutate(variable=case_when(variable=="bulk_ess"~"B. Bulk ESS",
                            variable=="tail_ess"~"C. Tail ESS",
                            variable=="rhat"~"A. Split-Rhat")) %>% 
  ggplot(aes(x=value)) +
  geom_histogram() + 
  facet_grid(vars(source), vars(relevel(as.factor(variable), "A. Split-Rhat", "B. Bulk ESS", "C. Tail ESS")), scales="free") + xlab("Value") +
  ylab("Count") +
  theme(text=element_text(size=12, colour="black"),
        axis.text=element_text(size=12, colour="black"),
        strip.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank())
ggsave("../output/wide_both_diagnostics.pdf", h1, width = 12, height = 8)
```

# Wide dataset - standard normal with 10,000 dimensions
```{r}
# N <- 10000
# A <- diag(nrow=N)
# model <- stan_model("mvt_wide.stan")
# fit <- sampling(model, data=list(N=N), iter=400, chains=4)
# saveRDS(fit, "../output/mvt_wide_400.rds")
fit <- readRDS("../output/mvt_wide_400.rds")

full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star <- r_star_dist(full_data)
a_df <- tibble(samples_400=r_star)
# saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)
```

1,000 samples
```{r}
N <- 10000
A <- diag(nrow=N)
model <- stan_model("mvt_wide.stan")
fit <- sampling(model, data=list(N=N), iter=1000, chains=4)
saveRDS(fit, "../output/mvt_wide_1000.rds")

fit <- readRDS("../output/mvt_wide_1000.rds")
full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
r_star1 <- r_star_dist(full_data)

a_df <- a_df %>%
  tibble(samples_1000=r_star1)
saveRDS(a_df, "../output/mvt_wide_summaries.rds")
mon <- monitor(fit)
mean(mon$Rhat>1.01)
mean(mon$Bulk_ESS<400)
mean(mon$Tail_ESS<400)
a_df <- readRDS("../output/mvt_wide_summaries.rds")
a_df <- tibble(samples_400=a_df$.$samples_400, samples_1000=a_df$samples_1000)
g1 <- 
  a_df %>% 
  melt() %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.8, bins=20) +
  scale_fill_grey("# of draws", labels=c("400", "1000")) +
  xlab(TeX("$R*$")) +
  ylab("Count") +
  geom_vline(xintercept = 1, linetype=2) +
  theme(text = element_text(size=14, colour="black"),
        legend.position = c(0.8, 0.8)) +
  ggtitle("B.")
# ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
```

```{r}
pdf("../output/mvt_wide_both.pdf", width = 12, height = 6)
multiplot(g, g1, cols = 2)
dev.off()
```

### 2020-05-03 write_up/ms.tex
Up to end of fig. 8

@@
 % [preprint] option:
-%     \usepackage[preprint]{neurips_2018}
+\usepackage[preprint]{neurips_2019}
 
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
+%\usepackage{neurips_2019}
 
@@
 In Fig.
-\ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a draw's value.
+\ref{fig:ar1}A, we show how a GBM fitted to these data classifies observations according to a draw's value.
 Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
@@
 In Fig.
-\ref{fig:ar1}B, we show that $R^*>1$ for all replicates indicating that the chains have not converged.
+\ref{fig:ar1}B, we show that $R^*>1$ for 997/1000 replicates, indicating that the chains have not converged in a vast majority of cases.
+In Fig.
+\ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate; as for $R^*$, this metric indicates the chains have not converged in all replicates.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows the calculated $R^*$ values across 1000 replicates; C shows the $R^*$ distributions for the two series described in the text.
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows 1000 $R^*$ draws as generated by Algorithm \ref{alg:R_star_uncertainty} across 1000 replicates; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows the $R^*$ distributions for the two series described in the text.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 In Fig.
-\ref{fig:ar1}C, we show the $R^*$ distributions in each case.
+\ref{fig:ar1}D, we show the $R^*$ distributions in each case.
 For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
@@
 The mean of this distribution is 1.18, and all $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
-By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals.
+By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
 
@@
 We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}.
-We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions.
+We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions.
 We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
-First, we used the ``centered'' parameterisation of this model, which is of the form,
+First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
 %
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\widehat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.03): in this case, $\widehat{R}<1.01$ for all parameters (Fig.
+\ref{fig:mvt}B), although 60\% of parameters had tail $\text{ESS}<400$ and 39\% had bulk $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
 
@@
 Fig.
-\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is centred on $R^*=1$ (mean $R^*=1.00$) indicating convergence.
-This was also reiterated by the fact that $\widehat{R}<1.01$ and $\text{ESS}>400$ for all parameters.
+\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is closer to $R^*=1$ (mean $R^*=1.01$).
+Fig.
+\ref{fig:mvt}B shows the $\widehat{R}$ values for each $x$ parameter in this model, and, echoing the result for $R^*$, $\widehat{R}<1.01$ in all cases; further, bulk- and tail-$\text{ESS}>400$ for all parameters.
 
@@
 For a GBM fitted to the centered model, we plot in Fig.
-\ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-From this plot, there appears a weak positive association between GBM's variable importance and $\widehat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$).
+\ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
+From this plot, there appears a weak positive association between GBM's variable importance and $\widehat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.10, S=2368138, p=0.11$).
 In Fig.
-\ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$).
-Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
+\ref{fig:mvt}D, we plot variable importance versus two measures: bulk ESS and tail ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk ESS: $\rho=-0.54, S=4146491, p<0.01$; tail ESS: $\rho=-0.55, S=4095040, p<0.01$).
+Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
 
@@
 	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\widehat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan.
+	\caption{\textbf{Multivariate normal example with 250 dimensions.} A shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 1000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 1000 draws); B shows the rank-normalised split-$\widehat{R}$ values for all parameters from the same MCMC runs as in A; C shows variable importance versus $\widehat{R}$ for each parameter; and D shows variable importance versus bulk- and tail-ESS as calculated by \cite{vehtari2019rank}.
+In A, 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} are shown for each MCMC run.
 In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.
@@
 \subsection{Wide datasets: multivariate normal}\label{sec:wide}
-As the number of parameter dimensions increases, it could be hypthesised that ML algorithms will overfit the data, and, hence, testing set classification would be poor; in being so, this could lead us to falsely conclude that convergence had occurred.
-To test this hypothesis, we investigated two scenarios based around the multivariate normal.
+As the number of parameter dimensions increases, it might be thought that ML algorithms will overfit the data, and, hence, testing set classification would be poor; leading to unreliable determinations of convergence.
+To test this hypothesis, we investigated two scenarios using a multivariate normal target.
 In the first of these, we used the 250-dimensional multivariate normal of eq.
-\ref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
+\eqref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
 Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
@@
 In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
-The same conclusion is reached if split-$\widehat{R}$ is used instead (Fig.
+The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
 \ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\widehat{R}>1.01$.
@@
 	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_wide_both.pdf}}
-	\caption{\textbf{Wide data examples: $R^*$ distributions.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations from Stan's NUTS algorithm across both model parameterisations.
+	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
+In all cases, 1000 draws of $R^*$ are plotted as generated by Alg.
+\ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
-	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
+	\caption{\textbf{Wide data 250-dimensional example: established diagnostics.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
 Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
+In both cases, the results correspond to a single MCMC run composed of 4 chains.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 In Fig.
-\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distributions: one with 400 iterations, the other with 1000.
+\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
 In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by split-$\widehat{R}$, with 28\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
+These results were also echoed by rank-normalised split-$\widehat{R}$, with 28\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
 
@@
 
-In Fig.
-\ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
-The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$.
-Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig.
-\ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5).
-In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
+In the top-left panel of Fig.
+\ref{fig:cauchy}, we show the $R^*$ distribution under both parameterisations.
+As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of four chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$, yet, still the mean is above this value.
+In the bottom-left panel of Fig.
+\ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs.
+The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}>1.01$ for all parameters.
+
+Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of four chains.
+In the right-hand column of Fig.
+\ref{fig:cauchy}, we show the results for these longer runs.
+In these, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
 Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
+The $\widehat{R}$ values are all below 1.01, indicating convergence in both cases.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} A shows 1000 MCMC draws and associated $R^*$ distribution generated under two model parameterisations; B shows the same but with 10,000 MCMC draws used (which are then thinned by a factor of 5).
+	\caption{\textbf{Cauchy example.} Columns show MCMC runs with 1000 (left) and 10,000 (right; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of four chains.
+Rows show the $R^*$ distributions (top) and rank-normalised split-$\widehat{R}$ values across all parameters (bottom).
+The different shadings indicate different model paramerisations as indicated in legend.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 
+\subsection{Many parameter models: ovarian and prostate analysis}
+In this section, we...
+
 \section{Discussion}

### 2020-05-03 src/r_star_monitor.R

@@ -60,7 +60,7 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
                              shrinkage=c(0.1),
                              n.minobsinnode=10)
   else
-    caretGrid <- expand.grid(caret_grid_default)
+    caretGrid <- expand.grid(caret_default)
   
   gbmFit <- train(chain ~ ., data = training_data, 
                    method = "gbm",
@@ -130,7 +130,7 @@ r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_default=N
                              shrinkage=c(0.1),
                              n.minobsinnode=10)
   else
-    caretGrid <- expand.grid(caret_grid_default)
+    caretGrid <- expand.grid(caret_default)
   
   gbmFit1 <- train(chain ~ ., data = training_data, 
                    method = "gbm",

### 2020-05-04 src/s_cauchy.Rmd
added new rhat plot for eight schools

@@
 ## 10,000 iterations
-Nominal model with 50000 iterations and thinning by 5
+Nominal model with 50000 iterations
 ```{r}
 fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
-            refresh = 0, iter = 50000, thin=5)
+            refresh = 0, iter = 50000)
 saveRDS(fit, "../output/fit_nominal_50000.rds")
-# fit <- readRDS("../output/fit_nominal_50000.rds")
+fit <- readRDS("../output/fit_nominal_50000.rds")
 full_data <- rstan::extract(fit, permuted=F)
@@
 
-Alternative model with 50000 iterations and thinning by 5
+Alternative model with 50000 iterations
 ```{r}
 fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
-            refresh = 0, iter = 60000, thin=3)
+            refresh = 0, iter = 50000)
 saveRDS(fit, "../output/fit_alt_50000.rds")
+fit <- readRDS("../output/fit_alt_50000.rds")
 full_data <- rstan::extract(fit, permuted=F)
@@
 
+## Look at how various measures indicate convergence with iteration
 Compare quantiles
@@
 fit <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
 r_squared <- function(x, lower=0.001){
@@
 
-log_last <- log10(25000)
+log_last <- log10(10000)
 log_first <- log10(100)
@@
 fit <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
 r_vals <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
@@
 g1
+```
+
+Cramer Von Mises -- seems very noisy
+```{r}
+library(goftest)
+f_cvm <- function(x){
+  temp <- cvm.test(x, null = "pcauchy")
+  return(temp$p.value)
+}
+
+f_cvm_all_dims <- function(x, max_iter){
+  vals <- map_dbl(seq(1, 50, 1), ~f_cvm(x[1:max_iter, , .]))
+  return(vals)
+}
+
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
+cvm_min <- map_dbl(iter, ~mean(f_cvm_all_dims(x, .)))
+
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
+cvm_min <- map_dbl(iter, ~mean(f_cvm_all_dims(x, .)))
+```
+
+Anderson-Darling test -- again, noisy
+```{r}
+library(goftest)
+f_ad <- function(x){
+  temp <- ad.test(x, null = "pcauchy")
+  return(temp$p.value)
+}
+
+f_ad_all_dims <- function(x, max_iter){
+  vals <- map_dbl(seq(1, 50, 1), ~f_ad(x[1:max_iter, , .]))
+  return(vals)
+}
+
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
+cvm_min <- map_dbl(iter, ~mean(f_ad_all_dims(x, .)<0.05))
+
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
+cvm_min <- map_dbl(iter, ~mean(f_ad_all_dims(x, .)<0.05))
 ```
@@
 g2
+```
+
+Progression of R* with ranked samples instead
+```{r}
+
 ```
@@
 ```
-
-Look at quantile accuracy for models converged by R* and Rhat
-```{r}
-fit <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x")
-quantile_errors <- function(x, lower=0.001){
-  print(dim(x))
-  qs <- seq(lower, 1-lower, lower)
-  q_actual <- map_dbl(qs, ~qcauchy(.))
-  errors <- matrix(nrow = length(qs),
-                   ncol=dim(x)[3])
-  for(i in 1:50){
-    q_approx <- quantile(x[, , i], qs)
-    errors[, i] <- abs(q_actual - q_approx)
-  }
-  errors <- errors %>% 
-    as.data.frame() %>% 
-    mutate(q=0.5-abs(0.5-qs)) %>% 
-    melt(id.vars="q")
-  return(errors)
-}
-
-errors_df <- quantile_errors(x, lower=0.001)
-errors_df %>% 
-  ggplot(aes(x=q, y=value)) +
-  geom_jitter(width = 0.05) +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_smooth(method="lm") +
-  xlab("Distance from centre") +
-  ylab("Error")
-
-aLM <- lm(data=errors_df, formula="log(value)~log(q)")
-summary(aLM)
-
-errors_df <- quantile_errors(x[1:iter[8], , ], lower=0.001)
-errors_df %>% 
-  ggplot(aes(x=q, y=value)) +
-  geom_jitter(width = 0.05) +
-  scale_x_log10() +
-  scale_y_log10() +
-  geom_smooth(method="lm") +
-  xlab("Distance from centre") +
-  ylab("Error")
-
-bLM <- lm(data=errors_df, formula="log(value)~log(q)")
-summary(bLM)
-```

### 2020-05-04 src/s_eight_schools.Rmd
added new rhat plot for eight schools

@@
 ## Break each chain into two then make flattened data
-Centered
 ```{r}
+# Centered
 x <- rstan::extract(fit_cp, permuted=F)
+r_star_centered <- r_star_dist(x)
+mon_cp <- monitor(split_data(x))
 
-r_star_centered <- r_star_dist(x)
-```
-
-Noncentered
-```{r}
+# Noncentered
 x <- rstan::extract(fit_ncp, permuted=F)
 r_star_noncentered <- r_star_dist(x)
-```
+mon_ncp <- monitor(split_data(x))
 
-Store both
-```{r}
 c_df <- tibble(centered=r_star_centered,
-               non_centered=r_star_noncentered)
+               non_centered=r_star_noncentered,
+               metric="R*",
+               simulations="split")
+c1_df <- tibble(centered=mon_cp$Rhat[1:10],
+                non_centered=mon_ncp$Rhat[c(1, 2, 11:18)],
+                metric="split-Rhat",
+               simulations="split")
 ```
@@
 x <- rstan::extract(fit_cp, permuted=F)
-r_star_centered <- r_star_dist(x)
-
+r_star_centered <- r_star_dist(x, split_chains = F)
+mon_cp <- monitor(fit_cp)
 # noncentered
 x <- rstan::extract(fit_ncp, permuted=F)
-r_star_noncentered <- r_star_dist(x)
-```
+r_star_noncentered <- r_star_dist(x, split_chains = F)
+mon_ncp <- monitor(fit_ncp)
 
-```{r}
 d_df <- tibble(centered=r_star_centered,
-               non_centered=r_star_noncentered)
+               non_centered=r_star_noncentered,
+               metric="R*",
+               simulations="unsplit")
+d1_df <- tibble(centered=mon_cp$Rhat[1:10],
+                non_centered=mon_ncp$Rhat[c(1, 2, 11:18)],
+                metric="split-Rhat",
+               simulations="unsplit")
 ```
@@
   c_df %>% 
-  melt(id.vars=NULL) %>% 
+  melt(id.vars=c("metric", "simulations")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   d_df %>% 
-  melt(id.vars=NULL) %>% 
+  melt(id.vars=c("metric", "simulations")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black"))
+  theme(text = element_text(size=14, colour="black"),
+        legend.position = "none")
 # a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
+```
 
-# pdf("../output/eight_schools.pdf", width = 12, height = 6)
-# multiplot(g1, g2, cols = 2)
-# dev.off()
+Look at Rhat distributions
+```{r}
+e_df <- c1_df %>% 
+  mutate(var_num=seq_along(metric)) %>% 
+  bind_rows(d1_df %>% mutate(var_num=seq_along(metric)))
+g3 <- e_df %>%
+  select(-metric) %>% 
+  melt(id.vars=c("simulations", "var_num")) %>% 
+  pivot_wider(id_cols=c("variable", "var_num"), values_from = "value", names_from = "simulations") %>% 
+  ggplot(aes(x=unsplit, y=split, colour=as.factor(variable))) +
+  geom_point() +
+  scale_x_continuous(limits=c(NA, 1.1)) +
+  scale_y_continuous(limits=c(NA, 1.1)) +
+  geom_vline(xintercept = 1.01, linetype=2) +
+  geom_hline(yintercept = 1.01, linetype=2) +
+  geom_abline(linetype=3) +
+  scale_color_grey("Series", labels=c("centered", "non-centered")) +
+  ylab("Split chains") +
+  xlab("Original chains") +
+  ggtitle("C.")
+
+pdf("../output/eight_schools.pdf", width = 12, height = 4)
+multiplot(g1, g2, g3, cols = 3)
+dev.off()
 ```

### 2020-05-04 write_up/ms.tex
added new rhat plot for eight schools

@@
 
+$\widehat{R}$, like $R^*$, is recommended to be calculated using split chains.
+In Fig.
+\ref{fig:eight_schools}C, we plot the values of values of this metric when using the original four chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
+These show that the values of $\widehat{R}$ for the non-centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, the value of $\widehat{R}$ is higher for the split case.
+Finally, as for $R^*$, all parameters for the centered models were below 1.01, indicating convergence.
+
 \begin{figure}[!htb]
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
+	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains vs those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
 In all cases, the plots show 1000 $R^*$ draws.

### 2020-05-04 src/s_ar1.Rmd
changed many results to use the rank split-Rhat

@@
   a_accuracy <- r_star(a_array)
-  mon <- monitor(temp)
-  return(list(r_star=a_accuracy, r_hat=mon$Rhat))
+  mon <- monitor_extra(split_data(a_array))
+  return(list(r_star=a_accuracy, r_hat=mon$zfsRhat))
 }

### 2020-05-04 src/s_cauchy.Rmd
changed many results to use the rank split-Rhat

@@
 
-mon <- monitor(fit_nom)
-a_rhat_df <- tibble(nominal=mon$Rhat[1:50])
+mon <- monitor_extra(split_data(x))
+a_rhat_df <- tibble(nominal=mon$zfsRhat[1:50])
 ```
@@
 r_star_alt <- r_star_dist(full_data)
-mon <- monitor(fit_alt1)
+mon <- monitor_extra(split_data(full_data))
 a_rhat_df <-a_rhat_df %>% 
-  mutate(alt=mon$Rhat[101:150],
+  mutate(alt=mon$zfsRhat[101:150],
          source="r_hat")
@@
                source="r_star") %>% 
-  bind_rows(a_rhat_df)
+  bind_rows(a_rhat_df) %>% 
+  mutate(source=if_else(source=="r_star", "R*", "Rhat"))
 g <-
@@
   scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab(TeX("$R*$")) +
+  xlab("Value") +
   ylab("Count") +
@@
 ```{r}
-fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
-            refresh = 0, iter = 50000)
-saveRDS(fit, "../output/fit_nominal_50000.rds")
+# fit <- stan(file = 'cauchy_nom.stan', seed = 7878,
+#             refresh = 0, iter = 50000)
+# saveRDS(fit, "../output/fit_nominal_50000.rds")
 fit <- readRDS("../output/fit_nominal_50000.rds")
@@
 r_star_nom <- r_star_dist(full_data)
-mon_nom <- monitor(fit)
+mon_nom <- monitor_extra(split_data(full_data))
 ```
@@
 ```{r}
-fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
-            refresh = 0, iter = 50000)
-saveRDS(fit, "../output/fit_alt_50000.rds")
+# fit <- stan(file = 'cauchy_alt_1.stan', seed = 7878,
+#             refresh = 0, iter = 50000)
+# saveRDS(fit, "../output/fit_alt_50000.rds")
 fit <- readRDS("../output/fit_alt_50000.rds")
@@
 r_star_alt <- r_star_dist(full_data)
-mon_alt <- monitor(fit)
+mon_alt <- monitor_extra(split_data(full_data))
 ```
@@
                source="r_star") %>% 
-  bind_rows(tibble(nominal=mon_nom$Rhat[1:50],
-                alt=mon_alt$Rhat[101:150],
+  bind_rows(tibble(nominal=mon_nom$zfsRhat[1:50],
+                alt=mon_alt$zfsRhat[101:150],
                 source="r_hat")) %>% 
@@
 
-log_last <- log10(10000)
+log_last <- log10(25000)
 log_first <- log10(100)
@@
 ```{r}
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+future::plan(multiprocess)
+unranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
 
+x <- z_scale(x)
+ranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
+cc_df <- tibble(ranked, unranked, iter, source="nominal")
+
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+future::plan(multiprocess)
+unranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
+
+x <- z_scale(x)
+ranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
+
+cc_df <- cc_df %>% 
+  bind_rows(tibble(ranked, unranked, iter, source="alt"))
+saveRDS(cc_df, "../output/cauchy_ranked_unranked.rds")
+
+g <- cc_df %>% 
+  mutate(source=if_else(source=="alt", "alternative", "nominal")) %>% 
+  melt(id.vars=c("iter", "source")) %>% 
+  mutate(variable=as.factor(variable)) %>% 
+  ggplot(aes(x=iter, y=value, colour=variable)) +
+  geom_line() +
+  facet_wrap(~source) +
+  scale_x_log10() +
+  scale_color_grey("Method") +
+  xlab("Draws") +
+  ylab("R*")
+ggsave("../output/cauchy_ranked_unranked.pdf", g)
 ```
@@
  nparams <- dim(x)[3] 
- rhat_vals <- map_dbl(seq(1, nparams, 1), ~rhat(split_chains(x[ , , .])))
+ rhat_vals <- map_dbl(seq(1, nparams, 1), ~rhat(split_chains(z_scale(x[ , , .]))))
  return(rhat_vals)
@@
 future::plan(multiprocess)
+iter <- a_df$iteration %>% unique()
 r_hat <- future_map_dbl(iter, ~max(r_hat_all_params(x[1:., , ])))
@@
          variable="nom"))
-# saveRDS(c_df, "../output/cauchy_r_hat.rds")
+saveRDS(c_df, "../output/cauchy_r_hat.rds")
 c_df <- readRDS("../output/cauchy_r_hat.rds")

### 2020-05-04 src/s_eight_schools.Rmd
changed many results to use the rank split-Rhat

@@
 r_star_centered <- r_star_dist(x)
-mon_cp <- monitor(split_data(x))
+mon_cp <- monitor_extra(split_data(x))
 
@@
 r_star_noncentered <- r_star_dist(x)
-mon_ncp <- monitor(split_data(x))
+mon_ncp <- monitor_extra(split_data(x))
 
@@
                simulations="split")
-c1_df <- tibble(centered=mon_cp$Rhat[1:10],
-                non_centered=mon_ncp$Rhat[c(1, 2, 11:18)],
+c1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
+                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
                 metric="split-Rhat",
                simulations="split")
+```
+
+## Look at GBM fitted to data for variable importance
+```{r}
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+x <- rstan::extract(fit_cp, permuted=F)
+x <- split_data(x)
+# Reshape data into form for R* estimation
+m_flattened <- matrix(nrow = nrow(x) * 8, ncol = 12)
+k <- 1
+for(i in 1:8){
+  for(j in 1:nrow(x)){
+    m_flattened[k, 1:11] <- x[j, i, ]
+    m_flattened[k, 12] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame() %>% 
+  rename(chain=V12) %>% 
+  mutate(chain=as.factor(chain))
+
+full_data <- m_flattened
+rand_samples <- sample(1:nrow(full_data), 0.7 * nrow(full_data))
+training_data <- full_data[rand_samples, ]
+testing_data <- full_data[-rand_samples, ]
+
+# Fit gradient-boosed model
+gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = caretGrid, verbose=FALSE)
+varImp(gbmFit1)
 ```
@@
 r_star_centered <- r_star_dist(x, split_chains = F)
-mon_cp <- monitor(fit_cp)
+mon_cp <- monitor_extra(fit_cp)
 # noncentered
@@
 r_star_noncentered <- r_star_dist(x, split_chains = F)
-mon_ncp <- monitor(fit_ncp)
+mon_ncp <- monitor_extra(fit_ncp)
 
@@
                simulations="unsplit")
-d1_df <- tibble(centered=mon_cp$Rhat[1:10],
-                non_centered=mon_ncp$Rhat[c(1, 2, 11:18)],
+d1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
+                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
                 metric="split-Rhat",

### 2020-05-04 src/s_mvt_normal.Rmd
changed many results to use the rank split-Rhat

@@
 ```{r}
-N <- 250
+# N <- 250
 # A <- rWishart(1, 250, diag(N))[,,1]
 # saveRDS(A, "../output/A_matrix.rds")
-A <- readRDS("../output/A_matrix.rds")
-model <- stan_model("mvt_250.stan")
+# A <- readRDS("../output/A_matrix.rds")
+# model <- stan_model("mvt_250.stan")
 # fit <- sampling(model, data=list(N=N, A=A), iter=10000, chains=4, thin=5)
@@
 # Collect Rhat
-a_sum <- monitor(fit)
-rhat <- a_sum$Rhat
+full_data <- rstan::extract(fit, permuted=F)
+a_sum <- monitor_extra(split_data(full_data))
+rhat <- a_sum$zfsRhat
 
 # Reshape data into form for R* estimation
-full_data <- rstan::extract(fit, permuted=F)
 m_flattened <- matrix(nrow = nrow(full_data) * 4, ncol = 252)
@@
 importance <- 100-varImp(gbmFit1)$importance$Overall
-ess_b <- a_sum$Bulk_ESS
-ess_t <- a_sum$Tail_ESS
+ess_b <- a_sum$zsseff
+ess_t <- a_sum$tailseff
 b_df <- tibble(importance, ess_b, ess_t, rhat, lp=c(rep(0, 250), 1))
-sum(rhat<1.01)
-mean(ess_b>400)
-mean(ess_t>400)
+mean(rhat<1.01)
+mean(ess_b<400)
+mean(ess_t<400)
 cor.test(b_df$importance, b_df$rhat, method="spearman")
@@
   theme(legend.position = "none")
-b_df <- tibble(samples_10000_cp=a_sum$Rhat[1:250])
+b_df <- tibble(samples_10000_cp=a_sum$zfsRhat[1:250])
 ```
@@
 ```{r}
+source("monitornew.R")
 # fit <- sampling(model, data=list(N=N, A=A), iter=400, chains=4)
-print(fit)
+# print(fit)
 # saveRDS(fit, "../output/mvt_fit_400.rds")
@@
 
-a_sum <- monitor(fit)
+a_sum <- monitor_extra(split_data(full_data))
 b_df <- b_df %>%
-  mutate(samples_400=a_sum$Rhat[1:250])
+  mutate(samples_400=a_sum$zfsRhat[1:250])
 ```
@@
 fit <- readRDS("../output/mvt_ncp_fit_10000.rds")
-a_sum <- monitor(fit)
-sum(a_sum$Bulk_ESS>400)
-sum(a_sum$Tail_ESS>400)
+full_data <- rstan::extract(fit, permuted=F)
+a_sum <- monitor_extra(split_data(full_data))
+sum(a_sum$zsseff>400)
+sum(a_sum$tailseff>400)
+sum(a_sum$zfsRhat<1.01)
 b_df <- b_df %>%
-  mutate(samples_10000_ncp=a_sum$Rhat[251:500])
+  mutate(samples_10000_ncp=a_sum$zfsRhat[251:500])
 saveRDS(b_df, "../output/mvt_fit_all_rhat.rds")
@@
 
-full_data <- rstan::extract(fit, permuted=F)
 r_star3 <- r_star_dist(full_data)

### 2020-05-04 src/s_wide_distributions.Rmd
changed many results to use the rank split-Rhat

@@
 
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
+mon <- monitor_extra(split_data(full_data))
+mean(mon$zfsRhat>1.01)
+mean(mon$zsseff<400)
+mean(mon$tailseff<400)
 
-b_df <- tibble(samples_cp=mon$Rhat[1:250])
+b_df <- tibble(samples_cp=mon$zfsRhat[1:250])
 
-ha <- tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat) %>% 
+ha <- tibble(bulk_ess=mon$zsseff,
+       tail_ess=mon$tailseff,
+       rhat=mon$zfsRhat) %>% 
   melt() %>% 
@@
 ha
-aa_df <- tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat,
+aa_df <- tibble(bulk_ess=mon$zsseff,
+       tail_ess=mon$tailseff,
+       rhat=mon$zfsRhat,
        source="Centered")
@@
 
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
+mon <- monitor_extra(split_data(full_data))
+mean(mon$zfsRhat>1.01)
+mean(mon$zsseff<400)
+mean(mon$tailseff<400)
 
-h <- tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat) %>% 
+h <- tibble(bulk_ess=mon$zsseff,
+       tail_ess=mon$tailseff,
+       rhat=mon$zfsRhat) %>% 
   melt() %>% 
@@
 aa_df <- aa_df %>%
-  bind_rows(tibble(bulk_ess=mon$Bulk_ESS,
-       tail_ess=mon$Tail_ESS,
-       rhat=mon$Rhat,
+  bind_rows(tibble(bulk_ess=mon$zsseff,
+       tail_ess=mon$tailseff,
+       rhat=mon$zfsRhat,
        source="Non-centered"))
@@
 # saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
+mon <- monitor_extra(split_data(full_data))
+mean(mon$zfsRhat>1.01)
+mean(mon$zsseff<400)
+mean(mon$tailseff<400)
 ```
@@
 ```{r}
-N <- 10000
-A <- diag(nrow=N)
-model <- stan_model("mvt_wide.stan")
-fit <- sampling(model, data=list(N=N), iter=1000, chains=4)
-saveRDS(fit, "../output/mvt_wide_1000.rds")
+# N <- 10000
+# A <- diag(nrow=N)
+# model <- stan_model("mvt_wide.stan")
+# fit <- sampling(model, data=list(N=N), iter=1000, chains=4)
+# saveRDS(fit, "../output/mvt_wide_1000.rds")
 
@@
 saveRDS(a_df, "../output/mvt_wide_summaries.rds")
-mon <- monitor(fit)
-mean(mon$Rhat>1.01)
-mean(mon$Bulk_ESS<400)
-mean(mon$Tail_ESS<400)
+mon <- monitor_extra(split_data(full_data))
+mean(mon$zfsRhat>1.01)
+mean(mon$zsseff<400)
+mean(mon$tailseff<400)
 a_df <- readRDS("../output/mvt_wide_summaries.rds")
@@
   ggtitle("B.")
-# ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
+ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
 ```

### 2020-05-04 write_up/ms.tex
changed many results to use the rank split-Rhat

@@
 For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically less than 1 second to calculate).
+For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically O(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
 It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.03): in this case, $\widehat{R}<1.01$ for all parameters (Fig.
-\ref{fig:mvt}B), although 60\% of parameters had tail $\text{ESS}<400$ and 39\% had bulk $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.03): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.
+\ref{fig:mvt}B), although 54\% had bulk $\text{ESS}<400$ and 13\% of parameters had tail $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
 
@@
 We now compare these with the more established metrics $\widehat{R}$ and ESS as calculated in Stan.
-For a GBM fitted to the centered model, we plot in Fig.
+For a GBM fitted to the centered model with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig.
 \ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-From this plot, there appears a weak positive association between GBM's variable importance and $\widehat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.10, S=2368138, p=0.11$).
+From this plot, there was a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$).
 In Fig.
-\ref{fig:mvt}D, we plot variable importance versus two measures: bulk ESS and tail ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk ESS: $\rho=-0.54, S=4146491, p<0.01$; tail ESS: $\rho=-0.55, S=4095040, p<0.01$).
+\ref{fig:mvt}D, we plot variable importance versus two measures: bulk ESS and tail ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk ESS: $\rho=-0.57, S=4142470, p<0.01$; tail ESS: $\rho=-0.56, S=4113709, p<0.01$).
 Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
@@
 In the first of these, we used the 250-dimensional multivariate normal of eq.
-\eqref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
+\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
 Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
@@
 The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
-\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\widehat{R}>1.01$.
+\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, some of the parameters had $\widehat{R}>1.01$.
 Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
@@
 In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by rank-normalised split-$\widehat{R}$, with 28\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
+These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.
 
@@
 
-As in \S\ref{sec:multivariate_normal}, we calculated a measure of variable importance to understand which variables contributed most to the predictive power of the model.
-For the GBM fitted to the centered model, the most important variable was $\tau$, followed by Stan's $lp$ variable (we also include this in our training and testing data), followed by $\mu$ and $\theta_2$.
-Interestingly, these were ordered differently according to Stan's $\widehat{R}$, with $lp$ having the highest value ($\hat R = 1.06$), followed by $\tau$ ($\hat R = 1.02$), then a host of variables with ($\hat R = 1.01$).
-
 In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains vs those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
+	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
 In all cases, the plots show 1000 $R^*$ draws.

### 2020-05-06 src/s_ar1.Rmd
added sensitivity plots

@@
 ```
+
+# Show performance as a function of hyperparameters
+```{r}
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid1 <- expand.grid(interaction.depth=c(5), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid2 <- expand.grid(interaction.depth=c(3), n.trees = 100,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid3 <- expand.grid(interaction.depth=c(5), n.trees = 100,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid4 <- expand.grid(interaction.depth=c(7), n.trees = 100,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid5 <- expand.grid(interaction.depth=c(7), n.trees = 200,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+lgrids <- list(caretGrid, caretGrid1, caretGrid2, caretGrid3, caretGrid4, caretGrid5)
+f_replicate_hyper <- function(lgrids){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  laccuracy <- vector(length = length(lgrids))
+  for(i in seq_along(lgrids))
+    laccuracy[i] <- r_star(a_array, caret_default=lgrids[[i]])
+  return(laccuracy)
+}
+
+nreplicates <- 200
+r_star_vals <- matrix(nrow = nreplicates, ncol=length(lgrids))
+for(i in 1:nreplicates){
+  if(i%%100==0)
+    print(i)
+
+  temp <- f_replicate_hyper(lgrids)
+  r_star_vals[i, ] <- temp
+}
+
+saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
+
+b_sum <-
+  r_star_vals %>% 
+  as.data.frame() %>% 
+  mutate(iterate=seq(1, nreplicates, 1)) %>%
+  melt(id.vars="iterate") %>% 
+  mutate(below=if_else(value < 1, 1, 0)) %>% 
+  group_by(variable) %>% 
+  summarise(below_per=round(mean(below)*100, 1),
+            lower=quantile(value, 0.25),
+            upper=quantile(value, 0.75),
+            middle=median(value)) %>%
+  mutate(below_per=paste0(below_per, "%")) %>% 
+  mutate(x_pos=seq(1, 6, 1)) %>% 
+  mutate(y_pos=0.95)
+
+g <- r_star_vals %>% 
+  as.data.frame() %>% 
+  mutate(iterate=seq(1, nreplicates, 1)) %>% 
+  melt(id.vars="iterate") %>% 
+  mutate(below=if_else(value < 1, 0, 1)) %>% 
+  left_join(b_sum, by = "variable") %>% 
+  mutate(variable=case_when(variable=="V1"~"int.depth=3\nn.trees=50",
+                            variable=="V2"~"int.depth=5\nn.trees=50",
+                            variable=="V3"~"int.depth=3\nn.trees=100",
+                            variable=="V4"~"int.depth=5\nn.trees=100",
+                            variable=="V5"~"int.depth=7\nn.trees=100",
+                            variable=="V6"~"int.depth=7\nn.trees=200")) %>% 
+  ggplot(aes(x=as.factor(variable), y=value, colour=as.factor(below))) +
+  geom_jitter() +
+  xlab("Hyperparameters") +
+  ylab("R*") +
+  geom_hline(yintercept = 1, linetype=2) +
+  annotate("text", x = b_sum$x_pos, y = b_sum$y_pos, label = b_sum$below_per, size=5) +
+  annotate("pointrange", x = 1:6, ymin=b_sum$lower, ymax=b_sum$upper, y = b_sum$middle, colour="black", size=1) +
+  coord_flip() +
+  scale_color_grey() +
+  theme(axis.text = element_text(colour="black", size=12),
+        axis.title = element_text(colour="black", size=14),
+        legend.position = "none")
+ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
+```

### 2020-05-06 src/s_ovarian_prostate.Rmd
added sensitivity plots

@@
 
-mon <- monitor(fit)
+x <- rstan::extract(fit, permuted=F)
+mon <- monitor_extra(split_data(x))
 saveRDS(mon, "../data/ovarian_long_mon.rds")
@@
 
-mon <- monitor(fit)
+x <- rstan::extract(fit, permuted=F)
+mon <- monitor_extra(split_data(x))
 saveRDS(mon, "../data/ovarian_short_mon.rds")
@@
 g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
-  bind_rows(tibble(long=mon_longer$Rhat, short=mon_shorter$Rhat, metric="B. Rhat")) %>% 
-  bind_rows(tibble(long=mon_longer$Bulk_ESS, short=mon_shorter$Bulk_ESS, metric="C. Bulk ESS")) %>% 
-  bind_rows(tibble(long=mon_longer$Tail_ESS, short=mon_shorter$Tail_ESS, metric="D. Tail ESS")) %>% 
+  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. split-Rhat")) %>% 
+  bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk ESS")) %>% 
+  bind_rows(tibble(long=mon_longer$tailseff, short=mon_shorter$tailseff, metric="D. Tail ESS")) %>% 
   melt(id.vars="metric") %>% 
-  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
+  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. split-Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
 # Prostate
-
 ```{r}
@@
 
-mon <- monitor(fit)
+x <- rstan::extract(fit, permuted=F)
+mon <- monitor_extra(split_data(x))
 saveRDS(mon, "../data/prostate_long_mon.rds")
@@
 
-mon <- monitor(fit)
+x <- rstan::extract(fit, permuted=F)
+mon <- monitor_extra(split_data(x))
 saveRDS(mon, "../data/prostate_short_mon.rds")

### 2020-05-06 write_up/ms.tex
added sensitivity plots

@@
 
+To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we perform a sensitivity analysis.
+In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data as were used to generate Fig.
+\ref{fig:ar1}.
+In particular, the GBM hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
+The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
+\ref{fig:ar1_sensitivity}.
+These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper).
+This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
+	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to different GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
+The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
+The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:ar1_sensitivity}
+\end{figure}
+
 \subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
@@
 
+Fig.
+\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in ML classification accuracy and falsely concluding that convergence has occurred.
+
 \begin{figure}[!htb]
@@
 \subsection{Many parameter models: ovarian and prostate analysis}
-In this section, we...
+In this section, we analyse two Bayesian models [Aki, can you provide some description of these?].
+These models have relatively many parameters (4719 for the ``ovarian'' model and 18,105 for the ``prostate'' model) and are known to be challenging to explore with MCMC.
+For each model, we consider two MCMC runs: one with 4 chains run with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
+
+For the ovarian model, we show the results in Fig.
+\ref{fig:ovarian}.
+In Fig.
+\ref{fig:ovarian}A, we show the $R^*$ distributions for each model run, which show that, whereas the ``long'' model run has converged, the ``short'' one has yet to do so.
+Whilst it is harder to discern, this pattern is mirrored in Fig.
+\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters [check this!], whereas the short model has a handful of parameters where this is not the case.
+Similarly, in Figs.
+\ref{fig:ovarian}C and \ref{fig:ovarian}D, which show the bulk ESS and tail ESS respectively, it is evident that, for the short model, there remain a few parameters with low effective sample sizes, whereas the long model has more consistent values for this metric.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/ovarian.pdf}}
+	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk ESS across all parameters; and D shows tail ESS across parameters.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:ovarian}
+\end{figure}
+
+For the prostate model, we show the results in Fig.
+\ref{fig:prostate}.
+Since this model has nearly four times as many parameters as the ovarian model, it was more computationally expensive to estimate $R^*$ for it.
+To handle this, we thinned down the parameters by a factor of 5 for the long model, recognising that, of course, this measure will make it more likely that we diagnose convergence.
+Despite this, both $R^*$ measures indicated that the MCMC runs had yet to converge (Fig.
+\ref{fig:prostate}A), which was mirrored by the other metrics considered (Figs.
+\ref{fig:prostate}B,C\&D).
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/prostate.pdf}}
+	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk ESS across all parameters; and D shows tail ESS across parameters.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:prostate}
+\end{figure}
 

### 2020-05-06 src/r_star_monitor.R

@@ -22,7 +22,7 @@ library(tidyverse)
 #' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
-r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
+r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL){
   
   if(split_chains)
     x <- split_data(x)
@@ -54,7 +54,7 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]
   
-  if(is.na(caret_default))
+  if(is.null(caret_default))
     caretGrid <- expand.grid(interaction.depth=c(3), 
                              n.trees = 50,
                              shrinkage=c(0.1),
@@ -96,7 +96,7 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NA){
 #' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
-r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_default=NA, nsim=1000){
+r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, nsim=1000){
   
   if(split_chains)
     x <- split_data(x)
@@ -124,7 +124,7 @@ r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_default=N
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]
   
-  if(is.na(caret_default))
+  if(is.null(caret_default))
     caretGrid <- expand.grid(interaction.depth=c(3), 
                              n.trees = 50,
                              shrinkage=c(0.1),

### 2020-05-06 src/s_ar1.Rmd
updated ar1 figures

--- 2020-05-03 src/s_ar1.Rmd
@@
 f_replicate <- function(){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
-  a_array <- array(dim=c(1000, 4, 1))
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 2000)
+  a_array <- array(dim=c(2000, 4, 1))
   a_array[,,1] <- temp
   a_accuracy <- r_star(a_array)
-  mon <- monitor(temp)
-  return(list(r_star=a_accuracy, r_hat=mon$Rhat))
+  mon <- monitor_extra(split_data(a_array))
+  return(list(r_star=a_accuracy, r_hat=mon$zfsRhat))
 }
@@
 # add in a column of noise since gbm requires >1 variables
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) %>% 
+full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 2000) %>% 
   melt() %>% 
@@
   rename(chain=Var2) %>% 
-  mutate(ran_val=rnorm(4000)) %>% 
+  mutate(ran_val=rnorm(8000)) %>% 
   mutate(chain=as.factor(chain))
@@
 
-Generate R* distribution based on 1000 iterations for run where chain has lower variance and another where all four chains have same variance
+Generate R* distribution based on 2000 iterations for run where chain has lower variance and another where all four chains have same variance
 ```{r}
-temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000) 
-temp1 <- matrix(rnorm(4000), nrow = 1000)
-a_array <- array(dim=c(1000, 4, 2))
+temp <- f_generate_lower_var_four(1/3, 0.3, 1, 2000) 
+temp1 <- matrix(rnorm(8000), nrow = 2000)
+a_array <- array(dim=c(2000, 4, 2))
 a_array[,,1] <- temp
@@
 
-temp <- f_generate_lower_var_four(1, 0.3, 1, 1000)
-temp1 <- matrix(rnorm(4000), nrow = 1000)
-a_array <- array(dim=c(1000, 4, 2))
+temp <- f_generate_lower_var_four(1, 0.3, 1, 2000)
+temp1 <- matrix(rnorm(8000), nrow = 1000)
+a_array <- array(dim=c(2000, 4, 2))
 a_array[,,1] <- temp
@@
 ```
+
+# Show performance as a function of hyperparameters
+```{r}
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid1 <- expand.grid(interaction.depth=c(5), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid2 <- expand.grid(interaction.depth=c(3), n.trees = 100,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid3 <- expand.grid(interaction.depth=c(5), n.trees = 100,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid4 <- expand.grid(interaction.depth=c(7), n.trees = 100,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+caretGrid5 <- expand.grid(interaction.depth=c(7), n.trees = 200,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
+lgrids <- list(caretGrid, caretGrid1, caretGrid2, caretGrid3, caretGrid4, caretGrid5)
+f_replicate_hyper <- function(lgrids){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  laccuracy <- vector(length = length(lgrids))
+  for(i in seq_along(lgrids))
+    laccuracy[i] <- r_star(a_array, caret_default=lgrids[[i]])
+  return(laccuracy)
+}
+
+nreplicates <- 200
+r_star_vals <- matrix(nrow = nreplicates, ncol=length(lgrids))
+for(i in 1:nreplicates){
+  if(i%%100==0)
+    print(i)
+
+  temp <- f_replicate_hyper(lgrids)
+  r_star_vals[i, ] <- temp
+}
+
+saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
+
+b_sum <-
+  r_star_vals %>% 
+  as.data.frame() %>% 
+  mutate(iterate=seq(1, nreplicates, 1)) %>%
+  melt(id.vars="iterate") %>% 
+  mutate(below=if_else(value < 1, 1, 0)) %>% 
+  group_by(variable) %>% 
+  summarise(below_per=round(mean(below)*100, 1),
+            lower=quantile(value, 0.25),
+            upper=quantile(value, 0.75),
+            middle=median(value)) %>%
+  mutate(below_per=paste0(below_per, "%")) %>% 
+  mutate(variable=case_when(variable=="V1"~"int.depth=3\nn.trees=50",
+                            variable=="V2"~"int.depth=5\nn.trees=50",
+                            variable=="V3"~"int.depth=3\nn.trees=100",
+                            variable=="V4"~"int.depth=5\nn.trees=100",
+                            variable=="V5"~"int.depth=7\nn.trees=100",
+                            variable=="V6"~"int.depth=7\nn.trees=200")) %>% 
+  arrange(variable) %>% 
+  mutate(x_pos=seq(1, 6, 1)) %>% 
+  mutate(y_pos=0.95)
+
+g <- r_star_vals %>% 
+  as.data.frame() %>% 
+  mutate(iterate=seq(1, nreplicates, 1)) %>% 
+  melt(id.vars="iterate") %>% 
+  mutate(below=if_else(value < 1, 0, 1)) %>% 
+  mutate(variable=case_when(variable=="V1"~"int.depth=3\nn.trees=50",
+                            variable=="V2"~"int.depth=5\nn.trees=50",
+                            variable=="V3"~"int.depth=3\nn.trees=100",
+                            variable=="V4"~"int.depth=5\nn.trees=100",
+                            variable=="V5"~"int.depth=7\nn.trees=100",
+                            variable=="V6"~"int.depth=7\nn.trees=200")) %>% 
+  left_join(b_sum, by = "variable") %>%
+  ggplot(aes(x=as.factor(variable), y=value, colour=as.factor(below))) +
+  geom_jitter() +
+  xlab("Hyperparameters") +
+  ylab("R*") +
+  geom_hline(yintercept = 1, linetype=2) +
+  annotate("text", x = b_sum$x_pos, y = b_sum$y_pos, label = b_sum$below_per, size=5) +
+  annotate("pointrange", x = 1:6, ymin=b_sum$lower, ymax=b_sum$upper, y = b_sum$middle, colour="black", size=1) +
+  coord_flip() +
+  scale_color_grey() +
+  theme(axis.text = element_text(colour="black", size=12),
+        axis.title = element_text(colour="black", size=14),
+        legend.position = "none")
+ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
+```

### 2020-05-06 src/s_mvt_normal.Rmd
updated ar1 figures

@@
   melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
   geom_histogram(position="identity", alpha=0.8, bins=20) +

### 2020-05-06 write_up/ms.tex
updated ar1 figures

@@
 Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
-Since the used machine learning classifier, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
+Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
 The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
@@
 MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions).
-Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small finite number of draws.
+Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small number of draws.
 
@@
 Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter draws rather than the raw values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}.
-Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
+Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
 Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.
@@
 By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between the chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
+To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
 Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
@@
 For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically O(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
+For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
 It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
@@
 
-The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \citep{vehtari2019rank} and elsewhere.
+The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank} and elsewhere.
 Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
-	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
+	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (with marginal distribution to the right of panel); B shows two chains that have not mixed.}
 	\label{fig:marginal}
@@
 We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing.
-Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters.
-In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).
+Our GBM model was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:results}).
+Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.
 
-We can also get class probabilities, which we leverage to produce an uncertainty distribution for $R^*$.
+From the GBM fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
 Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
@@
 		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N\bar{a}$
+		\STATE Calculate ratio to null model accuracy, $R^{*\{i\}} = \bar{a} / (1 / N) = N\bar{a}$
 		\ENDFOR
-		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
+		\RETURN $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$
 	\end{algorithmic}
@@
 %
-where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,1000$.
+where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$.
 Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others.
@@
 The fitted model is then used to classify draws in an independent test set according to the chain which generated them.
-For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}.
+For each replicate, we then calculated $R^*$ as described in Algorithm \ref{alg:R_star}.
 In Fig.
-\ref{fig:ar1}B, we show that $R^*>1$ for 997/1000 replicates, indicating that the chains have not converged in a vast majority of cases.
+\ref{fig:ar1}B, we show that $R^*>1$ for all replicates, indicating that the chains had not converged in all cases.
 In Fig.
-\ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate; as for $R^*$, this metric indicates the chains have not converged in all replicates.
+\ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate;  as for $R^*$, this metric indicates the chains had not converged in all replicates (as diagnosed by $R^*<1.01$).
 
@@
 
-To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we perform a sensitivity analysis.
-In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data as were used to generate Fig.
+To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we performed a sensitivity analysis.
+In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
 \ref{fig:ar1}.
-In particular, the GBM hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
+The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
 The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
-	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to different GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
+	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
 The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
@@
 To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density.
-In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
+In all 4 chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
 By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.
 
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS.
+First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail ESS.
 In all cases, the various quantities were calculated based on chains split into halves.
@@
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.03): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.
-\ref{fig:mvt}B), although 54\% had bulk $\text{ESS}<400$ and 13\% of parameters had tail $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.
+\ref{fig:mvt}B), although 54\% had bulk-$\text{ESS}<400$ and 13\% of parameters had tail-$\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
 
@@
 Fig.
-\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is closer to $R^*=1$ (mean $R^*=1.01$).
+\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is closer to $R^*=1$ (mean $R^*=1.00$).
 Fig.
@@
 
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power.
-We now compare these with the more established metrics $\widehat{R}$ and ESS as calculated in Stan.
+In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions.
+We now compare these with the more established metrics $\widehat{R}$ and ESS.
 For a GBM fitted to the centered model with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig.
@@
 In Fig.
-\ref{fig:mvt}D, we plot variable importance versus two measures: bulk ESS and tail ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk ESS: $\rho=-0.57, S=4142470, p<0.01$; tail ESS: $\rho=-0.56, S=4113709, p<0.01$).
+\ref{fig:mvt}D, we plot variable importance versus two measures: bulk-ESS and tail-ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk-ESS: $\rho=-0.57, S=4142470, p<0.01$; tail-ESS: $\rho=-0.56, S=4113709, p<0.01$).
 Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
@@
 In the first of these, we used the 250-dimensional multivariate normal of eq.
-\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
+\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
 Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
@@
 	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
-In all cases, 1000 draws of $R^*$ are plotted as generated by Alg.
-\ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.
+In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 \ref{fig:cauchy}, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of four chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
 The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$, yet, still the mean is above this value.
@@
 
-Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of four chains.
+Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of 4 chains.
 In the right-hand column of Fig.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} Columns show MCMC runs with 1000 (left) and 10,000 (right; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of four chains.
+	\caption{\textbf{Cauchy example.} Columns show MCMC runs with 1000 (left) and 10,000 (right; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
 Rows show the $R^*$ distributions (top) and rank-normalised split-$\widehat{R}$ values across all parameters (bottom).
-The different shadings indicate different model paramerisations as indicated in legend.
+Shadings indicate different model paramerisations as indicated in legend.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 This means that approximate KL-divergence is unreliable for these types of model.
-Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
+Additionally, we found that the Anderson-Darling and Cram\'er-Von Mises tests \citep{faraway2019goftest} behaved equally erratically and provided measures that were hard to intuit; instead, we chose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
 Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
@@
 	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
+Dashed lines indicate recommended thresholds for each convergence statistic.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 This example showed that split-$\widehat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
-Specifically, they first generated four chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling.
+Specifically, they first generated 4 chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling.
 We first repeat this analysis but using $R^*$ rather than $\widehat{R}$: in Fig.
 \ref{fig:trends_all_dim}, we show these results.
-In column A, we show the results for $R^*$ calculated on the four chains that ran; column B shows the same calculation but after the chains are split into two equal halves.
+In column A, we show the results for $R^*$ calculated on the 4 chains that ran; column B shows the same calculation but after the chains are split into two equal halves.
 The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_all_dim.pdf}}
-	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the four chains; column B  shows the same calculation after each chain is split into two halves.
+	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
 The rows present the differing sample sizes.
@@
 \subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \citep{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
 
-The model can be parameterised two ways, as described in \cite{vehtari2019rank}.
+The model can be parameterised in two ways, as described in \cite{vehtari2019rank}.
 The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
@@
 %
-The ``non-centered'' parameterisation recodes this model in a way which does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
+The ``non-centered'' parameterisation recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
 This can be written as,
@@
 Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
-We also used Stan with the same algorithm settings to sample from the non-centered model.
+We also used the same algorithm settings to sample from the non-centered model.
 
@@
 The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
-In this plot, the centered model shows signs of convergence, whereas the non-centered model does not.
+In this plot, the centered model is close to convergence, whereas the non-centered is not.
 
@@
 In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each draw, meaning that the centered model $R^*$ values are shifted leftwards.
-Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears converged.
+Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears near convergence.
 
-$\widehat{R}$, like $R^*$, is recommended to be calculated using split chains.
+It is recommended that $\widehat{R}$, like $R^*$, be calculated using split chains.
 In Fig.
-\ref{fig:eight_schools}C, we plot the values of values of this metric when using the original four chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the non-centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, the value of $\widehat{R}$ is higher for the split case.
-Finally, as for $R^*$, all parameters for the centered models were below 1.01, indicating convergence.
+\ref{fig:eight_schools}C, we plot the values of values of this metric when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
+These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
+All parameters for the non-centered models were below 1.01, indicating convergence.
 
@@
 	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
+The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
 In all cases, the plots show 1000 $R^*$ draws.
@@
 
-To understand the predictive power of the ML better, we investigated how predictive accuracy varies across parameter space.
+To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
 After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
@@
 Above the top panel, we show the marginal distributions for each chain.
-In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all the draws.
+In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
 In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
@@
 Whilst it is harder to discern, this pattern is mirrored in Fig.
-\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters [check this!], whereas the short model has a handful of parameters where this is not the case.
+\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 139 parameters where this is not the case.
 Similarly, in Figs.
-\ref{fig:ovarian}C and \ref{fig:ovarian}D, which show the bulk ESS and tail ESS respectively, it is evident that, for the short model, there remain a few parameters with low effective sample sizes, whereas the long model has more consistent values for this metric.
+\ref{fig:ovarian}C and \ref{fig:ovarian}D, which show the bulk-ESS and tail-ESS respectively, it is evident that, for the short model, there remain a few parameters with low effective sample sizes, whereas the long model has more consistent values for this metric.
 
@@
 	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk ESS across all parameters; and D shows tail ESS across parameters.
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk ESS across all parameters; and D shows tail ESS across parameters.
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects.
-Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise.
+Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
 This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
@@
 With gradient-boosted regression tree models, we found this not to be the case.
-Indeed, after a little experimenting, we found that a single set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples.
-To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and we recommend that applied users consider including this step in their workflow.
+Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples.
+To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
 
 Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
-In a number of examples, we trial this before calculating $R^*$, and find that this approach leads to more accurate chain prediction.
+In a number of examples, we trial this before calculating $R^*$ and find that this approach leads to more accurate chain prediction.
 As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
@@
 
+\section{Contributions}
+BL conceived of the original idea for $R^*$, carried out all analyses and wrote the original draft of the paper.
+AV reviewed the paper and made a host of recommendations: including suggesting a large range of new case studies on which to trial the method and also suggesting new visualisations for diagnosing how predictive performance depends on parameter quantiles; these collectively widened the scope of the original paper and substantially improved its quality.
+BL and AV reviewed and edited the draft of the paper.
+
 \bibliographystyle{chicago}

### 2020-05-06 src/s_cauchy.Rmd
updated cauchy distribution plot

@@
   bind_rows(a_rhat_df) %>% 
-  mutate(source=if_else(source=="r_star", "R*", "Rhat"))
+  mutate(source=if_else(source=="r_star", "R*", "split-Rhat"))
 g <-

### 2020-08-01 src/s_cauchy.Rmd
started on aki comments

@@
                 source="r_hat")) %>% 
-  mutate(iter=10000)
+  mutate(iter="10,000")
 
-c_df <- a_df %>% mutate(iter=1000) %>% 
+c_df <- a_df %>% mutate(iter="1000") %>% 
   bind_rows(b_df) %>% 
@@
   mutate(source=if_else(source=="r_star", "R*", source))
-
+saveRDS(c_df, "cauchy_rhat_rstar.rds")
+c_df <- readRDS("cauchy_rhat_rstar.rds")
 my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.03) else c(0.9, 3)}
@@
   ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
+  geom_histogram(bins=50, position="identity", alpha=0.8) +
   scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +

### 2020-08-01 src/s_mvt_normal.Rmd
started on aki comments

--- 2020-05-04 src/s_mvt_normal.Rmd
@@
   melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
+  geom_histogram(position="identity", alpha=0.8, bins=50) +
+  scale_fill_grey("# of draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
   xlab(TeX("$R*$")) +
@@
   melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
+  geom_histogram(position="identity", alpha=0.8, bins=60) +
+  scale_fill_grey("# of draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
   xlab(TeX("split-$\\hat{R}$")) +

### 2020-08-01 write_up/ms.tex
updated cauchy distribution plot
started on aki comments

@@
 
+\newcommand{\beginsupplement}{%
+	\setcounter{table}{0}
+	\renewcommand{\thetable}{S\arabic{table}}%
+	\setcounter{figure}{0}
+	\renewcommand{\thefigure}{S\arabic{figure}}%
+	\setcounter{section}{0}
+	\renewcommand{\thesection}{S\arabic{section}}%
+}
+
 \title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines}
@@
 To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each draw belongs.
-By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
+By evaluating its performance on a separate test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
 By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{gelman2013bayesian}.
@@
 To illustrate the versatility of $R^*$, we use a range of examples that demonstrate how this statistic fares across a range of scenarios.
+Table \ref{tab:results} summarises the examples and provides a rationale for their inclusion.
+Alongside the examples included in the main text, there are a number of supplementary text examples: in \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of samples; in \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain is diagnosed; in \S\ref{sec:prostate}, we show that $R^*$ performs well for a Bayesian logistic regression problem with a large number of parameters and a highly multimodal posterior.
+
+\begin{table}[]
+	\begin{tabular}{l|l|l}
+		Example &  Relevance & Section \\
+		\midrule
+		Autoregressive & Heterogeneous chain variance & \ref{sec:heterogeneity} \\
+		Bivariate normal & Converged marginals but unconverged joint distribution &  \ref{sec:multivariate_normal}\\
+		250D normal & High correlations between dims & \ref{sec:multivariate_normal}\\
+		Wide multivariate normal & \# samples $\sim$ \# dims & \ref{sec:wide}\\
+		Cauchy & Long-tailed distribution & \ref{sec:cauchy}\\
+		Non-stationary marginals & Time-varying sampling distribution & \ref{sec:non-stationary}\\
+		Eight schools model & Hierarchical Bayesian model slow convergence & \ref{sec:eight_shools}\\
+		Ovarian and prostate models & Bayesian models with many parameters & \ref{sec:prostate}
+	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
+\end{table}
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows 1000 $R^*$ draws as generated by Algorithm \ref{alg:R_star_uncertainty} across 1000 replicates; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows the $R^*$ distributions for the two series described in the text.
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows 1000 $R^*$ draws as generated by Algorithm \ref{alg:R_star_uncertainty} across 1000 replicates; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows the $R^*$ distributions for two series, the first being the same as used for figures A-C; the second where all chains have the same distribution as described in \S\ref{sec:heterogeneity_uncertainty}.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 
+\subsubsection{Uncertainty distribution for $R^*$}\label{sec:heterogeneity_uncertainty}
 Since GBMs return a probability simplex for each draw, which, for each chain, indicates the probability that the draw was generated by it, we can also generate a measure of uncertainty in $R^*$.
@@
 For the converged data, the distribution is centred on 1, indicating convergence.
+In the converged case, there are many draws where $R^*<1$: these indicate that, in that particular draw from the probability simplex, chain classification is actually worse than selecting a chain identification uniformly at random.
+Much like how it is possible for $\hat{R}<1$, this is a sample property, driven by the underlying uncertainty when sampling from the categorical distribution defined by the probability simplex.
 
-To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we performed a sensitivity analysis.
+\subsubsection{Robustness to hyperparameter choice}\label{sec:heterogeneity_robustness}
+The performance of GBM, like all ML methods, depends on its hyperparameters.
+To investigate the robustness of $R^*$ to the GBM hyperparameter choice, we performed a sensitivity analysis.
 In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
@@
 To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density.
-In all 4 chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
+In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
 By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.
 
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail ESS.
+First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail-ESS.
 In all cases, the various quantities were calculated based on chains split into halves.
@@
 
+\subsubsection{250-dimensional model}
 We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}.
@@
 
+\subsubsection{Variable importance}
 In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions.
@@
 	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example with 250 dimensions.} A shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 1000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 1000 draws); B shows the rank-normalised split-$\widehat{R}$ values for all parameters from the same MCMC runs as in A; C shows variable importance versus $\widehat{R}$ for each parameter; and D shows variable importance versus bulk- and tail-ESS as calculated by \cite{vehtari2019rank}.
+	\caption{\textbf{Multivariate normal example with 250 dimensions.} A shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws); B shows the rank-normalised split-$\widehat{R}$ values for all parameters from the same MCMC runs as in A; C shows variable importance versus $\widehat{R}$ for each parameter; and D shows variable importance versus bulk- and tail-ESS as calculated by \cite{vehtari2019rank}.
 In A, 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} are shown for each MCMC run.
@@
 
-\subsection{Wide datasets: multivariate normal}\label{sec:wide}
-As the number of parameter dimensions increases, it might be thought that ML algorithms will overfit the data, and, hence, testing set classification would be poor; leading to unreliable determinations of convergence.
-To test this hypothesis, we investigated two scenarios using a multivariate normal target.
-In the first of these, we used the 250-dimensional multivariate normal of eq.
-\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
-Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
-\ref{fig:mvt_wide_both}A.
-In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
-The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
-\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, some of the parameters had $\widehat{R}>1.01$.
-Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
-\ref{fig:wide_both_diagnostics}B\&C).
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_wide_both.pdf}}
-	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
-In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.
-See the R markdown file on the Github repository for code to reproduce this example.}
-	\label{fig:mvt_wide_both}
-\end{figure}
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
-	\caption{\textbf{Wide data 250-dimensional example: established diagnostics.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
-Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
-In both cases, the results correspond to a single MCMC run composed of 4 chains.
-See the R markdown file on the Github repository for code to reproduce this example.}
-	\label{fig:wide_both_diagnostics}
-\end{figure}
-
-We next consider a more challenging example -- a target distribution with 10,000 dimensions.
-In this case, we assume independent standard normals for each dimension.
-In Fig.
-\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
-In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.
-
-Overall, the examples in this section suggest that $R^*$ is robust to wide datasets; we also note that the statistic took comparable time to calculate relative to existing convergence diagnostics on a desktop computer.
-
-\subsection{Infinite variance: Cauchy example}
+\subsection{Infinite variance: Cauchy example}\label{sec:cauchy}
 We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
@@
 
+\subsubsection{Measuring convergence objectively}
 To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target.
@@
 This means that approximate KL-divergence is unreliable for these types of model.
-Additionally, we found that the Anderson-Darling and Cram\'er-Von Mises tests \citep{faraway2019goftest} behaved equally erratically and provided measures that were hard to intuit; instead, we chose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
+We decided not to use the Kolmogorov-Smirnov (KS) test, since it is most sensitive to differences between distributions around the median, whereas, here, we are interested in the behaviour in the tails.
+Additionally, we found that the Anderson-Darling and Cram\'er-Von Mises tests \citep{faraway2019goftest}, which do not suffer the same shortcomings as the KS, behaved equally erratically and provided measures that were hard to intuit.
+The Wasserstein distance was also trialled but had great uncertainty due to the long-tails of the Cauchy.
+Instead, we chose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
 Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
@@
 
-\subsection{Non-stationary marginals}
+\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
+We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+
+The model can be parameterised in two ways, as described in \cite{vehtari2019rank}.
+The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
+%
+\begin{align*}
+\theta_j &\sim \N(\mu, \tau), \\
+y_j &\sim \N(\theta_j, \sigma_j).
+\end{align*}
+%
+The ``non-centered'' parameterisation recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
+This can be written as,
+%
+\begin{align*}
+\tilde{\theta}_j &\sim \N(0, 1), \\
+\theta_j &= \mu + \tau \tilde{\theta}_j,\\
+y_j &\sim \N(\theta_j, \sigma_j).
+\end{align*}
+%
+In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
+of the distribution of these effects.
+In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
+
+We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
+Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
+We also used the same algorithm settings to sample from the non-centered model.
+
+To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
+Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
+The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
+In this plot, the centered model is close to convergence, whereas the non-centered is not.
+
+In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two.
+The results are shown in Fig.\ref{fig:eight_schools}B.
+In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each draw, meaning that the centered model $R^*$ values are shifted leftwards.
+Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears near convergence.
+
+It is recommended that $\widehat{R}$, like $R^*$, be calculated using split chains.
+In Fig.
+\ref{fig:eight_schools}C, we plot the values of values of this metric when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
+These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
+All parameters for the non-centered models were below 1.01, indicating convergence.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
+	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
+The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
+In all cases, the plots show 1000 $R^*$ draws.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:eight_schools}
+\end{figure}
+
+\subsubsection{Understanding chain classification}
+To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
+After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
+In Fig.
+\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
+In the top row of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
+Above the top panel, we show the marginal distributions for each chain.
+In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
+In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
+For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
+Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved ML predictive accuracy.
+
+Fig.
+\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in ML classification accuracy and falsely concluding that convergence has occurred.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools_r_star_quantiles.pdf}}
+	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
+In each column, we show the path of the four individual chains above and 40 $R^*$ draws for each parameter quantile below.
+Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
+Note that, in the bottom plots, jitter has been added to the data points.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:eight_schools_r_star_quantiles}
+\end{figure}
+
+\section{Discussion}
+If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a draw, it should be impossible to discern which chain generated it.
+Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
+By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
+By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
+Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
+
+The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target.
+$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each draw.
+It also naturally includes information about the joint distribution across all dimensions of the target.
+Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects.
+Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
+This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
+
+When first starting to develop $R^*$, it was unclear to us whether \textit{any} metric aligned with a supervised ML method would be overly sensitive to its hyperparameters.
+With gradient-boosted regression tree models, we found this not to be the case.
+Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples.
+To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
+
+Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
+In a number of examples, we trial this before calculating $R^*$ and find that this approach leads to more accurate chain prediction.
+As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
+Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
+If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.
+
+MCMC enables inference across a wide range of models encountered across the social, biological and physical sciences.
+Its ease of implementation, however, masks important underlying fragilities in the method.
+Namely, that unless the chains have converged to a truly stationary distribution, the draws generated are not faithful depictions of the posterior.
+In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far.
+$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
+
+\section{Contributions}
+BL conceived of the original idea for $R^*$, carried out all analyses and wrote the original draft of the paper.
+AV reviewed the paper and made a host of recommendations: including suggesting a large range of new case studies on which to trial the method and also suggesting new visualisations for diagnosing how predictive performance depends on parameter quantiles; these collectively widened the scope of the original paper and substantially improved its quality.
+BL and AV reviewed and edited the draft of the paper.
+
+\bibliographystyle{chicago}
+\bibliography{bibliography}
+
+\beginsupplement
+
+\section{Wide datasets: multivariate normal}\label{sec:wide}
+As the number of parameter dimensions increases, it might be thought that ML algorithms will overfit the data, and, hence, testing set classification would be poor; leading to unreliable determinations of convergence.
+To test this hypothesis, we investigated two scenarios using a multivariate normal target.
+In the first of these, we used the 250-dimensional multivariate normal of eq.
+\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
+Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
+\ref{fig:mvt_wide_both}A.
+In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
+The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
+\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, some of the parameters had $\widehat{R}>1.01$.
+Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
+\ref{fig:wide_both_diagnostics}B\&C).
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_wide_both.pdf}}
+	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
+In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:mvt_wide_both}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
+	\caption{\textbf{Wide data 250-dimensional example: established diagnostics.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
+Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
+In both cases, the results correspond to a single MCMC run composed of 4 chains.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:wide_both_diagnostics}
+\end{figure}
+
+We next consider a more challenging example -- a target distribution with 10,000 dimensions.
+In this case, we assume independent standard normals for each dimension.
+In Fig.
+\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
+In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
+These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.
+
+Overall, the examples in this section suggest that $R^*$ is a conservative measure of convergence: when there are not enough draws, it will tend to diagnose non-convergence.
+We also note that the statistic took comparable time to calculate relative to existing convergence diagnostics on a desktop computer.
+
+\section{Non-stationary marginals}\label{sec:non-stationary}
 To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}.
@@
 
-\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+\section{Many parameter models: ovarian and prostate analysis}\label{sec:prostate}
+In this section, we analyse two Bayesian models both fit to real data.
+The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
+Since there are relatively few data points relative to the number of predictors, regularised horseshoe priors are specified on the regression coefficients \cite{piironen2017sparsity} since most are expected to be zero.
+This dataset has been used for benchmarking in the past (see, for example, \cite{schummer1999comparative,hernandez2010expectation,paananen2019implicitly}) and is known to result in a multimodal posterior.
 
-The model can be parameterised in two ways, as described in \cite{vehtari2019rank}.
-The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
-%
-\begin{align*}
-\theta_j &\sim \N(\mu, \tau), \\
-y_j &\sim \N(\theta_j, \sigma_j).
-\end{align*}
-%
-The ``non-centered'' parameterisation recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
-This can be written as,
-%
-\begin{align*}
-\tilde{\theta}_j &\sim \N(0, 1), \\
-\theta_j &= \mu + \tau \tilde{\theta}_j,\\
-y_j &\sim \N(\theta_j, \sigma_j).
-\end{align*}
-%
-In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
-of the distribution of these effects.
-In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
+The other dataset we use is of similar form but for prostate cancer, which is also analysed using logistic regression with regularised horseshoe priors. It has 18,105 parameters in total.
 
-We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
-We also used the same algorithm settings to sample from the non-centered model.
-
-To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
-The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
-In this plot, the centered model is close to convergence, whereas the non-centered is not.
-
-In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two.
-The results are shown in Fig.\ref{fig:eight_schools}B.
-In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each draw, meaning that the centered model $R^*$ values are shifted leftwards.
-Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears near convergence.
-
-It is recommended that $\widehat{R}$, like $R^*$, be calculated using split chains.
-In Fig.
-\ref{fig:eight_schools}C, we plot the values of values of this metric when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
-All parameters for the non-centered models were below 1.01, indicating convergence.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
-	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
-The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In all cases, the plots show 1000 $R^*$ draws.
-See the R markdown file on the Github repository for code to reproduce this example.}
-	\label{fig:eight_schools}
-\end{figure}
-
-To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
-After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
-In Fig.
-\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
-In the top row of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
-Above the top panel, we show the marginal distributions for each chain.
-In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
-In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
-For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
-Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved ML predictive accuracy.
-
-Fig.
-\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in ML classification accuracy and falsely concluding that convergence has occurred.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools_r_star_quantiles.pdf}}
-	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
-In each column, we show the path of the four individual chains above and 40 $R^*$ draws for each parameter quantile below.
-Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
-Note that, in the bottom plots, jitter has been added to the data points.
-See the R markdown file on the Github repository for code to reproduce this example.}
-	\label{fig:eight_schools_r_star_quantiles}
-\end{figure}
-
-\subsection{Many parameter models: ovarian and prostate analysis}
-In this section, we analyse two Bayesian models [Aki, can you provide some description of these?].
-These models have relatively many parameters (4719 for the ``ovarian'' model and 18,105 for the ``prostate'' model) and are known to be challenging to explore with MCMC.
 For each model, we consider two MCMC runs: one with 4 chains run with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
@@
 
-\section{Discussion}
-If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a draw, it should be impossible to discern which chain generated it.
-Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
-By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
-By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
-Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
-
-The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target.
-$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each draw.
-It also naturally includes information about the joint distribution across all dimensions of the target.
-Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects.
-Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
-This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
-
-When first starting to develop $R^*$, it was unclear to us whether \textit{any} metric aligned with a supervised ML method would be overly sensitive to its hyperparameters.
-With gradient-boosted regression tree models, we found this not to be the case.
-Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples.
-To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
-
-Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
-In a number of examples, we trial this before calculating $R^*$ and find that this approach leads to more accurate chain prediction.
-As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
-Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
-If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.
-
-MCMC enables inference across a wide range of models encountered across the social, biological and physical sciences.
-Its ease of implementation, however, masks important underlying fragilities in the method.
-Namely, that unless the chains have converged to a truly stationary distribution, the draws generated are not faithful depictions of the posterior.
-In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far.
-$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
-
-\section{Contributions}
-BL conceived of the original idea for $R^*$, carried out all analyses and wrote the original draft of the paper.
-AV reviewed the paper and made a host of recommendations: including suggesting a large range of new case studies on which to trial the method and also suggesting new visualisations for diagnosing how predictive performance depends on parameter quantiles; these collectively widened the scope of the original paper and substantially improved its quality.
-BL and AV reviewed and edited the draft of the paper.
-
-\bibliographystyle{chicago}
-\bibliography{bibliography}
-
 \end{document}

### 2020-08-02 src/s_ar1.Rmd
changes to figures style and new section on supplementary experiments

@@
 ```
+
+# Performance as a function of number of chains
+```{r}
+f_generate_lower_var_n <- function(n_chains, var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = n_chains)
+  for(i in 1:(n_chains - 1))
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, n_chains] <- z
+  return(x)
+}
+
+f_replicate_n <- function(n_chains){
+  temp <- f_generate_lower_var_n(n_chains, 1/3, 0.3, 1, 2000)
+  a_array <- array(dim=c(2000, n_chains, 1))
+  a_array[,,1] <- temp
+  a_accuracy <- r_star(a_array)
+  mon <- monitor_extra(split_data(a_array))
+  return(list(r_star=a_accuracy, r_hat=mon$zfsRhat))
+}
+
+nreplicates <- 100
+lnum_chains <- c(32)
+mResults <- matrix(nrow = nreplicates * length(lnum_chains),
+                   ncol = 4)
+k <- 1
+for(j in seq_along(lnum_chains)){
+  print(j)
+  for(i in 1:nreplicates){
+    temp <- f_replicate_n(lnum_chains[j])
+    mResults[k, ] <- c(lnum_chains[j], i, temp$r_star, temp$r_hat)
+    k <- k + 1
+  }
+  saveRDS(mResults, "../data/ar1_numchains_extra1.rds")
+}
+
+colnames(mResults) <- c("nchains", "iterate", "R*", "split-Rhat")
+mResults <- as.data.frame(mResults) %>% 
+  filter(nchains < 64)
+saveRDS(mResults, "../data/ar1_numchains.rds")
+mResults <- readRDS("../data/ar1_numchains.rds")
+b_sum <- mResults %>% 
+  melt(id.vars=c("nchains", "iterate")) %>%
+  group_by(nchains, variable) %>% 
+  summarise(lower=quantile(value, 0.25),
+         upper=quantile(value, 0.75),
+         middle=quantile(value, 0.5))
+
+g <- mResults %>% 
+  melt(id.vars=c("nchains", "iterate")) %>% 
+  left_join(b_sum, by = c("nchains", "variable")) %>% 
+  ggplot(aes(x=as.factor(nchains), y=value)) +
+  geom_jitter(width = 0.2, colour="grey") +
+  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
+  facet_wrap(~variable) +
+  scale_x_discrete() +
+  xlab("# chains") +
+  ylab("Value") +
+  scale_color_grey() +
+  theme(axis.text = element_text(colour="black", size=12),
+        axis.title = element_text(colour="black", size=14),
+        strip.text.x = element_text(colour="black", size=14),
+        legend.position = "none")
+ggsave("../output/ar1_numchains.pdf", g, width = 12, height = 8)
+```

### 2020-08-02 src/s_cauchy.Rmd
changes to figures style and new section on supplementary experiments

@@
         legend.position = c(0.8, 0.7),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black"),
         axis.title.y = element_blank()) +
@@
   scale_y_log10() +
-  ggtitle(TeX("E. Tail ESS"))
+  ggtitle(TeX("E. Tail-ESS"))
 
@@
   scale_y_log10() +
-  ggtitle(TeX("F. Bulk ESS"))
+  ggtitle(TeX("F. Bulk-ESS"))
 ```

### 2020-08-02 src/s_eight_schools.Rmd
changes to figures style and new section on supplementary experiments

@@
 library(ggExtra)
+library(gridExtra)
 f <- ecdf(x[, , 2])
@@
 pdf("../output/eight_schools_r_star_quantiles.pdf", width = 6, height = 6)
-plot_grid(arrangeGrob(gb, top="A."), arrangeGrob(ga, top="B."), g1, g2,nrow = 2, rel_heights = c(1/2, 1/4))
+plot_grid(arrangeGrob(gb, top="A. mu"), arrangeGrob(ga, top="B. tau"), g1, g2,nrow = 2, rel_heights = c(1/2, 1/4))
 dev.off()

### 2020-08-02 src/s_ovarian_prostate.Rmd
changes to figures style and new section on supplementary experiments

@@
 x <- rstan::extract(fit, permuted=F)
+
 mon <- monitor_extra(split_data(x))
@@
   bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. split-Rhat")) %>% 
-  bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk ESS")) %>% 
-  bind_rows(tibble(long=mon_longer$tailseff, short=mon_shorter$tailseff, metric="D. Tail ESS")) %>% 
+  bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk-ESS")) %>% 
+  bind_rows(tibble(long=mon_longer$tailseff, short=mon_shorter$tailseff, metric="D. Tail-ESS")) %>% 
   melt(id.vars="metric") %>% 
-  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. split-Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
+  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. split-Rhat", "C. Bulk-ESS", "D. Tail-ESS")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +

### 2020-08-02 src/s_trends.Rmd
changes to figures style and new section on supplementary experiments

@@
 
-ggplot(data = res, aes(x = rho, y = Rhat)) + 
+ggplot(data = res, aes(x = rho, y = zsRhat)) + 
   geom_jitter() + 

### 2020-08-02 src/s_wide_distributions.Rmd
changes to figures style and new section on supplementary experiments

@@
   melt() %>% 
-  mutate(variable=case_when(variable=="bulk_ess"~"Bulk ESS",
-                            variable=="tail_ess"~"Tail ESS",
+  mutate(variable=case_when(variable=="bulk_ess"~"Bulk-ESS",
+                            variable=="tail_ess"~"Tail-ESS",
                             variable=="rhat"~"Split-Rhat")) %>% 
@@
   geom_histogram() + 
-  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk ESS", "Tail ESS"), scales="free") + xlab("Value") +
+  facet_wrap(~relevel(as.factor(variable), "Split-Rhat", "Bulk-ESS", "Tail-ESS"), scales="free") + xlab("Value") +
   ylab("Count") +
@@
   melt(id.vars="source") %>% 
-  mutate(variable=case_when(variable=="bulk_ess"~"B. Bulk ESS",
-                            variable=="tail_ess"~"C. Tail ESS",
+  mutate(variable=case_when(variable=="bulk_ess"~"B. Bulk-ESS",
+                            variable=="tail_ess"~"C. Tail-ESS",
                             variable=="rhat"~"A. Split-Rhat")) %>% 

### 2020-08-02 write_up/ms.tex
changes to figures style and new section on supplementary experiments

@@
 Table \ref{tab:results} summarises the examples and provides a rationale for their inclusion.
-Alongside the examples included in the main text, there are a number of supplementary text examples: in \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of samples; in \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain is diagnosed; in \S\ref{sec:prostate}, we show that $R^*$ performs well for a Bayesian logistic regression problem with a large number of parameters and a highly multimodal posterior.
+The experiments not detailed in the main text are briefly described in \S\ref{sec:further_experiments} and more fully in the relevant sections given in Table \ref{tab:results}.
 
@@
 Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done.
-We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data); and another, where all chains sample from the same distribution (we call this the ``converged'' data).
+We demonstrate this idea using two datasets: one generated as described in \S\ref{sec:heterogeneity}, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data); and another, where all chains sample from the same distribution (we call this the ``converged'' data).
 In Fig.
@@
 Much like how it is possible for $\hat{R}<1$, this is a sample property, driven by the underlying uncertainty when sampling from the categorical distribution defined by the probability simplex.
+
+\subsubsection{Sensitivity to number of chains}\label{sec:heterogeneity_numchains}
+We have so far focused on the sensitivity of $R^*$ to chain heterogeneity with a fixed number of chains: four.
+Since classification becomes a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
+For comparison, we also illustrate how the performance of rank-normalised split-$\widehat{R}$ varies with number of chains.
+To do so, we consider a problem similar to that described in \S\ref{sec:heterogeneity}: where all chains bar one have $\sigma=1$, and the remaining chain has $\sigma=1/3$.
+We consider cases with 2, 4, 8, 16, 32 and 64 chains.
+The other hyperparameters of the data generating process remain the same as in \S\ref{sec:heterogeneity}.
+
+In Fig.
+\ref{fig:ar1_numchains}, we show the results of these simulations.
+On the horizontal axis, we show the number of chains, and on the vertical axis, the value of each of the two convergence measures ($R^*$ in the left panel; $\widehat{R}$ in the right panel).
+Overall, both measures perform similarly as chain count increases.
+For $R^*$, this is because it is harder to classify chains when there are more of them.
+For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
+Still, when there were 64 chains, both measures remained above 1 across most replicates, with median $R^*=X$ and median $\widehat{R}=Y$.
+
+The decline of both of these measures with varying numbers of chains hints that perhaps a moving threshold for diagnosing convergence may be pertinent to avoid neglecting those minority of chains with differing information.
+Here, however, we do not make suggestions on what such guidelines could be and leave this for later work.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{../output/ar1_numchains.pdf}}
+	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
+The vertical axis shows the value of $R^*$ as by Algorithm \ref{alg:R_star} (left panel), and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
+Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
+The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 20 replicates at each number of chains.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:ar1_numchains}
+\end{figure}
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
-	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
+	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
 The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
@@
 
+\subsection{Further experiments}\label{sec:further_experiments}
+Alongside the examples included in the main text, there are a number of supplementary text examples, which we briefly outline here.
+
+In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of samples.
+Specifically, this was to test that the ML classification didn't become prone to overfitting in this limit.
+To test this hypothesis, we investigated two scenarios using a multivariate normal target: one with a 250-multivariate normal with high correlation between dimensions using 250 post-warm-up iterations; and another normal with 10,000 independent dimensions using up to 500 post-warm-up iterations.
+In both cases, sampling was done using Stan's NUTS algorithm.
+In both cases, $R^*$ and rank-normalised split-$\widehat{R}$ reached similar conclusions about convergence: namely, that more iterations were needed in all experiments considered.
+Overall, these experiments show that $R^*$ is a conservative convergence measure that will tend to diagnose convergence when there are insufficient draws.
+
+In \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain convergence is diagnosed.
+We illustrate this via three examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; and (c), a bivariate normal with a non-stationary covariance.
+The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue.
+The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish across the numbers of dimensions considered.
+
+In \S\ref{sec:prostate}, we show that $R^*$ performs well for a Bayesian logistic regression problem with a large number of parameters and a highly multimodal posterior.
+
 \section{Discussion}
@@
 Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
+Note, that it is possible for the ESS to exceed the actual sample size if there is negative autocorrelation in the Markov chains' values.
 In both cases, the results correspond to a single MCMC run composed of 4 chains.
@@
 
+\subsection{10,000-dimensional example}
 We next consider a more challenging example -- a target distribution with 10,000 dimensions.
@@
 
+\subsection{Trending mean in a single dimension}
 We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends.
@@
 
+\subsection{Trending covariance}
 Means that trend over time is one form of non-stationarity; another is a time-varying covariance.

### 2020-08-03 src/s_ar1.Rmd
halfway through runthrough of draft after incorporating Aki comments

@@
 ```{r}
+reses <- readRDS("../data/ar1_r_star_hat.rds")
+r_star_vals <- reses$r_star 
+r_hat_vals <- reses$r_hat 
 g2 <- tibble(accuracy=r_star_vals, iteration=seq(1, nreplicates, 1)) %>% 
@@
   geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("un-mixed", "mixed")) +
+  scale_fill_grey("Series", labels=c("unmixed", "mixed")) +
   xlab(TeX("$R*$")) +
@@
 
-nreplicates <- 100
+nreplicates <- 50
 lnum_chains <- c(32)
@@
   }
-  saveRDS(mResults, "../data/ar1_numchains_extra1.rds")
+  # saveRDS(mResults, "../data/ar1_numchains_extra2.rds")
 }
@@
         legend.position = "none")
-ggsave("../output/ar1_numchains.pdf", g, width = 12, height = 8)
+ggsave("../output/ar1_numchains.pdf", g, width = 8, height = 4)
 ```
+
+# Investigating stochasticity of R*
+```{r}
+f_replicate_samplesize <- function(temp, samplesize){
+  a_array <- array(dim=c(samplesize, 4, 1))
+  a_array[,,1] <- temp
+  a_accuracy <- r_star(a_array)
+  return(list(r_star=a_accuracy))
+}
+
+lsample_sizes <- c(500, 1000, 2000, 4000, 8000)
+nreplicates <- 100
+mResults <- matrix(nrow = nreplicates * length(lsample_sizes),
+                   ncol = 3)
+k <- 1
+for(j in seq_along(lsample_sizes)){
+  print(j)
+  x <- f_generate_lower_var_four(1/3, 0.3, 1, lsample_sizes[j])
+  for(i in 1:nreplicates){
+    temp <- f_replicate_samplesize(x, lsample_sizes[j])
+    mResults[k, ] <- c(lsample_sizes[j], i, temp$r_star)
+    k <- k + 1
+  }
+}
+
+colnames(mResults) <- c("n", "iterate", "R*")
+mResults <- as.data.frame(mResults)
+saveRDS(mResults, "../data/ar1_numsamples.rds")
+
+b_sum <- mResults %>% 
+  melt(id.vars=c("n", "iterate")) %>%
+  group_by(n, variable) %>% 
+  summarise(lower=quantile(value, 0.025),
+         upper=quantile(value, 0.975),
+         middle=quantile(value, 0.5))
+
+g <- mResults %>% 
+  melt(id.vars=c("n", "iterate")) %>% 
+  left_join(b_sum, by = c("n", "variable")) %>% 
+  filter(variable!="split-Rhat") %>% 
+  ggplot(aes(x=as.factor(n), y=value)) +
+  geom_jitter(width = 0.2, colour="grey") +
+  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
+  scale_x_discrete() +
+  xlab("# samples") +
+  ylab(TeX("split-R*")) +
+  scale_color_grey() +
+  theme(axis.text = element_text(colour="black", size=12),
+        axis.title = element_text(colour="black", size=14),
+        strip.text.x = element_text(colour="black", size=14),
+        legend.position = "none")
+ggsave("../output/ar1_samplesize.pdf", g, width = 8, height = 6)
+```

### 2020-08-03 src/s_trends.Rmd
halfway through runthrough of draft after incorporating Aki comments

@@
 
-ggplot(data = res, aes(x = rho, y = Rstar_split)) + 
+g <- ggplot(data = res, aes(x = rho, y = Rstar_split)) + 
   geom_jitter() + 
@@
   geom_hline(yintercept = 1) + 
-  ggtitle('Split-R*') +
-  ylim(0, 6)
+  ylab(TeX("split-R*")) +
+  xlab(TeX("$\\rho$")) +
+  ylim(0, 6) +
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black"))
+ggsave("../output/trends_ar1.pdf", g, width = 8, height = 6)
+
+b_sum <- res %>%
+  group_by(iters, rho) %>% 
+  summarise(lower=quantile(Rstar_split, 0.025),
+            upper=quantile(Rstar_split, 0.975),
+            Rstar_split=median(Rstar_split)) %>% 
+  mutate(type="summary")
+res1 <- res %>%
+  bind_rows(b_sum)
+
+g <- ggplot(data = filter(res1, is.na(type)), aes(x = as.factor(iters), y = Rstar_split, group=rho)) + 
+  geom_jitter(width = 0.2, height = 0, colour="grey") + 
+  geom_pointrange(data=filter(res1, !is.na(type)), aes(ymin=lower, ymax=upper)) +
+  geom_line(data=filter(res1, !is.na(type))) +
+  facet_wrap(~rho, scales="free") + 
+  geom_hline(yintercept = 1) + 
+  ylab(TeX("split-R*")) +
+  xlab("# iterations") +
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black"))
+
+ggsave("../output/trends_ar1_transposed.pdf", g, width = 8, height = 6)
 ```

### 2020-08-03 write_up/ms.tex
halfway through runthrough of draft after incorporating Aki comments

@@
 Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
-The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters.
+The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters.
 We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing.
-Our GBM model was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:results}).
+Our GBM model was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
 Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.
@@
 		\midrule
-		Autoregressive & Heterogeneous chain variance & \ref{sec:heterogeneity} \\
-		Bivariate normal & Converged marginals but unconverged joint distribution &  \ref{sec:multivariate_normal}\\
-		250D normal & High correlations between dims & \ref{sec:multivariate_normal}\\
+		Autoregressive & Examining $R^*$ and its sensitivities to its calculation & \ref{sec:heterogeneity} \\
+		& Detecting heterogeneous chain variance using $R^*$ & \ref{sec:heterogeneity_performance} \\
+		& Stochasticity in $R^*$ & \ref{sec:heterogeneity_stochasticity}\\
+		& Generating $R^*$ uncertainty measure & \ref{sec:heterogeneity_uncertainty}\\
+		& Sensitivity of $R^*$ to number of chains & \ref{sec:heterogeneity_numchains}\\
+		& Robustness of $R^*$ to ML model hyperparameters & \ref{sec:heterogeneity_robustness}\\
+		\midrule
+		Multivariate normals & Convergence in joint distribution &  \ref{sec:multivariate_normal}\\
+		& Unconverged joint distribution in bivariate normal & \ref{sec:multivariate_normal_bivariate}\\
+		& High correlations between dims in 250D normal  & \ref{sec:multivariate_normal_250}\\
+		& Measuring contributions of variables to poor convergence & \ref{sec:multivariate_normal_varimportance}\\
+		\midrule
+		Cauchy & Long-tailed distribution & \ref{sec:cauchy}\\
+		& Comparing $R^*$ and existing measures to objective convergence & \ref{sec:cauchy_objective}\\
+		\midrule
+		Eight schools model & Hierarchical Bayesian model slow convergence & \ref{sec:eight_shools}\\
+		\midrule 
 		Wide multivariate normal & \# samples $\sim$ \# dims & \ref{sec:wide}\\
-		Cauchy & Long-tailed distribution & \ref{sec:cauchy}\\
+		\midrule
 		Non-stationary marginals & Time-varying sampling distribution & \ref{sec:non-stationary}\\
-		Eight schools model & Hierarchical Bayesian model slow convergence & \ref{sec:eight_shools}\\
-		Ovarian and prostate models & Bayesian models with many parameters & \ref{sec:prostate}
+		& Trends in mean across all chains & \ref{sec:non-stationary_chains}\\
+		& Trends in mean in a single dimension & \ref{sec:non-stationary_single}\\
+		& Trends in covariance & \ref{sec:non-stationary_covariance}\\
+		& Sensitivity of $R^*$ to chain persistence & \ref{sec:non-stationary_persistence}\\
+		\midrule
+		Ovarian and prostate models & Bayesian models with many parameters and multimodal posteriors & \ref{sec:prostate}
 	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
@@
 \subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
-We generate four Markov chains, where each samples from an autoregressive order 1 (AR(1)) process of the form,
+In this section, we show how $R^*$ is able to detect heterogeneous variance across Markov chains.
+Apart from demonstrating how $R^*$ performs, this section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}); and, finally, to the ML model hyperparameters (\S\ref{sec:heterogeneity_robustness}).
+The experiments we use to study these issues are all of similar form to the following data generating process: four Markov chains are generated, where each samples from an autoregressive order 1 (AR(1)) process of the form,
 %
-\begin{equation}
+\begin{equation}\label{eq:ar1}
 X_t = \rho X_{t-1} + \epsilon_t
@@
 Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others.
-In Fig.
-\ref{fig:ar1}A, we show how a GBM fitted to these data classifies observations according to a draw's value.
-Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
 
+\subsubsection{Performance of $R^*$}\label{sec:heterogeneity_performance}
 To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e.
@@
 For each replicate, we then calculated $R^*$ as described in Algorithm \ref{alg:R_star}.
+
+In Fig.
+\ref{fig:ar1}A, we show how a GBM fitted to one such replicate dataset classifies observations according to a draw's value.
+Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
+
 In Fig.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows 1000 $R^*$ draws as generated by Algorithm \ref{alg:R_star_uncertainty} across 1000 replicates; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows the $R^*$ distributions for two series, the first being the same as used for figures A-C; the second where all chains have the same distribution as described in \S\ref{sec:heterogeneity_uncertainty}.
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the first dataset being the same as used for figures A-C; the second where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
+Note that, in D, only a single series of each series type is used to generate distribution.
 See the R markdown file on the Github repository for code to reproduce this example.}
 	\label{fig:ar1}
+\end{figure}
+
+\subsubsection{Stochasticity in $R^*$}\label{sec:heterogeneity_stochasticity}
+Unlike $\widehat{R}$, $R^*$ is a stochastic convergence measure due to randomness in creating training and testing sets (essentially a form of sampling variation) and randomness in the methods used to train the ML model.
+This means that even if the same sample is used, $R^*$ will return a different value each time it is calculated if the pseudorandom seed is not fixed.
+To probe the extent of this randomness, we generated data using the same process as in \S\ref{sec:heterogeneity} but now using varying sample sizes, including samples consisting of 500, 1000, 2000, 4000 and 8000 draws.
+For each dataset, we computed $R^*$ on it 100 times, allowing the pseudorandom seed to vary between calculations.
+We stress that, for each sample size, we used the same dataset (so there were 5 datasets created in total -- one for each sample size), so stochasticity comes from $R^*$ calculation, not that from the data generating process.
+
+In Fig.
+\ref{fig:ar1_samplesize}, we show the results of this study.
+In this figure, the horizontal axis shows the sample size, and the vertical axis, the value of $R^*$ in each repetition.
+This shows that as the number of samples increased, variation in $R^*$ declined.
+At a sample size of 500, there were four cases where $R^*<1$; in larger samples, there were none.
+Intuitively, the reduction in sampling variation when composing training and test sets from larger samples results in lower variability in ML model predictions.
+We also expect that larger sample sizes should lead to higher $R^*$ values since more training data leads to better ML models.
+We may start to see this here, since the median $R^*=1.25$ at a sample size of 8000 was greater than for the smaller samples.
+
+If randomness in $R^*$ calculation leads to different conclusions about convergence being drawn, this would be problematic.
+One potential remedy for this is to repeatedly calculate $R^*$ on a given sample, much as we have done here, and consider the distribution of $R^*$ values computed.
+The computational cost of doing this may, of course, be unreasonable.
+Instead, in \S\ref{sec:heterogeneity_uncertainty}, we consider an alternative approach based on bootstrapping a single ML model's predictions.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.6\textwidth]{../output/ar1_samplesize.pdf}}
+	\caption{\textbf{Autoregressive example: $R^*$ stochasticity.} The horizontal axis shows sample size; the vertical axis shows the value of $R^*$ calculated on chains split into two halves as per Algorithm \ref{alg:R_star}.
+Grey points show the value of $R^*$ for each replicate (jitter was added to point positions).
+Black points show the median $R^*$ value; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
+For each sample size, a single dataset was created and used for all $R^*$ calculations.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:ar1_samplesize}
 \end{figure}
@@
 Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done.
-We demonstrate this idea using two datasets: one generated as described in \S\ref{sec:heterogeneity}, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data); and another, where all chains sample from the same distribution (we call this the ``converged'' data).
+We demonstrate this idea using two datasets: one generated as described in \S\ref{sec:heterogeneity}, where one chain (out of four) has a lower variance than the others (we call this the ``unmixed'' data); and another, where all chains sample from the same distribution (we call this the ``mixed'' data).
 In Fig.
 \ref{fig:ar1}D, we show the $R^*$ distributions in each case.
-For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
-For the converged data, the distribution is centred on 1, indicating convergence.
-In the converged case, there are many draws where $R^*<1$: these indicate that, in that particular draw from the probability simplex, chain classification is actually worse than selecting a chain identification uniformly at random.
-Much like how it is possible for $\hat{R}<1$, this is a sample property, driven by the underlying uncertainty when sampling from the categorical distribution defined by the probability simplex.
+For the unmixed data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
+For the mixed data, the distribution is centred on 1, indicating convergence.
+In the mixed case, there are many draws where $R^*<1$: these indicate that, in that particular draw from the probability simplex, chain classification is actually worse than selecting a chain identification uniformly at random.
+Much like how it is possible for $\hat{R}<1$, this is a sample property, driven by the sampling distribution of the categorical distribution defined by the probability simplex.
+
+It is worth emphasising that the uncertainty distribution obtained by Algorithm \ref{alg:R_star_uncertainty} differs from that obtained from repeatedly calculating $R^*$ via Algorithm \ref{alg:R_star} as was done in \S\ref{sec:heterogeneity_stochasticity}.
+In Algorithm \ref{alg:R_star_uncertainty}, variation in $R^*$ comes from sampling from the probability simplex: if predicted chain probabilities are close to uniform, there will be greater uncertainty in $R^*$.
+Repeatedly calculating $R^*$ by applying Algorithm \ref{alg:R_star} to the same dataset yields a distribution whose width derives from sampling variation when forming training and testing sets and the stochasticity in training ML models.
+Collectively, these differences mean that the two measures of uncertainty will differ.
+
+There is an additional difference, though, in the central points of each distribution: the distribution obtained by Algorithm \ref{alg:R_star_uncertainty} will, in general, have a lower mean than that obtained by repeated application of Algorithm \ref{alg:R_star}.
+To see this, note that the darker-shaded $R^*$ distribution in Figure \ref{fig:ar1}D was generated via Algorithm \ref{alg:R_star_uncertainty} and has a mean around 1.07; the distribution shown in Figure \ref{fig:ar1}B was generated by repeatedly recomputing $R^*$ using Algorithm \ref{alg:R_star} and has a mean closer to 1.22.
+This difference in mean is expected since predictive performance when assigning chain identities stochastically when sampling from the categorical distribution of the probability simplex (as is done in Algorithm \ref{alg:R_star_uncertainty}) will generally result in worse prediction than when assigning each chain identity using whichever chain has the highest class probability (as is done in Algorithm \ref{alg:R_star}).
+Of course, we would prefer it if the uncertainty distribution generated by Algorithm \ref{alg:R_star_uncertainty} had a mean closer to the one obtained by repeated application of Algorithm \ref{alg:R_star}.
+Nonetheless, in practice, we have found that the mean of the $R^*$ distribution generated by Algorithm \ref{alg:R_star_uncertainty} provides a useful cheaper diagnostic.
 
@@
 For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
-Still, when there were 64 chains, both measures remained above 1 across most replicates, with median $R^*=X$ and median $\widehat{R}=Y$.
+Still, when there were 32 chains, both measures remained above 1 across most replicates, with median $R^*=X$ and median $\widehat{R}=Y$.
 
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{../output/ar1_numchains.pdf}}
+	\centerline{\includegraphics[width=0.8\textwidth]{../output/ar1_numchains.pdf}}
 	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
-The vertical axis shows the value of $R^*$ as by Algorithm \ref{alg:R_star} (left panel), and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
+The vertical axis shows the value of $R^*$ as by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
 Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
@@
 \subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
-To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density.
+In this section, we illustrate how $R^*$ can diagnose convergence issues in the joint target distribution.
+
+\subsubsection{Bivariate model}\label{sec:multivariate_normal_bivariate}
+First, we consider a bivariate normal density.
 In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
@@
 
-\subsubsection{250-dimensional model}
+\subsubsection{250-dimensional model}\label{sec:multivariate_normal_250}
 We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}.
@@
 
-\subsubsection{Variable importance}
+\subsubsection{Variable importance}\label{sec:multivariate_normal_varimportance}
 In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions.
@@
 
-\subsubsection{Measuring convergence objectively}
+\subsubsection{Measuring convergence objectively}\label{sec:cauchy_objective}
 To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target.
@@
 In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of samples.
-Specifically, this was to test that the ML classification didn't become prone to overfitting in this limit.
+Specifically, this was to test that ML classification didn't become prone to overfitting in this limit.
 To test this hypothesis, we investigated two scenarios using a multivariate normal target: one with a 250-multivariate normal with high correlation between dimensions using 250 post-warm-up iterations; and another normal with 10,000 independent dimensions using up to 500 post-warm-up iterations.
@@
 In both cases, $R^*$ and rank-normalised split-$\widehat{R}$ reached similar conclusions about convergence: namely, that more iterations were needed in all experiments considered.
-Overall, these experiments show that $R^*$ is a conservative convergence measure that will tend to diagnose convergence when there are insufficient draws.
+Overall, these experiments show that $R^*$ is a conservative convergence measure that will tend to diagnose unconvergence when there are insufficient draws.
 
@@
 The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue.
-The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish across the numbers of dimensions considered.
+The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish as the numbers of dimensions considered increased.
+Example (c) showed that split-$R^*$ opposed to split-$\widehat{R}$ is able to diagnose non-stationary covariance between dimensions of a target distribution.
 
-In \S\ref{sec:prostate}, we show that $R^*$ performs well for a Bayesian logistic regression problem with a large number of parameters and a highly multimodal posterior.
+In \S\ref{sec:prostate}, we show that $R^*$ performs well for two Bayesian logistic regression problems, each with a large number of parameters and highly multimodal posteriors.
 
@@
 To test this hypothesis, we investigated two scenarios using a multivariate normal target.
+
+\subsection{250-dimensional model}
 In the first of these, we used the 250-dimensional multivariate normal of eq.
@@
 
-\subsection{10,000-dimensional example}
+\subsection{10,000-dimensional model}
 We next consider a more challenging example -- a target distribution with 10,000 dimensions.
@@
 \section{Non-stationary marginals}\label{sec:non-stationary}
-To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}.
+If a Markov chain does not mix with itself, this also indicates that convergence has not occurred \cite{gelman2013bayesian}.
+In this section, we investigate whether $R^*$ can detect non-stationary sampling distributions.
+
+\subsection{Trending mean across all chains}\label{sec:non-stationary_chains}
+We first recapitulate an example from appendix A in \cite{vehtari2019rank}.
 This example showed that split-$\widehat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
@@
 
-\subsection{Trending mean in a single dimension}
+\subsection{Trending mean in a single dimension}\label{sec:non-stationary_single}
 We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends.
@@
 
-\subsection{Trending covariance}
+\subsection{Trending covariance}\label{sec:non-stationary_covariance}
 Means that trend over time is one form of non-stationarity; another is a time-varying covariance.
@@
 	\label{fig:trends_joint_distribution}
+\end{figure}
+
+\subsection{Chain persistence}\label{sec:non-stationary_persistence}
+When forming training and testing sets as part of the ML algorithm used to determine $R^*$, the testing set is effectively treated as a sort of ``independent'' hold-out dataset.
+Markov chains, in general, have persistence, meaning that the test set will not be truly independent and can -- according to the level of autocorrelation in the chains -- be highly related to the training set.
+In this section, we investigate how this autocorrelation affects the performance of $R^*$.
+The difficulty with this question is that higher chain autocorrelation typically means the sampling distribution is a rougher approximation of the target, so, $R^*$ should be higher due to the properties of the sampling distribution.
+It could also be higher because the training set is less distinct from the testing set.
+
+To do so, we generated AR(1) processes (as defined in eq.
+\ref{eq:ar1}) with autocorrelations, $\rho$, ranging from 0.8-1 and, in each case, calculated $R^*$ via Algorithm \ref{alg:R_star}.
+Note, that only when $|\rho|<1$ is the marginal distribution defined by this process itself stationary; at $\rho=1$, its variance increases linearly with time, so, by definition, is not converged.
+In Fig.
+\ref{fig:trends_ar1}, we show the results of these simulations across various numbers of iterations: 250, 1000 and 4000 (different panels).
+In all cases, $R^*>1$ whenever $\rho=1$, indicating lack of convergence.
+As $\rho$ declined, so did $R^*$.
+
+Notably, as the number of iterations increased, the values of $R^*$ for $\rho<1$ declined, whereas $R^*$ for $\rho=1$ actually increased: this is most easily seen by examining Fig.
+\ref{fig:trends_ar1_transposed} (which shows the same data as in Fig.
+\ref{fig:trends_ar1} but in a different way).
+This characteristic is exactly as desired: larger sample sizes should yield a sampling distribution closer to convergence for $\rho<1$; the same for $\rho$ produces distributions that are no closer to convergence, and more samples allows better determination of this non-convergence.
+
+Overall, it seems that $R^*$ is a conservative measure and with greater chain autocorrelation it suggests more draws are necessary for convergence.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_ar1.pdf}}
+	\caption{\textbf{Non-stationary distribution: AR(1) example} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
+At each parameter set, 10 replicates were run and jitter was added to the points.
+The black line shows $R^*=1$.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:trends_ar1}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_ar1_transposed.pdf}}
+	\caption{\textbf{Non-stationary distribution: AR(1) example} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
+At each parameter set, 10 replicates were run and jitter was added to the points.
+The black dots show the median $R^*$ values at each parameter set; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
+The black horizontal line shows $R^*=1$.
+See the R markdown file on the Github repository for code to reproduce this example.}
+	\label{fig:trends_ar1_transposed}
 \end{figure}

### 2020-08-04 src/s_ar1.Rmd
finished run through draft incorporating Aki's comments

@@
 nreplicates <- 50
-lnum_chains <- c(32)
+lnum_chains <- c(2, 4, 6, 8, 16, 32)
 mResults <- matrix(nrow = nreplicates * length(lnum_chains),
@@
   }
-  # saveRDS(mResults, "../data/ar1_numchains_extra2.rds")
 }
@@
 saveRDS(mResults, "../data/ar1_numchains.rds")
-mResults <- readRDS("../data/ar1_numchains.rds")
+mResults <- readRDS("../data/ar1_numchains.rds") %>% 
+  filter(iterate<=50)
 b_sum <- mResults %>% 
@@
 ggsave("../output/ar1_numchains.pdf", g, width = 8, height = 4)
+
 ```

### 2020-08-04 write_up/ms.tex
finished run through draft incorporating Aki's comments

@@
 In Fig.
-\ref{fig:ar1_numchains}, we show the results of these simulations.
+\ref{fig:ar1_numchains}, we show the results of these simulations with 50 replicates at each number of chains.
 On the horizontal axis, we show the number of chains, and on the vertical axis, the value of each of the two convergence measures ($R^*$ in the left panel; $\widehat{R}$ in the right panel).
@@
 For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
-Still, when there were 32 chains, both measures remained above 1 across most replicates, with median $R^*=X$ and median $\widehat{R}=Y$.
+Across the replicates we ran, median $\widehat{R}<1.01$ for 16 or more chains; a minimum of $R^*=1.10$ was obtained for 32 chains.
 
@@
 	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
-The vertical axis shows the value of $R^*$ as by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
+The vertical axis shows the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
 Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 20 replicates at each number of chains.
+The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 50 replicates at each number of chains.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
-The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
+The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
 The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.
@@
 In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
-By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.
+By construction, all chains target the same marginal distribution in each dimension, but the fourth chain has a different joint distribution.
 
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail-ESS.
+First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures that aim to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail-ESS.
 In all cases, the various quantities were calculated based on chains split into halves.
@@
 We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
+We use Stan's NUTS algorithm to sample from this target distribution and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
 First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
@@
 \ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-From this plot, there was a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$).
+In this plot, there is a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$).
 In Fig.
@@
 In A, 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} are shown for each MCMC run.
-In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.
+In plots C and D, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.
 See the R markdown file on the Github repository for code to reproduce this example.}
@@
 As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
-The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$, yet, still the mean is above this value.
+The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet, but its mean is above this value.
 In the bottom-left panel of Fig.
@@
 This shows that after c.10,000 iterations, the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
-Next, in Fig.
-\ref{fig:cauchy_convergence}B and Fig.
-\ref{fig:cauchy_convergence}C, we plot two measures of $\widehat{R}$, each calculated from splitting the 4 original chains into two equal halves.
+Next, in Figs.
+\ref{fig:cauchy_convergence}B\&C, we plot two measures of $\widehat{R}$, each calculated from splitting the 4 original chains into two equal halves.
 The first of these measures is the rank-normalised $\widehat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig.
@@
 \ref{fig:cauchy_convergence}C shows how this metric changes with sample size for both Cauchy model parameterisations.
-After c.
-1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more draws are needed.
+After c.1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more draws are needed.
 In Fig.
@@
 	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
-Dashed lines indicate recommended thresholds for each convergence statistic.
-See the R markdown file on the Github repository for code to reproduce this example.}
+Dashed lines indicate recommended thresholds for each convergence statistic.See the R markdown file on the Github repository for code to reproduce this example.}
 	\label{fig:cauchy_convergence}
@@
 \ref{fig:eight_schools}C, we plot the values of values of this metric when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all parameters, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
+These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all but a single parameter, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
 All parameters for the non-centered models were below 1.01, indicating convergence.
@@
 In \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain convergence is diagnosed.
-We illustrate this via three examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; and (c), a bivariate normal with a non-stationary covariance.
+We illustrate this via four examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; (c), a bivariate normal with a non-stationary covariance; and (d), an autocorrelated sampling distribution with a univariate normal target with a range of different autocorrelations.
 The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue.
@@
 Example (c) showed that split-$R^*$ opposed to split-$\widehat{R}$ is able to diagnose non-stationary covariance between dimensions of a target distribution.
+In (d), we show that $R^*$ is able to differentiate between distributions with non-stationary target distributions and stationary ones.
+It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there insufficient draws.
 
-In \S\ref{sec:prostate}, we show that $R^*$ performs well for two Bayesian logistic regression problems, each with a large number of parameters and highly multimodal posteriors.
+In \S\ref{sec:prostate}, we show that $R^*$ performs well for two Bayesian logistic regression problems with highly multimodal posteriors.
+Each of these models have 1000s of parameters, and we found that it was slow to compute both $\widehat{R}$ and $R^*$ for them.
+That said, the computational time for calculating $\widehat{R}$ was considerably less than was needed for $R^*$.
 
@@
 
+In \S\ref{sec:prostate}, we fit Bayesian models with many 1000s of parameters then used $R^*$ to diagnose convergence, finding that $R^*$ was considerably more expensive to calculate than $\widehat{R}$.
+The computational cost of training GBMs has been suggested to be $\mathcal{O}(N k)$, where $N$ is the number of training data points and $k$ is the number of dimensions of the target \citep{complexity2018}.
+If so, this suggests that larger models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
+
 Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
 In a number of examples, we trial this before calculating $R^*$ and find that this approach leads to more accurate chain prediction.
-As such, we recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
+We recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
 Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
@@
 Namely, that unless the chains have converged to a truly stationary distribution, the draws generated are not faithful depictions of the posterior.
-In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that we argue has received insufficient attention thus far.
+In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that has received insufficient attention thus far.
 $R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
@@
 In this section, we investigate how this autocorrelation affects the performance of $R^*$.
-The difficulty with this question is that higher chain autocorrelation typically means the sampling distribution is a rougher approximation of the target, so, $R^*$ should be higher due to the properties of the sampling distribution.
+The difficulty with this question is that higher chain autocorrelation typically means the sampling distribution is a rougher approximation of the target, so $R^*$ should be higher due to the properties of the sampling distribution.
 It could also be higher because the training set is less distinct from the testing set.
@@
 To do so, we generated AR(1) processes (as defined in eq.
-\ref{eq:ar1}) with autocorrelations, $\rho$, ranging from 0.8-1 and, in each case, calculated $R^*$ via Algorithm \ref{alg:R_star}.
+(\ref{eq:ar1})) with autocorrelations, $\rho$, ranging from 0.8-1 and, in each case, calculated $R^*$ via Algorithm \ref{alg:R_star}.
 Note, that only when $|\rho|<1$ is the marginal distribution defined by this process itself stationary; at $\rho=1$, its variance increases linearly with time, so, by definition, is not converged.
@@
 \ref{fig:trends_ar1} but in a different way).
-This characteristic is exactly as desired: larger sample sizes should yield a sampling distribution closer to convergence for $\rho<1$; the same for $\rho$ produces distributions that are no closer to convergence, and more samples allows better determination of this non-convergence.
+This characteristic is exactly as desired: larger sample sizes should yield a sampling distribution closer to convergence for $\rho<1$; the same for $\rho=1$ produces distributions that are no closer to convergence, and more samples allows better determination of this non-convergence.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_ar1.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
+	\caption{\textbf{Non-stationary distribution: AR(1) example.} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
 At each parameter set, 10 replicates were run and jitter was added to the points.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_ar1_transposed.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
+	\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
 At each parameter set, 10 replicates were run and jitter was added to the points.
@@
 The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
-Since there are relatively few data points relative to the number of predictors, regularised horseshoe priors are specified on the regression coefficients \cite{piironen2017sparsity} since most are expected to be zero.
+Since there are relatively few data points relative to the number of predictors, regularised horseshoe priors are specified on the regression coefficients \citep{piironen2017sparsity} since most are expected to be zero.
 This dataset has been used for benchmarking in the past (see, for example, \cite{schummer1999comparative,hernandez2010expectation,paananen2019implicitly}) and is known to result in a multimodal posterior.

### 2020-08-27 src/s_ar1.Rmd
cleaned up graph and addressed Aki's comments

@@
   ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black")) +
+  theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black")) +
   xlim(0.9, NA)
@@
   geom_histogram() +
-  xlab(TeX("split-$\\hat{R}$")) +
+  xlab(TeX("Split-$\\hat{R}$")) +
   ylab("Count") +
   ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black")) +
+  theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black")) +
   xlim(1, NA)
@@
   ylab("Chain classification") +
-  theme(text = element_text(size=14, colour="black")) +
+  theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black")) +
   coord_flip() +
@@
   geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Series", labels=c("unmixed", "mixed")) +
+  scale_fill_grey("Series", labels=c("Unmixed", "Mixed")) +
   xlab(TeX("$R*$")) +
@@
   theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = c(0.8, 0.8))
@@
   left_join(b_sum, by = c("nchains", "variable")) %>% 
+  mutate(variable=if_else(variable=="split-Rhat", "Split-Rhat", as.character(variable))) %>% 
   ggplot(aes(x=as.factor(nchains), y=value)) +
@@
 saveRDS(mResults, "../data/ar1_numsamples.rds")
+mResults <- readRDS("../data/ar1_numsamples.rds")
 
@@
   scale_x_discrete() +
-  xlab("# samples") +
-  ylab(TeX("split-R*")) +
+  xlab("Sample size") +
+  ylab(TeX("Split-R*")) +
   scale_color_grey() +

### 2020-08-27 src/s_bivariate_normal.Rmd
cleaned up graph and addressed Aki's comments

@@
 # calling function
-n <- 1000
+n <- 2000
 chain1 <- rmvrnorm2D(n, 0, 0, 1, 1, 0.0) %>% 
@@
   ylab("Count") +
-  theme(text=element_text(size=14, colour="black"))
-# ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
+  theme(text=element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black"))
+ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
 ```
@@
   temp <- chains_stacked_second %>% 
-    filter(iter==(i + 500))
+    filter(iter==(i + 1000))
   for(j in 5:8){

### 2020-08-27 src/s_cauchy.Rmd
cleaned up graph and addressed Aki's comments

@@
   melt(id.vars=c("source", "iter")) %>% 
+  mutate(source=if_else(source=="split-Rhat", "Split-Rhat", as.character(source))) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
   geom_histogram(bins=50, position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  scale_fill_grey("Parameterisation", labels=c("Nominal", "Alternative")) +
   ylab("Count") +
@@
   scale_x_log10() +
-  ggtitle(TeX("A. Quantile $R^2$"))
+  ggtitle(TeX("A. Quantile-$R^2$"))
 g1
@@
   ggtitle(TeX("C. Multivariate split-$\\hat{R}$"))
+g4
 ```

### 2020-08-27 src/s_eight_schools.Rmd
cleaned up graph and addressed Aki's comments

@@
   theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = "none")
@@
   geom_abline(linetype=3) +
-  scale_color_grey("Series", labels=c("centered", "non-centered")) +
+  scale_color_grey("Series", labels=c("Centered", "Non-centered")) +
   ylab("Split chains") +
   xlab("Original chains") +
-  ggtitle("C.")
+  ggtitle("C.") +
+  theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black", size=10))
 

### 2020-08-27 src/s_mvt_normal.Rmd
cleaned up graph and addressed Aki's comments

--- 2020-05-04 src/s_mvt_normal.Rmd
@@
   scale_shape("LP param", labels=c("False", "True")) +
-  theme(legend.position = c(0.85, 0.7))
+  theme(legend.position = c(0.85, 0.7),
+        text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black"))
 
@@
   xlab("Variable importance") +
-  ylab(TeX("split-$\\hat{R}$")) +
+  ylab(TeX("Split-$\\hat{R}$")) +
   ggtitle("C.") +
-  theme(legend.position = "none")
+  theme(legend.position = "none",
+        text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black"))
 b_df <- tibble(samples_10000_cp=a_sum$zfsRhat[1:250])
@@
   melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
+  geom_histogram(position="identity", alpha=0.8, bins=50) +
+  scale_fill_grey("# draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
   xlab(TeX("$R*$")) +
@@
   theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8)) +
+        legend.position = c(0.8, 0.8),
+        axis.text = element_text(colour="black")) +
   ggtitle("A.")
@@
   melt() %>% 
-  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
-  xlab(TeX("split-$\\hat{R}$")) +
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
+  geom_histogram(position="identity", alpha=0.8, bins=60) +
+  scale_fill_grey("# draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
+  xlab(TeX("Split-$\\hat{R}$")) +
   ylab("Count") +
@@
   theme(text = element_text(size=14, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = c(0.8, 0.8)) +

### 2020-08-27 src/s_ovarian_prostate.Rmd
cleaned up graph and addressed Aki's comments

@@
 g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
-  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. split-Rhat")) %>% 
+  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. Split-Rhat")) %>% 
   bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk-ESS")) %>% 
@@
   melt(id.vars="metric") %>% 
-  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. split-Rhat", "C. Bulk-ESS", "D. Tail-ESS")) %>% 
+  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. Split-Rhat", "C. Bulk-ESS", "D. Tail-ESS")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   facet_wrap(~metric, scales="free") +
-  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
-  theme(strip.text.x = element_text(size = 10, colour="black"),
-        text = element_text(size=12, colour="black"),
-        axis.text = element_text(size=10, colour="black"),
-        legend.text = element_text(size=12, colour="black"),
+  scale_fill_grey("MCMC detail", labels=c("4 chains: long", "16 chains: short")) +
+  theme(strip.text.x = element_text(size = 14, colour="black"),
+        text = element_text(size=14, colour="black"),
+        axis.text = element_text(size=14, colour="black"),
+        legend.text = element_text(size=14, colour="black"),
         axis.title.x = element_blank()) +

### 2020-08-27 src/s_trends.Rmd
cleaned up graph and addressed Aki's comments

--- 2020-08-02 src/s_trends.Rmd
@@
   mutate(variable=case_when(variable=="Rstar"~"A. R*",
-                            variable=="Rstar_split"~"B. split-R*")) %>% 
+                            variable=="Rstar_split"~"B. Split-R*")) %>% 
   ggplot(aes(y = value, x = trend)) + 
@@
   melt(c("rho", "iters"))  %>% 
-  mutate(variable=case_when(variable=="zsRhat"~"A. split-Rhat",
-                            variable=="Rstar_split"~"B. split-R*")) %>% 
+  mutate(variable=case_when(variable=="zsRhat"~"A. Split-Rhat",
+                            variable=="Rstar_split"~"B. Split-R*")) %>% 
   ggplot(aes(y = value, x = rho)) + 
@@
   geom_hline(yintercept = 1) + 
-  ggtitle('Split-R*') +
-  ylim(0, 6)
+  ylab(TeX("Split-R*")) +
+  xlab(TeX("$\\rho$")) +
+  ylim(0, 6) +
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black"))
+ggsave("../output/trends_ar1.pdf", g, width = 8, height = 6)
+
+b_sum <- res %>%
+  group_by(iters, rho) %>% 
+  summarise(lower=quantile(Rstar_split, 0.025),
+            upper=quantile(Rstar_split, 0.975),
+            Rstar_split=median(Rstar_split)) %>% 
+  mutate(type="summary")
+res1 <- res %>%
+  bind_rows(b_sum)
+
+g <- ggplot(data = filter(res1, is.na(type)), aes(x = as.factor(iters), y = Rstar_split, group=rho)) + 
+  geom_jitter(width = 0.2, height = 0, colour="grey") + 
+  geom_pointrange(data=filter(res1, !is.na(type)), aes(ymin=lower, ymax=upper)) +
+  geom_line(data=filter(res1, !is.na(type))) +
+  facet_wrap(~rho, scales="free") + 
+  geom_hline(yintercept = 1) + 
+  ylab(TeX("Split-R*")) +
+  xlab("# iterations") +
+  theme(strip.text.x = element_text(size=12, colour="black"),
+        strip.text.y = element_text(size=12, colour="black"),
+        axis.text.x = element_text(size=10, colour="black"),
+        axis.text.y = element_text(size=10, colour="black"))
+
+ggsave("../output/trends_ar1_transposed.pdf", g, width = 8, height = 6)
 ```

### 2020-08-27 src/s_wide_distributions.Rmd
cleaned up graph and addressed Aki's comments

@@
   geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("Model", labels=c("centered", "non-centered")) +
+  scale_fill_grey("Model", labels=c("Centered", "Non-centered")) +
   xlab(TeX("$R*$")) +

### 2020-08-27 write_up/ms.tex
cleaned up graph and addressed Aki's comments

@@
 \usepackage{bbm}
+\usepackage{caption} 
+\captionsetup[table]{skip=10pt}
 
@@
 By evaluating its performance on a separate test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{gelman2013bayesian}.
+By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
@@
 We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing.
-Our GBM model was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
+Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
 Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.
@@
 		\midrule
-		Ovarian and prostate models & Bayesian models with many parameters and multimodal posteriors & \ref{sec:prostate}
+		Ovarian and prostate models & Bayesian models with many parameters and multimodal posteriors & \ref{sec:prostate}\\
 	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
@@
 	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the first dataset being the same as used for figures A-C; the second where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
-Note that, in D, only a single series of each series type is used to generate distribution.
-See the R markdown file on the Github repository for code to reproduce this example.}
+Note that, in D, only a single series of each series type is used to generate distribution.}
 	\label{fig:ar1}
@@
 Black points show the median $R^*$ value; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
-For each sample size, a single dataset was created and used for all $R^*$ calculations.
-See the R markdown file on the Github repository for code to reproduce this example.}
+For each sample size, a single dataset was created and used for all $R^*$ calculations.}
 	\label{fig:ar1_samplesize}
@@
 Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 50 replicates at each number of chains.
-See the R markdown file on the Github repository for code to reproduce this example.}
+The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 50 replicates at each number of chains.}
 	\label{fig:ar1_numchains}
@@
 The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.
-See the R markdown file on the Github repository for code to reproduce this example.}
+The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.}
 	\label{fig:ar1_sensitivity}
@@
 First, we consider a bivariate normal density.
-In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
+In all four chains, we use independent sampling to generate 2000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
 By construction, all chains target the same marginal distribution in each dimension, but the fourth chain has a different joint distribution.
@@
 In all cases, the various quantities were calculated based on chains split into halves.
-For both dimensions, the two ESS measures were above 3700, and $\widehat{R}<1.001$, indicating no issues with convergence.
+For both dimensions, the two ESS measures were above 7000, and $\widehat{R}<1.001$, indicating no issues with convergence.
 
@@
 \ref{fig:bivariate}.
-The mean of this distribution is 1.18, and all $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
+The mean of this distribution is 1.14, and $>$99\% of $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
 By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
@@
 	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
-	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws. See the R markdown file on the Github repository for code to reproduce this example.}
+	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws.}
 	\label{fig:bivariate}
@@
 In A, 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} are shown for each MCMC run.
-In plots C and D, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.
-See the R markdown file on the Github repository for code to reproduce this example.}
+In plots C and D, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.}
 	\label{fig:mvt}
@@
 \ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs.
-The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}>1.01$ for all parameters.
+The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}<1.01$ for all parameters.
 
@@
 Rows show the $R^*$ distributions (top) and rank-normalised split-$\widehat{R}$ values across all parameters (bottom).
-Shadings indicate different model paramerisations as indicated in legend.
-See the R markdown file on the Github repository for code to reproduce this example.}
+Shadings indicate different model paramerisations as indicated in legend.}
 	\label{fig:cauchy}
@@
 In Fig.
-\ref{fig:cauchy_convergence}A, we plot this ``quantile $R^2$'' as a function of MCMC sample size for both parameterisations of the Cauchy model.
+\ref{fig:cauchy_convergence}A, we plot this \textit{quantile-$R^2$} as a function of MCMC sample size for both parameterisations of the Cauchy model.
 This shows that after c.10,000 iterations, the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
@@
 	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
-Dashed lines indicate recommended thresholds for each convergence statistic.See the R markdown file on the Github repository for code to reproduce this example.}
+Dashed lines indicate recommended thresholds for each convergence statistic.}
 	\label{fig:cauchy_convergence}
@@
 We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{hoffman2014no}, meaning that the resultant sampling distribution is likely to be biased.
+Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{betancourt2017conceptual}, meaning that the resultant sampling distribution is likely to be biased.
 We also used the same algorithm settings to sample from the non-centered model.
@@
 The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In all cases, the plots show 1000 $R^*$ draws.
-See the R markdown file on the Github repository for code to reproduce this example.}
+In all cases, the plots show 1000 $R^*$ draws.}
 	\label{fig:eight_schools}
@@
 Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
-Note that, in the bottom plots, jitter has been added to the data points.
-See the R markdown file on the Github repository for code to reproduce this example.}
+Note that, in the bottom plots, jitter has been added to the data points.}
 	\label{fig:eight_schools_r_star_quantiles}
@@
 
-In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of samples.
+In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of draws.
 Specifically, this was to test that ML classification didn't become prone to overfitting in this limit.
@@
 With gradient-boosted regression tree models, we found this not to be the case.
-Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples.
+Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples: perhaps, because the classification boundaries for this task are unlikely to be very complex compared to (say) machine vision tasks.
 To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
@@
 	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
-In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.
-See the R markdown file on the Github repository for code to reproduce this example.}
+In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.}
 	\label{fig:mvt_wide_both}
@@
 Note, that it is possible for the ESS to exceed the actual sample size if there is negative autocorrelation in the Markov chains' values.
-In both cases, the results correspond to a single MCMC run composed of 4 chains.
-See the R markdown file on the Github repository for code to reproduce this example.}
+In both cases, the results correspond to a single MCMC run composed of 4 chains.}
 	\label{fig:wide_both_diagnostics}
@@
 The horizontal axis measures (half) the change in mean across the whole sample: so a value ``1'' indicates the mean increases by 2 units from the start to end of sampling.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-See the R markdown file on the Github repository for code to reproduce this example.}
+At each parameter set, 10 replicates were run and jitter was added to the points.}
 	\label{fig:trends_all_dim}
@@
 The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-See the R markdown file on the Github repository for code to reproduce this example.}
+At each parameter set, 10 replicates were run and jitter was added to the points.}
 	\label{fig:trends_one_dim}
@@
 The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-See the R markdown file on the Github repository for code to reproduce this example.}
+At each parameter set, 10 replicates were run and jitter was added to the points.}
 	\label{fig:trends_joint_distribution}
@@
 At each parameter set, 10 replicates were run and jitter was added to the points.
-The black line shows $R^*=1$.
-See the R markdown file on the Github repository for code to reproduce this example.}
+The black line shows $R^*=1$.}
 	\label{fig:trends_ar1}
@@
 The black horizontal line shows $R^*=1$.
-See the R markdown file on the Github repository for code to reproduce this example.}
+}
 	\label{fig:trends_ar1_transposed}
@@
 	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.
-See the R markdown file on the Github repository for code to reproduce this example.}
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
 	\label{fig:ovarian}
@@
 	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.
-See the R markdown file on the Github repository for code to reproduce this example.}
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
 	\label{fig:prostate}

### 2020-08-28 src/s_cauchy.Rmd
ben run through and sent to Aki

@@
   mutate(source=if_else(source=="split-Rhat", "Split-Rhat", as.character(source))) %>% 
+  mutate(iter=fct_relevel(as.factor(iter), "1000", "10,000")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
   geom_vline(xintercept = 1, linetype=2) +
-  facet_wrap(~source+iter, scales="free") +
+  facet_wrap(~iter+source, scales="free") +
   scale_x_continuous(limits=my_lims) +
@@
         legend.text = element_text(size=14, colour="black"),
-        legend.title = element_text(size=14, colour="black")) +
+        legend.title = element_text(size=16, colour="black")) +
   xlab("Value")

### 2020-08-28 write_up/ms.tex
ben run through and sent to Aki

@@
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
+We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{S_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
 We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
@@
 	\begin{tabular}{l|l|l}
-		Example &  Relevance & Section \\
+		\textbf{Example }&  \textbf{Relevance} & \textbf{Section} \\
 		\midrule
-		Autoregressive & Examining $R^*$ and its sensitivities to its calculation & \ref{sec:heterogeneity} \\
+		Autoregressive & \textbf{Examining $R^*$ and sensitivities to its calculation} & \ref{sec:heterogeneity} \\
 		& Detecting heterogeneous chain variance using $R^*$ & \ref{sec:heterogeneity_performance} \\
@@
 		\midrule
-		Multivariate normals & Convergence in joint distribution &  \ref{sec:multivariate_normal}\\
+		Multivariate normals & \textbf{Detecting convergence in joint distributions} &  \ref{sec:multivariate_normal}\\
 		& Unconverged joint distribution in bivariate normal & \ref{sec:multivariate_normal_bivariate}\\
@@
 		\midrule
-		Cauchy & Long-tailed distribution & \ref{sec:cauchy}\\
+		Cauchy & \textbf{Detecting convergence for long-tailed distributions} & \ref{sec:cauchy}\\
 		& Comparing $R^*$ and existing measures to objective convergence & \ref{sec:cauchy_objective}\\
 		\midrule
-		Eight schools model & Hierarchical Bayesian model slow convergence & \ref{sec:eight_shools}\\
+		Eight schools model & \textbf{Hierarchical Bayesian model slow convergence} & \ref{sec:eight_shools}\\
 		\midrule 
-		Wide multivariate normal & \# samples $\sim$ \# dims & \ref{sec:wide}\\
+		Wide multivariate normal & \textbf{Detecting convergence when \# draws $\sim$ \# dims} & \ref{sec:wide}\\
 		\midrule
-		Non-stationary marginals & Time-varying sampling distribution & \ref{sec:non-stationary}\\
+		Non-stationary marginals & \textbf{Detecting time-varying sampling distributions} & \ref{sec:non-stationary}\\
 		& Trends in mean across all chains & \ref{sec:non-stationary_chains}\\
@@
 		\midrule
-		Ovarian and prostate models & Bayesian models with many parameters and multimodal posteriors & \ref{sec:prostate}\\
+		Ovarian and prostate models & \textbf{Bayesian models with many parameters and multimodal posteriors} & \ref{sec:prostate}\\
 	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
@@
 \begin{equation}\label{eq:ar1}
-X_t = \rho X_{t-1} + \epsilon_t
+X_t = \rho X_{t-1} + \epsilon_t,
 \end{equation}
@@
 where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$.
-Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others.
+Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) standard deviation of the others.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the first dataset being the same as used for figures A-C; the second where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
 Note that, in D, only a single series of each series type is used to generate distribution.}
@@
 Intuitively, the reduction in sampling variation when composing training and test sets from larger samples results in lower variability in ML model predictions.
-We also expect that larger sample sizes should lead to higher $R^*$ values since more training data leads to better ML models.
+We also expect that larger sample sizes should lead to higher $R^*$ values, since more training data leads to better ML models.
 We may start to see this here, since the median $R^*=1.25$ at a sample size of 8000 was greater than for the smaller samples.
@@
 	\centerline{\includegraphics[width=0.6\textwidth]{../output/ar1_samplesize.pdf}}
-	\caption{\textbf{Autoregressive example: $R^*$ stochasticity.} The horizontal axis shows sample size; the vertical axis shows the value of $R^*$ calculated on chains split into two halves as per Algorithm \ref{alg:R_star}.
+	\caption{\textbf{Autoregressive example: $R^*$ stochasticity.} The horizontal axis shows sample size; the vertical axis shows the value of $R^*$ calculated as per Algorithm \ref{alg:R_star} applied to chains split into two halves.
 Grey points show the value of $R^*$ for each replicate (jitter was added to point positions).
@@
 \subsubsection{Uncertainty distribution for $R^*$}\label{sec:heterogeneity_uncertainty}
-Since GBMs return a probability simplex for each draw, which, for each chain, indicates the probability that the draw was generated by it, we can also generate a measure of uncertainty in $R^*$.
-Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done.
+GBMs return a probability simplex for each draw, indicating the probability that the draw was generated by a given chain.
+We can use this simplex to generate a measure of uncertainty in $R^*$ as detailed in Algorithm \ref{alg:R_star_uncertainty}.
 We demonstrate this idea using two datasets: one generated as described in \S\ref{sec:heterogeneity}, where one chain (out of four) has a lower variance than the others (we call this the ``unmixed'' data); and another, where all chains sample from the same distribution (we call this the ``mixed'' data).
@@
 For comparison, we also illustrate how the performance of rank-normalised split-$\widehat{R}$ varies with number of chains.
-To do so, we consider a problem similar to that described in \S\ref{sec:heterogeneity}: where all chains bar one have $\sigma=1$, and the remaining chain has $\sigma=1/3$.
-We consider cases with 2, 4, 8, 16, 32 and 64 chains.
+To do so, we consider an autoregressive example similar to that described in \S\ref{sec:heterogeneity}: where all chains bar one have $\sigma=1$, and the remaining chain has $\sigma=1/3$.
+We consider cases with 2, 4, 8, 16, and 32 chains.
 The other hyperparameters of the data generating process remain the same as in \S\ref{sec:heterogeneity}.
@@
 On the horizontal axis, we show the number of chains, and on the vertical axis, the value of each of the two convergence measures ($R^*$ in the left panel; $\widehat{R}$ in the right panel).
-Overall, both measures perform similarly as chain count increases.
+In general, both measures decline with chain count.
 For $R^*$, this is because it is harder to classify chains when there are more of them.
@@
 
-The decline of both of these measures with varying numbers of chains hints that perhaps a moving threshold for diagnosing convergence may be pertinent to avoid neglecting those minority of chains with differing information.
+The decline of both of these measures when more chains are used hints that perhaps a moving threshold for diagnosing convergence may be pertinent to avoid neglecting those minority of chains with differing information.
 Here, however, we do not make suggestions on what such guidelines could be and leave this for later work.
@@
 The performance of GBM, like all ML methods, depends on its hyperparameters.
-To investigate the robustness of $R^*$ to the GBM hyperparameter choice, we performed a sensitivity analysis.
+To investigate the robustness of $R^*$ to GBM hyperparameter choice, we performed a sensitivity analysis.
 In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
@@
 Fig.
-\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is closer to $R^*=1$ (mean $R^*=1.00$).
+\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution has mean $R^*=1.00$.
 Fig.
@@
 We now compare these with the more established metrics $\widehat{R}$ and ESS.
-For a GBM fitted to the centered model with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig.
+For a GBM fitted to the centered model of eq.
+\eqref{eq:mvt_normal_250} with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig.
 \ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
@@
 As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
-The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet, but its mean is above this value.
-In the bottom-left panel of Fig.
+The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet its mean remains above this value.
+In the top-right panel of Fig.
 \ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs.
@@
 Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of 4 chains.
-In the right-hand column of Fig.
+In the bottom row of Fig.
 \ref{fig:cauchy}, we show the results for these longer runs.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} Columns show MCMC runs with 1000 (left) and 10,000 (right; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
-Rows show the $R^*$ distributions (top) and rank-normalised split-$\widehat{R}$ values across all parameters (bottom).
+	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
+Columns show the $R^*$ distributions (left) and rank-normalised split-$\widehat{R}$ values across all parameters (right).
 Shadings indicate different model paramerisations as indicated in legend.}
@@
 This means that approximate KL-divergence is unreliable for these types of model.
-We decided not to use the Kolmogorov-Smirnov (KS) test, since it is most sensitive to differences between distributions around the median, whereas, here, we are interested in the behaviour in the tails.
+We decided not to use the Kolmogorov-Smirnov (KS) test, since it is most sensitive to differences between distributions around the median, whereas, here, we are interested in behaviour in the tails.
 Additionally, we found that the Anderson-Darling and Cram\'er-Von Mises tests \citep{faraway2019goftest}, which do not suffer the same shortcomings as the KS, behaved equally erratically and provided measures that were hard to intuit.
@@
 Instead, we chose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
-Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
+Specifically, we calculate the $R^2$ for the linear regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
 In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.
@@
 In Fig.
-\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the nominal model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence.
+\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the alternative model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence.
 Finally, in Figs.
@@
 \subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
+We now examine a classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
 
@@
 In Fig.
-\ref{fig:eight_schools}C, we plot the values of values of this metric when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
+\ref{fig:eight_schools}C, we plot $\widehat{R}$ values obtained when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
 These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all but a single parameter, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
@@
 The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In all cases, the plots show 1000 $R^*$ draws.}
+In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} for each parameterisation.}
 	\label{fig:eight_schools}
@@
 	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
-In each column, we show the path of the four individual chains above and 40 $R^*$ draws for each parameter quantile below.
+In each column, we show the path of the four individual chains above and 40 $R^*$ draws obtained using Algorithm \ref{alg:R_star_uncertainty} for each parameter quantile below.
 Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
@@
 \section{Non-stationary marginals}\label{sec:non-stationary}
-If a Markov chain does not mix with itself, this also indicates that convergence has not occurred \cite{gelman2013bayesian}.
+If a Markov chain does not mix with itself, this also indicates that convergence has not occurred \citep{gelman2013bayesian}.
 In this section, we investigate whether $R^*$ can detect non-stationary sampling distributions.
@@
 This plot mirrors Fig.
-18 in \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does.
+4 in the supplementary materials of \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does.
 As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.
@@
 Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time.
-Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations, and use i.i.d.
+Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
 draws from the process across 4 ``chains''.
@@
 
-To do so, we generated AR(1) processes (as defined in eq.
+To investigate this, we generated AR(1) processes (as defined in eq.
 (\ref{eq:ar1})) with autocorrelations, $\rho$, ranging from 0.8-1 and, in each case, calculated $R^*$ via Algorithm \ref{alg:R_star}.
@@
 
-The other dataset we use is of similar form but for prostate cancer, which is also analysed using logistic regression with regularised horseshoe priors. It has 18,105 parameters in total.
+The other dataset we use is of similar form but for prostate cancer.
+The original form of the dataset is described here: \cite{singh2002gene}.
+We use a filtered version of the dataset as detailed here: \cite{yang2006stable}, which we also analysed using logistic regression with regularised horseshoe priors.
+It has 18,105 parameters in total.
 
-For each model, we consider two MCMC runs: one with 4 chains run with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
+For each model, we consider two MCMC runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
 
@@
 Whilst it is harder to discern, this pattern is mirrored in Fig.
-\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 139 parameters where this is not the case.
+\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 139 parameters where this was not the case.
 Similarly, in Figs.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/ovarian.pdf}}
-	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
+	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
 In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{../output/prostate.pdf}}
-	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).
+	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
 In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}

### 2020-09-01 src/s_ar1.Rmd
font size and readability changes to figures

@@
   ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black"),
-        axis.text = element_text(colour="black")) +
+  theme(text = element_text(size=18, colour="black"),
+        axis.text = element_text(colour="black", size=16)) +
   xlim(0.9, NA)
@@
   ggtitle("C.") +
-  theme(text = element_text(size=14, colour="black"),
-        axis.text = element_text(colour="black")) +
+  theme(text = element_text(size=18, colour="black"),
+        axis.text = element_text(colour="black", size=16)) +
   xlim(1, NA)
@@
   ylab("Chain classification") +
-  theme(text = element_text(size=14, colour="black"),
+  theme(text = element_text(size=18, colour="black"),
         axis.text = element_text(colour="black")) +
@@
   ggtitle("D.") +
-  theme(text = element_text(size=14, colour="black"),
+  theme(text = element_text(size=18, colour="black"),
         axis.text = element_text(colour="black"),
@@
 saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
+r_star_vals <- readRDS("../data/ar1_sensitivity.rds")
 
@@
   geom_hline(yintercept = 1, linetype=2) +
-  annotate("text", x = b_sum$x_pos, y = b_sum$y_pos, label = b_sum$below_per, size=5) +
+  annotate("text", x = b_sum$x_pos, y = b_sum$y_pos, label = b_sum$below_per, size=6) +
   annotate("pointrange", x = 1:6, ymin=b_sum$lower, ymax=b_sum$upper, y = b_sum$middle, colour="black", size=1) +
@@
   scale_color_grey() +
-  theme(axis.text = element_text(colour="black", size=12),
-        axis.title = element_text(colour="black", size=14),
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=20),
         legend.position = "none")
@@
   scale_color_grey() +
-  theme(axis.text = element_text(colour="black", size=12),
-        axis.title = element_text(colour="black", size=14),
-        strip.text.x = element_text(colour="black", size=14),
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.x = element_text(colour="black", size=16),
         legend.position = "none")
@@
   scale_color_grey() +
-  theme(axis.text = element_text(colour="black", size=12),
-        axis.title = element_text(colour="black", size=14),
-        strip.text.x = element_text(colour="black", size=14),
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=20),
+        strip.text.x = element_text(colour="black", size=16),
         legend.position = "none")

### 2020-09-01 src/s_bivariate_normal.Rmd
font size and readability changes to figures

@@
 r_star_vals <- r_star_dist(full_samples)
-
+mean(r_star_vals>1)
 g <- 
@@
   ylab("Count") +
-  theme(text=element_text(size=14, colour="black"),
+  theme(text=element_text(size=20, colour="black"),
         axis.text = element_text(colour="black"))

### 2020-09-01 src/s_cauchy.Rmd
font size and readability changes to figures

--- 2020-08-02 src/s_cauchy.Rmd
@@
   melt(id.vars=c("source", "iter")) %>% 
+  mutate(source=if_else(source=="split-Rhat", "Split-Rhat", as.character(source))) %>% 
+  mutate(iter=fct_relevel(as.factor(iter), "1000", "10,000")) %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
   geom_histogram(bins=50, position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+  scale_fill_grey("Parameterisation", labels=c("Nominal", "Alternative")) +
   ylab("Count") +
   geom_vline(xintercept = 1, linetype=2) +
-  facet_wrap(~source+iter, scales="free") +
+  facet_wrap(~iter+source, scales="free") +
   scale_x_continuous(limits=my_lims) +
-  theme(strip.text = element_text(size=14, colour="black"),
-        axis.text = element_text(size=12, colour="black"),
-        axis.title = element_text(size=14, colour="black"),
-        legend.text = element_text(size=14, colour="black"),
-        legend.title = element_text(size=14, colour="black")) +
+  theme(strip.text = element_text(size=16, colour="black"),
+        axis.text = element_text(size=16, colour="black"),
+        axis.title = element_text(size=20, colour="black"),
+        legend.text = element_text(size=16, colour="black"),
+        legend.title = element_text(size=16, colour="black"),
+        legend.position = "top") +
   xlab("Value")
@@
   xlab("Iteration") +
-  theme(text=element_text(size=11, colour="black"),
-        axis.text = element_text(size=10, colour="black"),
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = "none",
@@
   scale_x_log10() +
-  ggtitle(TeX("A. Quantile $R^2$"))
-g1
+  ggtitle(TeX("A. Quantile-$R^2$"))
 ```
@@
   xlab("Iteration") +
-  theme(text=element_text(size=11, colour="black"),
-        axis.text = element_text(size=10, colour="black"),
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = c(0.8, 0.7),
@@
 saveRDS(cc_df, "../output/cauchy_ranked_unranked.rds")
+cc_df <- readRDS("../output/cauchy_ranked_unranked.rds")
 

### 2020-09-01 src/s_eight_schools.Rmd
font size and readability changes to figures

--- 2020-08-02 src/s_eight_schools.Rmd
@@
   ggtitle("A.") +
-  theme(text = element_text(size=14, colour="black"),
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = "none")
@@
   ggtitle("B.") +
-  theme(text = element_text(size=14, colour="black"),
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = "none")
@@
   geom_point() +
-  scale_x_continuous(limits=c(NA, 1.1)) +
-  scale_y_continuous(limits=c(NA, 1.1)) +
+  scale_x_continuous(limits=c(NA, 1.035)) +
+  scale_y_continuous(limits=c(NA, 1.035)) +
   geom_vline(xintercept = 1.01, linetype=2) +
@@
   geom_abline(linetype=3) +
-  scale_color_grey("Series", labels=c("centered", "non-centered")) +
+  scale_color_grey("Series", labels=c("Centered", "Non-centered")) +
   ylab("Split chains") +
   xlab("Original chains") +
-  ggtitle("C.")
+  ggtitle("C.") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black", size=10),
+        legend.position = c(0.72, 0.3))
 

### 2020-09-01 src/s_mvt_normal.Rmd
font size and readability changes to figures

--- 2020-05-06 src/s_mvt_normal.Rmd
@@
   scale_shape("LP param", labels=c("False", "True")) +
-  theme(legend.position = c(0.85, 0.7))
+  theme(legend.position = c(0.85, 0.6),
+        text = element_text(size=18, colour="black"),
+        axis.text = element_text(colour="black"))
 
@@
   xlab("Variable importance") +
-  ylab(TeX("split-$\\hat{R}$")) +
+  ylab(TeX("Split-$\\hat{R}$")) +
   ggtitle("C.") +
-  theme(legend.position = "none")
+  theme(legend.position = "none",
+        text = element_text(size=18, colour="black"),
+        axis.text = element_text(colour="black"))
 b_df <- tibble(samples_10000_cp=a_sum$zfsRhat[1:250])
@@
   ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
+  geom_histogram(position="identity", alpha=0.8, bins=50) +
+  scale_fill_grey("# draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
   xlab(TeX("$R*$")) +
@@
   geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8)) +
+  theme(text = element_text(size=18, colour="black"),
+        legend.position = c(0.8, 0.75),
+        axis.text = element_text(colour="black")) +
   ggtitle("A.")
@@
   ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
-  geom_histogram(position="identity", alpha=0.8, bins=20) +
-  scale_fill_grey("# of draws", labels=c("400, cp", "10000, cp", "10000, ncp")) +
-  xlab(TeX("split-$\\hat{R}$")) +
+  geom_histogram(position="identity", alpha=0.8, bins=60) +
+  scale_fill_grey("# draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
+  xlab(TeX("Split-$\\hat{R}$")) +
   ylab("Count") +
   geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = c(0.8, 0.8)) +
+  theme(text = element_text(size=18, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.8, 0.75)) +
   ggtitle("B.")

### 2020-09-01 src/s_ovarian_prostate.Rmd
font size and readability changes to figures

@@
   scale_fill_grey("MCMC detail", labels=c("4 chains: long", "16 chains: short")) +
-  theme(strip.text.x = element_text(size = 14, colour="black"),
-        text = element_text(size=14, colour="black"),
-        axis.text = element_text(size=14, colour="black"),
-        legend.text = element_text(size=14, colour="black"),
-        axis.title.x = element_blank()) +
+  theme(strip.text.x = element_text(size = 20, colour="black"),
+        text = element_text(size=18, colour="black"),
+        axis.text = element_text(size=18, colour="black"),
+        legend.text = element_text(size=18, colour="black"),
+        axis.title = element_text(size=18, colour="black"),
+        axis.title.x = element_blank(),
+        legend.position = "top") +
   ylab("Count")
@@
   scale_fill_grey("MCMC detail", labels=c("4 chains: long", "16 chains: short")) +
-  theme(strip.text.x = element_text(size = 14, colour="black"),
-        text = element_text(size=14, colour="black"),
-        axis.text = element_text(size=14, colour="black"),
-        legend.text = element_text(size=14, colour="black"),
-        axis.title.x = element_blank()) +
+  theme(strip.text.x = element_text(size = 20, colour="black"),
+        text = element_text(size=18, colour="black"),
+        axis.text = element_text(size=18, colour="black"),
+        legend.text = element_text(size=18, colour="black"),
+        axis.title = element_text(size=18, colour="black"),
+        axis.title.x = element_blank(),
+        legend.position = "top") +
   ylab("Count") +

### 2020-09-01 src/s_trends.Rmd
font size and readability changes to figures

--- 2020-08-02 src/s_trends.Rmd
@@
   geom_hline(yintercept = 1) +
-  theme(strip.text.x = element_text(size=12, colour="black"),
-        strip.text.y = element_text(size=12, colour="black"),
-        axis.text.x = element_text(size=10, colour="black"),
-        axis.text.y = element_text(size=10, colour="black")) +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black")) +
   xlab("Trend") +
@@
   geom_hline(yintercept = 1) +
-  theme(strip.text.x = element_text(size=12, colour="black"),
-        strip.text.y = element_text(size=12, colour="black"),
-        axis.text.x = element_text(size=10, colour="black"),
-        axis.text.y = element_text(size=10, colour="black")) +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black")) +
   ylab("Value") +
@@
 
-ggplot(data = res, aes(x = rho, y = zsRhat)) + 
+# ggplot(data = res, aes(x = rho, y = zsRhat)) + 
+#   geom_jitter() + 
+#   facet_grid(vars(iters)) + 
+#   geom_hline(yintercept = 1) + 
+#   ggtitle('R* without splitting') +
+#   ylim(0, 6)
+
+g <- ggplot(data = res, aes(x = rho, y = Rstar_split)) + 
   geom_jitter() + 
@@
   geom_hline(yintercept = 1) + 
-  ggtitle('R* without splitting') +
-  ylim(0, 6)
+  ylab(TeX("Split-R*")) +
+  xlab(TeX("$\\rho$")) +
+  ylim(0, 6) +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black"))
+ggsave("../output/trends_ar1.pdf", g, width = 8, height = 6)
 
-ggplot(data = res, aes(x = rho, y = Rstar_split)) + 
-  geom_jitter() + 
-  facet_grid(vars(iters)) + 
+b_sum <- res %>%
+  group_by(iters, rho) %>% 
+  summarise(lower=quantile(Rstar_split, 0.025),
+            upper=quantile(Rstar_split, 0.975),
+            Rstar_split=median(Rstar_split)) %>% 
+  mutate(type="summary")
+res1 <- res %>%
+  bind_rows(b_sum)
+
+g <- ggplot(data = filter(res1, is.na(type)), aes(x = as.factor(iters), y = Rstar_split, group=rho)) + 
+  geom_jitter(width = 0.2, height = 0, colour="grey") + 
+  geom_pointrange(data=filter(res1, !is.na(type)), aes(ymin=lower, ymax=upper)) +
+  geom_line(data=filter(res1, !is.na(type))) +
+  facet_wrap(~rho, scales="free") + 
   geom_hline(yintercept = 1) + 
-  ggtitle('Split-R*') +
-  ylim(0, 6)
+  ylab(TeX("Split-R*")) +
+  xlab("# iterations") +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black"))
+
+ggsave("../output/trends_ar1_transposed.pdf", g, width = 8, height = 6)
 ```

### 2020-09-01 src/s_wide_distributions.Rmd
font size and readability changes to figures

--- 2020-08-02 src/s_wide_distributions.Rmd
@@
   mutate(samples_ncp=r_star)
-saveRDS(a_df, "../output/mvt_500_all.rds")
+# saveRDS(a_df, "../output/mvt_500_all.rds")
 a_df <- readRDS("../output/mvt_500_all.rds")
@@
   geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
+  theme(text = element_text(size=20, colour="black"),
+        axis.text = element_text(colour="black"),
         legend.position = c(0.8, 0.8)) +
@@
         axis.title.x = element_blank())
-ggsave("../output/wide_ncp_diagnostics.pdf", h, width = 12, height = 8)
+# ggsave("../output/wide_ncp_diagnostics.pdf", h, width = 12, height = 8)
 
@@
   ylab("Count") +
-  theme(text=element_text(size=12, colour="black"),
-        axis.text=element_text(size=12, colour="black"),
-        strip.text = element_text(size=12, colour="black"),
+  theme(text=element_text(size=18, colour="black"),
+        axis.text=element_text(size=18, colour="black"),
+        strip.text = element_text(size=18, colour="black"),
         axis.title.x = element_blank())

### 2020-09-01 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
arxiv submission done; BA submission underway

--- 2020-08-28 write_up/ms.tex
@@
 % [preprint] option:
-\usepackage[preprint]{neurips_2019}
+% \usepackage[preprint]{neurips_2019}
 
@@
 
+\usepackage{arxiv}
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \usepackage{caption} 
+\usepackage{natbib}
 \captionsetup[table]{skip=10pt}
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_1.pdf}}
 	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (with marginal distribution to the right of panel); B shows two chains that have not mixed.}
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
 	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{../output/ar1_samplesize.pdf}}
+	\centerline{\includegraphics[width=0.6\textwidth]{ar1_samplesize.pdf}}
 	\caption{\textbf{Autoregressive example: $R^*$ stochasticity.} The horizontal axis shows sample size; the vertical axis shows the value of $R^*$ calculated as per Algorithm \ref{alg:R_star} applied to chains split into two halves.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.8\textwidth]{../output/ar1_numchains.pdf}}
+	\centerline{\includegraphics[width=0.8\textwidth]{ar1_numchains.pdf}}
 	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{ar1_sensitivity.pdf}}
 	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{../output/bivariate.pdf}}
+	\centerline{\includegraphics[width=0.6\textwidth]{bivariate.pdf}}
 	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws.}
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
+	\centerline{\includegraphics[width=1\textwidth]{mvt_three.pdf}}
 	\caption{\textbf{Multivariate normal example with 250 dimensions.} A shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws); B shows the rank-normalised split-$\widehat{R}$ values for all parameters from the same MCMC runs as in A; C shows variable importance versus $\widehat{R}$ for each parameter; and D shows variable importance versus bulk- and tail-ESS as calculated by \cite{vehtari2019rank}.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
 	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{cauchy_convergence.pdf}}
 	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools.pdf}}
 	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
+The legend inset in panel C provides a key for all panels.
 The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/eight_schools_r_star_quantiles.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools_r_star_quantiles.pdf}}
 	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_wide_both.pdf}}
+	\centerline{\includegraphics[width=1\textwidth]{mvt_wide_both.pdf}}
 	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
+	\centerline{\includegraphics[width=1\textwidth]{wide_both_diagnostics.pdf}}
 	\caption{\textbf{Wide data 250-dimensional example: established diagnostics.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_all_dim.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{trends_all_dim.pdf}}
 	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_one_dim.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{trends_one_dim.pdf}}
 	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_joint_distribution.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{trends_joint_distribution.pdf}}
 	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_ar1.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1.pdf}}
 	\caption{\textbf{Non-stationary distribution: AR(1) example.} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_ar1_transposed.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1_transposed.pdf}}
 	\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/ovarian.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{ovarian.pdf}}
 	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{../output/prostate.pdf}}
+	\centerline{\includegraphics[width=1.0\textwidth]{prostate.pdf}}
 	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).

### 2020-09-01 write_up/arxiv_submission_aug_2020/ms_ba.tex
arxiv submission done; BA submission underway

--- 2020-09-01 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
@@
 \usepackage{natbib}
+\usepackage{xr}
+\externaldocument[]{supplementary_materials}
+
 \captionsetup[table]{skip=10pt}
@@
 \newcommand{\LF}{\mathrm{Leapfrog}}
-
-\newcommand{\beginsupplement}{%
-	\setcounter{table}{0}
-	\renewcommand{\thetable}{S\arabic{table}}%
-	\setcounter{figure}{0}
-	\renewcommand{\thefigure}{S\arabic{figure}}%
-	\setcounter{section}{0}
-	\renewcommand{\thesection}{S\arabic{section}}%
-}
 
@@
 
-\beginsupplement
-
-\section{Wide datasets: multivariate normal}\label{sec:wide}
-As the number of parameter dimensions increases, it might be thought that ML algorithms will overfit the data, and, hence, testing set classification would be poor; leading to unreliable determinations of convergence.
-To test this hypothesis, we investigated two scenarios using a multivariate normal target.
-
-\subsection{250-dimensional model}
-In the first of these, we used the 250-dimensional multivariate normal of eq.
-\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
-Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
-\ref{fig:mvt_wide_both}A.
-In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
-The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
-\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, some of the parameters had $\widehat{R}>1.01$.
-Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
-\ref{fig:wide_both_diagnostics}B\&C).
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{mvt_wide_both.pdf}}
-	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
-In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.}
-	\label{fig:mvt_wide_both}
-\end{figure}
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{wide_both_diagnostics.pdf}}
-	\caption{\textbf{Wide data 250-dimensional example: established diagnostics.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
-Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
-Note, that it is possible for the ESS to exceed the actual sample size if there is negative autocorrelation in the Markov chains' values.
-In both cases, the results correspond to a single MCMC run composed of 4 chains.}
-	\label{fig:wide_both_diagnostics}
-\end{figure}
-
-\subsection{10,000-dimensional model}
-We next consider a more challenging example -- a target distribution with 10,000 dimensions.
-In this case, we assume independent standard normals for each dimension.
-In Fig.
-\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
-In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.
-
-Overall, the examples in this section suggest that $R^*$ is a conservative measure of convergence: when there are not enough draws, it will tend to diagnose non-convergence.
-We also note that the statistic took comparable time to calculate relative to existing convergence diagnostics on a desktop computer.
-
-\section{Non-stationary marginals}\label{sec:non-stationary}
-If a Markov chain does not mix with itself, this also indicates that convergence has not occurred \citep{gelman2013bayesian}.
-In this section, we investigate whether $R^*$ can detect non-stationary sampling distributions.
-
-\subsection{Trending mean across all chains}\label{sec:non-stationary_chains}
-We first recapitulate an example from appendix A in \cite{vehtari2019rank}.
-This example showed that split-$\widehat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
-Specifically, they first generated 4 chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling.
-We first repeat this analysis but using $R^*$ rather than $\widehat{R}$: in Fig.
-\ref{fig:trends_all_dim}, we show these results.
-In column A, we show the results for $R^*$ calculated on the 4 chains that ran; column B shows the same calculation but after the chains are split into two equal halves.
-The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
-This plot mirrors Fig.
-4 in the supplementary materials of \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does.
-As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.
-
-These results make intuitive sense: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation.
-In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals.
-By splitting chains into two -- the first half being the early phase, and the second half being the later phase with higher mean -- this forces differences in the marginals.
-This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one.
-As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\widehat{R}$ \citep{carpenter2017stan,vehtari2019rank}.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{trends_all_dim.pdf}}
-	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
-The rows present the differing sample sizes.
-The horizontal axis measures (half) the change in mean across the whole sample: so a value ``1'' indicates the mean increases by 2 units from the start to end of sampling.
-At each parameter set, 10 replicates were run and jitter was added to the points.}
-	\label{fig:trends_all_dim}
-\end{figure}
-
-\subsection{Trending mean in a single dimension}\label{sec:non-stationary_single}
-We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends.
-In Fig.
-\ref{fig:trends_one_dim}, we show how $R^*$ performs across a range of target dimensions.
-In the simulations here, all dimensions bar one are stationary; the remaining dimension has a linear trend added to it.
-In all cases, split-$R^*$ increased with trend.
-Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered.
-This suggests split-$R^*$ can robustly determine chain identity if only a single dimension has a non-stationary sampling distribution.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{trends_one_dim.pdf}}
-	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
-The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
-At each parameter set, 10 replicates were run and jitter was added to the points.}
-	\label{fig:trends_one_dim}
-\end{figure}
-
-\subsection{Trending covariance}\label{sec:non-stationary_covariance}
-Means that trend over time is one form of non-stationarity; another is a time-varying covariance.
-Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time.
-Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
-draws from the process across 4 ``chains''.
-Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution.
-Similarly, split-$\widehat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig.
-\ref{fig:trends_joint_distribution}A).
-By contrast, split-$R^*$ can (Fig.
-\ref{fig:trends_joint_distribution}B), since it uses all information in the samples, including the covariance structure.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{trends_joint_distribution.pdf}}
-	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
-The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
-At each parameter set, 10 replicates were run and jitter was added to the points.}
-	\label{fig:trends_joint_distribution}
-\end{figure}
-
-\subsection{Chain persistence}\label{sec:non-stationary_persistence}
-When forming training and testing sets as part of the ML algorithm used to determine $R^*$, the testing set is effectively treated as a sort of ``independent'' hold-out dataset.
-Markov chains, in general, have persistence, meaning that the test set will not be truly independent and can -- according to the level of autocorrelation in the chains -- be highly related to the training set.
-In this section, we investigate how this autocorrelation affects the performance of $R^*$.
-The difficulty with this question is that higher chain autocorrelation typically means the sampling distribution is a rougher approximation of the target, so $R^*$ should be higher due to the properties of the sampling distribution.
-It could also be higher because the training set is less distinct from the testing set.
-
-To investigate this, we generated AR(1) processes (as defined in eq.
-(\ref{eq:ar1})) with autocorrelations, $\rho$, ranging from 0.8-1 and, in each case, calculated $R^*$ via Algorithm \ref{alg:R_star}.
-Note, that only when $|\rho|<1$ is the marginal distribution defined by this process itself stationary; at $\rho=1$, its variance increases linearly with time, so, by definition, is not converged.
-In Fig.
-\ref{fig:trends_ar1}, we show the results of these simulations across various numbers of iterations: 250, 1000 and 4000 (different panels).
-In all cases, $R^*>1$ whenever $\rho=1$, indicating lack of convergence.
-As $\rho$ declined, so did $R^*$.
-
-Notably, as the number of iterations increased, the values of $R^*$ for $\rho<1$ declined, whereas $R^*$ for $\rho=1$ actually increased: this is most easily seen by examining Fig.
-\ref{fig:trends_ar1_transposed} (which shows the same data as in Fig.
-\ref{fig:trends_ar1} but in a different way).
-This characteristic is exactly as desired: larger sample sizes should yield a sampling distribution closer to convergence for $\rho<1$; the same for $\rho=1$ produces distributions that are no closer to convergence, and more samples allows better determination of this non-convergence.
-
-Overall, it seems that $R^*$ is a conservative measure and with greater chain autocorrelation it suggests more draws are necessary for convergence.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example.} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-The black line shows $R^*=1$.}
-	\label{fig:trends_ar1}
-\end{figure}
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1_transposed.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-The black dots show the median $R^*$ values at each parameter set; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
-The black horizontal line shows $R^*=1$.
-}
-	\label{fig:trends_ar1_transposed}
-\end{figure}
-
-\section{Many parameter models: ovarian and prostate analysis}\label{sec:prostate}
-In this section, we analyse two Bayesian models both fit to real data.
-The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
-Since there are relatively few data points relative to the number of predictors, regularised horseshoe priors are specified on the regression coefficients \citep{piironen2017sparsity} since most are expected to be zero.
-This dataset has been used for benchmarking in the past (see, for example, \cite{schummer1999comparative,hernandez2010expectation,paananen2019implicitly}) and is known to result in a multimodal posterior.
-
-The other dataset we use is of similar form but for prostate cancer.
-The original form of the dataset is described here: \cite{singh2002gene}.
-We use a filtered version of the dataset as detailed here: \cite{yang2006stable}, which we also analysed using logistic regression with regularised horseshoe priors.
-It has 18,105 parameters in total.
-
-For each model, we consider two MCMC runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-
-For the ovarian model, we show the results in Fig.
-\ref{fig:ovarian}.
-In Fig.
-\ref{fig:ovarian}A, we show the $R^*$ distributions for each model run, which show that, whereas the ``long'' model run has converged, the ``short'' one has yet to do so.
-Whilst it is harder to discern, this pattern is mirrored in Fig.
-\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 139 parameters where this was not the case.
-Similarly, in Figs.
-\ref{fig:ovarian}C and \ref{fig:ovarian}D, which show the bulk-ESS and tail-ESS respectively, it is evident that, for the short model, there remain a few parameters with low effective sample sizes, whereas the long model has more consistent values for this metric.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ovarian.pdf}}
-	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
-	\label{fig:ovarian}
-\end{figure}
-
-For the prostate model, we show the results in Fig.
-\ref{fig:prostate}.
-Since this model has nearly four times as many parameters as the ovarian model, it was more computationally expensive to estimate $R^*$ for it.
-To handle this, we thinned down the parameters by a factor of 5 for the long model, recognising that, of course, this measure will make it more likely that we diagnose convergence.
-Despite this, both $R^*$ measures indicated that the MCMC runs had yet to converge (Fig.
-\ref{fig:prostate}A), which was mirrored by the other metrics considered (Figs.
-\ref{fig:prostate}B,C\&D).
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{prostate.pdf}}
-	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
-	\label{fig:prostate}
-\end{figure}
-
 \end{document}

### 2020-09-01 write_up/arxiv_submission_aug_2020/supplementary_materials.tex
arxiv submission done; BA submission underway

--- 2020-09-01 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
@@
 \usepackage{natbib}
+
+\usepackage{xr}
+\externaldocument[]{ms_ba}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thesection}{S\arabic{section}}
+
 \captionsetup[table]{skip=10pt}
@@
 
-\newcommand{\beginsupplement}{%
-	\setcounter{table}{0}
-	\renewcommand{\thetable}{S\arabic{table}}%
-	\setcounter{figure}{0}
-	\renewcommand{\thefigure}{S\arabic{figure}}%
-	\setcounter{section}{0}
-	\renewcommand{\thesection}{S\arabic{section}}%
-}
-
-\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines}
+\title{Supplementary materials: $R^*$ convergence diagnostic}
 
@@
 \begin{document}
-% \nipsfinalcopy is no longer used
-
 	\maketitle
-
-\begin{abstract}
-  Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
-Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
-The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
-Here, we introduce a new method for diagnosing convergence based on whether a machine learning classifier model can successfully discriminate the individual chains.
-We call this convergence measure $R^*$.
-In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
-Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
-The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
-\end{abstract}
-
-\section{Introduction}
-Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement.
-MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
-
-Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care.
-A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior.
-Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
-
-The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
-MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions).
-Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small number of draws.
-
-The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
-Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
-Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged.
-Like furious bees trapped in a room of a house \citep{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
-With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.
-
-To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}.
-If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge.
-This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}.
-The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\widehat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman1992inference}.
-If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed.
-Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter draws rather than the raw values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}.
-Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
-Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.
-
-Here, we introduce $R^*$, a new convergence metric.
-This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
-Rephrased, it should not be possible to predict the chain that \textit{caused} a draw.
-In this vein, we use a machine learning (ML) classifier to measure convergence.
-Specifically, we train a ML classifier to predict the chain that generated each observation.
-By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
-Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
-However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
-For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
-It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
-In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
-$R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
-
-The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank} and elsewhere.
-Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
-
-\section{Method}\label{sec:method}
-If Markov chains have not mixed, it is possible to determine to which chain a draw belongs from its value.
-This is possible if there are differences in the sampling distribution for any dimension, $\theta$, in the target distribution (Fig.
-\ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a draw belongs to.
-It  is also possible to predict the chain that generated a given draw if there are differences in the joint distribution of two (or more) dimensions of the target, even if the marginal distributions are the same (Fig.
-\ref{fig:joint}).
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_1.pdf}}
-	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (with marginal distribution to the right of panel); B shows two chains that have not mixed.}
-	\label{fig:marginal}
-\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_2.pdf}}
-	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
-	\label{fig:joint}
-\end{figure}
-
-These two cases, whilst simple, illustrate the basis of our approach.
-To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each draw belongs.
-By evaluating its performance on a separate test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
-In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
-Algorithm \ref{alg:R_star} gives a recipe for calculating $R^*$.
-
-\begin{algorithm}[tb]
-	\caption{$R^*$ calculation}
-	\label{alg:R_star}
-	\begin{algorithmic}
-		\STATE Given chain-wise draws from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
-		\FOR{$m=1$ to $N$}
-		\STATE Create train and test sets by random-sampling (w/o replacement), $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
-		\ENDFOR
-		\STATE Stack $X_\text{train} = (X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train})^T$
-		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
-		\STATE Train ML model to classify chain id from any draw, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
-		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
-		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
-		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
-		\ENDFOR
-		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N \bar{a}$
-		\RETURN $R^*$
-	\end{algorithmic}
-\end{algorithm}
-
-The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
-Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
-The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters.
-We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing.
-Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
-Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.
-
-From the GBM fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
-Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
-For each draw, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
-By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{S_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
-We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
-
-\begin{algorithm}[tb]
-	\caption{Procedure to generate $I$ draws of $R^*$}
-	\label{alg:R_star_uncertainty}
-	\begin{algorithmic}
-		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
-		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
-		\FOR{$i=1$ to $I$}
-		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
-		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
-		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
-		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
-		\ENDFOR
-		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^{*\{i\}} = \bar{a} / (1 / N) = N\bar{a}$
-		\ENDFOR
-		\RETURN $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$
-	\end{algorithmic}
-\end{algorithm}
-
-\section{Results}\label{sec:results}
-To illustrate the versatility of $R^*$, we use a range of examples that demonstrate how this statistic fares across a range of scenarios.
-Table \ref{tab:results} summarises the examples and provides a rationale for their inclusion.
-The experiments not detailed in the main text are briefly described in \S\ref{sec:further_experiments} and more fully in the relevant sections given in Table \ref{tab:results}.
-
-\begin{table}[]
-	\begin{tabular}{l|l|l}
-		\textbf{Example }&  \textbf{Relevance} & \textbf{Section} \\
-		\midrule
-		Autoregressive & \textbf{Examining $R^*$ and sensitivities to its calculation} & \ref{sec:heterogeneity} \\
-		& Detecting heterogeneous chain variance using $R^*$ & \ref{sec:heterogeneity_performance} \\
-		& Stochasticity in $R^*$ & \ref{sec:heterogeneity_stochasticity}\\
-		& Generating $R^*$ uncertainty measure & \ref{sec:heterogeneity_uncertainty}\\
-		& Sensitivity of $R^*$ to number of chains & \ref{sec:heterogeneity_numchains}\\
-		& Robustness of $R^*$ to ML model hyperparameters & \ref{sec:heterogeneity_robustness}\\
-		\midrule
-		Multivariate normals & \textbf{Detecting convergence in joint distributions} &  \ref{sec:multivariate_normal}\\
-		& Unconverged joint distribution in bivariate normal & \ref{sec:multivariate_normal_bivariate}\\
-		& High correlations between dims in 250D normal  & \ref{sec:multivariate_normal_250}\\
-		& Measuring contributions of variables to poor convergence & \ref{sec:multivariate_normal_varimportance}\\
-		\midrule
-		Cauchy & \textbf{Detecting convergence for long-tailed distributions} & \ref{sec:cauchy}\\
-		& Comparing $R^*$ and existing measures to objective convergence & \ref{sec:cauchy_objective}\\
-		\midrule
-		Eight schools model & \textbf{Hierarchical Bayesian model slow convergence} & \ref{sec:eight_shools}\\
-		\midrule 
-		Wide multivariate normal & \textbf{Detecting convergence when \# draws $\sim$ \# dims} & \ref{sec:wide}\\
-		\midrule
-		Non-stationary marginals & \textbf{Detecting time-varying sampling distributions} & \ref{sec:non-stationary}\\
-		& Trends in mean across all chains & \ref{sec:non-stationary_chains}\\
-		& Trends in mean in a single dimension & \ref{sec:non-stationary_single}\\
-		& Trends in covariance & \ref{sec:non-stationary_covariance}\\
-		& Sensitivity of $R^*$ to chain persistence & \ref{sec:non-stationary_persistence}\\
-		\midrule
-		Ovarian and prostate models & \textbf{Bayesian models with many parameters and multimodal posteriors} & \ref{sec:prostate}\\
-	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
-\end{table}
-
-\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
-In this section, we show how $R^*$ is able to detect heterogeneous variance across Markov chains.
-Apart from demonstrating how $R^*$ performs, this section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}); and, finally, to the ML model hyperparameters (\S\ref{sec:heterogeneity_robustness}).
-The experiments we use to study these issues are all of similar form to the following data generating process: four Markov chains are generated, where each samples from an autoregressive order 1 (AR(1)) process of the form,
-%
-\begin{equation}\label{eq:ar1}
-X_t = \rho X_{t-1} + \epsilon_t,
-\end{equation}
-%
-where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$.
-Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) standard deviation of the others.
-
-\subsubsection{Performance of $R^*$}\label{sec:heterogeneity_performance}
-To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e.
-where one chain has a lower variance).
-We then fit a GBM to a labelled training set.
-The fitted model is then used to classify draws in an independent test set according to the chain which generated them.
-For each replicate, we then calculated $R^*$ as described in Algorithm \ref{alg:R_star}.
-
-In Fig.
-\ref{fig:ar1}A, we show how a GBM fitted to one such replicate dataset classifies observations according to a draw's value.
-Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
-
-In Fig.
-\ref{fig:ar1}B, we show that $R^*>1$ for all replicates, indicating that the chains had not converged in all cases.
-In Fig.
-\ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate;  as for $R^*$, this metric indicates the chains had not converged in all replicates (as diagnosed by $R^*<1.01$).
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
-Note that, in D, only a single series of each series type is used to generate distribution.}
-	\label{fig:ar1}
-\end{figure}
-
-\subsubsection{Stochasticity in $R^*$}\label{sec:heterogeneity_stochasticity}
-Unlike $\widehat{R}$, $R^*$ is a stochastic convergence measure due to randomness in creating training and testing sets (essentially a form of sampling variation) and randomness in the methods used to train the ML model.
-This means that even if the same sample is used, $R^*$ will return a different value each time it is calculated if the pseudorandom seed is not fixed.
-To probe the extent of this randomness, we generated data using the same process as in \S\ref{sec:heterogeneity} but now using varying sample sizes, including samples consisting of 500, 1000, 2000, 4000 and 8000 draws.
-For each dataset, we computed $R^*$ on it 100 times, allowing the pseudorandom seed to vary between calculations.
-We stress that, for each sample size, we used the same dataset (so there were 5 datasets created in total -- one for each sample size), so stochasticity comes from $R^*$ calculation, not that from the data generating process.
-
-In Fig.
-\ref{fig:ar1_samplesize}, we show the results of this study.
-In this figure, the horizontal axis shows the sample size, and the vertical axis, the value of $R^*$ in each repetition.
-This shows that as the number of samples increased, variation in $R^*$ declined.
-At a sample size of 500, there were four cases where $R^*<1$; in larger samples, there were none.
-Intuitively, the reduction in sampling variation when composing training and test sets from larger samples results in lower variability in ML model predictions.
-We also expect that larger sample sizes should lead to higher $R^*$ values, since more training data leads to better ML models.
-We may start to see this here, since the median $R^*=1.25$ at a sample size of 8000 was greater than for the smaller samples.
-
-If randomness in $R^*$ calculation leads to different conclusions about convergence being drawn, this would be problematic.
-One potential remedy for this is to repeatedly calculate $R^*$ on a given sample, much as we have done here, and consider the distribution of $R^*$ values computed.
-The computational cost of doing this may, of course, be unreasonable.
-Instead, in \S\ref{sec:heterogeneity_uncertainty}, we consider an alternative approach based on bootstrapping a single ML model's predictions.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{ar1_samplesize.pdf}}
-	\caption{\textbf{Autoregressive example: $R^*$ stochasticity.} The horizontal axis shows sample size; the vertical axis shows the value of $R^*$ calculated as per Algorithm \ref{alg:R_star} applied to chains split into two halves.
-Grey points show the value of $R^*$ for each replicate (jitter was added to point positions).
-Black points show the median $R^*$ value; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
-For each sample size, a single dataset was created and used for all $R^*$ calculations.}
-	\label{fig:ar1_samplesize}
-\end{figure}
-
-\subsubsection{Uncertainty distribution for $R^*$}\label{sec:heterogeneity_uncertainty}
-GBMs return a probability simplex for each draw, indicating the probability that the draw was generated by a given chain.
-We can use this simplex to generate a measure of uncertainty in $R^*$ as detailed in Algorithm \ref{alg:R_star_uncertainty}.
-We demonstrate this idea using two datasets: one generated as described in \S\ref{sec:heterogeneity}, where one chain (out of four) has a lower variance than the others (we call this the ``unmixed'' data); and another, where all chains sample from the same distribution (we call this the ``mixed'' data).
-In Fig.
-\ref{fig:ar1}D, we show the $R^*$ distributions in each case.
-For the unmixed data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
-For the mixed data, the distribution is centred on 1, indicating convergence.
-In the mixed case, there are many draws where $R^*<1$: these indicate that, in that particular draw from the probability simplex, chain classification is actually worse than selecting a chain identification uniformly at random.
-Much like how it is possible for $\hat{R}<1$, this is a sample property, driven by the sampling distribution of the categorical distribution defined by the probability simplex.
-
-It is worth emphasising that the uncertainty distribution obtained by Algorithm \ref{alg:R_star_uncertainty} differs from that obtained from repeatedly calculating $R^*$ via Algorithm \ref{alg:R_star} as was done in \S\ref{sec:heterogeneity_stochasticity}.
-In Algorithm \ref{alg:R_star_uncertainty}, variation in $R^*$ comes from sampling from the probability simplex: if predicted chain probabilities are close to uniform, there will be greater uncertainty in $R^*$.
-Repeatedly calculating $R^*$ by applying Algorithm \ref{alg:R_star} to the same dataset yields a distribution whose width derives from sampling variation when forming training and testing sets and the stochasticity in training ML models.
-Collectively, these differences mean that the two measures of uncertainty will differ.
-
-There is an additional difference, though, in the central points of each distribution: the distribution obtained by Algorithm \ref{alg:R_star_uncertainty} will, in general, have a lower mean than that obtained by repeated application of Algorithm \ref{alg:R_star}.
-To see this, note that the darker-shaded $R^*$ distribution in Figure \ref{fig:ar1}D was generated via Algorithm \ref{alg:R_star_uncertainty} and has a mean around 1.07; the distribution shown in Figure \ref{fig:ar1}B was generated by repeatedly recomputing $R^*$ using Algorithm \ref{alg:R_star} and has a mean closer to 1.22.
-This difference in mean is expected since predictive performance when assigning chain identities stochastically when sampling from the categorical distribution of the probability simplex (as is done in Algorithm \ref{alg:R_star_uncertainty}) will generally result in worse prediction than when assigning each chain identity using whichever chain has the highest class probability (as is done in Algorithm \ref{alg:R_star}).
-Of course, we would prefer it if the uncertainty distribution generated by Algorithm \ref{alg:R_star_uncertainty} had a mean closer to the one obtained by repeated application of Algorithm \ref{alg:R_star}.
-Nonetheless, in practice, we have found that the mean of the $R^*$ distribution generated by Algorithm \ref{alg:R_star_uncertainty} provides a useful cheaper diagnostic.
-
-\subsubsection{Sensitivity to number of chains}\label{sec:heterogeneity_numchains}
-We have so far focused on the sensitivity of $R^*$ to chain heterogeneity with a fixed number of chains: four.
-Since classification becomes a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
-For comparison, we also illustrate how the performance of rank-normalised split-$\widehat{R}$ varies with number of chains.
-To do so, we consider an autoregressive example similar to that described in \S\ref{sec:heterogeneity}: where all chains bar one have $\sigma=1$, and the remaining chain has $\sigma=1/3$.
-We consider cases with 2, 4, 8, 16, and 32 chains.
-The other hyperparameters of the data generating process remain the same as in \S\ref{sec:heterogeneity}.
-
-In Fig.
-\ref{fig:ar1_numchains}, we show the results of these simulations with 50 replicates at each number of chains.
-On the horizontal axis, we show the number of chains, and on the vertical axis, the value of each of the two convergence measures ($R^*$ in the left panel; $\widehat{R}$ in the right panel).
-In general, both measures decline with chain count.
-For $R^*$, this is because it is harder to classify chains when there are more of them.
-For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
-Across the replicates we ran, median $\widehat{R}<1.01$ for 16 or more chains; a minimum of $R^*=1.10$ was obtained for 32 chains.
-
-The decline of both of these measures when more chains are used hints that perhaps a moving threshold for diagnosing convergence may be pertinent to avoid neglecting those minority of chains with differing information.
-Here, however, we do not make suggestions on what such guidelines could be and leave this for later work.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.8\textwidth]{ar1_numchains.pdf}}
-	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
-The vertical axis shows the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
-Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 50 replicates at each number of chains.}
-	\label{fig:ar1_numchains}
-\end{figure}
-
-\subsubsection{Robustness to hyperparameter choice}\label{sec:heterogeneity_robustness}
-The performance of GBM, like all ML methods, depends on its hyperparameters.
-To investigate the robustness of $R^*$ to GBM hyperparameter choice, we performed a sensitivity analysis.
-In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
-\ref{fig:ar1}.
-The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
-The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
-\ref{fig:ar1_sensitivity}.
-These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper).
-This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ar1_sensitivity.pdf}}
-	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
-The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.}
-	\label{fig:ar1_sensitivity}
-\end{figure}
-
-\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
-In this section, we illustrate how $R^*$ can diagnose convergence issues in the joint target distribution.
-
-\subsubsection{Bivariate model}\label{sec:multivariate_normal_bivariate}
-First, we consider a bivariate normal density.
-In all four chains, we use independent sampling to generate 2000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
-By construction, all chains target the same marginal distribution in each dimension, but the fourth chain has a different joint distribution.
-
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures that aim to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail-ESS.
-In all cases, the various quantities were calculated based on chains split into halves.
-For both dimensions, the two ESS measures were above 7000, and $\widehat{R}<1.001$, indicating no issues with convergence.
-
-Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty}, which is shown in Fig.
-\ref{fig:bivariate}.
-The mean of this distribution is 1.14, and $>$99\% of $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
-By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{bivariate.pdf}}
-	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws.}
-	\label{fig:bivariate}
-\end{figure}
-
-\subsubsection{250-dimensional model}\label{sec:multivariate_normal_250}
-We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}.
-We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm to sample from this target distribution and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
-First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
-%
-\begin{equation}\label{eq:mvt_normal_250}
-\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
-\end{equation}
-%
-where $\boldsymbol{x}\in\mathbb{R}^{250}$.
-For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig.
-\ref{fig:mvt}A.
-From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.
-\ref{fig:mvt}B), although 54\% had bulk-$\text{ESS}<400$ and 13\% of parameters had tail-$\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
-
-Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it.
-This model has the form,
-%
-\begin{align}
-\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
-\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
-z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
-\end{align}
-%
-where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$.
-Fig.
-\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution has mean $R^*=1.00$.
-Fig.
-\ref{fig:mvt}B shows the $\widehat{R}$ values for each $x$ parameter in this model, and, echoing the result for $R^*$, $\widehat{R}<1.01$ in all cases; further, bulk- and tail-$\text{ESS}>400$ for all parameters.
-
-\subsubsection{Variable importance}\label{sec:multivariate_normal_varimportance}
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions.
-We now compare these with the more established metrics $\widehat{R}$ and ESS.
-For a GBM fitted to the centered model of eq.
-\eqref{eq:mvt_normal_250} with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig.
-\ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-In this plot, there is a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$).
-In Fig.
-\ref{fig:mvt}D, we plot variable importance versus two measures: bulk-ESS and tail-ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk-ESS: $\rho=-0.57, S=4142470, p<0.01$; tail-ESS: $\rho=-0.56, S=4113709, p<0.01$).
-Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example with 250 dimensions.} A shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws); B shows the rank-normalised split-$\widehat{R}$ values for all parameters from the same MCMC runs as in A; C shows variable importance versus $\widehat{R}$ for each parameter; and D shows variable importance versus bulk- and tail-ESS as calculated by \cite{vehtari2019rank}.
-In A, 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} are shown for each MCMC run.
-In plots C and D, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.}
-	\label{fig:mvt}
-\end{figure}
-
-\subsection{Infinite variance: Cauchy example}\label{sec:cauchy}
-We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
-Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-dimensional vector $x$,
-%
-\begin{equation}
-x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50.
-\end{equation}
-%
-We call this parameterisation the ``nominal'' version of this model.
-
-In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \citep{vehtari2019rank},
-%
-\begin{align}
-a_j \sim  \N(0,1), \qquad
-b_j \sim  \text{Gamma}(0.5, 0.5), \qquad
-x_j =  a_j/\sqrt{b_j}.
-\end{align}
-%
-The distribution of the $x$ vector is the same under both parameterisations, although the thin-tailed $(a,b)$ vectors define a higher dimensional posterior that improves sampling efficiency.
-
-In the top-left panel of Fig.
-\ref{fig:cauchy}, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
-The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet its mean remains above this value.
-In the top-right panel of Fig.
-\ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs.
-The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}<1.01$ for all parameters.
-
-Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of 4 chains.
-In the bottom row of Fig.
-\ref{fig:cauchy}, we show the results for these longer runs.
-In these, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
-Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
-The $\widehat{R}$ values are all below 1.01, indicating convergence in both cases.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
-Columns show the $R^*$ distributions (left) and rank-normalised split-$\widehat{R}$ values across all parameters (right).
-Shadings indicate different model paramerisations as indicated in legend.}
-	\label{fig:cauchy}
-\end{figure}
-
-\subsubsection{Measuring convergence objectively}\label{sec:cauchy_objective}
-To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target.
-One measure of distributional ``closeness'' is the KL-divergence, which, in this case, could be used to measure the divergence from target to sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated for each dimension.
-The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails.
-This means that approximate KL-divergence is unreliable for these types of model.
-We decided not to use the Kolmogorov-Smirnov (KS) test, since it is most sensitive to differences between distributions around the median, whereas, here, we are interested in behaviour in the tails.
-Additionally, we found that the Anderson-Darling and Cram\'er-Von Mises tests \citep{faraway2019goftest}, which do not suffer the same shortcomings as the KS, behaved equally erratically and provided measures that were hard to intuit.
-The Wasserstein distance was also trialled but had great uncertainty due to the long-tails of the Cauchy.
-Instead, we chose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
-Specifically, we calculate the $R^2$ for the linear regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
-In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.
-
-In Fig.
-\ref{fig:cauchy_convergence}A, we plot this \textit{quantile-$R^2$} as a function of MCMC sample size for both parameterisations of the Cauchy model.
-This shows that after c.10,000 iterations, the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
-Next, in Figs.
-\ref{fig:cauchy_convergence}B\&C, we plot two measures of $\widehat{R}$, each calculated from splitting the 4 original chains into two equal halves.
-The first of these measures is the rank-normalised $\widehat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig.
-\ref{fig:cauchy_convergence}B, we show how the maximum of this measurement across all 50 dimensions changes with sample size.
-After c.550 iterations, the alternative parameterisation achieves $\widehat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum $\widehat{R}$ value: in both cases, these suggest convergence.
-The second measure is multivariate $\widehat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig.
-\ref{fig:cauchy_convergence}C shows how this metric changes with sample size for both Cauchy model parameterisations.
-After c.1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more draws are needed.
-In Fig.
-\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the alternative model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence.
-Finally, in Figs.
-\ref{fig:cauchy_convergence}E\&F, we plot the minimum across all the dimensions of tail- and bulk-ESS calculated as described in \cite{vehtari2019rank}.
-After c.180 iterations, the alternative parameterisation surpassed a tail-ESS of 400; after c.18,700, the nominal parameterisation did the same.
-Both models were quicker to pass 400 bulk-ESSs.
-
-Comparing our ``omniscient'' measure of convergence shown in Fig.
-\ref{fig:cauchy_convergence}A, with the various derived measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
-The rate at which these converge differs though, and $R^*$ (Fig.
-\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to our ``omniscient'' measure.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{cauchy_convergence.pdf}}
-	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
-Dashed lines indicate recommended thresholds for each convergence statistic.}
-	\label{fig:cauchy_convergence}
-\end{figure}
-
-\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-We now examine a classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
-
-The model can be parameterised in two ways, as described in \cite{vehtari2019rank}.
-The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
-%
-\begin{align*}
-\theta_j &\sim \N(\mu, \tau), \\
-y_j &\sim \N(\theta_j, \sigma_j).
-\end{align*}
-%
-The ``non-centered'' parameterisation recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
-This can be written as,
-%
-\begin{align*}
-\tilde{\theta}_j &\sim \N(0, 1), \\
-\theta_j &= \mu + \tau \tilde{\theta}_j,\\
-y_j &\sim \N(\theta_j, \sigma_j).
-\end{align*}
-%
-In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
-of the distribution of these effects.
-In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
-
-We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{betancourt2017conceptual}, meaning that the resultant sampling distribution is likely to be biased.
-We also used the same algorithm settings to sample from the non-centered model.
-
-To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
-The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
-In this plot, the centered model is close to convergence, whereas the non-centered is not.
-
-In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two.
-The results are shown in Fig.\ref{fig:eight_schools}B.
-In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each draw, meaning that the centered model $R^*$ values are shifted leftwards.
-Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears near convergence.
-
-It is recommended that $\widehat{R}$, like $R^*$, be calculated using split chains.
-In Fig.
-\ref{fig:eight_schools}C, we plot $\widehat{R}$ values obtained when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all but a single parameter, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
-All parameters for the non-centered models were below 1.01, indicating convergence.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools.pdf}}
-	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
-The legend inset in panel C provides a key for all panels.
-The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} for each parameterisation.}
-	\label{fig:eight_schools}
-\end{figure}
-
-\subsubsection{Understanding chain classification}
-To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
-After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
-In Fig.
-\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
-In the top row of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
-Above the top panel, we show the marginal distributions for each chain.
-In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
-In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
-For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
-Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved ML predictive accuracy.
-
-Fig.
-\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in ML classification accuracy and falsely concluding that convergence has occurred.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools_r_star_quantiles.pdf}}
-	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
-In each column, we show the path of the four individual chains above and 40 $R^*$ draws obtained using Algorithm \ref{alg:R_star_uncertainty} for each parameter quantile below.
-Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
-Note that, in the bottom plots, jitter has been added to the data points.}
-	\label{fig:eight_schools_r_star_quantiles}
-\end{figure}
-
-\subsection{Further experiments}\label{sec:further_experiments}
-Alongside the examples included in the main text, there are a number of supplementary text examples, which we briefly outline here.
-
-In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of draws.
-Specifically, this was to test that ML classification didn't become prone to overfitting in this limit.
-To test this hypothesis, we investigated two scenarios using a multivariate normal target: one with a 250-multivariate normal with high correlation between dimensions using 250 post-warm-up iterations; and another normal with 10,000 independent dimensions using up to 500 post-warm-up iterations.
-In both cases, sampling was done using Stan's NUTS algorithm.
-In both cases, $R^*$ and rank-normalised split-$\widehat{R}$ reached similar conclusions about convergence: namely, that more iterations were needed in all experiments considered.
-Overall, these experiments show that $R^*$ is a conservative convergence measure that will tend to diagnose unconvergence when there are insufficient draws.
-
-In \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain convergence is diagnosed.
-We illustrate this via four examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; (c), a bivariate normal with a non-stationary covariance; and (d), an autocorrelated sampling distribution with a univariate normal target with a range of different autocorrelations.
-The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue.
-The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish as the numbers of dimensions considered increased.
-Example (c) showed that split-$R^*$ opposed to split-$\widehat{R}$ is able to diagnose non-stationary covariance between dimensions of a target distribution.
-In (d), we show that $R^*$ is able to differentiate between distributions with non-stationary target distributions and stationary ones.
-It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there insufficient draws.
-
-In \S\ref{sec:prostate}, we show that $R^*$ performs well for two Bayesian logistic regression problems with highly multimodal posteriors.
-Each of these models have 1000s of parameters, and we found that it was slow to compute both $\widehat{R}$ and $R^*$ for them.
-That said, the computational time for calculating $\widehat{R}$ was considerably less than was needed for $R^*$.
-
-\section{Discussion}
-If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a draw, it should be impossible to discern which chain generated it.
-Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
-By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
-By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
-Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
-
-The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target.
-$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each draw.
-It also naturally includes information about the joint distribution across all dimensions of the target.
-Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects.
-Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
-This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
-
-When first starting to develop $R^*$, it was unclear to us whether \textit{any} metric aligned with a supervised ML method would be overly sensitive to its hyperparameters.
-With gradient-boosted regression tree models, we found this not to be the case.
-Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples: perhaps, because the classification boundaries for this task are unlikely to be very complex compared to (say) machine vision tasks.
-To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
-
-In \S\ref{sec:prostate}, we fit Bayesian models with many 1000s of parameters then used $R^*$ to diagnose convergence, finding that $R^*$ was considerably more expensive to calculate than $\widehat{R}$.
-The computational cost of training GBMs has been suggested to be $\mathcal{O}(N k)$, where $N$ is the number of training data points and $k$ is the number of dimensions of the target \citep{complexity2018}.
-If so, this suggests that larger models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
-
-Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
-In a number of examples, we trial this before calculating $R^*$ and find that this approach leads to more accurate chain prediction.
-We recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
-Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
-If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.
-
-MCMC enables inference across a wide range of models encountered across the social, biological and physical sciences.
-Its ease of implementation, however, masks important underlying fragilities in the method.
-Namely, that unless the chains have converged to a truly stationary distribution, the draws generated are not faithful depictions of the posterior.
-In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that has received insufficient attention thus far.
-$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
-
-\section{Contributions}
-BL conceived of the original idea for $R^*$, carried out all analyses and wrote the original draft of the paper.
-AV reviewed the paper and made a host of recommendations: including suggesting a large range of new case studies on which to trial the method and also suggesting new visualisations for diagnosing how predictive performance depends on parameter quantiles; these collectively widened the scope of the original paper and substantially improved its quality.
-BL and AV reviewed and edited the draft of the paper.
-
-\bibliographystyle{chicago}
-\bibliography{bibliography}
-
-\beginsupplement
 

### 2020-09-03 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
corrected typos

@@
 The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters.
-We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing.
+We split each chain's draws into randomly divided training and testing sets: here, we use 70\% of draws for training and 30\% for testing.
 Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
@@
 In (d), we show that $R^*$ is able to differentiate between distributions with non-stationary target distributions and stationary ones.
-It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there insufficient draws.
+It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there are insufficient draws.
 

### 2020-09-04 src/s_eight_schools.Rmd
addressing andrew gelman's comments including transposing fig 12

@@
 g1 <- plot_r_star_quantiles(1, x, nsim=40, split_chains = F) +
-  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6))
+  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6)) +
+  theme(axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        axis.ticks.y = element_blank()) +
+  coord_flip()
 
 g2 <- plot_r_star_quantiles(2, x, nsim=40, split_chains = F) +
-  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6))
+  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6)) +
+  theme(axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.ticks.y = element_blank()) +
+  coord_flip()
 
@@
 library(gridExtra)
-f <- ecdf(x[, , 2])
+f <- ecdf(as.vector(x[, , 2]))
 p <- x[, , 2] %>% 
@@
           axis.text = element_text(colour="black", size=10),
-        legend.position = "none",
-        axis.title.x = element_blank()) +
+        legend.position = "none") +
   xlab("Iteration") +
-  ylab("Quantile") +
-  ylim(0, 1) +
-  coord_flip()
-ga <- ggMarginal(p, type="density", margins="x", groupColour = T,
-                 xparams = list(bw=0.05))
+  ylab(TeX("B. $\\tau$ quantile")) +
+  ylim(0, 1)
+ga <- ggMarginal(p, type="density", margins="y", groupColour = T,
+                 yparams = list(bw=0.05))
 
-f <- ecdf(x[, , 1])
+f <- ecdf(as.vector(x[, , 1]))
 p <- x[, , 1] %>% 
@@
         legend.position = "none",
-        axis.title.x = element_blank()) +
+        axis.title.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
   xlab("Iteration") +
-  ylab("Quantile") +
-  ylim(0, 1) +
-  coord_flip()
-gb <- ggMarginal(p, type="density", margins="x", groupColour = T,
-                 xparams = list(bw=0.05))
+  ylab(TeX("A. $\\mu$ quantile")) +
+  ylim(0, 1)
+gb <- ggMarginal(p, type="density", margins="y", groupColour = T,
+                 yparams = list(bw=0.05))
 
@@
 
-pdf("../output/eight_schools_r_star_quantiles.pdf", width = 6, height = 6)
-plot_grid(arrangeGrob(gb, top="A. mu"), arrangeGrob(ga, top="B. tau"), g1, g2,nrow = 2, rel_heights = c(1/2, 1/4))
+pdf("../output/eight_schools_r_star_quantiles1.pdf", width = 6, height = 4)
+plot_grid(gb, g1, ga, g2, nrow = 2, rel_widths = c(1, 0.4), rel_heights = c(0.9, 1))
 dev.off()

### 2020-09-04 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
addressing andrew gelman's comments including transposing fig 12

@@
 Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
-Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged.
+Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged \citep{gelman1992single}.
 Like furious bees trapped in a room of a house \citep{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
@@
 This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
-Rephrased, it should not be possible to predict the chain that \textit{caused} a draw.
+Rephrased, it should not be possible to predict which draws come from which chain.
 In this vein, we use a machine learning (ML) classifier to measure convergence.
@@
 \section{Method}\label{sec:method}
-If Markov chains have not mixed, it is possible to determine to which chain a draw belongs from its value.
+If Markov chains have not mixed, it is possible to guess (with more accuracy than chance) to which chain a draw belongs from its value.
 This is possible if there are differences in the sampling distribution for any dimension, $\theta$, in the target distribution (Fig.
@@
 %
-where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$.
+where $\epsilon_t\stackrel{i.i.d.}{\sim}\text{normal}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$.
 Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) standard deviation of the others.
@@
 \boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
-z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
+z_j\sim \text{normal}(0, 1), \text{ for } j = 1,2,...,250.
 \end{align}
@@
 \begin{align}
-a_j \sim  \N(0,1), \qquad
+a_j \sim  \text{normal}(0,1), \qquad
 b_j \sim  \text{Gamma}(0.5, 0.5), \qquad
@@
 
-Comparing our ``omniscient'' measure of convergence shown in Fig.
-\ref{fig:cauchy_convergence}A, with the various derived measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
+Comparing our measure of convergence that requires knowing the actual target distribution (\textit{quantile-$R^2$}; in Fig.
+\ref{fig:cauchy_convergence}A), with the various heuristic measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
 The rate at which these converge differs though, and $R^*$ (Fig.
-\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to our ``omniscient'' measure.
+\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to \textit{quantile-$R^2$}.
 
@@
 \begin{align*}
-\theta_j &\sim \N(\mu, \tau), \\
-y_j &\sim \N(\theta_j, \sigma_j).
+\theta_j &\sim \text{normal}(\mu, \tau), \\
+y_j &\sim \text{normal}(\theta_j, \sigma_j).
 \end{align*}
@@
 \begin{align*}
-\tilde{\theta}_j &\sim \N(0, 1), \\
+\tilde{\theta}_j &\sim \text{normal}(0, 1), \\
 \theta_j &= \mu + \tau \tilde{\theta}_j,\\
-y_j &\sim \N(\theta_j, \sigma_j).
+y_j &\sim \text{normal}(\theta_j, \sigma_j).
 \end{align*}
@@
 \ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
-In the top row of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
-Above the top panel, we show the marginal distributions for each chain.
-In the bottom row, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
-In essence, the top rows explain the variation in $R^*$ in the bottom panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
+In the left-hand column of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
+To the right of each trace plot, we show the marginal distributions for each chain.
+In the right-hand column, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
+In essence, the left-hand panels explain the variation in $R^*$ in the right-hand panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
 For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools_r_star_quantiles.pdf}}
-	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Column A shows plots for $\mu$; Column B for $\tau$.
-In each column, we show the path of the four individual chains above and 40 $R^*$ draws obtained using Algorithm \ref{alg:R_star_uncertainty} for each parameter quantile below.
-Above the top row, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
-Note that, in the bottom plots, jitter has been added to the data points.}
+	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Row A shows plots for $\mu$; Row B for $\tau$.
+In each row, we show the path of the four individual chains above and 40 $R^*$ draws obtained using Algorithm \ref{alg:R_star_uncertainty} for each parameter quantile below.
+To the right of each trace plot, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
+Note that, in the right-hand plots, jitter has been added to the data points.}
 	\label{fig:eight_schools_r_star_quantiles}
@@
 If so, this suggests that larger models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
+That said, it is possible to reduce the runtime for $R^*$ using thinned draws (although this risks losing chain idiosyncracies) and using a subset of dimensions (although this risks losing problematic dimensions).
+Indeed, in \S\ref{sec:multivariate_normal_250}, \S\ref{sec:cauchy} and \S\ref{sec:prostate}, we use these strategies and, nonetheless, find that $R^*$ provides a stringent measure of convergence.
 
@@
 BL and AV reviewed and edited the draft of the paper.
+
+\section{Acknowledgements}
+The authors would like to thank Andrew Gelman and Hanne Oberman for comments on previous drafts of the paper that were submitted to arXiv.
 

### 2020-09-04 write_up/arxiv_submission_aug_2020/ms_ba.tex
updated journal file

@@
 
+\section{Acknowledgements}
+The authors would like to thank Andrew Gelman and Hanne Oberman for comments on previous drafts of the paper that were submitted to arXiv.
+
 \bibliographystyle{chicago}

### 2020-09-29 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
started to address ba reviewer comments

--- 2020-09-04 write_up/arxiv_submission_aug_2020/ms_ba.tex
@@
 \usepackage{xr}
+\usepackage{xcolor}
 \externaldocument[]{supplementary_materials}
@@
 	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
-Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
+Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about \textcolor{red}{its performance in finite time}.
 The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
-Here, we introduce a new method for diagnosing convergence based on whether a machine learning classifier model can successfully discriminate the individual chains.
+Here, we introduce a new method for diagnosing convergence based on \textcolor{red}{how well} a machine learning classifier model can successfully discriminate the individual chains.
 We call this convergence measure $R^*$.
 In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
+Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; \textcolor{red}{instead it uses} all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
 Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
@@
 
-The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
+The most important determination of whether MCMC has worked is \textcolor{red}{how closely} the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
 MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions).
@@
 
-The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
+The \textcolor{red}{one} diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
 Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
@@
 This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
-Rephrased, it should not be possible to predict which draws come from which chain.
+\textcolor{red}{Rephrased, it should not be possible to predict which draws come from which chain}.
 In this vein, we use a machine learning (ML) classifier to measure convergence.
@@
 Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
-However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
-For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
+\textcolor{red}{However, the ML classifier we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.} For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
 For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
@@
 
-Comparing our measure of convergence that requires knowing the actual target distribution (\textit{quantile-$R^2$}; in Fig.
+\textcolor{red}{Comparing our measure of convergence that requires knowing the actual target distribution (\textit{quantile-$R^2$}; in Fig.
 \ref{fig:cauchy_convergence}A), with the various heuristic measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
 The rate at which these converge differs though, and $R^*$ (Fig.
-\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to \textit{quantile-$R^2$}.
+\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to \textit{quantile-$R^2$}.}
 
@@
 Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
-This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
+\textcolor{red}{This indicates the complementarity of $R^*$ to existing measures.}
 

### 2020-09-29 src/s_discrete.Rmd
started discrete example

---
title: "R Notebook"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

Create a discrete Markov chain
```{r}
f_simulate_markov_chain <- function(L, P, state_0){
  states <- vector(length=L)
  states[1] <- state_0
  for(i in 2:L){
    p  <- P[states[i - 1], ]
    states[i] <-  which(rmultinom(1, 1, p) == 1)
  }
  return(states)
}

P <- matrix(nrow = 4, ncol = 4)
P[1, ] <- c(0, 1/2, 1/2, 0)
P[2, ] <- c(1/2, 0, 1/3, 1/6)
P[3, ] <- c(1/4, 1/4, 1/4, 1/4)
P[4, ] <- c(1/2, 0, 0, 1/2)

states <- f_simulate_markov_chain(100, P, 1)
plot(states, type="l")
```

Create 3 Markov chains with same transition matrix; one with a slightly different one
```{r}
f_generate_four <- function(L, P0, P1, state_0){
  x <- matrix(nrow = L, ncol = 4)
  for(i in 1:3)
    x[, i] <- f_simulate_markov_chain(L, P0, state_0)
  x[, 4] <- f_simulate_markov_chain(L, P1, state_0)
  return(x)
}

P0 <- P
P1 <- matrix(nrow = 4, ncol = 4)
P1[1, ] <- c(0, 1/2, 1/2, 0)
P1[2, ] <- c(1/2, 0, 1/3, 1/6)
P1[3, ] <- c(1/4, 1/4, 1/4, 1/4)
P1[4, ] <- c(1/2, 0, 1/6, 1/3)

f_generate_four(1000, P0, P1, 1) %>% 
  as.data.frame() %>% 
  rename(chain_1=V1, chain_2=V2, chain_3=V3, chain_4=V4) %>% 
  mutate(iter=seq_along(chain_1)) %>% 
  melt(id.vars="iter") %>% 
  group_by(variable) %>% 
  summarise(state_1=mean(value==1),
            state_2=mean(value==2),
            state_3=mean(value==3),
            state_4=mean(value==4)) %>% 
  pivot_longer(cols=state_1:state_4) %>% 
  ggplot(aes(x=as.factor(name), fill=as.factor(variable), y=value)) +
  geom_col(position = position_dodge()) +
  xlab("State") +
  ylab("Frequency") +
  scale_fill_brewer("Chain", palette = "Dark2")
```

Generate replicates and calculate R* on these
```{r}
f_replicate <- function(L, P0, P1, state_0){
  temp <- f_generate_four(L, P0, P1, state_0)
  a_array <- array(dim=c(L, 4, 1))
  a_array[,,1] <- temp
  a_accuracy <- r_star(a_array)
  mon <- monitor_extra(split_data(a_array))
  return(list(r_star=a_accuracy, r_hat=mon$zfsRhat))
}

nreplicates <- 20
Ls <- c(1000, 2000, 5000, 10000)

# create P1 matrices
f_create_P1_example <- function(last_val=1/2){
  P1 <- matrix(nrow = 4, ncol = 4)
  P1[1, ] <- c(0, 1/2, 1/2, 0)
  P1[2, ] <- c(1/2, 0, 1/3, 1/6)
  P1[3, ] <- c(1/4, 1/4, 1/4, 1/4)
  P1[4, ] <- c(1/2, 0, (1 - 1/2 - last_val), last_val)
  return(P1)
}

last_vals <- c(0, 0.25, 0.5)
P1s <- map(last_vals, ~f_create_P1_example(.))

r_star_vals <- matrix(nrow=nreplicates * length(Ls) * length(P1s), ncol=4)
r_hat_vals <- matrix(nrow=nreplicates * length(Ls) * length(P1s), ncol=4)
counter <- 1
for(k in 1:length(P1s)){
  for(j in seq_along(Ls)){
    print(paste0("k=", k, ", j=", j))
    for(i in 1:nreplicates){
      temp <- f_replicate(Ls[j], P0, P1s[[k]], 1)
      r_star_vals[counter, ] <- c(k, Ls[j], i, temp$r_star)
      r_hat_vals[counter, ] <- c(k, Ls[j], i, temp$r_hat)
      counter <- counter + 1
    }
  }
}
colnames(r_hat_vals) <- c("P1", "sample size", "iter", "value")
colnames(r_star_vals) <- c("P1", "sample size", "iter", "value")
r_hat_vals <- as.data.frame(r_hat_vals) %>% 
  mutate(type="Split-Rhat")
r_star_vals <- as.data.frame(r_star_vals) %>% 
  mutate(type="R*")

saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/discrete_r_star_hat.rds")
```

Plot
```{r}
a_sum <- r_star_vals %>%
  group_by(P1, `sample size`, type) %>%
  summarise(lower=quantile(value, 0.025),
         upper=quantile(value, 0.975),
         middle=quantile(value, 0.5))
r_star_vals %>% 
  left_join(a_sum) %>% 
  ggplot(aes(x=as.factor(`sample size`), y=value)) +
  geom_jitter(width = 0.2, colour="grey") +
  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
  facet_grid(vars(P1)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_y_continuous(limits=c(0.9, 1.2))
```

### 2020-09-29 src/r_star_monitor.R

@@ -120,6 +120,11 @@ r_star_dist <- function(x, split_chains=T, training_percent=0.7, caret_default=N
   r <- m_flattened %>% 
     mutate(chain=as.factor(chain))
   
+  # if only 1 param, add in a column of random noise since gbm requires >1 dims
+  if(nparams==1)
+    r <- r %>% 
+    mutate(V_new=rnorm(nrow(r)))
+  
   rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]

### 2020-09-29 src/s_discrete.Rmd
added discrete example

@@
 P[3, ] <- c(1/4, 1/4, 1/4, 1/4)
-P[4, ] <- c(1/2, 0, 0, 1/2)
+P[4, ] <- c(0, 1, 0, 0)
 
@@
 P1[2, ] <- c(1/2, 0, 1/3, 1/6)
-P1[3, ] <- c(1/4, 1/4, 1/4, 1/4)
-P1[4, ] <- c(1/2, 0, 1/6, 1/3)
+P1[3, ] <- c(1, 0, 0, 0)
+P1[4, ] <- c(1, 0, 0, 0)
 
@@
 
-nreplicates <- 20
+nreplicates <- 40
 Ls <- c(1000, 2000, 5000, 10000)
@@
 # create P1 matrices
-f_create_P1_example <- function(last_val=1/2){
+f_create_P1_example <- function(per_along){
   P1 <- matrix(nrow = 4, ncol = 4)
@@
   P1[2, ] <- c(1/2, 0, 1/3, 1/6)
-  P1[3, ] <- c(1/4, 1/4, 1/4, 1/4)
-  P1[4, ] <- c(1/2, 0, (1 - 1/2 - last_val), last_val)
+  v_diff <- c(1, 0, 0, 0) - c(1/4, 1/4, 1/4, 1/4)
+  P1[3, ] <- c(1/4, 1/4, 1/4, 1/4) + v_diff * per_along
+  v_diff <- c(1, 0, 0, 0) - c(0, 1, 0, 0)
+  P1[4, ] <- c(0, 1, 0, 0) + v_diff * per_along
   return(P1)
@@
 
-last_vals <- c(0, 0.25, 0.5)
+last_vals <- c(0, 0.5, 1)
 P1s <- map(last_vals, ~f_create_P1_example(.))
@@
 ```{r}
+temp <- readRDS("../data/discrete_r_star_hat.rds")
+r_star_vals <- temp$r_star %>% 
+  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
+  mutate(`sample size`=as.character(`sample size`)) %>% 
+  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
+  mutate(`sample size`=as.factor(`sample size`)) %>% 
+  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
 a_sum <- r_star_vals %>%
@@
          middle=quantile(value, 0.5))
-r_star_vals %>% 
+g1 <- r_star_vals %>% 
   left_join(a_sum) %>% 
@@
   geom_hline(yintercept = 1, linetype=2) +
-  scale_y_continuous(limits=c(0.9, 1.2))
+  scale_y_continuous(limits=c(NA, NA)) +
+  xlab("Sample size") +
+  ylab(TeX("$R*$")) +
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.y = element_text(colour="black", size=16),
+        legend.position = "none",
+        title = element_text(colour="black", size=16)) +
+  ggtitle("A.")
+r_star_vals <- temp$r_hat %>% 
+  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
+  mutate(`sample size`=as.character(`sample size`)) %>% 
+  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
+  mutate(`sample size`=as.factor(`sample size`)) %>% 
+  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
+a_sum <- r_star_vals %>%
+  group_by(P1, `sample size`, type) %>%
+  summarise(lower=quantile(value, 0.025),
+         upper=quantile(value, 0.975),
+         middle=quantile(value, 0.5))
+g2 <- r_star_vals %>% 
+  left_join(a_sum) %>% 
+  ggplot(aes(x=as.factor(`sample size`), y=value)) +
+  geom_jitter(width = 0.2, colour="grey") +
+  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
+  facet_grid(vars(P1)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_y_continuous(limits=c(NA, NA)) +
+  xlab("Sample size") +
+  ylab(TeX("Split-$\\hat{R}$")) +
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.y = element_text(colour="black", size=16),
+        legend.position = "none",
+        title = element_text(colour="black", size=16)) +
+  ggtitle("B.")
+
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  library(grid)
+
+  # Make a list from the ... arguments and plotlist
+  plots <- c(list(...), plotlist)
+
+  numPlots = length(plots)
+
+  # If layout is NULL, then use 'cols' to determine layout
+  if (is.null(layout)) {
+    # Make the panel
+    # ncol: Number of columns of plots
+    # nrow: Number of rows needed, calculated from # of cols
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                    ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+
+ if (numPlots==1) {
+    print(plots[[1]])
+
+  } else {
+    # Set up the page
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+
+    # Make each plot, in the correct location
+    for (i in 1:numPlots) {
+      # Get the i,j matrix positions of the regions that contain this subplot
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+  }
+}
+
+pdf("../output/discrete.pdf", width = 12, height = 8)
+multiplot(g1, g2, cols = 1)
+dev.off()
 ```
+
+Plotting stationary distributions
+```{r}
+m_stationary <- matrix(c(11/46, 15/46, 14/46, 6/46, 71/198, 17/66, 10/33, 8/99, 4/9, 2/9, 8/27, 1/27), nrow = 3, byrow = T) %>% 
+  as.data.frame() %>% 
+  mutate(matrix=c("P1", "P2", "P3"))
+colnames(m_stationary)[1:4] <- c("1", "2", "3", "4")
+g <- m_stationary %>% 
+  melt(id.vars="matrix") %>% 
+  ggplot(aes(x=as.factor(variable), y=value, fill=as.factor(matrix))) +
+  geom_col(position = position_dodge()) +
+  xlab("State") +
+  ylab("Probability") +
+  scale_fill_brewer("Transition\nprobability\nmatrix", palette = "Dark2", labels=unname(TeX(c("P_1", "P_2", "P_3")))) +
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.x = element_text(colour="black", size=16),
+        legend.text = element_text(colour="black", size=16),
+        legend.title = element_text(colour="black", size=16),
+        legend.position = c(0.85, 0.7))
+ggsave("../output/discrete_stationary.pdf", g, width = 10, height = 6)
+```

### 2020-09-29 write_up/arxiv_submission_aug_2020/supplementary_materials.tex
added discrete example

@@
 
+\section{Discrete distribution}\label{sec:discrete}
+In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a target distribution with a discrete dimension.
+To do so, we generate draws from four discrete Markov chains, following a 1st order Markov process with transition probabilities,
+%
+\begin{equation}
+p_{ij} = \text{Pr}(X_{n+1}=j|X_n=i),
+\end{equation}
+%
+where $p_{ij}$ is the probability of transitioning from state $i\in\{1,2,3,4\}$ at time $n$ to state $j\in\{1,2,3,4\}$ at time $n+1$, and $X(0)=1$.
+The target distribution is, hence, the stationary distribution of this process.
+
+In three of the four chains, we generate draws using a transition probability matrix:
+%
+\begin{gather}
+\boldsymbol{P}
+=
+\begin{bmatrix}
+0 & 1/2 & 1/2 & 0\\
+1/2 & 0 & 1/3 & 1/6\\
+1/4 & 1/4 & 1/4 & 1/4\\
+0 & 1 & 0 & 0
+\end{bmatrix},
+\end{gather}
+%
+which implies a stationary target distribution with the following probability simplex $\pi_1 = (11/46, 15/46, 14/46, 6/46)$.
+
+For the fourth chain, we specify differing transition probability matrices in each of three sets of results.
+In the first, we assume $\boldsymbol{P}_1=\boldsymbol{P}$ (i.e.
+the same as used for the other three chains); in the second and third cases, we assume this transition matrix is given by:
+%
+\begin{gather}
+\boldsymbol{P}_2
+=
+\begin{bmatrix}
+0 & 1/2 & 1/2 & 0\\
+1/2 & 0 & 1/3 & 1/6\\
+5/8 & 1/8 & 1/8 & 1/8\\
+1/2 & 1/2 & 0 & 0
+\end{bmatrix}\quad
+\boldsymbol{P}_3
+=
+\begin{bmatrix}
+0 & 1/2 & 1/2 & 0\\
+1/2 & 0 & 1/3 & 1/6\\
+1 & 0 & 0 & 0\\
+1 & 0 & 0 & 0
+\end{bmatrix}.
+\end{gather}
+%
+These correspond to stationary probability simplices $\pi_2 = (71/198, 17/66, 10/33, 8/99)$ and $\pi_3 = (4/9, 2/9, 8/27, 1/27)$.
+The three stationary distributions corresponding to each transition probability matrices are shown in Fig.
+\ref{fig:discrete_stationary}.
+This illustrates that there is an ordering among the stationary distributions $\pi_1\sim\pi_2\sim\pi_3$, meaning $\pi_1$ is most similar to $\pi_2$, and $\pi_3$ is most similar to $\pi_2$.
+Put another way, $\pi_3$ is most different from $p_1$.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary.pdf}}
+	\caption{\textbf{Discrete distribution example: stationary distributions.} The three colours illustrate the stationary distributions of the Markov chains corresponding to the matrices $P_1$, $P_2$ and $P_3$ defined in \S\ref{sec:discrete}.}
+	\label{fig:discrete_stationary}
+\end{figure}
+
+For each of these three cases, we consider a range of sample sizes across each of the four chains: 1000, 2000, 5000 and 10,000.
+At each combination of transition probability matrix (for the fourth chain) and sample size, we perform 40 replicates.
+In each replicate, we calculate both $R^*$ and rank-normalised split-$\widehat{R}$.
+The results of these experiments are shown in Fig.
+\ref{fig:discrete}.
+In the horizontal axes of both panels, we show the sample size.
+The vertical axis indicates the value of $R^*$ and split-$\widehat{R}$ in panels A and B, respectively.
+The stacked sub-panels within A and B indicate the transition matrix used for the fourth chain.
+Since $\boldsymbol{P}_1=\boldsymbol{P}$, the stationary distribution for the other three chains, both diagnostics should tend towards indicating convergence in these cases, and, for both statistics, this was typically the case.
+For $R^*$, the replicates were centred on $R^*=1$, although the stochasticity of replicates meant that, in some cases, $R^*>1$.
+Whereas, $\widehat{R}<1$ in all cases.
+Considering the simulations using $\boldsymbol{P}_2$ for the fourth chain transition probability matrix, $R^*>1$ across a majority of replicates, and this majority increased with sample size.
+In this case, $\widehat{R}<1$ for all replicates across the various sample sizes.
+For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; as for the $\boldsymbol{P}_2$ case, $\widehat{R}<1$, indicating convergence.
+
+This example hints that $R^*$ may be particularly adept at detecting poor convergence in small discrete parameter spaces.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{discrete.pdf}}
+	\caption{\textbf{Discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations.
+The panels within A.
+and B.
+indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$).
+In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates.
+In panel B, we show the same, but for split-$\widehat{R}$.
+The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
+	\label{fig:discrete}
+\end{figure}
+
 \end{document}

### 2020-09-29 src/s_discrete.Rmd
started higher dimensional example

@@
 library(gbm)
+library(MCMCpack)
 options(mc.cores=4)
@@
 ```
+
+# Higher dimension discrete Markov chain
+```{r}
+f_construct_random_matrix <- function(n_dim, alpha){
+  P <- rdirichlet(n_dim, alpha)
+  return(P)
+}
+
+P0 <- f_construct_random_matrix(20, rep(1, 20))
+P1 <- f_construct_random_matrix(20, rep(1, 20))
+
+saveRDS(list(P0=P0, P1=P1), "../data/discrete_higherd_matrices.rds")
+# show first few stationary frequencies for the chains
+g <- f_generate_four(100000, P0, P1, 1) %>% 
+  as.data.frame() %>% 
+  rename(chain_1=V1, chain_2=V2, chain_3=V3, chain_4=V4) %>% 
+  mutate(iter=seq_along(chain_1)) %>% 
+  melt(id.vars="iter") %>% 
+  group_by(variable) %>% 
+  summarise("1"=mean(value==1),
+            "2"=mean(value==2),
+            "3"=mean(value==3),
+            "4"=mean(value==4)) %>% 
+  pivot_longer(cols="1":"4") %>% 
+  ggplot(aes(x=as.factor(name), fill=as.factor(variable), y=value)) +
+  geom_col(position = position_dodge()) +
+  xlab("State") +
+  ylab("Frequency") +
+  scale_fill_brewer("Chain", palette = "Dark2") +
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.y = element_text(colour="black", size=16),
+        title = element_text(colour="black", size=16),
+        legend.text = element_text(colour="black", size=16))
+ggsave("../output/discrete_stationary_higherd.pdf", g, width = 10, height = 6)
+```
+
+Do same simulations as before but with this higher dimensional case
+```{r}
+nreplicates <- 40
+Ls <- c(1000, 2000, 5000, 10000)
+P1s <- list(P0, P1)
+
+r_star_vals <- matrix(nrow=nreplicates * length(Ls) * length(P1s), ncol=4)
+r_hat_vals <- matrix(nrow=nreplicates * length(Ls) * length(P1s), ncol=4)
+counter <- 1
+for(k in 1:length(P1s)){
+  for(j in seq_along(Ls)){
+    print(paste0("k=", k, ", j=", j))
+    for(i in 1:nreplicates){
+      temp <- f_replicate(Ls[j], P0, P1s[[k]], 1)
+      r_star_vals[counter, ] <- c(k, Ls[j], i, temp$r_star)
+      r_hat_vals[counter, ] <- c(k, Ls[j], i, temp$r_hat)
+      counter <- counter + 1
+    }
+  }
+}
+colnames(r_hat_vals) <- c("P1", "sample size", "iter", "value")
+colnames(r_star_vals) <- c("P1", "sample size", "iter", "value")
+r_hat_vals <- as.data.frame(r_hat_vals) %>% 
+  mutate(type="Split-Rhat")
+r_star_vals <- as.data.frame(r_star_vals) %>% 
+  mutate(type="R*")
+
+saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/discrete_r_star_hat_higherd.rds")
+
+temp <- readRDS("../data/discrete_r_star_hat_higherd.rds")
+r_star_vals <- temp$r_star %>% 
+  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
+  mutate(`sample size`=as.character(`sample size`)) %>% 
+  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
+  mutate(`sample size`=as.factor(`sample size`)) %>% 
+  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
+a_sum <- r_star_vals %>%
+  group_by(P1, `sample size`, type) %>%
+  summarise(lower=quantile(value, 0.025),
+         upper=quantile(value, 0.975),
+         middle=quantile(value, 0.5))
+g1 <- r_star_vals %>% 
+  left_join(a_sum) %>% 
+  ggplot(aes(x=as.factor(`sample size`), y=value)) +
+  geom_jitter(width = 0.2, colour="grey") +
+  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
+  facet_grid(vars(P1)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_y_continuous(limits=c(NA, NA)) +
+  xlab("Sample size") +
+  ylab(TeX("$R*$")) +
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.y = element_text(colour="black", size=16),
+        legend.position = "none",
+        title = element_text(colour="black", size=16)) +
+  ggtitle("A.")
+r_star_vals <- temp$r_hat %>% 
+  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
+  mutate(`sample size`=as.character(`sample size`)) %>% 
+  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
+  mutate(`sample size`=as.factor(`sample size`)) %>% 
+  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
+a_sum <- r_star_vals %>%
+  group_by(P1, `sample size`, type) %>%
+  summarise(lower=quantile(value, 0.025),
+         upper=quantile(value, 0.975),
+         middle=quantile(value, 0.5))
+g2 <- r_star_vals %>% 
+  left_join(a_sum) %>% 
+  ggplot(aes(x=as.factor(`sample size`), y=value)) +
+  geom_jitter(width = 0.2, colour="grey") +
+  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
+  facet_grid(vars(P1)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_y_continuous(limits=c(NA, NA)) +
+  xlab("Sample size") +
+  ylab(TeX("Split-$\\hat{R}$")) +
+  theme(axis.text = element_text(colour="black", size=16),
+        axis.title = element_text(colour="black", size=16),
+        strip.text.y = element_text(colour="black", size=16),
+        legend.position = "none",
+        title = element_text(colour="black", size=16)) +
+  ggtitle("B.")
+
+pdf("../output/discrete_higherd.pdf", width = 12, height = 8)
+multiplot(g1, g2, cols = 1)
+dev.off()
+```

### 2020-09-29 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
started higher dimensional example

--- 2020-09-29 write_up/arxiv_submission_aug_2020/supplementary_materials.tex
@@
 \usepackage{natbib}
+\usepackage{xcolor}
 
 \usepackage{xr}
-\externaldocument[]{ms_ba}
+\externaldocument[]{ms_ba_marked}
 
@@
 
+\color{red}
 \section{Discrete distribution}\label{sec:discrete}
@@
 %
-where $p_{ij}$ is the probability of transitioning from state $i\in\{1,2,3,4\}$ at time $n$ to state $j\in\{1,2,3,4\}$ at time $n+1$, and $X(0)=1$.
-The target distribution is, hence, the stationary distribution of this process.
+where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$. The target distribution is, hence, the stationary distribution of this process.
 
+\subsection{Small state-space}\label{sec:discrete_small}
 In three of the four chains, we generate draws using a transition probability matrix:
@@
 
+\subsection{Large state-space}\label{sec:discrete_large}
+We next consider a larger discrete parameter space: a univariate distribution with 20 states.
+As in \S\ref{sec:discrete_small}, we generate draws from three Markov chains that share a common transition probability matrix, $\boldsymbol{P}$.
+For the fourth (and final) chain, we consider two cases: one where it shares the same matrix, $\boldsymbol{P}_1=\boldsymbol{P}$; the other, when its Markov process has a different matrix, $\boldsymbol{P}_2=\boldsymbol{P}$.
+Rather than write down a given $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we generated these stochastically: by random sampling from a Dirichlet process with shape vectors consisting of ones  to generate each of its rows.
+
+Include plot of simulated frequencies (already generated).
+
+\color{black}
+
 \end{document}

### 2020-10-01 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
started higher dimensional example
first draft of discrete distributions section

@@
 \usepackage{xcolor}
-\externaldocument[]{supplementary_materials}
+\externaldocument[]{supplementary_materials_marked}
 
@@
 \section{Acknowledgements}
-The authors would like to thank Andrew Gelman and Hanne Oberman for comments on previous drafts of the paper that were submitted to arXiv.
+The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.
 

### 2020-10-01 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
first draft of discrete distributions section

--- 2020-09-29 write_up/arxiv_submission_aug_2020/supplementary_materials.tex
@@
 
+\color{red}
 \section{Discrete distribution}\label{sec:discrete}
-In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a target distribution with a discrete dimension.
+In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a univariate discrete target distribution.
 To do so, we generate draws from four discrete Markov chains, following a 1st order Markov process with transition probabilities,
@@
 %
-where $p_{ij}$ is the probability of transitioning from state $i\in\{1,2,3,4\}$ at time $n$ to state $j\in\{1,2,3,4\}$ at time $n+1$, and $X(0)=1$.
-The target distribution is, hence, the stationary distribution of this process.
+where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$. The target distribution is, hence, the stationary distribution of this process.
 
-In three of the four chains, we generate draws using a transition probability matrix:
+\subsection{Small state-space}\label{sec:discrete_small}
+We first consider a univariate discrete distribution with four states. In three of the four chains, we generate draws using a transition probability matrix:
 %
@@
 1/2 & 1/2 & 0 & 0
-\end{bmatrix}\quad
+\end{bmatrix},\quad
 \boldsymbol{P}_3
@@
 %
-These correspond to stationary probability simplices $\pi_2 = (71/198, 17/66, 10/33, 8/99)$ and $\pi_3 = (4/9, 2/9, 8/27, 1/27)$.
-The three stationary distributions corresponding to each transition probability matrices are shown in Fig.
+These correspond to stationary probability simplices:
+%
+\begin{equation}
+\pi_2 = (71/198, 17/66, 10/33, 8/99), \quad \pi_3 = (4/9, 2/9, 8/27, 1/27).
+\end{equation}
+%
+The three stationary distributions corresponding to each transition probability matrix are shown in Fig.
 \ref{fig:discrete_stationary}.
@@
 In this case, $\widehat{R}<1$ for all replicates across the various sample sizes.
-For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; as for the $\boldsymbol{P}_2$ case, $\widehat{R}<1$, indicating convergence.
+For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; as for the previous case, $\widehat{R}<1$, indicating convergence.
 
@@
 
+\subsection{Large state-space}\label{sec:discrete_large}
+We next consider a larger discrete parameter space: a univariate distribution with 20 states.
+As in \S\ref{sec:discrete_small}, we generate draws from three discrete Markov chains that share a common transition probability matrix, $\boldsymbol{P}$.
+For the fourth (and final) chain, we consider two cases: one where it shares the same matrix, $\boldsymbol{P}_1=\boldsymbol{P}$; the other, when its Markov process has a different matrix, $\boldsymbol{P}_2$.
+Rather than write down a given $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we generated these stochastically: for each of the rows, we randomly sampled from from a Dirichlet process with shape vectors consisting of ones (of length 20).
+
+To illustrate the differences between the two transition probability matrices, $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we simulate 100,000 draws from each Markov process.
+In Fig.
+\ref{fig:discrete_stationary_higherd}, we show the marginal distribution obtained across all four chains: where the first three used $\boldsymbol{P}$, and the last used $\boldsymbol{P}_2$.
+In this plot, we show only the first four states of the 20-state model.
+The differences in marginal distributions between the first three chains and the last indicate that there are substantial differences in the target distributions.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary_higherd.pdf}}
+	\caption{\textbf{Large state-space discrete distribution example: approximate stationary distribution.} The distribution over the first four states of a 20-state discrete model achieved after drawing 100,000 samples from each of four Markov chains.
+Here, the first three chains shared the same transition probability matrix; the last chain used a different matrix.}
+	\label{fig:discrete_stationary_higherd}
+\end{figure}
+
+To probe the ability of $R^*$ and $\widehat{R}$ to detect poor convergence, we repeat the same exercise as in \S\ref{sec:discrete_small} although, here, we consider the new $\boldsymbol{P}_1$ and $\boldsymbol{P}_2$ cases.
+In Fig.
+\ref{fig:discrete_higherd}, we show the results of this analysis.
+This shows that $R^*\sim 1$ for the $\boldsymbol{P}_1$ case, correctly indicating no convergence issues; for the $\boldsymbol{P}_2$ case, $R^*>1$ across all replicates at sample sizes of 5000 and above.
+By contrast, $\widehat{R}$ struggles to differentiate between the $\boldsymbol{P}_1$ and $\boldsymbol{P}_2$ cases: in both instances, it signifies convergence.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{discrete_higherd.pdf}}
+	\caption{\textbf{Large state-space discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations.
+The panels within A.
+and B.
+indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$).
+In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates.
+In panel B, we show the same, but for split-$\widehat{R}$.
+The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
+	\label{fig:discrete_higherd}
+\end{figure}
+
+Collectively, the results of \S\ref{sec:discrete_small} and \S\ref{sec:discrete_large} hint that $R^*$ may be more reliable at detecting poor convergence than $\widehat{R}$, although further empirical investigation is needed.
+
+\color{black}
+
 \end{document}

### 2020-10-01 src/s_discrete.Rmd
started work on ml comparison

@@
 ---
-title: "R Notebook"
+title: "Discrete target distribution examples"
 output: html_notebook
@@
 
-P0 <- f_construct_random_matrix(20, rep(1, 20))
-P1 <- f_construct_random_matrix(20, rep(1, 20))
+# P0 <- f_construct_random_matrix(20, rep(1, 20))
+# P1 <- f_construct_random_matrix(20, rep(1, 20))
 
 saveRDS(list(P0=P0, P1=P1), "../data/discrete_higherd_matrices.rds")
+temp <- readRDS("../data/discrete_higherd_matrices.rds")
+P0 <- temp$P0
+P1 <- temp$P1
 # show first few stationary frequencies for the chains
@@
   pivot_longer(cols="1":"4") %>% 
+  mutate(variable=if_else(variable=="chain_1", "1", if_else(variable=="chain_2", "2", if_else(variable=="chain_3", "3", "4")))) %>% 
   ggplot(aes(x=as.factor(name), fill=as.factor(variable), y=value)) +

### 2020-10-01 src/s_ml_comparison.Rmd
started work on ml comparison

---
title: "ML model comparison"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

# AR1 model
```{r}
f_ar1 <- function(rho, sigma, L){
  x <- vector(length = L)
  x[1] <- rnorm(1, 0, sd=sigma)
  for(i in 2:L)
    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
  return(x)
}

# Generates three chains with same var; one with a different var
f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
  x <- matrix(nrow = L, ncol = 4)
  for(i in 1:3)
    x[, i] <- f_ar1(rho, sigma, L)
  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
  x[, 4] <- z
  return(x)
}

r_star_ml_variety <- function(x, split_chains=T, training_percent=0.7, method_="gbm", caretGrid=NULL){

  if(split_chains)
    x <- split_data(x)

  nparams <- dim(x)[3]
  nchains <- dim(x)[2]
  niter <- dim(x)[1]
  m_flattened <- matrix(nrow = nrow(x) * nchains,
                        ncol = (nparams + 1))
  k <- 1
  for(i in 1:nchains){
    for(j in 1:nrow(x)){
      m_flattened[k, 1:nparams] <- x[j, i, ]
      m_flattened[k, (nparams + 1)] <- i
      k <- k + 1
    }
  }
  m_flattened <- m_flattened %>% 
    as.data.frame()
  colnames(m_flattened)[nparams + 1] <- "chain"
  r <- m_flattened %>% 
    mutate(chain=as.factor(chain))
  # if only 1 param, add in a column of random noise since gbm requires >1 dims
  if(nparams==1)
    r <- r %>% 
      mutate(V_new=rnorm(nrow(r)))

  rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
  training_data <- r[rand_samples, ]
  testing_data <- r[-rand_samples, ]

  if(!is.null(caretGrid))
    gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_,
                   tuneGrid = caretGrid)
  else
    gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_)
  plda <- predict(object=gbmFit, newdata=testing_data)
  a_accuracy <- 
    tibble(predicted=plda, actual=testing_data$chain) %>%
    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
    summarise(mean(correct)) %>% 
    pull()
  return(a_accuracy * n_distinct(training_data$chain))
}

f_replicate_ml_variety <- function(x, method_, caretGrid_=NULL){
  temp <- x
  a_array <- array(dim=c(2000, 4, 1))
  a_array[,,1] <- temp
  a_accuracy <- r_star_ml_variety(a_array, method_=method_, caretGrid = caretGrid_)
  return(a_accuracy)
}

f_replicate_rhat <- function(x){
  temp <- x
  a_array <- array(dim=c(2000, 4, 1))
  a_array[,,1] <- temp
  mon <- monitor_extra(split_data(a_array))
  a_accuracy <- mon$zfsRhat
  return(a_accuracy)
}

f_replicate_ml_variety_gridded <- function(x, method_="gbm", caretGrids_){
  n_things <- nrow(caretGrids_)
  v_accuracy <- length(n_things)
  for(i in 1:n_things){
    caretGrid <- caretGrids_[i, ]
    v_accuracy[i] <- f_replicate_ml_variety(x, method_, caretGrid)
  }
  return(list(a_accuracy=max(v_accuracy), hyper=caretGrids_[which.max(v_accuracy),]))
}

tunegrid_rf <- tibble(mtry = 1:2)
tunegrid_gbm <- expand.grid(interaction.depth=c(3), 
                             n.trees = 50,
                             shrinkage=c(0.1),
                             n.minobsinnode=10)
tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
tunegrid_svm <- tibble(C = c(0.25, 0.5, 0.75))
tunegrid_nnet <- tibble(size=c(2, 3, 4, 5), decay=c(0.5, 0.5, 0.5, 0.5))
niterates <- 50
m_res <- matrix(nrow=niterates, ncol = 6)
l_rf_hyper <- vector(length = niterates, mode="list")
l_knn_hyper <- vector(length = niterates, mode="list")
l_svm_hyper <- vector(length = niterates, mode="list")
l_nnet_hyper <- vector(length = niterates, mode="list")
for(i in 1:niterates){
  print(i)
  x <- f_generate_lower_var_four(1/3, 0.3, 1, 2000)
  m_res[i, 1] <- f_replicate_ml_variety(x, "gbm", tunegrid_gbm)
  m_res[i, 2] <- f_replicate_rhat(x)
  temp <- f_replicate_ml_variety_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
  l_rf_hyper[[i]] <- temp$hyper
  m_res[i, 3] <- temp$a_accuracy
  temp <- f_replicate_ml_variety_gridded(x, "knn", tunegrid_knn)
  l_knn_hyper[[i]] <- temp$hyper
  m_res[i, 4] <- temp$a_accuracy
  temp <- f_replicate_ml_variety_gridded(x, "svmLinear", tunegrid_svm)
  l_svm_hyper[[i]] <- temp$hyper
  m_res[i, 5] <- temp$a_accuracy
  temp <- f_replicate_ml_variety_gridded(x, "nnet", tunegrid_nnet)
  l_nnet_hyper[[i]] <- temp$hyper
  m_res[i, 6] <- temp$a_accuracy
}

colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "nnet")
m_res1 <- m_res %>%
  as.data.frame() %>% 
  mutate(iter=seq_along(gbm))
saveRDS(m_res1, "../data/ml_comparison.rds")
m_res1 <- readRDS("../data/ml_comparison.rds")
g <- m_res1 %>% 
  dplyr::select(-one_of("rhat")) %>% 
  melt(id.vars="iter") %>% 
  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value)) +
  geom_boxplot() +
  xlab("ML model") +
  ylab("R*") +
  theme(text=element_text(colour="black", size=14))
ggsave("../output/ml_comparison_ar1.pdf", g, width = 12, height = 8)
```

# Eight schools model
```{r}
source("eight_schools.data.R")
eight_schools <- list(J=J, y=y, sigma=sigma)
model_cp <- stan_model("eight_schools_cp.stan")
model_ncp <- stan_model("eight_schools_ncp.stan")

fit_cp <- sampling(
  model_cp, data = eight_schools,
  iter = 4000, chains = 4, seed = 483892929, refresh = 0,
  control = list(adapt_delta = 0.95)
)

fit_ncp <- sampling(
  model_ncp, data = eight_schools,
  iter = 4000, chains = 4, seed = 483892929, refresh = 0,
  control = list(adapt_delta = 0.95)
)

x_cp <- rstan::extract(fit_cp, permuted=F)
x_ncp <- rstan::extract(fit_ncp, permuted=F)
```

Try out different ML models
```{r}
f_replicate_ml_simple <- function(x, method_, caretGrid_=NULL){
  a_array <- x
  a_accuracy <- r_star_ml_variety(a_array, method_=method_, caretGrid = caretGrid_)
  return(a_accuracy)
}

f_replicate_rhat_simple <- function(x){
  a_array <- x
  mon <- monitor_extra(split_data(a_array))
  a_accuracy <- mon$zfsRhat
  return(a_accuracy)
}

f_replicate_ml_simple_gridded <- function(x, method_="gbm", caretGrids_){
  n_things <- nrow(caretGrids_)
  v_accuracy <- length(n_things)
  for(i in 1:n_things){
    caretGrid <- caretGrids_[i, ]
    v_accuracy[i] <- f_replicate_ml_simple(x, method_, caretGrid)
  }
  return(list(a_accuracy=max(v_accuracy), hyper=caretGrids_[which.max(v_accuracy),]))
}

m_res <- matrix(nrow=2, ncol = 6)
l_rf_hyper <- vector(length = 2, mode="list")
l_knn_hyper <- vector(length = 2, mode="list")
l_svm_hyper <- vector(length = 2, mode="list")
l_nnet_hyper <- vector(length = 2, mode="list")
x <- x_cp
i <- 1
m_res[i, 1] <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
m_res[i, 2] <- max(f_replicate_rhat_simple(x))
temp <- f_replicate_ml_simple_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
l_rf_hyper[[i]] <- temp$hyper
m_res[i, 3] <- temp$a_accuracy
temp <- f_replicate_ml_simple_gridded(x, "knn", tunegrid_knn)
l_knn_hyper[[i]] <- temp$hyper
m_res[i, 4] <- temp$a_accuracy
temp <- f_replicate_ml_simple_gridded(x, "svmLinear", tunegrid_svm)
l_svm_hyper[[i]] <- temp$hyper
m_res[i, 5] <- temp$a_accuracy
temp <- f_replicate_ml_simple_gridded(x, "nnet", tunegrid_nnet)
l_nnet_hyper[[i]] <- temp$hyper
m_res[i, 6] <- temp$a_accuracy
x <- x_ncp
i <- 2
m_res[i, 1] <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
m_res[i, 2] <- max(f_replicate_rhat_simple(x))
temp <- f_replicate_ml_simple_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
l_rf_hyper[[i]] <- temp$hyper
m_res[i, 3] <- temp$a_accuracy
temp <- f_replicate_ml_simple_gridded(x, "knn", tunegrid_knn)
l_knn_hyper[[i]] <- temp$hyper
m_res[i, 4] <- temp$a_accuracy
temp <- f_replicate_ml_simple_gridded(x, "svmLinear", tunegrid_svm)
l_svm_hyper[[i]] <- temp$hyper
m_res[i, 5] <- temp$a_accuracy
temp <- f_replicate_ml_simple_gridded(x, "nnet", tunegrid_nnet)
l_nnet_hyper[[i]] <- temp$hyper
m_res[i, 6] <- temp$a_accuracy

colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "nnet")
m_res <- m_res %>%
  as.data.frame() %>% 
  mutate(model=c("cp", "ncp"))
# saveRDS(m_res, "../data/ml_comparison_8_schools.rds")
m_res %>% 
  dplyr::select(-rhat) %>% 
  melt(id.vars="model") %>% 
  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value, colour=as.factor(model))) +
  geom_point(size=4) +
  xlab("ML model") +
  scale_color_brewer("8 schools model", palette = "Dark2") +
  ylab("R*") +
  theme(text=element_text(colour="black", size=14))
# ggsave("../output/ml_comparison_8_schools.pdf", width = 12, height = 8)

x_ncp[,,11] %>%
  as.data.frame() %>%
  mutate(iter=seq(1,2000,1)) %>%
  melt(id.vars="iter") %>%
  ggplot(aes(x=value, fill=as.factor(variable))) +
  stat_density(position="identity", alpha=0.2)
```

### 2020-10-03 src/s_ml_comparison.Rmd
Update s_ml_comparison.Rmd

@@
   temp <- x
-  a_array <- array(dim=c(2000, 4, 1))
+  a_array <- array(dim=c(nrow(x), 4, 1))
   a_array[,,1] <- temp
@@
 tunegrid_nnet <- tibble(size=c(2, 3, 4, 5), decay=c(0.5, 0.5, 0.5, 0.5))
+tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
 niterates <- 50
-m_res <- matrix(nrow=niterates, ncol = 6)
+m_res <- matrix(nrow=niterates, ncol = 7)
 l_rf_hyper <- vector(length = niterates, mode="list")
@@
 l_nnet_hyper <- vector(length = niterates, mode="list")
+l_multinom_hyper <- vector(length = niterates, mode="list")
 for(i in 1:niterates){
@@
   m_res[i, 6] <- temp$a_accuracy
+  temp <- f_replicate_ml_variety_gridded(x, "multinom", tunegrid_multinom)
+  l_nnet_hyper[[i]] <- temp$hyper
+  m_res[i, 7] <- temp$a_accuracy
 }
@@
   model_cp, data = eight_schools,
-  iter = 4000, chains = 4, seed = 483892929, refresh = 0,
+  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
   control = list(adapt_delta = 0.95)
@@
   model_ncp, data = eight_schools,
-  iter = 4000, chains = 4, seed = 483892929, refresh = 0,
+  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
   control = list(adapt_delta = 0.95)
@@
 ```
+
+Looking into rf performance on 8 schools model more closely
+```{r}
+x <- x_cp
+x <- split_data(x)
+
+nparams <- dim(x)[3]
+nchains <- dim(x)[2]
+niter <- dim(x)[1]
+m_flattened <- matrix(nrow = nrow(x) * nchains,
+                      ncol = (nparams + 1))
+k <- 1
+for(i in 1:nchains){
+  for(j in 1:nrow(x)){
+    m_flattened[k, 1:nparams] <- x[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame()
+colnames(m_flattened)[nparams + 1] <- "chain"
+r <- m_flattened %>% 
+  mutate(chain=as.factor(chain))
+
+rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
+training_data <- r[rand_samples, ]
+testing_data <- r[-rand_samples, ]
+gbmFit <- train(chain ~ ., data = training_data, 
+               trControl = trainControl(method = 'none'),
+               method="rf",
+               tuneGrid = tibble(mtry=1))
+plda <- predict(object=gbmFit, newdata=testing_data)
+a_accuracy <- 
+  tibble(predicted=plda, actual=testing_data$chain) %>%
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct)) %>% 
+  pull()
+a_accuracy * n_distinct(training_data$chain)
+varImp(gbmFit)
+
+x[ , , 9] %>%
+  as.data.frame() %>%
+  mutate(iter=seq(1,1000,1)) %>%
+  melt(id.vars="iter") %>%
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  stat_density(position="identity", alpha=0.2)
+
+tibble(var1=testing_data$V1, var2=testing_data$V9, chain=plda) %>% 
+  ggplot(aes(x=var1, y=var2, colour=as.factor(chain))) +
+  geom_jitter()
+```
+
+Cauchy model
+```{r}
+fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878,
+                refresh = 0)
+x <- rstan::extract(fit_nom, permuted=F)
+temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
+temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
+temp$a_accuracy
+
+fit_alt1 <- stan(file = 'cauchy_alt_1.stan',
+                 seed = 7878, refresh = 0)
+x <- rstan::extract(fit_alt1, permuted=F)
+temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
+temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
+temp$a_accuracy
+```
+
+AR1 vs independent
+```{r}
+f_generate_lower_var_four_independent <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- rnorm(L, 0, sqrt(sigma^2/(1-rho^2)))
+  z <- rnorm(L, 0, sqrt(var_ratio * sigma^2/(1-rho^2)))
+  x[, 4] <- z
+  return(x)
+}
+
+rho <- 0.95
+lgbm <- vector(length=2)
+lrf <- vector(length=2)
+x <- f_generate_lower_var_four(1/3, rho, 1, 10000)
+var(x[, 4])
+lgbm[1] <- f_replicate_ml_variety(x, "gbm", tunegrid_gbm)
+lrf[1] <- f_replicate_ml_variety(x, "rf", tibble(mtry=1))
+
+x <- f_generate_lower_var_four_independent(1/3, rho, 1, 10000)
+var(x[, 4])
+lgbm[2] <- f_replicate_ml_variety(x, "gbm", tunegrid_gbm)
+lrf[2] <- f_replicate_ml_variety(x, "rf", tibble(mtry=1))
+```
+
+Try optimal Bayes classifier: two normals
+```{r}
+L <- 10000
+x1 <- rnorm(L, 0, 1)
+x2 <- rnorm(L, 1, 1)
+x <- matrix(c(x1, x2), byrow = F, ncol=2)
+
+a_array <- array(dim=c(nrow(x), 2, 1))
+a_array[,,1] <- x
+r_star_ml_variety(a_array, method_ ="rf", caretGrid = tibble(mtry=2))
+```
+
+Optimal classifier any dimension data
+```{r}
+# generate a correlation matrix via LKJ
+rlkj <- function(n, nu, d){
+  r_list <- vector(length = n, mode = 'list')
+  for(i in 1:n){
+    if(d==1)
+      r = as.array(1)
+    else if(d==2){
+      rho <- 2 * rbeta(1, nu, nu) - 1
+      r <- matrix(c(1, rho, rho, 1), ncol = 2)
+    }else{
+      beta <- nu + (d - 2) / 2
+      u <- rbeta(1, beta, beta)
+      r_12 <- 2 * u - 1
+      r <- matrix(c(1, r_12, r_12, 1), ncol = 2)
+      for(m in 2:(d - 1)){
+        beta <- beta - 0.5
+        y <- rbeta(1, m / 2, beta)
+        a <- rnorm(m)
+        anorm <- sqrt(sum(a^2))
+        u <- a / anorm
+        w <- sqrt(y) * u
+        A <- chol(r)
+        z <- w%*%A
+        r <-cbind(r, t(z))
+        r <- rbind(r, c(z, 1))
+      }
+    }
+    r_list[[i]] <- r
+  }
+  return(r_list)
+}
+
+nd <- 50
+df <- 200
+Sigma1 <- rlkj(1, 50, nd)[[1]]
+Sigma2 <- rlkj(1, 50, nd)[[1]]
+
+# generate any dimensional data
+library(mvtnorm)
+n <- 2000
+x1 <- rmvnorm(n, rep(0, nd), Sigma1)
+x2 <- rmvnorm(n, rep(0, nd), Sigma2)
+
+# calculate likelihood from each mode for each point
+lp11 <- map_dbl(seq(1, nrow(x1), 1), ~dmvnorm(x1[., ], rep(0, nd), Sigma1, log = T))
+lp12 <- map_dbl(seq(1, nrow(x1), 1), ~dmvnorm(x1[., ], rep(0, nd), Sigma2, log = T))
+lp21 <- map_dbl(seq(1, nrow(x2), 1), ~dmvnorm(x2[., ], rep(0, nd), Sigma1, log = T))
+lp22 <- map_dbl(seq(1, nrow(x2), 1), ~dmvnorm(x2[., ], rep(0, nd), Sigma2, log = T))
+r_star_opt <- 0.5 * (mean(lp11>lp12) + mean(lp22>lp21)) / 0.5
+
+# calculate R* from gbm and rf
+x <- array(dim=c(nrow(x1), 2, ncol(x1)))
+x[, 1, ] <- x1
+x[, 2, ] <- x2
+r_star(x, split_chains = T)
+
+f_replicate_ml_variety_simple <- function(x, method_, caretGrid_=NULL){
+  temp <- x
+  a_accuracy <- r_star_ml_variety(x, method_=method_, caretGrid = caretGrid_)
+  return(a_accuracy)
+}
+f_replicate_ml_variety_simple(x, "rf", tibble(mtry=20))
+f_replicate_ml_variety_simple(x, "gbm",
+                              tibble(interaction.depth=c(15), 
+                              n.trees = 200,
+                              shrinkage=c(0.1),
+                              n.minobsinnode=40))
+```

### 2020-10-03 src/s_ml_comparison.Rmd
Update s_ml_comparison.Rmd

@@
 ```{r}
-x <- x_cp
+x <- x_ncp
 x <- split_data(x)
@@
 
-x[ , , 9] %>%
-  as.data.frame() %>%
-  mutate(iter=seq(1,1000,1)) %>%
-  melt(id.vars="iter") %>%
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  stat_density(position="identity", alpha=0.2)
+# x[ , , 9] %>%
+#   as.data.frame() %>%
+#   mutate(iter=seq(1,1000,1)) %>%
+#   melt(id.vars="iter") %>%
+#   ggplot(aes(x=value, fill=as.factor(variable))) +
+#   stat_density(position="identity", alpha=0.2)
 
-tibble(var1=testing_data$V1, var2=testing_data$V9, chain=plda) %>% 
+tibble(var1=testing_data$V6, var2=testing_data$V10, chain=plda) %>% 
   ggplot(aes(x=var1, y=var2, colour=as.factor(chain))) +
@@
 r_star(x, split_chains = T)
+tmp <- monitor(x)
 

### 2020-10-04 src/s_ml_comparison.Rmd
more 8 schools analysis

@@
   model_ncp, data = eight_schools,
-  iter = 2000, chains = 4, seed = 483892929, refresh = 0,
+  iter = 40000, chains = 4, seed = 483892929, refresh = 0,
   control = list(adapt_delta = 0.95)
@@
 
-tibble(var1=testing_data$V6, var2=testing_data$V10, chain=plda) %>% 
+tibble(var1=testing_data$V7, var2=testing_data$V15, chain=plda) %>% 
   ggplot(aes(x=var1, y=var2, colour=as.factor(chain))) +
-  geom_jitter()
+  geom_jitter() +
+  geom_smooth(method="lm", se=F)
+
+tibble(var1=testing_data$V7, var2=testing_data$V15, chain=plda) %>% 
+  ggplot(aes(x=var1, colour=as.factor(chain))) +
+  stat_density(fill="none", position="identity")
+
+hist(map_dbl(seq(1, 8, 1), function(i) cor(testing_data %>% filter(chain==i) %>% select(-chain) %>% as.matrix())[7, 15]))
 ```
 
-Cauchy model
+Visualise between-chain correlation
 ```{r}
-fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878,
-                refresh = 0)
-x <- rstan::extract(fit_nom, permuted=F)
-temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
-temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
-temp$a_accuracy
+x <- split_data(x_ncp)
+cors <- map(seq(1, dim(x)[2], 1), ~cor(x[, ., ]))
+m_sd <- matrix(nrow = 19, ncol = 19)
+for(i in 1:19)
+  for(j in 1:19)
+    m_sd[i, j] <- sd(map_dbl(seq(1, dim(x)[2], 1), ~cors[[.]][i, j]))
+tmp <- which(m_sd == max(m_sd), arr.ind = TRUE)
+a <- tmp[1,1]
+b <- tmp[1,2]
+m_sd[a, b]
+colnames(r)[a] <- "a"
+colnames(r)[b] <- "b"
+
+r %>% 
+  ggplot(aes(x=a, y=b, colour=as.factor(chain))) +
+  geom_jitter(alpha=0.1) +
+  geom_smooth(method="lm", se=F)
+
+r %>% 
+  select(chain, a, b) %>% 
+  melt(id.vars="chain") %>% 
+  ggplot(aes(x=value, colour=as.factor(chain), group=as.factor(chain))) +
+  stat_density(fill="none", position = "identity") +
+  facet_wrap(~variable)
+
+r %>% 
+  select(chain, a, b) %>% 
+  melt(id.vars="chain") %>% 
+  group_by(variable, chain) %>% 
+  summarise(value=mean(value)) %>% 
+  ggplot(aes(x=as.factor(chain), y=value)) +
+  geom_point() +
+  facet_wrap(~variable)
+
+r %>% 
+  select(chain, a, b) %>% 
+  melt(id.vars="chain") %>% 
+  group_by(variable, chain) %>% 
+  summarise(value=quantile(value, 0.95)) %>% 
+  ggplot(aes(x=as.factor(chain), y=value)) +
+  geom_point() +
+  facet_wrap(~variable)
+```
+
+How R* varies with hyperparameters
+```{r}
+x <- x_ncp
+
+f_run_gbm <- function(x, n.trees_, int.depth_){
+  return(f_replicate_ml_simple(x, "gbm", expand.grid(interaction.depth=c(int.depth_), 
+                             n.trees = n.trees_,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)))
+}
+
+hypers <- tibble(
+       n.trees=c(50, 50, 100, 200, 200, 200, 200, 400, 400, 1000),
+       int.depth=c(3,5,3,3,5,7,9,5,9,11))
+k <- 1
+niter <- 20
+m_results <- matrix(nrow = nrow(hypers) * niter, ncol=4)
+for(i in 1:nrow(hypers)){
+  for(j in 1:niter){
+    tree <- hypers$n.trees[i]
+    depth <- hypers$int.depth[i]
+    temp <- f_run_gbm(x, tree, depth)
+    m_results[k, ] <- c(tree, depth, j, temp)
+    k <- k + 1
+  }
+}
+colnames(m_results) <- c("n.tree", "int.depth", "iter", "r_star")
+m_results <- m_results %>% 
+  as.data.frame()
+saveRDS(m_results, "../data/8_schools_gbm_hyper.rds")
+m_results %>% 
+  group_by(n.tree, int.depth) %>% 
+  summarise(lower=quantile(r_star, 0.25),
+            middle=median(r_star),
+            upper=quantile(r_star, 0.75)) %>% 
+  ggplot(aes(x=as.factor(n.tree), y=middle, color=as.factor(int.depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position = position_dodge2(width=0.2)) +
+  ylim(0.95, NA) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int.depth", palette = "Dark2") +
+  xlab("# tree") +
+  ylab("R*")
+```
+
+Try gbm function itself to optimise hyperparameters
+```{r}
+x <- x_ncp
+x <- split_data(x)
+
+nparams <- dim(x)[3]
+nchains <- dim(x)[2]
+niter <- dim(x)[1]
+m_flattened <- matrix(nrow = nrow(x) * nchains,
+                      ncol = (nparams + 1))
+k <- 1
+for(i in 1:nchains){
+  for(j in 1:nrow(x)){
+    m_flattened[k, 1:nparams] <- x[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame()
+colnames(m_flattened)[nparams + 1] <- "chain"
+r <- m_flattened %>% 
+  mutate(chain=as.factor(chain))
+
+rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
+training_data <- r[rand_samples, ]
+testing_data <- r[-rand_samples, ]
+
+gbmFit <- train(chain ~ ., data = training_data, 
+               method="gbm",
+               tuneGrid = expand.grid(
+       n.trees=c(50, 200, 2000),
+       interaction.depth=c(3,11, 17),
+       shrinkage=c(0.1),
+             n.minobsinnode=c(10)))
+plda <- predict(object=gbmFit, newdata=testing_data)
+a_accuracy <- 
+  tibble(predicted=plda, actual=testing_data$chain) %>%
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct)) %>% 
+  pull()
+a_accuracy * n_distinct(training_data$chain)
+varImp(gbmFit)
+```
+
+# Cauchy model
+```{r}
+# fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878,
+#                 refresh = 0)
+# x <- rstan::extract(fit_nom, permuted=F)
+# temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
+# temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
+# temp$a_accuracy
 
@@
 x <- rstan::extract(fit_alt1, permuted=F)
-temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
-temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
-temp$a_accuracy
+# temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
+# temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
+# temp$a_accuracy
+
+hypers <- tibble(
+       n.trees=c(50, 50, 100, 200, 200, 200, 200, 400, 400, 1000),
+       int.depth=c(3,5,3,3,5,7,9,5,9,11))
+k <- 1
+niter <- 3
+m_results <- matrix(nrow = nrow(hypers) * niter, ncol=4)
+for(i in 1:nrow(hypers)){
+  for(j in 1:niter){
+    tree <- hypers$n.trees[i]
+    depth <- hypers$int.depth[i]
+    temp <- f_run_gbm(x, tree, depth)
+    m_results[k, ] <- c(tree, depth, j, temp)
+    k <- k + 1
+  }
+}
 ```

### 2020-10-09 src/s_hyperparameters.Rmd
set up infrastructure for hyperparameter runs and ml comparison

---
title: "Hyperparameter sensitivity"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

# AR1 model
```{r}
f_ar1 <- function(rho, sigma, L){
  x <- vector(length = L)
  x[1] <- rnorm(1, 0, sd=sigma)
  for(i in 2:L)
    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
  return(x)
}

# Generates three chains with same var; one with a different var
f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
  x <- matrix(nrow = L, ncol = 4)
  for(i in 1:3)
    x[, i] <- f_ar1(rho, sigma, L)
  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
  x[, 4] <- z
  return(x)
}
```

```{r}
f_data_ar1 <- function(){
  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
  a_array <- array(dim=c(1000, 4, 1))
  a_array[,,1] <- temp
  return(a_array)
}

f_replicate_hyper <- function(m_parameters, f_data_generator){

  a_array <- f_data_generator()

  laccuracy <- vector(length = nrow(m_parameters))
  for(i in 1:nrow(m_parameters))
    laccuracy[i] <- r_star(a_array,
                           caret_default=expand.grid(interaction.depth=m_parameters$int.depth[i],
                                                     n.trees = m_parameters$n.trees[i],
                                                     shrinkage=0.1,
                                                     n.minobsinnode=10))
  return(laccuracy)
}

f_replicate_r_star <- function(nreplicates, m_parameters, f_data_generator){
  r_star_vals <- matrix(nrow = nreplicates, ncol=nrow(m_parameters))
  for(i in 1:nreplicates){
      print(i)

    temp <- f_replicate_hyper(m_parameters, f_data_generator)
    r_star_vals[i, ] <- temp
  }
  r_star_vals <- r_star_vals %>% 
    t() %>% 
    as.data.frame()
  r_star_vals1 <- m_parameters %>% cbind(r_star_vals)
  return(r_star_vals1)
}

m_parameters <- expand_grid(int.depth=seq(1, 7, 2),
                            n.trees=c(1, 10, 50, 100))
r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_ar1)

saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
r_star_vals <- readRDS("../data/ar1_sensitivity.rds")

b_sum <-
  r_star_vals %>% 
  as.data.frame() %>% 
  mutate(iterate=seq(1, nreplicates, 1)) %>%
  melt(id.vars="iterate") %>% 
  mutate(below=if_else(value < 1, 1, 0)) %>% 
  group_by(variable) %>% 
  summarise(below_per=round(mean(below)*100, 1),
            lower=quantile(value, 0.25),
            upper=quantile(value, 0.75),
            middle=median(value)) %>%
  mutate(below_per=paste0(below_per, "%")) %>% 
  mutate(x_pos=seq(1, nrow(m_parameters), 1)) %>% 
  mutate(y_pos=0.95)

g <- b_sum %>% 
  cbind(m_parameters) %>% 
  ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
  geom_line() +
  xlab("# trees") +
  ylab("R*") +
  theme(text = element_text(size=16, colour="black"),
        axis.text = element_text(colour="black"),
        legend.position = c(0.7, 0.25)) +
  scale_y_continuous(limits=c(0.98, NA)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_color_brewer("Int. depth", palette = "Dark2") +
  guides(colour = guide_legend(nrow = 2, byrow = T))
ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
```

# 8 schools
```{r}
source("eight_schools.data.R")
eight_schools <- list(J=J, y=y, sigma=sigma)
model_cp <- stan_model("eight_schools_cp.stan")
model_ncp <- stan_model("eight_schools_ncp.stan")

f_data_8_schools <- function(){
  fit_ncp <- sampling(
    model_ncp, data = eight_schools,
    iter = 2000, chains = 4, refresh = 0,
    control = list(adapt_delta = 0.95)
    )
  x_ncp <- rstan::extract(fit_ncp, permuted=F)
}

m_parameters <- expand_grid(int.depth=c(1, 3, 5),
                            n.trees=c(1, 50, 200))
r_star_vals <- f_replicate_r_star(20, m_parameters, f_data_ar1)

r_star_vals %>% 
  melt(id.vars=c("n.trees", "int.depth")) %>% 
  group_by(n.trees, int.depth) %>% 
  summarise(lower=quantile(value, 0.25),
            upper=quantile(value, 0.75),
            middle=median(value)) %>% 
  ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
  geom_line() +
  xlab("# trees") +
  ylab("R*") +
  theme(text = element_text(size=16, colour="black"),
        axis.text = element_text(colour="black"),
        legend.position = c(0.7, 0.25)) +
  scale_y_continuous(limits=c(0.98, NA)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_color_brewer("Int. depth", palette = "Dark2") +
  guides(colour = guide_legend(nrow = 2, byrow = T))
```

# Cauchy
```{r}
f_data_cauchy <- function(){
  fit_nom <- stan(file = 'cauchy_alt_1.stan',
                refresh = 0)
  x <- rstan::extract(fit_nom, permuted=F)
  return(x)
}

m_parameters <- expand_grid(int.depth=c(3, 7, 11),
                            n.trees=c(50, 100, 400))
r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_cauchy)
```

### 2020-10-09 src/s_ml.Rmd
set up infrastructure for hyperparameter runs and ml comparison

---
title: ML comparison"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

Function to fit any ML model in Caret
```{r}
r_star_ml_variety <- function(x, method_, caretGrid){

  split_chains=T
  training_percent=0.7
  if(split_chains)
    x <- split_data(x)
  nparams <- dim(x)[3]
  nchains <- dim(x)[2]
  niter <- dim(x)[1]
  m_flattened <- matrix(nrow = nrow(x) * nchains,
                        ncol = (nparams + 1))
  k <- 1
  for(i in 1:nchains){
    for(j in 1:nrow(x)){
      m_flattened[k, 1:nparams] <- x[j, i, ]
      m_flattened[k, (nparams + 1)] <- i
      k <- k + 1
    }
  }
  m_flattened <- m_flattened %>% 
    as.data.frame()
  colnames(m_flattened)[nparams + 1] <- "chain"
  r <- m_flattened %>% 
    mutate(chain=as.factor(chain))
  # if only 1 param, add in a column of random noise since gbm requires >1 dims
  if(nparams==1)
    r <- r %>% 
      mutate(V_new=rnorm(nrow(r)))

  rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
  training_data <- r[rand_samples, ]
  testing_data <- r[-rand_samples, ]

  if(method_!="multinom")
      gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_,
                   tuneGrid = caretGrid)
  else
    gbmFit <- train(chain ~ ., data = training_data, 
                 trControl = trainControl(method = 'none'),
                 method=method_,
                 tuneGrid = caretGrid,
                 MaxNWts=10000
                 )
  plda <- predict(object=gbmFit, newdata=testing_data)
  a_accuracy <- 
    tibble(predicted=plda, actual=testing_data$chain) %>%
    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
    summarise(mean(correct)) %>% 
    pull()
  return(a_accuracy * n_distinct(training_data$chain))
}

f_replicate_gridded <- function(x, method_, caretGrids_){
  r_stars <- vector(length = nrow(caretGrids_))
  for(i in 1:nrow(caretGrids_))
    r_stars[i] <- r_star_ml_variety(x, method_, caretGrids_[i, ])
  return(r_stars)
}

f_run_all <- function(methods, list_of_caret_grids, f_data_generator){
  x <- f_data_generator()
  r_stars <- vector(length = length(methods))
  for(i in seq_along(r_stars))
    r_stars[i] <- max(f_replicate_gridded(x, methods[i], list_of_caret_grids[[i]]))
  return(tibble(r_star=r_stars, method=methods))
}

f_run_all_replicates <- function(niterations, methods, list_of_caret_grids, f_data_generator){
  for(i in 1:niterations){
    print(i)
    temp_df <- f_run_all(methods, list_of_caret_grids, f_data_generator) %>% mutate(iter=i)
    if(i==1)
      big_df <- temp_df
    else
      big_df <- big_df %>% bind_rows(temp_df)
  }
  return(big_df)
}

tunegrid_gbm <- expand.grid(interaction.depth=3, 
                             n.trees = 50,
                             shrinkage=0.1,
                             n.minobsinnode=10)
tunegrid_rf <- tibble(mtry = 1:2)
tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
tunegrid_svm <- tibble(C = c(0.25, 0.5, 0.75))
tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
```

# AR1
```{r}
f_ar1 <- function(rho, sigma, L){
  x <- vector(length = L)
  x[1] <- rnorm(1, 0, sd=sigma)
  for(i in 2:L)
    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
  return(x)
}

# Generates three chains with same var; one with a different var
f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
  x <- matrix(nrow = L, ncol = 4)
  for(i in 1:3)
    x[, i] <- f_ar1(rho, sigma, L)
  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
  x[, 4] <- z
  return(x)
}

f_data_ar1 <- function(){
  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
  a_array <- array(dim=c(1000, 4, 1))
  a_array[,,1] <- temp
  return(a_array)
}

temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
          list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
          f_data_ar1)
```

# 8 schools
```{r}
source("eight_schools.data.R")
eight_schools <- list(J=J, y=y, sigma=sigma)
model_cp <- stan_model("eight_schools_cp.stan")
model_ncp <- stan_model("eight_schools_ncp.stan")

f_data_8_schools <- function(){
  fit_ncp <- sampling(
    model_ncp, data = eight_schools,
    iter = 2000, chains = 4, refresh = 0,
    control = list(adapt_delta = 0.95)
    )
  x_ncp <- rstan::extract(fit_ncp, permuted=F)
}

temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
          list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
          f_data_8_schools)
```

# Cauchy
```{r}
f_data_cauchy <- function(){
  fit_nom <- stan(file = 'cauchy_alt_1.stan',
                refresh = 0)
  x <- rstan::extract(fit_nom, permuted=F)
  return(x)
}

temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
          list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
          f_data_cauchy)
```

# Multivariate normal
```{r}

```

### 2020-10-09 src/s_ml_comparison.Rmd
set up infrastructure for hyperparameter runs and ml comparison

--- 2020-10-03 src/s_ml_comparison.Rmd
@@
   if(!is.null(caretGrid))
+    if(method_!="multinom")
       gbmFit <- train(chain ~ ., data = training_data, 
@@
                    tuneGrid = caretGrid)
+    else
+      gbmFit <- train(chain ~ ., data = training_data, 
+                   trControl = trainControl(method = 'none'),
+                   method=method_,
+                   tuneGrid = caretGrid,
+                   MaxNWts=10000
+                   )
+
   else
@@
 
-colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "nnet")
+colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "nnet", "multinom")
 m_res1 <- m_res %>%
@@
 g <- m_res1 %>% 
-  dplyr::select(-one_of("rhat")) %>% 
+  dplyr::select(-one_of("rhat", "nnet")) %>% 
   melt(id.vars="iter") %>% 
@@
 
-tibble(var1=testing_data$V6, var2=testing_data$V10, chain=plda) %>% 
+tibble(var1=testing_data$V7, var2=testing_data$V15, chain=plda) %>% 
   ggplot(aes(x=var1, y=var2, colour=as.factor(chain))) +
-  geom_jitter()
+  geom_jitter() +
+  geom_smooth(method="lm", se=F)
+
+tibble(var1=testing_data$V7, var2=testing_data$V15, chain=plda) %>% 
+  ggplot(aes(x=var1, colour=as.factor(chain))) +
+  stat_density(fill="none", position="identity")
+
+hist(map_dbl(seq(1, 8, 1), function(i) cor(testing_data %>% filter(chain==i) %>% select(-chain) %>% as.matrix())[7, 15]))
 ```
 
-Cauchy model
+Visualise between-chain correlation
+```{r}
+x <- split_data(x_ncp)
+cors <- map(seq(1, dim(x)[2], 1), ~cor(x[, ., ]))
+m_sd <- matrix(nrow = 19, ncol = 19)
+for(i in 1:19)
+  for(j in 1:19)
+    m_sd[i, j] <- sd(map_dbl(seq(1, dim(x)[2], 1), ~cors[[.]][i, j]))
+tmp <- which(m_sd == max(m_sd), arr.ind = TRUE)
+a <- tmp[1,1]
+b <- tmp[1,2]
+m_sd[a, b]
+colnames(r)[a] <- "a"
+colnames(r)[b] <- "b"
+
+g <- r %>% 
+  ggplot(aes(x=a, y=b, colour=as.factor(chain))) +
+  geom_jitter(alpha=0.1) +
+  geom_smooth(method="lm", se=F) +
+  scale_color_brewer("Chain", palette = "Dark2")
+ggsave("../output/ml_comparison_8_schools_ncp_1.pdf", g, width = 12, height = 8)
+
+g <- r %>% 
+  select(chain, a, b) %>% 
+  melt(id.vars="chain") %>% 
+  ggplot(aes(x=value, colour=as.factor(chain), group=as.factor(chain))) +
+  stat_density(fill="none", position = "identity") +
+  facet_wrap(~variable)
+ggsave("../output/ml_comparison_8_schools_ncp_2.pdf", g, width = 12, height = 8)
+
+g <- r %>% 
+  select(chain, a, b) %>% 
+  melt(id.vars="chain") %>% 
+  group_by(variable, chain) %>% 
+  summarise(value=mean(value)) %>% 
+  ggplot(aes(x=as.factor(chain), y=value)) +
+  geom_point() +
+  facet_wrap(~variable)
+ggsave("../output/ml_comparison_8_schools_ncp_3.pdf", g, width = 12, height = 8)
+
+g <- r %>% 
+  select(chain, a, b) %>% 
+  melt(id.vars="chain") %>% 
+  group_by(variable, chain) %>% 
+  summarise(value=quantile(value, 0.95)) %>% 
+  ggplot(aes(x=as.factor(chain), y=value)) +
+  geom_point() +
+  facet_wrap(~variable)
+ggsave("../output/ml_comparison_8_schools_ncp_4.pdf", g, width = 12, height = 8)
+```
+
+How R* varies with hyperparameters
+```{r}
+x <- x_ncp
+
+f_run_gbm <- function(x, n.trees_, int.depth_){
+  return(f_replicate_ml_simple(x, "gbm", expand.grid(interaction.depth=c(int.depth_), 
+                             n.trees = n.trees_,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)))
+}
+
+hypers <- tibble(
+       n.trees=c(50, 50, 100, 200, 200, 200, 200, 400, 400, 1000),
+       int.depth=c(3,5,3,3,5,7,9,5,9,11))
+k <- 1
+niter <- 20
+m_results <- matrix(nrow = nrow(hypers) * niter, ncol=4)
+for(i in 1:nrow(hypers)){
+  for(j in 1:niter){
+    tree <- hypers$n.trees[i]
+    depth <- hypers$int.depth[i]
+    temp <- f_run_gbm(x, tree, depth)
+    m_results[k, ] <- c(tree, depth, j, temp)
+    k <- k + 1
+  }
+}
+colnames(m_results) <- c("n.tree", "int.depth", "iter", "r_star")
+m_results <- m_results %>% 
+  as.data.frame()
+saveRDS(m_results, "../data/8_schools_gbm_hyper.rds")
+m_results <- readRDS("../data/8_schools_gbm_hyper.rds")
+g <- m_results %>% 
+  group_by(n.tree, int.depth) %>% 
+  summarise(lower=quantile(r_star, 0.25),
+            middle=median(r_star),
+            upper=quantile(r_star, 0.75)) %>% 
+  ggplot(aes(x=as.factor(n.tree), y=middle, color=as.factor(int.depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position = position_dodge2(width=0.2)) +
+  ylim(0.95, NA) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int.depth", palette = "Dark2") +
+  xlab("# tree") +
+  ylab("R*")
+ggsave("../output/ml_comparison_8_schools_hyper.pdf", g, width = 12, height = 8)
+```
+
+Try gbm function itself to optimise hyperparameters
+```{r}
+x <- x_ncp
+x <- split_data(x)
+
+nparams <- dim(x)[3]
+nchains <- dim(x)[2]
+niter <- dim(x)[1]
+m_flattened <- matrix(nrow = nrow(x) * nchains,
+                      ncol = (nparams + 1))
+k <- 1
+for(i in 1:nchains){
+  for(j in 1:nrow(x)){
+    m_flattened[k, 1:nparams] <- x[j, i, ]
+    m_flattened[k, (nparams + 1)] <- i
+    k <- k + 1
+  }
+}
+m_flattened <- m_flattened %>% 
+  as.data.frame()
+colnames(m_flattened)[nparams + 1] <- "chain"
+r <- m_flattened %>% 
+  mutate(chain=as.factor(chain))
+
+rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
+training_data <- r[rand_samples, ]
+testing_data <- r[-rand_samples, ]
+
+gbmFit <- train(chain ~ ., data = training_data, 
+               method="gbm",
+               tuneGrid = expand.grid(
+       n.trees=c(50, 200, 2000),
+       interaction.depth=c(3,11, 17),
+       shrinkage=c(0.1),
+             n.minobsinnode=c(10)))
+plda <- predict(object=gbmFit, newdata=testing_data)
+a_accuracy <- 
+  tibble(predicted=plda, actual=testing_data$chain) %>%
+  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+  summarise(mean(correct)) %>% 
+  pull()
+a_accuracy * n_distinct(training_data$chain)
+varImp(gbmFit)
+```
+
+# Cauchy model
 ```{r}
@@
 x <- rstan::extract(fit_nom, permuted=F)
-temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
-temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
-temp$a_accuracy
+
+tunegrid_rf <- tibble(mtry = 1:2)
+tunegrid_gbm <- expand.grid(interaction.depth=c(3), 
+                             n.trees = 50,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)
+tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
+tunegrid_svm <- tibble(C = c(0.25, 0.5, 0.75))
+tunegrid_nnet <- tibble(size=c(2, 3, 4, 5), decay=c(0.5, 0.5, 0.5, 0.5))
+tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
+niterates <- 10
+m_res <- matrix(nrow=niterates, ncol = 6)
+l_rf_hyper <- vector(length = niterates, mode="list")
+l_knn_hyper <- vector(length = niterates, mode="list")
+l_svm_hyper <- vector(length = niterates, mode="list")
+l_nnet_hyper <- vector(length = niterates, mode="list")
+l_multinom_hyper <- vector(length = niterates, mode="list")
+for(i in 1:niterates){
+  m_res[i, 1] <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
+  m_res[i, 2] <- max(f_replicate_rhat_simple(x))
+  temp <- f_replicate_ml_simple_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
+  l_rf_hyper[[i]] <- temp$hyper
+  m_res[i, 3] <- temp$a_accuracy
+  temp <- f_replicate_ml_simple_gridded(x, "knn", tunegrid_knn)
+  l_knn_hyper[[i]] <- temp$hyper
+  m_res[i, 4] <- temp$a_accuracy
+  temp <- f_replicate_ml_simple_gridded(x, "svmLinear", tunegrid_svm)
+  l_svm_hyper[[i]] <- temp$hyper
+  m_res[i, 5] <- temp$a_accuracy
+  temp <- f_replicate_ml_simple_gridded(x, "multinom", tunegrid_multinom)
+  l_nnet_hyper[[i]] <- temp$hyper
+  m_res[i, 6] <- temp$a_accuracy
+}
+
+colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "multinom")
+m_res1 <- m_res %>%
+  as.data.frame() %>% 
+  mutate(iter=seq_along(gbm))
+saveRDS(m_res1, "../data/ml_comparison_cauchy_nom.rds")
+
+g <- m_res1 %>% 
+  dplyr::select(-one_of("rhat")) %>% 
+  melt(id.vars="iter") %>% 
+  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value)) +
+  geom_boxplot() +
+  xlab("ML model") +
+  ylab("R*") +
+  theme(text=element_text(colour="black", size=14))
+ggsave("../output/ml_comparison_cauchy_nom.pdf", g, width = 12, height = 8)
 
@@
 x <- rstan::extract(fit_alt1, permuted=F)
-temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
-temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
-temp$a_accuracy
+
+m_res <- matrix(nrow=niterates, ncol = 6)
+l_rf_hyper <- vector(length = niterates, mode="list")
+l_knn_hyper <- vector(length = niterates, mode="list")
+l_svm_hyper <- vector(length = niterates, mode="list")
+l_nnet_hyper <- vector(length = niterates, mode="list")
+l_multinom_hyper <- vector(length = niterates, mode="list")
+for(i in 1:niterates){
+  m_res[i, 1] <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
+  m_res[i, 2] <- max(f_replicate_rhat_simple(x))
+  temp <- f_replicate_ml_simple_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
+  l_rf_hyper[[i]] <- temp$hyper
+  m_res[i, 3] <- temp$a_accuracy
+  temp <- f_replicate_ml_simple_gridded(x, "knn", tunegrid_knn)
+  l_knn_hyper[[i]] <- temp$hyper
+  m_res[i, 4] <- temp$a_accuracy
+  temp <- f_replicate_ml_simple_gridded(x, "svmLinear", tunegrid_svm)
+  l_svm_hyper[[i]] <- temp$hyper
+  m_res[i, 5] <- temp$a_accuracy
+  temp <- f_replicate_ml_simple_gridded(x, "multinom", tunegrid_multinom)
+  l_multinom_hyper[[i]] <- temp$hyper
+  m_res[i, 6] <- temp$a_accuracy
+}
+
+colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "multinom")
+m_res1 <- m_res %>%
+  as.data.frame() %>% 
+  mutate(iter=seq_along(gbm))
+saveRDS(m_res1, "../data/ml_comparison_cauchy_alt.rds")
+m_res1 <- readRDS("../data/ml_comparison_cauchy_alt.rds")
+m_res2 <- readRDS("../data/ml_comparison_cauchy_nom.rds") %>% 
+  mutate(type="nominal")
+
+g <- m_res1 %>% 
+  mutate(type="alternative") %>% 
+  bind_rows(m_res2) %>% 
+  dplyr::select(-one_of("rhat")) %>% 
+  melt(id.vars=c("iter", "type")) %>% 
+  mutate(value=as.numeric(as.character(value))) %>% 
+  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value, fill=as.factor(type))) +
+  geom_boxplot() +
+  xlab("ML model") +
+  ylab("R*") +
+  theme(text=element_text(colour="black", size=14)) +
+  scale_fill_brewer("Parameterisation", palette = "Dark2") +
+  scale_y_continuous(limits=c(0.8, NA)) +
+  geom_hline(yintercept = 1, linetype=2)
+ggsave("../output/ml_comparison_cauchy.pdf", g, width = 12, height = 8)
 ```

### 2020-10-09 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
set up infrastructure for hyperparameter runs and ml comparison

@@
 Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
-The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
+\textcolor{red}{The method is straightforward to implement and could be a complementary additional check on MCMC convergence for applied analyses.}
 \end{abstract}
@@
 For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
-It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
-In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
+\textcolor{red}{The statistic depends on the hyperparameters of GBM, although even for ``default'' settings, $R^*$ provides a stringent test of accuracy, giving a reliable measure of various Markov chain pathologies.} In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
 $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
@@
 
-\subsubsection{Robustness to hyperparameter choice}\label{sec:heterogeneity_robustness}
-The performance of GBM, like all ML methods, depends on its hyperparameters.
-To investigate the robustness of $R^*$ to GBM hyperparameter choice, we performed a sensitivity analysis.
-In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
-\ref{fig:ar1}.
-The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
-The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
-\ref{fig:ar1_sensitivity}.
-These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper).
-This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ar1_sensitivity.pdf}}
-	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
-The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.}
-	\label{fig:ar1_sensitivity}
-\end{figure}
-
 \subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
@@
 
+\color{red}
+\subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
+The performance of GBM, like all ML methods, depends on its hyperparameters.
+In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree \cite{greenwell2019package}.
+The number of such regression trees is thus a hyperparameter of the model.
+Each regression tree can split amongst any integer number of variables (the \textit{interaction depth}).
+In this section, we investigate the sensitivity of $R^*$ to these hyperparameters across a range of examples.
+
+\subsubsection{Autoregressive example}\label{sec:heterogeneity_robustness}
+In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
+\ref{fig:ar1}.
+The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
+The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
+\ref{fig:ar1_sensitivity}.
+These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper).
+This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{ar1_sensitivity.pdf}}
+	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
+The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
+The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.}
+	\label{fig:ar1_sensitivity}
+\end{figure}
+
+\subsubsection{Eight schools: non-centered parameterisation}
+
+\subsubsection{Cauchy model: alternative paramerisation}
+
+\subsubsection{Multivariate normal: 250-dimensional model}
+
+\subsection{ML model sensitivity}\label{sec:ml_model}
+
+\color{black}
+
 \subsection{Further experiments}\label{sec:further_experiments}

### 2020-10-09 src/s_jobScheduler_hypers.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+chmod +x *.R
+chmod +x *.sh
+
+
+Rscript s_run_hyperparameters.R 1 200
+Rscript s_run_hyperparameters.R 2 50
+Rscript s_run_hyperparameters.R 3 20
+

### 2020-10-09 src/s_jobScheduler_master.sh

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+chmod +x *.R
+chmod +x *.sh
+
+
+sh s_jobScheduler_hypers.sh
+sh s_jobScheduler_ml.sh
+

### 2020-10-09 src/s_jobScheduler_ml.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+chmod +x *.R
+chmod +x *.sh
+
+
+Rscript s_run_ml.R 1 50
+Rscript s_run_ml.R 2 20
+Rscript s_run_ml.R 3 20
+

### 2020-10-09 src/s_run_hyperparameters.R

@@ -1,0 +1,113 @@
+setwd("/Volumes/Samsung1.5TB/Github/ml-mcmc-convergence/src")
+rm(list=ls())
+library(tidyverse)
+library(reshape2)
+library(rstan)
+library(latex2exp)
+library(caret)
+library(gbm)
+options(mc.cores=4)
+rstan_options(auto_write = TRUE)
+source("monitornew.R")
+source("r_star_monitor.R")
+source("eight_schools.data.R")
+eight_schools <- list(J=J, y=y, sigma=sigma)
+model_cp <- stan_model("eight_schools_cp.stan")
+model_ncp <- stan_model("eight_schools_ncp.stan")
+
+
+# helper functions ------
+f_ar1 <- function(rho, sigma, L){
+  x <- vector(length = L)
+  x[1] <- rnorm(1, 0, sd=sigma)
+  for(i in 2:L)
+    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
+  return(x)
+}
+
+f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, 4] <- z
+  return(x)
+}
+
+f_data_ar1 <- function(){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  return(a_array)
+}
+
+f_replicate_hyper <- function(m_parameters, f_data_generator){
+  
+  a_array <- f_data_generator()
+  
+  laccuracy <- vector(length = nrow(m_parameters))
+  for(i in 1:nrow(m_parameters))
+    laccuracy[i] <- r_star(a_array,
+                           caret_default=expand.grid(interaction.depth=m_parameters$int.depth[i],
+                                                     n.trees = m_parameters$n.trees[i],
+                                                     shrinkage=0.1,
+                                                     n.minobsinnode=10))
+  return(laccuracy)
+}
+
+
+f_replicate_r_star <- function(nreplicates, m_parameters, f_data_generator){
+  r_star_vals <- matrix(nrow = nreplicates, ncol=nrow(m_parameters))
[…]

### 2020-10-09 src/s_run_ml.R

@@ -1,0 +1,199 @@
+setwd("/Volumes/Samsung1.5TB/Github/ml-mcmc-convergence/src")
+rm(list=ls())
+library(tidyverse)
+library(reshape2)
+library(rstan)
+library(latex2exp)
+library(caret)
+library(gbm)
+options(mc.cores=4)
+rstan_options(auto_write = TRUE)
+source("monitornew.R")
+source("r_star_monitor.R")
+source("eight_schools.data.R")
+eight_schools <- list(J=J, y=y, sigma=sigma)
+model_cp <- stan_model("eight_schools_cp.stan")
+model_ncp <- stan_model("eight_schools_ncp.stan")
+
+
+# helper functions ------
+f_ar1 <- function(rho, sigma, L){
+  x <- vector(length = L)
+  x[1] <- rnorm(1, 0, sd=sigma)
+  for(i in 2:L)
+    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
+  return(x)
+}
+
+f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, 4] <- z
+  return(x)
+}
+
+f_data_ar1 <- function(){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  return(a_array)
+}
+
+f_replicate_hyper <- function(m_parameters, f_data_generator){
+  
+  a_array <- f_data_generator()
+  
+  laccuracy <- vector(length = nrow(m_parameters))
+  for(i in 1:nrow(m_parameters))
+    laccuracy[i] <- r_star(a_array,
+                           caret_default=expand.grid(interaction.depth=m_parameters$int.depth[i],
+                                                     n.trees = m_parameters$n.trees[i],
+                                                     shrinkage=0.1,
+                                                     n.minobsinnode=10))
+  return(laccuracy)
+}
+
+
+f_replicate_r_star <- function(nreplicates, m_parameters, f_data_generator){
+  r_star_vals <- matrix(nrow = nreplicates, ncol=nrow(m_parameters))
+  for(i in 1:nreplicates){
+    print(i)
+    
[…]

### 2020-10-10 src/s_ml.Rmd
first runs done of ml comparison

@@
   r_stars <- vector(length = length(methods))
-  for(i in seq_along(r_stars))
+  for(i in seq_along(r_stars)){
+    print(methods[i])
     r_stars[i] <- max(f_replicate_gridded(x, methods[i], list_of_caret_grids[[i]]))
+  }
   return(tibble(r_star=r_stars, method=methods))
@@
 tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
+tunegrid_xgbtree <- expand_grid(nrounds = c(1, 10),
+                       max_depth = c(1, 4),
+                       eta = c(.1, .4),
+                       gamma = 0,
+                       colsample_bytree = .7,
+                       min_child_weight = 1,
+                       subsample = c(.8, 1))
 ```
@@
           f_data_ar1)
+temp_df <- readRDS("../data/ml_comp_ar1.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
 ```
@@
           f_data_8_schools)
+
+temp_df <- readRDS("../data/ml_comp_8_schools.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
 ```
@@
           f_data_cauchy)
+temp_df <- readRDS("../data/ml_comp_cauchy.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
 ```
@@
 ```{r}
+f_data_normal <- function(){
+  N <- 250
+  A <- rWishart(1, 250, diag(N))[,,1]
+  model <- stan_model("mvt_250_ncp.stan")
+  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
+  x <- rstan::extract(fit, permuted=F)
+  return(x)
+}
 
+temp_df <- f_run_all_replicates(20, c("xgbTree", "rf", "gbm"),
+          list(tunegrid_xgbtree, tunegrid_rf, tunegrid_gbm),
+          f_data_normal)
 ```

### 2020-10-10 src/s_jobScheduler_hypers.sh

@@ -4,7 +4,7 @@ chmod +x *.R
 chmod +x *.sh
 
 
-Rscript s_run_hyperparameters.R 1 200
-Rscript s_run_hyperparameters.R 2 50
+# Rscript s_run_hyperparameters.R 1 200
+# Rscript s_run_hyperparameters.R 2 50
 Rscript s_run_hyperparameters.R 3 20
 

### 2020-10-10 src/s_jobScheduler_ml.sh

@@ -4,7 +4,9 @@ chmod +x *.R
 chmod +x *.sh
 
 
-Rscript s_run_ml.R 1 50
-Rscript s_run_ml.R 2 20
-Rscript s_run_ml.R 3 20
+Rscript s_run_ml.R 1 2
+# Rscript s_run_ml.R 2 20
+# Rscript s_run_ml.R 3 20
+# Rscript s_run_ml.R 4 20
+# Rscript s_run_ml.R 5 20
 

### 2020-10-10 src/s_run_hyperparameters.R

@@ -103,8 +103,8 @@ if(model_num==1){
   f_data_gen <- f_data_8_schools
   name <- "8_schools"
 }else if(model_num==3){
-  m_parameters <- expand_grid(int.depth=c(3, 7, 11),
-                              n.trees=c(50, 100, 400))
+  m_parameters <- expand_grid(int.depth=c(3, 7, 11, 13),
+                              n.trees=c(50, 100, 400, 1000))
   f_data_gen <- f_data_cauchy
   name <- "cauchy"
 }

### 2020-10-10 src/s_run_ml.R

@@ -87,6 +87,21 @@ f_data_cauchy <- function(){
   return(x)
 }
 
+f_data_normal <- function(){
+  N <- 250
+  A <- rWishart(1, 250, diag(N))[,,1]
+  model <- stan_model("mvt_250_ncp.stan")
+  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
+  x <- rstan::extract(fit, permuted=F)
+  return(x)
+}
+
+f_data_normal_10000 <- function(){
+  fit <- readRDS("../output/mvt_wide_1000.rds")
+  x <- rstan::extract(fit, permuted=F)
+  return(x)
+}
+
 r_star_ml_variety <- function(x, method_, caretGrid){
   
   split_chains=T
@@ -143,17 +158,30 @@ r_star_ml_variety <- function(x, method_, caretGrid){
 
 f_replicate_gridded <- function(x, method_, caretGrids_){
   r_stars <- vector(length = nrow(caretGrids_))
-  for(i in 1:nrow(caretGrids_))
+  times <- vector(length = nrow(caretGrids_))
+  for(i in 1:nrow(caretGrids_)){
+    start.time <- Sys.time()
     r_stars[i] <- r_star_ml_variety(x, method_, caretGrids_[i, ])
-  return(r_stars)
+    end.time <- Sys.time()
+    times[i] <- end.time - start.time
+  }
+  return(list(r_star=r_stars, time=times))
 }
 
 f_run_all <- function(methods, list_of_caret_grids, f_data_generator){
   x <- f_data_generator()
   r_stars <- vector(length = length(methods))
-  for(i in seq_along(r_stars))
-    r_stars[i] <- max(f_replicate_gridded(x, methods[i], list_of_caret_grids[[i]]))
-  return(tibble(r_star=r_stars, method=methods))
+  times <- vector(length = length(methods))
+  hypers <- vector(length = length(methods), mode="list")
+  for(i in seq_along(r_stars)){
+    a_grid <- list_of_caret_grids[[i]]
+    a <- f_replicate_gridded(x, methods[i], a_grid)
+    r_stars[i] <- max(a$r_star)
+    a_ind <- which.max(a$star)
+    times[i] <- a$times[a_ind]
+    hypers[[i]] <- a_grid[a_ind, ]
+  }
+  return(tibble(r_star=r_stars, method=methods, time=times, hyper=hypers))
 }
 
 f_run_all_replicates <- function(niterations, methods, list_of_caret_grids, f_data_generator){
[…]

### 2020-10-11 src/s_ml.Rmd
runs of ml comparison done

--- 2020-10-09 src/s_ml.Rmd
@@
   r_stars <- vector(length = nrow(caretGrids_))
-  for(i in 1:nrow(caretGrids_))
+  times <- vector(length = nrow(caretGrids_))
+  for(i in 1:nrow(caretGrids_)){
+    start.time <- Sys.time()
     r_stars[i] <- r_star_ml_variety(x, method_, caretGrids_[i, ])
-  return(r_stars)
+    end.time <- Sys.time()
+    times[i] <- end.time - start.time
+  }
+  return(list(r_star=r_stars, time=times))
 }
@@
   r_stars <- vector(length = length(methods))
-  for(i in seq_along(r_stars))
-    r_stars[i] <- max(f_replicate_gridded(x, methods[i], list_of_caret_grids[[i]]))
-  return(tibble(r_star=r_stars, method=methods))
+  times <- vector(length = length(methods))
+  hypers <- vector(length = length(methods), mode="list")
+  for(i in seq_along(r_stars)){
+    print(methods[i])
+    a_grid <- list_of_caret_grids[[i]]
+    a <- f_replicate_gridded(x, methods[i], a_grid)
+    r_stars[i] <- max(a$r_star)
+    a_ind <- which.max(a$r_star)
+    times[i] <- a$time[a_ind]
+    hypers[[i]] <- a_grid[a_ind, ]
+  }
+  return(tibble(r_star=r_stars, method=methods, time=times, hyper=hypers))
 }
@@
 
-tunegrid_gbm <- expand.grid(interaction.depth=3, 
-                             n.trees = 50,
-                             shrinkage=0.1,
-                             n.minobsinnode=10)
-tunegrid_rf <- tibble(mtry = 1:2)
+tunegrid_gbm <- expand.grid(interaction.depth=1, 
+                             n.trees = c(400),
+                             shrinkage=c(0.1),
+                             n.minobsinnode=c(1))
+tunegrid_rf <- tibble(mtry = c(1, 5, 15))
 tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
@@
           f_data_ar1)
+temp_df <- readRDS("../data/ml_comp_hypers_set_ar1.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
 ```
@@
           f_data_8_schools)
+
+temp_df <- readRDS("../data/ml_comp_hypers_set_8_schools.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
+temp_df %>%
+  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
+  geom_jitter()
 ```
@@
           f_data_cauchy)
+temp_df <- readRDS("../data/ml_comp_hypers_set_cauchy.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
+temp_df %>%
+  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
+  geom_jitter()
 ```
@@
 ```{r}
+f_data_normal <- function(){
+  N <- 250
+  A <- rWishart(1, 250, diag(N))[,,1]
+  model <- stan_model("mvt_250_ncp.stan")
+  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
+  x <- rstan::extract(fit, permuted=F)
+  return(x)
+}
 
+temp_df <- f_replicate_gridded(x, "rf", tunegrid_rf)
+
+temp_df <- readRDS("../data/ml_comp_hypers_set_normal_250.rds")
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
+temp_df %>%
+  ggplot(aes(x=as.factor(method), y=time)) +
+  geom_boxplot()
+
+temp_df %>%
+  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
+  geom_jitter()
 ```
+
+# Check for false positives with random forests
+```{r}
+f_data_converged <- function(){
+  d <- 10
+  n <- 2000
+  nchain <- 4
+
+  x <- rnorm(d * n * nchain)
+  x <- array(x, dim=c(n, nchain, d))
+  return(x)
+}
+
+temp_df <- f_run_all_replicates(10, c("gbm", "rf"),
+          list(tunegrid_gbm, tunegrid_rf),
+          f_data_converged)
+temp_df %>% 
+  ggplot(aes(x=as.factor(method), y=r_star)) +
+  geom_boxplot()
+```
+
+# Setting hyperparameters for random forest:
+"Influence of Hyperparameters on Random Forest Accuracy"
+Simon Bernard, Laurent Heutte, Sebastien Adam, 2009
+Use mtry=sqrt(number features)
+
+whereas for gbms / xgboost, performance depends on lots of things

### 2020-10-11 src/s_jobScheduler_ml.sh

@@ -5,8 +5,8 @@ chmod +x *.sh
 
 
 Rscript s_run_ml.R 1 2
-# Rscript s_run_ml.R 2 20
-# Rscript s_run_ml.R 3 20
-# Rscript s_run_ml.R 4 20
-# Rscript s_run_ml.R 5 20
+Rscript s_run_ml.R 2 2
+Rscript s_run_ml.R 3 2
+Rscript s_run_ml.R 4 2
+Rscript s_run_ml.R 5 2
 

### 2020-10-11 src/s_jobScheduler_ml_hypers_set.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+chmod +x *.R
+chmod +x *.sh
+
+
+Rscript s_run_ml_hypers_set.R 1 100
+Rscript s_run_ml_hypers_set.R 2 100
+Rscript s_run_ml_hypers_set.R 3 50
+Rscript s_run_ml_hypers_set.R 4 30
+Rscript s_run_ml_hypers_set.R 5 10
+

### 2020-10-11 src/s_run_ml.R

@@ -174,11 +174,12 @@ f_run_all <- function(methods, list_of_caret_grids, f_data_generator){
   times <- vector(length = length(methods))
   hypers <- vector(length = length(methods), mode="list")
   for(i in seq_along(r_stars)){
+    print(methods[i])
     a_grid <- list_of_caret_grids[[i]]
     a <- f_replicate_gridded(x, methods[i], a_grid)
     r_stars[i] <- max(a$r_star)
-    a_ind <- which.max(a$star)
-    times[i] <- a$times[a_ind]
+    a_ind <- which.max(a$r_star)
+    times[i] <- a$time[a_ind]
     hypers[[i]] <- a_grid[a_ind, ]
   }
   return(tibble(r_star=r_stars, method=methods, time=times, hyper=hypers))
@@ -204,7 +205,7 @@ tunegrid_rf <- tibble(mtry = 1:2)
 tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
 tunegrid_svm <- tibble(C = c(0.25, 0.5, 0.75))
 tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
-tunegrid_xgbtree <- expand_grid(nrounds = c(1, 10),
+tunegrid_xgbtree <- expand_grid(nrounds = c(1, 10, 100),
                                 max_depth = c(1, 4),
                                 eta = c(.1, .4),
                                 gamma = 0,

### 2020-10-11 src/s_run_ml_hypers_set.R

@@ -1,0 +1,251 @@
+setwd("/Volumes/Samsung1.5TB/Github/ml-mcmc-convergence/src")
+rm(list=ls())
+library(tidyverse)
+library(reshape2)
+library(rstan)
+library(latex2exp)
+library(caret)
+library(gbm)
+options(mc.cores=4)
+rstan_options(auto_write = TRUE)
+source("monitornew.R")
+source("r_star_monitor.R")
+source("eight_schools.data.R")
+eight_schools <- list(J=J, y=y, sigma=sigma)
+model_cp <- stan_model("eight_schools_cp.stan")
+model_ncp <- stan_model("eight_schools_ncp.stan")
+
+
+# helper functions ------
+f_ar1 <- function(rho, sigma, L){
+  x <- vector(length = L)
+  x[1] <- rnorm(1, 0, sd=sigma)
+  for(i in 2:L)
+    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
+  return(x)
+}
+
+f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, 4] <- z
+  return(x)
+}
+
+f_data_ar1 <- function(){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  return(a_array)
+}
+
+f_replicate_hyper <- function(m_parameters, f_data_generator){
+  
+  a_array <- f_data_generator()
+  
+  laccuracy <- vector(length = nrow(m_parameters))
+  for(i in 1:nrow(m_parameters))
+    laccuracy[i] <- r_star(a_array,
+                           caret_default=expand.grid(interaction.depth=m_parameters$int.depth[i],
+                                                     n.trees = m_parameters$n.trees[i],
+                                                     shrinkage=0.1,
+                                                     n.minobsinnode=10))
+  return(laccuracy)
+}
+
+
+f_replicate_r_star <- function(nreplicates, m_parameters, f_data_generator){
+  r_star_vals <- matrix(nrow = nreplicates, ncol=nrow(m_parameters))
+  for(i in 1:nreplicates){
+    print(i)
+    
[…]

### 2020-10-12 src/s_ar1.Rmd
added lots of things ml-related

@@
 
-nreplicates <- 1000
+nreplicates <- 100
 r_star_vals <- vector(length = nreplicates)
@@
 for(i in 1:nreplicates){
-  if(i%%100==0)
     print(i)
@@
 
-saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/ar1_r_star_hat.rds")
+# saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/ar1_r_star_hat.rds")
 ```
@@
 ```{r}
-reses <- readRDS("../data/ar1_r_star_hat.rds")
+# reses <- readRDS("../data/ar1_r_star_hat.rds")
 r_star_vals <- reses$r_star 
@@
 
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
+caretGrid <- tibble(mtry=2)
 
@@
 gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
+                 method = "rf",
                  trControl = trainControl(method = 'none'), 
@@
 dev.off()
-```
-
-# Show performance as a function of hyperparameters
-```{r}
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-caretGrid1 <- expand.grid(interaction.depth=c(5), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-caretGrid2 <- expand.grid(interaction.depth=c(3), n.trees = 100,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-caretGrid3 <- expand.grid(interaction.depth=c(5), n.trees = 100,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-caretGrid4 <- expand.grid(interaction.depth=c(7), n.trees = 100,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-caretGrid5 <- expand.grid(interaction.depth=c(7), n.trees = 200,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-lgrids <- list(caretGrid, caretGrid1, caretGrid2, caretGrid3, caretGrid4, caretGrid5)
-f_replicate_hyper <- function(lgrids){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
-  a_array <- array(dim=c(1000, 4, 1))
-  a_array[,,1] <- temp
-  laccuracy <- vector(length = length(lgrids))
-  for(i in seq_along(lgrids))
-    laccuracy[i] <- r_star(a_array, caret_default=lgrids[[i]])
-  return(laccuracy)
-}
-
-nreplicates <- 200
-r_star_vals <- matrix(nrow = nreplicates, ncol=length(lgrids))
-for(i in 1:nreplicates){
-  if(i%%100==0)
-    print(i)
-
-  temp <- f_replicate_hyper(lgrids)
-  r_star_vals[i, ] <- temp
-}
-
-saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
-r_star_vals <- readRDS("../data/ar1_sensitivity.rds")
-
-b_sum <-
-  r_star_vals %>% 
-  as.data.frame() %>% 
-  mutate(iterate=seq(1, nreplicates, 1)) %>%
-  melt(id.vars="iterate") %>% 
-  mutate(below=if_else(value < 1, 1, 0)) %>% 
-  group_by(variable) %>% 
-  summarise(below_per=round(mean(below)*100, 1),
-            lower=quantile(value, 0.25),
-            upper=quantile(value, 0.75),
-            middle=median(value)) %>%
-  mutate(below_per=paste0(below_per, "%")) %>% 
-  mutate(variable=case_when(variable=="V1"~"int.depth=3\nn.trees=50",
-                            variable=="V2"~"int.depth=5\nn.trees=50",
-                            variable=="V3"~"int.depth=3\nn.trees=100",
-                            variable=="V4"~"int.depth=5\nn.trees=100",
-                            variable=="V5"~"int.depth=7\nn.trees=100",
-                            variable=="V6"~"int.depth=7\nn.trees=200")) %>% 
-  arrange(variable) %>% 
-  mutate(x_pos=seq(1, 6, 1)) %>% 
-  mutate(y_pos=0.95)
-
-g <- r_star_vals %>% 
-  as.data.frame() %>% 
-  mutate(iterate=seq(1, nreplicates, 1)) %>% 
-  melt(id.vars="iterate") %>% 
-  mutate(below=if_else(value < 1, 0, 1)) %>% 
-  mutate(variable=case_when(variable=="V1"~"int.depth=3\nn.trees=50",
-                            variable=="V2"~"int.depth=5\nn.trees=50",
-                            variable=="V3"~"int.depth=3\nn.trees=100",
-                            variable=="V4"~"int.depth=5\nn.trees=100",
-                            variable=="V5"~"int.depth=7\nn.trees=100",
-                            variable=="V6"~"int.depth=7\nn.trees=200")) %>% 
-  left_join(b_sum, by = "variable") %>%
-  ggplot(aes(x=as.factor(variable), y=value, colour=as.factor(below))) +
-  geom_jitter() +
-  xlab("Hyperparameters") +
-  ylab("R*") +
-  geom_hline(yintercept = 1, linetype=2) +
-  annotate("text", x = b_sum$x_pos, y = b_sum$y_pos, label = b_sum$below_per, size=6) +
-  annotate("pointrange", x = 1:6, ymin=b_sum$lower, ymax=b_sum$upper, y = b_sum$middle, colour="black", size=1) +
-  coord_flip() +
-  scale_color_grey() +
-  theme(axis.text = element_text(colour="black", size=16),
-        axis.title = element_text(colour="black", size=20),
-        legend.position = "none")
-ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
 ```

### 2020-10-12 src/s_cauchy.Rmd
added lots of things ml-related

@@
        nom=laccuracy)
-# saveRDS(a_df, "../output/cauchy_r_star.rds")
+saveRDS(a_df, "../output/cauchy_r_star_rf.rds")
 ```
@@
 ```{r}
-a_df <- readRDS("../output/cauchy_r_star.rds")
+# a_df <- readRDS("../output/cauchy_r_star.rds")
 g2 <- 
@@
 g2
+```
+
+Add in GBM R*
+```{r}
+a_df <- readRDS("../output/cauchy_r_star.rds") %>% 
+  mutate(model="gbm")
+a_df1 <- readRDS("../output/cauchy_r_star_rf.rds") %>% 
+  mutate(model="rf")
+g2 <- a_df %>%
+  bind_rows(a_df1) %>% 
+  melt(id.vars=c("iteration", "model")) %>% 
+  ggplot(aes(x=iteration, y=value, colour=as.factor(variable), linetype=as.factor(model))) +
+  geom_point() +
+  geom_line() +
+  ylab(TeX("$R^*$")) +
+  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  xlab("Iteration") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black"),
+        axis.title.y = element_blank()) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_x_log10() +
+  scale_y_continuous(limits=c(0.8, NA), breaks = seq(1, 7, 1)) +
+  ggtitle(TeX("D. Split-$R^*$"))+
+  scale_linetype_discrete("Model") +
+  guides(colour = guide_legend(nrow = 1, byrow = T),
+         linetype=guide_legend(nrow = 1, byrow = T))
+ggsave("../output/fig_9d.pdf", g2, width = 12, height = 8)
 ```

### 2020-10-12 src/s_eight_schools.Rmd
added lots of things ml-related

@@
   geom_histogram(position="identity", alpha=0.8) +
-  xlim(0.78, 1.6) +
+  xlim(0.78, 2) +
   scale_fill_grey("Series", labels=c("centered", "non-centered")) +
@@
   scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlim(0.78, 1.6) +
+  xlim(0.78, 2) +
   xlab(TeX("$R*$")) +
@@
 
-pdf("../output/eight_schools.pdf", width = 12, height = 4)
+pdf("../output/fig_10.pdf", width = 12, height = 4)
 multiplot(g1, g2, g3, cols = 3)
@@
 g1 <- plot_r_star_quantiles(1, x, nsim=40, split_chains = F) +
-  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6)) +
+  scale_y_continuous(limits = c(0.6, 2.6)) +
   theme(axis.title.y = element_blank(),
@@
 g2 <- plot_r_star_quantiles(2, x, nsim=40, split_chains = F) +
-  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6)) +
+  scale_y_continuous(limits = c(0.6, 2.6)) +
   theme(axis.title.y = element_blank(),
@@
 
-pdf("../output/eight_schools_r_star_quantiles1.pdf", width = 6, height = 4)
+pdf("../output/fig_11.pdf", width = 6, height = 4)
 plot_grid(gb, g1, ga, g2, nrow = 2, rel_widths = c(1, 0.4), rel_heights = c(0.9, 1))

### 2020-10-12 src/s_hyperparameters.Rmd
added lots of things ml-related

@@
 
-saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
-r_star_vals <- readRDS("../data/ar1_sensitivity.rds")
-
-b_sum <-
-  r_star_vals %>% 
-  as.data.frame() %>% 
-  mutate(iterate=seq(1, nreplicates, 1)) %>%
-  melt(id.vars="iterate") %>% 
-  mutate(below=if_else(value < 1, 1, 0)) %>% 
-  group_by(variable) %>% 
-  summarise(below_per=round(mean(below)*100, 1),
-            lower=quantile(value, 0.25),
-            upper=quantile(value, 0.75),
-            middle=median(value)) %>%
-  mutate(below_per=paste0(below_per, "%")) %>% 
-  mutate(x_pos=seq(1, nrow(m_parameters), 1)) %>% 
-  mutate(y_pos=0.95)
-
-g <- b_sum %>% 
-  cbind(m_parameters) %>% 
+# saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
+g <- readRDS("../data/hypers_ar1.rds")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
   ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
@@
   guides(colour = guide_legend(nrow = 2, byrow = T))
-ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
+ggsave("../output/hypers_ar1.pdf", g, width = 12, height = 8)
 ```
@@
 
-r_star_vals %>% 
-  melt(id.vars=c("n.trees", "int.depth")) %>% 
-  group_by(n.trees, int.depth) %>% 
+readRDS("../data/hypers_8_schools.rds")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
   summarise(lower=quantile(value, 0.25),
-            upper=quantile(value, 0.75),
-            middle=median(value)) %>% 
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
   ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
@@
         legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.98, NA)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
   geom_hline(yintercept = 1, linetype=2) +
@@
 r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_cauchy)
+
+readRDS("../data/hypers_cauchy.rds")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# trees") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
 ```
+
+# RF hyperparameters
+AR1
+```{r}
+temp <- readRDS("../data/hypers_rf_ar1.rds")$r_star
+temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
+```
+
+8 schools
+```{r}
+temp <- readRDS("../data/hypers_rf_8_schools.rds")$r_star
+temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
+```
+Cauchy
+```{r}
+temp <- readRDS("../data/hypers_rf_cauchy.rds")$r_star
+temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
+```
+
+Normal
+```{r}
+temp <- readRDS("../data/hypers_rf_")$r_star
+temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
+```
+
+# xgbtree
+AR1
+```{r}
+temp <- readRDS("../data/hypers_xgbtree_ar1.rds")$r_star
+temp %>%
+  select(-one_of("gamma", "colsample_bytree", "min_child_weight", "subsample")) %>% 
+  melt(id.vars=c("nrounds", "max_depth", "eta")) %>%
+  group_by(nrounds, max_depth, eta) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(nrounds), y=middle, colour=as.factor(max_depth), group=as.factor(max_depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# rounds") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black")) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  facet_wrap(~eta)
+```
+
+```{r}
+temp <- readRDS("../data/hypers_xgbtree_8_schools.rds")$r_star
+temp %>%
+  select(-one_of("gamma", "colsample_bytree", "min_child_weight", "subsample")) %>% 
+  melt(id.vars=c("nrounds", "max_depth", "eta")) %>%
+  group_by(nrounds, max_depth, eta) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(nrounds), y=middle, colour=as.factor(max_depth), group=as.factor(max_depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# rounds") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black")) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  facet_wrap(~eta)
+```

### 2020-10-12 src/s_ml.Rmd
added lots of things ml-related

@@
           f_data_ar1)
-temp_df <- readRDS("../data/ml_comp_hypers_set_ar1.rds")
-temp_df %>%
+
+temp_df <- readRDS("../data/ml_comp_hypers_set_ar1.rds") %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, orders))
+g1 <- temp_df %>%
   ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("R*") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("A.")
+g2 <- temp_df %>%
+  ggplot(aes(x=as.factor(method), y=time)) +
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("Time") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("B.")
+pdf("../output/ml_comparison_ar1.pdf", width = 12, height = 8)
+multiplot(g1, g2, cols = 2)
+dev.off()
 ```
@@
 
-temp_df <- readRDS("../data/ml_comp_hypers_set_8_schools.rds")
-temp_df %>%
+temp_df <- readRDS("../data/ml_comp_hypers_set_8_schools.rds") %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, orders))
+g1 <- temp_df %>%
   ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-temp_df %>%
-  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
-  geom_jitter()
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("R*") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("A.")
+g2 <- temp_df %>%
+  ggplot(aes(x=as.factor(method), y=time)) +
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("Time") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("B.")
+pdf("../output/ml_comparison_8_schools.pdf", width = 12, height = 8)
+multiplot(g1, g2, cols = 2)
+dev.off()
 ```
@@
           f_data_cauchy)
-temp_df <- readRDS("../data/ml_comp_hypers_set_cauchy.rds")
-temp_df %>%
+temp_df <- readRDS("../data/ml_comp_hypers_set_cauchy.rds") %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, orders))
+g1 <- temp_df %>%
   ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-temp_df %>%
-  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
-  geom_jitter()
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("R*") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("A.")
+g2 <- temp_df %>%
+  ggplot(aes(x=as.factor(method), y=time)) +
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("Time") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("B.")
+pdf("../output/ml_comparison_cauchy.pdf", width = 12, height = 8)
+multiplot(g1, g2, cols = 2)
+dev.off()
 ```
@@
 
-temp_df <- readRDS("../data/ml_comp_hypers_set_normal_250.rds")
-temp_df %>%
+orders <- c("rf", "xgbTree", "gbm", "knn", "multinom", "svmLinear")
+
+temp_df <- readRDS("../data/ml_comp_hypers_set_normal_250.rds") %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, orders))
+g1 <- temp_df %>%
   ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-temp_df %>%
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("R*") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("A.")
+g2 <- temp_df %>%
   ggplot(aes(x=as.factor(method), y=time)) +
-  geom_boxplot()
-
-temp_df %>%
-  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
-  geom_jitter()
+  geom_boxplot() +
+  xlab("Method") +
+  ylab("Time") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  ggtitle("B.")
+pdf("../output/ml_comparison_mvt_normal_250.pdf", width = 12, height = 8)
+multiplot(g1, g2, cols = 2)
+dev.off()
 ```

### 2020-10-12 src/s_xgbtree_vs_rf.Rmd
added lots of things ml-related

---
title: "Comparing xgbtree vs rf (default parameterisations)"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

Generate a large correlation matrix
```{r}
rlkj <- function(n, nu, d){
  r_list <- vector(length = n, mode = 'list')
  for(i in 1:n){
    if(d==1)
      r = as.array(1)
    else if(d==2){
      rho <- 2 * rbeta(1, nu, nu) - 1
      r <- matrix(c(1, rho, rho, 1), ncol = 2)
    }else{
      beta <- nu + (d - 2) / 2
      u <- rbeta(1, beta, beta)
      r_12 <- 2 * u - 1
      r <- matrix(c(1, r_12, r_12, 1), ncol = 2)
      for(m in 2:(d - 1)){
        beta <- beta - 0.5
        y <- rbeta(1, m / 2, beta)
        a <- rnorm(m)
        anorm <- sqrt(sum(a^2))
        u <- a / anorm
        w <- sqrt(y) * u
        A <- chol(r)
        z <- w%*%A
        r <-cbind(r, t(z))
        r <- rbind(r, c(z, 1))
      }
    }
    r_list[[i]] <- r
  }
  return(r_list)
}
# d <- 32
# A0 <- rlkj(1, 1, d)[[1]]
# A1 <- rlkj(1, 1, d)[[1]]
# saveRDS(list(A0=A0, A1=A1), "../data/xgbTree_vs_rf_normals.rds")
temp <- readRDS("../data/xgbTree_vs_rf_normals.rds")
A0 <- temp$A0
A1 <- temp$A1

library(mvtnorm)
f_data_normals <- function(A0, A1, d, niter=2000){
  x <- array(dim=c(niter, 4, d))
  if(d > 1){
    for(i in 1:3)
      x[, i, ] <- rmvnorm(niter, rep(0, d), A0[1:d, 1:d])
    x[ , 4, ] <- rmvnorm(niter, rep(0, d), A1[1:d, 1:d])
  }else{
    for(i in 1:3)
      x[, i, ] <- rnorm(niter, rep(0, d), A0[1:d, 1:d])
    x[ , 4, ] <- rnorm(niter, rep(0, d), A1[1:d, 1:d])
  }
  return(x)
}

f_compare_rf_xgbtree <- function(d, A0, A1){
  lrstar <- vector(length=2)
  ltimes <- vector(length=2)
  x <- f_data_normals(A0, A1, d)
  start <- Sys.time()
  lrstar[1] <- r_star(x, method="rf")
  end <- Sys.time()
  ltimes[1] <- end-start
  start <- Sys.time()
  lrstar[2] <- r_star(x, method="xgbTree")
  end <- Sys.time()
  ltimes[2] <- end-start
  return(list(rstar=lrstar, time=ltimes))
}

f_compare_iter <- function(nreplicate, d, A0, A1){
  m_res <- matrix(nrow=nreplicate, ncol=6)
  for(i in 1:nreplicate){
    temp <- f_compare_rf_xgbtree(d, A0, A1)
    rstars <- temp$rstar
    times <- temp$time
    m_res[i, ] <- c(i, d, rstars[1], rstars[2], times[1], times[2])
  }
  colnames(m_res) <- c("iter", "d", "rstar_rf", "rstar_xgbTree", "time_rf", "time_xgbTree")
  m_res <- m_res %>% 
    as.data.frame()
  return(m_res)
}

nreplicates <- 20
ds <- c(1, 2, 4, 8, 16, 32)
for(i in seq_along(ds)){
  print(i)
  temp_df <- f_compare_iter(nreplicates, ds[i], A0, A1)
  if(i==1)
    big_df <- temp_df
  else
    big_df <- big_df %>% bind_rows(temp_df)
}
saveRDS(big_df, "../data/xgbTree_vs_rf_normals_results.rds")

big_df %>%
  select(-contains("time")) %>% 
  melt(id.vars=c("iter", "d")) %>% 
  ggplot(aes(x=as.factor(d), y=value, fill=as.factor(variable))) +
  geom_boxplot(position = position_dodge2(width=0.3)) +
  geom_hline(yintercept = 1, linetype=2)
```

Same but increase fatness of tails
```{r}
library(LaplacesDemon)
f_data_student_ts <- function(nu1, nu2, d, A0, niter=2000){
  x <- array(dim=c(niter, 4, d))
  # ensures variance stays same
  sigma1 <- A0[1:d, 1:d] * (nu1 - 2) / nu1
  sigma2 <- A0[1:d, 1:d] * (nu2 - 2) / nu2
  if(d > 1){
    for(i in 1:3)
      x[, i, ] <- mvtnorm::rmvt(niter, sigma=sigma1, df=nu1, delta=rep(0, d))
    x[ , 4, ] <- mvtnorm::rmvt(niter, sigma=sigma2, df=nu2, delta=rep(0, d))
  }else{
    for(i in 1:3)
      x[, i, ] <- LaplacesDemon::rst(niter, 0, sqrt(sigma1), nu1)
    x[ , 4, ] <- LaplacesDemon::rst(niter, 0, sqrt(sigma2), nu2)
  }
  return(x)
}

# x <- f_data_student_ts(3, 100, 1, A0, niter=10000)
library(e1071)   
# kurtosis(x[, 2, 1])
# kurtosis(x[, 1, 1])
# kurtosis(x[, 3, 1])
# kurtosis(x[, 4, 1])
# sd(x[, 2, 1])
# sd(x[, 1, 1])
# sd(x[, 3, 1])
# sd(x[, 4, 1])

f_compare_rf_xgbtree_student_ts <- function(nu1, nu2, d, A0){
  lrstar <- vector(length=2)
  ltimes <- vector(length=2)
  x <- f_data_student_ts(nu1, nu2, d, A0)
  start <- Sys.time()
  lrstar[1] <- r_star(x, method="rf")
  end <- Sys.time()
  ltimes[1] <- end-start
  start <- Sys.time()
  lrstar[2] <- r_star(x, method="xgbTree")
  end <- Sys.time()
  ltimes[2] <- end-start
  return(list(rstar=lrstar, time=ltimes))
}

f_compare_iter_student_t <- function(nreplicate, nu1, nu2, d, A0){
  m_res <- matrix(nrow=nreplicate, ncol=8)
  for(i in 1:nreplicate){
    temp <- f_compare_rf_xgbtree_student_ts(nu1, nu2, d, A0)
    temp <- f_compare_rf_xgbtree(d, A0, A1)
    rstars <- temp$rstar
    times <- temp$time
    m_res[i, ] <- c(i, d, nu1, nu2, rstars[1], rstars[2], times[1], times[2])
  }
  colnames(m_res) <- c("iter", "d", "nu1", "nu2", "rstar_rf", "rstar_xgbTree", "time_rf", "time_xgbTree")
  m_res <- m_res %>% 
    as.data.frame()
  return(m_res)
}

nreplicates <- 20
nu1 <- 3
nu2 <- c(4, 8, 16, 32)
ds <- c(1, 2, 4, 8, 16, 32)
m_parameters <- expand_grid(nu2, ds)

for(i in 1:nrow(m_parameters)){
  print(i)
  temp_df <- f_compare_iter_student_t(nreplicates, nu1, m_parameters$nu2[i], m_parameters$ds[i], A0)
  if(i==1)
    big_df <- temp_df
  else
    big_df <- big_df %>% bind_rows(temp_df)
}
saveRDS(big_df, "../data/xgbTree_vs_rf_studentts_results.rds")
```

### 2020-10-12 src/r_star_monitor.R

@@ -22,7 +22,8 @@ library(tidyverse)
 #' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
-r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL){
+r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, method=NULL,
+                   uncertainty=F){
   
   if(split_chains)
     x <- split_data(x)
@@ -54,102 +55,45 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL){
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]
   
-  if(is.null(caret_default))
-    caretGrid <- expand.grid(interaction.depth=c(3), 
-                             n.trees = 50,
-                             shrinkage=c(0.1),
-                             n.minobsinnode=10)
+  if(is.null(caret_default)&&method=="xgbTree")
+    caretGrid <- tibble(nrounds = c(100),
+                        max_depth = c(4),
+                        eta = c(.4),
+                        gamma = 0,
+                        colsample_bytree = .7,
+                        min_child_weight = 1,
+                        subsample = 0.8)
+  else if(is.null(caret_default))
+    caretGrid <- tibble(mtry=floor(sqrt(nparams)))
   else
     caretGrid <- expand.grid(caret_default)
-  
-  gbmFit <- train(chain ~ ., data = training_data, 
-                   method = "gbm",
-                   trControl = trainControl(method = 'none'), 
-                   tuneGrid = caretGrid, verbose=FALSE)
-  plda <- predict(object=gbmFit, newdata=testing_data)
-  a_accuracy <- 
-    tibble(predicted=plda, actual=testing_data$chain) %>%
-    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-    summarise(mean(correct)) %>% 
-    pull()
-  return(a_accuracy * n_distinct(training_data$chain))
-}
+  if(is.null(method))
+    method <- "rf"
 
-#' Compute estimate of R* distribution using a gradient-boosted model (GBM)
-#'
[…]

### 2020-10-12 src/s_jobScheduler_hypers_rf.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+
+chmod +x *.R
+chmod +x *.sh
+
+
+# Rscript s_run_hyperparameters_rf.R 1 200
+Rscript s_run_hyperparameters_rf.R 2 50
+# Rscript s_run_hyperparameters_rf.R 3 20
+Rscript s_run_hyperparameters_rf.R 4 20
+

### 2020-10-12 src/s_jobScheduler_hypers_xgbtree.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+chmod +x *.R
+chmod +x *.sh
+
+
+Rscript s_run_hyperparameters_xgbtree.R 1 200
+Rscript s_run_hyperparameters_xgbtree.R 2 50
+Rscript s_run_hyperparameters_xgbtree.R 3 20
+

### 2020-10-12 src/s_jobScheduler_ml_hypers_set.sh

@@ -4,9 +4,9 @@ chmod +x *.R
 chmod +x *.sh
 
 
-Rscript s_run_ml_hypers_set.R 1 100
-Rscript s_run_ml_hypers_set.R 2 100
-Rscript s_run_ml_hypers_set.R 3 50
-Rscript s_run_ml_hypers_set.R 4 30
+# Rscript s_run_ml_hypers_set.R 1 100
+# Rscript s_run_ml_hypers_set.R 2 100
+# Rscript s_run_ml_hypers_set.R 3 50
+# Rscript s_run_ml_hypers_set.R 4 30
 Rscript s_run_ml_hypers_set.R 5 10
 

### 2020-10-12 src/s_run_hyperparameters_rf.R

@@ -1,0 +1,120 @@
+setwd("/Volumes/Samsung1.5TB/Github/ml-mcmc-convergence/src")
+rm(list=ls())
+library(tidyverse)
+library(reshape2)
+library(rstan)
+library(latex2exp)
+library(caret)
+library(gbm)
+options(mc.cores=4)
+rstan_options(auto_write = TRUE)
+source("monitornew.R")
+source("r_star_monitor.R")
+source("eight_schools.data.R")
+eight_schools <- list(J=J, y=y, sigma=sigma)
+model_cp <- stan_model("eight_schools_cp.stan")
+model_ncp <- stan_model("eight_schools_ncp.stan")
+
+
+# helper functions ------
+f_ar1 <- function(rho, sigma, L){
+  x <- vector(length = L)
+  x[1] <- rnorm(1, 0, sd=sigma)
+  for(i in 2:L)
+    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
+  return(x)
+}
+
+f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, 4] <- z
+  return(x)
+}
+
+f_data_ar1 <- function(){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  return(a_array)
+}
+
+f_replicate_hyper <- function(m_parameters, f_data_generator){
+  
+  a_array <- f_data_generator()
+  
+  laccuracy <- vector(length = nrow(m_parameters))
+  for(i in 1:nrow(m_parameters))
+    laccuracy[i] <- r_star(a_array,
+                           caret_default=expand.grid(mtry=m_parameters$mtry[i]))
+  return(laccuracy)
+}
+
+
+f_replicate_r_star <- function(nreplicates, m_parameters, f_data_generator){
+  r_star_vals <- matrix(nrow = nreplicates, ncol=nrow(m_parameters))
+  for(i in 1:nreplicates){
+    print(i)
+    
+    temp <- f_replicate_hyper(m_parameters, f_data_generator)
+    r_star_vals[i, ] <- temp
+  }
[…]

### 2020-10-12 src/s_run_hyperparameters_xgbtree.R

@@ -1,0 +1,132 @@
+setwd("/Volumes/Samsung1.5TB/Github/ml-mcmc-convergence/src")
+rm(list=ls())
+library(tidyverse)
+library(reshape2)
+library(rstan)
+library(latex2exp)
+library(caret)
+library(gbm)
+options(mc.cores=4)
+rstan_options(auto_write = TRUE)
+source("monitornew.R")
+source("r_star_monitor.R")
+source("eight_schools.data.R")
+eight_schools <- list(J=J, y=y, sigma=sigma)
+model_cp <- stan_model("eight_schools_cp.stan")
+model_ncp <- stan_model("eight_schools_ncp.stan")
+
+
+# helper functions ------
+f_ar1 <- function(rho, sigma, L){
+  x <- vector(length = L)
+  x[1] <- rnorm(1, 0, sd=sigma)
+  for(i in 2:L)
+    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
+  return(x)
+}
+
+f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
+  x <- matrix(nrow = L, ncol = 4)
+  for(i in 1:3)
+    x[, i] <- f_ar1(rho, sigma, L)
+  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
+  x[, 4] <- z
+  return(x)
+}
+
+f_data_ar1 <- function(){
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
+  a_array <- array(dim=c(1000, 4, 1))
+  a_array[,,1] <- temp
+  return(a_array)
+}
+
+f_replicate_hyper <- function(m_parameters, f_data_generator){
+  
+  a_array <- f_data_generator()
+  
+  laccuracy <- vector(length = nrow(m_parameters))
+  for(i in 1:nrow(m_parameters))
+    laccuracy[i] <- r_star(a_array,
+                           caret_default=tibble(nrounds = m_parameters$nrounds[i],
+                                                max_depth = m_parameters$max_depth[i],
+                                                eta = m_parameters$eta[i],
+                                                gamma = m_parameters$gamma[i],
+                                                colsample_bytree = m_parameters$colsample_bytree[i],
+                                                min_child_weight = m_parameters$min_child_weight[i],
+                                                subsample = m_parameters$subsample[i]),
+                           method="xgbTree")
+  return(laccuracy)
+}
+
+
[…]

### 2020-10-12 src/s_run_ml_hypers_set.R

@@ -145,7 +145,7 @@ r_star_ml_variety <- function(x, method_, caretGrid){
                     trControl = trainControl(method = 'none'),
                     method=method_,
                     tuneGrid = caretGrid,
-                    MaxNWts=10000
+                    MaxNWts=100000
     )
   plda <- predict(object=gbmFit, newdata=testing_data)
   a_accuracy <- 

### 2020-10-12 src/s_ar1.Rmd
started adding ar1 results

@@
 
-Calculate R* and Rhat across 1000 replicates
+Calculate R* and Rhat across 200 replicates
 ```{r}
@@
   a_array[,,1] <- temp
-  a_accuracy <- r_star(a_array)
+  a_accuracy <- r_star(a_array, method = "xgbTree")
   mon <- monitor_extra(split_data(a_array))
@@
 
-nreplicates <- 100
+nreplicates <- 200
 r_star_vals <- vector(length = nreplicates)
@@
 
-# saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/ar1_r_star_hat.rds")
+saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/ar1_r_star_hat.rds")
 ```
@@
 ```{r}
-# reses <- readRDS("../data/ar1_r_star_hat.rds")
+reses <- readRDS("../data/ar1_r_star_hat.rds")
 r_star_vals <- reses$r_star 
@@
   mutate(chain=as.factor(chain))
-
-caretGrid <- tibble(mtry=2)
+caretGrid <- tibble(nrounds = c(100),
+                        max_depth = c(4),
+                        eta = c(.4),
+                        gamma = 0,
+                        colsample_bytree = .7,
+                        min_child_weight = 1,
+                        subsample = 0.8)
 
@@
 gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "rf",
+                 method = "xgbTree",
                  trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
+                 verbose=FALSE, tuneGrid = caretGrid)
 
@@
 a_array[,,2] <- temp1
-unmixed <- r_star_dist(a_array)
+unmixed <- r_star(a_array, uncertainty = T, method = "xgbTree")
 
@@
 a_array[,,2] <- temp1
-mixed <- r_star_dist(a_array)
+mixed <- r_star(a_array, uncertainty = T, method = "xgbTree")
 
@@
   a_array[,,1] <- temp
-  a_accuracy <- r_star(a_array)
+  a_accuracy <- r_star(a_array, method="xgbTree")
   mon <- monitor_extra(split_data(a_array))

### 2020-10-12 src/s_hyperparameters.Rmd
started adding ar1 results

@@
 ```
+
+```{r}
+temp <- readRDS("../data/hypers_xgbtree_cauchy.rds")$r_star
+temp %>%
+  select(-one_of("gamma", "colsample_bytree", "min_child_weight", "subsample")) %>% 
+  melt(id.vars=c("nrounds", "max_depth", "eta")) %>%
+  group_by(nrounds, max_depth, eta) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(nrounds), y=middle, colour=as.factor(max_depth), group=as.factor(max_depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# rounds") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black")) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  facet_wrap(~eta)
+```

### 2020-10-12 src/s_xgbtree_vs_rf.Rmd
started adding ar1 results

@@
 
+# work out optimal classification rates
+f_optimal_normals <- function(niter, A0, A1, d){
+  x <- f_data_normals(A0, A1, d, niter)
+  A0_temp <- A0[1:d, 1:d]
+  A1_temp <- A1[1:d, 1:d]
+  if(d > 1){
+    f_first <- function(x) dmvnorm(x, rep(0, d), A0_temp)
+    f_second <- function(x) dmvnorm(x, rep(0, d), A1_temp)
+  }else{
+    f_first <- function(x) dnorm(x, 0, A0_temp)
+    f_second <- function(x) dnorm(x, 0, A1_temp)
+  }
+  m_class <- matrix(nrow = dim(x)[1], ncol = dim(x)[2])
+  for(i in 1:dim(x)[1]){
+    for(j in 1:dim(x)[2]){
+      prob0 <- f_first(x[i, j,])
+      prob1 <- f_second(x[i, j,])
+      if(prob0 > prob1)
+        m_class[i, j] <- 0
+      else
+        m_class[i, j] <- 1
+    }
+  }
+  a0 <- 1/3 * 1/3 * (mean(m_class[, 1]==0) + mean(m_class[, 2]==0) + mean(m_class[, 3]==0))
+  a1 <- mean(m_class[, 4]==1)
+  return(1/4 * (3 * a0 + a1) / 0.25)
+}
+
+r_star_opts <- map_dbl(ds[2:length(ds)], ~f_optimal_normals(10000, A0, A1, .))
+r_star_opts <- c(1, r_star_opts)
+r_star_opts <- tibble(d=ds, optimal=r_star_opts)
+temp_df <- 
   big_df %>%
+  left_join(r_star_opts) %>% 
   select(-contains("time")) %>% 
   melt(id.vars=c("iter", "d")) %>% 
-  ggplot(aes(x=as.factor(d), y=value, fill=as.factor(variable))) +
-  geom_boxplot(position = position_dodge2(width=0.3)) +
-  geom_hline(yintercept = 1, linetype=2)
+  group_by(variable, d) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  ungroup() %>% 
+  mutate(variable=as.character(variable)) %>% 
+  mutate(variable=if_else(variable=="rstar_rf", "rf", if_else(variable=="rstar_xgbTree", "xgbtree", "optimal")))
+saveRDS(temp_df, "../data/xgbTree_vs_rf_normals_results_optimals.rds")
+temp_df <- readRDS("../data/xgbTree_vs_rf_normals_results_optimals.rds")
+g <- ggplot(data=filter(temp_df, variable!="optimal"),
+       aes(x=as.factor(d), y=middle, colour=as.factor(variable))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper),
+                  position = position_dodge2(width=0.3)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  geom_point(data=filter(temp_df, variable=="optimal")) +
+  scale_color_brewer("Method", palette = "Dark2") +
+  xlab("# dimensions") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"))
+ggsave("../output/xgbtree_vs_rf_normal.pdf", g, width = 10, height = 6)
 ```
@@
     temp <- f_compare_rf_xgbtree_student_ts(nu1, nu2, d, A0)
-    temp <- f_compare_rf_xgbtree(d, A0, A1)
     rstars <- temp$rstar
@@
 saveRDS(big_df, "../data/xgbTree_vs_rf_studentts_results.rds")
+
+f_optimal_studentts <- function(niter, nu1, nu2, d, A0){
+  x <- f_data_student_ts(nu1, nu2, d, A0, niter = niter)
+  sigma1 <- A0[1:d, 1:d] * (nu1 - 2) / nu1
+  sigma2 <- A0[1:d, 1:d] * (nu2 - 2) / nu2
+  if(d > 1){
+    f_first <- function(x) dmvt(x, rep(0, d), sigma1, nu1)
+    f_second <- function(x) dmvt(x, rep(0, d), sigma2, nu2)
+  }else{
+    f_first <- function(x) dst(x, 0, sigma1, nu1)
+    f_second <- function(x) dst(x, 0, sigma2, nu2)
+  }
+  m_class <- matrix(nrow = dim(x)[1], ncol = dim(x)[2])
+  for(i in 1:dim(x)[1]){
+    for(j in 1:dim(x)[2]){
+      prob0 <- f_first(x[i, j,])
+      prob1 <- f_second(x[i, j,])
+      if(prob0 > prob1)
+        m_class[i, j] <- 0
+      else
+        m_class[i, j] <- 1
+    }
+  }
+  a0 <- 1/3 * 1/3 * (mean(m_class[, 1]==0) + mean(m_class[, 2]==0) + mean(m_class[, 3]==0))
+  a1 <- mean(m_class[, 4]==1)
+  return(1/4 * (3 * a0 + a1) / 0.25)
+}
+
+r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(1000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))
+
+r_star_opts <- tibble(d=m_parameters$ds, nu2=m_parameters$nu2, optimal=r_star_opts1)
+temp_df <- 
+  big_df %>%
+  left_join(r_star_opts) %>% 
+  select(-contains("time")) %>% 
+  select(-one_of("nu1")) %>% 
+  melt(id.vars=c("iter", "d", "nu2")) %>% 
+  group_by(variable, d, nu2) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  ungroup() %>% 
+  mutate(variable=as.character(variable)) %>% 
+  mutate(variable=if_else(variable=="rstar_rf", "rf", if_else(variable=="rstar_xgbTree", "xgbtree", "optimal")))
+saveRDS(temp_df, "../data/xgbTree_vs_rf_studentts_results_optimal.rds")
+temp_df <- readRDS("../data/xgbTree_vs_rf_studentts_results_optimal.rds")
+g2 <- ggplot(data=filter(temp_df, variable!="optimal"),
+       aes(x=as.factor(nu2), y=middle, colour=as.factor(variable))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper),
+                  position = position_dodge2(width=0.3)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  geom_point(data=filter(temp_df, variable=="optimal")) +
+  scale_color_brewer("Method", palette = "Dark2") +
+  xlab("# degrees of freedom") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black")) +
+  facet_wrap(~d)
+ggsave("../output/xgbtree_vs_rf_studentt.pdf", g2, width = 12, height = 8)
 ```

### 2020-10-12 src/r_star_monitor.R

@@ -23,7 +23,7 @@ library(tidyverse)
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
 r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, method=NULL,
-                   uncertainty=F){
+                   uncertainty=F, nsim=1000){
   
   if(split_chains)
     x <- split_data(x)
@@ -55,6 +55,8 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, 
   training_data <- r[rand_samples, ]
   testing_data <- r[-rand_samples, ]
   
+  if(is.null(method))
+    method <- "rf"
   if(is.null(caret_default)&&method=="xgbTree")
     caretGrid <- tibble(nrounds = c(100),
                         max_depth = c(4),
@@ -67,16 +69,14 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, 
     caretGrid <- tibble(mtry=floor(sqrt(nparams)))
   else
     caretGrid <- expand.grid(caret_default)
-  if(is.null(method))
-    method <- "rf"
 
-  gbmFit <- train(chain ~ ., data = training_data, 
+  fit <- train(chain ~ ., data = training_data, 
                     method = method,
                     trControl = trainControl(method = 'none'), 
                     tuneGrid = caretGrid, verbose=FALSE)
   
   if(uncertainty){
-    plda <- predict(object=gbmFit1, newdata=testing_data, type = "prob")
+    plda <- predict(object=fit, newdata=testing_data, type = "prob")
     
     mAccuracy <- matrix(nrow = nrow(plda),
                         ncol = nsim)
@@ -86,7 +86,7 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, 
     }
     return(colMeans(mAccuracy) * n_distinct(testing_data$chain))
   } else{
-    plda <- predict(object=gbmFit, newdata=testing_data)
+    plda <- predict(object=fit, newdata=testing_data)
     a_accuracy <- 
       tibble(predicted=plda, actual=testing_data$chain) %>%
       mutate(correct=if_else(predicted==actual, 1, 0)) %>% 

### 2020-10-12 src/s_jobScheduler_hypers_xgbtree.sh

@@ -5,6 +5,6 @@ chmod +x *.sh
 
 
 Rscript s_run_hyperparameters_xgbtree.R 1 200
-Rscript s_run_hyperparameters_xgbtree.R 2 50
-Rscript s_run_hyperparameters_xgbtree.R 3 20
+# Rscript s_run_hyperparameters_xgbtree.R 2 50
+# Rscript s_run_hyperparameters_xgbtree.R 3 20
 

### 2020-10-12 src/s_run_hyperparameters_xgbtree.R

@@ -35,8 +35,8 @@ f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
 }
 
 f_data_ar1 <- function(){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
-  a_array <- array(dim=c(1000, 4, 1))
+  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 2000)
+  a_array <- array(dim=c(2000, 4, 1))
   a_array[,,1] <- temp
   return(a_array)
 }

### 2020-10-15 src/s_ar1.Rmd
went back to gbm because of xgbtree poor results in trends

@@
 f_replicate <- function(){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 2000)
+  temp <- f_generate_lower_var_four(1/4, 0.3, 1, 2000)
   a_array <- array(dim=c(2000, 4, 1))
@@
 # add in a column of noise since gbm requires >1 variables
-full_data <- f_generate_lower_var_four(1/3, 0.3, 1, 2000) %>% 
+full_data <- f_generate_lower_var_four(1/4, 0.3, 1, 2000) %>% 
   melt() %>% 
@@
 ```{r}
-temp <- f_generate_lower_var_four(1/3, 0.3, 1, 2000) 
+temp <- f_generate_lower_var_four(1/4, 0.3, 1, 2000) 
 temp1 <- matrix(rnorm(8000), nrow = 2000)
@@
 f_replicate_n <- function(n_chains){
-  temp <- f_generate_lower_var_n(n_chains, 1/3, 0.3, 1, 2000)
+  temp <- f_generate_lower_var_n(n_chains, 1/4, 0.3, 1, 2000)
   a_array <- array(dim=c(2000, n_chains, 1))
@@
 nreplicates <- 50
-lnum_chains <- c(2, 4, 6, 8, 16, 32)
+lnum_chains <- c(2, 4, 6, 8, 16, 32, 64)
 mResults <- matrix(nrow = nreplicates * length(lnum_chains),
@@
 colnames(mResults) <- c("nchains", "iterate", "R*", "split-Rhat")
-mResults <- as.data.frame(mResults) %>% 
-  filter(nchains < 64)
+mResults <- as.data.frame(mResults)
 saveRDS(mResults, "../data/ar1_numchains.rds")
-mResults <- readRDS("../data/ar1_numchains.rds") %>% 
-  filter(iterate<=50)
+mResults <- readRDS("../data/ar1_numchains.rds")
 b_sum <- mResults %>% 
@@
 ggsave("../output/ar1_numchains.pdf", g, width = 8, height = 4)
-
 ```
@@
   print(j)
-  x <- f_generate_lower_var_four(1/3, 0.3, 1, lsample_sizes[j])
+  x <- f_generate_lower_var_four(1/4, 0.3, 1, lsample_sizes[j])
   for(i in 1:nreplicates){

### 2020-10-15 src/s_cauchy.Rmd
went back to gbm because of xgbtree poor results in trends

@@
 saveRDS(a_df, "../output/cauchy_r_star_rf.rds")
+
+fit <- readRDS("../output/fit_nominal_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+
+laccuracy <- future_map_dbl(iter, function(i) r_star(x[1:i, , ], method="xgbTree"))
+
+fit <- readRDS("../output/fit_alt_50000.rds")
+x <- rstan::extract(fit, permuted=F, pars="x")
+
+laccuracy1 <- future_map_dbl(iter, function(i) r_star(x[1:i, , ], method="xgbTree"))
+
+a_df <- 
+  tibble(iteration=iter,
+         alt=laccuracy1,
+       nom=laccuracy)
+saveRDS(a_df, "../output/cauchy_r_star_xgbtree.rds")
 ```
@@
 ```{r}
-a_df <- readRDS("../output/cauchy_r_star.rds") %>% 
+a_df <- readRDS("../output/cauchy_r_star_xgbtree.rds") %>% 
   mutate(model="gbm")

### 2020-10-15 src/s_eight_schools.Rmd
went back to gbm because of xgbtree poor results in trends

@@
 ```
-
+# Random forest
 ## Break each chain into two then make flattened data
@@
 ```{r}
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
 x <- rstan::extract(fit_cp, permuted=F)
@@
 # Fit gradient-boosed model
-gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
+fit <- train(chain ~ ., data = training_data, 
+                 method = "rf",
                  trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
-varImp(gbmFit1)
+                 verbose=FALSE)
+varImp(fit)
 ```
@@
 ```
+# xgbTree
 

### 2020-10-15 src/s_gbm_vs_rf.Rmd
went back to gbm because of xgbtree poor results in trends

--- 2020-10-12 src/s_xgbtree_vs_rf.Rmd
@@
 ---
-title: "Comparing xgbtree vs rf (default parameterisations)"
+title: "Comparing gbm vs rf (default parameterisations)"
 output: html_notebook
@@
 
-f_compare_rf_xgbtree <- function(d, A0, A1){
+f_compare_rf_gbm <- function(d, A0, A1){
   lrstar <- vector(length=2)
@@
   start <- Sys.time()
-  lrstar[2] <- r_star(x, method="xgbTree")
+  lrstar[2] <- r_star(x, method="gbm")
   end <- Sys.time()
@@
   for(i in 1:nreplicate){
-    temp <- f_compare_rf_xgbtree(d, A0, A1)
+    temp <- f_compare_rf_gbm(d, A0, A1)
     rstars <- temp$rstar
@@
   }
-  colnames(m_res) <- c("iter", "d", "rstar_rf", "rstar_xgbTree", "time_rf", "time_xgbTree")
+  colnames(m_res) <- c("iter", "d", "rstar_rf", "rstar_gbm", "time_rf", "time_gbm")
   m_res <- m_res %>% 
@@
 }
-saveRDS(big_df, "../data/xgbTree_vs_rf_normals_results.rds")
+saveRDS(big_df, "../data/gbm_vs_rf_normals_results.rds")
 
+# work out optimal classification rates
+f_optimal_normals <- function(niter, A0, A1, d){
+  x <- f_data_normals(A0, A1, d, niter)
+  A0_temp <- A0[1:d, 1:d]
+  A1_temp <- A1[1:d, 1:d]
+  if(d > 1){
+    f_first <- function(x) dmvnorm(x, rep(0, d), A0_temp)
+    f_second <- function(x) dmvnorm(x, rep(0, d), A1_temp)
+  }else{
+    f_first <- function(x) dnorm(x, 0, A0_temp)
+    f_second <- function(x) dnorm(x, 0, A1_temp)
+  }
+  m_class <- matrix(nrow = dim(x)[1], ncol = dim(x)[2])
+  for(i in 1:dim(x)[1]){
+    for(j in 1:dim(x)[2]){
+      prob0 <- f_first(x[i, j,])
+      prob1 <- f_second(x[i, j,])
+      if(prob0 > prob1)
+        m_class[i, j] <- 0
+      else
+        m_class[i, j] <- 1
+    }
+  }
+  a0 <- 1/3 * 1/3 * (mean(m_class[, 1]==0) + mean(m_class[, 2]==0) + mean(m_class[, 3]==0))
+  a1 <- mean(m_class[, 4]==1)
+  return(1/4 * (3 * a0 + a1) / 0.25)
+}
+
+r_star_opts <- map_dbl(ds[2:length(ds)], ~f_optimal_normals(10000, A0, A1, .))
+r_star_opts <- c(1, r_star_opts)
+r_star_opts <- tibble(d=ds, optimal=r_star_opts)
+temp_df <- 
   big_df %>%
+  left_join(r_star_opts) %>% 
   select(-contains("time")) %>% 
   melt(id.vars=c("iter", "d")) %>% 
-  ggplot(aes(x=as.factor(d), y=value, fill=as.factor(variable))) +
-  geom_boxplot(position = position_dodge2(width=0.3)) +
-  geom_hline(yintercept = 1, linetype=2)
+  group_by(variable, d) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  ungroup() %>% 
+  mutate(variable=as.character(variable)) %>% 
+  mutate(variable=if_else(variable=="rstar_rf", "rf", if_else(variable=="rstar_gbm", "gbm", "optimal")))
+saveRDS(temp_df, "../data/gbm_vs_rf_normals_results_optimals.rds")
+temp_df <- readRDS("../data/gbm_vs_rf_normals_results_optimals.rds")
+g <- ggplot(data=filter(temp_df, variable!="optimal"),
+       aes(x=as.factor(d), y=middle, colour=as.factor(variable))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper),
+                  position = position_dodge2(width=0.3)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  geom_point(data=filter(temp_df, variable=="optimal")) +
+  scale_color_brewer("Method", palette = "Dark2") +
+  xlab("# dimensions") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"))
+ggsave("../output/gbm_vs_rf_normal.pdf", g, width = 10, height = 6)
 ```
@@
 
-f_compare_rf_xgbtree_student_ts <- function(nu1, nu2, d, A0){
+f_compare_rf_gbm_student_ts <- function(nu1, nu2, d, A0){
   lrstar <- vector(length=2)
@@
   for(i in 1:nreplicate){
-    temp <- f_compare_rf_xgbtree_student_ts(nu1, nu2, d, A0)
-    temp <- f_compare_rf_xgbtree(d, A0, A1)
+    temp <- f_compare_rf_gbm_student_ts(nu1, nu2, d, A0)
     rstars <- temp$rstar
@@
   }
-  colnames(m_res) <- c("iter", "d", "nu1", "nu2", "rstar_rf", "rstar_xgbTree", "time_rf", "time_xgbTree")
+  colnames(m_res) <- c("iter", "d", "nu1", "nu2", "rstar_rf", "rstar_gbm", "time_rf", "time_gbm")
   m_res <- m_res %>% 
@@
 }
-saveRDS(big_df, "../data/xgbTree_vs_rf_studentts_results.rds")
+saveRDS(big_df, "../data/gbm_vs_rf_studentts_results.rds")
+
+f_optimal_studentts <- function(niter, nu1, nu2, d, A0){
+  x <- f_data_student_ts(nu1, nu2, d, A0, niter = niter)
+  sigma1 <- A0[1:d, 1:d] * (nu1 - 2) / nu1
+  sigma2 <- A0[1:d, 1:d] * (nu2 - 2) / nu2
+  if(d > 1){
+    f_first <- function(x) dmvt(x, rep(0, d), sigma1, nu1)
+    f_second <- function(x) dmvt(x, rep(0, d), sigma2, nu2)
+  }else{
+    f_first <- function(x) dst(x, 0, sigma1, nu1)
+    f_second <- function(x) dst(x, 0, sigma2, nu2)
+  }
+  m_class <- matrix(nrow = dim(x)[1], ncol = dim(x)[2])
+  for(i in 1:dim(x)[1]){
+    for(j in 1:dim(x)[2]){
+      prob0 <- f_first(x[i, j,])
+      prob1 <- f_second(x[i, j,])
+      if(prob0 > prob1)
+        m_class[i, j] <- 0
+      else
+        m_class[i, j] <- 1
+    }
+  }
+  a0 <- 1/3 * 1/3 * (mean(m_class[, 1]==0) + mean(m_class[, 2]==0) + mean(m_class[, 3]==0))
+  a1 <- mean(m_class[, 4]==1)
+  return(1/4 * (3 * a0 + a1) / 0.25)
+}
+
+r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(1000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))
+
+r_star_opts <- tibble(d=m_parameters$ds, nu2=m_parameters$nu2, optimal=r_star_opts1)
+temp_df <- 
+  big_df %>%
+  left_join(r_star_opts) %>% 
+  select(-contains("time")) %>% 
+  select(-one_of("nu1")) %>% 
+  melt(id.vars=c("iter", "d", "nu2")) %>% 
+  group_by(variable, d, nu2) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  ungroup() %>% 
+  mutate(variable=as.character(variable)) %>% 
+  mutate(variable=if_else(variable=="rstar_rf", "rf", if_else(variable=="rstar_gbm", "gbm", "optimal")))
+saveRDS(temp_df, "../data/gbm_vs_rf_studentts_results_optimal.rds")
+temp_df <- readRDS("../data/gbm_vs_rf_studentts_results_optimal.rds")
+g2 <- ggplot(data=filter(temp_df, variable!="optimal"),
+       aes(x=as.factor(nu2), y=middle, colour=as.factor(variable))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper),
+                  position = position_dodge2(width=0.3)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  geom_point(data=filter(temp_df, variable=="optimal")) +
+  scale_color_brewer("Method", palette = "Dark2") +
+  xlab("# degrees of freedom") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black")) +
+  facet_wrap(~d)
+ggsave("../output/gbm_vs_rf_studentt.pdf", g2, width = 12, height = 8)
 ```

### 2020-10-15 src/s_trends.Rmd
went back to gbm because of xgbtree poor results in trends

@@
 
-caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
-                   shrinkage=c(0.1),
-                   n.minobsinnode=10)
-
-f_fit_gbm_predict <- function(r){
+f_fit_rf_predict <- function(r){
   rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
+  nvars <- ncol(r) - 1
   training_data <- r[rand_samples, ]
@@
   gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "gbm",
+                 method = "rf",
                  trControl = trainControl(method = 'none'), 
-                    tuneGrid = caretGrid, verbose=FALSE)
+                    tuneGrid = tibble(mtry=nvars), verbose=FALSE)
   plda <- predict(object=gbmFit1, newdata=testing_data)
@@
 }
+
+f_fit_gbm_predict <- function(r){
+  rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
+  nvars <- ncol(r) - 1
+  training_data <- r[rand_samples, ]
+  testing_data <- r[-rand_samples, ]
+
+  grids <- tibble(interaction.depth=c(3), 
+                             n.trees = 50,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)
+
+  gbmFit1 <- train(chain ~ ., data = training_data, 
+                 method = "gbm",
+                 trControl = trainControl(method = 'none'), 
+                    tuneGrid = grids, verbose=FALSE)
+  plda <- predict(object=gbmFit1, newdata=testing_data)
+  a_accuracy <- 
+    tibble(predicted=plda, actual=testing_data$chain) %>%
+    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
+    summarise(mean(correct)) %>% 
+    pull()
+  return(a_accuracy)
+}
+
+f_fit_both <- function(r){
+  r1 <- f_fit_rf_predict(r)
+  r2 <- f_fit_gbm_predict(r)
+  return(list(rf=r1, gbm=r2))
+}
+models <- c("gbm", "rf")
 ```
@@
 )
-res <- vector("list", nrow(conds))
+res <- vector("list", 2 * nrow(conds))
 chains = 4
+kkk <- 1
+for(mm in 1:2){
   for (i in 1:nrow(conds)) {
+    print(i)
     iters <- conds[i, "iters"]
@@
     mutate(ones=rnorm(length(chain)))
+    if(mm==1)
       a_accuracy <- f_fit_gbm_predict(r1)
+    else
+      a_accuracy <- f_fit_rf_predict(r1)
 
@@
     mutate(ones=rnorm(length(chain)))
+    if(mm==1)
       b_accuracy <- f_fit_gbm_predict(r2)
+    else
+      b_accuracy <- f_fit_rf_predict(r2)
 
-  res[[i]] <- cbind(iters, trend, rep, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
-                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
+    res[[kkk]] <- cbind(iters, trend, rep, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
+                      "Rstar_split"=b_accuracy * n_distinct(r2$chain),
+                      method=models[mm])
+    kkk <- kkk + 1
+  }
 }
@@
 ```{r}
-res <- readRDS("../output/experiment_trends_all_dim.rds")
-g <- 
-  res %>%
-  select(iters, trend, Rstar, Rstar_split) %>% 
-  melt(id.vars=c("iters", "trend")) %>% 
-  mutate(variable=case_when(variable=="Rstar"~"A. R*",
-                            variable=="Rstar_split"~"B. Split-R*")) %>% 
-  ggplot(aes(y = value, x = trend)) + 
-  geom_point() + 
-  geom_jitter() + 
-  facet_grid(vars(iters), vars(variable)) + 
-  geom_hline(yintercept = 1) +
-  theme(strip.text.x = element_text(size=16, colour="black"),
-        strip.text.y = element_text(size=16, colour="black"),
-        axis.text.x = element_text(size=16, colour="black"),
-        axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black")) +
-  xlab("Trend") +
-  ylab("Value")
-ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)
+# res <- readRDS("../output/experiment_trends_all_dim.rds")
+# g <- 
+#   res %>%
+#   select(iters, trend, Rstar, Rstar_split) %>% 
+#   melt(id.vars=c("iters", "trend")) %>% 
+#   mutate(variable=case_when(variable=="Rstar"~"A. R*",
+#                             variable=="Rstar_split"~"B. Split-R*")) %>% 
+#   ggplot(aes(y = value, x = trend)) + 
+#   geom_point() + 
+#   geom_jitter() + 
+#   facet_grid(vars(iters), vars(variable)) + 
+#   geom_hline(yintercept = 1) +
+#   theme(strip.text.x = element_text(size=16, colour="black"),
+#         strip.text.y = element_text(size=16, colour="black"),
+#         axis.text.x = element_text(size=16, colour="black"),
+#         axis.text.y = element_text(size=16, colour="black"),
+#         axis.title = element_text(size=16, colour="black")) +
+#   xlab("Trend") +
+#   ylab("Value")
+# ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)
 ```
@@
 chains = 4
-
+kkk <- 1
+for(mm in 1:2){
   for (i in 1:nrow(conds)) {
+    print(i)
     iters <- conds[i, "iters"]
@@
       r1 <- r1 %>% mutate(ones=rnorm(length(chain)))
+    if(mm==1)
       a_accuracy <- f_fit_gbm_predict(r1)
+    else
+      a_accuracy <- f_fit_rf_predict(r1)
 
@@
       r2 <- r2 %>% mutate(ones=rnorm(length(chain)))
+    if(mm==1)
       b_accuracy <- f_fit_gbm_predict(r2)
+    else
+      b_accuracy <- f_fit_rf_predict(r2)
 
-  res[[i]] <- cbind(iters, trend, rep, dimension, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
-                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
+    res[[kkk]] <- cbind(iters, trend, rep, dimension, rs,
+                        "Rstar"=a_accuracy * n_distinct(r1$chain),
+                        "Rstar_split"=b_accuracy * n_distinct(r2$chain),
+                        method=models[mm])
+    kkk <- kkk + 1
+  }
 }
@@
 ```{r}
-g <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
-  select(trend, iters, rep, dimension, Rstar_split) %>% 
-  unique() %>%
-  select(-rep) %>% 
-  melt(c("trend", "iters", "dimension")) %>% 
-  filter(variable=="Rstar_split") %>% 
-  ggplot(aes(y = value, x = trend)) + 
-  geom_point() + 
-  geom_jitter() + 
-  facet_grid(vars(iters), vars(dimension)) + 
-  geom_hline(yintercept = 1) +
-  theme(strip.text.x = element_text(size=16, colour="black"),
-        strip.text.y = element_text(size=16, colour="black"),
-        axis.text.x = element_text(size=16, colour="black"),
-        axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black")) +
-  xlab("Trend") +
-  ylab("Value")
-g
-
-ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
+# g <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
+#   select(trend, iters, rep, dimension, Rstar_split, method) %>% 
+#   unique() %>%
+#   select(-rep) %>% 
+#   melt(c("trend", "iters", "dimension", "method")) %>% 
+#   filter(variable=="Rstar_split") %>% 
+#   ggplot(aes(y = value, x = trend)) + 
+#   geom_jitter(aes(colour=as.factor(method))) + 
+#   facet_grid(vars(iters), vars(dimension)) + 
+#   geom_hline(yintercept = 1) +
+#   theme(strip.text.x = element_text(size=16, colour="black"),
+#         strip.text.y = element_text(size=16, colour="black"),
+#         axis.text.x = element_text(size=16, colour="black"),
+#         axis.text.y = element_text(size=16, colour="black"),
+#         axis.title = element_text(size=16, colour="black")) +
+#   xlab("Trend") +
+#   ylab("Value")
+# g
+# 
+# ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
 ```
@@
       mutate(chain=as.factor(chain))
+    if(mm==1)
         a_accuracy <- f_fit_gbm_predict(r1)
+      else
+        a_accuracy <- f_fit_rf_predict(r1)
 
@@
       mutate(chain=as.factor(chain))
+    if(mm==1)
         b_accuracy <- f_fit_gbm_predict(r2)
+      else
+        b_accuracy <- f_fit_rf_predict(r2)
 
-  res[[i]] <- cbind(iters, rep, rho, t(rs), "Rstar"=a_accuracy * n_distinct(r1$chain),
-                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
+    res[[kkk]] <- cbind(iters, rep, rho, t(rs), "Rstar"=a_accuracy * n_distinct(r1$chain),
+                      "Rstar_split"=b_accuracy * n_distinct(r2$chain),
+                      method=models[mm])
+    kkk <- kkk + 1
+  }
 }
@@
   as.data.frame() %>%
-  select(iters, rho, rep, zsRhat, Rstar_split) %>% 
+  select(iters, rho, rep, zsRhat, Rstar_split, method) %>%
   unique() %>%
   select(-rep) %>%
-  melt(c("rho", "iters"))  %>% 
+  melt(c("rho", "iters", "method")) %>%
+  mutate(value=as.numeric(as.character(value))) %>% 
   mutate(variable=case_when(variable=="zsRhat"~"A. Split-Rhat",
@@
   ggplot(aes(y = value, x = rho)) +
-  geom_jitter() + 
+  geom_jitter(aes(colour=as.factor(method))) +
   facet_grid(vars(iters), vars(variable)) +
@@
 sigma <- 1
-
+kkk <- 1
+for(mm in 1:2){
   for (i in 1:nrow(conds)) {
+    print(i)
     iters <- conds[i, "iters"]
@@
       r1 <- r1 %>% mutate(ones=rnorm(length(chain)))
+    if(mm==1)
         a_accuracy <- f_fit_gbm_predict(r1)
+      else
+        a_accuracy <- f_fit_rf_predict(r1)
 
@@
       r2 <- r2 %>% mutate(ones=rnorm(length(chain)))
+    if(mm==1)
         b_accuracy <- f_fit_gbm_predict(r2)
+      else
+        b_accuracy <- f_fit_rf_predict(r2)
 
-  res[[i]] <- cbind(iters, trend, rep, rho, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
-                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
+    res[[kkk]] <- cbind(iters, trend, rep, rho, rs, "Rstar"=a_accuracy * n_distinct(r1$chain),
+                      "Rstar_split"=b_accuracy * n_distinct(r2$chain),
+                      method=models[mm])
+    kkk <- kkk + 1
+  }
 }
@@
 g <- ggplot(data = res, aes(x = rho, y = Rstar_split)) +
-  geom_jitter() + 
+  geom_jitter(aes(colour=as.factor(method))) +
   facet_grid(vars(iters)) +

### 2020-10-15 src/r_star_monitor.R

@@ -65,6 +65,11 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, 
                         colsample_bytree = .7,
                         min_child_weight = 1,
                         subsample = 0.8)
+  if(is.null(caret_default)&&method=="gbm")
+    caretGrid <- tibble(interaction.depth=c(3), 
+                             n.trees = 50,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)
   else if(is.null(caret_default))
     caretGrid <- tibble(mtry=floor(sqrt(nparams)))
   else

### 2020-10-17 src/s_trends.Rmd
started to rework manuscript

@@
 ```{r}
-# res <- readRDS("../output/experiment_trends_all_dim.rds")
-# g <- 
-#   res %>%
-#   select(iters, trend, Rstar, Rstar_split) %>% 
-#   melt(id.vars=c("iters", "trend")) %>% 
-#   mutate(variable=case_when(variable=="Rstar"~"A. R*",
-#                             variable=="Rstar_split"~"B. Split-R*")) %>% 
-#   ggplot(aes(y = value, x = trend)) + 
-#   geom_point() + 
-#   geom_jitter() + 
-#   facet_grid(vars(iters), vars(variable)) + 
-#   geom_hline(yintercept = 1) +
-#   theme(strip.text.x = element_text(size=16, colour="black"),
-#         strip.text.y = element_text(size=16, colour="black"),
-#         axis.text.x = element_text(size=16, colour="black"),
-#         axis.text.y = element_text(size=16, colour="black"),
-#         axis.title = element_text(size=16, colour="black")) +
-#   xlab("Trend") +
-#   ylab("Value")
+res <- readRDS("../output/experiment_trends_all_dim.rds")
+g <-
+  res %>%
+  select(iters, trend, Rstar, Rstar_split, method) %>%
+  melt(id.vars=c("iters", "trend", "method")) %>%
+  mutate(variable=case_when(variable=="Rstar"~"A. R*",
+                            variable=="Rstar_split"~"B. Split-R*")) %>%
+  ggplot(aes(y = value, x = trend)) +
+  geom_jitter(aes(colour=as.factor(method))) +
+  facet_grid(vars(iters), vars(variable)) +
+  geom_hline(yintercept = 1) +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black")) +
+  xlab("Trend") +
+  ylab("Value")
 # ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)

### 2020-10-17 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
started to rework manuscript

@@
 
-\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines}
+\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers}
 
@@
 Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; \textcolor{red}{instead it uses} all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
-Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
+\textcolor{red}{We recommend calculating $R^*$ using two different machine learning classifiers — gradient-boosted regression trees and random forests — which each work well in models of different dimensions.
+Because each of these methods outputs a classification probability}, as a byproduct, we obtain uncertainty in $R^*$.
 \textcolor{red}{The method is straightforward to implement and could be a complementary additional check on MCMC convergence for applied analyses.}
@@
 
-Here, we introduce $R^*$, a new convergence metric.
+\textcolor{red}{Here, we introduce $R^*$, a new convergence metric.
 This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
-\textcolor{red}{Rephrased, it should not be possible to predict which draws come from which chain}.
-In this vein, we use a machine learning (ML) classifier to measure convergence.
-Specifically, we train a ML classifier to predict the chain that generated each observation.
-By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
-Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
-\textcolor{red}{However, the ML classifier we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.} For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
+Rephrased, it should not be possible to predict which draws come from which chain.
+In this vein, we use machine learning (ML) classifiers to measure convergence.
+Specifically, we train classifiers to predict the chain that generated each observation.
+By evaluating the performance of classifiers on a held-out test set, this provides a new convergence metric.
+To maximise predictive accuracy, our chosen classifiers naturally exploit differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
+Our statistic, unlike its $\widehat{R}$ cousins, is scalar-valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
+However, the ML classifiers we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.}
+
+\textcolor{red}{There are, of course, a huge variety of possible ML classifiers.
+For a method to be useful and widely adopted, however, it needs to satisfy a number of criteria: first, across a range of examples, it should consistently be able to detect poor MCMC convergence; second, the ML classifier should be trainable in a reasonable amount of time; lastly, the methods should be tuned (via their hyperparameters) so as to be standardised, so that $R^*$ computed by one analyst is comparable to that computed by another.
+Here, we perform comparisons across some of the most popular methods in the ML literature and recommend two tree-based ML classifiers: gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM'') and random forest classifiers \citep{breiman2001random} (``RF'').
+Both of these methods performed consistently well across our range of examples, and the models were relatively efficient to train.
+We recommend calculating $R^*$ using both of these methods: GBMs tend to perform best for low dimensional cases; in moderate dimensional cases upwards, RF dominates.
+This difference in performance is partly due to hyperparameter tuning: echoing previous results in the literature \cite[Chapters~11\&12]{boehmke2019hands}, our results show that GBMs, in general, are more sensitive to hyperparameter choice than are RFs.
+For a GBM classifier to perform well in higher dimensions, more rounds of boosting are required, which substantially increases training time; instead, we fix the hyperparameters of GBMs to specific values to ensure practical usefulness.
+For RF classifiers, empirical studies have demonstrated reliable heuristics for selecting robust hyperparameters \citep{bernard2009influence}, and, in any case, across our set of test cases, these classifiers are relatively insensitive to hyperparameter choice.}
+
 For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
-\textcolor{red}{The statistic depends on the hyperparameters of GBM, although even for ``default'' settings, $R^*$ provides a stringent test of accuracy, giving a reliable measure of various Markov chain pathologies.} In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
+In addition, since ML classifiers can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
 $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
@@
 
-The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
-Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
+\textcolor{red}{To identify promising candidate ML methods, we run a series of experiments using a number of popular classifiers (see Section X).
+Two methods performed consistently well across our examples: gradient-boosted regression trees (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}) and random forest classifiers \citep{breiman2001random} (``RF'').
+Both of these approaches are based on decision trees: GBMs use an iterative approach known as ``boosting'', where each subsequent decision tree aims to predict the residuals from the previous one; RFs use an approach known as ``bagging'' which trains decision trees on many bootstrapped copies of the training data, yielding a collection of independently trained trees whose individual decisions, when aggregated, yield a class.
+For the implementations of these classifiers, we use those available in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' \citep{greenwell2019package} and ``randomForest'' \citep{liaw2002classification} packages.}
+
 The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters.
 We split each chain's draws into randomly divided training and testing sets: here, we use 70\% of draws for training and 30\% for testing.
-Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
-Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.
+Both approaches typically took $\mathcal{O}$(seconds) on a desktop computer to execute training then prediction on a testing set for most models we consider in \S\ref{sec:results}.
 
-From the GBM fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
+\textcolor{red}{Since different posteriors present different challenges to MCMC samplers, the nature of classification boundaries is problem-specific.
+Because of this, there is no unique optimal classifier across all problems, and the performance of the algorithms we investigate depends on their hyperparameters (see Section X).
+GBM models have a number of hyperparameters and have previously been demonstrated to be hard to tune \cite[Chapters~12]{boehmke2019hands}.
+To ensure practical run time for this method, we suggest running it using a default hyperparameter set: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees be grown.
+This choice of hyperparameters was chosen to present a balance between training cost and classification accuracy, producing an $R^*$  that was a stringent measure of convergence — in general, more stringent than $\widehat{R}$.
+RFs are generally less sensitive to variation in their single hyperparameter: $m_{\text{try}}$, the size of the subset of all features over which to search for an optimal split \citep{boehmke2019hands}.
+Our experiments replicate these results (see Section X), and we suggest running RFs using the heuristic $m_{\text{try}}=\sqrt{K}$, which has previously been shown to be a choice resulting in robust classification performance \citep{bernard2009influence,boehmke2019hands}.
+Unless otherwise stated, in the examples explored in \S\ref{sec:results}, the hyperparameters for these two classifiers were set at these defaults.}
+
+From a \textcolor{red}{classifier} fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
 Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
-For each draw, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
+For each draw, $s$, in our testing set, \textcolor{red}{classifiers} outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.

### 2020-10-17 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
started to update results

@@
 \subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
-In this section, we show how $R^*$ is able to detect heterogeneous variance across Markov chains.
-Apart from demonstrating how $R^*$ performs, this section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}); and, finally, to the ML model hyperparameters (\S\ref{sec:heterogeneity_robustness}).
+\textcolor{red}{In this section, we use a simple example to illustrate how $R^*$ works.
+Specifically, we show how $R^*$ can detect heterogeneous variance across Markov chains.
+This example aims to illustrate the basic mechanics behind how $R^*$ works, so we use only the GBM classifier here.} This section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model-predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); and to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}).
 The experiments we use to study these issues are all of similar form to the following data generating process: four Markov chains are generated, where each samples from an autoregressive order 1 (AR(1)) process of the form,

### 2020-10-17 src/s_ar1.Rmd
added older versions of the ar1 experiments

--- 2020-10-12 src/s_ar1.Rmd
@@
 
-nreplicates <- 100
+nreplicates <- 1000
 r_star_vals <- vector(length = nreplicates)
@@
 for(i in 1:nreplicates){
+  if(i%%100==0)
     print(i)
@@
 
-caretGrid <- tibble(mtry=2)
+caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
+                   shrinkage=c(0.1),
+                   n.minobsinnode=10)
 
@@
 gbmFit1 <- train(chain ~ ., data = training_data, 
-                 method = "rf",
+                 method = "gbm",
                  trControl = trainControl(method = 'none'), 

### 2020-10-17 src/s_ar1.Rmd
changed ar1 to use new functions

@@
   a_array[,,1] <- temp
-  a_accuracy <- r_star(a_array)
+  a_accuracy <- r_star(a_array, method = "gbm")
   mon <- monitor_extra(split_data(a_array))
@@
 a_array[,,2] <- temp1
-unmixed <- r_star_dist(a_array)
+unmixed <- r_star(a_array, method = "gbm", uncertainty = T)
 
@@
 a_array[,,2] <- temp1
-mixed <- r_star_dist(a_array)
+mixed <- r_star(a_array, method = "gbm", uncertainty = T)
 
@@
   a_array[,,1] <- temp
-  a_accuracy <- r_star(a_array)
+  a_accuracy <- r_star(a_array, method = "gbm")
   return(list(r_star=a_accuracy))
@@
 ```
+
+Examine probability calibration
+```{r}
+
+```

### 2020-10-17 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
changed ar1 to use new functions

@@
 	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
-Note that, in D, only a single series of each series type is used to generate distribution.}
+Note that, in D, only a single series of each series type is used to generate distribution.
+\textcolor{red}{All examples used a GBM for classification using the default hyperparameter values given in \S\ref{sec:method}.}}
 	\label{fig:ar1}

### 2020-10-18 src/s_ar1.Rmd
started to add gbm vs rf section

@@
   ggtitle("A.")
+```
+
+Examine probability calibration
+```{r}
+predict(object=gbmFit1, newdata=testing_data, type = "prob") %>% 
+  select(`4`) %>% 
+  rename(prob_4=`4`) %>% 
+  mutate(actual=testing_data$chain) %>% 
+  mutate(prob_4_cat=cut(prob_4, seq(0, 1, 0.05), ordered_result = T)) %>% 
+  group_by(prob_4_cat) %>% 
+  summarise(prob_4_actual=mean(actual==4)) %>% 
+  ggplot(aes(x=prob_4_cat, y=prob_4_actual)) +
+  geom_point()
 ```
@@
 ```
-
-Examine probability calibration
-```{r}
-
-```

### 2020-10-18 src/s_gbm_vs_rf.Rmd
started to add gbm vs rf section

@@
 saveRDS(temp_df, "../data/gbm_vs_rf_normals_results_optimals.rds")
-temp_df <- readRDS("../data/gbm_vs_rf_normals_results_optimals.rds")
-g <- ggplot(data=filter(temp_df, variable!="optimal"),
-       aes(x=as.factor(d), y=middle, colour=as.factor(variable))) +
+temp_df <- readRDS("../data/gbm_vs_rf_normals_results_optimals.rds") %>% 
+  mutate(variable=as.character(variable)) %>% 
+  mutate(variable=if_else(variable=="gbm", "GBM",
+                          if_else(variable=="rf", "RF", "Optimal"))) %>% 
+  mutate(variable=fct_relevel(as.factor(variable), "GBM", "RF", "Optimal"))
+g <- ggplot(temp_df,
+       aes(x=as.factor(d), y=middle, colour=as.factor(variable),
+           shape=as.factor(variable))) +
   geom_pointrange(aes(ymin=lower, ymax=upper),
@@
   geom_hline(yintercept = 1, linetype=2) +
-  geom_point(data=filter(temp_df, variable=="optimal")) +
-  scale_color_brewer("Method", palette = "Dark2") +
+  scale_color_grey("Method", start=0.8, end=0) +
+  scale_shape("Method") +
   xlab("# dimensions") +
@@
 saveRDS(temp_df, "../data/gbm_vs_rf_studentts_results_optimal.rds")
-temp_df <- readRDS("../data/gbm_vs_rf_studentts_results_optimal.rds")
-g2 <- ggplot(data=filter(temp_df, variable!="optimal"),
-       aes(x=as.factor(nu2), y=middle, colour=as.factor(variable))) +
+temp_df <- readRDS("../data/gbm_vs_rf_studentts_results_optimal.rds") %>% 
+  mutate(variable=as.character(variable)) %>% 
+  mutate(variable=if_else(variable=="gbm", "GBM",
+                          if_else(variable=="rf", "RF", "Optimal"))) %>% 
+  mutate(variable=fct_relevel(as.factor(variable), "GBM", "RF", "Optimal"))
+g2 <- ggplot(data=temp_df,
+       aes(x=as.factor(nu2), y=middle, colour=as.factor(variable),
+           shape=as.factor(variable))) +
   geom_pointrange(aes(ymin=lower, ymax=upper),
@@
   geom_hline(yintercept = 1, linetype=2) +
-  geom_point(data=filter(temp_df, variable=="optimal")) +
-  scale_color_brewer("Method", palette = "Dark2") +
+  scale_color_grey("Method", start=0.8, end=0) +
+  scale_shape("Method") +
   xlab("# degrees of freedom") +

### 2020-10-18 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
started to add gbm vs rf section

@@
 
+\subsection{Comparing GBMs and RFs}
+In section SX, we perform a series of experiments using popular ML methods and find that GBMs and RFs perform most consistently across them.
+In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
+In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}.
+In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution.
+In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains.
+Since we use independent sampling to generate draws, we are also able to calculate the Bayes optimal classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}.
+In this classifier, categories are assigned according to the one which has the maximum posterior probability of having generated the data.
+Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws according only to which chain maximises the likelihood of that observation.
+
+\subsubsection{Convergence in joint distribution}\label{sec:joint_distribution}
+Here, we consider a multivariate normal target distribution over a range of target dimensions ranging from 1-32.
+For three 'chains', we generate draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, draws are generated in the same way except the covariance matrix differs.
+To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
+To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^d$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^d=A^{32}_{1:d, 1:d}$.
+For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
+
+The results of these experiments are shown in Fig. X.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
+	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using GBM}
+	\label{fig:gbm_vs_rf_normal}
+\end{figure}
+
+\subsubsection{Fat tails}\label{sec:tail_fatness}
+
 \subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}

### 2020-10-18 src/s_gbm_vs_rf.Rmd
first draft of gbm vs rf sections

@@
 
-r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(1000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))
+r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(10000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))
 
 r_star_opts <- tibble(d=m_parameters$ds, nu2=m_parameters$nu2, optimal=r_star_opts1)
+big_df <- readRDS("../data/gbm_vs_rf_studentts_results.rds")
 temp_df <- 

### 2020-10-18 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
first draft of gbm vs rf sections

--- 2020-10-17 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
@@
 
+\color{red}
+\subsection{Comparing GBMs and RFs}
+In section SX, we perform a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
+In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
+In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}.
+In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution.
+In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains.
+Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}.
+This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw.
+Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.
+
+\subsubsection{Joint distribution}\label{sec:joint_distribution}
+Here, we consider a multivariate normal target distribution over a range of target dimensions from 1-32.
+For three `chains', we generate 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
+To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
+To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$.
+For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
+In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and determine the predictive accuracy using the Bayes optimal classifier.
+
+The results of these experiments are shown in Fig.
+\ref{fig:gbm_vs_rf_normal}.
+In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic.
+When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$.
+In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence.
+As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows.
+$R^*$ as calculated by GBMs and RFs also increases.
+For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification.
+For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
+	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
+For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
+The dashed line shows $R^*=1$.
+Here, $R^*$ was calculated using chains split into two halves.}
+	\label{fig:gbm_vs_rf_normal}
+\end{figure}
+
+\subsubsection{Fat tails}\label{sec:tail_fatness}
+We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution.
+To do so, we again consider four chains.
+In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom.
+In the remaining chain, we generate the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
+We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$.
+We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness.
+Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$.
+To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$.
+For each combination of dimensions and $\nu$, we perform 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
+Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.
+
+The results of these experiments are shown in Fig.
+\ref{fig:gbm_vs_rf_studdentt}.
+Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions.
+Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$.
+The point colour and shadings give the method used to calculate $R^*$.
+In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method.
+As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases.
+In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers.
+In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate.
+If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.
+
+Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
+The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
+In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
+These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
+This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and section X (hyperparameters)): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
+Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
+RFs are less sensitive to hyperparameter variation (section X), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
+Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
+	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution.
+The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
+For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
+The dashed line shows $R^*=1$.
+Here, $R^*$ was calculated using chains split into two halves.}
+	\label{fig:gbm_vs_rf_studdentt}
+\end{figure}
+
+\color{black}
 \subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
@@
 \section{Acknowledgements}
-The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.
+\textcolor{red}{The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.}
 

### 2020-10-19 src/s_hyperparameters.Rmd
started to add hyperparameters section

--- 2020-10-09 src/s_hyperparameters.Rmd
@@
 
-saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
-r_star_vals <- readRDS("../data/ar1_sensitivity.rds")
-
-b_sum <-
-  r_star_vals %>% 
-  as.data.frame() %>% 
-  mutate(iterate=seq(1, nreplicates, 1)) %>%
-  melt(id.vars="iterate") %>% 
-  mutate(below=if_else(value < 1, 1, 0)) %>% 
-  group_by(variable) %>% 
-  summarise(below_per=round(mean(below)*100, 1),
-            lower=quantile(value, 0.25),
-            upper=quantile(value, 0.75),
-            middle=median(value)) %>%
-  mutate(below_per=paste0(below_per, "%")) %>% 
-  mutate(x_pos=seq(1, nrow(m_parameters), 1)) %>% 
-  mutate(y_pos=0.95)
-
-g <- b_sum %>% 
-  cbind(m_parameters) %>% 
+# saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
+g1 <- readRDS("../data/hypers_ar1.rds")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
   ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
@@
         legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.98, NA)) +
+  scale_y_continuous(limits=c(0.95, 1.3)) +
   geom_hline(yintercept = 1, linetype=2) +
   scale_color_brewer("Int. depth", palette = "Dark2") +
-  guides(colour = guide_legend(nrow = 2, byrow = T))
-ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  ggtitle("A. GBM")
 ```
@@
 
-r_star_vals %>% 
-  melt(id.vars=c("n.trees", "int.depth")) %>% 
-  group_by(n.trees, int.depth) %>% 
+g1a <- readRDS("../data/hypers_8_schools.rds")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
   summarise(lower=quantile(value, 0.25),
-            upper=quantile(value, 0.75),
-            middle=median(value)) %>% 
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
   ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
@@
         axis.text = element_text(colour="black"),
-        legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.98, NA)) +
+        legend.position = c(0.7, 0.6)) +
+  scale_y_continuous(limits=c(0.95, 1.6)) +
   geom_hline(yintercept = 1, linetype=2) +
   scale_color_brewer("Int. depth", palette = "Dark2") +
-  guides(colour = guide_legend(nrow = 2, byrow = T))
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  ggtitle("A. GBM")
 ```
@@
 r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_cauchy)
+
+readRDS("../data/hypers_cauchy.rds")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# trees") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
 ```
+
+# RF hyperparameters
+AR1
+```{r}
+temp <- readRDS("../data/hypers_rf_ar1.rds")$r_star
+g2 <- temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=mtry, y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab(TeX("$m_{try}$")) +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, 1.3)) +
+  scale_x_continuous(breaks=c(1, 2), limits = c(0.5, 2.5)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  ggtitle("B. RF")
+pdf("../output/hypers_ar1.pdf", width = 12, height = 6)
+multiplot(g1, g2, cols=2)
+dev.off()
+```
+
+8 schools
+```{r}
+g2a <- readRDS("../data/hypers_rf_8_schools.rds")$r_star %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line(aes(group=1)) +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, 1.6)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  ggtitle("B. RF")
+
+pdf("../output/hypers_8_schools.pdf", width = 12, height = 6)
+multiplot(g1a, g2a, cols=2)
+dev.off()
+```
+Cauchy
+```{r}
+temp <- readRDS("../data/hypers_rf_cauchy.rds")$r_star
+temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
+```
+
+Normal
+```{r}
+temp <- readRDS("../data/hypers_rf_")$r_star
+temp %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, NA)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T))
+```

### 2020-10-19 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
made changes in title red
started to add hyperparameters section

@@
 
-\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers}
+\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using \textcolor{red}{decision tree classifiers}}
 
@@
 \color{red}
-\subsection{Comparing GBMs and RFs}
+\subsection{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
 In section SX, we perform a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
@@
 \subsubsection{Joint distribution}\label{sec:joint_distribution}
-Here, we consider a multivariate normal target distribution over a range of target dimensions from 1-32.
+Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32.
 For three `chains', we generate 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
@@
 For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
-In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and determine the predictive accuracy using the Bayes optimal classifier.
+In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.
 
@@
 
-\color{red}
-\subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
-The performance of GBM, like all ML methods, depends on its hyperparameters.
-In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree \cite{greenwell2019package}.
-The number of such regression trees is thus a hyperparameter of the model.
-Each regression tree can split amongst any integer number of variables (the \textit{interaction depth}).
-In this section, we investigate the sensitivity of $R^*$ to these hyperparameters across a range of examples.
-
-\subsubsection{Autoregressive example}\label{sec:heterogeneity_robustness}
-In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
-\ref{fig:ar1}.
-The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
-The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
-\ref{fig:ar1_sensitivity}.
-These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper).
-This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ar1_sensitivity.pdf}}
-	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
-The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.}
-	\label{fig:ar1_sensitivity}
-\end{figure}
-
-\subsubsection{Eight schools: non-centered parameterisation}
-
-\subsubsection{Cauchy model: alternative paramerisation}
-
-\subsubsection{Multivariate normal: 250-dimensional model}
-
-\subsection{ML model sensitivity}\label{sec:ml_model}
-
-\color{black}
-
 \subsection{Further experiments}\label{sec:further_experiments}

### 2020-10-19 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
started to add hyperparameters section

@@
 
+\color{red}
+\subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
+The performance of GBMs and RFs, like all ML methods, depends on their hyperparameters.
+In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree \cite{greenwell2019package}.
+The number of such regression trees is thus a hyperparameter of the model.
+Each regression tree can split amongst any integer number of variables (the \textit{interaction depth}).
+For GBMs, we analyse how their performance depends on these two hyperparameters.
+For RFs, there is a single hyperpameter: $m_{\text{try}}$, the number of members of the subset of all features over which to search for an optimal split \citep{boehmke2019hands} when forming a decision tree.
+
+In this section, we investigate the sensitivity of $R^*$ calculated by GBMs and RFs to each of these sets of hyperparameters across a range of examples.
+The test cases picked correspond to examples already introduced throughout the manuscript: in general, we chose to investigate cases where $R^*$ was nearer 1, since, for these, small changes in calculated $R^*$ could lead to different decisions about convergence being reached.
+
+\subsubsection{Autoregressive example}\label{sec:hyperparameters_ar1}
+In this section, we assume the same data generating process as described in \S\ref{sec:heterogeneity}.
+For each generated dataset, we fit GBMs with all 16 possible combinations of interaction depth: $(1,3,5,7)$; and number of regression trees: $(1,10,50,100)$.
+Using the same data generating process, we also fit RFs with the two allowable $m_{\text{try}}$ values: $(1,2)$.
+At each combination of classifier and hyperparameters, with performed 200 replicates: in each case, calculating the $R^*$ value.
+
+In Fig.
+\ref{fig:hypers_ar1}, we show the results of this analysis.
+In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
+In panel B, we show the same but for RFs.
+In panel A, we see that $R^*$ does vary according to hyperparameters: of the set investigated, an interaction depth of 1 appears optimal with 10+ trees.
+In panel B, the values of $R^*$ are lower generally than those achieved by GBMs: again, RFs appear less well-suited to lower dimensional problems as shown in \S\ref{sec:comparison_gbm_rf}.
+Across the two hyperparameter values investigated, there was not a substantial difference between the values of $R^*$ calculated.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{hypers_ar1.pdf}}
+	\caption{\textbf{Hyperparameter sensitivity: AR(1) example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the AR(1) example described in \S\ref{sec:hyperparameters_ar1}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.}
+	\label{fig:hypers_ar1}
+\end{figure}
+
+\subsubsection{Eight schools: non-centered parameterisation}\label{sec:hyperparameters_8_schools}
+In this section, we revisit the non-centered parameterisation of the eight schools model described in \S\ref{non-centered parameterisation}.
+We investigated a range of hyperparameters for GBM and RF classifiers.
+At each combination of classifier and hyperparameter setting, we performed 50 replicates: in each replicate, we used Stan's NUTS algorithm with adapt delta set to 0.95 to generate posterior samples; these were then used to calculate $R^*$.
+
+In Fig.
+\ref{fig:hypers_8_schools}, we show the results of these experiments.
+Note that, due to length of computational runtime, we chose not to explore the full range of hyperparameters needed to optimise the performance of each classifier.
+Panel A shows how $R^*$ calculated using a GBM varies with its two hyperparameters; panel B shows the same but for RFs.
+This shows that, in this case, the value of $R^*$ calculated by RFs was higher than that for GBMs.
+Further, whereas GBMs were relatively sensitive to their hyperparameter values, RFs were less so.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{hypers_8_schools.pdf}}
+	\caption{\textbf{Hyperparameter sensitivity: eight schools example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the eight schools example described in \S\ref{sec:hyperparameters_8_schools}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.}
+	\label{fig:hypers_8_schools}
+\end{figure}
+
+\subsubsection{Cauchy model: alternative paramerisation}
+
+\subsubsection{Multivariate normal: 250-dimensional model}
+
+\subsection{ML model sensitivity}\label{sec:ml_model}
+
+\color{black}
+
+\color{black}
+
 \end{document}

### 2020-10-19 src/s_jobScheduler_hypers_rf.sh

@@ -5,7 +5,7 @@ chmod +x *.sh
 
 
 # Rscript s_run_hyperparameters_rf.R 1 200
-Rscript s_run_hyperparameters_rf.R 2 50
+Rscript s_run_hyperparameters_rf.R 2 5
 # Rscript s_run_hyperparameters_rf.R 3 20
-Rscript s_run_hyperparameters_rf.R 4 20
+# Rscript s_run_hyperparameters_rf.R 4 20
 

### 2020-10-19 src/s_run_hyperparameters_rf.R

@@ -103,7 +103,7 @@ if(model_num==1){
   f_data_gen <- f_data_ar1
   name <- "ar1"
 }else if(model_num==2){
-  m_parameters <- expand_grid(mtry=seq(2, 16, 2))
+  m_parameters <- expand_grid(mtry=c(2, 4, 8, 16, 32, 64))
   f_data_gen <- f_data_8_schools
   name <- "8_schools"
 }else if(model_num==3){

### 2020-10-19 src/s_hyperparameters.Rmd
hyperparameters section nearly complete

@@
             upper=quantile(value, 0.75)) %>% 
+  mutate(default=if_else(int.depth==3&n.trees==50, 1, 0)) %>% 
   ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
-  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper, shape=as.factor(default)),
+                  position=position_dodge2(width=0.1)) +
   geom_line() +
@@
   guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  scale_shape(guide="none") +
   ggtitle("A. GBM")
@@
         axis.text = element_text(colour="black"),
-        legend.position = c(0.7, 0.6)) +
+        legend.position = c(0.7, 0.7)) +
   scale_y_continuous(limits=c(0.95, 1.6)) +
@@
 
-m_parameters <- expand_grid(int.depth=c(3, 7, 11),
-                            n.trees=c(50, 100, 400))
-r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_cauchy)
+# m_parameters <- expand_grid(int.depth=c(3, 7, 11),
+#                             n.trees=c(50, 100, 400))
+# r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_cauchy)
 
-readRDS("../data/hypers_cauchy.rds")$r_star %>% 
+g1b <- readRDS("../data/hypers_cauchy.rds")$r_star %>% 
   melt(id.vars=c("int.depth", "n.trees")) %>% 
@@
         legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.95, NA)) +
+  scale_y_continuous(limits=c(0.95, 2.8)) +
   geom_hline(yintercept = 1, linetype=2) +
   scale_color_brewer("Int. depth", palette = "Dark2") +
-  guides(colour = guide_legend(nrow = 2, byrow = T))
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  scale_shape(guide="none") +
+  ggtitle("A. GBM")
+```
+
+Normal
+```{r}
+readRDS("../data/hypers_")$r_star %>% 
+  melt(id.vars=c("int.depth", "n.trees")) %>% 
+  group_by(int.depth, n.trees) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  mutate(default=if_else(int.depth==3&n.trees==50, 1, 0)) %>% 
+  ggplot(aes(x=as.factor(n.trees), y=middle, colour=as.factor(int.depth), group=as.factor(int.depth))) +
+  geom_pointrange(aes(ymin=lower, ymax=upper, shape=as.factor(default)),
+                  position=position_dodge2(width=0.1)) +
+  geom_line() +
+  xlab("# trees") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, 2.8)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  scale_shape(guide="none") +
+  ggtitle("A. GBM")
 ```
@@
 ```{r}
-temp <- readRDS("../data/hypers_rf_ar1.rds")$r_star
-g2 <- temp %>%
+g2 <- readRDS("../data/hypers_rf_ar1.rds")$r_star %>%
   melt(id.vars="mtry") %>%
@@
             upper=quantile(value, 0.75)) %>%
+  mutate(default=if_else(mtry==1, 1, 0)) %>% 
   ggplot(aes(x=mtry, y=middle)) +
-  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper, shape=as.factor(default)),
+                  position=position_dodge2(width=0.1)) +
   geom_line() +
@@
   guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  scale_shape(guide="none") +
   ggtitle("B. RF")
@@
             upper=quantile(value, 0.75)) %>%
+  mutate(default=if_else(mtry==4, 1, 0)) %>% 
   ggplot(aes(x=as.factor(mtry), y=middle)) +
-  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper, shape=as.factor(default)),
+                  position=position_dodge2(width=0.1)) +
   geom_line(aes(group=1)) +
@@
 ```{r}
-temp <- readRDS("../data/hypers_rf_cauchy.rds")$r_star
-temp %>%
+g2b <- readRDS("../data/hypers_rf_cauchy.rds")$r_star %>%
+  melt(id.vars="mtry") %>%
+  group_by(mtry) %>%
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>%
+  mutate(default=if_else(mtry==12, 1, 0)) %>% 
+  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper, shape=as.factor(default)),
+                  position=position_dodge2(width=0.1)) +
+  geom_line(aes(group=1)) +
+  xlab("# features") +
+  ylab("R*") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.7, 0.25)) +
+  scale_y_continuous(limits=c(0.95, 2.8)) +
+  geom_hline(yintercept = 1, linetype=2) +
+  scale_color_brewer("Int. depth", palette = "Dark2") +
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  scale_shape(guide="none") +
+  ggtitle("B. RF")
+pdf("../output/hypers_cauchy.pdf", width = 12, height = 6)
+multiplot(g1b, g2b, cols=2)
+dev.off()
+```
+
+Normal
+```{r}
+readRDS("../data/hypers_rf_normal_250.rds")$r_star %>%
   melt(id.vars="mtry") %>%
@@
 ```
-
-Normal
-```{r}
-temp <- readRDS("../data/hypers_rf_")$r_star
-temp %>%
-  melt(id.vars="mtry") %>%
-  group_by(mtry) %>%
-  summarise(lower=quantile(value, 0.25),
-            middle=quantile(value, 0.5),
-            upper=quantile(value, 0.75)) %>%
-  ggplot(aes(x=as.factor(mtry), y=middle)) +
-  geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
-  geom_line() +
-  xlab("# features") +
-  ylab("R*") +
-  theme(text = element_text(size=16, colour="black"),
-        axis.text = element_text(colour="black"),
-        legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.95, NA)) +
-  geom_hline(yintercept = 1, linetype=2) +
-  scale_color_brewer("Int. depth", palette = "Dark2") +
-  guides(colour = guide_legend(nrow = 2, byrow = T))
-```

### 2020-10-19 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
hyperparameters section nearly complete

@@
 \color{red}
+\subsection{ML model sensitivity}\label{sec:ml_model}
+In this section, we investigate how classification accuracy depends on choice of ML classifier.
+
+\color{black}
+
+\color{red}
 \subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
-The performance of GBMs and RFs, like all ML methods, depends on their hyperparameters.
-In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree \cite{greenwell2019package}.
+The performance of GBMs and RFs, like most ML methods, depends on their hyperparameters.
+In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree.
 The number of such regression trees is thus a hyperparameter of the model.
@@
 For GBMs, we analyse how their performance depends on these two hyperparameters.
-For RFs, there is a single hyperpameter: $m_{\text{try}}$, the number of members of the subset of all features over which to search for an optimal split \citep{boehmke2019hands} when forming a decision tree.
+For RFs, there is a single hyperpameter: $m_{\text{try}}$, the number of members of the subset of all features over which to search for an optimal split when forming a decision tree.
+Note, that the maximum allowed value of $m_{\text{try}}=K$ — the dimensionality of the target distribution.
 
@@
 In panel B, we show the same but for RFs.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
 In panel A, we see that $R^*$ does vary according to hyperparameters: of the set investigated, an interaction depth of 1 appears optimal with 10+ trees.
-In panel B, the values of $R^*$ are lower generally than those achieved by GBMs: again, RFs appear less well-suited to lower dimensional problems as shown in \S\ref{sec:comparison_gbm_rf}.
-Across the two hyperparameter values investigated, there was not a substantial difference between the values of $R^*$ calculated.
+In panel B, the values of $R^*$ are generally lower than those achieved by GBMs: again, RFs appear less well-suited to lower dimensional problems as shown in \S\ref{sec:comparison_gbm_rf}.
+Across the two possible hyperparameter values for RFs, there was not a substantial difference in the values of $R^*$ calculated.
 
@@
 
+\subsubsection{Multivariate normal: 250-dimensional model}
+
+\subsubsection{Cauchy model: alternative parameterisation}\label{sec:hyperparameters_cauchy}
+In this section, we revisit the alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy} and investigate how $R^*$ as calculated by GBM and RF models depends on hyperparameters.
+Note, that for the GBM model, we did not investigate a full range of hyperparameters needed to optimise $R^*$ due to the extensive training time needed to do so.
+For each combination of model and hyperparameters, we generated samples from the model as described in \S\ref{sec:cauchy} and performed 20 replicates at each of these combinations.
+
+In Fig.
+\ref{fig:hypers_ar1}, we show the results of this analysis.
+In panel A, we show $R^*$ as calculated by a GBM classifier; in panel B, the same but using a RF classifier.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
+Across the range of hyperparameters investigated, the RF classifier outperformed the GBM and was less sensitive to variation in its hyperparameters.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{hypers_cauchy.pdf}}
+	\caption{\textbf{Hyperparameter sensitivity: Cauchy example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the Cauchy example described in \S\ref{sec:hyperparameters_cauchy}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.}
+	\label{fig:hypers_cauchy}
+\end{figure}
+
 \subsubsection{Eight schools: non-centered parameterisation}\label{sec:hyperparameters_8_schools}
-In this section, we revisit the non-centered parameterisation of the eight schools model described in \S\ref{non-centered parameterisation}.
+In this section, we revisit the non-centered parameterisation of the eight schools model described in \S\ref{sec:eight_shools}.
 We investigated a range of hyperparameters for GBM and RF classifiers.
@@
 \ref{fig:hypers_8_schools}, we show the results of these experiments.
-Note that, due to length of computational runtime, we chose not to explore the full range of hyperparameters needed to optimise the performance of each classifier.
+Note that, due to length of computational runtime, we chose not to explore the full range of hyperparameters needed to optimise the performance for GBMs; for RFs, the maximum allowed value of $m_{\text{try}}=19$.
 Panel A shows how $R^*$ calculated using a GBM varies with its two hyperparameters; panel B shows the same but for RFs.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
 This shows that, in this case, the value of $R^*$ calculated by RFs was higher than that for GBMs.
@@
 
-\subsubsection{Cauchy model: alternative paramerisation}
-
-\subsubsection{Multivariate normal: 250-dimensional model}
-
-\subsection{ML model sensitivity}\label{sec:ml_model}
-
 \color{black}
 
-\color{black}
+\bibliographystyle{chicago}
+\bibliography{bibliography}
 

### 2020-10-19 src/s_jobScheduler_hypers.sh

@@ -6,5 +6,6 @@ chmod +x *.sh
 
 # Rscript s_run_hyperparameters.R 1 200
 # Rscript s_run_hyperparameters.R 2 50
-Rscript s_run_hyperparameters.R 3 20
+# Rscript s_run_hyperparameters.R 3 20
+Rscript s_run_hyperparameters.R 4 20
 

### 2020-10-19 src/s_jobScheduler_hypers_rf.sh

@@ -5,7 +5,7 @@ chmod +x *.sh
 
 
 # Rscript s_run_hyperparameters_rf.R 1 200
-Rscript s_run_hyperparameters_rf.R 2 5
+Rscript s_run_hyperparameters_rf.R 2 50
 # Rscript s_run_hyperparameters_rf.R 3 20
 # Rscript s_run_hyperparameters_rf.R 4 20
 

### 2020-10-19 src/s_run_hyperparameters.R

@@ -47,7 +47,7 @@ f_replicate_hyper <- function(m_parameters, f_data_generator){
   
   laccuracy <- vector(length = nrow(m_parameters))
   for(i in 1:nrow(m_parameters))
-    laccuracy[i] <- r_star(a_array,
+    laccuracy[i] <- r_star(a_array, method="gbm",
                            caret_default=expand.grid(interaction.depth=m_parameters$int.depth[i],
                                                      n.trees = m_parameters$n.trees[i],
                                                      shrinkage=0.1,
@@ -87,6 +87,16 @@ f_data_cauchy <- function(){
   return(x)
 }
 
+f_data_normal <- function(){
+  N <- 250
+  A <- rWishart(1, 250, diag(N))[,,1]
+  model <- stan_model("mvt_250_ncp.stan")
+  fit <- sampling(model, data=list(N=N, A=A), iter=10, chains=4)
+  x <- rstan::extract(fit, permuted=F)
+  return(x)
+}
+
+
 # command line arguments ------
 args <- commandArgs(trailingOnly=TRUE)
 model_num <- as.numeric(args[1])
@@ -107,6 +117,11 @@ if(model_num==1){
                               n.trees=c(50, 100, 400, 1000))
   f_data_gen <- f_data_cauchy
   name <- "cauchy"
+}else if(model_num==4){
+  m_parameters <- expand_grid(int.depth=c(3, 7, 11),
+                              n.trees=c(50, 100, 400))
+  f_data_gen <- f_data_cauchy
+  name <- "normal_250"
 }
 
 r_star_vals <- f_replicate_r_star(num_iter, m_parameters, f_data_gen)

### 2020-10-19 src/s_run_hyperparameters_rf.R

@@ -88,7 +88,7 @@ f_data_normal <- function(){
   N <- 250
   A <- rWishart(1, 250, diag(N))[,,1]
   model <- stan_model("mvt_250_ncp.stan")
-  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
+  fit <- sampling(model, data=list(N=N, A=A), iter=10, chains=4)
   x <- rstan::extract(fit, permuted=F)
   return(x)
 }
@@ -103,7 +103,7 @@ if(model_num==1){
   f_data_gen <- f_data_ar1
   name <- "ar1"
 }else if(model_num==2){
-  m_parameters <- expand_grid(mtry=c(2, 4, 8, 16, 32, 64))
+  m_parameters <- expand_grid(mtry=c(2, 4, 8, 16, 19))
   f_data_gen <- f_data_8_schools
   name <- "8_schools"
 }else if(model_num==3){

### 2020-10-19 src/s_hyperparameters.Rmd
gbm normal hypers run

@@
 
-# AR1 model
+# GBM
+AR1 model
 ```{r}
@@
 
-# 8 schools
+8 schools
 ```{r}
@@
 
-# Cauchy
+Cauchy
 ```{r}
@@
 ```{r}
-readRDS("../data/hypers_")$r_star %>% 
+readRDS("../data/hypers_normal_250.rds")$r_star %>% 
   melt(id.vars=c("int.depth", "n.trees")) %>% 
@@
             upper=quantile(value, 0.75)) %>%
-  ggplot(aes(x=as.factor(mtry), y=middle)) +
+  ggplot(aes(x=as.factor(mtry), y=middle, group=1)) +
   geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +

### 2020-10-20 src/s_hyperparameters.Rmd
added new hyperparameter plots

@@
 
-m_parameters <- expand_grid(int.depth=seq(1, 7, 2),
-                            n.trees=c(1, 10, 50, 100))
-r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_ar1)
+# m_parameters <- expand_grid(int.depth=seq(1, 7, 2),
+#                             n.trees=c(1, 10, 50, 100))
+# r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_ar1)
 
@@
                   position=position_dodge2(width=0.1)) +
-  geom_line() +
+  geom_line(position=position_dodge2(width=0.1)) +
   xlab("# trees") +
@@
 ```{r}
-source("eight_schools.data.R")
-eight_schools <- list(J=J, y=y, sigma=sigma)
-model_cp <- stan_model("eight_schools_cp.stan")
-model_ncp <- stan_model("eight_schools_ncp.stan")
-
-f_data_8_schools <- function(){
-  fit_ncp <- sampling(
-    model_ncp, data = eight_schools,
-    iter = 2000, chains = 4, refresh = 0,
-    control = list(adapt_delta = 0.95)
-    )
-  x_ncp <- rstan::extract(fit_ncp, permuted=F)
-}
-
-m_parameters <- expand_grid(int.depth=c(1, 3, 5),
-                            n.trees=c(1, 50, 200))
-r_star_vals <- f_replicate_r_star(20, m_parameters, f_data_ar1)
+# source("eight_schools.data.R")
+# eight_schools <- list(J=J, y=y, sigma=sigma)
+# model_cp <- stan_model("eight_schools_cp.stan")
+# model_ncp <- stan_model("eight_schools_ncp.stan")
+# 
+# f_data_8_schools <- function(){
+#   fit_ncp <- sampling(
+#     model_ncp, data = eight_schools,
+#     iter = 2000, chains = 4, refresh = 0,
+#     control = list(adapt_delta = 0.95)
+#     )
+#   x_ncp <- rstan::extract(fit_ncp, permuted=F)
+# }
+# 
+# m_parameters <- expand_grid(int.depth=c(1, 3, 5),
+#                             n.trees=c(1, 50, 200))
+# r_star_vals <- f_replicate_r_star(20, m_parameters, f_data_ar1)
 
@@
 ```{r}
-readRDS("../data/hypers_normal_250.rds")$r_star %>% 
+g1c <- readRDS("../data/hypers_normal_250.rds")$r_star %>% 
   melt(id.vars=c("int.depth", "n.trees")) %>% 
@@
                   position=position_dodge2(width=0.1)) +
-  geom_line() +
+  geom_line(position=position_dodge2(width=0.1)) +
   xlab(TeX("$m_{try}$")) +
@@
                   position=position_dodge2(width=0.1)) +
-  geom_line(aes(group=1)) +
-  xlab("# features") +
+  geom_line(aes(group=1), position=position_dodge2(width=0.1)) +
+  xlab(TeX("$m_{try}$")) +
   ylab("R*") +
@@
 ```{r}
-readRDS("../data/hypers_rf_normal_250.rds")$r_star %>%
+g2c <- readRDS("../data/hypers_rf_normal_250.rds")$r_star %>%
   melt(id.vars="mtry") %>%
@@
   geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
-  geom_line() +
-  xlab("# features") +
+  geom_line(group=1, position=position_dodge2(width=0.1)) +
+  xlab(TeX("$m_{try}$")) +
   ylab("R*") +
@@
   scale_color_brewer("Int. depth", palette = "Dark2") +
-  guides(colour = guide_legend(nrow = 2, byrow = T))
+  guides(colour = guide_legend(nrow = 2, byrow = T)) +
+  ggtitle("B. RF")
+pdf("../output/hypers_normal_250.pdf", width = 12, height = 6)
+multiplot(g1c, g2c, cols=2)
+dev.off()
 ```

### 2020-10-20 src/s_ml.Rmd
started to add ml comparison section

--- 2020-10-11 src/s_ml.Rmd
@@
 
-temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
-          list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
-          f_data_ar1)
-temp_df <- readRDS("../data/ml_comp_hypers_set_ar1.rds")
-temp_df %>%
-  ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
+# temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
+#           list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
+#           f_data_ar1)
+
+orders <- c("rf", "xgbTree", "gbm", "knn", "multinom", "svmLinear")
+lookup <- tibble(method=orders,
+                 new=c("RF", "xgbTree", "GBM", "KNN", "GLM", "SVM"))
+
+temp_df <- readRDS("../data/ml_comp_hypers_set_ar1.rds")  %>% 
+  filter(method!="xgbTree") %>%
+  select(method, iter, r_star, time) %>% 
+  left_join(lookup, by="method") %>% 
+  select(-method) %>% 
+  rename(method=new) %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, lookup$new)) %>% 
+  rename(`R*`=r_star, `Time`=time) %>% 
+  melt(id.vars=c("method", "iter")) %>% 
+  group_by(method, variable) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  mutate(model="A. AR(1)")
 ```
@@
 
-temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
-          list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
-          f_data_8_schools)
+# temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
+#           list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
+#           f_data_8_schools)
 
-temp_df <- readRDS("../data/ml_comp_hypers_set_8_schools.rds")
-temp_df %>%
-  ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-temp_df %>%
-  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
-  geom_jitter()
+temp_df1 <- readRDS("../data/ml_comp_hypers_set_8_schools.rds")  %>% 
+  filter(method!="xgbTree") %>%
+  select(method, iter, r_star, time) %>% 
+  left_join(lookup, by="method") %>% 
+  select(-method) %>% 
+  rename(method=new) %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, lookup$new)) %>% 
+  rename(`R*`=r_star, `Time`=time) %>% 
+  melt(id.vars=c("method", "iter")) %>% 
+  group_by(method, variable) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  mutate(model="D. 8 schools")
 ```
@@
 
-temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
-          list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
-          f_data_cauchy)
-temp_df <- readRDS("../data/ml_comp_hypers_set_cauchy.rds")
-temp_df %>%
-  ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-temp_df %>%
-  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
-  geom_jitter()
+# temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
+#           list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
+#           f_data_cauchy)
+temp_df2 <- readRDS("../data/ml_comp_hypers_set_cauchy.rds") %>% 
+  filter(method!="xgbTree") %>%
+  select(method, iter, r_star, time) %>% 
+  left_join(lookup, by="method") %>% 
+  select(-method) %>% 
+  rename(method=new) %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, lookup$new)) %>% 
+  rename(`R*`=r_star, `Time`=time) %>% 
+  melt(id.vars=c("method", "iter")) %>% 
+  group_by(method, variable) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  mutate(model="C. Cauchy")
 ```
@@
 
-temp_df <- f_replicate_gridded(x, "rf", tunegrid_rf)
+# temp_df <- f_replicate_gridded(x, "rf", tunegrid_rf)
 
-temp_df <- readRDS("../data/ml_comp_hypers_set_normal_250.rds")
-temp_df %>%
-  ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-temp_df %>%
-  ggplot(aes(x=as.factor(method), y=time)) +
-  geom_boxplot()
+temp_df3 <- readRDS("../data/ml_comp_hypers_set_normal_250.rds") %>% 
+  filter(method!="xgbTree") %>%
+  select(method, iter, r_star, time) %>% 
+  left_join(lookup, by="method") %>% 
+  select(-method) %>% 
+  rename(method=new) %>% 
+  mutate(method=as.factor(method)) %>% 
+  mutate(method=fct_relevel(method, lookup$new)) %>% 
+  rename(`R*`=r_star, `Time`=time) %>% 
+  melt(id.vars=c("method", "iter")) %>% 
+  group_by(method, variable) %>% 
+  summarise(lower=quantile(value, 0.25),
+            middle=quantile(value, 0.5),
+            upper=quantile(value, 0.75)) %>% 
+  mutate(model="B. 250D normal")
 
-temp_df %>%
-  ggplot(aes(x=time, y=r_star, colour=as.factor(method))) +
-  geom_jitter()
+g1 <- temp_df %>%
+  bind_rows(temp_df1) %>% 
+  bind_rows(temp_df2) %>% 
+  bind_rows(temp_df3) %>% 
+  mutate(model=fct_relevel(as.factor(model), "A. AR(1)", "B. 250D normal", "C. Cauchy", "D. 8 schools")) %>%
+  ggplot(aes(x=as.factor(method), y=middle)) +
+  geom_pointrange(aes(ymin=lower, ymax=upper)) +
+  xlab("Method") +
+  ylab("Value") +
+  theme(text=element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = c(0.83, 0.79),
+        legend.text = element_text(size=12, colour="black"),
+        legend.title = element_text(size=14, colour="black")) +
+  facet_grid(vars(variable), vars(model), scales="free")
+ggsave("../output/ml_comparison_all.pdf", g1, width = 12, height = 6)
 ```

### 2020-10-20 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
started to add ml comparison section

@@
 \color{red}
-\subsection{ML model sensitivity}\label{sec:ml_model}
+\section{ML sensitivity}\label{sec:ML_sensitivity}
+In this section, we investigate how two decisions about classifiers — which classifier to use (in \S\ref{sec:ml_model}) and what hyperparameters to use for it (in \S\ref{sec:hyperparameters}) — affect calculation of $R^*$.
+The test cases used for empirical evaluation were selected from our pool of examples introduced in the text.
+Specifically, they were chosen to represent cases where $R^*$ was nearer 1, since, for these, small changes in calculated $R^*$ could lead to different decisions about convergence being reached.
+The four examples we used were:
+
+\begin{itemize}
+	\item The AR(1) example described in \S\ref{sec:heterogeneity}.
+	\item The 250-dimensional non-centered multivariate normal example introduced in \S\ref{sec:multivariate_normal_250}.
+In each replicate, we generate 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
+	\item The alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy}.
+For each replicate, we generated samples from the model using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
+	\item The non-centered parameterisation of the eight schools model described in \S\ref{sec:eight_shools}.
+In each replicate, we used Stan's NUTS algorithm with adapt delta set to 0.95 to generate 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains; these were then used to calculate $R^*$.
+\end{itemize}
+
+\subsection{ML classifier comparison}\label{sec:ml_model}
 In this section, we investigate how classification accuracy depends on choice of ML classifier.
+We restricted our analysis to popular classifiers with relatively few hyperparameters to simplify the comparison amongst them.
+Notably, we do not consider neural network models, since the structure of nets effectively defines a high dimensional hyperparameter space.
+
+The ML classifiers we compared were GBMs, RFs, k-nearest neighbours (KNNs), SVMs with linear kernels and a generalised linear model approach (GLM).
+All these models were called through the \textbf{\textsf{R}}'s Caret package \citep{kuhn2008building}: the GBM is implemented using the \textit{gbm} model in Caret which uses the ``gbm'' package \citep{greenwell2019package}; RFs are implemented using the \textit{rf} model in Caret which uses the ``randomForest'' \citep{liaw2002classification} package; KNN is natively implemented in Caret and called using the \textit{knn} model; SVMs are implemented in Caret using the \textit{svmLinear} model which uses ``kernlab'' package \citep{karatzoglou2004kernlab}; and the GLM is implemented through the \textit{multinom} model in the ``nnet'' package \citep{ripley2016package}.
+
+Each of these classifiers has hyperparameters, and, due to the differences between the examples we consider (defined in \S\ref{sec:ML_sensitivity}), the optimal hyperparameters were likely to differ between examples.
+For each classifier-example combination, we first performed a single replicate of each experiment and choose those hyperparameters which maximised classification accuracy.
+Hyperparameter sets were chosen so that training time for each classifier remained reasonable and were the same across all examples:
+
+\begin{itemize}
+	\item GBM: 
+\end{itemize}
+
+These hyperparameters were then used in subsequent replicates.
+We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 100 replicates; for the multivariate normal example, we used 100; for the Cauchy model, 50 replicates; and for the eight schools model, 30 replicates.
+
+The results of these experiments are shown in Fig. \ref{fig:ml_comparison_all}.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{ml_comparison_all.pdf}}
+	\caption{\textbf{ML classifier comparison.} Top row shows $R^*$ values calculated using Algorithm \ref{alg:R_star}; bottom row show the time taken to run each method on a desktop computer.
+The different columns show the examples described in \S\ref{sec:ML_sensitivity}.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates; note, that the number of replicates varies across examples as given in \S\ref{sec:ml_model}.}
+	\label{fig:ml_comparison_all}
+\end{figure}
 
@@
 \subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
+In this section, we considered only the two classifiers — GBMs and RFs — that performed best on our comparison between ML methods in \S\ref{sec:ml_model}.
 The performance of GBMs and RFs, like most ML methods, depends on their hyperparameters.
@@
 
-In this section, we investigate the sensitivity of $R^*$ calculated by GBMs and RFs to each of these sets of hyperparameters across a range of examples.
-The test cases picked correspond to examples already introduced throughout the manuscript: in general, we chose to investigate cases where $R^*$ was nearer 1, since, for these, small changes in calculated $R^*$ could lead to different decisions about convergence being reached.
+In this section, we investigate the sensitivity of $R^*$ calculated by GBMs and RFs to each of these sets of hyperparameters across a range of examples (defined in \S\ref{sec:ML_sensitivity}).
+Since each example is of different dimensionalities, we compare different sets of hyperparameters in each, although include amongst the sets the default values we suggest for each classifier in \S\ref{sec:method}.
+We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 200 replicates; for the multivariate normal example, we used 20; for the Cauchy model, 20 replicates; and for the eight schools model, 50 replicates.
 
 \subsubsection{Autoregressive example}\label{sec:hyperparameters_ar1}
-In this section, we assume the same data generating process as described in \S\ref{sec:heterogeneity}.
-For each generated dataset, we fit GBMs with all 16 possible combinations of interaction depth: $(1,3,5,7)$; and number of regression trees: $(1,10,50,100)$.
-Using the same data generating process, we also fit RFs with the two allowable $m_{\text{try}}$ values: $(1,2)$.
-At each combination of classifier and hyperparameters, with performed 200 replicates: in each case, calculating the $R^*$ value.
-
 In Fig.
@@
 \subsubsection{Cauchy model: alternative parameterisation}\label{sec:hyperparameters_cauchy}
-In this section, we revisit the alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy} and investigate how $R^*$ as calculated by GBM and RF models depends on hyperparameters.
 Note, that for the GBM model, we did not investigate a full range of hyperparameters needed to optimise $R^*$ due to the extensive training time needed to do so.
-For each combination of model and hyperparameters, we generated samples from the model as described in \S\ref{sec:cauchy} and performed 20 replicates at each of these combinations.
 
@@
 \subsubsection{Eight schools: non-centered parameterisation}\label{sec:hyperparameters_8_schools}
-In this section, we revisit the non-centered parameterisation of the eight schools model described in \S\ref{sec:eight_shools}.
-We investigated a range of hyperparameters for GBM and RF classifiers.
-At each combination of classifier and hyperparameter setting, we performed 50 replicates: in each replicate, we used Stan's NUTS algorithm with adapt delta set to 0.95 to generate posterior samples; these were then used to calculate $R^*$.
-
 In Fig.

### 2020-10-20 src/s_jobScheduler_hypers_rf.sh

@@ -5,7 +5,8 @@ chmod +x *.sh
 
 
 # Rscript s_run_hyperparameters_rf.R 1 200
-Rscript s_run_hyperparameters_rf.R 2 50
+# Rscript s_run_hyperparameters_rf.R 2 50
 # Rscript s_run_hyperparameters_rf.R 3 20
-# Rscript s_run_hyperparameters_rf.R 4 20
+Rscript s_run_hyperparameters_rf.R 4 20
+Rscript s_run_hyperparameters.R 4 20
 

### 2020-10-20 src/s_run_hyperparameters.R

@@ -91,7 +91,7 @@ f_data_normal <- function(){
   N <- 250
   A <- rWishart(1, 250, diag(N))[,,1]
   model <- stan_model("mvt_250_ncp.stan")
-  fit <- sampling(model, data=list(N=N, A=A), iter=10, chains=4)
+  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
   x <- rstan::extract(fit, permuted=F)
   return(x)
 }
@@ -120,7 +120,7 @@ if(model_num==1){
 }else if(model_num==4){
   m_parameters <- expand_grid(int.depth=c(3, 7, 11),
                               n.trees=c(50, 100, 400))
-  f_data_gen <- f_data_cauchy
+  f_data_gen <- f_data_normal
   name <- "normal_250"
 }
 

### 2020-10-20 src/s_run_hyperparameters_rf.R

@@ -88,7 +88,7 @@ f_data_normal <- function(){
   N <- 250
   A <- rWishart(1, 250, diag(N))[,,1]
   model <- stan_model("mvt_250_ncp.stan")
-  fit <- sampling(model, data=list(N=N, A=A), iter=10, chains=4)
+  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
   x <- rstan::extract(fit, permuted=F)
   return(x)
 }
@@ -111,7 +111,7 @@ if(model_num==1){
   f_data_gen <- f_data_cauchy
   name <- "cauchy"
 }else if(model_num==4){
-  m_parameters <- expand_grid(mtry=seq(7, 30, 4))
+  m_parameters <- expand_grid(mtry=seq(10, 30, 4))
   f_data_gen <- f_data_normal
   name <- "normal_250"
 }

### 2020-10-22 src/s_bivariate_normal.Rmd
nearly through trends

--- 2020-08-27 src/s_bivariate_normal.Rmd
@@
 # Calculate R* dist
-r_star_vals <- r_star_dist(full_samples)
+r_star_vals_gbm <- r_star(full_samples, uncertainty = T, method = "gbm")
+r_star_vals_rf <- r_star(full_samples, uncertainty = T, method = "rf")
+mean(r_star_vals_gbm>1)
+mean(r_star_vals_gbm)
+mean(r_star_vals_rf>1)
+mean(r_star_vals_rf)
 
 g <- 
-  qplot(r_star_vals) +
+  tibble(RF=r_star_vals_rf, GBM=r_star_vals_gbm, iter=seq_along(GBM)) %>%
+  melt(id.vars="iter") %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position = "identity", alpha=0.8) +
   xlab(TeX("$R*$")) +
@@
   ylab("Count") +
-  theme(text=element_text(size=14, colour="black"),
-        axis.text = element_text(colour="black"))
+  theme(text=element_text(size=20, colour="black"),
+        axis.text = element_text(colour="black")) +
+  scale_fill_grey("Classifier")
 ggsave("../output/bivariate.pdf", g, width = 8, height = 6)

### 2020-10-22 src/s_hyperparameters.Rmd
nearly through trends

@@
         legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.95, 2.8)) +
+  scale_y_continuous(limits=c(0.95, 4.1)) +
   geom_hline(yintercept = 1, linetype=2) +
@@
             upper=quantile(value, 0.75)) %>%
-  ggplot(aes(x=as.factor(mtry), y=middle, group=1)) +
+  mutate(default=if_else(mtry==22, 1, 0)) %>%
+  ggplot(aes(x=as.factor(mtry), y=middle, group=1, shape=as.factor(default))) +
   geom_pointrange(aes(ymin=lower, ymax=upper), position=position_dodge2(width=0.1)) +
@@
         legend.position = c(0.7, 0.25)) +
-  scale_y_continuous(limits=c(0.95, NA)) +
+  scale_y_continuous(limits=c(0.95, 4.1)) +
   geom_hline(yintercept = 1, linetype=2) +

### 2020-10-22 src/s_mvt_normal.Rmd
nearly through trends

@@
 full_data <- rstan::extract(fit, permuted=F)
-a_df <- tibble(samples_10000_cp=r_star_dist(full_data))
+a_df <- tibble(samples_10000_cp_gbm=r_star(full_data, uncertainty = T, method="gbm"),
+               samples_10000_cp_rf=r_star(full_data, uncertainty = T, method="rf"))
 
@@
 full_data <- rstan::extract(fit, permuted=F)
-r_star2 <- r_star_dist(full_data)
+r_star2_gbm <- r_star(full_data, uncertainty = T, method="gbm")
+r_star2_rf <- r_star(full_data, uncertainty = T, method="rf")
 
 a_df <- a_df %>%
-  mutate(samples_400=r_star2)
+  mutate(samples_400_gbm=r_star2_gbm,
+         samples_400_rf=r_star2_rf)
 
@@
 
-r_star3 <- r_star_dist(full_data)
+r_star3_gbm <- r_star(full_data, uncertainty = T, method = "gbm")
+r_star3_rf <- r_star(full_data, uncertainty = T, method = "rf")
 
 a_df <- a_df %>%
-  mutate(samples_10000_ncp=r_star3)
+  mutate(samples_10000_ncp_gbm=r_star3_gbm,
+         samples_10000_ncp_rf=r_star3_rf)
 
@@
 a_df <- readRDS("../output/mvt_fit_all_accuracy.rds")
+a_df_gbm <- a_df %>%
+  select(contains("gbm")) %>% 
+  rename(samples_10000_cp=samples_10000_cp_gbm,
+         samples_400=samples_400_gbm,
+         samples_10000_ncp=samples_10000_ncp_gbm) %>% 
+  mutate(Classifier="A. GBM")
+a_df_rf <- a_df %>%
+  select(contains("rf")) %>% 
+  rename(samples_10000_cp=samples_10000_cp_rf,
+         samples_400=samples_400_rf,
+         samples_10000_ncp=samples_10000_ncp_rf) %>% 
+  mutate(Classifier="B. RF")
 
 ga <- 
-  a_df %>% 
+  a_df_gbm %>% 
   melt() %>% 
@@
 dev.off()
+
+# provide GBM vs RF results in supplementary
+ga1 <- 
+  a_df_gbm %>% 
+  bind_rows(a_df_rf) %>% 
+  melt() %>% 
+  ggplot(aes(x=value, fill=fct_relevel(as.factor(variable), "samples_400", "samples_10000_cp", "samples_10000_ncp"))) +
+  geom_histogram(position="identity", alpha=0.8, bins=50) +
+  scale_fill_grey("# draws", labels=c("400, cp", "10,000, cp", "10,000, ncp")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  theme(text = element_text(size=18, colour="black"),
+        legend.position = c(0.8, 0.75),
+        axis.text = element_text(colour="black")) +
+  facet_wrap(~Classifier)
+ggsave("../output/mvt_gbm_vs_rf.pdf", ga1, width = 12, height = 8)
+
+a_df_gbm %>% 
+  bind_rows(a_df_rf) %>% 
+  melt() %>% 
+  group_by(Classifier, variable) %>% 
+  summarise(value=mean(value)) %>% 
+  pivot_wider(id_cols = "variable", names_from = "Classifier", values_from = "value")
 ```

### 2020-10-22 src/s_trends.Rmd
nearly through trends

@@
 ```{r}
-# g <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
-#   select(trend, iters, rep, dimension, Rstar_split, method) %>% 
-#   unique() %>%
-#   select(-rep) %>% 
-#   melt(c("trend", "iters", "dimension", "method")) %>% 
-#   filter(variable=="Rstar_split") %>% 
-#   ggplot(aes(y = value, x = trend)) + 
-#   geom_jitter(aes(colour=as.factor(method))) + 
-#   facet_grid(vars(iters), vars(dimension)) + 
-#   geom_hline(yintercept = 1) +
-#   theme(strip.text.x = element_text(size=16, colour="black"),
-#         strip.text.y = element_text(size=16, colour="black"),
-#         axis.text.x = element_text(size=16, colour="black"),
-#         axis.text.y = element_text(size=16, colour="black"),
-#         axis.title = element_text(size=16, colour="black")) +
-#   xlab("Trend") +
-#   ylab("Value")
-# g
-# 
+g <- readRDS("../output/experiment_trends_one_dim.rds") %>%
+  select(trend, iters, rep, dimension, Rstar_split, method) %>%
+  unique() %>%
+  select(-rep) %>%
+  melt(c("trend", "iters", "dimension", "method")) %>%
+  filter(variable=="Rstar_split") %>%
+  ggplot(aes(y = value, x = trend)) +
+  geom_jitter(aes(colour=as.factor(method))) +
+  facet_grid(vars(iters), vars(dimension)) +
+  geom_hline(yintercept = 1) +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black")) +
+  xlab("Trend") +
+  ylab("Value")
+g
+
 # ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
@@
                             variable=="Rstar_split"~"B. Split-R*")) %>% 
+  mutate(iters=as.numeric(as.character(iters))) %>% 
+  mutate(method=as.character(method)) %>% 
+  mutate(method=if_else(method=="rf", "RF", "GBM")) %>% 
   ggplot(aes(y = value, x = rho)) +
@@
   ylab("Value") +
-  xlab(TeX("$\\rho$"))
+  xlab(TeX("$\\rho$")) +
+  scale_colour_grey("Classifier")
 g

### 2020-10-22 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
nearly through trends

@@
 By evaluating the performance of classifiers on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, our chosen classifiers naturally exploit differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
+To maximise predictive accuracy, our chosen classifiers naturally exploit differences in the full joint distributions between chains, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
 Our statistic, unlike its $\widehat{R}$ cousins, is scalar-valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
@@
 
-\textcolor{red}{To identify promising candidate ML methods, we run a series of experiments using a number of popular classifiers (see Section X).
+\textcolor{red}{To identify promising candidate ML methods, we run a series of experiments using a number of popular classifiers (see \S\ref{sec:ml_model}).
 Two methods performed consistently well across our examples: gradient-boosted regression trees (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}) and random forest classifiers \citep{breiman2001random} (``RF'').
@@
 \textcolor{red}{Since different posteriors present different challenges to MCMC samplers, the nature of classification boundaries is problem-specific.
-Because of this, there is no unique optimal classifier across all problems, and the performance of the algorithms we investigate depends on their hyperparameters (see Section X).
-GBM models have a number of hyperparameters and have previously been demonstrated to be hard to tune \cite[Chapters~12]{boehmke2019hands}.
+Because of this, there is no unique optimal classifier across all problems, and the performance of the algorithms we investigate depends on their hyperparameters (see \S\ref{sec:hyperparameters}).
+GBM models have a number of hyperparameters and have previously been demonstrated to be hard to tune \cite[Chapter~12]{boehmke2019hands}.
 To ensure practical run time for this method, we suggest running it using a default hyperparameter set: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees be grown.
 This choice of hyperparameters was chosen to present a balance between training cost and classification accuracy, producing an $R^*$  that was a stringent measure of convergence — in general, more stringent than $\widehat{R}$.
-RFs are generally less sensitive to variation in their single hyperparameter: $m_{\text{try}}$, the size of the subset of all features over which to search for an optimal split \citep{boehmke2019hands}.
-Our experiments replicate these results (see Section X), and we suggest running RFs using the heuristic $m_{\text{try}}=\sqrt{K}$, which has previously been shown to be a choice resulting in robust classification performance \citep{bernard2009influence,boehmke2019hands}.
+RFs are generally less sensitive to variation in their single hyperparameter: $m_{\text{try}}$, the size of the subset of all features over which to search for an optimal split \cite[Chapter~11]{boehmke2019hands}.
+Our experiments replicate these results (see \S\ref{sec:hyperparameters}), and we suggest running RFs using the heuristic $m_{\text{try}}=\sqrt{K}$, which has previously been shown to be a choice resulting in robust classification performance \citep{bernard2009influence,boehmke2019hands}.
 Unless otherwise stated, in the examples explored in \S\ref{sec:results}, the hyperparameters for these two classifiers were set at these defaults.}
@@
 \subsubsection{Performance of $R^*$}\label{sec:heterogeneity_performance}
-To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e.
+To illustrate the consistency of $R^*$, we performed 1000 replicates where, in each case, we generated four $\{X_t\}$ series as described (i.e.
 where one chain has a lower variance).
@@
 \subsection{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
-In section SX, we perform a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
+In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
 In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
@@
 Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32.
-For three `chains', we generate 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
+For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
 To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
@@
 In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom.
-In the remaining chain, we generate the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
+In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
 We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$.
@@
 To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$.
-For each combination of dimensions and $\nu$, we perform 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
+For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
 Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.
@@
 These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
-This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and section X (hyperparameters)): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
+This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
 Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
-RFs are less sensitive to hyperparameter variation (section X), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
+RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
 Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
@@
 
-Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty}, which is shown in Fig.
+Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty} \textcolor{red}{using both GBM and RF classifiers.
+These distributions are shown in Fig.
 \ref{fig:bivariate}.
-The mean of this distribution is 1.14, and $>$99\% of $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
-By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
+The mean of the GBM-$R^*$ distribution is 1.14, and $>$99\% of $R^*$ draws are above 1; the mean of the RF-$R^*$ distribution was 1.27 and all draws were above 1.
+Collectively, these measures indicate that the sampling distribution has not converged.} By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
 
@@
 	\centerline{\includegraphics[width=0.6\textwidth]{bivariate.pdf}}
-	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws.}
+	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws as calculated using Algorithm \ref{alg:R_star_uncertainty} for both the GBM and RF classifiers.}
 	\label{fig:bivariate}
@@
 where $\boldsymbol{x}\in\mathbb{R}^{250}$.
-For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig.
-\ref{fig:mvt}A.
-From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
+For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} with \textcolor{red}{a GBM classifier to generate an uncertainty distribution for $R^*$, which is shown in Fig.
+\ref{fig:mvt}A (the equivalent plot for a RF classifier is similar and shown in Fig.
+\ref{fig:mvt_gbm_vs_rf}).} From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
 Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.

### 2020-10-22 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
nearly through trends

@@
 	\maketitle
+
+\color{red}
+\section{Diagnosing convergence in joint distributions: multivariate normal models supplementary}\label{sec:multivariate_normal_supp}
+
+In Fig.
+\ref{fig:mvt_gbm_vs_rf}, we compare the performance of GBM and RF classifiers on the multivariate normal example described in \S\ref{sec:multivariate_normal_250}.
+The two classifiers produced similar $R^*$ distributions across the various examples: for the 400 draw example, the GBM $R^*$ distribution had a mean of 2.86 and the RF equivalent was 3.06; for the example with 10,000 samples from the centered parameterisation, the GBM classifier had a $R^*$ distribution mean of 1.07 versus 1.08 from the RF; for the example with 10,000 samples from the non-centered parameterisation, both classifiers produced a mean of 1.00.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{mvt_gbm_vs_rf.pdf}}
+	\caption{\textbf{Multivariate normal example: GBM (A) versus RF (B) classifiers.} Shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws).
+Here, we show 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} for each MCMC run.}
+	\label{fig:mvt_gbm_vs_rf}
+\end{figure}
+
+\color{black}
 
@@
 In this section, we investigate how classification accuracy depends on choice of ML classifier.
-We restricted our analysis to popular classifiers with relatively few hyperparameters to simplify the comparison amongst them.
+We restricted our analysis to popular classifiers with relatively few hyperparameters to simplify comparison amongst them.
 Notably, we do not consider neural network models, since the structure of nets effectively defines a high dimensional hyperparameter space.
@@
 For each classifier-example combination, we first performed a single replicate of each experiment and choose those hyperparameters which maximised classification accuracy.
-Hyperparameter sets were chosen so that training time for each classifier remained reasonable and were the same across all examples:
+The set of all hyperparameters we selected between were chosen so that training time for each classifier remained reasonable.
+These sets were the same across all examples:
 
 \begin{itemize}
-	\item GBM: 
+	\item GBM: Cartesian product of $\text{int.depth}=(3, 7)$; $\text{\# trees} = (50, 100)$;
+	\item RF: $m_{\text{try}} = (1, 2)$.
+(Note, in \S\ref{sec:hyperparameters}, we investigate a dimension-specific approach to $m_{\text{try}}$ but, here, for ease of comparison with the other methods, we keep this static.)
+	\item KNN: $K=(5,10,15, 20,40)$;
+	\item SVM: $C=(0.25, 0.5, 0.75)$;
+	\item GLM: $\text{decay}=(0.1, 0.2, 0.5, 1)$.
 \end{itemize}
@@
 
-The results of these experiments are shown in Fig. \ref{fig:ml_comparison_all}.
+The results of these experiments are shown in Fig.
+\ref{fig:ml_comparison_all}.
+The top row shows $R^*$ as calculated by each classifier (in each case, chains split into two halves); the bottom row shows the time (in seconds) taken for $R^*$ calculation.
+The panels correspond to the examples described in \S\ref{sec:ML_sensitivity}.
+Across the examples, the classification accuracy of RFs and GBMs were, generally, highest; followed by KNNs.
+An exception was for the 250-dimensional normal where KNN performed best.
+In the AR(1) example, GBM outperformed RF.
+In higher dimensional cases, RF outperformed GBM.
+In all cases, RF and GBM classifiers took the greatest time to train; in higher dimensional examples, RF-based $R^*$ were cheaper to calculate than those from GBMs.
 
@@
 	\caption{\textbf{ML classifier comparison.} Top row shows $R^*$ values calculated using Algorithm \ref{alg:R_star}; bottom row show the time taken to run each method on a desktop computer.
-The different columns show the examples described in \S\ref{sec:ML_sensitivity}.
+The different columns correspond to the examples described in \S\ref{sec:ML_sensitivity}.
 Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates; note, that the number of replicates varies across examples as given in \S\ref{sec:ml_model}.}
@@
 In this section, we investigate the sensitivity of $R^*$ calculated by GBMs and RFs to each of these sets of hyperparameters across a range of examples (defined in \S\ref{sec:ML_sensitivity}).
-Since each example is of different dimensionalities, we compare different sets of hyperparameters in each, although include amongst the sets the default values we suggest for each classifier in \S\ref{sec:method}.
+Since each example is of different dimensionalities, we compare different sets of hyperparameters in each, although include amongst the sets the default values we suggest for each classifier in \S\ref{sec:method} (these are shown as triangles on the plots).
+Note, that for the GBM model, we did not investigate a full range of hyperparameters needed to optimise $R^*$ for many of the examples due to the extensive training time needed to do so.
+For RFs, runtime was less restrictive and we were better able to survey the sensitivity across hyperparameter space.
+
 We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 200 replicates; for the multivariate normal example, we used 20; for the Cauchy model, 20 replicates; and for the eight schools model, 50 replicates.
@@
 In Fig.
-\ref{fig:hypers_ar1}, we show the results of this analysis.
+\ref{fig:hypers_ar1}, we show the results of the AR(1) analysis.
 In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
@@
 In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
-In panel A, we see that $R^*$ does vary according to hyperparameters: of the set investigated, an interaction depth of 1 appears optimal with 10+ trees.
-In panel B, the values of $R^*$ are generally lower than those achieved by GBMs: again, RFs appear less well-suited to lower dimensional problems as shown in \S\ref{sec:comparison_gbm_rf}.
+For the GBM model, $R^*$ varied according to hyperparameters: of the set investigated, an interaction depth of 1 appears optimal with 10+ trees.
+In panel B, the values of $R^*$ are generally lower than those achieved by GBMs.
+As shown in \S\ref{sec:comparison_gbm_rf}, RFs appear less well-suited to lower dimensional problems.
 Across the two possible hyperparameter values for RFs, there was not a substantial difference in the values of $R^*$ calculated.
@@
 Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.}
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
 	\label{fig:hypers_ar1}
@@
 
-\subsubsection{Multivariate normal: 250-dimensional model}
+\subsubsection{Multivariate normal: 250-dimensional model}\label{sec:hyperparameters_normal}
+In Fig.
+\ref{fig:hypers_normal}, we show the results of the analysis for the 250-dimensional normal example.
+In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
+In panel B, we show the same but for RFs.
+This shows that $R^*$ was higher as calculated by RFs and that performance of GBMs depended strongly on hyperparameters.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1.0\textwidth]{hypers_normal_250.pdf}}
+	\caption{\textbf{Hyperparameter sensitivity: 250-dimensional normal example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the Cauchy example described in \S\ref{sec:hyperparameters_normal}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
+	\label{fig:hypers_normal}
+\end{figure}
 
 \subsubsection{Cauchy model: alternative parameterisation}\label{sec:hyperparameters_cauchy}
-Note, that for the GBM model, we did not investigate a full range of hyperparameters needed to optimise $R^*$ due to the extensive training time needed to do so.
-
 In Fig.
-\ref{fig:hypers_ar1}, we show the results of this analysis.
+\ref{fig:hypers_cauchy}, we show the results of the Cauchy model analysis.
 In panel A, we show $R^*$ as calculated by a GBM classifier; in panel B, the same but using a RF classifier.
@@
 Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.}
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
 	\label{fig:hypers_cauchy}
@@
 In Fig.
-\ref{fig:hypers_8_schools}, we show the results of these experiments.
-Note that, due to length of computational runtime, we chose not to explore the full range of hyperparameters needed to optimise the performance for GBMs; for RFs, the maximum allowed value of $m_{\text{try}}=19$.
+\ref{fig:hypers_8_schools}, we show the results of the experiments on the eight schools model.
 Panel A shows how $R^*$ calculated using a GBM varies with its two hyperparameters; panel B shows the same but for RFs.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
 This shows that, in this case, the value of $R^*$ calculated by RFs was higher than that for GBMs.
@@
 Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.}
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
 	\label{fig:hypers_8_schools}

### 2020-10-22 src/s_cauchy.Rmd
small changes

@@
 
-laccuracy <- future_map_dbl(iter, function(i) r_star(x[1:i, , ], method="xgbTree"))
+laccuracy <- future_map_dbl(iter, function(i) r_star(x[1:i, , ], method="gbm"))
 
@@
 
-laccuracy1 <- future_map_dbl(iter, function(i) r_star(x[1:i, , ], method="xgbTree"))
+laccuracy1 <- future_map_dbl(iter, function(i) r_star(x[1:i, , ], method="gbm"))
 
@@
        nom=laccuracy)
-saveRDS(a_df, "../output/cauchy_r_star_xgbtree.rds")
+saveRDS(a_df, "../output/cauchy_r_star_gbm.rds")
 ```
@@
 ```{r}
-a_df <- readRDS("../output/cauchy_r_star_xgbtree.rds") %>% 
+a_df <- readRDS("../output/cauchy_r_star_gbm.rds") %>% 
   mutate(model="gbm")

### 2020-10-22 src/s_wide_distributions.Rmd
small changes

@@
 full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star <- r_star_dist(full_data)
+r_star_gbm <- r_star(full_data, uncertainty = T, method="gbm")
+r_star_rf <- r_star(full_data, uncertainty = T, method="rf")
 
-a_df <- tibble(samples_cp=r_star)
+a_df <- tibble(samples_cp_gbm=r_star_gbm,
+               samples_cp_rf=r_star_rf)
 
@@
 full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star <- r_star_dist(full_data)
+r_star_gbm <- r_star(full_data, uncertainty = T, method="gbm")
+r_star_rf <- r_star(full_data, uncertainty = T, method="rf")
 
 a_df <- a_df %>%
-  mutate(samples_ncp=r_star)
-# saveRDS(a_df, "../output/mvt_500_all.rds")
+  mutate(samples_ncp_gbm=r_star_gbm,
+         samples_ncp_rf=r_star_rf)
+saveRDS(a_df, "../output/mvt_500_all.rds")
 a_df <- readRDS("../output/mvt_500_all.rds")
+a_df_gbm <- a_df %>%
+  select(contains("gbm")) %>% 
+  rename(samples_cp=samples_cp_gbm,
+         samples_ncp=samples_ncp_gbm) %>% 
+  mutate(Classifier="A. GBM")
+a_df_rf <- a_df %>%
+  select(contains("rf")) %>% 
+  rename(samples_cp=samples_cp_rf,
+         samples_ncp=samples_ncp_rf) %>% 
+  mutate(Classifier="B. RF")
+
 g <- 
-  a_df %>% 
-  melt() %>% 
+  a_df_gbm %>%
+  bind_rows(a_df_rf) %>% 
+  melt(id.vars="Classifier") %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
         legend.position = c(0.8, 0.8)) +
-  ggtitle("A.")
+  facet_wrap(~Classifier)
 # ggsave("../output/wide_data.pdf", g, width = 12, height = 8)
@@
 full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star <- r_star_dist(full_data)
-a_df <- tibble(samples_400=r_star)
-# saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
+r_star_gbm <- r_star(full_data, uncertainty = T, method="gbm")
+r_star_rf <- r_star(full_data, uncertainty = T, method="rf")
+a_df <- tibble(samples_400_gbm=r_star_gbm,
+               samples_400_rf=r_star_rf)
+saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
 mon <- monitor_extra(split_data(full_data))
@@
 full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
-r_star1 <- r_star_dist(full_data)
+r_star_gbm <- r_star(full_data, uncertainty = T, method="gbm")
+r_star_rf <- r_star(full_data, uncertainty = T, method="rf")
 
+a_df <- readRDS("../output/mvt_wide_400_summary.rds")
 a_df <- a_df %>% 
-  tibble(samples_1000=r_star1)
+    mutate(samples_1000_gbm=r_star_gbm,
+           samples_1000_rf=r_star_rf)
 saveRDS(a_df, "../output/mvt_wide_summaries.rds")
@@
 a_df <- readRDS("../output/mvt_wide_summaries.rds")
+a_df_gbm <- a_df %>%
+  select(contains("gbm")) %>% 
+  rename(samples_1000cp=samples_cp_gbm,
+         samples_ncp=samples_ncp_gbm) %>% 
+  mutate(Classifier="A. GBM")
+a_df_rf <- a_df %>%
+  select(contains("rf")) %>% 
+  rename(samples_cp=samples_cp_rf,
+         samples_ncp=samples_ncp_rf) %>% 
+  mutate(Classifier="B. RF")
+
 a_df <- tibble(samples_400=a_df$.$samples_400, samples_1000=a_df$samples_1000)

### 2020-10-22 src/s_cauchy.Rmd
added trends

@@
 
-r_star_nom <- r_star_dist(x)
+r_star_nom_gbm <- r_star(x, method="gbm", uncertainty = T)
+r_star_nom_rf <- r_star(x, method="rf", uncertainty = T)
 
@@
 
-r_star_alt <- r_star_dist(full_data)
+r_star_alt_gbm <- r_star(x, method="gbm", uncertainty = T)
+r_star_alt_rf <- r_star(x, method="rf", uncertainty = T)
+
 mon <- monitor_extra(split_data(full_data))
@@
 ```{r}
-a_df <- tibble(nominal=r_star_nom,
-               alt=r_star_alt,
+a_df <- tibble(nominal_gbm=r_star_nom_gbm,
+               nominal_rf=r_star_nom_rf,
+               alt_gbm=r_star_alt_gbm,
+               alt_rf=r_star_alt_rf,
                source="r_star") %>% 
@@
   mutate(source=if_else(source=="r_star", "R*", "split-Rhat"))
-g <-
-  a_df %>% 
-  melt() %>% 
-  ggplot(aes(x=value, fill=as.factor(variable))) +
-  geom_histogram(position="identity", alpha=0.8) +
-  scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
-  xlab("Value") +
-  ylab("Count") +
-  geom_vline(xintercept = 1, linetype=2) +
-  theme(text = element_text(size=14, colour="black"),
-        legend.position = "none") +
-  facet_wrap(~source, scales = "free")
-g
+# g <-
+#   a_df %>% 
+#   melt() %>% 
+#   ggplot(aes(x=value, fill=as.factor(variable))) +
+#   geom_histogram(position="identity", alpha=0.8) +
+#   scale_fill_grey("Parameterisation", labels=c("nominal", "alternative")) +
+#   xlab("Value") +
+#   ylab("Count") +
+#   geom_vline(xintercept = 1, linetype=2) +
+#   theme(text = element_text(size=14, colour="black"),
+#         legend.position = "none") +
+#   facet_wrap(~source, scales = "free")
+# g
 ```
@@
 
-r_star_nom <- r_star_dist(full_data)
+r_star_nom_gbm <- r_star(full_data, method="gbm", uncertainty = T)
+r_star_nom_rf <- r_star(full_data, method="rf", uncertainty = T)
 mon_nom <- monitor_extra(split_data(full_data))
@@
 
-r_star_alt <- r_star_dist(full_data)
+r_star_alt_gbm <- r_star(full_data, method="gbm", uncertainty = T)
+r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)
 mon_alt <- monitor_extra(split_data(full_data))
@@
 ```{r}
-b_df <- tibble(nominal=r_star_nom,
-               alt=r_star_alt,
+b_df <- tibble(nominal_gbm=r_star_nom_gbm,
+               nominal_rf=r_star_nom_rf,
+               alt_gbm=r_star_alt_gbm,
+               alt_rf=r_star_alt_rf,
                source="r_star") %>% 
@@
 
-```{r}
-# a_df <- readRDS("../output/cauchy_r_star.rds")
-g2 <- 
-  a_df %>% 
-  melt(id.vars="iteration") %>% 
-  ggplot(aes(x=iteration, y=value, colour=as.factor(variable))) +
-  geom_point() +
-  geom_line() +
-  ylab(TeX("$R^*$")) +
-  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
-  xlab("Iteration") +
-  theme(text=element_text(size=16, colour="black"),
-        axis.text = element_text(colour="black"),
-        legend.position = c(0.8, 0.7),
-        legend.text = element_text(size=12, colour="black"),
-        legend.title = element_text(size=14, colour="black"),
-        axis.title.y = element_blank()) +
-  geom_hline(yintercept = 1, linetype=2) +
-  scale_x_log10() +
-  scale_y_continuous(limits=c(0.8, NA), breaks = seq(1, 7, 1)) +
-  ggtitle(TeX("D. Split-$R^*$"))
-g2
-```
-
 Add in GBM R*
@@
 a_df <- readRDS("../output/cauchy_r_star_gbm.rds") %>% 
-  mutate(model="gbm")
+  mutate(model="GBM")
 a_df1 <- readRDS("../output/cauchy_r_star_rf.rds") %>% 
-  mutate(model="rf")
+  mutate(model="RF")
 g2 <- a_df %>%
@@
   ylab(TeX("$R^*$")) +
-  scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
+  scale_color_grey(guide="none") +
   xlab("Iteration") +
@@
         axis.text = element_text(colour="black"),
-        legend.position = c(0.83, 0.79),
+        legend.position = c(0.83, 0.77),
         legend.text = element_text(size=12, colour="black"),
@@
   ggtitle(TeX("D. Split-$R^*$"))+
-  scale_linetype_discrete("Model") +
-  guides(colour = guide_legend(nrow = 1, byrow = T),
+  scale_linetype_discrete("Classifier") +
+  guides(
          linetype=guide_legend(nrow = 1, byrow = T))
-ggsave("../output/fig_9d.pdf", g2, width = 12, height = 8)
 ```
@@
   scale_color_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
-  scale_fill_grey("Parameterisation", labels=c("Alternative", "Nominal")) +
   xlab("Iteration") +
@@
         axis.text = element_text(colour="black"),
-        legend.position = "none",
+        legend.position = c(0.8, 0.6),
         axis.title.y = element_blank()) +

### 2020-10-22 src/s_trends.Rmd
added trends

--- 2020-10-17 src/s_trends.Rmd
@@
                             variable=="Rstar_split"~"B. Split-R*")) %>%
+  mutate(method=as.character(method)) %>% 
+  mutate(method=if_else(method=="gbm", "GBM", "RF")) %>% 
   ggplot(aes(y = value, x = trend)) +
-  geom_jitter(aes(colour=as.factor(method))) +
+  geom_jitter(aes(colour=as.factor(method), shape=as.factor(method))) +
   facet_grid(vars(iters), vars(variable)) +
@@
   xlab("Trend") +
-  ylab("Value")
-# ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)
+  ylab("Value") +
+  scale_colour_grey("Classifier") +
+  scale_shape("Classifier")
+ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)
 ```
@@
 ```{r}
-# g <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
-#   select(trend, iters, rep, dimension, Rstar_split, method) %>% 
-#   unique() %>%
-#   select(-rep) %>% 
-#   melt(c("trend", "iters", "dimension", "method")) %>% 
-#   filter(variable=="Rstar_split") %>% 
-#   ggplot(aes(y = value, x = trend)) + 
-#   geom_jitter(aes(colour=as.factor(method))) + 
-#   facet_grid(vars(iters), vars(dimension)) + 
-#   geom_hline(yintercept = 1) +
-#   theme(strip.text.x = element_text(size=16, colour="black"),
-#         strip.text.y = element_text(size=16, colour="black"),
-#         axis.text.x = element_text(size=16, colour="black"),
-#         axis.text.y = element_text(size=16, colour="black"),
-#         axis.title = element_text(size=16, colour="black")) +
-#   xlab("Trend") +
-#   ylab("Value")
-# g
-# 
-# ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
+g <- readRDS("../output/experiment_trends_one_dim.rds") %>%
+  select(trend, iters, rep, dimension, Rstar_split, method) %>%
+  unique() %>%
+  select(-rep) %>%
+  melt(c("trend", "iters", "dimension", "method")) %>%
+  filter(variable=="Rstar_split") %>%
+  mutate(method=as.character(method)) %>% 
+  mutate(method=if_else(method=="gbm", "GBM", "RF")) %>% 
+  ggplot(aes(y = value, x = trend)) +
+  geom_jitter(aes(colour=as.factor(method),
+                  shape=as.factor(method))) +
+  facet_grid(vars(iters), vars(dimension)) +
+  geom_hline(yintercept = 1) +
+  theme(strip.text.x = element_text(size=16, colour="black"),
+        strip.text.y = element_text(size=16, colour="black"),
+        axis.text.x = element_text(size=16, colour="black"),
+        axis.text.y = element_text(size=16, colour="black"),
+        axis.title = element_text(size=16, colour="black")) +
+  xlab("Trend") +
+  ylab("Value") +
+  scale_colour_grey("Classifier") +
+  scale_shape("Classifier")
+g
+
+ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
 ```
@@
                             variable=="Rstar_split"~"B. Split-R*")) %>% 
+  mutate(iters=as.numeric(as.character(iters))) %>% 
+  mutate(method=as.character(method)) %>% 
+  mutate(method=if_else(method=="rf", "RF", "GBM")) %>% 
   ggplot(aes(y = value, x = rho)) +
-  geom_jitter(aes(colour=as.factor(method))) +
+  geom_jitter(aes(colour=as.factor(method),
+                  shape=as.factor(method))) +
   facet_grid(vars(iters), vars(variable)) +
@@
   ylab("Value") +
-  xlab(TeX("$\\rho$"))
+  xlab(TeX("$\\rho$"))  +
+  scale_colour_grey("Classifier") +
+  scale_shape("Classifier")
 g
@@
 ```{r}
-res <- readRDS("../output/experiment_trends_rho_nearer1.rds")
+res <- readRDS("../output/experiment_trends_rho_nearer1.rds") %>%
+  mutate(method=as.character(method)) %>% 
+  mutate(method=if_else(method=="gbm", "GBM", "RF"))
 
@@
 g <- ggplot(data = res, aes(x = rho, y = Rstar_split)) +
-  geom_jitter(aes(colour=as.factor(method))) +
+  geom_jitter(aes(colour=as.factor(method),
+                  shape=as.factor(method))) +
   facet_grid(vars(iters)) +
@@
         axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black"))
+        axis.title = element_text(size=16, colour="black"))  +
+  scale_colour_grey("Classifier") +
+  scale_shape("Classifier")
 ggsave("../output/trends_ar1.pdf", g, width = 8, height = 6)
@@
 b_sum <- res %>%
-  group_by(iters, rho) %>%
+  group_by(iters, rho, method) %>%
   summarise(lower=quantile(Rstar_split, 0.025),
@@
   geom_line(data=filter(res1, !is.na(type))) +
-  facet_wrap(~rho, scales="free") +
+  facet_grid(vars(rho), vars(method), scales="free") +
   geom_hline(yintercept = 1) +

### 2020-10-22 src/s_wide_distributions.Rmd
added trends

@@
         axis.text = element_text(colour="black"),
-        legend.position = c(0.8, 0.8)) +
+        legend.position = c(0.45, 0.5)) +
   facet_wrap(~Classifier)
@@
   select(contains("gbm")) %>% 
-  rename(samples_1000cp=samples_cp_gbm,
-         samples_ncp=samples_ncp_gbm) %>% 
-  mutate(Classifier="A. GBM")
+  rename(samples_400=samples_400_gbm,
+         samples_1000=samples_1000_gbm) %>% 
+  mutate(Classifier="C. GBM")
 a_df_rf <- a_df %>%
   select(contains("rf")) %>% 
-  rename(samples_cp=samples_cp_rf,
-         samples_ncp=samples_ncp_rf) %>% 
-  mutate(Classifier="B. RF")
+  rename(samples_400=samples_400_rf,
+         samples_1000=samples_1000_rf) %>% 
+  mutate(Classifier="D. RF")
 
-a_df <- tibble(samples_400=a_df$.$samples_400, samples_1000=a_df$samples_1000)
 g1 <- 
-  a_df %>% 
-  melt() %>% 
+  a_df_gbm %>%
+  bind_rows(a_df_rf) %>% 
+  melt(id.vars="Classifier") %>% 
   ggplot(aes(x=value, fill=as.factor(variable))) +
@@
         axis.text = element_text(colour="black"),
-        legend.position = c(0.8, 0.8)) +
-  ggtitle("B.")
+        legend.position = c(0.5, 0.5)) +
+  facet_wrap(~Classifier)
 ggsave("../output/very_wide_data.pdf", g, width = 12, height = 8)
@@
 pdf("../output/mvt_wide_both.pdf", width = 12, height = 6)
-multiplot(g, g1, cols = 2)
+multiplot(g, g1, cols = 1)
 dev.off()

### 2020-10-22 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
added trends

@@
 After c.1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more draws are needed.
-In Fig.
-\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the alternative model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence.
-Finally, in Figs.
+\textcolor{red}{In Fig.
+\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models and for both GBM and RF classifiers: these indicate that, after 25,000 iterations, for the alternative model, $R^*\approx 1.05$ for the GBM classifier and $R^*\approx 1.74$ for the RF classifier; for the nominal model, $R^*>2$ for the GBM classifier $R^*>3$ for the RF classifier: all these $R^*$ values suggest lack of convergence.} Finally, in Figs.
 \ref{fig:cauchy_convergence}E\&F, we plot the minimum across all the dimensions of tail- and bulk-ESS calculated as described in \cite{vehtari2019rank}.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{cauchy_convergence.pdf}}
-	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS.
+	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star} for both the GBM and RF classifiers; and E and F show tail- and bulk-ESS.
 Dashed lines indicate recommended thresholds for each convergence statistic.}

### 2020-10-22 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
added trends

@@
 \eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
-Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
-\ref{fig:mvt_wide_both}A.
-In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
-The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
+Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide''.
+\textcolor{red}{We also calculated $R^*$ distributions using both the GBM and RF classifiers: the $R^*$ distribution in each case is shown in Figs.
+\ref{fig:mvt_wide_both}A\&B.
+This figure shows that, across both parameterisations, convergence has not yet been reached since all $R^*$ distributions are shifted rightwards of 1.} The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
 \ref{fig:wide_both_diagnostics}A), since, for both parameterisations, some of the parameters had $\widehat{R}>1.01$.
@@
 	\centerline{\includegraphics[width=1\textwidth]{mvt_wide_both.pdf}}
-	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
+	\caption{\textbf{Wide data examples.} Top row shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; bottom row shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
+In the left column, results for the GBM classifier are shown; in the right, we show the same for the RF classifier.
 In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.}
@@
 In this case, we assume independent standard normals for each dimension.
-In Fig.
-\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
-In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
+\textcolor{red}{In Figs.
+\ref{fig:mvt_wide_both}C\&D, we plot the $R^*$ distributions from GBM and RF classifiers for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.} In all cases, the distributions were right of $R^*=1$, indicating non-convergence.
 These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.

### 2020-10-22 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
started to change trends text

@@
 In this section, we investigate whether $R^*$ can detect non-stationary sampling distributions.
+\textcolor{red}{In all the examples, we present results for both GBM and RF classifiers.}
 
@@
 The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
-This plot mirrors Fig.
+\textcolor{red}{The points and triangles show $R^*$ derived from GBM and RF classifiers, respectively.} This plot mirrors Fig.
 4 in the supplementary materials of \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{trends_all_dim.pdf}}
-	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
+	\caption{\textbf{Univariate trends example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
 The rows present the differing sample sizes.
@@
 In the simulations here, all dimensions bar one are stationary; the remaining dimension has a linear trend added to it.
-In all cases, split-$R^*$ increased with trend.
+In all cases, \textcolor{red}{across both GBM and RF classifiers}, split-$R^*$ increased with trend.
 Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{trends_one_dim.pdf}}
-	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
+	\caption{\textbf{Multivariate trends example with split-$R^*$.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
 The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{trends_joint_distribution.pdf}}
-	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
+	\caption{\textbf{Bivariate normal with trending correlation example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
 The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example.} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
+	\caption{\textbf{Non-stationary distribution: AR(1) example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
 At each parameter set, 10 replicates were run and jitter was added to the points.
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1_transposed.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
+	\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} Column A shows the results for a GBM classifier; column B for a RF classifier.
+The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
 At each parameter set, 10 replicates were run and jitter was added to the points.

### 2020-10-23 src/s_cauchy.Rmd
added cauchy and sorted trends section

@@
 
-r_star_alt_gbm <- r_star(x, method="gbm", uncertainty = T)
-r_star_alt_rf <- r_star(x, method="rf", uncertainty = T)
+r_star_alt_gbm <- r_star(full_data, method="gbm", uncertainty = T)
+r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)
 
@@
 mon_alt <- monitor_extra(split_data(full_data))
-```
 
-Plot
-```{r}
 b_df <- tibble(nominal_gbm=r_star_nom_gbm,
@@
 saveRDS(c_df, "cauchy_rhat_rstar.rds")
+```
+
+Plot
+```{r}
 c_df <- readRDS("cauchy_rhat_rstar.rds")
-my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.03) else c(0.9, 3)}
+c_df_rhat <- c_df %>% 
+  filter(source=="split-Rhat") %>% 
+  select(source, nominal, alt, iter)
+c_df_rstar <- c_df %>% 
+  filter(source!="split-Rhat")
+c_df_gbm <- c_df_rstar %>%
+  select(contains("gbm"), source, iter) %>% 
+  rename(nominal=nominal_gbm,
+         alt=alt_gbm) %>% 
+  mutate(source="R*: GBM")
+c_df_rf <- c_df_rstar %>%
+  select(contains("rf"), source, iter) %>% 
+  rename(nominal=nominal_rf,
+         alt=alt_rf) %>% 
+  mutate(source="R*: RF")
 
-g <- c_df %>% 
+my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.06) else c(0.9, 3)}
+g <- c_df_rhat %>% 
+  bind_rows(c_df_gbm) %>% 
+  bind_rows(c_df_rf) %>% 
   melt(id.vars=c("source", "iter")) %>% 

### 2020-10-23 src/s_trends.Rmd
started to change trends text
added cauchy and sorted trends section

@@
 ```{r}
+# remove double iterates for rhat from rfs
 g <- readRDS("../output/experiment_trends_joint_distribution.rds") %>%
@@
   melt(c("rho", "iters", "method")) %>%
+  filter(!((method=="rf")&(variable=="zsRhat"))) %>% 
   mutate(value=as.numeric(as.character(value))) %>% 
@@
 
+res <- readRDS("../output/experiment_trends_rho_nearer1.rds") %>%
+  mutate(method=as.character(method)) %>% 
+  mutate(method=if_else(method=="gbm", "A. GBM", "B. RF"))
+
 b_sum <- res %>%

### 2020-10-23 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
added cauchy and sorted trends section

@@
 		& Sensitivity of $R^*$ to number of chains & \ref{sec:heterogeneity_numchains}\\
-		& Robustness of $R^*$ to ML model hyperparameters & \ref{sec:heterogeneity_robustness}\\
 		\midrule
@@
 		Ovarian and prostate models & \textbf{Bayesian models with many parameters and multimodal posteriors} & \ref{sec:prostate}\\
+		\midrule		
+		\textcolor{red}{Discrete Markov model} & \textbf{Evaluating $R^*$ on discrete parameter models} & \ref{sec:discrete}\\
+		& Small state-space & \ref{sec:discrete_small}\\
+		& Large state-space & \ref{sec:discrete_large}\\
+		\midrule
+		\textcolor{red}{Various} & \textbf{Sensitivity of $R^*$ to ML model} & \ref{sec:ML_sensitivity}\\
+		& Comparing different ML classifiers & \ref{sec:ml_model}\\
+		& Sensitivity of $R^*$ to GBM and RF hyperparameters & \ref{sec:hyperparameters}\\
+		\midrule
+		\textcolor{red}{Multivariate normal \& Student-t dists.} & \textbf{Comparing GBM and RF classifiers} & \ref{sec:comparison_gbm_rf}\\
+		& Detecting convergence in joint distributions & \ref{sec:joint_distribution}\\
+		& Tail convergence & \ref{sec:tail_fatness}\\
 	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
@@
 
-\color{red}
-\subsection{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
-In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
-In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
-In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}.
-In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution.
-In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains.
-Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}.
-This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw.
-Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.
-
-\subsubsection{Joint distribution}\label{sec:joint_distribution}
-Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32.
-For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
-To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
-To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$.
-For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
-In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.
-
-The results of these experiments are shown in Fig.
-\ref{fig:gbm_vs_rf_normal}.
-In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic.
-When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$.
-In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence.
-As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows.
-$R^*$ as calculated by GBMs and RFs also increases.
-For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification.
-For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
-	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
-The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
-For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
-The dashed line shows $R^*=1$.
-Here, $R^*$ was calculated using chains split into two halves.}
-	\label{fig:gbm_vs_rf_normal}
-\end{figure}
-
-\subsubsection{Fat tails}\label{sec:tail_fatness}
-We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution.
-To do so, we again consider four chains.
-In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom.
-In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
-We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$.
-We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness.
-Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$.
-To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$.
-For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
-Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.
-
-The results of these experiments are shown in Fig.
-\ref{fig:gbm_vs_rf_studdentt}.
-Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions.
-Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$.
-The point colour and shadings give the method used to calculate $R^*$.
-In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method.
-As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases.
-In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers.
-In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate.
-If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.
-
-Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
-The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
-In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
-These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
-This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
-Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
-RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
-Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
-	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution.
-The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
-The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
-For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
-The dashed line shows $R^*=1$.
-Here, $R^*$ was calculated using chains split into two halves.}
-	\label{fig:gbm_vs_rf_studdentt}
-\end{figure}
-
 \color{black}
@@
 
-In the top-left panel of Fig.
-\ref{fig:cauchy}, we show the $R^*$ distribution under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
+\textcolor{red}{In the top-left and top-middle panel of Fig.
+\ref{fig:cauchy}, we show the $R^*$ distribution for GBM and RF classifiers under both parameterisations.} As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
 The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet its mean remains above this value.
@@
 \ref{fig:cauchy}, we show the results for these longer runs.
-In these, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
-Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
+\textcolor{red}{In these, the alternative parameterisation now has an $R^*$ distribution centred on $R^*=1$ for the GBM classifier although the RF classifier $R^*$ distribution remains slightly rightwards of this target indicating that convergence is nearer but more iterations are likely still required.} Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
 The $\widehat{R}$ values are all below 1.01, indicating convergence in both cases.
@@
 	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
-Columns show the $R^*$ distributions (left) and rank-normalised split-$\widehat{R}$ values across all parameters (right).
+Columns show the $R^*$ distributions from GBM (left) and RF (middle) and rank-normalised split-$\widehat{R}$ values across all parameters (right).
 Shadings indicate different model paramerisations as indicated in legend.}

### 2020-10-23 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
added cauchy and sorted trends section

@@
 
+\color{red}
+\section{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
+In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
+In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
+In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}.
+In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution.
+In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains.
+Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}.
+This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw.
+Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.
+
+\subsection{Joint distribution}\label{sec:joint_distribution}
+Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32.
+For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
+To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
+To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$.
+For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
+In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.
+
+The results of these experiments are shown in Fig.
+\ref{fig:gbm_vs_rf_normal}.
+In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic.
+When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$.
+In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence.
+As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows.
+$R^*$ as calculated by GBMs and RFs also increases.
+For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification.
+For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
+	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
+For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
+The dashed line shows $R^*=1$.
+Here, $R^*$ was calculated using chains split into two halves.}
+	\label{fig:gbm_vs_rf_normal}
+\end{figure}
+
+\subsection{Fat tails}\label{sec:tail_fatness}
+We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution.
+To do so, we again consider four chains.
+In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom.
+In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
+We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$.
+We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness.
+Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$.
+To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$.
+For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
+Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.
+
+The results of these experiments are shown in Fig.
+\ref{fig:gbm_vs_rf_studdentt}.
+Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions.
+Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$.
+The point colour and shadings give the method used to calculate $R^*$.
+In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method.
+As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases.
+In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers.
+In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate.
+If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.
+
+Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
+The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
+In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
+These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
+This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
+Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
+RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
+Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
+	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution.
+The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
+For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
+The dashed line shows $R^*=1$.
+Here, $R^*$ was calculated using chains split into two halves.}
+	\label{fig:gbm_vs_rf_studdentt}
+\end{figure}
+\color{black}
+
 \bibliographystyle{chicago}

### 2020-10-23 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
small change to ovarian examples

@@
 In this section, we analyse two Bayesian models both fit to real data.
-The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
+\textcolor{red}{In both examples, we illustrate $R^*$ distributions derived only from GBM classifiers.} The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
 Since there are relatively few data points relative to the number of predictors, regularised horseshoe priors are specified on the regression coefficients \citep{piironen2017sparsity} since most are expected to be zero.
@@
 	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) using the GBM classifier are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
 	\label{fig:ovarian}
@@
 	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) using the GBM classifier are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
 	\label{fig:prostate}

### 2020-10-23 src/s_eight_schools.Rmd
added rf results for 8 schools model

@@
 ```
-# Random forest
 ## Break each chain into two then make flattened data
@@
 x <- rstan::extract(fit_cp, permuted=F)
-r_star_centered <- r_star_dist(x)
+r_star_centered <- r_star(x, uncertainty = T)
 mon_cp <- monitor_extra(split_data(x))
@@
 x <- rstan::extract(fit_ncp, permuted=F)
-r_star_noncentered <- r_star_dist(x)
+r_star_noncentered <- r_star(x, uncertainty = T)
 mon_ncp <- monitor_extra(split_data(x))
@@
 # Fit gradient-boosed model
+caretGrid <- tibble(interaction.depth=c(3), 
+                             n.trees = 50,
+                             shrinkage=c(0.1),
+                             n.minobsinnode=10)
 fit <- train(chain ~ ., data = training_data, 
-                 method = "rf",
+                 method = "gbm",
                  trControl = trainControl(method = 'none'), 
-                 verbose=FALSE)
+                 verbose=FALSE, tuneGrid=caretGrid)
 varImp(fit)
@@
 x <- rstan::extract(fit_cp, permuted=F)
-r_star_centered <- r_star_dist(x, split_chains = F)
+r_star_centered <- r_star(x, split_chains = F, uncertainty = T, method="gbm")
 mon_cp <- monitor_extra(fit_cp)
@@
 x <- rstan::extract(fit_ncp, permuted=F)
-r_star_noncentered <- r_star_dist(x, split_chains = F)
+r_star_noncentered <- r_star(x, split_chains = F, uncertainty = T, method="gbm")
 mon_ncp <- monitor_extra(fit_ncp)
@@
 
-pdf("../output/fig_10.pdf", width = 12, height = 4)
+pdf("../output/eight_schools.pdf", width = 12, height = 4)
 multiplot(g1, g2, g3, cols = 3)
@@
 ```
-# xgbTree
+
+# Same but for RF
+```{r}
+# Centered
+x <- rstan::extract(fit_cp, permuted=F)
+r_star_centered <- r_star(x, uncertainty = T, method="rf")
+mon_cp <- monitor_extra(split_data(x))
+
+# Noncentered
+x <- rstan::extract(fit_ncp, permuted=F)
+r_star_noncentered <- r_star(x, uncertainty = T, method="rf")
+mon_ncp <- monitor_extra(split_data(x))
+
+c_df <- tibble(centered=r_star_centered,
+               non_centered=r_star_noncentered,
+               metric="R*",
+               simulations="split")
+c1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
+                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
+                metric="split-Rhat",
+               simulations="split")
+```
+
+```{r}
+x <- rstan::extract(fit_cp, permuted=F)
+r_star_centered <- r_star(x, split_chains = F, uncertainty = T, method="rf")
+mon_cp <- monitor_extra(fit_cp)
+# noncentered
+x <- rstan::extract(fit_ncp, permuted=F)
+r_star_noncentered <- r_star(x, split_chains = F, uncertainty = T, method="rf")
+mon_ncp <- monitor_extra(fit_ncp)
+
+d_df <- tibble(centered=r_star_centered,
+               non_centered=r_star_noncentered,
+               metric="R*",
+               simulations="unsplit")
+d1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
+                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
+                metric="split-Rhat",
+               simulations="unsplit")
+```
+
+Plot two R* distributions
+```{r}
+g1 <- 
+  c_df %>% 
+  melt(id.vars=c("metric", "simulations")) %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  xlim(0.78, 2) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("A.") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = "none")
+g2 <- 
+  d_df %>% 
+  melt(id.vars=c("metric", "simulations")) %>% 
+  ggplot(aes(x=value, fill=as.factor(variable))) +
+  geom_histogram(position="identity", alpha=0.8) +
+  scale_fill_grey("Series", labels=c("centered", "non-centered")) +
+  xlim(0.78, 2) +
+  xlab(TeX("$R*$")) +
+  ylab("Count") +
+  geom_vline(xintercept = 1, linetype=2) +
+  ggtitle("B.") +
+  theme(text = element_text(size=16, colour="black"),
+        axis.text = element_text(colour="black"),
+        legend.position = "none")
+# a_df <- readRDS("../output/eight_schools_replicates_4_index.rds")
+```
+
+Look at Rhat distributions
+```{r}
+pdf("../output/eight_schools_rf.pdf", width = 12, height = 4)
+multiplot(g1, g2, cols = 2)
+dev.off()
+```
 
@@
 
-pdf("../output/fig_11.pdf", width = 6, height = 4)
+pdf("../output/eight_schools_r_star_quantiles.pdf", width = 6, height = 4)
 plot_grid(gb, g1, ga, g2, nrow = 2, rel_widths = c(1, 0.4), rel_heights = c(0.9, 1))

### 2020-10-23 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
added rf results for 8 schools model

@@
 To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
+Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models \textcolor{red}{using a GBM classifier}.
 The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
@@
 Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears near convergence.
+
+\textcolor{red}{Fig.
+\ref{fig:eight_schools_rf} shows the equivalent of Fig.
+\ref{fig:eight_schools} except using a RF classifier.
+The results are similar, although the $R^*$ distributions are shifted slightly rightwards: indicating, for example, that more that 2000 draws from the non-centered model may be required for convergence.}
 
@@
 The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} for each parameterisation.}
+In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} using a GBM classifier for each parameterisation.}
 	\label{fig:eight_schools}

### 2020-10-23 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
added rf results for 8 schools model

@@
 	\label{fig:mvt_gbm_vs_rf}
+\end{figure}
+
+\color{black}
+
+\color{red}
+\section{Hierarchical model: Eight schools model supplementary}\label{sec:8_schools_supp}
+
+Fig. \ref{fig:eight_schools_rf} shows the equivalent of Fig. \ref{fig:eight_schools} except using a RF classifier.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=1\textwidth]{eight_schools_rf.pdf}}
+	\caption{\textbf{Eight schools example: $R^*$ distributions for a RF classifier.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
+The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
+In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} for each parameterisation.}
+	\label{fig:eight_schools_rf}
 \end{figure}

### 2020-10-23 src/s_eight_schools.Rmd
finished first draft incorporating both gbm and rf classifiers

@@
 x <- rstan::extract(fit_cp, permuted=F)
-g1 <- plot_r_star_quantiles(1, x, nsim=40, split_chains = F) +
-  scale_y_continuous(limits = c(0.6, 2.6)) +
+g1 <- plot_r_star_quantiles(1, x, nsim=40, split_chains = F, method="gbm") +
+  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6)) +
   theme(axis.title.y = element_blank(),
@@
 
-g2 <- plot_r_star_quantiles(2, x, nsim=40, split_chains = F) +
-  scale_y_continuous(limits = c(0.6, 2.6)) +
+g2 <- plot_r_star_quantiles(2, x, nsim=40, split_chains = F, method="gbm") +
+  scale_y_continuous(limits = c(0.6, 1.7), breaks = c(0.6, 0.8, 1, 1.2, 1.4, 1.6)) +
   theme(axis.title.y = element_blank(),
@@
 
-pdf("../output/eight_schools_r_star_quantiles.pdf", width = 6, height = 4)
+pdf("../output/eight_schools_r_star_quantiles1.pdf", width = 6, height = 4)
 plot_grid(gb, g1, ga, g2, nrow = 2, rel_widths = c(1, 0.4), rel_heights = c(0.9, 1))

### 2020-10-23 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
finished first draft incorporating both gbm and rf classifiers

@@
 %
-The ``non-centered'' parameterisation recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
+The ``non-centered'' parameterisation (first introduced in \cite{van2001art}) recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
 This can be written as,
@@
 %
-In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation 
-of the distribution of these effects.
+In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation of the distribution of these effects.
 In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
@@
 In Fig.
-\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
+\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$ \textcolor{red}{using a GBM classifier}.
 In the left-hand column of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
@@
 That said, the computational time for calculating $\widehat{R}$ was considerably less than was needed for $R^*$.
+
+\textcolor{red}{In \S\ref{sec:discrete}, we evaluate $R^*$ on univariate discrete examples: one with four states (in \S\ref{sec:discrete_small}); another, with a larger state-space consisting of 20 states (in \S\ref{sec:discrete_large}).
+In these examples, we use a discrete Markov model to generate draws from a given target.
+The small and larger state-space cases, show that $R^*$ behaves as expected: given a sufficient sample size, it is able to detect differences in the transition probability matrix between chains that result in differences in the target distribution.}
+
+\textcolor{red}{In \S\ref{sec:ML_sensitivity}, we investigate how two decisions about classifiers — which classifier to use (in \S\ref{sec:ml_model}) and what hyperparameters to use for it (in \S\ref{sec:hyperparameters}) — affect calculation of $R^*$.
+In \S\ref{sec:ml_model}, we test a range of popular classifiers: GBMs, RFs, k-nearest-neighbour models, support vector machines and generalised linear models across examples.
+This indicated that GBMs and RFs consistently had the highest classification accuracy across the examples: in higher dimensional problems, RFs tended to best GBMs.
+In \S\ref{sec:hyperparameters}, we show how these two best classifiers — GBMs and RFs — depend on their hyperparameters.
+Across the examples we test, GBMs are more sensitive to hyperparameter choices than are RFs.
+Additionally, GBMs require typically more rounds of boosting in higher dimensional problems, leading to more extensive algorithm runtime.
+Because of this, we suggest fixing GBM's hyperparameters to values that result in reasonable performance and reasonable runtime.
+For RFs, we suggest using a heuristic for choosing its hyperparameters that was derived in a previous empirical evaluation of RFs \citep{bernard2009influence}.}
+
+\textcolor{red}{In \S\ref{sec:comparison_gbm_rf}, we then calculate $R^*$ using both GBM and RF classifiers with our suggested default hyperparameter values (given in \S\ref{sec:method}).
+In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution (using a multivariate normal target).
+In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains (using Student-t targets).
+We both examples, our draws are generated by independent sampling meaning that we can also calculate an optimal $R^*$ based on the Bayes optimal classifier (see, for example, \citep{devroye2013probabilistic} and \S\ref{sec:comparison_gbm_rf} for more information).
+Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
+The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
+In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
+These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
+This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
+Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
+RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
+Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.}
 

### 2020-10-23 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
finished first draft incorporating both gbm and rf classifiers

@@
 
-Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
-The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
-In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
-These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
-This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
-Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
-RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
-Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
-
 \begin{figure}[!htb]

### 2020-10-24 src/s_bivariate_normal.Rmd
first draft done

@@
   scale_fill_grey("Classifier")
-ggsave("../output/bivariate.pdf", g, width = 8, height = 6)
+ggsave("../output/bivariate.pdf", g, width = 8, height = 4)
 ```

### 2020-10-24 src/s_mvt_normal.Rmd
first draft done

@@
   facet_wrap(~Classifier)
-ggsave("../output/mvt_gbm_vs_rf.pdf", ga1, width = 12, height = 8)
+ggsave("../output/mvt_gbm_vs_rf.pdf", ga1, width = 12, height = 4)
 

### 2020-10-24 src/s_trends.Rmd
first draft done

@@
         axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black")) +
+        axis.title = element_text(size=16, colour="black"),
+        legend.text = element_text(size=16, colour="black"),
+        legend.title = element_text(size=16, colour="black")) +
   xlab("Trend") +
@@
         axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black")) +
+        axis.title = element_text(size=16, colour="black"),
+        legend.text = element_text(size=16, colour="black"),
+        legend.title = element_text(size=16, colour="black")) +
   ylab("Value") +
@@
         axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black"))  +
+        axis.title = element_text(size=16, colour="black"),
+        legend.text = element_text(size=16, colour="black"),
+        legend.title = element_text(size=16, colour="black"))  +
   scale_colour_grey("Classifier") +
@@
         axis.text.y = element_text(size=16, colour="black"),
-        axis.title = element_text(size=16, colour="black"))
+        axis.title = element_text(size=16, colour="black"),
+        legend.text = element_text(size=16, colour="black"),
+        legend.title = element_text(size=16, colour="black"))
 

### 2020-10-24 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
first draft done

@@
 Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
-For each draw, $s$, in our testing set, \textcolor{red}{classifiers} outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
+For each draw, $s$, in our testing set, \textcolor{red}{classifiers} output a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
@@
 \ref{fig:eight_schools} except using a RF classifier.
-The results are similar, although the $R^*$ distributions are shifted slightly rightwards: indicating, for example, that more that 2000 draws from the non-centered model may be required for convergence.}
+The results are similar, although the $R^*$ distributions are shifted slightly rightwards: indicating, for example, that more than 2000 draws from the non-centered model may be required for convergence.}
 
@@
 In Fig.
-\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$ \textcolor{red}{using a GBM classifier}.
+\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
 In the left-hand column of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
@@
 
-\textcolor{red}{In \S\ref{sec:comparison_gbm_rf}, we then calculate $R^*$ using both GBM and RF classifiers with our suggested default hyperparameter values (given in \S\ref{sec:method}).
+\textcolor{red}{In \S\ref{sec:comparison_gbm_rf}, we use a slew of examples to compare $R^*$ calculated using GBM and RF classifiers using our suggested default hyperparameter sets (given in \S\ref{sec:method}).
 In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution (using a multivariate normal target).
 In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains (using Student-t targets).
-We both examples, our draws are generated by independent sampling meaning that we can also calculate an optimal $R^*$ based on the Bayes optimal classifier (see, for example, \citep{devroye2013probabilistic} and \S\ref{sec:comparison_gbm_rf} for more information).
+In both examples, draws are generated by independent sampling meaning that an optimal $R^*$ can be calculated based on the Bayes optimal classifier (see, for example, \citep{devroye2013probabilistic} and \S\ref{sec:comparison_gbm_rf} for more information).
 Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
@@
 In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
-These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-higher ones.
+These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-high ones.
 This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
@@
 
-When first starting to develop $R^*$, it was unclear to us whether \textit{any} metric aligned with a supervised ML method would be overly sensitive to its hyperparameters.
-With gradient-boosted regression tree models, we found this not to be the case.
-Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples: perhaps, because the classification boundaries for this task are unlikely to be very complex compared to (say) machine vision tasks.
-To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
+\color{red}
+Different target distributions present different challenges to sampling.
+Because of this, there is not a unique optimal ML classifier across all cases: this is just a manifestation of the no free lunch theorem \citep{wolpert1997no}.
+Across a range of examples we tested (see \S\ref{sec:ml_model}), GBM and RF classifiers performed consistently well, and we then used these across all other illustrative examples.
+It is possible — indeed, likely — that another ML model may exist or be invented that consistently outperforms both these classifiers.
+We do not see this as a problem: across the examples we considered, $R^*$ calculated using both classifiers tended to provide a measure of convergence as or more stringent than existing diagnostics.
+In that sense, it is a step in the right direction.
+If a better classifier is found, the same apparatus we develop here can be used, and this will present a harsher test of convergence.
+
+A different question is, ``When is such a measure of convergence of practical use?''.
+In our examples, $R^*$ is able to diagnose poor convergence in the tails of marginal distributions: likely of practical relevance for many applications that require tail quantiles.
+$R^*$ is also able to diagnose lack of convergence in the joint distribution — even if the marginals appear converged.
+Indeed, it is less clear how the joint distribution can be unconverged when the marginals appear so, but we found examples where this appeared true.
+A consequence of poor convergence of the joint distribution would be for prediction and, by corollary, model comparison.
+Further work examining these consequences further is needed, and $R^*$ can help to identify and monitor fruitful candidate systems.
+\color{black}
 
@@
 \section{Acknowledgements}
-\textcolor{red}{The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.}
+\textcolor{red}{The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.
+We would also like to thank Paul B{\"u}rkner and Jonah Gabry, with whom useful discussions were had during the preparation of this manuscript.}
 

### 2020-10-24 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
first draft done

@@
 This illustrates that there is an ordering among the stationary distributions $\pi_1\sim\pi_2\sim\pi_3$, meaning $\pi_1$ is most similar to $\pi_2$, and $\pi_3$ is most similar to $\pi_2$.
-Put another way, $\pi_3$ is most different from $p_1$.
+Put another way, $\pi_3$ is most different from $\pi_1$.
 
@@
 	\item The 250-dimensional non-centered multivariate normal example introduced in \S\ref{sec:multivariate_normal_250}.
-In each replicate, we generate 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
+In each replicate, we generated 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
 	\item The alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy}.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{gbm_vs_rf_normal.pdf}}
-	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+	\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_normal.pdf}}
+	\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
 The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
@@
 	\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution.
-The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
 The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.

### 2020-10-24 src/s_ar1.Rmd
redraft sent to aki

@@
         axis.text = element_text(colour="black", size=16)) +
-  xlim(0.9, NA)
+  xlim(0.8, 1.45)
 
@@
         axis.text = element_text(colour="black"),
-        legend.position = c(0.8, 0.8))
+        legend.position = c(0.8, 0.8)) +
+  scale_x_continuous(limits=c(0.8, 1.45))
 

### 2020-10-24 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
redraft sent to aki

@@
 Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan}, PyMC3 \citep{salvatier2016probabilistic}, \textcolor{red}{Turing \citep{ge2018turing}, TensorFlow Probability  \citep{dillon2017tensorflow} and Pyro \citep{bingham2019pyro}} implement.
 MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior.
-Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
+Because of this, software packages (\textcolor{red}{Stan \citep{carpenter2017stan}, for example}), go to great lengths to communicate to users any issues with sampling.
 
@@
 
-\textcolor{red}{Here, we introduce $R^*$, a new convergence metric.
+\textcolor{red}{Here, we introduce $R^*$, a new convergence diagnostic metric.
 This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
@@
 
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
+For the types of problem we \textcolor{red}{tested}, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
 In addition, since ML classifiers can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
@@
 
-The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank} and elsewhere.
+The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ \textcolor{red}{across a range of scenarios}.
 Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
@@
 If Markov chains have not mixed, it is possible to guess (with more accuracy than chance) to which chain a draw belongs from its value.
-This is possible if there are differences in the sampling distribution for any dimension, $\theta$, in the target distribution (Fig.
+This is possible if there are differences in the sampling distribution for any dimension in the target distribution (Fig.
 \ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a draw belongs to.
@@
 These two cases, whilst simple, illustrate the basis of our approach.
-To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each draw belongs.
+To determine if a set of Markov chains has converged to the same distribution, we train a supervised \textcolor{red}{ML} model to classify the chain to which each draw belongs.
 By evaluating its performance on a separate test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
+By taking the ratio of \textcolor{red}{classification} accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
@@
 Intuitively, the reduction in sampling variation when composing training and test sets from larger samples results in lower variability in ML model predictions.
-We also expect that larger sample sizes should lead to higher $R^*$ values, since more training data leads to better ML models.
+We also expect that larger sample sizes should lead to higher $R^*$ values \textcolor{red}{with lower variance}, since more training data leads to \textcolor{red}{ML models with lower generalisation error}.
 We may start to see this here, since the median $R^*=1.25$ at a sample size of 8000 was greater than for the smaller samples.
@@
 We have so far focused on the sensitivity of $R^*$ to chain heterogeneity with a fixed number of chains: four.
-Since classification becomes a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
+Since classification \textcolor{red}{may become} a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
 For comparison, we also illustrate how the performance of rank-normalised split-$\widehat{R}$ varies with number of chains.
@@
 In general, both measures decline with chain count.
-For $R^*$, this is because it is harder to classify chains when there are more of them.
+For $R^*$, this \textcolor{red}{may be} because it is harder to classify chains when there are more of them.
 For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
@@
 We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm to sample from this target distribution and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
+We use Stan's NUTS algorithm \citep{betancourt2017conceptual} to sample from this target distribution and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
 First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
@@
 \subsubsection{Variable importance}\label{sec:multivariate_normal_varimportance}
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions.
+In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), \textcolor{red}{which allows us} to determine which variables were most informative for predictions.
 We now compare these with the more established metrics $\widehat{R}$ and ESS.
@@
 
-The model can be parameterised in two ways, as described in \cite{vehtari2019rank}.
+The model can be parameterised in two ways, as described in \cite{vehtari2019rank} (and introduced in \cite{van2001art}).
 The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
@@
 We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{betancourt2017conceptual}, meaning that the resultant sampling distribution is likely to be biased.
+Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for \textcolor{red}{the dynamic HMC algorithm \citep{betancourt2017conceptual} (called using the ``NUTS'' option in Stan)}, meaning that the resultant sampling distribution is likely to be biased.
 We also used the same algorithm settings to sample from the non-centered model.
@@
 To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
-After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
+After fitting the GBM model, \textcolor{red}{we group MCMC draws in the test set} into deciles and draw from the $R^*$ distribution for each decile.
 In Fig.
@@
 For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
-Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved ML predictive accuracy.
+Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved predictive accuracy.
 
 Fig.
-\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in ML classification accuracy and falsely concluding that convergence has occurred.
+\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in classification accuracy and falsely concluding that convergence has occurred.
 
@@
 In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of draws.
-Specifically, this was to test that ML classification didn't become prone to overfitting in this limit.
+Specifically, this was to test that classification didn't become prone to overfitting in this limit.
 To test this hypothesis, we investigated two scenarios using a multivariate normal target: one with a 250-multivariate normal with high correlation between dimensions using 250 post-warm-up iterations; and another normal with 10,000 independent dimensions using up to 500 post-warm-up iterations.
@@
 Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
-By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
-By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
+By taking the ratio of model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
+By extracting \textcolor{red}{classifier-predicted} chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
 Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
@@
 In \S\ref{sec:prostate}, we fit Bayesian models with many 1000s of parameters then used $R^*$ to diagnose convergence, finding that $R^*$ was considerably more expensive to calculate than $\widehat{R}$.
-The computational cost of training GBMs has been suggested to be $\mathcal{O}(N k)$, where $N$ is the number of training data points and $k$ is the number of dimensions of the target \citep{complexity2018}.
-If so, this suggests that larger models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
+\textcolor{red}{The time complexity of training RFs is thought to be $\mathcal{O}(m_{\text{try}} n_{tree} N \text{log} N)$ \cite[chapter~5]{louppe2014understanding}, and given the similarities with GBMs (which also builds many decision trees), it is likely to be similar.} If so, this suggests that larger \textcolor{red}{statistical} models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
 That said, it is possible to reduce the runtime for $R^*$ using thinned draws (although this risks losing chain idiosyncracies) and using a subset of dimensions (although this risks losing problematic dimensions).
@@
 We recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
-Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
+Additionally, our \textcolor{red}{non-parametric} calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
 If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.

### 2020-10-24 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
redraft sent to aki

@@
 \color{red}
-\section{Discrete distribution}\label{sec:discrete}
+\section{Discrete distributions}\label{sec:discrete}
 In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a univariate discrete target distribution.

### 2020-10-26 src/r_star_monitor.R
first draft of function

@@ -1,10 +1,11 @@
 library(caret)
 library(gbm)
 library(tidyverse)
+library(posterior)
 
 #' Estimate R star
 #'
-#' Compute estimate of R* using a gradient-boosted model (GBM)
+#' Compute estimate of R* using a ML classifier
 #'
 #' @param x a 3D array of samples of dimensions: # iter x # chains x # parameters
 #'
@@ -14,20 +15,26 @@ library(tidyverse)
 #' @param training_percent proportion of iterations used to train GBM model
 #' (default is 0.7)
 #'
-#' @param caret_default GBM hyperparameter settings given as a list (defaults
-#' are interaction.depth=3, n.trees=50, shrinkage=0.1, nminobsinmode=10)
+#' @param method ML classifer available in Caret R package (defaults to rf model)
+#'
+#' @param caret_default hyperparameter settings for ML classifier given as a list (defaults
+#' to NULL)
+#'
+#' @param uncertainty whether to provide a list of R* values (if true) or a single value (if false) (defaults to false)
+#'
+#' @param nsim number of values in list of R* vals
 #'
 #' @return estimated R* value
-#' 
+#'
 #' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
 r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, method=NULL,
                    uncertainty=F, nsim=1000){
-  
+
   if(split_chains)
     x <- split_data(x)
-  
+
   nparams <- dim(x)[3]
   nchains <- dim(x)[2]
   niter <- dim(x)[1]
@@ -41,32 +48,24 @@ r_star <- function(x, split_chains=T, training_percent=0.7, caret_default=NULL, 
       k <- k + 1
     }
   }
-  m_flattened <- m_flattened %>% 
+  m_flattened <- m_flattened %>%
     as.data.frame()
   colnames(m_flattened)[nparams + 1] <- "chain"
-  r <- m_flattened %>% 
+  r <- m_flattened %>%
     mutate(chain=as.factor(chain))
   # if only 1 param, add in a column of random noise since gbm requires >1 dims
   if(nparams==1)
-    r <- r %>% 
+    r <- r %>%
       mutate(V_new=rnorm(nrow(r)))
-  
+
[…]

### 2020-10-26 src/s_test_draws_df.R

@@ -1,0 +1,141 @@
+rm(list=ls())
+library(tidyverse)
+library(reshape2)
+library(posteriordb)
+library(posterior)
+source("r_star_monitor.R")
+source("monitornew.R")
+
+pd <- pdb_default() # Posterior database connection
+pn <- posterior_names(pd)
+po <- pdb_posterior("eight_schools-eight_schools_noncentered", pdb = pd)
+rpd <- reference_posterior_draws(po)
+
+temp_df <- as_draws_df(rpd)
+temp_df$.chain
+
+split_chains <- function(x) {
+  x <- as.matrix(x)
+  niter <- NROW(x)
+  if (niter == 1L) {
+    return(x)
+  }
+  half <- niter / 2
+  cbind(x[1:floor(half), ], x[ceiling(half + 1):niter, ])
+}
+
+split_chains(temp_df$`theta[1]`)
+
+# splits chains in `draws_df` object
+split_chains_draws_df <- function(x) {
+  nchain <- nchains(x)
+  k <- 1
+  for(i in 1:nchain) {
+    tmp <- x$.chain[x$.chain == i]
+    niter <- length(tmp)
+    if(niter > 1) {
+      half <- niter / 2
+      tmp[1:floor(half)] <- k
+      k <- k + 1
+      tmp[ceiling(half + 1):niter] <- k
+      k <- k + 1
+    }
+    if(i == 1)
+      chains <- tmp
+    else
+      chains <- c(chains, tmp)
+  }
+  x$.chain <- chains
+  return(x)
+}
+
+test <- split_chains_draws_df(x)
+
+#' Estimate R star
+#'
+#' Compute estimate of R* using a ML classifier
+#'
+#' @param x a draws_df object from posterior package
+#'
+#' @param split_chains a Boolean (defaults to true) indicating whether to split
+#' chains into two equal halves
+#'
+#' @param uncertainty whether to provide a list of R* values (if true) or a single value (if false) (defaults to false)
+#'
+#' @param method ML classifer available in Caret R package (defaults to rf model)
+#'
+#' @param hyperparameters hyperparameter settings for ML classifier given as a list (defaults
+#' to NULL)
+#'
+#' @param nsim number of values in list of R* vals
+#'
+#' @param training_percent proportion of iterations used to train GBM model
+#' (default is 0.7)
+#'
+#' @return estimated R* value
+#'
+#' @references Ben Lambert, Aki Vehtari (2020) R*: A robust MCMC convergence
[…]

### 2020-10-26 src/s_test_draws_df.R
started to add test cases

@@ -78,7 +78,7 @@ test <- split_chains_draws_df(x)
 #' diagnostic with uncertainty using gradient-boosted machines
 #' \emph{arXiv preprint} \code{arXiv:TBD}
 r_star_draws_df <- function(x, split_chains=T, uncertainty=F, method=NULL, hyperparameters=NULL,
-                            training_percent=0.7, nsim=1000){
+                            training_percent=0.7, nsim=1000, ...){
 
   if(split_chains)
     x <- split_chains_draws_df(x)
@@ -115,7 +115,7 @@ r_star_draws_df <- function(x, split_chains=T, uncertainty=F, method=NULL, hyper
                method = method,
                trControl = trainControl(method = 'none'),
                tuneGrid = caret_grid,
-               verbose = FALSE)
+               ...)
 
   # calculate classification accuracy then R*
   nchains <- length(unique(testing_data$.chain))
@@ -137,5 +137,17 @@ r_star_draws_df <- function(x, split_chains=T, uncertainty=F, method=NULL, hyper
   }
 }
 
-r_star_draws_df(temp_df, uncertainty=T, method="gbm")
+start <- Sys.time()
+vals1 <- r_star_draws_df(temp_df, uncertainty=T, method="gbm")
+end <- Sys.time()
+start-end
+start <- Sys.time()
+vals2 <- r_star_draws_df(temp_df, uncertainty=T, method="rf")
+end <- Sys.time()
+start-end
+start <- Sys.time()
+vals3 <- r_star_draws_df(temp_df, uncertainty=F, method="knn",
+                         hyperparameters = tibble(k=5))
+end <- Sys.time()
+start-end
 r_star(as_draws_array(temp_df))

### 2020-11-04 src/s_test_draws_df.R
added a plot using vanilla rhat for discrete case (to show to Aki)

@@ -79,7 +79,7 @@ test <- split_chains_draws_df(x)
 #' \emph{arXiv preprint} \code{arXiv:TBD}
 r_star_draws_df <- function(x, split_chains=T, uncertainty=F, method=NULL, hyperparameters=NULL,
                             training_percent=0.7, nsim=1000, ...){
-
+  x <- as_draws_df(x)
   if(split_chains)
     x <- split_chains_draws_df(x)
 
@@ -97,13 +97,12 @@ r_star_draws_df <- function(x, split_chains=T, uncertainty=F, method=NULL, hyper
   # choose hyperparameters
   if(is.null(method)) {
     method <- "rf"
+    caret_grid <- tibble(mtry=floor(sqrt(nparams)))
   } else if(is.null(hyperparameters) && method=="gbm") {
     caret_grid <- tibble(interaction.depth=c(3),
                          n.trees = 50,
                          shrinkage=c(0.1),
                          n.minobsinnode=10)
-  } else if(is.null(hyperparameters)) {
-    caret_grid <- tibble(mtry=floor(sqrt(nparams)))
   } else {
     caret_grid <- hyperparameters
   }
@@ -146,8 +145,20 @@ vals2 <- r_star_draws_df(temp_df, uncertainty=T, method="rf")
 end <- Sys.time()
 start-end
 start <- Sys.time()
-vals3 <- r_star_draws_df(temp_df, uncertainty=F, method="knn",
+vals3 <- r_star_draws_df(temp_df, uncertainty=T, method="knn",
                          hyperparameters = tibble(k=5))
 end <- Sys.time()
 start-end
 r_star(as_draws_array(temp_df))
+
+
+# build simple example
+x <- example_draws("eight_schools")
+r_star_draws_df(x)
+r_star_draws_df(x, split_chains = F)
+r_star_draws_df(x, method = "gbm", verbose=F)
+
+hist(r_star_draws_df(x, uncertainty = T))
+hist(r_star_draws_df(x, uncertainty = T, nsim = 100))
+
+

### 2020-11-06 src/s_discrete_aki.Rmd
Create s_discrete_aki.Rmd

---
title: "Discrete target distribution examples"
output: html_notebook
---

```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)
library(rstan)
library(latex2exp)
library(caret)
library(gbm)
library(MCMCpack)
options(mc.cores=4)
rstan_options(auto_write = TRUE)
source("monitornew.R")
source("r_star_monitor.R")
```

Create a discrete Markov chain
```{r}
f_simulate_markov_chain <- function(L, P, state_0){
  states <- vector(length=L)
  states[1] <- state_0
  for(i in 2:L){
    p  <- P[states[i - 1], ]
    states[i] <-  which(rmultinom(1, 1, p) == 1)
  }
  return(states)
}

P <- matrix(nrow = 4, ncol = 4)
P[1, ] <- c(0, 1/2, 1/2, 0)
P[2, ] <- c(1/2, 0, 1/3, 1/6)
P[3, ] <- c(1/4, 1/4, 1/4, 1/4)
P[4, ] <- c(0, 1, 0, 0)
```

Create 3 Markov chains with same transition matrix; one with a slightly different one
```{r}
f_generate_four <- function(L, P0, P1, state_0){
  x <- matrix(nrow = L, ncol = 4)
  for(i in 1:3)
    x[, i] <- f_simulate_markov_chain(L, P0, state_0)
  x[, 4] <- f_simulate_markov_chain(L, P1, state_0)
  return(x)
}

P0 <- P
P1 <- matrix(nrow = 4, ncol = 4)
P1[1, ] <- c(0, 1/2, 1/2, 0)
P1[2, ] <- c(1/2, 0, 1/3, 1/6)
P1[3, ] <- c(1, 0, 0, 0)
P1[4, ] <- c(1, 0, 0, 0)
```

Generate replicates for both zfs-Rhat and s-Rhat
```{r}
f_replicate <- function(L, P0, P1, state_0){
  temp <- f_generate_four(L, P0, P1, state_0)
  a_array <- array(dim=c(L, 4, 1))
  a_array[,,1] <- temp
  mon <- monitor_extra(split_data(a_array))
  return(list(r_hat_zfs=mon$zfsRhat, r_hat_s=mon$sRhat))
}

nreplicates <- 5
Ls <- c(1000, 2000, 5000, 10000)

# create P1 matrices
f_create_P1_example <- function(per_along){
  P1 <- matrix(nrow = 4, ncol = 4)
  P1[1, ] <- c(0, 1/2, 1/2, 0)
  P1[2, ] <- c(1/2, 0, 1/3, 1/6)
  v_diff <- c(1, 0, 0, 0) - c(1/4, 1/4, 1/4, 1/4)
  P1[3, ] <- c(1/4, 1/4, 1/4, 1/4) + v_diff * per_along
  v_diff <- c(1, 0, 0, 0) - c(0, 1, 0, 0)
  P1[4, ] <- c(0, 1, 0, 0) + v_diff * per_along
  return(P1)
}

last_vals <- c(0, 0.5, 1)
P1s <- map(last_vals, ~f_create_P1_example(.))

r_hat_vals_zfs <- matrix(nrow=nreplicates * length(Ls) * length(P1s), ncol=4)
r_hat_vals_s <- matrix(nrow=nreplicates * length(Ls) * length(P1s), ncol=4)
counter <- 1
for(k in 1:length(P1s)){
  for(j in seq_along(Ls)){
    print(paste0("k=", k, ", j=", j))
    for(i in 1:nreplicates){
      temp <- f_replicate(Ls[j], P0, P1s[[k]], 1)
      r_hat_vals_zfs[counter, ] <- c(k, Ls[j], i, temp$r_hat_zfs)
      r_hat_vals_s[counter, ] <- c(k, Ls[j], i, temp$r_hat_s)
      counter <- counter + 1
    }
  }
}
colnames(r_hat_vals_zfs) <- c("P1", "sample size", "iter", "value")
colnames(r_hat_vals_s) <- c("P1", "sample size", "iter", "value")
r_hat_vals_zfs <- as.data.frame(r_hat_vals_zfs) %>% 
  mutate(type="zfs-Rhat")
r_hat_vals_s <- as.data.frame(r_hat_vals_s) %>% 
  mutate(type="s-Rhat")

temp <- r_hat_vals_zfs %>% 
  bind_rows(r_hat_vals_s)
```

Plot
```{r}
r_star_vals <- temp %>% 
  mutate(P1=if_else(P1==1, "P1", if_else(P1==2, "P2", "P3"))) %>% 
  mutate(`sample size`=as.character(`sample size`)) %>% 
  mutate(`sample size`=if_else(`sample size`=="10000", "10,000", `sample size`)) %>% 
  mutate(`sample size`=as.factor(`sample size`)) %>% 
  mutate(`sample size`=fct_relevel(`sample size`, "1000", "2000", "5000", "10,000"))
a_sum <- r_star_vals %>%
  group_by(P1, `sample size`, type) %>%
  summarise(lower=quantile(value, 0.025),
         upper=quantile(value, 0.975),
         middle=quantile(value, 0.5))
r_star_vals %>% 
  left_join(a_sum) %>% 
  ggplot(aes(x=as.factor(`sample size`), y=value)) +
  geom_jitter(width = 0.2, colour="grey") +
  geom_pointrange(aes(ymin=lower, y=middle, ymax=upper)) +
  facet_grid(vars(P1), vars(type)) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_y_continuous(limits=c(NA, NA)) +
  xlab("Sample size") +
  ylab(TeX("$R*$")) +
  theme(axis.text = element_text(colour="black", size=16),
        axis.title = element_text(colour="black", size=16),
        strip.text.y = element_text(colour="black", size=16),
        legend.position = "none",
        title = element_text(colour="black", size=16))
```

### 2020-11-08 src/s_ar1.Rmd
started to make rhat changes

@@
   a_accuracy <- r_star(a_array, method = "gbm")
-  mon <- monitor_extra(split_data(a_array))
-  return(list(r_star=a_accuracy, r_hat=mon$zfsRhat))
+  mon <- monitor_extra(a_array)
+  rhat <- max(mon$zsRhat, mon$zfsRhat)
+  return(list(r_star=a_accuracy, r_hat=rhat))
 }
@@
         axis.text = element_text(colour="black", size=16)) +
-  xlim(1, NA)
+  xlim(1, 1.05)
 ```
@@
 nreplicates <- 50
-lnum_chains <- c(2, 4, 6, 8, 16, 32)
+lnum_chains <- c(2, 4, 8, 16, 32)
 mResults <- matrix(nrow = nreplicates * length(lnum_chains),

### 2020-11-08 src/s_bivariate_normal.Rmd
started to make rhat changes

@@
 
-# Use monitor package functionality -- split chains first
-half_iter <- max(chains_stacked$iter) / 2
+# Use monitor package functionality
+half_iter <- max(chains_stacked$iter)
 m_samples <- array(dim = c(half_iter,
-                           2*n_distinct(chains_stacked$chain),
+                           n_distinct(chains_stacked$chain),
                            2))
@@
   filter(iter<=half_iter)
-chains_stacked_second <- chains_stacked %>% 
-  filter(iter>half_iter)
 
@@
 
-for(i in 1:half_iter){
-  temp <- chains_stacked_second %>% 
-    filter(iter==(i + 1000))
-  for(j in 5:8){
-    temp1 <- temp %>%
-      filter(chain==(j - 4))
-    m_samples[i, j, ] <- as.numeric(temp1[1, 1:2])
-    k <- k + 1
-  }
-}
-
-mon <- monitor(m_samples)
+mon <- monitor_extra(m_samples)
+mon$zsRhat
+mon$zfsRhat
+max(mon$zsRhat, mon$zfsRhat)
 ```

### 2020-11-08 src/s_cauchy.Rmd
started to make rhat changes

@@
 
-mon <- monitor_extra(split_data(x))
-a_rhat_df <- tibble(nominal=mon$zfsRhat[1:50])
+mon <- monitor_extra(x)
+rhat <- mon$zfsRhat
+rhat1 <- mon$zsRhat
+rhat <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
+
+a_rhat_df <- tibble(nominal=rhat[1:50])
 ```
@@
 
-mon <- monitor_extra(split_data(full_data))
+mon <- monitor_extra(full_data)
+rhat <- mon$zfsRhat
+rhat1 <- mon$zsRhat
+rhat <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
 a_rhat_df <-a_rhat_df %>% 
-  mutate(alt=mon$zfsRhat[101:150],
+  mutate(alt=rhat[101:150],
          source="r_hat")
@@
 r_star_nom_rf <- r_star(full_data, method="rf", uncertainty = T)
-mon_nom <- monitor_extra(split_data(full_data))
+mon_nom <- monitor_extra(full_data)
+
+rhat <- mon_nom$zfsRhat
+rhat1 <- mon_nom$zsRhat
+rhat_nom <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
 ```
@@
 r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)
-mon_alt <- monitor_extra(split_data(full_data))
+mon_alt <- monitor_extra(full_data)
+rhat <- mon_alt$zfsRhat
+rhat1 <- mon_alt$zsRhat
+rhat_alt <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
 
@@
                source="r_star") %>% 
-  bind_rows(tibble(nominal=mon_nom$zfsRhat[1:50],
-                alt=mon_alt$zfsRhat[101:150],
+  bind_rows(tibble(nominal=rhat_nom[1:50],
+                alt=rhat_alt[101:150],
                 source="r_hat")) %>% 
@@
 
-my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.06) else c(0.9, 3)}
+my_lims <- function(x) { if (max(x) < 1.2) c(0.985, 1.04) else c(0.9, 3)}
 g <- c_df_rhat %>% 
@@
 fit <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=F)
 r_squared <- function(x, lower=0.001){
@@
 fit <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=F)
 r_vals <- map_dbl(iter, ~median(r_squared(x[1:., , ])))
@@
          variable="nom"))
-# saveRDS(b_df, "../output/cauchy_r_squared.rds")
-b_df <- readRDS("../output/cauchy_r_squared.rds")
+saveRDS(b_df, "../output/cauchy_r_squared.rds")
+# b_df <- readRDS("../output/cauchy_r_squared.rds")
 g1 <- b_df %>% 
@@
   ggtitle(TeX("A. Quantile-$R^2$"))
-```
-
-Cramer Von Mises -- seems very noisy
-```{r}
-library(goftest)
-f_cvm <- function(x){
-  temp <- cvm.test(x, null = "pcauchy")
-  return(temp$p.value)
-}
-
-f_cvm_all_dims <- function(x, max_iter){
-  vals <- map_dbl(seq(1, 50, 1), ~f_cvm(x[1:max_iter, , .]))
-  return(vals)
-}
-
-fit <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
-cvm_min <- map_dbl(iter, ~mean(f_cvm_all_dims(x, .)))
-
-fit <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
-cvm_min <- map_dbl(iter, ~mean(f_cvm_all_dims(x, .)))
-```
-
-Anderson-Darling test -- again, noisy
-```{r}
-library(goftest)
-f_ad <- function(x){
-  temp <- ad.test(x, null = "pcauchy")
-  return(temp$p.value)
-}
-
-f_ad_all_dims <- function(x, max_iter){
-  vals <- map_dbl(seq(1, 50, 1), ~f_ad(x[1:max_iter, , .]))
-  return(vals)
-}
-
-fit <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
-cvm_min <- map_dbl(iter, ~mean(f_ad_all_dims(x, .)<0.05))
-
-fit <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=T)
-cvm_min <- map_dbl(iter, ~mean(f_ad_all_dims(x, .)<0.05))
 ```
@@
 
-Progression of R* with ranked samples instead
-```{r}
-fit <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x")
-future::plan(multiprocess)
-unranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
-
-x <- z_scale(x)
-ranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
-cc_df <- tibble(ranked, unranked, iter, source="nominal")
-
-fit <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x")
-future::plan(multiprocess)
-unranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
-
-x <- z_scale(x)
-ranked <- future_map_dbl(iter, function(i) r_star(x[1:i, , ]))
-
-cc_df <- cc_df %>% 
-  bind_rows(tibble(ranked, unranked, iter, source="alt"))
-saveRDS(cc_df, "../output/cauchy_ranked_unranked.rds")
-cc_df <- readRDS("../output/cauchy_ranked_unranked.rds")
-
-g <- cc_df %>% 
-  mutate(source=if_else(source=="alt", "alternative", "nominal")) %>% 
-  melt(id.vars=c("iter", "source")) %>% 
-  mutate(variable=as.factor(variable)) %>% 
-  ggplot(aes(x=iter, y=value, colour=variable)) +
-  geom_line() +
-  facet_wrap(~source) +
-  scale_x_log10() +
-  scale_color_grey("Method") +
-  xlab("Draws") +
-  ylab("R*")
-ggsave("../output/cauchy_ranked_unranked.pdf", g)
-```
-
 Rhat
@@
  nparams <- dim(x)[3] 
- rhat_vals <- map_dbl(seq(1, nparams, 1), ~rhat(split_chains(z_scale(x[ , , .]))))
+ rhat_vals <- map_dbl(seq(1, nparams, 1), ~posterior::rhat((x[ , , .])))
  return(rhat_vals)
@@
 fit <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit, permuted=F, pars="x")
+x <- rstan::extract(fit, permuted=F, pars="x", inc_warmup=F)
 future::plan(multiprocess)
-iter <- a_df$iteration %>% unique()
 r_hat <- future_map_dbl(iter, ~max(r_hat_all_params(x[1:., , ])))
@@
 fit1 <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit1, permuted=F, pars="x")
+x <- rstan::extract(fit1, permuted=F, pars="x", inc_warmup=F)
 r_hat1 <- future_map_dbl(iter, ~max(r_hat_all_params(x[1:., , ])))
@@
 fit1 <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit1, permuted=F, pars="x")
+x <- rstan::extract(fit1, permuted=F, pars="x", inc_warmup=F)
 Rhat_alt <- map_dbl(iter, ~r_hat_multivariate(x[1:., , ]))
@@
 fit1 <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit1, permuted=F, pars="x")
+x <- rstan::extract(fit1, permuted=F, pars="x", inc_warmup=F)
 Rhat_nom <- map_dbl(iter, ~r_hat_multivariate(x[1:., , ]))
@@
        nom=Rhat_nom) 
+saveRDS(g_df, "../output/cauchy_mvt_rhat.rds")
 g4 <- 
@@
 fit1 <- readRDS("../output/fit_alt_50000.rds")
-x <- rstan::extract(fit1, permuted=F, pars="x")
+x <- rstan::extract(fit1, permuted=F, pars="x", inc_warmup=F)
 tail_ess_alt <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_tail(x[1:i, , .]))))
@@
 fit1 <- readRDS("../output/fit_nominal_50000.rds")
-x <- rstan::extract(fit1, permuted=F, pars="x")
+x <- rstan::extract(fit1, permuted=F, pars="x", inc_warmup=F)
 tail_ess_nom <- map_dbl(iter, function(i) min(map_dbl(seq(1, 50, 1), ~ess_tail(x[1:i, , .]))))

### 2020-11-08 src/s_eight_schools.Rmd
started to make rhat changes

@@
 r_star_centered <- r_star(x, uncertainty = T)
-mon_cp <- monitor_extra(split_data(x))
+mon_cp <- monitor_extra(x)
 
@@
 r_star_noncentered <- r_star(x, uncertainty = T)
-mon_ncp <- monitor_extra(split_data(x))
+mon_ncp <- monitor_extra(x)
 
@@
                simulations="split")
+
 c1_df <- tibble(centered=mon_cp$zfsRhat[1:10],

### 2020-11-08 src/s_mvt_normal.Rmd
started to make rhat changes

@@
 full_data <- rstan::extract(fit, permuted=F)
-a_sum <- monitor_extra(split_data(full_data))
+a_sum <- monitor_extra(full_data)
 rhat <- a_sum$zfsRhat
+rhat1 <- a_sum$zsRhat
+rhat <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
 
@@
         axis.text = element_text(colour="black"))
-b_df <- tibble(samples_10000_cp=a_sum$zfsRhat[1:250])
+b_df <- tibble(samples_10000_cp=rhat[1:250])
 ```
@@
          samples_400_rf=r_star2_rf)
-
-a_sum <- monitor_extra(split_data(full_data))
+a_sum <- monitor_extra(full_data)
+rhat <- a_sum$zfsRhat
+rhat1 <- a_sum$zsRhat
+rhat <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
 b_df <- b_df %>%
-  mutate(samples_400=a_sum$zfsRhat[1:250])
+  mutate(samples_400=rhat[1:250])
 ```
@@
 full_data <- rstan::extract(fit, permuted=F)
-a_sum <- monitor_extra(split_data(full_data))
+a_sum <- monitor_extra(full_data)
 sum(a_sum$zsseff>400)
@@
 sum(a_sum$zfsRhat<1.01)
+sum(a_sum$zsRhat<1.01)
+rhat <- a_sum$zfsRhat
+rhat1 <- a_sum$zsRhat
+rhat <- tibble(rhat, rhat1) %>% 
+  apply(1, max)
 b_df <- b_df %>%
-  mutate(samples_10000_ncp=a_sum$zfsRhat[251:500])
+  mutate(samples_10000_ncp=rhat[251:500])
 saveRDS(b_df, "../output/mvt_fit_all_rhat.rds")

### 2020-11-08 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
started to make rhat changes

@@
 \ref{fig:mvt_gbm_vs_rf}).} From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ \textcolor{red}{for all parameters} (Fig.
 \ref{fig:mvt}B), although 54\% had bulk-$\text{ESS}<400$ and 13\% of parameters had tail-$\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
@@
 \ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-In this plot, there is a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$).
+In this plot, there is a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.17, S=2185680, p<0.01$).
 In Fig.

### 2020-11-09 src/s_ar1.Rmd
cleaning up repo and rerunning rhat experiments

@@
   ggtitle("A.")
-```
-
-Examine probability calibration
-```{r}
-predict(object=gbmFit1, newdata=testing_data, type = "prob") %>% 
-  select(`4`) %>% 
-  rename(prob_4=`4`) %>% 
-  mutate(actual=testing_data$chain) %>% 
-  mutate(prob_4_cat=cut(prob_4, seq(0, 1, 0.05), ordered_result = T)) %>% 
-  group_by(prob_4_cat) %>% 
-  summarise(prob_4_actual=mean(actual==4)) %>% 
-  ggplot(aes(x=prob_4_cat, y=prob_4_actual)) +
-  geom_point()
 ```

### 2020-11-09 src/s_cauchy.Rmd
cleaning up repo and rerunning rhat experiments

@@
 mon <- monitor_extra(x)
+
+# calculate rhat as in posterior package
 rhat <- mon$zfsRhat
@@
 
+# calculate rhat as in posterior package
 mon <- monitor_extra(full_data)
@@
 
+# calculate R*
 r_star_nom_gbm <- r_star(full_data, method="gbm", uncertainty = T)
 r_star_nom_rf <- r_star(full_data, method="rf", uncertainty = T)
+
+# calculate rhat as in posterior package
 mon_nom <- monitor_extra(full_data)
-
 rhat <- mon_nom$zfsRhat
@@
 
+# calculate R*
 r_star_alt_gbm <- r_star(full_data, method="gbm", uncertainty = T)
 r_star_alt_rf <- r_star(full_data, method="rf", uncertainty = T)
+
+# calculate rhat as in posterior package
 mon_alt <- monitor_extra(full_data)

### 2020-11-09 src/s_discrete.Rmd
cleaning up repo and rerunning rhat experiments

@@
   a_array[,,1] <- temp
-  a_accuracy <- r_star(a_array)
-  mon <- monitor_extra(split_data(a_array))
-  return(list(r_star=a_accuracy, r_hat=mon$zfsRhat))
+  a_accuracy <- r_star(a_array, method = "gbm")
+  mon <- monitor_extra(a_array)
+  rhat <- max(mon$zsRhat, mon$zfsRhat)
+  return(list(r_star=a_accuracy, r_hat=rhat))
 }
@@
 
-saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/discrete_r_star_hat.rds")
+saveRDS(list(r_star=r_star_vals, r_hat=r_hat_vals), "../data/discrete_r_star_hat.rds")
 ```
@@
 
-saveRDS(list(P0=P0, P1=P1), "../data/discrete_higherd_matrices.rds")
+# saveRDS(list(P0=P0, P1=P1), "../data/discrete_higherd_matrices.rds")
 temp <- readRDS("../data/discrete_higherd_matrices.rds")
@@
         legend.text = element_text(colour="black", size=16))
-ggsave("../output/discrete_stationary_higherd.pdf", g, width = 10, height = 6)
+# ggsave("../output/discrete_stationary_higherd.pdf", g, width = 10, height = 6)
 ```
@@
 
-saveRDS(tibble(r_star=r_star_vals, r_hat=r_hat_vals), "../data/discrete_r_star_hat_higherd.rds")
+saveRDS(list(r_star=r_star_vals, r_hat=r_hat_vals), "../data/discrete_r_star_hat_higherd.rds")
 

### 2020-11-09 src/s_eight_schools.Rmd
cleaning up repo and rerunning rhat experiments

--- 2020-10-23 src/s_eight_schools.Rmd
@@
 ---
-title: "R Notebook"
+title: "8 schools"
 output: html_notebook
@@
 x <- rstan::extract(fit_cp, permuted=F)
-r_star_centered <- r_star(x, uncertainty = T)
-mon_cp <- monitor_extra(split_data(x))
+r_star_centered <- r_star(x, uncertainty = T, method = "gbm")
+mon_cp <- monitor_extra(x)
 
@@
 x <- rstan::extract(fit_ncp, permuted=F)
-r_star_noncentered <- r_star(x, uncertainty = T)
-mon_ncp <- monitor_extra(split_data(x))
+r_star_noncentered <- r_star(x, uncertainty = T, method = "gbm")
+mon_ncp <- monitor_extra(x)
 
@@
                simulations="split")
-c1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
-                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
+
+# calculate Rhat as max of zfs and zs Rhat
+f_rhat_max <- function(mon, ids) {
+  rhat <- mon$zfsRhat[ids]
+  rhat1 <- mon$zsRhat[ids]
+  rhat <- tibble(rhat, rhat1) %>% 
+    apply(1, max)
+  return(rhat)
+}
+
+c1_df <- tibble(centered=f_rhat_max(mon_cp, 1:10),
+                non_centered=f_rhat_max(mon_ncp, c(1, 2, 11:18)),
                 metric="split-Rhat",
@@
 r_star_centered <- r_star(x, split_chains = F, uncertainty = T, method="gbm")
-mon_cp <- monitor_extra(fit_cp)
+
+# do calculations using unsplit chains
+rhat_cp <- vector(length = 10)
+for(i in 1:10) {
+  sims_i <- x[, , i]
+  zsims_split <- z_scale(sims_i)
+  zsplit_rhat <- rhat_rfun(zsims_split)
+  sims_centered <- sims_i - median(sims_i)
+  sims_folded <- abs(sims_centered)
+  zsims_folded_split <- z_scale(sims_folded)
+  zfsplit_rhat <- rhat_rfun(zsims_folded_split)
+  rhat_cp[i] <- max(zsplit_rhat, zfsplit_rhat)
+}
+
 # noncentered
@@
 r_star_noncentered <- r_star(x, split_chains = F, uncertainty = T, method="gbm")
-mon_ncp <- monitor_extra(fit_ncp)
+
+ids <- c(1, 2, 11:18)
+rhat_ncp <- vector(length = 10)
+k <- 1
+for(i in ids) {
+  sims_i <- x[, , i]
+  zsims_split <- z_scale(sims_i)
+  zsplit_rhat <- rhat_rfun(zsims_split)
+  sims_centered <- sims_i - median(sims_i)
+  sims_folded <- abs(sims_centered)
+  zsims_folded_split <- z_scale(sims_folded)
+  zfsplit_rhat <- rhat_rfun(zsims_folded_split)
+  rhat_ncp[k] <- max(zsplit_rhat, zfsplit_rhat)
+  k <- k + 1
+}
 
@@
                simulations="unsplit")
-d1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
-                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
+d1_df <- tibble(centered=rhat_cp,
+                non_centered=rhat_ncp,
                 metric="split-Rhat",
@@
   geom_histogram(position="identity", alpha=0.8) +
-  xlim(0.78, 2) +
+  xlim(0.78, 1.7) +
   scale_fill_grey("Series", labels=c("centered", "non-centered")) +
@@
   scale_fill_grey("Series", labels=c("centered", "non-centered")) +
-  xlim(0.78, 2) +
+  xlim(0.78, 1.7) +
   xlab(TeX("$R*$")) +
@@
   geom_point() +
-  scale_x_continuous(limits=c(NA, 1.035)) +
-  scale_y_continuous(limits=c(NA, 1.035)) +
+  scale_x_continuous(limits=c(NA, 1.02)) +
+  scale_y_continuous(limits=c(NA, 1.02)) +
   geom_vline(xintercept = 1.01, linetype=2) +
@@
 r_star_noncentered <- r_star(x, uncertainty = T, method="rf")
-mon_ncp <- monitor_extra(split_data(x))
 
@@
                metric="R*",
-               simulations="split")
-c1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
-                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
-                metric="split-Rhat",
                simulations="split")
@@
 r_star_noncentered <- r_star(x, split_chains = F, uncertainty = T, method="rf")
-mon_ncp <- monitor_extra(fit_ncp)
 
@@
                metric="R*",
-               simulations="unsplit")
-d1_df <- tibble(centered=mon_cp$zfsRhat[1:10],
-                non_centered=mon_ncp$zfsRhat[c(1, 2, 11:18)],
-                metric="split-Rhat",
                simulations="unsplit")

### 2020-11-09 src/s_gbm_vs_rf.Rmd
cleaning up repo and rerunning rhat experiments

--- 2020-10-15 src/s_gbm_vs_rf.Rmd
@@
 
-Generate a large correlation matrix
+# Different covariance matrices
 ```{r}
@@
 
+# Run replicates with one chain with a different cov matrix
 nreplicates <- 20
@@
 r_star_opts <- tibble(d=ds, optimal=r_star_opts)
+
+# plot results
 temp_df <- 
@@
 
-Same but increase fatness of tails
+# Tail fatness
 ```{r}
@@
 
+# Run replicates
 nreplicates <- 20
@@
 
+# calculate optimal classification rates
 f_optimal_studentts <- function(niter, nu1, nu2, d, A0){
@@
 
-r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(1000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))
+r_star_opts1 <- map_dbl(seq(1, nrow(m_parameters), 1), ~f_optimal_studentts(10000, nu1, m_parameters$nu2[.], m_parameters$ds[.], A0))
 
+# plot results
 r_star_opts <- tibble(d=m_parameters$ds, nu2=m_parameters$nu2, optimal=r_star_opts1)
+big_df <- readRDS("../data/gbm_vs_rf_studentts_results.rds")
 temp_df <- 

### 2020-11-09 src/s_hyperparameters.Rmd
cleaning up repo and rerunning rhat experiments

@@
 ---
+
+Note that the results used in this file were generated using various shell scripts: s_jobScheduler_hypers.sh (for the GBM classifier) and s_jobScheduler_hypers_rf.sh (for the RF classifier)
 
@@
 ```{r}
-f_ar1 <- function(rho, sigma, L){
-  x <- vector(length = L)
-  x[1] <- rnorm(1, 0, sd=sigma)
-  for(i in 2:L)
-    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
-  return(x)
-}
-
-# Generates three chains with same var; one with a different var
-f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:3)
-    x[, i] <- f_ar1(rho, sigma, L)
-  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
-  x[, 4] <- z
-  return(x)
-}
-```
-
-```{r}
-f_data_ar1 <- function(){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
-  a_array <- array(dim=c(1000, 4, 1))
-  a_array[,,1] <- temp
-  return(a_array)
-}
-
-f_replicate_hyper <- function(m_parameters, f_data_generator){
-
-  a_array <- f_data_generator()
-
-  laccuracy <- vector(length = nrow(m_parameters))
-  for(i in 1:nrow(m_parameters))
-    laccuracy[i] <- r_star(a_array,
-                           caret_default=expand.grid(interaction.depth=m_parameters$int.depth[i],
-                                                     n.trees = m_parameters$n.trees[i],
-                                                     shrinkage=0.1,
-                                                     n.minobsinnode=10))
-  return(laccuracy)
-}
-
-f_replicate_r_star <- function(nreplicates, m_parameters, f_data_generator){
-  r_star_vals <- matrix(nrow = nreplicates, ncol=nrow(m_parameters))
-  for(i in 1:nreplicates){
-      print(i)
-
-    temp <- f_replicate_hyper(m_parameters, f_data_generator)
-    r_star_vals[i, ] <- temp
-  }
-  r_star_vals <- r_star_vals %>% 
-    t() %>% 
-    as.data.frame()
-  r_star_vals1 <- m_parameters %>% cbind(r_star_vals)
-  return(r_star_vals1)
-}
-
-# m_parameters <- expand_grid(int.depth=seq(1, 7, 2),
-#                             n.trees=c(1, 10, 50, 100))
-# r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_ar1)
-
-# saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")
 g1 <- readRDS("../data/hypers_ar1.rds")$r_star %>% 
@@
 ```{r}
-# source("eight_schools.data.R")
-# eight_schools <- list(J=J, y=y, sigma=sigma)
-# model_cp <- stan_model("eight_schools_cp.stan")
-# model_ncp <- stan_model("eight_schools_ncp.stan")
-# 
-# f_data_8_schools <- function(){
-#   fit_ncp <- sampling(
-#     model_ncp, data = eight_schools,
-#     iter = 2000, chains = 4, refresh = 0,
-#     control = list(adapt_delta = 0.95)
-#     )
-#   x_ncp <- rstan::extract(fit_ncp, permuted=F)
-# }
-# 
-# m_parameters <- expand_grid(int.depth=c(1, 3, 5),
-#                             n.trees=c(1, 50, 200))
-# r_star_vals <- f_replicate_r_star(20, m_parameters, f_data_ar1)
-
 g1a <- readRDS("../data/hypers_8_schools.rds")$r_star %>% 
@@
 ```{r}
-f_data_cauchy <- function(){
-  fit_nom <- stan(file = 'cauchy_alt_1.stan',
-                refresh = 0)
-  x <- rstan::extract(fit_nom, permuted=F)
-  return(x)
-}
-
-# m_parameters <- expand_grid(int.depth=c(3, 7, 11),
-#                             n.trees=c(50, 100, 400))
-# r_star_vals <- f_replicate_r_star(2, m_parameters, f_data_cauchy)
-
 g1b <- readRDS("../data/hypers_cauchy.rds")$r_star %>% 

### 2020-11-09 src/s_ml.Rmd
cleaning up repo and rerunning rhat experiments

@@
 ---
+
+Note that, to generate the results in this section, we used the shell file 
+s_jobScheduler_ml_hypers_set.sh
 
@@
 
-Function to fit any ML model in Caret
-```{r}
-r_star_ml_variety <- function(x, method_, caretGrid){
-
-  split_chains=T
-  training_percent=0.7
-  if(split_chains)
-    x <- split_data(x)
-  nparams <- dim(x)[3]
-  nchains <- dim(x)[2]
-  niter <- dim(x)[1]
-  m_flattened <- matrix(nrow = nrow(x) * nchains,
-                        ncol = (nparams + 1))
-  k <- 1
-  for(i in 1:nchains){
-    for(j in 1:nrow(x)){
-      m_flattened[k, 1:nparams] <- x[j, i, ]
-      m_flattened[k, (nparams + 1)] <- i
-      k <- k + 1
-    }
-  }
-  m_flattened <- m_flattened %>% 
-    as.data.frame()
-  colnames(m_flattened)[nparams + 1] <- "chain"
-  r <- m_flattened %>% 
-    mutate(chain=as.factor(chain))
-  # if only 1 param, add in a column of random noise since gbm requires >1 dims
-  if(nparams==1)
-    r <- r %>% 
-      mutate(V_new=rnorm(nrow(r)))
-
-  rand_samples <- sample(1:nrow(r), training_percent * nrow(r))
-  training_data <- r[rand_samples, ]
-  testing_data <- r[-rand_samples, ]
-
-  if(method_!="multinom")
-      gbmFit <- train(chain ~ ., data = training_data, 
-                   trControl = trainControl(method = 'none'),
-                   method=method_,
-                   tuneGrid = caretGrid)
-  else
-    gbmFit <- train(chain ~ ., data = training_data, 
-                 trControl = trainControl(method = 'none'),
-                 method=method_,
-                 tuneGrid = caretGrid,
-                 MaxNWts=10000
-                 )
-  plda <- predict(object=gbmFit, newdata=testing_data)
-  a_accuracy <- 
-    tibble(predicted=plda, actual=testing_data$chain) %>%
-    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
-    summarise(mean(correct)) %>% 
-    pull()
-  return(a_accuracy * n_distinct(training_data$chain))
-}
-
-f_replicate_gridded <- function(x, method_, caretGrids_){
-  r_stars <- vector(length = nrow(caretGrids_))
-  times <- vector(length = nrow(caretGrids_))
-  for(i in 1:nrow(caretGrids_)){
-    start.time <- Sys.time()
-    r_stars[i] <- r_star_ml_variety(x, method_, caretGrids_[i, ])
-    end.time <- Sys.time()
-    times[i] <- end.time - start.time
-  }
-  return(list(r_star=r_stars, time=times))
-}
-
-f_run_all <- function(methods, list_of_caret_grids, f_data_generator){
-  x <- f_data_generator()
-  r_stars <- vector(length = length(methods))
-  times <- vector(length = length(methods))
-  hypers <- vector(length = length(methods), mode="list")
-  for(i in seq_along(r_stars)){
-    print(methods[i])
-    a_grid <- list_of_caret_grids[[i]]
-    a <- f_replicate_gridded(x, methods[i], a_grid)
-    r_stars[i] <- max(a$r_star)
-    a_ind <- which.max(a$r_star)
-    times[i] <- a$time[a_ind]
-    hypers[[i]] <- a_grid[a_ind, ]
-  }
-  return(tibble(r_star=r_stars, method=methods, time=times, hyper=hypers))
-}
-
-f_run_all_replicates <- function(niterations, methods, list_of_caret_grids, f_data_generator){
-  for(i in 1:niterations){
-    print(i)
-    temp_df <- f_run_all(methods, list_of_caret_grids, f_data_generator) %>% mutate(iter=i)
-    if(i==1)
-      big_df <- temp_df
-    else
-      big_df <- big_df %>% bind_rows(temp_df)
-  }
-  return(big_df)
-}
-
-tunegrid_gbm <- expand.grid(interaction.depth=1, 
-                             n.trees = c(400),
-                             shrinkage=c(0.1),
-                             n.minobsinnode=c(1))
-tunegrid_rf <- tibble(mtry = c(1, 5, 15))
-tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
-tunegrid_svm <- tibble(C = c(0.25, 0.5, 0.75))
-tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
-tunegrid_xgbtree <- expand_grid(nrounds = c(1, 10),
-                       max_depth = c(1, 4),
-                       eta = c(.1, .4),
-                       gamma = 0,
-                       colsample_bytree = .7,
-                       min_child_weight = 1,
-                       subsample = c(.8, 1))
-```
-
 # AR1
 ```{r}
-f_ar1 <- function(rho, sigma, L){
-  x <- vector(length = L)
-  x[1] <- rnorm(1, 0, sd=sigma)
-  for(i in 2:L)
-    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
-  return(x)
-}
-
-# Generates three chains with same var; one with a different var
-f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:3)
-    x[, i] <- f_ar1(rho, sigma, L)
-  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
-  x[, 4] <- z
-  return(x)
-}
-
-f_data_ar1 <- function(){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
-  a_array <- array(dim=c(1000, 4, 1))
-  a_array[,,1] <- temp
-  return(a_array)
-}
-
-# temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
-#           list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
-#           f_data_ar1)
-
 orders <- c("rf", "xgbTree", "gbm", "knn", "multinom", "svmLinear")
@@
 ```{r}
-source("eight_schools.data.R")
-eight_schools <- list(J=J, y=y, sigma=sigma)
-model_cp <- stan_model("eight_schools_cp.stan")
-model_ncp <- stan_model("eight_schools_ncp.stan")
-
-f_data_8_schools <- function(){
-  fit_ncp <- sampling(
-    model_ncp, data = eight_schools,
-    iter = 2000, chains = 4, refresh = 0,
-    control = list(adapt_delta = 0.95)
-    )
-  x_ncp <- rstan::extract(fit_ncp, permuted=F)
-}
-
-# temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
-#           list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
-#           f_data_8_schools)
-
 temp_df1 <- readRDS("../data/ml_comp_hypers_set_8_schools.rds")  %>% 
@@
 ```{r}
-f_data_cauchy <- function(){
-  fit_nom <- stan(file = 'cauchy_alt_1.stan',
-                refresh = 0)
-  x <- rstan::extract(fit_nom, permuted=F)
-  return(x)
-}
-
-# temp_df <- f_run_all_replicates(2, c("gbm", "rf", "knn", "svmLinear", "multinom"),
-#           list(tunegrid_gbm, tunegrid_rf, tunegrid_knn, tunegrid_svm, tunegrid_multinom),
-#           f_data_cauchy)
 temp_df2 <- readRDS("../data/ml_comp_hypers_set_cauchy.rds") %>% 
@@
 ```{r}
-f_data_normal <- function(){
-  N <- 250
-  A <- rWishart(1, 250, diag(N))[,,1]
-  model <- stan_model("mvt_250_ncp.stan")
-  fit <- sampling(model, data=list(N=N, A=A), iter=1000, chains=4)
-  x <- rstan::extract(fit, permuted=F)
-  return(x)
-}
-
-# temp_df <- f_replicate_gridded(x, "rf", tunegrid_rf)
-
 temp_df3 <- readRDS("../data/ml_comp_hypers_set_normal_250.rds") %>% 
@@
 ```
-
-# Check for false positives with random forests
-```{r}
-f_data_converged <- function(){
-  d <- 10
-  n <- 2000
-  nchain <- 4
-
-  x <- rnorm(d * n * nchain)
-  x <- array(x, dim=c(n, nchain, d))
-  return(x)
-}
-
-temp_df <- f_run_all_replicates(10, c("gbm", "rf"),
-          list(tunegrid_gbm, tunegrid_rf),
-          f_data_converged)
-temp_df %>% 
-  ggplot(aes(x=as.factor(method), y=r_star)) +
-  geom_boxplot()
-```
-
-# Setting hyperparameters for random forest:
-"Influence of Hyperparameters on Random Forest Accuracy"
-Simon Bernard, Laurent Heutte, Sebastien Adam, 2009
-Use mtry=sqrt(number features)
-
-whereas for gbms / xgboost, performance depends on lots of things

### 2020-11-09 src/s_mvt_normal.Rmd
cleaning up repo and rerunning rhat experiments

@@
 
-# Collect Rhat
+# Collect Rhat and calculate as in posterior package
 full_data <- rstan::extract(fit, permuted=F)

### 2020-11-09 src/s_ovarian_prostate.Rmd
cleaning up repo and rerunning rhat experiments

@@
 
-mon <- monitor_extra(split_data(x))
+mon <- monitor_extra(x)
 saveRDS(mon, "../data/ovarian_long_mon.rds")
 full_data <- rstan::extract(fit, permuted=F)
-r_star <- r_star_dist(full_data)
+r_star <- r_star(full_data, method = "gbm", uncertainty = T)
 saveRDS(r_star, "../data/ovarian_long_r_star.rds")
@@
 x <- rstan::extract(fit, permuted=F)
-mon <- monitor_extra(split_data(x))
+mon <- monitor_extra(x)
 saveRDS(mon, "../data/ovarian_short_mon.rds")
 full_data <- rstan::extract(fit, permuted=F)
-r_star <- r_star_dist(full_data)
+r_star <- r_star(full_data, method = "gbm", uncertainty = T)
 saveRDS(r_star, "../data/ovarian_short_r_star.rds")
@@
 
+# find max of rank normalized split-Rhat and rank normalized folded-split-Rhat
+f_rhat_max <- function(mon) {
+  rhat <- mon$zsRhat
+  rhat1 <- mon$zfsRhat
+  rhat <- tibble(rhat, rhat1) %>% 
+    apply(1, max)
+  return(rhat)
+}
+
 g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
-  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. Split-Rhat")) %>% 
+  bind_rows(tibble(long=f_rhat_max(mon_longer), short=f_rhat_max(mon_shorter), metric="B. Split-Rhat"))%>% 
   bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk-ESS")) %>% 
@@
 
-mon <- monitor_extra(split_data(x))
+mon <- monitor_extra(x)
 saveRDS(mon, "../data/prostate_long_mon.rds")
@@
 # Thin data again
-r_star <- r_star_dist(full_data[, , seq(1, 18106, 5)])
+r_star <- r_star(full_data[, , seq(1, 18106, 5)], method = "gbm", uncertainty = T)
 saveRDS(r_star, "../data/prostate_long_r_star.rds")
@@
 x <- rstan::extract(fit, permuted=F)
-mon <- monitor_extra(split_data(x))
+mon <- monitor_extra(x)
 saveRDS(mon, "../data/prostate_short_mon.rds")
 full_data <- rstan::extract(fit, permuted=F)
-r_star <- r_star_dist(full_data[, , seq(1, 18106, 5)])
+r_star <- r_star(full_data[, , seq(1, 18106, 5)], method = "gbm", uncertainty = T)
 saveRDS(r_star, "../data/prostate_short_r_star.rds")
@@
 g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
-  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. Split-Rhat")) %>% 
+  bind_rows(tibble(long=f_rhat_max(mon_longer), short=f_rhat_max(mon_shorter), metric="B. Split-Rhat")) %>% 
   bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk-ESS")) %>% 

### 2020-11-09 src/s_trends.Rmd
cleaning up repo and rerunning rhat experiments

@@
 ---
-title: "Examining the effect of a trend on R*"
+title: "Examining the effect of sampler trends on R*"
 output: html_notebook
@@
 
+Plot
 ```{r}
@@
   as.data.frame() %>%
-  select(iters, rho, rep, zsRhat, Rstar_split, method) %>%
+  dplyr::select(iters, rho, rep, zsRhat, zfsRhat, Rstar_split, method) %>%
   unique() %>%
-  select(-rep) %>%
+  dplyr::select(-rep) %>%
+  mutate(zsRhat=max(as.numeric(as.character(zsRhat), as.numeric(as.character(zfsRhat))))) %>% 
+  dplyr::select(-zfsRhat) %>% 
   melt(c("rho", "iters", "method")) %>%
@@
   mutate(method=if_else(method=="gbm", "GBM", "RF"))
-
-# ggplot(data = res, aes(x = rho, y = zsRhat)) +
-#   geom_jitter() +
-#   facet_grid(vars(iters)) +
-#   geom_hline(yintercept = 1) +
-#   ggtitle('R* without splitting') +
-#   ylim(0, 6)
 

### 2020-11-09 src/s_wide_distributions.Rmd
cleaning up repo and rerunning rhat experiments

@@
 full_data <- rstan::extract(fit, permuted=F, inc_warmup=F)
+
+## R* calcs
 r_star_gbm <- r_star(full_data, uncertainty = T, method="gbm")
@@
 
-mon <- monitor_extra(split_data(full_data))
-mean(mon$zfsRhat>1.01)
+## Rhat cals and ESSs
+mon <- monitor_extra(full_data)
+
+f_rhat_max <- function(mon) {
+  rhat <- mon$zfsRhat
+  rhat1 <- mon$zsRhat
+  rhat <- tibble(rhat, rhat1) %>% 
+    apply(1, max)
+  return(rhat)
+}
+
+rhats <- f_rhat_max(mon)
+mean(rhats>1.01)
 mean(mon$zsseff<400)
@@
 
-b_df <- tibble(samples_cp=mon$zfsRhat[1:250])
-
+## plot traditional
 ha <- tibble(bulk_ess=mon$zsseff,
        tail_ess=mon$tailseff,
-       rhat=mon$zfsRhat) %>% 
+       rhat=rhats) %>% 
   melt() %>% 
@@
        tail_ess=mon$tailseff,
-       rhat=mon$zfsRhat,
+       rhat=rhats,
        source="Centered")
@@
 
-mon <- monitor_extra(split_data(full_data))
-mean(mon$zfsRhat>1.01)
+mon <- monitor_extra(full_data)
+rhats <- f_rhat_max(mon)
+mean(rhats>1.01)
 mean(mon$zsseff<400)
@@
        tail_ess=mon$tailseff,
-       rhat=mon$zfsRhat) %>% 
+       rhat=rhats) %>% 
   melt() %>% 
@@
 
+## plot all traditional measures
 aa_df <- aa_df %>%
@@
        tail_ess=mon$tailseff,
-       rhat=mon$zfsRhat,
+       rhat=rhats,
        source="Non-centered"))
@@
 saveRDS(a_df, "../output/mvt_wide_400_summary.rds")
-mon <- monitor_extra(split_data(full_data))
-mean(mon$zfsRhat>1.01)
+mon <- monitor_extra(full_data)
+rhats <- f_rhat_max(mon)
+mean(rhats>1.01)
 mean(mon$zsseff<400)
@@
 saveRDS(a_df, "../output/mvt_wide_summaries.rds")
-mon <- monitor_extra(split_data(full_data))
-mean(mon$zfsRhat>1.01)
+mon <- monitor_extra(full_data)
+rhats <- f_rhat_max(mon)
+mean(rhats>1.01)
 mean(mon$zsseff<400)

### 2020-11-09 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
cleaning up repo and rerunning rhat experiments

@@
 \ref{fig:eight_schools}C, we plot $\widehat{R}$ values obtained when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all but a single parameter, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
-All parameters for the non-centered models were below 1.01, indicating convergence.
+\textcolor{red}{These show that the values of $\widehat{R}$ for the centered model using the unsplit chains were below 1.01; when using the chains split into two halves, $\widehat{R}>1.01$ for all but a single parameter.
+All parameters for the non-centered models were below 1.01, indicating convergence.}
 

### 2020-11-09 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
cleaning up repo and rerunning rhat experiments

@@
 \ref{fig:mvt_wide_both}C\&D, we plot the $R^*$ distributions from GBM and RF classifiers for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.} In all cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.
+These results were also echoed by rank-normalised split-$\widehat{R}$, with \textcolor{red}{19\%} of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and \textcolor{red}{3\%} for the 1000 iteration case.
 
@@
 Whilst it is harder to discern, this pattern is mirrored in Fig.
-\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 139 parameters where this was not the case.
+\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had \textcolor{red}{83} parameters where this was not the case.
 Similarly, in Figs.
@@
 At each combination of transition probability matrix (for the fourth chain) and sample size, we perform 40 replicates.
-In each replicate, we calculate both $R^*$ and rank-normalised split-$\widehat{R}$.
+In each replicate, we calculate both $R^*$ using a GBM classifier and rank-normalised split-$\widehat{R}$.
 The results of these experiments are shown in Fig.
@@
 For $R^*$, the replicates were centred on $R^*=1$, although the stochasticity of replicates meant that, in some cases, $R^*>1$.
-Whereas, $\widehat{R}<1$ in all cases.
+Whereas, $\widehat{R}$ was similarly close to 1 in all cases.
 Considering the simulations using $\boldsymbol{P}_2$ for the fourth chain transition probability matrix, $R^*>1$ across a majority of replicates, and this majority increased with sample size.
-In this case, $\widehat{R}<1$ for all replicates across the various sample sizes.
-For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; as for the previous case, $\widehat{R}<1$, indicating convergence.
+In this case, $1.01>\widehat{R}>1$ for all replicates across the various sample sizes.
+For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; in this case, $\widehat{R}>1.01$ in all replicates.
 
-This example hints that $R^*$ may be particularly adept at detecting poor convergence in small discrete parameter spaces.
+This example shows that $R^*$ can detect poor convergence for discrete parameter spaces, like $\widehat{R}$.
 
@@
 
-Collectively, the results of \S\ref{sec:discrete_small} and \S\ref{sec:discrete_large} hint that $R^*$ may be more reliable at detecting poor convergence than $\widehat{R}$, although further empirical investigation is needed.
+Collectively, the results of \S\ref{sec:discrete_small} and \S\ref{sec:discrete_large} show that $R^*$ is able to detect convergence issues in discrete target distributions.
+The results of \S\ref{sec:discrete_large} hint that $R^*$ may be superior than $\widehat{R}$ in larger discrete spaces, although further empirical investigation is needed.
 

### 2020-11-09 src/s_jobScheduler_hypers_rf.sh

@@ -8,5 +8,4 @@ chmod +x *.sh
 # Rscript s_run_hyperparameters_rf.R 2 50
 # Rscript s_run_hyperparameters_rf.R 3 20
 Rscript s_run_hyperparameters_rf.R 4 20
-Rscript s_run_hyperparameters.R 4 20
 

### 2020-11-09 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
run through draft

@@
 The experiments not detailed in the main text are briefly described in \S\ref{sec:further_experiments} and more fully in the relevant sections given in Table \ref{tab:results}.
+\textcolor{red}{In all cases where $\widehat{R}$ was calculated, unless otherwise stated, we followed the approach in \cite{vehtari2019rank}, by calculating it as the maximum of rank-normalised split-$\widehat{R}$ and rank-normalised folded split-$\widehat{R}$: for simplicity, we refer to this as rank-normalised split-$\widehat{R}$.}
 
@@
 In Fig.
-\ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate;  as for $R^*$, this metric indicates the chains had not converged in all replicates (as diagnosed by $R^*<1.01$).
+\ref{fig:ar1}C, we show rank-normalised split-$\widehat{R}$ for each replicate;  as for $R^*$, this metric indicates the chains had not converged in all replicates because $\widehat{R}>1.01$.
 
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
 Note that, in D, only a single series of each series type is used to generate distribution.
@@
 	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
-The vertical axis shows the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} (right panel).
+The vertical axis shows the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ (right panel).
 Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
@@
 \ref{fig:cauchy_convergence}B, we show how the maximum of this measurement across all 50 dimensions changes with sample size.
-After c.550 iterations, the alternative parameterisation achieves $\widehat{R}<1.01$ for all target dimensions, and, after c.10,400 iterations, the nominal model achieves the same maximum $\widehat{R}$ value: in both cases, these suggest convergence.
+After c.500 iterations, the alternative parameterisation achieves $\widehat{R}<1.01$ for all target dimensions, and, after c.10,000 iterations, the nominal model achieves the same maximum $\widehat{R}$ value: in both cases, these suggest convergence.
 The second measure is multivariate $\widehat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig.
@@
 In \S\ref{sec:prostate}, we fit Bayesian models with many 1000s of parameters then used $R^*$ to diagnose convergence, finding that $R^*$ was considerably more expensive to calculate than $\widehat{R}$.
-\textcolor{red}{The time complexity of training RFs is thought to be $\mathcal{O}(m_{\text{try}} n_{tree} N \text{log} N)$ \cite[chapter~5]{louppe2014understanding}, and given the similarities with GBMs (which also builds many decision trees), it is likely to be similar.} If so, this suggests that larger \textcolor{red}{statistical} models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
+\textcolor{red}{The time complexity of training RFs is thought to be $\mathcal{O}(m_{\text{try}} n_{\text{tree}} N \text{log} N)$ \cite[chapter~5]{louppe2014understanding}, and given the similarities with GBMs (which also builds many decision trees), it is likely to be similar.} If so, this suggests that larger \textcolor{red}{statistical} models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
 That said, it is possible to reduce the runtime for $R^*$ using thinned draws (although this risks losing chain idiosyncracies) and using a subset of dimensions (although this risks losing problematic dimensions).

### 2020-11-09 write_up/arxiv_submission_aug_2020/supplementary_materials_marked.tex
run through draft

@@
 Means that trend over time is one form of non-stationarity; another is a time-varying covariance.
-Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time.
+Next, we consider a bivariate normal with (constant) standard normal marginals but where the correlation between dimensions trends over time.
 Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
@@
 \begin{itemize}
-	\item The AR(1) example described in \S\ref{sec:heterogeneity}.
+	\item The AR(1) example described in \S\ref{sec:heterogeneity}, except using 1000 draws per chain.
 	\item The 250-dimensional non-centered multivariate normal example introduced in \S\ref{sec:multivariate_normal_250}.
-In each replicate, we generated 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
+In each replicate, we generated 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm across each of 4 chains.
 	\item The alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy}.
@@
 In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate.
-If the number of dimensions is 16 or greater, the order switches, and RFs outperform GBMs.
+If the number of dimensions is 16 or greater, the order switches, and RFs often outperform GBMs.
 

### 2020-11-09 src/s_jobScheduler_hypers.sh

@@ -4,8 +4,8 @@ chmod +x *.R
 chmod +x *.sh
 
 
-# Rscript s_run_hyperparameters.R 1 200
-# Rscript s_run_hyperparameters.R 2 50
-# Rscript s_run_hyperparameters.R 3 20
+Rscript s_run_hyperparameters.R 1 200
+Rscript s_run_hyperparameters.R 2 50
+Rscript s_run_hyperparameters.R 3 20
 Rscript s_run_hyperparameters.R 4 20
 

### 2020-11-09 src/s_jobScheduler_hypers_rf.sh

@@ -4,8 +4,8 @@ chmod +x *.R
 chmod +x *.sh
 
 
-# Rscript s_run_hyperparameters_rf.R 1 200
-# Rscript s_run_hyperparameters_rf.R 2 50
-# Rscript s_run_hyperparameters_rf.R 3 20
+Rscript s_run_hyperparameters_rf.R 1 200
+Rscript s_run_hyperparameters_rf.R 2 50
+Rscript s_run_hyperparameters_rf.R 3 20
 Rscript s_run_hyperparameters_rf.R 4 20
 

### 2020-11-09 src/s_jobScheduler_hypers_xgbtree.sh

@@ -1,10 +1,0 @@
-#!/bin/bash
-
-chmod +x *.R
-chmod +x *.sh
-
-
-Rscript s_run_hyperparameters_xgbtree.R 1 200
-# Rscript s_run_hyperparameters_xgbtree.R 2 50
-# Rscript s_run_hyperparameters_xgbtree.R 3 20
-

### 2020-11-09 src/s_jobScheduler_ml_hypers_set.sh

@@ -4,9 +4,8 @@ chmod +x *.R
 chmod +x *.sh
 
 
-# Rscript s_run_ml_hypers_set.R 1 100
-# Rscript s_run_ml_hypers_set.R 2 100
-# Rscript s_run_ml_hypers_set.R 3 50
-# Rscript s_run_ml_hypers_set.R 4 30
-Rscript s_run_ml_hypers_set.R 5 10
+Rscript s_run_ml_hypers_set.R 1 100
+Rscript s_run_ml_hypers_set.R 2 100
+Rscript s_run_ml_hypers_set.R 3 50
+Rscript s_run_ml_hypers_set.R 4 30
 

### 2020-11-09 src/s_run_hyperparameters_xgbtree.R

@@ -1,132 +1,0 @@
-setwd("/Volumes/Samsung1.5TB/Github/ml-mcmc-convergence/src")
-rm(list=ls())
-library(tidyverse)
-library(reshape2)
-library(rstan)
-library(latex2exp)
-library(caret)
-library(gbm)
-options(mc.cores=4)
-rstan_options(auto_write = TRUE)
-source("monitornew.R")
-source("r_star_monitor.R")
-source("eight_schools.data.R")
-eight_schools <- list(J=J, y=y, sigma=sigma)
-model_cp <- stan_model("eight_schools_cp.stan")
-model_ncp <- stan_model("eight_schools_ncp.stan")
-
-
-# helper functions ------
-f_ar1 <- function(rho, sigma, L){
-  x <- vector(length = L)
-  x[1] <- rnorm(1, 0, sd=sigma)
-  for(i in 2:L)
-    x[i] = rho * x[i - 1] + rnorm(1, 0, sd=sigma)
-  return(x)
-}
-
-f_generate_lower_var_four <- function(var_ratio, rho, sigma, L){
-  x <- matrix(nrow = L, ncol = 4)
-  for(i in 1:3)
-    x[, i] <- f_ar1(rho, sigma, L)
-  z <- f_ar1(rho, sigma * sqrt(var_ratio), L)
-  x[, 4] <- z
-  return(x)
-}
-
-f_data_ar1 <- function(){
-  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 2000)
-  a_array <- array(dim=c(2000, 4, 1))
-  a_array[,,1] <- temp
-  return(a_array)
-}
-
-f_replicate_hyper <- function(m_parameters, f_data_generator){
-  
-  a_array <- f_data_generator()
-  
-  laccuracy <- vector(length = nrow(m_parameters))
-  for(i in 1:nrow(m_parameters))
-    laccuracy[i] <- r_star(a_array,
-                           caret_default=tibble(nrounds = m_parameters$nrounds[i],
-                                                max_depth = m_parameters$max_depth[i],
-                                                eta = m_parameters$eta[i],
-                                                gamma = m_parameters$gamma[i],
-                                                colsample_bytree = m_parameters$colsample_bytree[i],
-                                                min_child_weight = m_parameters$min_child_weight[i],
-                                                subsample = m_parameters$subsample[i]),
-                           method="xgbTree")
-  return(laccuracy)
-}
-
-
[…]

### 2020-11-18 write_up/ba_final_submission/badoc.tex
added ba submission draft

\documentclass[ba]{imsart}
%
\pubyear{0000}
\volume{00}
\issue{0}
\doi{0000}
\arxiv{2010.00000}
\firstpage{1}
\lastpage{1}

%
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue,filecolor=blue,backref=page]{hyperref}
\usepackage{graphicx}

\startlocaldefs
\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\endlocaldefs

\begin{document}

\begin{frontmatter}
\title{A Sample Document\thanksref{T1}}
\runtitle{A Sample Document}
\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}

\begin{aug}
\author{\fnms{First} \snm{Author}\thanksref{addr1,t1,t2,m1}\ead[label=e1]{first@somewhere.com}},
\author{\fnms{Second} \snm{Author}\thanksref{addr1,t3,m1,m2}\ead[label=e2]{second@somewhere.com}}
\and
\author{\fnms{Third} \snm{Author}\thanksref{addr2,t1,m2}%
\ead[label=e3]{third@somewhere.com}%
\ead[label=u1,url]{http://www.foo.com}}

\runauthor{F. Author et al.}

\address[addr1]{Address of the First and Second authors
     Usually a few lines long 
    \printead{e1} % print email address of "e1"
    \printead*{e2}
}

\address[addr2]{Address of the Third author
    Usually a few lines long
    Usually a few lines long
    \printead{e3}
    \printead{u1}
}

\thankstext{t1}{Some comment}
\thankstext{t2}{First supporter of the project}
\thankstext{t3}{Second supporter of the project}

\end{aug}

\begin{abstract}
The abstract should summarize the contents of the paper. It 
should be clear, descriptive, self-explanatory and not longer
than 200 words. It should also be suitable for publication in
abstracting services. Please avoid using math formulas as 
much as possible.

This is a sample input file.  Comparing it with the output it
generates can show you how to produce a simple document 
of your own.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{60K35}
\kwd{60K35}
\kwd[; secondary ]{60K35}
\end{keyword}

\begin{keyword}
\kwd{sample}
\kwd{\LaTeXe}
\end{keyword}

\end{frontmatter}

\section{Ordinary text}

The ends of words and sentences are marked by spaces. It doesn't
matter how many spaces you type; one is as good as 100.  The end of a
line counts as a space.

One or more blank lines denote the end of a paragraph.

Since any number of consecutive spaces are treated like a single one,
the formatting of the input file makes no difference to
\TeX, % The \TeX command generates the TeX logo.
but it makes a difference to you.  When you use
\LaTeX, % The \LaTeX command generates the LaTeX logo.
making your input file as easy to read as possible will be a great
help as you write your document and when you change it.  This sample
file shows how you can add comments to your own input file.

Because printing is different from typewriting, there are a number of
things that you have to do differently when preparing an input file
than if you were just typing the document directly.  Quotation marks
like ``this'' have to be handled specially, as do quotes within
quotes: ``\,`this' % \, separates the double and single quote.
is what I just wrote, not `that.'\,''

Dashes come in three sizes: a hyphen, %an intra-word dash,
a medium dash for number ranges like 1--2, and a punctuation dash in
place of a comma, semicolon, colon or parentheses ---like this.

A sentence-ending space should be larger than the space between words
within a sentence.  You sometimes have to type special commands in
conjunction with punctuation characters to get this right, as in the
following sentence.  Gnats, gnus,
etc.\ % `\ ' makes an inter-word space.
all begin with G\@.  % \@ marks end-of-sentence punctuation.
You should check the spaces after periods when reading your output to
make sure you haven't forgotten any special cases.  Generating an
ellipsis \ldots\ % `\ ' needed because TeX ignores spaces after
          % command names like \ldots made from \ + letters.
          %
          % Note how a `%' character causes TeX to ignore the
          % end of the input line, so these blank lines do not
          % start a new paragraph.
with the right spacing around the periods
requires a special  command.

\TeX\ interprets some common characters as commands, so you must type
special commands to generate them.  These characters include the
following:
%       \$ - arXiv do not like it...
        \& \% \# \{ and~\}.

In printing, text is emphasized by using an {\em
          italic\/} % The \/ command produces the tiny extra space that
               % should be added between a slanted and a following
               % unslanted letter.
type style.

\begin{em}
  A long segment of text can also be emphasized in this way.  Text
  within such a segment given additional emphasis with\/ {\em Roman}
  type.  Italic type loses its ability to emphasize and become simply
  distracting when used excessively.
\end{em}

It is sometimes necessary to prevent \TeX\ from breaking a line where
it might otherwise do so.  This may be at a space, as between the
``Mr.'' and ``Jones'' in
``Mr.~Jones'', % ~ produces an unbreakable interword space.
or within a word---especially when the word is a symbol like \mbox{\em
  itemnum\/} that makes little sense when hyphenated across lines.

\TeX\ is good at typesetting mathematical formulas like \( x-3y = 7 \)
or \( a_{1} > x^{2n} / y^{2n} > x' \).  Remember that a letter like
$x$ % $ ... $ and \( ... \) are equivalent
is a formula when it denotes a mathematical symbol, and should be
treated as one.

\section{Notes}

Footnotes\footnote{This is an example of a footnote.}  pose no
problem\footnote{And another one}.

\section{Displayed text}

Text is displayed by indenting it from the left margin.  Quotations
are commonly displayed.  There are short quotations
\begin{quote}
  This is a short a quotation.  It consists of a single paragraph of
  text.  There is no paragraph indentation.
\end{quote}
and longer ones.
\begin{quotation}
  This is a longer quotation.  It consists of two paragraphs of text.
  The beginning of each paragraph is indicated by an extra
  indentation.

  This is the second paragraph of the quotation.  It is just as dull
  as the first paragraph.
\end{quotation}
Another frequently displayed structure is a list.  The following is an
example of an {\em itemized} list, four levels deep.
\begin{itemize}
\item This is the first item of an itemized list.  Each item in the
  list is marked with a ``tick.''  The document style determines what
  kind of tick mark is used.
\item This is the second item of the list.  It contains another list
  nested inside it.  The three inner lists are an {\em itemized} list.
    \begin{itemize}
    \item This is the first item of an enumerated list that is nested
      within the itemized list.
    \item This is the second item of the inner list.  \LaTeX\ allows
      you to nest lists deeper than you really should.
    \end{itemize}
    This is the rest of the second item of the outer list.  It is no
    more interesting than any other part of the item.
   \item  This is the third item of the list.
\end{itemize}

The following is an example of an {\em enumerated} list, four levels
deep.
\begin{enumerate}
\item This is the first item of an enumerated list.  Each item in the
  list is marked with a ``tick.''  The document style determines what
  kind of tick mark is used.
\item This is the second item of the list.  It contains another list
  nested inside it.  The three inner lists are an {\em enumerated}
  list.
  \begin{enumerate}
  \item This is the first item of an enumerated list that is nested
    within the enumerated list.
  \item This is the second item of the inner list.  \LaTeX\ allows you
    to nest lists deeper than you really should.
  \end{enumerate}
  This is the rest of the second item of the outer list.  It is no
  more interesting than any other part of the item.
\item This is the third item of the list.
\end{enumerate}

The following is an example of a {\em description} list.
\begin{description}
\item[Cow] Highly intelligent animal that can produce milk out of grass.
\item[Horse] Less intelligent animal renowned for its legs.
\item[Human being] Not so intelligent animal convinced it can think.
\end{description}

You can even display poetry.
\begin{verse}
  There is an environment for verse
  \\ % The \\ command separates lines
  Whose features some poets will curse.  % within a stanza.

               % One or more blank lines separate stanzas.

  For instead of making\\
  Them do {\em all\/} line breaking, \\
  It allows them to put too many words on a line when they'd rather be
  forced to be terse.
\end{verse}

Mathematical formulas may also be displayed.  A displayed formula is
one-line long:

   \[  x' + y^{2} = z_{i}^{2};\]

   multiline formulas require special formatting instructions.

   Do not start a paragraph with a displayed equation, nor make one a
   paragraph by itself.

Example of a theorem:

\begin{thm}
All conjectures are interesting, but some conjectures are more
interesting than others.
\end{thm}

\begin{proof}
Obvious.
\end{proof}

\section{Tables and figures}
Cross reference to labeled table: As you can see in Table~\ref{sphericcase} on
page~\pageref{sphericcase} and also in Table~\ref{parset} on page~\pageref{parset}.

\begin{table*}
\begin{tabular}{crrrrc}
\hline
Equil. \\
Points & \multicolumn{1}{c}{$x$} & \multicolumn{1}{c}{$y$} & \multicolumn{1}{c}{$z$} & \multicolumn{1}{c}{$C$} &
S \\
\hline
$~~L_1$ & $-$2.485252241 & 0.000000000 & 0.017100631 & 8.230711648 & U \\
$~~L_2$ &    0.000000000 & 0.000000000 & 3.068883732 & 0.000000000 & S \\
$~~L_3$ &    0.009869059 & 0.000000000 & 4.756386544 & $-$0.000057922 & U \\
$~~L_4$ &    0.210589855 & 0.000000000 & $-$0.007021459 & 9.440510897 & U \\
$~~L_5$ &    0.455926604 & 0.000000000 & $-$0.212446624 & 7.586126667 & U \\
$~~L_6$ &    0.667031314 & 0.000000000 & 0.529879957 & 3.497660052 & U \\
$~~L_7$ &    2.164386674 & 0.000000000 & $-$0.169308438 & 6.866562449 & U \\
$~~L_8$ &    0.560414471 & 0.421735658 & $-$0.093667445 & 9.241525367 & U \\
$~~L_9$ &    0.560414471 & $-$0.421735658 & $-$0.093667445 & 9.241525367 & U
\\
$~~L_{10}$ & 1.472523232 & 1.393484549 & $-$0.083801333 & 6.733436505 & U \\
$~~L_{11}$ & 1.472523232 & $-$1.393484549 & $-$0.083801333 & 6.733436505 & U
\\ \hline
\end{tabular}
\caption{The spherical case ($I_1=0$, $I_2=0$)}
\label{sphericcase}
\end{table*}

A major point of difference lies in the value of the specific
production rate $\pi$ for large values of the specific growth rate
$\mu$.  Already in the early publications \cite{akaike,dyke,greene} it
appeared that high glucose concentrations in the production phase are
well correlated with a low penicillin yield (the ``glucose
effect''). It has been confirmed recently
\cite{akaike,dyke,greene,hilbe} that high glucose concentrations
inhibit the synthesis of the enzymes of the penicillin pathway, but
not the actual penicillin biosynthesis.  In other words, glucose
represses (and not inhibits) the penicillin biosynthesis.

These findings do not contradict the results of \cite{akaike} and of
\cite{hilbe} which were obtained for continuous culture fermentations.
Because for high values of the specific growth rate $\mu$ it is most
likely (as shall be discussed below) that maintenance metabolism
occurs, it can be shown that in steady state continuous culture
conditions, and with $\mu$ described by a Monod kinetics
\begin{equation}
    C_{s}  =  K_{M} \frac{\mu/\mu_{x}}{1-\mu/\mu_{x}} \label{cs}
\end{equation}
Pirt \& Rhigelato determined $\pi$ for $\mu$ between
$0.023$ and $0.086$ h$^{-1}$.
They also reported a value $\mu_{x} \approx 0.095$
h$^{-1}$, so that for their experiments $\mu/\mu_{x}$ is in the range
of $0.24$ to $0.9$.
Substituting $K _M$ in equation (\ref{cs}) by
the value $K_{M}=1$ g/L as used by \cite{akaike}, one finds
with the above equation $0.3 < C_{s} < 9$ g/L. This agrees with
the work of  \cite{hilbe}, who reported that penicillin biosynthesis
repression only occurs at glucose concentrations from $C_{s}=10$ g/L on.
The conclusion is that the glucose concentrations in the experiments of
Pirt \& Rhigelato probably were too low for glucose repression to be
detected. The experimental data published by Ryu \& Hospodka
are not detailed sufficiently to permit a similar analysis.

\begin{table}
\begin{tabular}{lrll}
\hline
\multicolumn{2}{l}{\it parameter} & {\it Set 1} & {\it Set 2}\\
\hline
$\mu_{x}$           & [h$^{-1}$]  & 0.092       & 0.11          \\
$K_{x}$             & [g/g DM]     & 0.15        & 0.006         \\
$\mu_{p}$           & [g/g DM h]  & 0.005       & 0.004         \\
$K_{p}$             & [g/L]        & 0.0002      & 0.0001        \\
$K_{i}$             & [g/L]        & 0.1         & 0.1           \\
$Y_{x/s}$           & [g DM/g]     & 0.45        & 0.47          \\
$Y_{p/s}$           & [g/g]        & 0.9         & 1.2           \\
$k_{h}$             & [h$^{-1}$]  & 0.04        & 0.01          \\
$m_{s}$             & [g/g DM h]  & 0.014       & 0.029         \\
\hline
\end{tabular}
\caption{Parameter sets used by Bajpai \& Reu\ss\ }\label{parset}
\end{table}

Bajpai \& Reu\ss\ decided to disregard the differences between time
constants for the two regulation mechanisms (glucose repression or
inhibition) because of the relatively very long fermentation times,
and therefore proposed a Haldane expression for $\pi$.

It is interesting that simulations with the \cite{hilbe} model for the
initial conditions given by these authors indicate that, when the
remaining substrate is fed at a constant rate, a considerable and
unrealistic amount of penicillin is produced when the glucose
concentration is still very high \cite{dyke,greene,hilbe} Simulations
with the Bajpai \& Reu\ss\ model correctly predict almost no
penicillin production in similar conditions.

% figure 1
\begin{figure} 
\includegraphics{unmixed_1.pdf}
\caption[]{Pathway of the penicillin G biosynthesis.}
\label{penG}
\end{figure}

Sample of cross-reference to figure.
Figure~\ref{penG} shows that is not easy to get something on paper.

\section{Headings}

\subsection{Subsection}

Carr--Goldstein based their model on balancing methods and biochemical
know\-ledge. The original model (1980) contained an equation for the
oxygen dynamics which has been omitted in a second paper (1981). This
simplified model shall be discussed here.

\subsubsection{Subsubsection}

Carr--Goldstein based their model on balancing methods and biochemical
know\-ledge. The original model (1980) contained an equation for the
oxygen dynamics which has been omitted in a second paper (1981). This
simplified model shall be discussed here.

\section{Equations and the like}

Two equations:
\begin{equation}
    C_{s}  =  K_{M} \frac{\mu/\mu_{x}}{1-\mu/\mu_{x}} \label{ccs}
\end{equation}
and
\begin{equation}
    G = \frac{P_{\rm opt} - P_{\rm ref}}{P_{\rm ref}} \mbox{\ }100 \mbox{\ }(\%).
\end{equation}

Two equation arrays:

\begin{eqnarray}
  \frac{dS}{dt} & = & - \sigma X + s_{F} F\\
  \frac{dX}{dt} & = &   \mu    X\\
  \frac{dP}{dt} & = &   \pi    X - k_{h} P\\
  \frac{dV}{dt} & = &   F
\end{eqnarray}
and
\begin{eqnarray}
 \mu_{\rm substr} & = & \mu_{x} \frac{C_{s}}{K_{x}C_{x}+C_{s}}  \\
 \mu              & = & \mu_{\rm substr} - Y_{x/s}(1-H(C_{s}))(m_{s}+\pi /Y_{p/s}) \\
 \sigma           & = & \mu_{\rm substr}/Y_{x/s}+ H(C_{s}) (m_{s}+ \pi /Y_{p/s}).
\end{eqnarray}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, if any, should   %%
%% be provided in {supplement} environment  %%
%% with title and short description.        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Title of Supplement A}
\sdescription{Short description of Supplement A.}
\end{supplement}
\begin{supplement}
\stitle{Title of Supplement B}
\sdescription{Short description of Supplement B.}
\end{supplement}

The following are some other citations to check that the bibliographc style
file is working correctly: \citet{akaike}, \citet*{akivarsq}, \citet{dyke},
\citet{greene}, \citet*{kstuart}, \citet{hilbetech}, \citet{hilbe},
\citet{hilbeglm}, \citet{maddalacntrsq}, and \citet*{companion}.

\bibliographystyle{ba}
\bibliography{sample}

\begin{acknowledgement}
And this is an acknowledgements section with a heading that was produced by the
$\backslash$section* command. Thank you all for helping me writing this
\LaTeX\ sample file.
\end{acknowledgement}

\end{document}

### 2020-11-18 write_up/ba_final_submission/ms_final.tex
added ba submission draft

--- 2020-09-04 write_up/arxiv_submission_aug_2020/ms_arxiv_2020.tex
@@
-\documentclass{article}
-\pdfoutput=1
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2018
-
-% ready for submission
-% \usepackage{neurips_2018}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-% \usepackage[preprint]{neurips_2019}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-%\usepackage{neurips_2019}
-
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2018}
-
-\usepackage{arxiv}
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\hypersetup{unicode=true,
-            pdfborder={0 0 0},
-            breaklinks=true,
-            colorlinks=true,
-            linkcolor=blue,
-            citecolor=blue,
-            filecolor=blue,
-            urlcolor=blue}
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
+\documentclass[ba]{imsart}
+%
+\pubyear{0000}
+\volume{00}
+\issue{0}
+\doi{0000}
+\arxiv{2003.07900}
+\firstpage{1}
+\lastpage{1}
+
+%
+\usepackage{amsthm}
 \usepackage{amsmath}
-\usepackage{algorithm,algorithmic}
+\usepackage{natbib}
+\usepackage[colorlinks,citecolor=blue,urlcolor=blue,filecolor=blue,backref=page]{hyperref}
 \usepackage{graphicx}
 \usepackage{bbm}
-\usepackage{caption} 
-\usepackage{natbib}
-\captionsetup[table]{skip=10pt}
-
-\newcommand{\aki}[1]{\textcolor{red}{[Aki: #1]}}
-
-\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}
-\newcommand{\Dir}[0]{\textrm{Dirichlet}}
-\newcommand{\Ray}[0]{\textrm{Rayleigh}}
-\newcommand{\gam}[0]{\textrm{Gamma}}
-\newcommand{\dgamma}[0]{\textrm{Gamma}}
-\newcommand{\dpoisson}[0]{\textrm{Poisson}}
-\newcommand{\dbeta}[0]{\textrm{Beta}}
-\newcommand{\dbern}[0]{\textrm{Bernoulli}}
-\newcommand{\dunif}[0]{\mathrm{Uniform}}
-\newcommand{\dgig}[0]{\textrm{GIG}}
-\newcommand{\dnormal}[0]{\mathrm{Normal}}
-\newcommand{\dt}[0]{\mathrm{t}}
-\newcommand{\igamma}[0]{\textrm{Gamma}^{-1}}
-\newcommand{\rayl}[0]{\textrm{Rayleigh}}
-\newcommand{\Exp}[0]{\textrm{Exponential}}
-\newcommand{\Bet}[0]{\textrm{Beta}}
-\newcommand{\GEM}[0]{\textrm{GEM}}
-\newcommand{\DP}[0]{\textrm{DP}}
-\newcommand{\ESS}[0]{\mathrm{ESS}}
-\newcommand{\bm}[1]{\boldsymbol{#1}}
-\newcommand{\bbeta}{\bm{\beta}}
-\newcommand{\bpi}{\bm{\pi}}
-\newcommand{\bomega}{\bm{\omega}}
-\newcommand{\bgamma}{\bm{\gamma}}
-\newcommand{\blambda}{\bm{\lambda}}
-\newcommand{\bphi}{\bm{\phi}}
-\newcommand{\btheta}{\bm{\theta}}
-\newcommand{\bmu}{\bm{\mu}}
-\newcommand{\bb}{\bm{b}}
-\newcommand{\bk}{\bm{k}}
-\newcommand{\bl}{\bm{l}}
-\newcommand{\bn}{\bm{n}}
-\newcommand{\bw}{\bm{w}}
-\newcommand{\bz}{\bm{z}}
-\newcommand{\bx}{\bm{x}}
-\newcommand{\bX}{\bm{X}}
-\newcommand{\by}{\bm{y}}
-\newcommand{\bZ}{\bm{Z}}
-\newcommand{\bW}{\bm{W}}
-\newcommand{\bS}{\bm{S}}
-\newcommand{\bH}{\bm{H}}
-\newcommand{\Mult}{\textrm{Multinomial}}
-\newcommand{\N}{\mathcal{N}}
-\newcommand{\NEW}{\textrm{\tiny new}}
-\newcommand{\OLD}{\textrm{\tiny old}}
-\newcommand{\sigmat}{\sigma^2}
-\newcommand{\IBP}{\textrm{IBP}}
-\newcommand{\E}{\mathbb{E}}
-\newcommand{\V}{\mathbb{V}}
-\newcommand{\Eq}{\mathbb{E}_q}
-\newcommand{\cL}{\mathcal{L}}
-\newcommand{\cB}{\mathcal{B}}
-\newcommand{\cC}{\mathcal{C}}
-\newcommand{\cV}{\mathcal{V}}
-\newcommand{\cHq}{\mathcal{H}_q}
-\newcommand{\test}[1]{\mbox{$#1$}^{\small \mbox{test}}}
-\newcommand{\alphaW}{\alpha^{(W)}}
-\newcommand{\alphaH}{\alpha^{(H)}}
-\newcommand{\betaW}{\beta^{(W)}}
-\newcommand{\betaH}{\beta^{(H)}}
-\newcommand{\gammaW}{\gamma^{(W)}}
-\newcommand{\gammaH}{\gamma^{(H)}}
-\newcommand{\gammaT}{\gamma^{(\theta)}}
-\newcommand{\rhoW}{\rho^{(W)}}
-\newcommand{\rhoH}{\rho^{(H)}}
-\newcommand{\rhoT}{\rho^{(\theta)}}
-\newcommand{\tauW}{\tau^{(W)}}
-\newcommand{\tauH}{\tau^{(H)}}
-\newcommand{\tauT}{\tau^{(\theta)}}
-\newcommand{\muW}{\hat{W}}
-\newcommand{\muH}{\hat{H}}
-\newcommand{\Var}{\textrm{Var}}
-\newcommand{\LF}{\mathrm{Leapfrog}}
-
-\newcommand{\beginsupplement}{%
-	\setcounter{table}{0}
-	\renewcommand{\thetable}{S\arabic{table}}%
-	\setcounter{figure}{0}
-	\renewcommand{\thefigure}{S\arabic{figure}}%
-	\setcounter{section}{0}
-	\renewcommand{\thesection}{S\arabic{section}}%
+\usepackage{algorithm,algorithmic}
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{booktabs}
+
+\startlocaldefs
+\numberwithin{equation}{section}
+\theoremstyle{plain}
+\newtheorem{thm}{Theorem}[section]
+\endlocaldefs
+
+\begin{document}
+
+	\begin{frontmatter}
+		\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers}
+		\runtitle{A MCMC convergence diagnostic using decision tree classifiers}
+
+		\begin{aug}
+			\author{\fnms{Ben} \snm{Lambert}\thanksref{addr1}\ead[label=e1]{ben.c.lambert@gmail.com}}
+			\and
+			\author{\fnms{Aki} \snm{Vehtari}\thanksref{addr2}%
+				\ead[label=e3]{aki.vehtari@aalto.fi}%
+				\ead[label=u1,url]{http://www.foo.com}}
+
+			\runauthor{Lambert and Vehtari}
+
+			\address[addr1]{MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London,W2 1PG, United Kingdom
+				\printead{e1} % print email address of "e1"
 			}
 
-\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{%
-	 Ben Lambert\\
-	 MRC Centre for Global Infectious Disease Analysis\\
-	 School of Public Health\\
-	 Imperial College London\\
-	 W2 1PG, United Kingdom\\
-	 \texttt{ben.c.lambert@gmail.com} \\
-	 \And
-	 Aki Vehtari \\
-	 Department of Computer Science\\
-	 Aalto University\\
-	 Finland\\
-	 \texttt{aki.vehtari@aalto.fi}
+			\address[addr2]{Department of Computer Science, Aalto University, Finland
+				\printead{e3}
 			}
 
-\begin{document}
-% \nipsfinalcopy is no longer used
-
-\maketitle
+		\end{aug}
 
@@
 			Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
-Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance.
+Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its performance in finite time.
 The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
-Here, we introduce a new method for diagnosing convergence based on whether a machine learning classifier model can successfully discriminate the individual chains.
+Here, we introduce a new method for diagnosing convergence based on how well a machine learning classifier model can successfully discriminate the individual chains.
 We call this convergence measure $R^*$.
 In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
-Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$.
-The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
+Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead it uses all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
+We recommend calculating $R^*$ using two different machine learning classifiers — gradient-boosted regression trees and random forests — which each work well in models of different dimensions.
+Because each of these methods outputs a classification probability, as a byproduct, we obtain uncertainty in $R^*$.
+The method is straightforward to implement and could be a complementary additional check on MCMC convergence for applied analyses.
 		\end{abstract}
+
+	\end{frontmatter}
 
@@
 	Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement.
+In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan}, PyMC3 \citep{salvatier2016probabilistic}, Turing \citep{ge2018turing}, TensorFlow Probability  \citep{dillon2017tensorflow} and Pyro \citep{bingham2019pyro} implement.
 MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
@@
 A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior.
-Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
-
-The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
+Because of this, software packages (Stan \citep{carpenter2017stan}, for example), go to great lengths to communicate to users any issues with sampling.
+
+	The most important determination of whether MCMC has worked is how closely the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
 MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions).
@@
 
-The predominant diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
+	The one diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
 Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
@@
 
-Here, we introduce $R^*$, a new convergence metric.
+	Here, we introduce $R^*$, a new convergence diagnostic metric.
 This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
 Rephrased, it should not be possible to predict which draws come from which chain.
-In this vein, we use a machine learning (ML) classifier to measure convergence.
-Specifically, we train a ML classifier to predict the chain that generated each observation.
-By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
-Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
-However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy.
-For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}.
-For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
-It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies.
-In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
+In this vein, we use machine learning (ML) classifiers to measure convergence.
+Specifically, we train classifiers to predict the chain that generated each observation.
+By evaluating the performance of classifiers on a held-out test set, this provides a new convergence metric.
+To maximise predictive accuracy, our chosen classifiers naturally exploit differences in the full joint distributions between chains, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
+Our statistic, unlike its $\widehat{R}$ cousins, is scalar-valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
+However, the ML classifiers we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.
+
+	There are, of course, a huge variety of possible ML classifiers.
+For a method to be useful and widely adopted, however, it needs to satisfy a number of criteria: first, across a range of examples, it should consistently be able to detect poor MCMC convergence; second, the ML classifier should be trainable in a reasonable amount of time; lastly, the methods should be tuned (via their hyperparameters) so as to be standardised, so that $R^*$ computed by one analyst is comparable to that computed by another.
+Here, we perform comparisons across some of the most popular methods in the ML literature and recommend two tree-based ML classifiers: gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM'') and random forest classifiers \citep{breiman2001random} (``RF'').
+Both of these methods performed consistently well across our range of examples, and the models were relatively efficient to train.
+We recommend calculating $R^*$ using both of these methods: GBMs tend to perform best for low dimensional cases; in moderate dimensional cases upwards, RF dominates.
+This difference in performance is partly due to hyperparameter tuning: echoing previous results in the literature \cite[Chapters~11\&12]{boehmke2019hands}, our results show that GBMs, in general, are more sensitive to hyperparameter choice than are RFs.
+For a GBM classifier to perform well in higher dimensions, more rounds of boosting are required, which substantially increases training time; instead, we fix the hyperparameters of GBMs to specific values to ensure practical usefulness.
+For RF classifiers, empirical studies have demonstrated reliable heuristics for selecting robust hyperparameters \citep{bernard2009influence}, and, in any case, across our set of test cases, these classifiers are relatively insensitive to hyperparameter choice.
+
+	For the types of problem we tested, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
+In addition, since ML classifiers can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
 $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
 
-The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank} and elsewhere.
+	The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios.
 Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
@@
 	If Markov chains have not mixed, it is possible to guess (with more accuracy than chance) to which chain a draw belongs from its value.
-This is possible if there are differences in the sampling distribution for any dimension, $\theta$, in the target distribution (Fig.
+This is possible if there are differences in the sampling distribution for any dimension in the target distribution (Fig.
 \ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a draw belongs to.
@@
 These two cases, whilst simple, illustrate the basis of our approach.
-To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each draw belongs.
+To determine if a set of Markov chains has converged to the same distribution, we train a supervised ML model to classify the chain to which each draw belongs.
 By evaluating its performance on a separate test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
+By taking the ratio of classification accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
 In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
@@
 
-The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours.
-Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}.
+To identify promising candidate ML methods, we run a series of experiments using a number of popular classifiers (see \S\ref{sec:ml_model}).
+Two methods performed consistently well across our examples: gradient-boosted regression trees (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}) and random forest classifiers \citep{breiman2001random} (``RF'').
+Both of these approaches are based on decision trees: GBMs use an iterative approach known as ``boosting'', where each subsequent decision tree aims to predict the residuals from the previous one; RFs use an approach known as ``bagging'' which trains decision trees on many bootstrapped copies of the training data, yielding a collection of independently trained trees whose individual decisions, when aggregated, yield a class.
+For the implementations of these classifiers, we use those available in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' \citep{greenwell2019package} and ``randomForest'' \citep{liaw2002classification} packages.
+
 The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters.
 We split each chain's draws into randomly divided training and testing sets: here, we use 70\% of draws for training and 30\% for testing.
-Our GBM was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:heterogeneity_robustness}).
-Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.
-
-From the GBM fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
+Both approaches typically took $\mathcal{O}$(seconds) on a desktop computer to execute training then prediction on a testing set for most models we consider in \S\ref{sec:results}.
+
+Since different posteriors present different challenges to MCMC samplers, the nature of classification boundaries is problem-specific.
+Because of this, there is no unique optimal classifier across all problems, and the performance of the algorithms we investigate depends on their hyperparameters (see \S\ref{sec:hyperparameters}).
+GBM models have a number of hyperparameters and have previously been demonstrated to be hard to tune \cite[Chapter~12]{boehmke2019hands}.
+To ensure practical run time for this method, we suggest running it using a default hyperparameter set: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees be grown.
+This choice of hyperparameters was chosen to present a balance between training cost and classification accuracy, producing an $R^*$  that was a stringent measure of convergence — in general, more stringent than $\widehat{R}$.
+RFs are generally less sensitive to variation in their single hyperparameter: $m_{\text{try}}$, the size of the subset of all features over which to search for an optimal split \cite[Chapter~11]{boehmke2019hands}.
+Our experiments replicate these results (see \S\ref{sec:hyperparameters}), and we suggest running RFs using the heuristic $m_{\text{try}}=\sqrt{K}$, which has previously been shown to be a choice resulting in robust classification performance \citep{bernard2009influence,boehmke2019hands}.
+Unless otherwise stated, in the examples explored in \S\ref{sec:results}, the hyperparameters for these two classifiers were set at these defaults.
+
+From a classifier fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
 Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
-For each draw, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
+For each draw, $s$, in our testing set, classifiers output a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
 By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
@@
 The experiments not detailed in the main text are briefly described in \S\ref{sec:further_experiments} and more fully in the relevant sections given in Table \ref{tab:results}.
+In all cases where $\widehat{R}$ was calculated, unless otherwise stated, we followed the approach in \cite{vehtari2019rank}, by calculating it as the maximum of rank-normalised split-$\widehat{R}$ and rank-normalised folded split-$\widehat{R}$: for simplicity, we refer to this as rank-normalised split-$\widehat{R}$.
 
 \begin{table}[]
+	\footnotesize
 	\begin{tabular}{l|l|l}
@@
 		Cauchy & \textbf{Detecting convergence for long-tailed distributions} & \ref{sec:cauchy}\\
-		& Comparing $R^*$ and existing measures to objective convergence & \ref{sec:cauchy_objective}\\
+		& Comparing $R^*$ and existing measures to & \ref{sec:cauchy_objective}\\
+		& objective convergence &\\
 		\midrule
@@
 		\midrule
-		Ovarian and prostate models & \textbf{Bayesian models with many parameters and multimodal posteriors} & \ref{sec:prostate}\\
+		Ovarian and prostate models & \textbf{Bayesian models with many parameters} & \ref{sec:prostate}\\
+		&  \textbf{and multimodal posteriors} &\\
+		\midrule		
+		Discrete Markov model & \textbf{Evaluating $R^*$ on discrete parameter models} & \ref{sec:discrete}\\
+		& Small state-space & \ref{sec:discrete_small}\\
+		& Large state-space & \ref{sec:discrete_large}\\
+		\midrule
+		Various & \textbf{Sensitivity of $R^*$ to ML model} & \ref{sec:ML_sensitivity}\\
+		& Comparing different ML classifiers & \ref{sec:ml_model}\\
+		& Sensitivity of $R^*$ to GBM and RF hyperparameters & \ref{sec:hyperparameters}\\
+		\midrule
+		Multivariate normal \& & \textbf{Comparing GBM and RF classifiers} & \ref{sec:comparison_gbm_rf}\\
+		Student-t dists. & &\\
+		& Detecting convergence in joint distributions & \ref{sec:joint_distribution}\\
+		& Tail convergence & \ref{sec:tail_fatness}\\
 	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
@@
 \subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
-In this section, we show how $R^*$ is able to detect heterogeneous variance across Markov chains.
-Apart from demonstrating how $R^*$ performs, this section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}); and, finally, to the ML model hyperparameters (\S\ref{sec:heterogeneity_robustness}).
+In this section, we use a simple example to illustrate how $R^*$ works.
+Specifically, we show how $R^*$ can detect heterogeneous variance across Markov chains.
+This example aims to illustrate the basic mechanics behind how $R^*$ works, so we use only the GBM classifier here.
+This section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model-predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); and to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}).
 The experiments we use to study these issues are all of similar form to the following data generating process: four Markov chains are generated, where each samples from an autoregressive order 1 (AR(1)) process of the form,
@@
 	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values as calculated in \cite{vehtari2019rank} for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
-Note that, in D, only a single series of each series type is used to generate distribution.}
+	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
+Note that, in D, only a single series of each series type is used to generate distribution.
+All examples used a GBM for classification using the default hyperparameter values given in \S\ref{sec:method}.}
 	\label{fig:ar1}
@@
 Intuitively, the reduction in sampling variation when composing training and test sets from larger samples results in lower variability in ML model predictions.
-We also expect that larger sample sizes should lead to higher $R^*$ values, since more training data leads to better ML models.
+We also expect that larger sample sizes should lead to higher $R^*$ values with lower variance, since more training data leads to ML models with lower generalisation error.
 We may start to see this here, since the median $R^*=1.25$ at a sample size of 8000 was greater than for the smaller samples.
@@
 We have so far focused on the sensitivity of $R^*$ to chain heterogeneity with a fixed number of chains: four.
-Since classification becomes a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
+Since classification may become a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
 For comparison, we also illustrate how the performance of rank-normalised split-$\widehat{R}$ varies with number of chains.
@@
 In general, both measures decline with chain count.
-For $R^*$, this is because it is harder to classify chains when there are more of them.
+For $R^*$, this may be because it is harder to classify chains when there are more of them.
 For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
@@
 
-\subsubsection{Robustness to hyperparameter choice}\label{sec:heterogeneity_robustness}
-The performance of GBM, like all ML methods, depends on its hyperparameters.
-To investigate the robustness of $R^*$ to GBM hyperparameter choice, we performed a sensitivity analysis.
-In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig.
-\ref{fig:ar1}.
-The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200).
-The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig.
-\ref{fig:ar1_sensitivity}.
-These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper).
-This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ar1_sensitivity.pdf}}
-	\caption{\textbf{Autoregressive example: sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases).
-The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star} using split chains; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model.}
-	\label{fig:ar1_sensitivity}
-\end{figure}
-
+\color{black}
 \subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
@@
 
-Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty}, which is shown in Fig.
+Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty} using both GBM and RF classifiers.
+These distributions are shown in Fig.
 \ref{fig:bivariate}.
-The mean of this distribution is 1.14, and $>$99\% of $R^*$ draws are above 1, indicating that the sampling distribution has not converged.
+The mean of the GBM-$R^*$ distribution is 1.14, and $>$99\% of $R^*$ draws are above 1; the mean of the RF-$R^*$ distribution was 1.27 and all draws were above 1.
+Collectively, these measures indicate that the sampling distribution has not converged.
 By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
@@
 where $\boldsymbol{x}\in\mathbb{R}^{250}$.
-For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig.
-\ref{fig:mvt}A.
+For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} with a GBM classifier to generate an uncertainty distribution for $R^*$, which is shown in Fig.
+\ref{fig:mvt}A (the equivalent plot for a RF classifier is similar and shown in Fig.
+\ref{fig:mvt_gbm_vs_rf}).
 From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig.
+Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for all parameters (Fig.
 \ref{fig:mvt}B), although 54\% had bulk-$\text{ESS}<400$ and 13\% of parameters had tail-$\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
@@
 \subsubsection{Variable importance}\label{sec:multivariate_normal_varimportance}
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions.
+In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), which allows us to determine which variables were most informative for predictions.
 We now compare these with the more established metrics $\widehat{R}$ and ESS.
@@
 
-In the top-left panel of Fig.
-\ref{fig:cauchy}, we show the $R^*$ distribution under both parameterisations.
+In the top-left and top-middle panel of Fig.
+\ref{fig:cauchy}, we show the $R^*$ distribution for GBM and RF classifiers under both parameterisations.
 As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
@@
 \ref{fig:cauchy}, we show the results for these longer runs.
-In these, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred.
+In these, the alternative parameterisation now has an $R^*$ distribution centred on $R^*=1$ for the GBM classifier although the RF classifier $R^*$ distribution remains slightly rightwards of this target indicating that convergence is nearer but more iterations are likely still required.
 Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
@@
 In Fig.
-\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models: these indicate that, after 25,000 iterations, $R^*\approx 1.05$ for the alternative model, and $R^*>2$ for the nominal parameterisation: both these $R^*$ values suggest lack of convergence.
+\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models and for both GBM and RF classifiers: these indicate that, after 25,000 iterations, for the alternative model, $R^*\approx 1.05$ for the GBM classifier and $R^*\approx 1.74$ for the RF classifier; for the nominal model, $R^*>2$ for the GBM classifier $R^*>3$ for the RF classifier: all these $R^*$ values suggest lack of convergence.
 Finally, in Figs.
@@
 We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the NUTS algorithm \citep{betancourt2017conceptual}, meaning that the resultant sampling distribution is likely to be biased.
+Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the dynamic HMC algorithm \citep{betancourt2017conceptual} (called using the ``NUTS'' option in Stan), meaning that the resultant sampling distribution is likely to be biased.
 We also used the same algorithm settings to sample from the non-centered model.
@@
 To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models.
+Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models using a GBM classifier.
 The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
@@
 
+Fig.
+\ref{fig:eight_schools_rf} shows the equivalent of Fig.
+\ref{fig:eight_schools} except using a RF classifier.
+The results are similar, although the $R^*$ distributions are shifted slightly rightwards: indicating, for example, that more than 2000 draws from the non-centered model may be required for convergence.
+
 It is recommended that $\widehat{R}$, like $R^*$, be calculated using split chains.
@@
 \ref{fig:eight_schools}C, we plot $\widehat{R}$ values obtained when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the centered model for both the split and unsplit cases were, for all but a single parameter, above 1.01, echoing the results for $R^*$; also, like $R^*$, $\widehat{R}$ is higher for the split case.
+These show that the values of $\widehat{R}$ for the centered model using the unsplit chains were below 1.01; when using the chains split into two halves, $\widehat{R}>1.01$ for all but a single parameter.
 All parameters for the non-centered models were below 1.01, indicating convergence.
@@
 To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
-After fitting the GBM model, we group MCMC draws into deciles and draw from the $R^*$ distribution for each decile.
+After fitting the GBM model, we group MCMC draws in the test set into deciles and draw from the $R^*$ distribution for each decile.
 In Fig.
@@
 
+In \S\ref{sec:discrete}, we evaluate $R^*$ on univariate discrete examples: one with four states (in \S\ref{sec:discrete_small}); another, with a larger state-space consisting of 20 states (in \S\ref{sec:discrete_large}).
+In these examples, we use a discrete Markov model to generate draws from a given target.
+The small and larger state-space cases, show that $R^*$ behaves as expected: given a sufficient sample size, it is able to detect differences in the transition probability matrix between chains that result in differences in the target distribution.
+
+In \S\ref{sec:ML_sensitivity}, we investigate how two decisions about classifiers — which classifier to use (in \S\ref{sec:ml_model}) and what hyperparameters to use for it (in \S\ref{sec:hyperparameters}) — affect calculation of $R^*$.
+In \S\ref{sec:ml_model}, we test a range of popular classifiers: GBMs, RFs, k-nearest-neighbour models, support vector machines and generalised linear models across examples.
+This indicated that GBMs and RFs consistently had the highest classification accuracy across the examples: in higher dimensional problems, RFs tended to best GBMs.
+In \S\ref{sec:hyperparameters}, we show how these two best classifiers — GBMs and RFs — depend on their hyperparameters.
+Across the examples we test, GBMs are more sensitive to hyperparameter choices than are RFs.
+Additionally, GBMs require typically more rounds of boosting in higher dimensional problems, leading to more extensive algorithm runtime.
+Because of this, we suggest fixing GBM's hyperparameters to values that result in reasonable performance and reasonable runtime.
+For RFs, we suggest using a heuristic for choosing its hyperparameters that was derived in a previous empirical evaluation of RFs \citep{bernard2009influence}.
+
+In \S\ref{sec:comparison_gbm_rf}, we use a slew of examples to compare $R^*$ calculated using GBM and RF classifiers using our suggested default hyperparameter sets (given in \S\ref{sec:method}).
+In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution (using a multivariate normal target).
+In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains (using Student-t targets).
+In both examples, draws are generated by independent sampling meaning that an optimal $R^*$ can be calculated based on the Bayes optimal classifier (see, for example, \citep{devroye2013probabilistic} and \S\ref{sec:comparison_gbm_rf} for more information).
+Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
+The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
+In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
+These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-high ones.
+This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
+Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
+RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
+Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
+
 \section{Discussion}
@@
 Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
-By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
-By extracting ML-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
+By taking the ratio of model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
+By extracting classifier-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
 Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
@@
 Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
-This is not a sleight on existing measures, more that this illustrates the complementarity of $R^*$ to them.
-
-When first starting to develop $R^*$, it was unclear to us whether \textit{any} metric aligned with a supervised ML method would be overly sensitive to its hyperparameters.
-With gradient-boosted regression tree models, we found this not to be the case.
-Indeed, after a little experimenting, we found that a default set of hyperparameters (which we report in \S\ref{sec:method}) sufficed across all our examples: perhaps, because the classification boundaries for this task are unlikely to be very complex compared to (say) machine vision tasks.
-To ensure maximal predictive capability of any fitted ML model, however, it is prudent to optimise it over choice of hyperparameters, and further work is needed to determine the importance of this step.
+This indicates the complementarity of $R^*$ to existing measures.
+
+Different target distributions present different challenges to sampling.
+Because of this, there is not a unique optimal ML classifier across all cases: this is just a manifestation of the no free lunch theorem \citep{wolpert1997no}.
+Across a range of examples we tested (see \S\ref{sec:ml_model}), GBM and RF classifiers performed consistently well, and we then used these across all other illustrative examples.
+It is possible — indeed, likely — that another ML model may exist or be invented that consistently outperforms both these classifiers.
+We do not see this as a problem: across the examples we considered, $R^*$ calculated using both classifiers tended to provide a measure of convergence as or more stringent than existing diagnostics.
+In that sense, it is a step in the right direction.
+If a better classifier is found, the same apparatus we develop here can be used, and this will present a harsher test of convergence.
+
+A different question is, ``When is such a measure of convergence of practical use?''.
+In our examples, $R^*$ is able to diagnose poor convergence in the tails of marginal distributions: likely of practical relevance for many applications that require tail quantiles.
+$R^*$ is also able to diagnose lack of convergence in the joint distribution — even if the marginals appear converged.
+Indeed, it is less clear how the joint distribution can be unconverged when the marginals appear so, but we found examples where this appeared true.
+A consequence of poor convergence of the joint distribution would be for prediction and, by corollary, model comparison.
+Further work examining these consequences further is needed, and $R^*$ can help to identify and monitor fruitful candidate systems.
 
 In \S\ref{sec:prostate}, we fit Bayesian models with many 1000s of parameters then used $R^*$ to diagnose convergence, finding that $R^*$ was considerably more expensive to calculate than $\widehat{R}$.
-The computational cost of training GBMs has been suggested to be $\mathcal{O}(N k)$, where $N$ is the number of training data points and $k$ is the number of dimensions of the target \citep{complexity2018}.
-If so, this suggests that larger models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
+The time complexity of training RFs is thought to be $\mathcal{O}(m_{\text{try}} n_{\text{tree}} N \text{log} N)$ \cite[chapter~5]{louppe2014understanding}, and given the similarities with GBMs (which also builds many decision trees), it is likely to be similar.
+If so, this suggests that larger statistical models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
 That said, it is possible to reduce the runtime for $R^*$ using thinned draws (although this risks losing chain idiosyncracies) and using a subset of dimensions (although this risks losing problematic dimensions).
@@
 We recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
-Additionally, our ML calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
+Additionally, our non-parametric calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
 If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.
@@
 
-\section{Contributions}
-BL conceived of the original idea for $R^*$, carried out all analyses and wrote the original draft of the paper.
-AV reviewed the paper and made a host of recommendations: including suggesting a large range of new case studies on which to trial the method and also suggesting new visualisations for diagnosing how predictive performance depends on parameter quantiles; these collectively widened the scope of the original paper and substantially improved its quality.
-BL and AV reviewed and edited the draft of the paper.
-
-\section{Acknowledgements}
-The authors would like to thank Andrew Gelman and Hanne Oberman for comments on previous drafts of the paper that were submitted to arXiv.
-
-\bibliographystyle{chicago}
-\bibliography{bibliography}
-
-\beginsupplement
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	%% Supplementary Material, if any, should   %%
+	%% be provided in {supplement} environment  %%
+	%% with title and short description.        %%
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	\begin{supplement}
+		\stitle{Supplementary materials: $R^*$ convergence diagnostic}
+		\sdescription{Further experiments using $R^*$}
+
+		\section{Diagnosing convergence in joint distributions: multivariate normal models supplementary}\label{sec:multivariate_normal_supp}
+
+		In Fig.
+\ref{fig:mvt_gbm_vs_rf}, we compare the performance of GBM and RF classifiers on the multivariate normal example described in \S\ref{sec:multivariate_normal_250}.
+The two classifiers produced similar $R^*$ distributions across the various examples: for the 400 draw example, the GBM $R^*$ distribution had a mean of 2.86 and the RF equivalent was 3.06; for the example with 10,000 samples from the centered parameterisation, the GBM classifier had a $R^*$ distribution mean of 1.07 versus 1.08 from the RF; for the example with 10,000 samples from the non-centered parameterisation, both classifiers produced a mean of 1.00.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1\textwidth]{mvt_gbm_vs_rf.pdf}}
+			\caption{\textbf{Multivariate normal example: GBM (A) versus RF (B) classifiers.} Shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws).
+Here, we show 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} for each MCMC run.}
+			\label{fig:mvt_gbm_vs_rf}
+		\end{figure}
+
+		\section{Hierarchical model: Eight schools model supplementary}\label{sec:8_schools_supp}
+
+		Fig. \ref{fig:eight_schools_rf} shows the equivalent of Fig. \ref{fig:eight_schools} except using a RF classifier.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1\textwidth]{eight_schools_rf.pdf}}
+			\caption{\textbf{Eight schools example: $R^*$ distributions for a RF classifier.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
+The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
+In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} for each parameterisation.}
+			\label{fig:eight_schools_rf}
+		\end{figure}
 
@@
 \eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
-Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig.
-\ref{fig:mvt_wide_both}A.
-In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence.
+Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide''.
+We also calculated $R^*$ distributions using both the GBM and RF classifiers: the $R^*$ distribution in each case is shown in Figs.
+\ref{fig:mvt_wide_both}A\&B.
+This figure shows that, across both parameterisations, convergence has not yet been reached since all $R^*$ distributions are shifted rightwards of 1.
 The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
@@
 			\centerline{\includegraphics[width=1\textwidth]{mvt_wide_both.pdf}}
-	\caption{\textbf{Wide data examples.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; B shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
+			\caption{\textbf{Wide data examples.} Top row shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; bottom row shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
+In the left column, results for the GBM classifier are shown; in the right, we show the same for the RF classifier.
 In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.}
@@
 In this case, we assume independent standard normals for each dimension.
-In Fig.
-\ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
-In both cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by rank-normalised split-$\widehat{R}$, with 65\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 19\% for the 1000 iteration case.
+In Figs.
+\ref{fig:mvt_wide_both}C\&D, we plot the $R^*$ distributions from GBM and RF classifiers for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
+In all cases, the distributions were right of $R^*=1$, indicating non-convergence.
+These results were also echoed by rank-normalised split-$\widehat{R}$, with 19\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
 
@@
 In this section, we investigate whether $R^*$ can detect non-stationary sampling distributions.
+In all the examples, we present results for both GBM and RF classifiers.
 
@@
 The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
+The points and triangles show $R^*$ derived from GBM and RF classifiers, respectively.
 This plot mirrors Fig.
@@
 			\centerline{\includegraphics[width=1.0\textwidth]{trends_all_dim.pdf}}
-	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
+			\caption{\textbf{Univariate trends example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
 The rows present the differing sample sizes.
@@
 In the simulations here, all dimensions bar one are stationary; the remaining dimension has a linear trend added to it.
-In all cases, split-$R^*$ increased with trend.
+In all cases, across both GBM and RF classifiers, split-$R^*$ increased with trend.
 Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered.
@@
 			\centerline{\includegraphics[width=1.0\textwidth]{trends_one_dim.pdf}}
-	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
+			\caption{\textbf{Multivariate trends example with split-$R^*$.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
 The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
@@
 		Means that trend over time is one form of non-stationarity; another is a time-varying covariance.
-Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time.
+Next, we consider a bivariate normal with (constant) standard normal marginals but where the correlation between dimensions trends over time.
 Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
@@
 			\centerline{\includegraphics[width=1.0\textwidth]{trends_joint_distribution.pdf}}
-	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
+			\caption{\textbf{Bivariate normal with trending correlation example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
 The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
@@
 			\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example.} The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
+			\caption{\textbf{Non-stationary distribution: AR(1) example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
+The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
 At each parameter set, 10 replicates were run and jitter was added to the points.
@@
 			\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1_transposed.pdf}}
-	\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
+			\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} Column A shows the results for a GBM classifier; column B for a RF classifier.
+The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
 At each parameter set, 10 replicates were run and jitter was added to the points.
@@
 		In this section, we analyse two Bayesian models both fit to real data.
+In both examples, we illustrate $R^*$ distributions derived only from GBM classifiers.
 The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
@@
 Whilst it is harder to discern, this pattern is mirrored in Fig.
-\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 139 parameters where this was not the case.
+\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 83 parameters where this was not the case.
 Similarly, in Figs.
@@
 			\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) using the GBM classifier are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
 			\label{fig:ovarian}
@@
 			\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
+In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) using the GBM classifier are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
 			\label{fig:prostate}
@@
 
+		\section{Discrete distributions}\label{sec:discrete}
+		In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a univariate discrete target distribution.
+To do so, we generate draws from four discrete Markov chains, following a 1st order Markov process with transition probabilities,
+		%
+		\begin{equation}
+		p_{ij} = \text{Pr}(X_{n+1}=j|X_n=i),
+		\end{equation}
+		%
+		where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$.
+The target distribution is, hence, the stationary distribution of this process.
+
+		\subsection{Small state-space}\label{sec:discrete_small}
+		We first consider a univariate discrete distribution with four states. In three of the four chains, we generate draws using a transition probability matrix:
+		%
+		\begin{gather}
+		\boldsymbol{P}
+		=
+		\begin{bmatrix}
+		0 & 1/2 & 1/2 & 0\\
+		1/2 & 0 & 1/3 & 1/6\\
+		1/4 & 1/4 & 1/4 & 1/4\\
+		0 & 1 & 0 & 0
+		\end{bmatrix},
+		\end{gather}
+		%
+		which implies a stationary target distribution with the following probability simplex $\pi_1 = (11/46, 15/46, 14/46, 6/46)$.
+
+		For the fourth chain, we specify differing transition probability matrices in each of three sets of results.
+In the first, we assume $\boldsymbol{P}_1=\boldsymbol{P}$ (i.e.
+the same as used for the other three chains); in the second and third cases, we assume this transition matrix is given by:
+		%
+		\begin{gather}
+		\boldsymbol{P}_2
+		=
+		\begin{bmatrix}
+		0 & 1/2 & 1/2 & 0\\
+		1/2 & 0 & 1/3 & 1/6\\
+		5/8 & 1/8 & 1/8 & 1/8\\
+		1/2 & 1/2 & 0 & 0
+		\end{bmatrix},\quad
+		\boldsymbol{P}_3
+		=
+		\begin{bmatrix}
+		0 & 1/2 & 1/2 & 0\\
+		1/2 & 0 & 1/3 & 1/6\\
+		1 & 0 & 0 & 0\\
+		1 & 0 & 0 & 0
+		\end{bmatrix}.
+		\end{gather}
+		%
+		These correspond to stationary probability simplices:
+		%
+		\begin{equation}
+		\pi_2 = (71/198, 17/66, 10/33, 8/99), \quad \pi_3 = (4/9, 2/9, 8/27, 1/27).
+		\end{equation}
+		%
+		The three stationary distributions corresponding to each transition probability matrix are shown in Fig.
+\ref{fig:discrete_stationary}.
+This illustrates that there is an ordering among the stationary distributions $\pi_1\sim\pi_2\sim\pi_3$, meaning $\pi_1$ is most similar to $\pi_2$, and $\pi_3$ is most similar to $\pi_2$.
+Put another way, $\pi_3$ is most different from $\pi_1$.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary.pdf}}
+			\caption{\textbf{Discrete distribution example: stationary distributions.} The three colours illustrate the stationary distributions of the Markov chains corresponding to the matrices $P_1$, $P_2$ and $P_3$ defined in \S\ref{sec:discrete}.}
+			\label{fig:discrete_stationary}
+		\end{figure}
+
+		For each of these three cases, we consider a range of sample sizes across each of the four chains: 1000, 2000, 5000 and 10,000.
+At each combination of transition probability matrix (for the fourth chain) and sample size, we perform 40 replicates.
+In each replicate, we calculate both $R^*$ using a GBM classifier and rank-normalised split-$\widehat{R}$.
+The results of these experiments are shown in Fig.
+\ref{fig:discrete}.
+In the horizontal axes of both panels, we show the sample size.
+The vertical axis indicates the value of $R^*$ and split-$\widehat{R}$ in panels A and B, respectively.
+The stacked sub-panels within A and B indicate the transition matrix used for the fourth chain.
+Since $\boldsymbol{P}_1=\boldsymbol{P}$, the stationary distribution for the other three chains, both diagnostics should tend towards indicating convergence in these cases, and, for both statistics, this was typically the case.
+For $R^*$, the replicates were centred on $R^*=1$, although the stochasticity of replicates meant that, in some cases, $R^*>1$.
+Whereas, $\widehat{R}$ was similarly close to 1 in all cases.
+Considering the simulations using $\boldsymbol{P}_2$ for the fourth chain transition probability matrix, $R^*>1$ across a majority of replicates, and this majority increased with sample size.
+In this case, $1.01>\widehat{R}>1$ for all replicates across the various sample sizes.
+For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; in this case, $\widehat{R}>1.01$ in all replicates.
+
+		This example shows that $R^*$ can detect poor convergence for discrete parameter spaces, like $\widehat{R}$.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{discrete.pdf}}
+			\caption{\textbf{Discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations.
+The panels within A.
+and B.
+indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$).
+In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates.
+In panel B, we show the same, but for split-$\widehat{R}$.
+The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
+			\label{fig:discrete}
+		\end{figure}
+
+		\subsection{Large state-space}\label{sec:discrete_large}
+		We next consider a larger discrete parameter space: a univariate distribution with 20 states.
+As in \S\ref{sec:discrete_small}, we generate draws from three discrete Markov chains that share a common transition probability matrix, $\boldsymbol{P}$.
+For the fourth (and final) chain, we consider two cases: one where it shares the same matrix, $\boldsymbol{P}_1=\boldsymbol{P}$; the other, when its Markov process has a different matrix, $\boldsymbol{P}_2$.
+Rather than write down a given $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we generated these stochastically: for each of the rows, we randomly sampled from from a Dirichlet process with shape vectors consisting of ones (of length 20).
+
+		To illustrate the differences between the two transition probability matrices, $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we simulate 100,000 draws from each Markov process.
+In Fig.
+\ref{fig:discrete_stationary_higherd}, we show the marginal distribution obtained across all four chains: where the first three used $\boldsymbol{P}$, and the last used $\boldsymbol{P}_2$.
+In this plot, we show only the first four states of the 20-state model.
+The differences in marginal distributions between the first three chains and the last indicate that there are substantial differences in the target distributions.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary_higherd.pdf}}
+			\caption{\textbf{Large state-space discrete distribution example: approximate stationary distribution.} The distribution over the first four states of a 20-state discrete model achieved after drawing 100,000 samples from each of four Markov chains.
+Here, the first three chains shared the same transition probability matrix; the last chain used a different matrix.}
+			\label{fig:discrete_stationary_higherd}
+		\end{figure}
+
+		To probe the ability of $R^*$ and $\widehat{R}$ to detect poor convergence, we repeat the same exercise as in \S\ref{sec:discrete_small} although, here, we consider the new $\boldsymbol{P}_1$ and $\boldsymbol{P}_2$ cases.
+In Fig.
+\ref{fig:discrete_higherd}, we show the results of this analysis.
+This shows that $R^*\sim 1$ for the $\boldsymbol{P}_1$ case, correctly indicating no convergence issues; for the $\boldsymbol{P}_2$ case, $R^*>1$ across all replicates at sample sizes of 5000 and above.
+By contrast, $\widehat{R}$ struggles to differentiate between the $\boldsymbol{P}_1$ and $\boldsymbol{P}_2$ cases: in both instances, it signifies convergence.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{discrete_higherd.pdf}}
+			\caption{\textbf{Large state-space discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations.
+The panels within A.
+and B.
+indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$).
+In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates.
+In panel B, we show the same, but for split-$\widehat{R}$.
+The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
+			\label{fig:discrete_higherd}
+		\end{figure}
+
+		Collectively, the results of \S\ref{sec:discrete_small} and \S\ref{sec:discrete_large} show that $R^*$ is able to detect convergence issues in discrete target distributions.
+The results of \S\ref{sec:discrete_large} hint that $R^*$ may be superior than $\widehat{R}$ in larger discrete spaces, although further empirical investigation is needed.
+
+		\section{ML sensitivity}\label{sec:ML_sensitivity}
+		In this section, we investigate how two decisions about classifiers — which classifier to use (in \S\ref{sec:ml_model}) and what hyperparameters to use for it (in \S\ref{sec:hyperparameters}) — affect calculation of $R^*$.
+The test cases used for empirical evaluation were selected from our pool of examples introduced in the text.
+Specifically, they were chosen to represent cases where $R^*$ was nearer 1, since, for these, small changes in calculated $R^*$ could lead to different decisions about convergence being reached.
+The four examples we used were:
+
+		\begin{itemize}
+			\item The AR(1) example described in \S\ref{sec:heterogeneity}, except using 1000 draws per chain.
+			\item The 250-dimensional non-centered multivariate normal example introduced in \S\ref{sec:multivariate_normal_250}.
+In each replicate, we generated 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm across each of 4 chains.
+			\item The alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy}.
+For each replicate, we generated samples from the model using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
+			\item The non-centered parameterisation of the eight schools model described in \S\ref{sec:eight_shools}.
+In each replicate, we used Stan's NUTS algorithm with adapt delta set to 0.95 to generate 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains; these were then used to calculate $R^*$.
+		\end{itemize}
+
+		\subsection{ML classifier comparison}\label{sec:ml_model}
+		In this section, we investigate how classification accuracy depends on choice of ML classifier.
+We restricted our analysis to popular classifiers with relatively few hyperparameters to simplify comparison amongst them.
+Notably, we do not consider neural network models, since the structure of nets effectively defines a high dimensional hyperparameter space.
+
+		The ML classifiers we compared were GBMs, RFs, k-nearest neighbours (KNNs), SVMs with linear kernels and a generalised linear model approach (GLM).
+All these models were called through the \textbf{\textsf{R}}'s Caret package \citep{kuhn2008building}: the GBM is implemented using the \textit{gbm} model in Caret which uses the ``gbm'' package \citep{greenwell2019package}; RFs are implemented using the \textit{rf} model in Caret which uses the ``randomForest'' \citep{liaw2002classification} package; KNN is natively implemented in Caret and called using the \textit{knn} model; SVMs are implemented in Caret using the \textit{svmLinear} model which uses ``kernlab'' package \citep{karatzoglou2004kernlab}; and the GLM is implemented through the \textit{multinom} model in the ``nnet'' package \citep{ripley2016package}.
+
+		Each of these classifiers has hyperparameters, and, due to the differences between the examples we consider (defined in \S\ref{sec:ML_sensitivity}), the optimal hyperparameters were likely to differ between examples.
+For each classifier-example combination, we first performed a single replicate of each experiment and choose those hyperparameters which maximised classification accuracy.
+The set of all hyperparameters we selected between were chosen so that training time for each classifier remained reasonable.
+These sets were the same across all examples:
+
+		\begin{itemize}
+			\item GBM: Cartesian product of $\text{int.depth}=(3, 7)$; $\text{\# trees} = (50, 100)$;
+			\item RF: $m_{\text{try}} = (1, 2)$.
+(Note, in \S\ref{sec:hyperparameters}, we investigate a dimension-specific approach to $m_{\text{try}}$ but, here, for ease of comparison with the other methods, we keep this static.)
+			\item KNN: $K=(5,10,15, 20,40)$;
+			\item SVM: $C=(0.25, 0.5, 0.75)$;
+			\item GLM: $\text{decay}=(0.1, 0.2, 0.5, 1)$.
+		\end{itemize}
+
+		These hyperparameters were then used in subsequent replicates.
+We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 100 replicates; for the multivariate normal example, we used 100; for the Cauchy model, 50 replicates; and for the eight schools model, 30 replicates.
+
+		The results of these experiments are shown in Fig.
+\ref{fig:ml_comparison_all}.
+The top row shows $R^*$ as calculated by each classifier (in each case, chains split into two halves); the bottom row shows the time (in seconds) taken for $R^*$ calculation.
+The panels correspond to the examples described in \S\ref{sec:ML_sensitivity}.
+Across the examples, the classification accuracy of RFs and GBMs were, generally, highest; followed by KNNs.
+An exception was for the 250-dimensional normal where KNN performed best.
+In the AR(1) example, GBM outperformed RF.
+In higher dimensional cases, RF outperformed GBM.
+In all cases, RF and GBM classifiers took the greatest time to train; in higher dimensional examples, RF-based $R^*$ were cheaper to calculate than those from GBMs.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{ml_comparison_all.pdf}}
+			\caption{\textbf{ML classifier comparison.} Top row shows $R^*$ values calculated using Algorithm \ref{alg:R_star}; bottom row show the time taken to run each method on a desktop computer.
+The different columns correspond to the examples described in \S\ref{sec:ML_sensitivity}.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates; note, that the number of replicates varies across examples as given in \S\ref{sec:ml_model}.}
+			\label{fig:ml_comparison_all}
+		\end{figure}
+
+		\subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
+		In this section, we considered only the two classifiers — GBMs and RFs — that performed best on our comparison between ML methods in \S\ref{sec:ml_model}.
+The performance of GBMs and RFs, like most ML methods, depends on their hyperparameters.
+In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree.
+The number of such regression trees is thus a hyperparameter of the model.
+Each regression tree can split amongst any integer number of variables (the \textit{interaction depth}).
+For GBMs, we analyse how their performance depends on these two hyperparameters.
+For RFs, there is a single hyperpameter: $m_{\text{try}}$, the number of members of the subset of all features over which to search for an optimal split when forming a decision tree.
+Note, that the maximum allowed value of $m_{\text{try}}=K$ — the dimensionality of the target distribution.
+
+		In this section, we investigate the sensitivity of $R^*$ calculated by GBMs and RFs to each of these sets of hyperparameters across a range of examples (defined in \S\ref{sec:ML_sensitivity}).
+Since each example is of different dimensionalities, we compare different sets of hyperparameters in each, although include amongst the sets the default values we suggest for each classifier in \S\ref{sec:method} (these are shown as triangles on the plots).
+Note, that for the GBM model, we did not investigate a full range of hyperparameters needed to optimise $R^*$ for many of the examples due to the extensive training time needed to do so.
+For RFs, runtime was less restrictive and we were better able to survey the sensitivity across hyperparameter space.
+
+		We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 200 replicates; for the multivariate normal example, we used 20; for the Cauchy model, 20 replicates; and for the eight schools model, 50 replicates.
+
+		\subsubsection{Autoregressive example}\label{sec:hyperparameters_ar1}
+		In Fig.
+\ref{fig:hypers_ar1}, we show the results of the AR(1) analysis.
+In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
+In panel B, we show the same but for RFs.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
+For the GBM model, $R^*$ varied according to hyperparameters: of the set investigated, an interaction depth of 1 appears optimal with 10+ trees.
+In panel B, the values of $R^*$ are generally lower than those achieved by GBMs.
+As shown in \S\ref{sec:comparison_gbm_rf}, RFs appear less well-suited to lower dimensional problems.
+Across the two possible hyperparameter values for RFs, there was not a substantial difference in the values of $R^*$ calculated.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{hypers_ar1.pdf}}
+			\caption{\textbf{Hyperparameter sensitivity: AR(1) example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the AR(1) example described in \S\ref{sec:hyperparameters_ar1}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
+			\label{fig:hypers_ar1}
+		\end{figure}
+
+		\subsubsection{Multivariate normal: 250-dimensional model}\label{sec:hyperparameters_normal}
+		In Fig.
+\ref{fig:hypers_normal}, we show the results of the analysis for the 250-dimensional normal example.
+In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
+In panel B, we show the same but for RFs.
+This shows that $R^*$ was higher as calculated by RFs and that performance of GBMs depended strongly on hyperparameters.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{hypers_normal_250.pdf}}
+			\caption{\textbf{Hyperparameter sensitivity: 250-dimensional normal example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the Cauchy example described in \S\ref{sec:hyperparameters_normal}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
+			\label{fig:hypers_normal}
+		\end{figure}
+
+		\subsubsection{Cauchy model: alternative parameterisation}\label{sec:hyperparameters_cauchy}
+		In Fig.
+\ref{fig:hypers_cauchy}, we show the results of the Cauchy model analysis.
+In panel A, we show $R^*$ as calculated by a GBM classifier; in panel B, the same but using a RF classifier.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
+Across the range of hyperparameters investigated, the RF classifier outperformed the GBM and was less sensitive to variation in its hyperparameters.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{hypers_cauchy.pdf}}
+			\caption{\textbf{Hyperparameter sensitivity: Cauchy example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the Cauchy example described in \S\ref{sec:hyperparameters_cauchy}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
+			\label{fig:hypers_cauchy}
+		\end{figure}
+
+		\subsubsection{Eight schools: non-centered parameterisation}\label{sec:hyperparameters_8_schools}
+		In Fig.
+\ref{fig:hypers_8_schools}, we show the results of the experiments on the eight schools model.
+Panel A shows how $R^*$ calculated using a GBM varies with its two hyperparameters; panel B shows the same but for RFs.
+This shows that, in this case, the value of $R^*$ calculated by RFs was higher than that for GBMs.
+Further, whereas GBMs were relatively sensitive to their hyperparameter values, RFs were less so.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1.0\textwidth]{hypers_8_schools.pdf}}
+			\caption{\textbf{Hyperparameter sensitivity: eight schools example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the eight schools example described in \S\ref{sec:hyperparameters_8_schools}; Panel B shows the same but for a RF classifier.
+Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
+In Panel A, the coloured lines shows the different interaction depths investigated.
+In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
+			\label{fig:hypers_8_schools}
+		\end{figure}
+
+		\section{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
+		In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
+In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
+In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}.
+In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution.
+In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains.
+Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}.
+This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw.
+Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.
+
+		\subsection{Joint distribution}\label{sec:joint_distribution}
+		Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32.
+For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
+To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
+To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$.
+For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
+In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.
+
+		The results of these experiments are shown in Fig.
+\ref{fig:gbm_vs_rf_normal}.
+In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic.
+When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$.
+In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence.
+As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows.
+$R^*$ as calculated by GBMs and RFs also increases.
+For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification.
+For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_normal.pdf}}
+			\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
+For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
+The dashed line shows $R^*=1$.
+Here, $R^*$ was calculated using chains split into two halves.}
+			\label{fig:gbm_vs_rf_normal}
+		\end{figure}
+
+		\subsection{Fat tails}\label{sec:tail_fatness}
+		We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution.
+To do so, we again consider four chains.
+In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom.
+In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
+We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$.
+We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness.
+Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$.
+To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$.
+For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
+Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.
+
+		The results of these experiments are shown in Fig.
+\ref{fig:gbm_vs_rf_studdentt}.
+Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions.
+Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$.
+The point colour and shadings give the method used to calculate $R^*$.
+In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method.
+As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases.
+In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers.
+In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate.
+If the number of dimensions is 16 or greater, the order switches, and RFs often outperform GBMs.
+
+		\begin{figure}[!htb]
+			\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
+			\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution.
+The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
+The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
+For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
+The dashed line shows $R^*=1$.
+Here, $R^*$ was calculated using chains split into two halves.}
+			\label{fig:gbm_vs_rf_studdentt}
+		\end{figure}
+
+	\end{supplement}
+
+	\bibliographystyle{ba}
+	\bibliography{bibliography}
+
+	\begin{acknowledgement}
+		The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.
+We would also like to thank Paul B{\"u}rkner and Jonah Gabry, with whom useful discussions were had during the preparation of this manuscript.
+	\end{acknowledgement}
+
 \end{document}

### 2020-11-19 write_up/arxiv_submission_aug_2020/ms_arxiv_final_preprint.tex
added submission to arxiv

--- 2020-11-18 write_up/ba_final_submission/ms_final.tex
@@
-\documentclass[ba]{imsart}
-%
-\pubyear{0000}
-\volume{00}
-\issue{0}
-\doi{0000}
-\arxiv{2003.07900}
-\firstpage{1}
-\lastpage{1}
-
-%
-\usepackage{amsthm}
+\documentclass{article}
+\pdfoutput=1
+% if you need to pass options to natbib, use, e.g.:
+%     \PassOptionsToPackage{numbers, compress}{natbib}
+% before loading neurips_2018
+
+% ready for submission
+% \usepackage{neurips_2018}
+
+% to compile a preprint version, e.g., for submission to arXiv, add add the
+% [preprint] option:
+% \usepackage[preprint]{neurips_2019}
+
+% to compile a camera-ready version, add the [final] option, e.g.:
+%\usepackage{neurips_2019}
+
+% to avoid loading the natbib package, add option nonatbib:
+%     \usepackage[nonatbib]{neurips_2018}
+
+\usepackage{arxiv}
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\hypersetup{unicode=true,
+            pdfborder={0 0 0},
+            breaklinks=true,
+            colorlinks=true,
+            linkcolor=blue,
+            citecolor=blue,
+            filecolor=blue,
+            urlcolor=blue}
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
 \usepackage{amsmath}
-\usepackage{natbib}
-\usepackage[colorlinks,citecolor=blue,urlcolor=blue,filecolor=blue,backref=page]{hyperref}
+\usepackage{algorithm,algorithmic}
 \usepackage{graphicx}
 \usepackage{bbm}
-\usepackage{algorithm,algorithmic}
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{booktabs}
-
-\startlocaldefs
-\numberwithin{equation}{section}
-\theoremstyle{plain}
-\newtheorem{thm}{Theorem}[section]
-\endlocaldefs
+\usepackage{caption} 
+\usepackage{natbib}
+\usepackage{xr}
+\usepackage{xcolor}
+\externaldocument[]{supplementary_materials_marked}
+
+\captionsetup[table]{skip=10pt}
+
+\newcommand{\aki}[1]{\textcolor{red}{[Aki: #1]}}
+
+\newcommand{\fracpartial}[2]{\frac{\partial #1}{\partial  #2}}
+\newcommand{\Dir}[0]{\textrm{Dirichlet}}
+\newcommand{\Ray}[0]{\textrm{Rayleigh}}
+\newcommand{\gam}[0]{\textrm{Gamma}}
+\newcommand{\dgamma}[0]{\textrm{Gamma}}
+\newcommand{\dpoisson}[0]{\textrm{Poisson}}
+\newcommand{\dbeta}[0]{\textrm{Beta}}
+\newcommand{\dbern}[0]{\textrm{Bernoulli}}
+\newcommand{\dunif}[0]{\mathrm{Uniform}}
+\newcommand{\dgig}[0]{\textrm{GIG}}
+\newcommand{\dnormal}[0]{\mathrm{Normal}}
+\newcommand{\dt}[0]{\mathrm{t}}
+\newcommand{\igamma}[0]{\textrm{Gamma}^{-1}}
+\newcommand{\rayl}[0]{\textrm{Rayleigh}}
+\newcommand{\Exp}[0]{\textrm{Exponential}}
+\newcommand{\Bet}[0]{\textrm{Beta}}
+\newcommand{\GEM}[0]{\textrm{GEM}}
+\newcommand{\DP}[0]{\textrm{DP}}
+\newcommand{\ESS}[0]{\mathrm{ESS}}
+\newcommand{\bm}[1]{\boldsymbol{#1}}
+\newcommand{\bbeta}{\bm{\beta}}
+\newcommand{\bpi}{\bm{\pi}}
+\newcommand{\bomega}{\bm{\omega}}
+\newcommand{\bgamma}{\bm{\gamma}}
+\newcommand{\blambda}{\bm{\lambda}}
+\newcommand{\bphi}{\bm{\phi}}
+\newcommand{\btheta}{\bm{\theta}}
+\newcommand{\bmu}{\bm{\mu}}
+\newcommand{\bb}{\bm{b}}
+\newcommand{\bk}{\bm{k}}
+\newcommand{\bl}{\bm{l}}
+\newcommand{\bn}{\bm{n}}
+\newcommand{\bw}{\bm{w}}
+\newcommand{\bz}{\bm{z}}
+\newcommand{\bx}{\bm{x}}
+\newcommand{\bX}{\bm{X}}
+\newcommand{\by}{\bm{y}}
+\newcommand{\bZ}{\bm{Z}}
+\newcommand{\bW}{\bm{W}}
+\newcommand{\bS}{\bm{S}}
+\newcommand{\bH}{\bm{H}}
+\newcommand{\Mult}{\textrm{Multinomial}}
+\newcommand{\N}{\mathcal{N}}
+\newcommand{\NEW}{\textrm{\tiny new}}
+\newcommand{\OLD}{\textrm{\tiny old}}
+\newcommand{\sigmat}{\sigma^2}
+\newcommand{\IBP}{\textrm{IBP}}
+\newcommand{\E}{\mathbb{E}}
+\newcommand{\V}{\mathbb{V}}
+\newcommand{\Eq}{\mathbb{E}_q}
+\newcommand{\cL}{\mathcal{L}}
+\newcommand{\cB}{\mathcal{B}}
+\newcommand{\cC}{\mathcal{C}}
+\newcommand{\cV}{\mathcal{V}}
+\newcommand{\cHq}{\mathcal{H}_q}
+\newcommand{\test}[1]{\mbox{$#1$}^{\small \mbox{test}}}
+\newcommand{\alphaW}{\alpha^{(W)}}
+\newcommand{\alphaH}{\alpha^{(H)}}
+\newcommand{\betaW}{\beta^{(W)}}
+\newcommand{\betaH}{\beta^{(H)}}
+\newcommand{\gammaW}{\gamma^{(W)}}
+\newcommand{\gammaH}{\gamma^{(H)}}
+\newcommand{\gammaT}{\gamma^{(\theta)}}
+\newcommand{\rhoW}{\rho^{(W)}}
+\newcommand{\rhoH}{\rho^{(H)}}
+\newcommand{\rhoT}{\rho^{(\theta)}}
+\newcommand{\tauW}{\tau^{(W)}}
+\newcommand{\tauH}{\tau^{(H)}}
+\newcommand{\tauT}{\tau^{(\theta)}}
+\newcommand{\muW}{\hat{W}}
+\newcommand{\muH}{\hat{H}}
+\newcommand{\Var}{\textrm{Var}}
+\newcommand{\LF}{\mathrm{Leapfrog}}
+
+\newcommand{\beginsupplement}{%
+	\setcounter{table}{0}
+	\renewcommand{\thetable}{S\arabic{table}}%
+	\setcounter{figure}{0}
+	\renewcommand{\thefigure}{S\arabic{figure}}%
+	\setcounter{section}{0}
+	\renewcommand{\thesection}{S\arabic{section}}%
+}
+
+\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers}
+
+% The \author macro works with any number of authors. There are two commands
+% used to separate the names and addresses of multiple authors: \And and \AND.
+%
+% Using \And between authors leaves it to LaTeX to determine where to break the
+% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
+% authors names on the first line, and the last on the second line, try using
+% \AND instead of \And before the third author name.
+
+\author{%
+	 Ben Lambert\\
+	 MRC Centre for Global Infectious Disease Analysis\\
+	 School of Public Health\\
+	 Imperial College London\\
+	 W2 1PG, United Kingdom\\
+	 \texttt{ben.c.lambert@gmail.com} \\
+	 \And
+	 Aki Vehtari \\
+	 Department of Computer Science\\
+	 Aalto University\\
+	 Finland\\
+	 \texttt{aki.vehtari@aalto.fi}
+}
 
 \begin{document}
-
-	\begin{frontmatter}
-		\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers}
-		\runtitle{A MCMC convergence diagnostic using decision tree classifiers}
-
-		\begin{aug}
-			\author{\fnms{Ben} \snm{Lambert}\thanksref{addr1}\ead[label=e1]{ben.c.lambert@gmail.com}}
-			\and
-			\author{\fnms{Aki} \snm{Vehtari}\thanksref{addr2}%
-				\ead[label=e3]{aki.vehtari@aalto.fi}%
-				\ead[label=u1,url]{http://www.foo.com}}
-
-			\runauthor{Lambert and Vehtari}
-
-			\address[addr1]{MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London,W2 1PG, United Kingdom
-				\printead{e1} % print email address of "e1"
-			}
-
-			\address[addr2]{Department of Computer Science, Aalto University, Finland
-				\printead{e3}
-			}
-
-		\end{aug}
+% \nipsfinalcopy is no longer used
+
+\maketitle
 
@@
 \end{abstract}
-
-	\end{frontmatter}
 
@@
 
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	%% Supplementary Material, if any, should   %%
-	%% be provided in {supplement} environment  %%
-	%% with title and short description.        %%
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\begin{supplement}
-		\stitle{Supplementary materials: $R^*$ convergence diagnostic}
-		\sdescription{Further experiments using $R^*$}
-
+	\section{Contributions}
+	BL conceived of the original idea for $R^*$, carried out all analyses and wrote the original draft of the paper.
+AV reviewed the paper and made a host of recommendations: including suggesting a large range of new case studies on which to trial the method and also suggesting new visualisations for diagnosing how predictive performance depends on parameter quantiles; these collectively widened the scope of the original paper and substantially improved its quality.
+BL and AV reviewed and edited the draft of the paper.
+
+	\section{Acknowledgements}
+	The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.
+We would also like to thank Paul B{\"u}rkner and Jonah Gabry, with whom useful discussions were had during the preparation of this manuscript.
+
+	\bibliographystyle{chicago}
+	\bibliography{bibliography}
+
+	\beginsupplement
 	\section{Diagnosing convergence in joint distributions: multivariate normal models supplementary}\label{sec:multivariate_normal_supp}
@@
 	%
-		where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$.
-The target distribution is, hence, the stationary distribution of this process.
+	where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$. The target distribution is, hence, the stationary distribution of this process.
 
@@
 
-	\end{supplement}
-
-	\bibliographystyle{ba}
-	\bibliography{bibliography}
-
-	\begin{acknowledgement}
-		The authors would like to thank the anonymous reviewers for comments on previous drafts of the paper that lead to significant improvements.
-We would also like to thank Paul B{\"u}rkner and Jonah Gabry, with whom useful discussions were had during the preparation of this manuscript.
-	\end{acknowledgement}
-
 \end{document}

### 2020-11-20 write_up/ba_final_submission/ms_final.tex
added break out of files

@@
 \usepackage{booktabs}
+\usepackage{xr}
+\externaldocument[]{supplementary}
 
@@
 
-		\section{Diagnosing convergence in joint distributions: multivariate normal models supplementary}\label{sec:multivariate_normal_supp}
-
-		In Fig.
-\ref{fig:mvt_gbm_vs_rf}, we compare the performance of GBM and RF classifiers on the multivariate normal example described in \S\ref{sec:multivariate_normal_250}.
-The two classifiers produced similar $R^*$ distributions across the various examples: for the 400 draw example, the GBM $R^*$ distribution had a mean of 2.86 and the RF equivalent was 3.06; for the example with 10,000 samples from the centered parameterisation, the GBM classifier had a $R^*$ distribution mean of 1.07 versus 1.08 from the RF; for the example with 10,000 samples from the non-centered parameterisation, both classifiers produced a mean of 1.00.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1\textwidth]{mvt_gbm_vs_rf.pdf}}
-			\caption{\textbf{Multivariate normal example: GBM (A) versus RF (B) classifiers.} Shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws).
-Here, we show 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} for each MCMC run.}
-			\label{fig:mvt_gbm_vs_rf}
-		\end{figure}
-
-		\section{Hierarchical model: Eight schools model supplementary}\label{sec:8_schools_supp}
-
-		Fig. \ref{fig:eight_schools_rf} shows the equivalent of Fig. \ref{fig:eight_schools} except using a RF classifier.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1\textwidth]{eight_schools_rf.pdf}}
-			\caption{\textbf{Eight schools example: $R^*$ distributions for a RF classifier.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains.
-The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} for each parameterisation.}
-			\label{fig:eight_schools_rf}
-		\end{figure}
-
-		\section{Wide datasets: multivariate normal}\label{sec:wide}
-		As the number of parameter dimensions increases, it might be thought that ML algorithms will overfit the data, and, hence, testing set classification would be poor; leading to unreliable determinations of convergence.
-To test this hypothesis, we investigated two scenarios using a multivariate normal target.
-
-		\subsection{250-dimensional model}
-		In the first of these, we used the 250-dimensional multivariate normal of eq.
-\eqref{eq:mvt_normal_250} with 250 post-warm-up iterations (after 250 warm-up iterations) for each of 4 chains from Stan's NUTS to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}.
-Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide''.
-We also calculated $R^*$ distributions using both the GBM and RF classifiers: the $R^*$ distribution in each case is shown in Figs.
-\ref{fig:mvt_wide_both}A\&B.
-This figure shows that, across both parameterisations, convergence has not yet been reached since all $R^*$ distributions are shifted rightwards of 1.
-The same conclusion is reached if rank-normalised split-$\widehat{R}$ is used instead (Fig.
-\ref{fig:wide_both_diagnostics}A), since, for both parameterisations, some of the parameters had $\widehat{R}>1.01$.
-Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig.
-\ref{fig:wide_both_diagnostics}B\&C).
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1\textwidth]{mvt_wide_both.pdf}}
-			\caption{\textbf{Wide data examples.} Top row shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations per chain from Stan's NUTS algorithm across both model parameterisations; bottom row shows $R^*$ distribution for the 10,000-dimensional example with 400 and 1000 MCMC iterations per chain (although the first half of these were discarded as warm-up).
-In the left column, results for the GBM classifier are shown; in the right, we show the same for the RF classifier.
-In all cases, 1000 draws of $R^*$ are plotted as generated by Algorithm \ref{alg:R_star_uncertainty} for a single MCMC run composed of 4 chains.}
-			\label{fig:mvt_wide_both}
-		\end{figure}
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1\textwidth]{wide_both_diagnostics.pdf}}
-			\caption{\textbf{Wide data 250-dimensional example: established diagnostics.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered.
-Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up draws from Stan's NUTS algorithm.
-Note, that it is possible for the ESS to exceed the actual sample size if there is negative autocorrelation in the Markov chains' values.
-In both cases, the results correspond to a single MCMC run composed of 4 chains.}
-			\label{fig:wide_both_diagnostics}
-		\end{figure}
-
-		\subsection{10,000-dimensional model}
-		We next consider a more challenging example -- a target distribution with 10,000 dimensions.
-In this case, we assume independent standard normals for each dimension.
-In Figs.
-\ref{fig:mvt_wide_both}C\&D, we plot the $R^*$ distributions from GBM and RF classifiers for two MCMC runs targeting this distribution: one with 400 iterations, the other with 1000.
-In all cases, the distributions were right of $R^*=1$, indicating non-convergence.
-These results were also echoed by rank-normalised split-$\widehat{R}$, with 19\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
-
-		Overall, the examples in this section suggest that $R^*$ is a conservative measure of convergence: when there are not enough draws, it will tend to diagnose non-convergence.
-We also note that the statistic took comparable time to calculate relative to existing convergence diagnostics on a desktop computer.
-
-		\section{Non-stationary marginals}\label{sec:non-stationary}
-		If a Markov chain does not mix with itself, this also indicates that convergence has not occurred \citep{gelman2013bayesian}.
-In this section, we investigate whether $R^*$ can detect non-stationary sampling distributions.
-In all the examples, we present results for both GBM and RF classifiers.
-
-		\subsection{Trending mean across all chains}\label{sec:non-stationary_chains}
-		We first recapitulate an example from appendix A in \cite{vehtari2019rank}.
-This example showed that split-$\widehat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean.
-Specifically, they first generated 4 chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling.
-We first repeat this analysis but using $R^*$ rather than $\widehat{R}$: in Fig.
-\ref{fig:trends_all_dim}, we show these results.
-In column A, we show the results for $R^*$ calculated on the 4 chains that ran; column B shows the same calculation but after the chains are split into two equal halves.
-The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates.
-The points and triangles show $R^*$ derived from GBM and RF classifiers, respectively.
-This plot mirrors Fig.
-4 in the supplementary materials of \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does.
-As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.
-
-		These results make intuitive sense: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation.
-In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals.
-By splitting chains into two -- the first half being the early phase, and the second half being the later phase with higher mean -- this forces differences in the marginals.
-This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one.
-As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\widehat{R}$ \citep{carpenter2017stan,vehtari2019rank}.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{trends_all_dim.pdf}}
-			\caption{\textbf{Univariate trends example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
-Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the 4 chains; column B  shows the same calculation after each chain is split into two halves.
-The rows present the differing sample sizes.
-The horizontal axis measures (half) the change in mean across the whole sample: so a value ``1'' indicates the mean increases by 2 units from the start to end of sampling.
-At each parameter set, 10 replicates were run and jitter was added to the points.}
-			\label{fig:trends_all_dim}
-		\end{figure}
-
-		\subsection{Trending mean in a single dimension}\label{sec:non-stationary_single}
-		We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends.
-In Fig.
-\ref{fig:trends_one_dim}, we show how $R^*$ performs across a range of target dimensions.
-In the simulations here, all dimensions bar one are stationary; the remaining dimension has a linear trend added to it.
-In all cases, across both GBM and RF classifiers, split-$R^*$ increased with trend.
-Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered.
-This suggests split-$R^*$ can robustly determine chain identity if only a single dimension has a non-stationary sampling distribution.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{trends_one_dim.pdf}}
-			\caption{\textbf{Multivariate trends example with split-$R^*$.} Points and triangles represent results for the GBM and RF classifiers, respectively.
-The columns present different dimensionalities of the target distribution; the rows present different sample sizes.
-The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions.
-At each parameter set, 10 replicates were run and jitter was added to the points.}
-			\label{fig:trends_one_dim}
-		\end{figure}
-
-		\subsection{Trending covariance}\label{sec:non-stationary_covariance}
-		Means that trend over time is one form of non-stationarity; another is a time-varying covariance.
-Next, we consider a bivariate normal with (constant) standard normal marginals but where the correlation between dimensions trends over time.
-Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d.
-draws from the process across 4 ``chains''.
-Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution.
-Similarly, split-$\widehat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig.
-\ref{fig:trends_joint_distribution}A).
-By contrast, split-$R^*$ can (Fig.
-\ref{fig:trends_joint_distribution}B), since it uses all information in the samples, including the covariance structure.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{trends_joint_distribution.pdf}}
-			\caption{\textbf{Bivariate normal with trending correlation example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
-Column A shows the results for split-$\widehat{R}$; column B for $R^*$; the rows present the differing sample sizes.
-The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling.
-At each parameter set, 10 replicates were run and jitter was added to the points.}
-			\label{fig:trends_joint_distribution}
-		\end{figure}
-
-		\subsection{Chain persistence}\label{sec:non-stationary_persistence}
-		When forming training and testing sets as part of the ML algorithm used to determine $R^*$, the testing set is effectively treated as a sort of ``independent'' hold-out dataset.
-Markov chains, in general, have persistence, meaning that the test set will not be truly independent and can -- according to the level of autocorrelation in the chains -- be highly related to the training set.
-In this section, we investigate how this autocorrelation affects the performance of $R^*$.
-The difficulty with this question is that higher chain autocorrelation typically means the sampling distribution is a rougher approximation of the target, so $R^*$ should be higher due to the properties of the sampling distribution.
-It could also be higher because the training set is less distinct from the testing set.
-
-		To investigate this, we generated AR(1) processes (as defined in eq.
-(\ref{eq:ar1})) with autocorrelations, $\rho$, ranging from 0.8-1 and, in each case, calculated $R^*$ via Algorithm \ref{alg:R_star}.
-Note, that only when $|\rho|<1$ is the marginal distribution defined by this process itself stationary; at $\rho=1$, its variance increases linearly with time, so, by definition, is not converged.
-In Fig.
-\ref{fig:trends_ar1}, we show the results of these simulations across various numbers of iterations: 250, 1000 and 4000 (different panels).
-In all cases, $R^*>1$ whenever $\rho=1$, indicating lack of convergence.
-As $\rho$ declined, so did $R^*$.
-
-		Notably, as the number of iterations increased, the values of $R^*$ for $\rho<1$ declined, whereas $R^*$ for $\rho=1$ actually increased: this is most easily seen by examining Fig.
-\ref{fig:trends_ar1_transposed} (which shows the same data as in Fig.
-\ref{fig:trends_ar1} but in a different way).
-This characteristic is exactly as desired: larger sample sizes should yield a sampling distribution closer to convergence for $\rho<1$; the same for $\rho=1$ produces distributions that are no closer to convergence, and more samples allows better determination of this non-convergence.
-
-		Overall, it seems that $R^*$ is a conservative measure and with greater chain autocorrelation it suggests more draws are necessary for convergence.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1.pdf}}
-			\caption{\textbf{Non-stationary distribution: AR(1) example.} Points and triangles represent results for the GBM and RF classifiers, respectively.
-The horizontal axis measures the autocorrelation of the AR(1) processes; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different number of iterations.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-The black line shows $R^*=1$.}
-			\label{fig:trends_ar1}
-		\end{figure}
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{trends_ar1_transposed.pdf}}
-			\caption{\textbf{Non-stationary distribution: AR(1) example alternative view.} Column A shows the results for a GBM classifier; column B for a RF classifier.
-The horizontal axis measures the number of iterations; the vertical axis shows the value of $R^*$ calculated on chains split in half; each panel shows a different value of autocorrelation.
-At each parameter set, 10 replicates were run and jitter was added to the points.
-The black dots show the median $R^*$ values at each parameter set; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
-The black horizontal line shows $R^*=1$.
-}
-			\label{fig:trends_ar1_transposed}
-		\end{figure}
-
-		\section{Many parameter models: ovarian and prostate analysis}\label{sec:prostate}
-		In this section, we analyse two Bayesian models both fit to real data.
-In both examples, we illustrate $R^*$ distributions derived only from GBM classifiers.
-The first is a logistic regression model fit to microarray ovarian cancer data with 54 data points and 1536 predictor variables; overall, this model has 4719 parameters.
-Since there are relatively few data points relative to the number of predictors, regularised horseshoe priors are specified on the regression coefficients \citep{piironen2017sparsity} since most are expected to be zero.
-This dataset has been used for benchmarking in the past (see, for example, \cite{schummer1999comparative,hernandez2010expectation,paananen2019implicitly}) and is known to result in a multimodal posterior.
-
-		The other dataset we use is of similar form but for prostate cancer.
-The original form of the dataset is described here: \cite{singh2002gene}.
-We use a filtered version of the dataset as detailed here: \cite{yang2006stable}, which we also analysed using logistic regression with regularised horseshoe priors.
-It has 18,105 parameters in total.
-
-		For each model, we consider two MCMC runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-
-		For the ovarian model, we show the results in Fig.
-\ref{fig:ovarian}.
-In Fig.
-\ref{fig:ovarian}A, we show the $R^*$ distributions for each model run, which show that, whereas the ``long'' model run has converged, the ``short'' one has yet to do so.
-Whilst it is harder to discern, this pattern is mirrored in Fig.
-\ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters, whereas the short model had 83 parameters where this was not the case.
-Similarly, in Figs.
-\ref{fig:ovarian}C and \ref{fig:ovarian}D, which show the bulk-ESS and tail-ESS respectively, it is evident that, for the short model, there remain a few parameters with low effective sample sizes, whereas the long model has more consistent values for this metric.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{ovarian.pdf}}
-			\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) using the GBM classifier are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
-			\label{fig:ovarian}
-		\end{figure}
-
-		For the prostate model, we show the results in Fig.
-\ref{fig:prostate}.
-Since this model has nearly four times as many parameters as the ovarian model, it was more computationally expensive to estimate $R^*$ for it.
-To handle this, we thinned down the parameters by a factor of 5 for the long model, recognising that, of course, this measure will make it more likely that we diagnose convergence.
-Despite this, both $R^*$ measures indicated that the MCMC runs had yet to converge (Fig.
-\ref{fig:prostate}A), which was mirrored by the other metrics considered (Figs.
-\ref{fig:prostate}B,C\&D).
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{prostate.pdf}}
-			\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains run with 800 thinned post-warm-up iterations (9000 total iterations with 1000 discarded as warm-up; 8000 post-warm-up iterations thinned by a factor of 10); another with 16 chains run with 1000 post-warm-up iterations each (500 warm-up iterations discarded and no thinning).
-In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) using the GBM classifier are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk-ESS across all parameters; and D shows tail-ESS across parameters.}
-			\label{fig:prostate}
-		\end{figure}
-
-		\section{Discrete distributions}\label{sec:discrete}
-		In this section, we compare the performance of $R^*$ and $\widehat{R}$ on a univariate discrete target distribution.
-To do so, we generate draws from four discrete Markov chains, following a 1st order Markov process with transition probabilities,
-		%
-		\begin{equation}
-		p_{ij} = \text{Pr}(X_{n+1}=j|X_n=i),
-		\end{equation}
-		%
-		where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$.
-The target distribution is, hence, the stationary distribution of this process.
-
-		\subsection{Small state-space}\label{sec:discrete_small}
-		We first consider a univariate discrete distribution with four states. In three of the four chains, we generate draws using a transition probability matrix:
-		%
-		\begin{gather}
-		\boldsymbol{P}
-		=
-		\begin{bmatrix}
-		0 & 1/2 & 1/2 & 0\\
-		1/2 & 0 & 1/3 & 1/6\\
-		1/4 & 1/4 & 1/4 & 1/4\\
-		0 & 1 & 0 & 0
-		\end{bmatrix},
-		\end{gather}
-		%
-		which implies a stationary target distribution with the following probability simplex $\pi_1 = (11/46, 15/46, 14/46, 6/46)$.
-
-		For the fourth chain, we specify differing transition probability matrices in each of three sets of results.
-In the first, we assume $\boldsymbol{P}_1=\boldsymbol{P}$ (i.e.
-the same as used for the other three chains); in the second and third cases, we assume this transition matrix is given by:
-		%
-		\begin{gather}
-		\boldsymbol{P}_2
-		=
-		\begin{bmatrix}
-		0 & 1/2 & 1/2 & 0\\
-		1/2 & 0 & 1/3 & 1/6\\
-		5/8 & 1/8 & 1/8 & 1/8\\
-		1/2 & 1/2 & 0 & 0
-		\end{bmatrix},\quad
-		\boldsymbol{P}_3
-		=
-		\begin{bmatrix}
-		0 & 1/2 & 1/2 & 0\\
-		1/2 & 0 & 1/3 & 1/6\\
-		1 & 0 & 0 & 0\\
-		1 & 0 & 0 & 0
-		\end{bmatrix}.
-		\end{gather}
-		%
-		These correspond to stationary probability simplices:
-		%
-		\begin{equation}
-		\pi_2 = (71/198, 17/66, 10/33, 8/99), \quad \pi_3 = (4/9, 2/9, 8/27, 1/27).
-		\end{equation}
-		%
-		The three stationary distributions corresponding to each transition probability matrix are shown in Fig.
-\ref{fig:discrete_stationary}.
-This illustrates that there is an ordering among the stationary distributions $\pi_1\sim\pi_2\sim\pi_3$, meaning $\pi_1$ is most similar to $\pi_2$, and $\pi_3$ is most similar to $\pi_2$.
-Put another way, $\pi_3$ is most different from $\pi_1$.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary.pdf}}
-			\caption{\textbf{Discrete distribution example: stationary distributions.} The three colours illustrate the stationary distributions of the Markov chains corresponding to the matrices $P_1$, $P_2$ and $P_3$ defined in \S\ref{sec:discrete}.}
-			\label{fig:discrete_stationary}
-		\end{figure}
-
-		For each of these three cases, we consider a range of sample sizes across each of the four chains: 1000, 2000, 5000 and 10,000.
-At each combination of transition probability matrix (for the fourth chain) and sample size, we perform 40 replicates.
-In each replicate, we calculate both $R^*$ using a GBM classifier and rank-normalised split-$\widehat{R}$.
-The results of these experiments are shown in Fig.
-\ref{fig:discrete}.
-In the horizontal axes of both panels, we show the sample size.
-The vertical axis indicates the value of $R^*$ and split-$\widehat{R}$ in panels A and B, respectively.
-The stacked sub-panels within A and B indicate the transition matrix used for the fourth chain.
-Since $\boldsymbol{P}_1=\boldsymbol{P}$, the stationary distribution for the other three chains, both diagnostics should tend towards indicating convergence in these cases, and, for both statistics, this was typically the case.
-For $R^*$, the replicates were centred on $R^*=1$, although the stochasticity of replicates meant that, in some cases, $R^*>1$.
-Whereas, $\widehat{R}$ was similarly close to 1 in all cases.
-Considering the simulations using $\boldsymbol{P}_2$ for the fourth chain transition probability matrix, $R^*>1$ across a majority of replicates, and this majority increased with sample size.
-In this case, $1.01>\widehat{R}>1$ for all replicates across the various sample sizes.
-For simulations using $\boldsymbol{P}_3$ for the fourth chain transition probability matrix, $R^*>1$ across all replicates and the magnitude of median $R^*$ increased with sample size; in this case, $\widehat{R}>1.01$ in all replicates.
-
-		This example shows that $R^*$ can detect poor convergence for discrete parameter spaces, like $\widehat{R}$.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{discrete.pdf}}
-			\caption{\textbf{Discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations.
-The panels within A.
-and B.
-indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$).
-In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates.
-In panel B, we show the same, but for split-$\widehat{R}$.
-The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
-			\label{fig:discrete}
-		\end{figure}
-
-		\subsection{Large state-space}\label{sec:discrete_large}
-		We next consider a larger discrete parameter space: a univariate distribution with 20 states.
-As in \S\ref{sec:discrete_small}, we generate draws from three discrete Markov chains that share a common transition probability matrix, $\boldsymbol{P}$.
-For the fourth (and final) chain, we consider two cases: one where it shares the same matrix, $\boldsymbol{P}_1=\boldsymbol{P}$; the other, when its Markov process has a different matrix, $\boldsymbol{P}_2$.
-Rather than write down a given $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we generated these stochastically: for each of the rows, we randomly sampled from from a Dirichlet process with shape vectors consisting of ones (of length 20).
-
-		To illustrate the differences between the two transition probability matrices, $\boldsymbol{P}$ and $\boldsymbol{P}_2$, we simulate 100,000 draws from each Markov process.
-In Fig.
-\ref{fig:discrete_stationary_higherd}, we show the marginal distribution obtained across all four chains: where the first three used $\boldsymbol{P}$, and the last used $\boldsymbol{P}_2$.
-In this plot, we show only the first four states of the 20-state model.
-The differences in marginal distributions between the first three chains and the last indicate that there are substantial differences in the target distributions.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{discrete_stationary_higherd.pdf}}
-			\caption{\textbf{Large state-space discrete distribution example: approximate stationary distribution.} The distribution over the first four states of a 20-state discrete model achieved after drawing 100,000 samples from each of four Markov chains.
-Here, the first three chains shared the same transition probability matrix; the last chain used a different matrix.}
-			\label{fig:discrete_stationary_higherd}
-		\end{figure}
-
-		To probe the ability of $R^*$ and $\widehat{R}$ to detect poor convergence, we repeat the same exercise as in \S\ref{sec:discrete_small} although, here, we consider the new $\boldsymbol{P}_1$ and $\boldsymbol{P}_2$ cases.
-In Fig.
-\ref{fig:discrete_higherd}, we show the results of this analysis.
-This shows that $R^*\sim 1$ for the $\boldsymbol{P}_1$ case, correctly indicating no convergence issues; for the $\boldsymbol{P}_2$ case, $R^*>1$ across all replicates at sample sizes of 5000 and above.
-By contrast, $\widehat{R}$ struggles to differentiate between the $\boldsymbol{P}_1$ and $\boldsymbol{P}_2$ cases: in both instances, it signifies convergence.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{discrete_higherd.pdf}}
-			\caption{\textbf{Large state-space discrete distribution example: convergence diagnostics.} On the horizontal axis, we show the sample size of each set of simulations.
-The panels within A.
-and B.
-indicate the transition probability matrix used for the fourth chain (the other three chains always used $\boldsymbol{P}=\boldsymbol{P}_1$).
-In panel A, grey points indicate $R^*$ from each replicate calculated using Algorithm \ref{alg:R_star}) , and the black point-ranges indicate 2.5\% , 50\% and 97.5\% quantiles across the 40 replicates.
-In panel B, we show the same, but for split-$\widehat{R}$.
-The dashed horizontal lines at 1 indicate the convergence threshold in both cases.}
-			\label{fig:discrete_higherd}
-		\end{figure}
-
-		Collectively, the results of \S\ref{sec:discrete_small} and \S\ref{sec:discrete_large} show that $R^*$ is able to detect convergence issues in discrete target distributions.
-The results of \S\ref{sec:discrete_large} hint that $R^*$ may be superior than $\widehat{R}$ in larger discrete spaces, although further empirical investigation is needed.
-
-		\section{ML sensitivity}\label{sec:ML_sensitivity}
-		In this section, we investigate how two decisions about classifiers — which classifier to use (in \S\ref{sec:ml_model}) and what hyperparameters to use for it (in \S\ref{sec:hyperparameters}) — affect calculation of $R^*$.
-The test cases used for empirical evaluation were selected from our pool of examples introduced in the text.
-Specifically, they were chosen to represent cases where $R^*$ was nearer 1, since, for these, small changes in calculated $R^*$ could lead to different decisions about convergence being reached.
-The four examples we used were:
-
-		\begin{itemize}
-			\item The AR(1) example described in \S\ref{sec:heterogeneity}, except using 1000 draws per chain.
-			\item The 250-dimensional non-centered multivariate normal example introduced in \S\ref{sec:multivariate_normal_250}.
-In each replicate, we generated 500 post-warm-up draws from the posterior (using 500 warm-up steps) using Stan's NUTS algorithm across each of 4 chains.
-			\item The alternative parameterisation of the Cauchy model described in \S\ref{sec:cauchy}.
-For each replicate, we generated samples from the model using Stan's NUTS algorithm with 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains.
-			\item The non-centered parameterisation of the eight schools model described in \S\ref{sec:eight_shools}.
-In each replicate, we used Stan's NUTS algorithm with adapt delta set to 0.95 to generate 1000 post-warm-up draws (and 1000 warm-up draws) across each of 4 chains; these were then used to calculate $R^*$.
-		\end{itemize}
-
-		\subsection{ML classifier comparison}\label{sec:ml_model}
-		In this section, we investigate how classification accuracy depends on choice of ML classifier.
-We restricted our analysis to popular classifiers with relatively few hyperparameters to simplify comparison amongst them.
-Notably, we do not consider neural network models, since the structure of nets effectively defines a high dimensional hyperparameter space.
-
-		The ML classifiers we compared were GBMs, RFs, k-nearest neighbours (KNNs), SVMs with linear kernels and a generalised linear model approach (GLM).
-All these models were called through the \textbf{\textsf{R}}'s Caret package \citep{kuhn2008building}: the GBM is implemented using the \textit{gbm} model in Caret which uses the ``gbm'' package \citep{greenwell2019package}; RFs are implemented using the \textit{rf} model in Caret which uses the ``randomForest'' \citep{liaw2002classification} package; KNN is natively implemented in Caret and called using the \textit{knn} model; SVMs are implemented in Caret using the \textit{svmLinear} model which uses ``kernlab'' package \citep{karatzoglou2004kernlab}; and the GLM is implemented through the \textit{multinom} model in the ``nnet'' package \citep{ripley2016package}.
-
-		Each of these classifiers has hyperparameters, and, due to the differences between the examples we consider (defined in \S\ref{sec:ML_sensitivity}), the optimal hyperparameters were likely to differ between examples.
-For each classifier-example combination, we first performed a single replicate of each experiment and choose those hyperparameters which maximised classification accuracy.
-The set of all hyperparameters we selected between were chosen so that training time for each classifier remained reasonable.
-These sets were the same across all examples:
-
-		\begin{itemize}
-			\item GBM: Cartesian product of $\text{int.depth}=(3, 7)$; $\text{\# trees} = (50, 100)$;
-			\item RF: $m_{\text{try}} = (1, 2)$.
-(Note, in \S\ref{sec:hyperparameters}, we investigate a dimension-specific approach to $m_{\text{try}}$ but, here, for ease of comparison with the other methods, we keep this static.)
-			\item KNN: $K=(5,10,15, 20,40)$;
-			\item SVM: $C=(0.25, 0.5, 0.75)$;
-			\item GLM: $\text{decay}=(0.1, 0.2, 0.5, 1)$.
-		\end{itemize}
-
-		These hyperparameters were then used in subsequent replicates.
-We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 100 replicates; for the multivariate normal example, we used 100; for the Cauchy model, 50 replicates; and for the eight schools model, 30 replicates.
-
-		The results of these experiments are shown in Fig.
-\ref{fig:ml_comparison_all}.
-The top row shows $R^*$ as calculated by each classifier (in each case, chains split into two halves); the bottom row shows the time (in seconds) taken for $R^*$ calculation.
-The panels correspond to the examples described in \S\ref{sec:ML_sensitivity}.
-Across the examples, the classification accuracy of RFs and GBMs were, generally, highest; followed by KNNs.
-An exception was for the 250-dimensional normal where KNN performed best.
-In the AR(1) example, GBM outperformed RF.
-In higher dimensional cases, RF outperformed GBM.
-In all cases, RF and GBM classifiers took the greatest time to train; in higher dimensional examples, RF-based $R^*$ were cheaper to calculate than those from GBMs.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{ml_comparison_all.pdf}}
-			\caption{\textbf{ML classifier comparison.} Top row shows $R^*$ values calculated using Algorithm \ref{alg:R_star}; bottom row show the time taken to run each method on a desktop computer.
-The different columns correspond to the examples described in \S\ref{sec:ML_sensitivity}.
-Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates; note, that the number of replicates varies across examples as given in \S\ref{sec:ml_model}.}
-			\label{fig:ml_comparison_all}
-		\end{figure}
-
-		\subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
-		In this section, we considered only the two classifiers — GBMs and RFs — that performed best on our comparison between ML methods in \S\ref{sec:ml_model}.
-The performance of GBMs and RFs, like most ML methods, depends on their hyperparameters.
-In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree.
-The number of such regression trees is thus a hyperparameter of the model.
-Each regression tree can split amongst any integer number of variables (the \textit{interaction depth}).
-For GBMs, we analyse how their performance depends on these two hyperparameters.
-For RFs, there is a single hyperpameter: $m_{\text{try}}$, the number of members of the subset of all features over which to search for an optimal split when forming a decision tree.
-Note, that the maximum allowed value of $m_{\text{try}}=K$ — the dimensionality of the target distribution.
-
-		In this section, we investigate the sensitivity of $R^*$ calculated by GBMs and RFs to each of these sets of hyperparameters across a range of examples (defined in \S\ref{sec:ML_sensitivity}).
-Since each example is of different dimensionalities, we compare different sets of hyperparameters in each, although include amongst the sets the default values we suggest for each classifier in \S\ref{sec:method} (these are shown as triangles on the plots).
-Note, that for the GBM model, we did not investigate a full range of hyperparameters needed to optimise $R^*$ for many of the examples due to the extensive training time needed to do so.
-For RFs, runtime was less restrictive and we were better able to survey the sensitivity across hyperparameter space.
-
-		We also used a different number of replicates at each unique combination of classifier and hyperparameters across the experiments due to the differing demands of each example: for the AR(1) example, we used 200 replicates; for the multivariate normal example, we used 20; for the Cauchy model, 20 replicates; and for the eight schools model, 50 replicates.
-
-		\subsubsection{Autoregressive example}\label{sec:hyperparameters_ar1}
-		In Fig.
-\ref{fig:hypers_ar1}, we show the results of the AR(1) analysis.
-In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
-In panel B, we show the same but for RFs.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
-For the GBM model, $R^*$ varied according to hyperparameters: of the set investigated, an interaction depth of 1 appears optimal with 10+ trees.
-In panel B, the values of $R^*$ are generally lower than those achieved by GBMs.
-As shown in \S\ref{sec:comparison_gbm_rf}, RFs appear less well-suited to lower dimensional problems.
-Across the two possible hyperparameter values for RFs, there was not a substantial difference in the values of $R^*$ calculated.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{hypers_ar1.pdf}}
-			\caption{\textbf{Hyperparameter sensitivity: AR(1) example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the AR(1) example described in \S\ref{sec:hyperparameters_ar1}; Panel B shows the same but for a RF classifier.
-Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
-			\label{fig:hypers_ar1}
-		\end{figure}
-
-		\subsubsection{Multivariate normal: 250-dimensional model}\label{sec:hyperparameters_normal}
-		In Fig.
-\ref{fig:hypers_normal}, we show the results of the analysis for the 250-dimensional normal example.
-In panel A, we show the sensitivity of $R^*$ as calculated by GBMs to variation in hyperparameters.
-In panel B, we show the same but for RFs.
-This shows that $R^*$ was higher as calculated by RFs and that performance of GBMs depended strongly on hyperparameters.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{hypers_normal_250.pdf}}
-			\caption{\textbf{Hyperparameter sensitivity: 250-dimensional normal example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the Cauchy example described in \S\ref{sec:hyperparameters_normal}; Panel B shows the same but for a RF classifier.
-Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
-			\label{fig:hypers_normal}
-		\end{figure}
-
-		\subsubsection{Cauchy model: alternative parameterisation}\label{sec:hyperparameters_cauchy}
-		In Fig.
-\ref{fig:hypers_cauchy}, we show the results of the Cauchy model analysis.
-In panel A, we show $R^*$ as calculated by a GBM classifier; in panel B, the same but using a RF classifier.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.
-Across the range of hyperparameters investigated, the RF classifier outperformed the GBM and was less sensitive to variation in its hyperparameters.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{hypers_cauchy.pdf}}
-			\caption{\textbf{Hyperparameter sensitivity: Cauchy example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the Cauchy example described in \S\ref{sec:hyperparameters_cauchy}; Panel B shows the same but for a RF classifier.
-Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
-			\label{fig:hypers_cauchy}
-		\end{figure}
-
-		\subsubsection{Eight schools: non-centered parameterisation}\label{sec:hyperparameters_8_schools}
-		In Fig.
-\ref{fig:hypers_8_schools}, we show the results of the experiments on the eight schools model.
-Panel A shows how $R^*$ calculated using a GBM varies with its two hyperparameters; panel B shows the same but for RFs.
-This shows that, in this case, the value of $R^*$ calculated by RFs was higher than that for GBMs.
-Further, whereas GBMs were relatively sensitive to their hyperparameter values, RFs were less so.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1.0\textwidth]{hypers_8_schools.pdf}}
-			\caption{\textbf{Hyperparameter sensitivity: eight schools example.} Panel A shows the sensitivity of GBM-derived $R^*$ to hyperparameter variation for the eight schools example described in \S\ref{sec:hyperparameters_8_schools}; Panel B shows the same but for a RF classifier.
-Black point-ranges indicate 25\% , 50\% and 75\% quantiles across all replicates.
-In Panel A, the coloured lines shows the different interaction depths investigated.
-In both panels, the results for the parameterisation we suggest as defaults is shown as a triangle.}
-			\label{fig:hypers_8_schools}
-		\end{figure}
-
-		\section{Comparing GBMs and RFs}\label{sec:comparison_gbm_rf}
-		In \S\ref{sec:ML_sensitivity}, we present a series of experiments using popular ML methods and find that, among these methods, GBMs and RFs perform most consistently across them.
-In this section, we compare $R^*$ derived from using a GBM classifier with that from a RF classifier.
-In these experiments, we fix the hyperparameters of each as given in \S\ref{sec:method}.
-In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution.
-In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains.
-Since we use independent sampling to generate draws, we are also able to calculate the \textit{Bayes optimal} classification accuracy, which is the classifier that minimises the probability of misclassification error \citep{devroye2013probabilistic}.
-This classifier assigns chain identities according to the one which has the maximum posterior probability of having generated a draw.
-Here, we assume that \textit{a priori} all chains are equally likely to have generated the data, so the classifier categorises draws to the chain which maximises the likelihood of that observation.
-
-		\subsection{Joint distribution}\label{sec:joint_distribution}
-		Here, we consider a multivariate normal target distribution with dimensionality ranging from 1 to 32.
-For three `chains', we generated 2000 draws by independently sampling from a multivariate normal with zero mean and given covariance matrix; for the fourth chain, the same number of draws are generated in the same way except with a different covariance matrix.
-To generate the covariance matrices, we randomly sampled two 32-dimensional covariance matrices from the LKJ distribution with degrees of freedom 1 \citep{lewandowski2009generating}: note, that this results in matrices with unit diagonal values, meaning the marginal distributions of each dimension are standard normals.
-To obtain covariance matrices for the targets with dimensions, $d<32$, submatrices, $A^{\{d\}}$, were constructed from the 32-dimensional covariance matrices by taking leading blocks from them: $A^{\{d\}}:=A^{\{32\}}_{1:d, 1:d}$.
-For each target, we performed 20 replicates, where in each case, $R^*$ was calculated using GBM and RF classifiers.
-In addition, we determined an $R^*$ using the Bayes optimal classifier: to do so, we generated 10,000 draws from the four-chain process and assign chain identities to the maximum likelihood class.
-
-		The results of these experiments are shown in Fig.
-\ref{fig:gbm_vs_rf_normal}.
-In this plot, the bottom axis shows the dimensionality of the target distribution; the vertical axis shows $R^*$, and the point colour and shadings give the method used to calculate this statistic.
-When the target is unidimensional, the four chains have the same target distribution and the optimal $R^*=1$.
-In this case, both the GBM and RF methods produce classification rates that overlap with this value, indicating convergence.
-As the number of dimensions increases, it becomes easier to differentiate between samples from the two processes and the optimal $R^*$ grows.
-$R^*$ as calculated by GBMs and RFs also increases.
-For a two-dimensional target, the GBM method outperforms the RF one, getting closer to optimal classification.
-For higher dimensional targets, the RF classifier outperforms, achieving near-optimal $R^*$ in 32 dimensions.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_normal.pdf}}
-			\caption{\textbf{GBM versus RF: multivariate normal.} The horizontal axis gives the dimensions of the target distribution; the vertical axis gives the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
-The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
-For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
-The dashed line shows $R^*=1$.
-Here, $R^*$ was calculated using chains split into two halves.}
-			\label{fig:gbm_vs_rf_normal}
-		\end{figure}
-
-		\subsection{Fat tails}\label{sec:tail_fatness}
-		We now compare how the GBM and RF methods are able to diagnose lack of convergence in the tails of a target distribution.
-To do so, we again consider four chains.
-In three of these chains, we generated 2000 draws by randomly sampling from a multivariate Student-t distribution with mean zero and 3 degrees of freedom.
-In the remaining chain, we generated the same number of independent draws but use a mean-zero multivariate Student-t distribution with differing degrees of freedom, $\nu$.
-We consider a range of target dimensions: to generate the shape matrix for the first three chains, we use the same approach as in \S\ref{sec:joint_distribution} to build up appropriate shape matrices for each target dimensions, $A^{\{d\}}$.
-We examine a range of values of $\nu=\{4, 8, 16, 32\}$, representing decreasing tail fatness.
-Across the range of $\nu$ considered, the covariance matrix for the first three chains is given by $\nu/{(\nu-2)} A^{\{d\}}$.
-To ensure that the fourth chain has the same covariance as the first three, we use a shape matrix ${(\nu-2)}/\nu A^{\{d\}}$.
-For each combination of dimensions and $\nu$, we performed 20 replicates, where in each, we again calculate $R^*$ for both the GBM and RF classifiers.
-Additionally, we estimate an optimal $R^*$ using the Bayes optimal classifier by calculating the predictive accuracy of 10,000 draws of the four-chain process.
-
-		The results of these experiments are shown in Fig.
-\ref{fig:gbm_vs_rf_studdentt}.
-Here, each panel shows a target distribution of different dimensionality: 1, 2, 4, 8, 16 and 32 dimensions.
-Within each panel, the horizontal axis shows the degrees of freedom, $\nu$, of the fourth chain; the vertical axis gives $R^*$.
-The point colour and shadings give the method used to calculate $R^*$.
-In each panel, increases in $\nu$ make it easier to differentiate draws from the fourth chain from those of the first three: accordingly $R^*$ tends to increase for each method.
-As the target dimensionality increases (panels left-right along each row), it also becomes easier to contrast draws from the fourth chain, and $R^*$ increases.
-In all cases, the optimal $R^*$ exceeds those using GBM or RF classifiers.
-In 1-4 dimensions, the GBM approach outperforms the RF one, getting closer to the optimal classification rate.
-If the number of dimensions is 16 or greater, the order switches, and RFs often outperform GBMs.
-
-		\begin{figure}[!htb]
-			\centerline{\includegraphics[width=1\textwidth]{gbm_vs_rf_studentt.pdf}}
-			\caption{\textbf{GBM versus RF: multivariate Student-t.} Different panels show the various dimensions of the target distribution.
-The horizontal axis gives the degrees of freedom of the multivariate Student-t distribution for the fourth chain; the vertical axis gives the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} using a GBM (light grey points), a RF (dark grey triangles) and the Bayes optimal classifier (black squares).
-The upper and lower point ranges show the 75\% and 25\% quantiles and the filled shapes show the medians.
-For the Bayes optimal classifier, there is only a point estimate since the optimal accuracy was estimated by Monte Carlo sampling using 10,000 draws from the four chain process.
-The dashed line shows $R^*=1$.
-Here, $R^*$ was calculated using chains split into two halves.}
-			\label{fig:gbm_vs_rf_studdentt}
-		\end{figure}
-
 	\end{supplement}

### 2020-11-20 write_up/ba_final_submission/supplementary.tex
added break out of files

--- 2020-11-18 write_up/ba_final_submission/ms_final.tex
@@
 \usepackage{booktabs}
+\usepackage{xr}
+\externaldocument[]{ms_final}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thesection}{S\arabic{section}}
 
@@
 	\begin{frontmatter}
-		\title{$R^*$: A robust MCMC convergence diagnostic with uncertainty using decision tree classifiers}
-		\runtitle{A MCMC convergence diagnostic using decision tree classifiers}
+		\title{Supplementary materials: $R^*$ convergence diagnostic}
+		\runtitle{Supplementary materials: $R^*$ convergence diagnostic}
 
@@
 
-		\begin{abstract}
-			Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
-Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its performance in finite time.
-The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
-Here, we introduce a new method for diagnosing convergence based on how well a machine learning classifier model can successfully discriminate the individual chains.
-We call this convergence measure $R^*$.
-In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined.
-Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead it uses all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches.
-We recommend calculating $R^*$ using two different machine learning classifiers — gradient-boosted regression trees and random forests — which each work well in models of different dimensions.
-Because each of these methods outputs a classification probability, as a byproduct, we obtain uncertainty in $R^*$.
-The method is straightforward to implement and could be a complementary additional check on MCMC convergence for applied analyses.
-		\end{abstract}
-
 	\end{frontmatter}
-
-	\section{Introduction}
-	Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years.
-In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan}, PyMC3 \citep{salvatier2016probabilistic}, Turing \citep{ge2018turing}, TensorFlow Probability  \citep{dillon2017tensorflow} and Pyro \citep{bingham2019pyro} implement.
-MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.
-
-	Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care.
-A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior.
-Because of this, software packages (Stan \citep{carpenter2017stan}, for example), go to great lengths to communicate to users any issues with sampling.
-
-	The most important determination of whether MCMC has worked is how closely the sampling distribution has converged to the posterior \citep{brooks2011handbook}.
-MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions).
-Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small number of draws.
-
-	The one diagnostic method for determining whether practical convergence has occurred relies on the fact that the posterior distribution is the unique stationary distribution for an MCMC sampler.
-Therefore, it would appear that, if an MCMC sampling distribution stops changing, then convergence has occurred.
-Unfortunately, anyone who uses MCMC knows that it is full of false dawns: chains can easily become stuck in areas of parameter space, and observation over short intervals mean the sampling distribution \textit{appears} converged \citep{gelman1992single}.
-Like furious bees trapped in a room of a house \citep{lambertbees}, MCMC samplers may fail to move due to the narrow gaps that join neighbouring areas.
-With MCMC, absence of evidence of new areas of high posterior density is, time and again, not evidence of their absence.
-
-	To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}.
-If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge.
-This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}.
-The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\widehat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman1992inference}.
-If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed.
-Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter draws rather than the raw values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}.
-Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}.
-Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.
-
-	Here, we introduce $R^*$, a new convergence diagnostic metric.
-This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it.
-Rephrased, it should not be possible to predict which draws come from which chain.
-In this vein, we use machine learning (ML) classifiers to measure convergence.
-Specifically, we train classifiers to predict the chain that generated each observation.
-By evaluating the performance of classifiers on a held-out test set, this provides a new convergence metric.
-To maximise predictive accuracy, our chosen classifiers naturally exploit differences in the full joint distributions between chains, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics.
-Our statistic, unlike its $\widehat{R}$ cousins, is scalar-valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution.
-However, the ML classifiers we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.
-
-	There are, of course, a huge variety of possible ML classifiers.
-For a method to be useful and widely adopted, however, it needs to satisfy a number of criteria: first, across a range of examples, it should consistently be able to detect poor MCMC convergence; second, the ML classifier should be trainable in a reasonable amount of time; lastly, the methods should be tuned (via their hyperparameters) so as to be standardised, so that $R^*$ computed by one analyst is comparable to that computed by another.
-Here, we perform comparisons across some of the most popular methods in the ML literature and recommend two tree-based ML classifiers: gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM'') and random forest classifiers \citep{breiman2001random} (``RF'').
-Both of these methods performed consistently well across our range of examples, and the models were relatively efficient to train.
-We recommend calculating $R^*$ using both of these methods: GBMs tend to perform best for low dimensional cases; in moderate dimensional cases upwards, RF dominates.
-This difference in performance is partly due to hyperparameter tuning: echoing previous results in the literature \cite[Chapters~11\&12]{boehmke2019hands}, our results show that GBMs, in general, are more sensitive to hyperparameter choice than are RFs.
-For a GBM classifier to perform well in higher dimensions, more rounds of boosting are required, which substantially increases training time; instead, we fix the hyperparameters of GBMs to specific values to ensure practical usefulness.
-For RF classifiers, empirical studies have demonstrated reliable heuristics for selecting robust hyperparameters \citep{bernard2009influence}, and, in any case, across our set of test cases, these classifiers are relatively insensitive to hyperparameter choice.
-
-	For the types of problem we tested, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer.
-In addition, since ML classifiers can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence.
-$R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.
-
-	The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios.
-Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.
-
-	\section{Method}\label{sec:method}
-	If Markov chains have not mixed, it is possible to guess (with more accuracy than chance) to which chain a draw belongs from its value.
-This is possible if there are differences in the sampling distribution for any dimension in the target distribution (Fig.
-\ref{fig:marginal}): in this case, if the marginal distributions differ between chains, this information can be used to predict which chain a draw belongs to.
-It  is also possible to predict the chain that generated a given draw if there are differences in the joint distribution of two (or more) dimensions of the target, even if the marginal distributions are the same (Fig.
-\ref{fig:joint}).
-
-	\begin{figure}[!htb]
-		\centerline{\includegraphics[width=1.0\textwidth]{unmixed_1.pdf}}
-		\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (with marginal distribution to the right of panel); B shows two chains that have not mixed.}
-		\label{fig:marginal}
-	\end{figure}
-
-\begin{figure}[h]
-	\centerline{\includegraphics[width=1.0\textwidth]{unmixed_2.pdf}}
-	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
-	\label{fig:joint}
-\end{figure}
-
-These two cases, whilst simple, illustrate the basis of our approach.
-To determine if a set of Markov chains has converged to the same distribution, we train a supervised ML model to classify the chain to which each draw belongs.
-By evaluating its performance on a separate test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains.
-By taking the ratio of classification accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{vehtari2019rank}.
-In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence.
-Algorithm \ref{alg:R_star} gives a recipe for calculating $R^*$.
-
-\begin{algorithm}[tb]
-	\caption{$R^*$ calculation}
-	\label{alg:R_star}
-	\begin{algorithmic}
-		\STATE Given chain-wise draws from the target, $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}$ and a test set length, $S_\text{test}$:
-		\FOR{$m=1$ to $N$}
-		\STATE Create train and test sets by random-sampling (w/o replacement), $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
-		\ENDFOR
-		\STATE Stack $X_\text{train} = (X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train})^T$
-		\STATE Stack $X_\text{test} = (X^{\{1\}}_\text{test},X^{\{2\}}_\text{test},...,X^{\{N\}}_\text{test})^T$
-		\STATE Train ML model to classify chain id from any draw, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
-		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
-		\STATE Predict chain id, $c^{\{s\}} = \text{ML}(x^{\{s\}}|X_\text{train})$
-		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
-		\ENDFOR
-		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^* = \bar{a} / (1 / N) = N \bar{a}$
-		\RETURN $R^*$
-	\end{algorithmic}
-\end{algorithm}
-
-To identify promising candidate ML methods, we run a series of experiments using a number of popular classifiers (see \S\ref{sec:ml_model}).
-Two methods performed consistently well across our examples: gradient-boosted regression trees (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}) and random forest classifiers \citep{breiman2001random} (``RF'').
-Both of these approaches are based on decision trees: GBMs use an iterative approach known as ``boosting'', where each subsequent decision tree aims to predict the residuals from the previous one; RFs use an approach known as ``bagging'' which trains decision trees on many bootstrapped copies of the training data, yielding a collection of independently trained trees whose individual decisions, when aggregated, yield a class.
-For the implementations of these classifiers, we use those available in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' \citep{greenwell2019package} and ``randomForest'' \citep{liaw2002classification} packages.
-
-The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint), and $K$ is the number of parameters.
-We split each chain's draws into randomly divided training and testing sets: here, we use 70\% of draws for training and 30\% for testing.
-Both approaches typically took $\mathcal{O}$(seconds) on a desktop computer to execute training then prediction on a testing set for most models we consider in \S\ref{sec:results}.
-
-Since different posteriors present different challenges to MCMC samplers, the nature of classification boundaries is problem-specific.
-Because of this, there is no unique optimal classifier across all problems, and the performance of the algorithms we investigate depends on their hyperparameters (see \S\ref{sec:hyperparameters}).
-GBM models have a number of hyperparameters and have previously been demonstrated to be hard to tune \cite[Chapter~12]{boehmke2019hands}.
-To ensure practical run time for this method, we suggest running it using a default hyperparameter set: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees be grown.
-This choice of hyperparameters was chosen to present a balance between training cost and classification accuracy, producing an $R^*$  that was a stringent measure of convergence — in general, more stringent than $\widehat{R}$.
-RFs are generally less sensitive to variation in their single hyperparameter: $m_{\text{try}}$, the size of the subset of all features over which to search for an optimal split \cite[Chapter~11]{boehmke2019hands}.
-Our experiments replicate these results (see \S\ref{sec:hyperparameters}), and we suggest running RFs using the heuristic $m_{\text{try}}=\sqrt{K}$, which has previously been shown to be a choice resulting in robust classification performance \citep{bernard2009influence,boehmke2019hands}.
-Unless otherwise stated, in the examples explored in \S\ref{sec:results}, the hyperparameters for these two classifiers were set at these defaults.
-
-From a classifier fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$.
-Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words.
-For each draw, $s$, in our testing set, classifiers output a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$.
-By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct.
-We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{S_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$.
-We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.
-
-\begin{algorithm}[tb]
-	\caption{Procedure to generate $I$ draws of $R^*$}
-	\label{alg:R_star_uncertainty}
-	\begin{algorithmic}
-		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
-		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
-		\FOR{$i=1$ to $I$}
-		\FOR{$s=1$ to $S_\text{test}$}
-		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
-		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
-		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
-		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
-		\ENDFOR
-		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
-		\STATE Calculate ratio to null model accuracy, $R^{*\{i\}} = \bar{a} / (1 / N) = N\bar{a}$
-		\ENDFOR
-		\RETURN $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$
-	\end{algorithmic}
-\end{algorithm}
-
-\section{Results}\label{sec:results}
-To illustrate the versatility of $R^*$, we use a range of examples that demonstrate how this statistic fares across a range of scenarios.
-Table \ref{tab:results} summarises the examples and provides a rationale for their inclusion.
-The experiments not detailed in the main text are briefly described in \S\ref{sec:further_experiments} and more fully in the relevant sections given in Table \ref{tab:results}.
-In all cases where $\widehat{R}$ was calculated, unless otherwise stated, we followed the approach in \cite{vehtari2019rank}, by calculating it as the maximum of rank-normalised split-$\widehat{R}$ and rank-normalised folded split-$\widehat{R}$: for simplicity, we refer to this as rank-normalised split-$\widehat{R}$.
-
-\begin{table}[]
-	\footnotesize
-	\begin{tabular}{l|l|l}
-		\textbf{Example }&  \textbf{Relevance} & \textbf{Section} \\
-		\midrule
-		Autoregressive & \textbf{Examining $R^*$ and sensitivities to its calculation} & \ref{sec:heterogeneity} \\
-		& Detecting heterogeneous chain variance using $R^*$ & \ref{sec:heterogeneity_performance} \\
-		& Stochasticity in $R^*$ & \ref{sec:heterogeneity_stochasticity}\\
-		& Generating $R^*$ uncertainty measure & \ref{sec:heterogeneity_uncertainty}\\
-		& Sensitivity of $R^*$ to number of chains & \ref{sec:heterogeneity_numchains}\\
-		\midrule
-		Multivariate normals & \textbf{Detecting convergence in joint distributions} &  \ref{sec:multivariate_normal}\\
-		& Unconverged joint distribution in bivariate normal & \ref{sec:multivariate_normal_bivariate}\\
-		& High correlations between dims in 250D normal  & \ref{sec:multivariate_normal_250}\\
-		& Measuring contributions of variables to poor convergence & \ref{sec:multivariate_normal_varimportance}\\
-		\midrule
-		Cauchy & \textbf{Detecting convergence for long-tailed distributions} & \ref{sec:cauchy}\\
-		& Comparing $R^*$ and existing measures to & \ref{sec:cauchy_objective}\\
-		& objective convergence &\\
-		\midrule
-		Eight schools model & \textbf{Hierarchical Bayesian model slow convergence} & \ref{sec:eight_shools}\\
-		\midrule 
-		Wide multivariate normal & \textbf{Detecting convergence when \# draws $\sim$ \# dims} & \ref{sec:wide}\\
-		\midrule
-		Non-stationary marginals & \textbf{Detecting time-varying sampling distributions} & \ref{sec:non-stationary}\\
-		& Trends in mean across all chains & \ref{sec:non-stationary_chains}\\
-		& Trends in mean in a single dimension & \ref{sec:non-stationary_single}\\
-		& Trends in covariance & \ref{sec:non-stationary_covariance}\\
-		& Sensitivity of $R^*$ to chain persistence & \ref{sec:non-stationary_persistence}\\
-		\midrule
-		Ovarian and prostate models & \textbf{Bayesian models with many parameters} & \ref{sec:prostate}\\
-		&  \textbf{and multimodal posteriors} &\\
-		\midrule		
-		Discrete Markov model & \textbf{Evaluating $R^*$ on discrete parameter models} & \ref{sec:discrete}\\
-		& Small state-space & \ref{sec:discrete_small}\\
-		& Large state-space & \ref{sec:discrete_large}\\
-		\midrule
-		Various & \textbf{Sensitivity of $R^*$ to ML model} & \ref{sec:ML_sensitivity}\\
-		& Comparing different ML classifiers & \ref{sec:ml_model}\\
-		& Sensitivity of $R^*$ to GBM and RF hyperparameters & \ref{sec:hyperparameters}\\
-		\midrule
-		Multivariate normal \& & \textbf{Comparing GBM and RF classifiers} & \ref{sec:comparison_gbm_rf}\\
-		Student-t dists. & &\\
-		& Detecting convergence in joint distributions & \ref{sec:joint_distribution}\\
-		& Tail convergence & \ref{sec:tail_fatness}\\
-	\end{tabular}\caption{\textbf{Summarising the example problems and reasons for their inclusion.}}\label{tab:results}
-\end{table}
-
-\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
-In this section, we use a simple example to illustrate how $R^*$ works.
-Specifically, we show how $R^*$ can detect heterogeneous variance across Markov chains.
-This example aims to illustrate the basic mechanics behind how $R^*$ works, so we use only the GBM classifier here.
-This section also investigates the sensitivity of this measure: across different training and testing sets (\S\ref{sec:heterogeneity_stochasticity}) and different draws from the ML model-predicted probability simplex (\S\ref{sec:heterogeneity_uncertainty}); and to differing numbers of chains (\S\ref{sec:heterogeneity_numchains}).
-The experiments we use to study these issues are all of similar form to the following data generating process: four Markov chains are generated, where each samples from an autoregressive order 1 (AR(1)) process of the form,
-%
-\begin{equation}\label{eq:ar1}
-X_t = \rho X_{t-1} + \epsilon_t,
-\end{equation}
-%
-where $\epsilon_t\stackrel{i.i.d.}{\sim}\text{normal}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$.
-Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) standard deviation of the others.
-
-\subsubsection{Performance of $R^*$}\label{sec:heterogeneity_performance}
-To illustrate the consistency of $R^*$, we performed 1000 replicates where, in each case, we generated four $\{X_t\}$ series as described (i.e.
-where one chain has a lower variance).
-We then fit a GBM to a labelled training set.
-The fitted model is then used to classify draws in an independent test set according to the chain which generated them.
-For each replicate, we then calculated $R^*$ as described in Algorithm \ref{alg:R_star}.
-
-In Fig.
-\ref{fig:ar1}A, we show how a GBM fitted to one such replicate dataset classifies observations according to a draw's value.
-Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.
-
-In Fig.
-\ref{fig:ar1}B, we show that $R^*>1$ for all replicates, indicating that the chains had not converged in all cases.
-In Fig.
-\ref{fig:ar1}C, we show rank-normalised split-$\widehat{R}$ for each replicate;  as for $R^*$, this metric indicates the chains had not converged in all replicates because $\widehat{R}>1.01$.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{ar1.pdf}}
-	\caption{\textbf{Autoregressive example.} A shows how the GBM's classifications vary according to the draw's value for an example model fit; B shows $R^*$ values generated by Algorithm \ref{alg:R_star} across 1000 replicate datasets; C shows corresponding rank-normalised split-$\widehat{R}$ values for each of the 1000 replicates; and D shows 1000 $R^*$ samples as generated by Algorithm \ref{alg:R_star_uncertainty} for two series: the ``unmixed'' dataset being the same as used for figures A-C; the ``mixed'' where all chains have the same distribution as described in \S\ref{sec:heterogeneity}.
-Note that, in D, only a single series of each series type is used to generate distribution.
-All examples used a GBM for classification using the default hyperparameter values given in \S\ref{sec:method}.}
-	\label{fig:ar1}
-\end{figure}
-
-\subsubsection{Stochasticity in $R^*$}\label{sec:heterogeneity_stochasticity}
-Unlike $\widehat{R}$, $R^*$ is a stochastic convergence measure due to randomness in creating training and testing sets (essentially a form of sampling variation) and randomness in the methods used to train the ML model.
-This means that even if the same sample is used, $R^*$ will return a different value each time it is calculated if the pseudorandom seed is not fixed.
-To probe the extent of this randomness, we generated data using the same process as in \S\ref{sec:heterogeneity} but now using varying sample sizes, including samples consisting of 500, 1000, 2000, 4000 and 8000 draws.
-For each dataset, we computed $R^*$ on it 100 times, allowing the pseudorandom seed to vary between calculations.
-We stress that, for each sample size, we used the same dataset (so there were 5 datasets created in total -- one for each sample size), so stochasticity comes from $R^*$ calculation, not that from the data generating process.
-
-In Fig.
-\ref{fig:ar1_samplesize}, we show the results of this study.
-In this figure, the horizontal axis shows the sample size, and the vertical axis, the value of $R^*$ in each repetition.
-This shows that as the number of samples increased, variation in $R^*$ declined.
-At a sample size of 500, there were four cases where $R^*<1$; in larger samples, there were none.
-Intuitively, the reduction in sampling variation when composing training and test sets from larger samples results in lower variability in ML model predictions.
-We also expect that larger sample sizes should lead to higher $R^*$ values with lower variance, since more training data leads to ML models with lower generalisation error.
-We may start to see this here, since the median $R^*=1.25$ at a sample size of 8000 was greater than for the smaller samples.
-
-If randomness in $R^*$ calculation leads to different conclusions about convergence being drawn, this would be problematic.
-One potential remedy for this is to repeatedly calculate $R^*$ on a given sample, much as we have done here, and consider the distribution of $R^*$ values computed.
-The computational cost of doing this may, of course, be unreasonable.
-Instead, in \S\ref{sec:heterogeneity_uncertainty}, we consider an alternative approach based on bootstrapping a single ML model's predictions.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{ar1_samplesize.pdf}}
-	\caption{\textbf{Autoregressive example: $R^*$ stochasticity.} The horizontal axis shows sample size; the vertical axis shows the value of $R^*$ calculated as per Algorithm \ref{alg:R_star} applied to chains split into two halves.
-Grey points show the value of $R^*$ for each replicate (jitter was added to point positions).
-Black points show the median $R^*$ value; upper and lower whiskers show 2.5\% and 97.5\% quantiles.
-For each sample size, a single dataset was created and used for all $R^*$ calculations.}
-	\label{fig:ar1_samplesize}
-\end{figure}
-
-\subsubsection{Uncertainty distribution for $R^*$}\label{sec:heterogeneity_uncertainty}
-GBMs return a probability simplex for each draw, indicating the probability that the draw was generated by a given chain.
-We can use this simplex to generate a measure of uncertainty in $R^*$ as detailed in Algorithm \ref{alg:R_star_uncertainty}.
-We demonstrate this idea using two datasets: one generated as described in \S\ref{sec:heterogeneity}, where one chain (out of four) has a lower variance than the others (we call this the ``unmixed'' data); and another, where all chains sample from the same distribution (we call this the ``mixed'' data).
-In Fig.
-\ref{fig:ar1}D, we show the $R^*$ distributions in each case.
-For the unmixed data, the distribution has its bulk of mass away from 1, indicating lack of convergence.
-For the mixed data, the distribution is centred on 1, indicating convergence.
-In the mixed case, there are many draws where $R^*<1$: these indicate that, in that particular draw from the probability simplex, chain classification is actually worse than selecting a chain identification uniformly at random.
-Much like how it is possible for $\hat{R}<1$, this is a sample property, driven by the sampling distribution of the categorical distribution defined by the probability simplex.
-
-It is worth emphasising that the uncertainty distribution obtained by Algorithm \ref{alg:R_star_uncertainty} differs from that obtained from repeatedly calculating $R^*$ via Algorithm \ref{alg:R_star} as was done in \S\ref{sec:heterogeneity_stochasticity}.
-In Algorithm \ref{alg:R_star_uncertainty}, variation in $R^*$ comes from sampling from the probability simplex: if predicted chain probabilities are close to uniform, there will be greater uncertainty in $R^*$.
-Repeatedly calculating $R^*$ by applying Algorithm \ref{alg:R_star} to the same dataset yields a distribution whose width derives from sampling variation when forming training and testing sets and the stochasticity in training ML models.
-Collectively, these differences mean that the two measures of uncertainty will differ.
-
-There is an additional difference, though, in the central points of each distribution: the distribution obtained by Algorithm \ref{alg:R_star_uncertainty} will, in general, have a lower mean than that obtained by repeated application of Algorithm \ref{alg:R_star}.
-To see this, note that the darker-shaded $R^*$ distribution in Figure \ref{fig:ar1}D was generated via Algorithm \ref{alg:R_star_uncertainty} and has a mean around 1.07; the distribution shown in Figure \ref{fig:ar1}B was generated by repeatedly recomputing $R^*$ using Algorithm \ref{alg:R_star} and has a mean closer to 1.22.
-This difference in mean is expected since predictive performance when assigning chain identities stochastically when sampling from the categorical distribution of the probability simplex (as is done in Algorithm \ref{alg:R_star_uncertainty}) will generally result in worse prediction than when assigning each chain identity using whichever chain has the highest class probability (as is done in Algorithm \ref{alg:R_star}).
-Of course, we would prefer it if the uncertainty distribution generated by Algorithm \ref{alg:R_star_uncertainty} had a mean closer to the one obtained by repeated application of Algorithm \ref{alg:R_star}.
-Nonetheless, in practice, we have found that the mean of the $R^*$ distribution generated by Algorithm \ref{alg:R_star_uncertainty} provides a useful cheaper diagnostic.
-
-\subsubsection{Sensitivity to number of chains}\label{sec:heterogeneity_numchains}
-We have so far focused on the sensitivity of $R^*$ to chain heterogeneity with a fixed number of chains: four.
-Since classification may become a harder problem when there are more categories, we now demonstrate how $R^*$ (as calculated by Algorithm \ref{alg:R_star}) performs across various number of chains.
-For comparison, we also illustrate how the performance of rank-normalised split-$\widehat{R}$ varies with number of chains.
-To do so, we consider an autoregressive example similar to that described in \S\ref{sec:heterogeneity}: where all chains bar one have $\sigma=1$, and the remaining chain has $\sigma=1/3$.
-We consider cases with 2, 4, 8, 16, and 32 chains.
-The other hyperparameters of the data generating process remain the same as in \S\ref{sec:heterogeneity}.
-
-In Fig.
-\ref{fig:ar1_numchains}, we show the results of these simulations with 50 replicates at each number of chains.
-On the horizontal axis, we show the number of chains, and on the vertical axis, the value of each of the two convergence measures ($R^*$ in the left panel; $\widehat{R}$ in the right panel).
-In general, both measures decline with chain count.
-For $R^*$, this may be because it is harder to classify chains when there are more of them.
-For $\widehat{R}$, this is because between-chain variance becomes relatively lower to that within them when there are more chains and only one of them differs in its marginal distribution.
-Across the replicates we ran, median $\widehat{R}<1.01$ for 16 or more chains; a minimum of $R^*=1.10$ was obtained for 32 chains.
-
-The decline of both of these measures when more chains are used hints that perhaps a moving threshold for diagnosing convergence may be pertinent to avoid neglecting those minority of chains with differing information.
-Here, however, we do not make suggestions on what such guidelines could be and leave this for later work.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.8\textwidth]{ar1_numchains.pdf}}
-	\caption{\textbf{Autoregressive example: sensitivity to number of chains.} The horizontal axis shows the number of chains used in the data generating process described in \S\ref{sec:heterogeneity_numchains}.
-The vertical axis shows the value of $R^*$ as calculated by Algorithm \ref{alg:R_star} (left panel) on chains split into two halves, and rank-normalised split-$\widehat{R}$ (right panel).
-Grey points indicate the values of both convergence measures calculated for each replicate; horizontal jitter has been added to points.
-The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles across 50 replicates at each number of chains.}
-	\label{fig:ar1_numchains}
-\end{figure}
-
-\color{black}
-\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
-In this section, we illustrate how $R^*$ can diagnose convergence issues in the joint target distribution.
-
-\subsubsection{Bivariate model}\label{sec:multivariate_normal_bivariate}
-First, we consider a bivariate normal density.
-In all four chains, we use independent sampling to generate 2000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions.
-By construction, all chains target the same marginal distribution in each dimension, but the fourth chain has a different joint distribution.
-
-First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures that aim to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail-ESS.
-In all cases, the various quantities were calculated based on chains split into halves.
-For both dimensions, the two ESS measures were above 7000, and $\widehat{R}<1.001$, indicating no issues with convergence.
-
-Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty} using both GBM and RF classifiers.
-These distributions are shown in Fig.
-\ref{fig:bivariate}.
-The mean of the GBM-$R^*$ distribution is 1.14, and $>$99\% of $R^*$ draws are above 1; the mean of the RF-$R^*$ distribution was 1.27 and all draws were above 1.
-Collectively, these measures indicate that the sampling distribution has not converged.
-By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only marginals.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.6\textwidth]{bivariate.pdf}}
-	\caption{\textbf{Bivariate normal example.} The distribution for $R^*$ across 1000 draws as calculated using Algorithm \ref{alg:R_star_uncertainty} for both the GBM and RF classifiers.}
-	\label{fig:bivariate}
-\end{figure}
-
-\subsubsection{250-dimensional model}\label{sec:multivariate_normal_250}
-We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}.
-We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions.
-We use Stan's NUTS algorithm \citep{betancourt2017conceptual} to sample from this target distribution and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5).
-First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
-%
-\begin{equation}\label{eq:mvt_normal_250}
-\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
-\end{equation}
-%
-where $\boldsymbol{x}\in\mathbb{R}^{250}$.
-For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} with a GBM classifier to generate an uncertainty distribution for $R^*$, which is shown in Fig.
-\ref{fig:mvt}A (the equivalent plot for a RF classifier is similar and shown in Fig.
-\ref{fig:mvt_gbm_vs_rf}).
-From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution.
-Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for all parameters (Fig.
-\ref{fig:mvt}B), although 54\% had bulk-$\text{ESS}<400$ and 13\% of parameters had tail-$\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.
-
-Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it.
-This model has the form,
-%
-\begin{align}
-\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
-\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
-z_j\sim \text{normal}(0, 1), \text{ for } j = 1,2,...,250.
-\end{align}
-%
-where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$.
-Fig.
-\ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution has mean $R^*=1.00$.
-Fig.
-\ref{fig:mvt}B shows the $\widehat{R}$ values for each $x$ parameter in this model, and, echoing the result for $R^*$, $\widehat{R}<1.01$ in all cases; further, bulk- and tail-$\text{ESS}>400$ for all parameters.
-
-\subsubsection{Variable importance}\label{sec:multivariate_normal_varimportance}
-In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), which allows us to determine which variables were most informative for predictions.
-We now compare these with the more established metrics $\widehat{R}$ and ESS.
-For a GBM fitted to the centered model of eq.
-\eqref{eq:mvt_normal_250} with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig.
-\ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle).
-In this plot, there is a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.17, S=2185680, p<0.01$).
-In Fig.
-\ref{fig:mvt}D, we plot variable importance versus two measures: bulk-ESS and tail-ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk-ESS: $\rho=-0.57, S=4142470, p<0.01$; tail-ESS: $\rho=-0.56, S=4113709, p<0.01$).
-Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1\textwidth]{mvt_three.pdf}}
-	\caption{\textbf{Multivariate normal example with 250 dimensions.} A shows $R^*$ distributions obtained for two MCMC samples (of differing numbers of draws: 400 and 10,000) from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''; with 10,000 draws); B shows the rank-normalised split-$\widehat{R}$ values for all parameters from the same MCMC runs as in A; C shows variable importance versus $\widehat{R}$ for each parameter; and D shows variable importance versus bulk- and tail-ESS as calculated by \cite{vehtari2019rank}.
-In A, 1000 $R^*$ draws by Algorithm \ref{alg:R_star_uncertainty} are shown for each MCMC run.
-In plots C and D, horizontal jitter was added to the points and a loess fit line with standard errors overlaid.}
-	\label{fig:mvt}
-\end{figure}
-
-\subsection{Infinite variance: Cauchy example}\label{sec:cauchy}
-We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance.
-Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-dimensional vector $x$,
-%
-\begin{equation}
-x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50.
-\end{equation}
-%
-We call this parameterisation the ``nominal'' version of this model.
-
-In addition, we also use Stan to sample from an ``alternative'' parameterisation of the Cauchy, based on a scale mixture of Gaussians \citep{vehtari2019rank},
-%
-\begin{align}
-a_j \sim  \text{normal}(0,1), \qquad
-b_j \sim  \text{Gamma}(0.5, 0.5), \qquad
-x_j =  a_j/\sqrt{b_j}.
-\end{align}
-%
-The distribution of the $x$ vector is the same under both parameterisations, although the thin-tailed $(a,b)$ vectors define a higher dimensional posterior that improves sampling efficiency.
-
-In the top-left and top-middle panel of Fig.
-\ref{fig:cauchy}, we show the $R^*$ distribution for GBM and RF classifiers under both parameterisations.
-As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 1000 MCMC post-warm-up iterations (with 1000 warm-up iterations discarded) across each of 4 chains, draws still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$.
-The alternative parameterisation fares better, and the $R^*$ distribution is nearer $R^*=1$, yet its mean remains above this value.
-In the top-right panel of Fig.
-\ref{fig:cauchy}, we show the rank-normalised split-$\widehat{R}$ values across each of the 50 parameters for the same MCMC runs.
-The nominal parameterisation has some parameters with $\widehat{R}>1.01$, indicative non-convergence, whereas the alternative has $\widehat{R}<1.01$ for all parameters.
-
-Since the $R^*$ distribution indicated non-convergence for both parameterisations, we ran each model for sixty-times as long, although thinned by a factor of 3, resulting in 10,000 post-warm-up iterations across each of 4 chains.
-In the bottom row of Fig.
-\ref{fig:cauchy}, we show the results for these longer runs.
-In these, the alternative parameterisation now has an $R^*$ distribution centred on $R^*=1$ for the GBM classifier although the RF classifier $R^*$ distribution remains slightly rightwards of this target indicating that convergence is nearer but more iterations are likely still required.
-Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.
-The $\widehat{R}$ values are all below 1.01, indicating convergence in both cases.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{cauchy.pdf}}
-	\caption{\textbf{Cauchy example.} Rows show convergence results for MCMC runs with 1000 (top) and 10,000 (bottom; obtained by thinning iterations by a factor of 3) post-warm-up iterations (each with half iterations discarded as warm-up) for each of 4 chains.
-Columns show the $R^*$ distributions from GBM (left) and RF (middle) and rank-normalised split-$\widehat{R}$ values across all parameters (right).
-Shadings indicate different model paramerisations as indicated in legend.}
-	\label{fig:cauchy}
-\end{figure}
-
-\subsubsection{Measuring convergence objectively}\label{sec:cauchy_objective}
-To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target.
-One measure of distributional ``closeness'' is the KL-divergence, which, in this case, could be used to measure the divergence from target to sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated for each dimension.
-The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails.
-This means that approximate KL-divergence is unreliable for these types of model.
-We decided not to use the Kolmogorov-Smirnov (KS) test, since it is most sensitive to differences between distributions around the median, whereas, here, we are interested in behaviour in the tails.
-Additionally, we found that the Anderson-Darling and Cram\'er-Von Mises tests \citep{faraway2019goftest}, which do not suffer the same shortcomings as the KS, behaved equally erratically and provided measures that were hard to intuit.
-The Wasserstein distance was also trialled but had great uncertainty due to the long-tails of the Cauchy.
-Instead, we chose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents.
-Specifically, we calculate the $R^2$ for the linear regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities.
-In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.
-
-In Fig.
-\ref{fig:cauchy_convergence}A, we plot this \textit{quantile-$R^2$} as a function of MCMC sample size for both parameterisations of the Cauchy model.
-This shows that after c.10,000 iterations, the alternative parameterisation approaches $R^2\approx 1$; at the same number of iterations, the nominal parameterisation still provides a poor measure of tail quantiles.
-Next, in Figs.
-\ref{fig:cauchy_convergence}B\&C, we plot two measures of $\widehat{R}$, each calculated from splitting the 4 original chains into two equal halves.
-The first of these measures is the rank-normalised $\widehat{R}$ \citep{vehtari2019rank}, which provides a separate measurement for each target dimension; in Fig.
-\ref{fig:cauchy_convergence}B, we show how the maximum of this measurement across all 50 dimensions changes with sample size.
-After c.500 iterations, the alternative parameterisation achieves $\widehat{R}<1.01$ for all target dimensions, and, after c.10,000 iterations, the nominal model achieves the same maximum $\widehat{R}$ value: in both cases, these suggest convergence.
-The second measure is multivariate $\widehat{R}$ \citep{brooks1998general}, which, like $R^*$, yields a single measurement across all dimensions; Fig.
-\ref{fig:cauchy_convergence}C shows how this metric changes with sample size for both Cauchy model parameterisations.
-After c.1800 iterations, multivariate $\widehat{R}<1.01$ for the alternative parameterisation, whilst after 25,000 iterations multivariate $\widehat{R}>1.07$ indicating more draws are needed.
-In Fig.
-\ref{fig:cauchy_convergence}D, we plot $R^*$ against iteration for both models and for both GBM and RF classifiers: these indicate that, after 25,000 iterations, for the alternative model, $R^*\approx 1.05$ for the GBM classifier and $R^*\approx 1.74$ for the RF classifier; for the nominal model, $R^*>2$ for the GBM classifier $R^*>3$ for the RF classifier: all these $R^*$ values suggest lack of convergence.
-Finally, in Figs.
-\ref{fig:cauchy_convergence}E\&F, we plot the minimum across all the dimensions of tail- and bulk-ESS calculated as described in \cite{vehtari2019rank}.
-After c.180 iterations, the alternative parameterisation surpassed a tail-ESS of 400; after c.18,700, the nominal parameterisation did the same.
-Both models were quicker to pass 400 bulk-ESSs.
-
-Comparing our measure of convergence that requires knowing the actual target distribution (\textit{quantile-$R^2$}; in Fig.
-\ref{fig:cauchy_convergence}A), with the various heuristic measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence.
-The rate at which these converge differs though, and $R^*$ (Fig.
-\ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to \textit{quantile-$R^2$}.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{cauchy_convergence.pdf}}
-	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\widehat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\widehat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star} for both the GBM and RF classifiers; and E and F show tail- and bulk-ESS.
-Dashed lines indicate recommended thresholds for each convergence statistic.}
-	\label{fig:cauchy_convergence}
-\end{figure}
-
-\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
-We now examine a classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
-
-The model can be parameterised in two ways, as described in \cite{vehtari2019rank} (and introduced in \cite{van2001art}).
-The simplest way is referred to as the ``centered'' parameterisation and exactly mirrors the underlying statistical model,
-%
-\begin{align*}
-\theta_j &\sim \text{normal}(\mu, \tau), \\
-y_j &\sim \text{normal}(\theta_j, \sigma_j).
-\end{align*}
-%
-The ``non-centered'' parameterisation (first introduced in \cite{van2001art}) recodes this model in a way that does not affect the joint distribution of $(\theta, \mu, \tau, \sigma)$ but makes it easier to sample from it, by introducing auxillary variables, $\tilde \theta_j$.
-This can be written as,
-%
-\begin{align*}
-\tilde{\theta}_j &\sim \text{normal}(0, 1), \\
-\theta_j &= \mu + \tau \tilde{\theta}_j,\\
-y_j &\sim \text{normal}(\theta_j, \sigma_j).
-\end{align*}
-%
-In both cases, $\theta_j$ are the treatment effects in the eight schools, and $(\mu, \tau)$ represent the population mean and standard deviation of the distribution of these effects.
-In the centered parameterization, the $\theta_j$ are parameters, whereas in the non-centered parameterization, the $\tilde{\theta}_j$ are parameters and $\theta_j$ is a derived quantity.
-
-We first used Stan \citep{carpenter2017stan} to sample from the centered model using 4 chains.
-Like \cite{vehtari2019rank}, we used settings that reduce the chance of divergent iterations for the dynamic HMC algorithm \citep{betancourt2017conceptual} (called using the ``NUTS'' option in Stan), meaning that the resultant sampling distribution is likely to be biased.
-We also used the same algorithm settings to sample from the non-centered model.
-
-To see how $R^*$ performed on this example, we first split each of the (post-warm-up) chains in two, as is done by default in Stan \citep{carpenter2017stan} and in \cite{vehtari2019rank}, resulting in 500 iterations across 8 chains.
-Following the same approach as in Algorithm \ref{alg:R_star_uncertainty}, we generated $R^*$ distributions for both the centered and non-centered models using a GBM classifier.
-The resultant distributions for $R^*$ are shown in Fig.\ref{fig:eight_schools}A.
-In this plot, the centered model is close to convergence, whereas the non-centered is not.
-
-In addition, to illustrate the power of $R^*$, we also repeat the analysis but, this time, do not split the chains in two.
-The results are shown in Fig.\ref{fig:eight_schools}B.
-In this case, because the unsplit chains do not mix with themselves, it is harder to accurately predict the chain that generated each draw, meaning that the centered model $R^*$ values are shifted leftwards.
-Despite this, however, the centered model distribution for $R^*$ still does not strongly overlap with $R^*=1$, indicating that the model has not converged, contrasting with the non-centered model which appears near convergence.
-
-Fig.
-\ref{fig:eight_schools_rf} shows the equivalent of Fig.
-\ref{fig:eight_schools} except using a RF classifier.
-The results are similar, although the $R^*$ distributions are shifted slightly rightwards: indicating, for example, that more than 2000 draws from the non-centered model may be required for convergence.
-
-It is recommended that $\widehat{R}$, like $R^*$, be calculated using split chains.
-In Fig.
-\ref{fig:eight_schools}C, we plot $\widehat{R}$ values obtained when using the original 4 chains (horizontal axis) versus those when using the split chains (vertical axis) for the ten parameters in this model; we do this for both the centered and non-centered models.
-These show that the values of $\widehat{R}$ for the centered model using the unsplit chains were below 1.01; when using the chains split into two halves, $\widehat{R}>1.01$ for all but a single parameter.
-All parameters for the non-centered models were below 1.01, indicating convergence.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools.pdf}}
-	\caption{\textbf{Eight schools example: $R^*$ distributions.} A shows draws from the $R^*$ distribution when splitting chains in two (resulting in 8 chains); B shows the same but using the 4 original chains; C shows rank-normalised $\widehat{R}$ for original 4 chains versus those for the 8 chains case for all ten parameters defined by the centered model -- in this case, we plot horizontal and vertical dashed lines to illustrate the $\widehat{R}=1.01$ cutoff and a $y=x$ line.
-The legend inset in panel C provides a key for all panels.
-The MCMC samples comprised 2000 draws in all cases, with 1000 used as post-warm-up iterations.
-In panels A and B, the plots show 1000 $R^*$ draws using Algorithm \ref{alg:R_star_uncertainty} using a GBM classifier for each parameterisation.}
-	\label{fig:eight_schools}
-\end{figure}
-
-\subsubsection{Understanding chain classification}
-To probe the predictive power of the ML classifier, we investigated how predictive accuracy varies across parameter space.
-After fitting the GBM model, we group MCMC draws in the test set into deciles and draw from the $R^*$ distribution for each decile.
-In Fig.
-\ref{fig:eight_schools_r_star_quantiles}, we show the results of this exercise for (A) $\mu$ and (B) $\tau$.
-In the left-hand column of this figure, we show the path of four MCMC chains (here we did not split chains when calculating $R^*$ to simplify visualisations) across the quantiles of each parameter space.
-To the right of each trace plot, we show the marginal distributions for each chain.
-In the right-hand column, we show 40 $R^*$ draws for each decile, which were generated according to Algorithm \ref{alg:R_star_uncertainty} using a GBM fit to all draws.
-In essence, the left-hand panels explain the variation in $R^*$ in the right-hand panels: if chains become stuck in regions of parameter space, this causes differences between the marginal distributions of the chains; these differences, in turn, allow a ML model to predict the generative chain in those same sticky regions.
-For example, for $\mu$, the purple chain became stuck around the middle quantile, forcing a difference in its marginal distribution in that region, which resulted in $R^*>1$ for the corresponding decile.
-Similarly, for $\tau$, the purple chain became stuck in the lowest quantiles, elevating its marginal distribution there and resulting in improved predictive accuracy.
-
-Fig.
-\ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in classification accuracy and falsely concluding that convergence has occurred.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=1.0\textwidth]{eight_schools_r_star_quantiles.pdf}}
-	\caption{\textbf{Eight schools example: quantile $R^*$ plots.} Row A shows plots for $\mu$; Row B for $\tau$.
-In each row, we show the path of the four individual chains above and 40 $R^*$ draws obtained using Algorithm \ref{alg:R_star_uncertainty} for each parameter quantile below.
-To the right of each trace plot, we show the marginal distribution of each chain estimated via kernel density estimation using Gaussian kernels.
-Note that, in the right-hand plots, jitter has been added to the data points.}
-	\label{fig:eight_schools_r_star_quantiles}
-\end{figure}
-
-\subsection{Further experiments}\label{sec:further_experiments}
-Alongside the examples included in the main text, there are a number of supplementary text examples, which we briefly outline here.
-
-In \S\ref{sec:wide}, we illustrate how $R^*$ can provide a reasonable measure of convergence when the number of dimensions of a distribution is comparable to the number of draws.
-Specifically, this was to test that classification didn't become prone to overfitting in this limit.
-To test this hypothesis, we investigated two scenarios using a multivariate normal target: one with a 250-multivariate normal with high correlation between dimensions using 250 post-warm-up iterations; and another normal with 10,000 independent dimensions using up to 500 post-warm-up iterations.
-In both cases, sampling was done using Stan's NUTS algorithm.
-In both cases, $R^*$ and rank-normalised split-$\widehat{R}$ reached similar conclusions about convergence: namely, that more iterations were needed in all experiments considered.
-Overall, these experiments show that $R^*$ is a conservative convergence measure that will tend to diagnose unconvergence when there are insufficient draws.
-
-In \S\ref{sec:non-stationary}, we illustrate the importance of splitting chains before calculating $R^*$ to ensure poor within-chain convergence is diagnosed.
-We illustrate this via four examples: (a) sampling from a univariate normal and adding a linear trend over sampling time, to ensure that the sampling distributions were non-stationary; (b), similar to (a) but across a range of target distribution dimensions where only a single dimension had a non-stationary mean; (c), a bivariate normal with a non-stationary covariance; and (d), an autocorrelated sampling distribution with a univariate normal target with a range of different autocorrelations.
-The results of (a) echoed those presented in \cite{vehtari2019rank} for $\widehat{R}$ and showed that $R^*$ is insensitive to sampling non-convergence if it occurs within chains; splitting chains into two halves alleviates this issue.
-The results of (b) show that $R^*$ calculated on split chains is able to diagnose non-stationarity in mean in a single dimension in a way that did not diminish as the numbers of dimensions considered increased.
-Example (c) showed that split-$R^*$ opposed to split-$\widehat{R}$ is able to diagnose non-stationary covariance between dimensions of a target distribution.
-In (d), we show that $R^*$ is able to differentiate between distributions with non-stationary target distributions and stationary ones.
-It also shows that $R^*$ still functions reasonably at higher levels of chain persistence: yielding a conservative convergence measure when there are insufficient draws.
-
-In \S\ref{sec:prostate}, we show that $R^*$ performs well for two Bayesian logistic regression problems with highly multimodal posteriors.
-Each of these models have 1000s of parameters, and we found that it was slow to compute both $\widehat{R}$ and $R^*$ for them.
-That said, the computational time for calculating $\widehat{R}$ was considerably less than was needed for $R^*$.
-
-In \S\ref{sec:discrete}, we evaluate $R^*$ on univariate discrete examples: one with four states (in \S\ref{sec:discrete_small}); another, with a larger state-space consisting of 20 states (in \S\ref{sec:discrete_large}).
-In these examples, we use a discrete Markov model to generate draws from a given target.
-The small and larger state-space cases, show that $R^*$ behaves as expected: given a sufficient sample size, it is able to detect differences in the transition probability matrix between chains that result in differences in the target distribution.
-
-In \S\ref{sec:ML_sensitivity}, we investigate how two decisions about classifiers — which classifier to use (in \S\ref{sec:ml_model}) and what hyperparameters to use for it (in \S\ref{sec:hyperparameters}) — affect calculation of $R^*$.
-In \S\ref{sec:ml_model}, we test a range of popular classifiers: GBMs, RFs, k-nearest-neighbour models, support vector machines and generalised linear models across examples.
-This indicated that GBMs and RFs consistently had the highest classification accuracy across the examples: in higher dimensional problems, RFs tended to best GBMs.
-In \S\ref{sec:hyperparameters}, we show how these two best classifiers — GBMs and RFs — depend on their hyperparameters.
-Across the examples we test, GBMs are more sensitive to hyperparameter choices than are RFs.
-Additionally, GBMs require typically more rounds of boosting in higher dimensional problems, leading to more extensive algorithm runtime.
-Because of this, we suggest fixing GBM's hyperparameters to values that result in reasonable performance and reasonable runtime.
-For RFs, we suggest using a heuristic for choosing its hyperparameters that was derived in a previous empirical evaluation of RFs \citep{bernard2009influence}.
-
-In \S\ref{sec:comparison_gbm_rf}, we use a slew of examples to compare $R^*$ calculated using GBM and RF classifiers using our suggested default hyperparameter sets (given in \S\ref{sec:method}).
-In \S\ref{sec:joint_distribution}, we compare how both methods are able to diagnose lack of convergence in a joint distribution (using a multivariate normal target).
-In \S\ref{sec:tail_fatness}, we compare the ability of both approaches to diagnose differences in the tails of the marginal distributions between chains (using Student-t targets).
-In both examples, draws are generated by independent sampling meaning that an optimal $R^*$ can be calculated based on the Bayes optimal classifier (see, for example, \citep{devroye2013probabilistic} and \S\ref{sec:comparison_gbm_rf} for more information).
-Collectively, the results of \S\ref{sec:joint_distribution} and \S\ref{sec:tail_fatness} suggest that both GBM and RF classifiers can detect differences in sampling distributions, should they exist, between chains.
-The classification rates achieved by these two approaches exhibited similar trends to the optimal classifier, albeit with lower predictive accuracy.
-In higher dimensions, it is likely that the difference between optimal classification rates and those from the GBM or RF will increase: particularly, when searching for between-chain differences in tail fatness.
-These results also suggest a different region of optimality for each classifier: GBMs tend to perform best for low dimensional targets and RFs for moderate-high ones.
-This is a function of the different rules used to set the hyperparameters of each classifier (see \S\ref{sec:method} and \S\ref{sec:hyperparameters}): for GBMs to perform well in higher dimensions, they need more rounds of tree boosting which substantially increases training time.
-Because of this, we fixed the hyperparameters of the GBM to values that yield reasonable computation time.
-RFs are less sensitive to hyperparameter variation (\S\ref{sec:hyperparameters}), and useful heuristics for adapting these with target dimensions are known, which we follow here, as described in \S\ref{sec:method}.
-Despite this dynamic choice of hyperparameters for RFs, its runtime remained reasonable over the range of examples we test in this paper.
-
-\section{Discussion}
-If an MCMC sampler has converged on the target distribution, the chains must be well-``mixed'', that is, given a draw, it should be impossible to discern which chain generated it.
-Based on this observation, we used supervised machine learning (ML) classifiers to quantify the information about the generative chain identity contained in draws.
-By taking the ratio of model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic.
-By extracting classifier-predicted chain probabilities from each prediction in the test set, we can additionally generate an uncertainty distribution for $R^*$.
-Across a range of previously published examples, $R^*$ was shown to be predictive of whether chains had converged.
-
-The predominant methods for diagnosing MCMC convergence rely heavily on looking for between-chain differences in the marginal distributions along each dimension of the target.
-$R^*$ naturally includes this information in building a model capable of predicting the chain that generated each draw.
-It also naturally includes information about the joint distribution across all dimensions of the target.
-Since converged chains should have similar joint distributions (implying similar marginals), any measure of convergence should account for both of these aspects.
-Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate convergence whereas $R^*$ shows otherwise.
-This indicates the complementarity of $R^*$ to existing measures.
-
-Different target distributions present different challenges to sampling.
-Because of this, there is not a unique optimal ML classifier across all cases: this is just a manifestation of the no free lunch theorem \citep{wolpert1997no}.
-Across a range of examples we tested (see \S\ref{sec:ml_model}), GBM and RF classifiers performed consistently well, and we then used these across all other illustrative examples.
-It is possible — indeed, likely — that another ML model may exist or be invented that consistently outperforms both these classifiers.
-We do not see this as a problem: across the examples we considered, $R^*$ calculated using both classifiers tended to provide a measure of convergence as or more stringent than existing diagnostics.
-In that sense, it is a step in the right direction.
-If a better classifier is found, the same apparatus we develop here can be used, and this will present a harsher test of convergence.
-
-A different question is, ``When is such a measure of convergence of practical use?''.
-In our examples, $R^*$ is able to diagnose poor convergence in the tails of marginal distributions: likely of practical relevance for many applications that require tail quantiles.
-$R^*$ is also able to diagnose lack of convergence in the joint distribution — even if the marginals appear converged.
-Indeed, it is less clear how the joint distribution can be unconverged when the marginals appear so, but we found examples where this appeared true.
-A consequence of poor convergence of the joint distribution would be for prediction and, by corollary, model comparison.
-Further work examining these consequences further is needed, and $R^*$ can help to identify and monitor fruitful candidate systems.
-
-In \S\ref{sec:prostate}, we fit Bayesian models with many 1000s of parameters then used $R^*$ to diagnose convergence, finding that $R^*$ was considerably more expensive to calculate than $\widehat{R}$.
-The time complexity of training RFs is thought to be $\mathcal{O}(m_{\text{try}} n_{\text{tree}} N \text{log} N)$ \cite[chapter~5]{louppe2014understanding}, and given the similarities with GBMs (which also builds many decision trees), it is likely to be similar.
-If so, this suggests that larger statistical models (usually needing more MCMC iterations) may currently be beyond the reach of $R^*$.
-That said, it is possible to reduce the runtime for $R^*$ using thinned draws (although this risks losing chain idiosyncracies) and using a subset of dimensions (although this risks losing problematic dimensions).
-Indeed, in \S\ref{sec:multivariate_normal_250}, \S\ref{sec:cauchy} and \S\ref{sec:prostate}, we use these strategies and, nonetheless, find that $R^*$ provides a stringent measure of convergence.
-
-Many implementations of $\widehat{R}$ suggest splitting chains in two before calculating it.
-In a number of examples, we trial this before calculating $R^*$ and find that this approach leads to more accurate chain prediction.
-We recommend that this practice be adopted whenever $R^*$ is calculated to ensure that this measure is maximised.
-Additionally, our non-parametric calculation method for $R^*$ makes it possible to include any covariates which may be useful features for prediction, such as an ``iteration block'' indicator variable taking values $1, 2, ..., K$ in each of $K$ blocks of contiguous iterations.
-If each chain is thoroughly mixed with itself, including this additional information shouldn't change $R^*$; by contrast, if the chains are random walk-like, this information should boost $R^*$.
-
-MCMC enables inference across a wide range of models encountered across the social, biological and physical sciences.
-Its ease of implementation, however, masks important underlying fragilities in the method.
-Namely, that unless the chains have converged to a truly stationary distribution, the draws generated are not faithful depictions of the posterior.
-In this paper, we introduce a new metric, $R^*$, that is especially good at diagnosing poor convergence in the joint sampling distribution -- an area that has received insufficient attention thus far.
-$R^*$ can straightforwardly be introduced into existing MCMC libraries and could provide a measure of convergence complementary to existing metrics.
-
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	%% Supplementary Material, if any, should   %%
-	%% be provided in {supplement} environment  %%
-	%% with title and short description.        %%
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-	\begin{supplement}
-		\stitle{Supplementary materials: $R^*$ convergence diagnostic}
-		\sdescription{Further experiments using $R^*$}
 
@@
 %
-		where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$.
-The target distribution is, hence, the stationary distribution of this process.
+where $p_{ij}$ is the probability of transitioning from state $i$ at time $n$ to state $j$ at time $n+1$, and $X(0)=1$. The target distribution is, hence, the stationary distribution of this process.
 
